### 2016-09-13 paper.tex
initial draft

\documentclass[10pt]{sigplanconf}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{listings}
\usepackage[noline]{algorithm2e}
\usepackage{comment}
\usepackage{color}
\usepackage{times}
\usepackage{multirow}
\usepackage{balance}
\usepackage{adjustbox}
\usepackage{alltt}
\usepackage{textcomp}
\usepackage[usenames]{xcolor}

\setcounter{tocdepth}{3}

% configures style of source code listings
\lstset{
	basicstyle=\small\sffamily
	%basicstyle=\scriptsize\ttfamily
	%numbers=left
}

\lstdefinelanguage{java}{
	morekeywords={abstract,continue,for,new,switch,assert,default,goto,package,synchronized,boolean,stream,event,filter,when,do,if,private,this,break,double,implements,protected,throw,byte,else,import,public,throws,case,enum,instanceof,return,transient,catch,extends,int,short,try,char,final,interface,static,void,class,finally,long,strictfp,volatile,const,float,native,super,while},
	sensitive=false,
	morestring=[b]",
	morestring=[b]',
	morecomment=[l]{//},
	morecomment=[s]{/*}{*/)},
	escapeinside=??,
	moredelim=[is][\textit]{__}{__},
	moredelim=[is][\textbf]{_*}{*_}
}
\lstnewenvironment{displayjava}
	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b,xleftmargin=1em,xrightmargin=1em}}{}
\lstnewenvironment{java}
	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b}}{}

\newcommand{\code}[1]{{\sf \small #1}\xspace}

\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\mr}[1]{\mathrm{#1}}
\newcommand{\mt}[1]{\mathtt{#1}}

\newcommand{\todo}[1]{{\bfseries #1}}

% for writing comments
\newcommand{\PP}[1]{\textcolor{blue}{\bfseries PP: #1}} % Pavel Parizek
\newcommand{\PV}[1]{\textcolor{red}{\bfseries PV: #1}} % Premek Vysoky
\newcommand{\JB}[1]{\textcolor{orange}{\bfseries JB: #1}} % JetBrains

% colors for the ANTLR environment
\definecolor{antlrbgcolor}{rgb}{.962, .962, .962}
\definecolor{antlrparserrulecolor}{RGB}{11,83,148}
\definecolor{antlrlexerrulecolor}{RGB}{191,144,0}
\definecolor{antlrliteralcolor}{RGB}{133,32,12}
\definecolor{antlrregexcolor}{RGB}{56,118,29}

% colors for the MPS environment
\definecolor{mpsbgcolor}{rgb}{.962, .962, .962}
\definecolor{mpsconceptcolor}{RGB}{42, 140, 0}
\definecolor{mpsinterfacecolor}{RGB}{170,0,229}

% coloring commands
\newcommand{\antlrap}{\textquotesingle}
\newcommand{\antlrparserrule}[1]{\textcolor{antlrparserrulecolor}{#1}}
\newcommand{\antlrlexerrule}[1]{\textcolor{antlrlexerrulecolor}{#1}}
\newcommand{\antlrliteralnoap}[1]{\textcolor{antlrliteralcolor}{#1}}
\newcommand{\antlrliteral}[1]{\antlrap\antlrliteralnoap{#1}\antlrap}
\newcommand{\antlrregex}[1]{\textcolor{antlrregexcolor}{#1}}
% TODO mozna zmenit "mpsconcept" na "mpslangelem"
\newcommand{\mpsconcept}[1]{\textcolor{mpsconceptcolor}{#1}}
\newcommand{\mpsinterface}[1]{\textcolor{mpsinterfacecolor}{#1}}

% TODO zmenit barvu pozadi na bilou
% environment for the ANTLR code (grammars)
\newenvironment{antlr}
{
	\vspace{3mm}
	\par
	\noindent
	\adjustbox{margin=3mm,bgcolor=antlrbgcolor,frame,minipage=394pt,inner=\linewidth}
	\bgroup
	\alltt
}
{
	\endalltt
	\egroup
	\\
	\vspace{2mm}
}

% environment for the MPS code
\newenvironment{mps}
{
	\vspace{3mm}
	\par
	\noindent
	\adjustbox{margin=3mm,bgcolor=mpsbgcolor,frame,minipage=394pt,inner=\linewidth}
	\bgroup
	\alltt
}
{
	\endalltt
	\egroup
	\\
	\vspace{2mm}
}

\begin{document}

% TODO
%\conferenceinfo{OOPSLA'12,} {October 19--26, 2012, Tuscon, Arizona, USA.}
%\CopyrightYear{2012}
%\copyrightdata{978-1-4503-1561-6/12/10}

% TODO
\title{\todo{INGRID: Creating Languages in MPS by Importing ANTLR Grammars}}
% alternative: "INGRID: Importing Grammars Into MPS"

\authorinfo{P\v{r}emysl Vysok\'y \and Pavel Par\'izek}
	{Charles University, Faculty of Mathematics and Physics, Prague, Czech Republic}
	{premek.vysoky@gmail.com, parizek@d3s.mff.cuni.cz}

% TODO pridat nekoho dalsiho z JetBrains
% TODO upravit affiliation pro JetBrains -> doplnit "s.r.o" nebo radsi "Inc."
% TODO muzeme taky napsat jejich emaily pokud budou lidi z JetBrains souhlasit
\authorinfo{V\'aclav Pech}
	{JetBrains}

\maketitle

\begin{abstract}
% TODO
\todo{write later}

\todo{Copy from master thesis.}
JetBrains MPS is a language workbench focusing on domain-specific languages.
Unlike many other language workbenches and IDEs, it uses a projectional editor for code.
The developer directly manipulates the program in its tree form (AST) and not by editing a text source code.
This brings many advantages, but on the other hand requires time-consuming and complicated MPS language definition.
The thesis elaborates on the possibility of automating the process of creating MPS language definition from its grammar description.
It introduces the MPS editor, evaluates approaches of related projects and describes author's efforts to implement an MPS plugin that allows this import.
The chosen approach and the selection of tools used for implementation are justified in the thesis.
We point out important problems that any similar project might deal with and we introduce some possible solutions.
Furthermore, the thesis contains examples of imported languages, showing the potency of the chosen approach.
The thesis also aims to lay groundwork for future extensions and suggest possible improvements.
\end{abstract}

% TODO possible keywords: JetBrains MPS, programming language, grammar, projectional editor

\section{Introduction}

Prevailing approach to the definition of valid syntax of programming languages is through/using grammars written, for example, using notations based on (or similar to) EBNF [reference]. Many popular tools exist for automated generation of parsers from the grammars, such as ANTLR and Bison/Flex. Source code of programs in the language is then written in a text editor, which is a part of some IDE.

An alternative emerged recently in the form of projectional editors that are used mainly in special IDEs for domain specific languages, which are called language workbenches. A popular tool of this kind is MPS [odkaz] developed by JetBrains. In MPS, you can design a custom language, and then use it as an IDE for writing programs in the language and compiling them into executables. The key idea behind projectional editors is that a developer directly manipulates the AST representation of the source code, and not a plain text. \todo{Kratce zminit vyhody a nevyhody toho postupu.} One advantage of projectional editors is that developers practically cannot make syntax errors (MPS IDE will not allow them to make syntax errors). Furthermore, languages can be defined in a modular way (object-oriented, hierarchical) and multiple languages can be easily combined together. 
The mechanisms supported by MPS also give us, for instance, the possibility to extend existing languages very easily, just by inheritance.
Nice illustration of the abilities of MPS is the mbeddr project [citation] which extends C with domain-specific constructs and data types.
We provide more details about MPS in Section 2.1.

\todo{From thesis.
	One of these tools is the JetBrains MPS~\cite{ref:MPS}, which approaches languages from a~different point of view.
	MPS doesn't use the textual representation of code as usual, but rather works with the actual abstract syntax tree that holds the code's structure.
	This has some positive implications as it gives us a~very powerful tool rich on features but also introduces some new problems we haven't encountered before.
	One of them being the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
	MPS also becomes a~very powerful tool once the language has already been created and you start using it inside MPS.
	There is a~huge variety of new languages constructed for MPS that usually solve domain specific problems.
}

\todo{Pouzivat jen terminy "AST nodes" a "language elements" (to spis). Nepouzivat pojem "concepts" abychom ctenare nematli.}
In order to use a language inside MPS, it has to be defined using the MPS concepts. In particular, one has to create the structure (types of AST nodes), editor (projection of AST), and text generation (to enable creation of plain text files used as input for compilers). We provide more details in Section X (Background about MPS).

However, it is a lot of work to create a definition of a programming language in MPS. Therefore, very few mainstream languages are now fully supported by MPS (through some extensions).
\todo{From thesis.
	However, building these languages is a~complicated and time-consuming process and a~lot of effort must be put into them before they are ready for use.

	These examples show that recreating a~full language inside MPS is not an easy task and a~lot of time must be spent on implementing all aspects of the language.
	A big part of this effort is, however, quite straightforward and can be possibly automated, as we show in this paper. Automation speeds up the process of adoption of new languages to a great degree.
}

\todo{From thesis.
	We will look into the possibility of automatic import of already existing general purpose languages into MPS using a grammar description of their syntax.
}
We present INGRID --- a method and a tool for semi-automated construction of MPS definition of programming languages based on their ANTLR grammars.
We also discuss challenges that we encountered during this process and our solution to them. Most of these challenges are general, i.e. not MPS specific, as they apply to the relation between grammars defined in EBNF-like style (rules and tokens, everything defined in plain text) and structured (object-oriented) approach used in MPS (and other language workbenches).

\todo{Hodne zduraznit automation, tedy ze chceme maximalizovat automatizaci.}
The main goal of our/the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible.

\todo{From thesis.
	Automatic import of already existing languages into MPS from the grammar description of their syntax.
}
\todo{Mozna prodat INGRID jako automatickou konverzi gramatik do podoby MPS languages, tedy jako nastroj ktery umoznuje portaci jazyku do prostredi MPS a jejich nasledne pouziti v MPS (treba kombinace s dalsimi jazyky).}

Our primary motivation is to enable better support for mainstream and/or general purpose programming languages (XML, C, C++, JavaScript, Python) in MPS, especially to enable writing programs in these languages using the projectional editor and to combine them with domain-specific languages natively created in MPS.

\todo{Briefly mention the biggest challenges.}
The are related to the way grammar encoding/description of languages are usually written.
They do not hold any information about the code layout, and contain many rules that do not directly correspond to AST nodes (programming language constructs).
We discuss them in more detail later (Section X).

\todo{From thesis.
	INGRID allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
	The biggest benefit is the possibility to code in these mainstream general purpose languages using the projectional editor.
}

However, we do not target just the very complex general purpose languages such as JavaScript.
INGRID can be used also for simple languages.
We believe that, for simple languages, it is less time consuming to create the grammar by hand (e.g., using the ANLTR notation and toolset) and then create the language definition in MPS by using the INGRID tool, than to create it manually using the MPS GUI --- this is another benefit of INGRID.

Structure of the paper.
In the beginning, we will describe the MPS editor and related/similar projects.
Then, we describe our approach to creating language definition in MPS from their grammars, and discuss the major challenges together with our solution.
We will also show some example languages imported with INGRID and discuss our experience.

\section{Background and Related Work}

We will briefly introduce fundamental basics of the MPS environment.
We will analyze existing related projects that are trying to accomplish similar goals, and then we will weigh advantages and disadvantages of these approaches.

\subsection{JetBrains MPS}

JetBrains MPS (Meta Programming System) is a development environment that allows building/creating/defining custom (domain specific) languages or extending existing ones.
Developers can use the newly created languages inside MPS, and for their programs they can also instruct MPS to generate code in a given target language such as Java that is compiled to make an executable.

We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
Focus on basic/fundamental and relevant features.
In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.

MPS differs from mainstream/typical IDEs in one important aspect.
In MPS, user/developer does not work with the textual representation of the source code directly, but rather with its AST (abstract syntax tree) that is the model of the code.
The code itself, which is later compiled and run, is built out of this tree.
When users are creating programs using MPS and projectional editor, they are basically assembling the tree (AST) out of predefined building blocks of the language.
The definition of the language dictates, which statements or elements can be nested inside each other and what structure the AST can have.

Main building blocks of any MPS language are nodes of the AST.
Code of a program is built from the AST nodes.

Keeping the model of the code in the AST form has several advantages.
One of them is that MPS then only allows composing the program strictly following the syntax of the language, which results in an inability to actually write syntactically invalid code.
The code is generated out of the always-valid AST beyond user's reach.

Another advantage is that every node of the AST can be represented in any visual form and take any shape.
It is just up to the designer of the language what shape will each node take inside the projectional editor.
For a better understanding, we will give a~small example.
Let's assume we have a language with typical if/else structure similar to the one in Java:

\begin{center}
	\begin{minipage}{.38\textwidth}
		\begin{alltt}
			if (\textit{condition})
			    \textit{then statement block}
			else
			    \textit{else statement block}
		\end{alltt}
	\end{minipage}
\end{center}

Its abstract syntax tree representation then might look something like shown on Figure~\ref{fig:if_ast}.

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.6]{./images/if_statement_ast.png}
  \caption{"If statement" abstract syntax tree}
  \label{fig:if_ast}
\end{figure}

The underlying child nodes have their own children and so on.
The type of nodes can be restricted by the language structure definition. For example, we can restrict the condition in the if/else statement to be an expression, statement block to be a list of statements, etc.

Alternative representations are possible, and they are not bound to be just the textual representation at all.
For example, blocks of both branches can be aligned next to each other like shown in Figure~\ref{fig:if_editor}.

\begin{figure}[ht]
	\centering
	\hspace{-4mm}
	\includegraphics[scale=0.75]{./images/if_statement_editor.png}
	\caption{"If statement" projectional editor example}
	\label{fig:if_editor}
\end{figure}

The definition of a language element/construct (a type of an AST node) consists of several aspects where each aspect represents a different point of view.
The most important aspects are the following: structure (basic properties, where the node can appear in the AST (tree)), editor (how does it look like, how it is projected/presented), textgen (what code in the underlying base language is generated out of it).
Editor, textgen, and other aspects of a language (see below) must be defined in order to make the language comfortably usable inside MPS.

The three most important aspects (components) of a language definition in MPS are the following: structure (AST nodes and their possible relationships/hierarchy/tree), editor, text generation. They are also most relevant for the contribution of this paper, so we describe them here (and neglect/ignore the other aspects such as type system and data-flow).

\todo{Highlight the object-oriented hierarchical nature of language definitions in MPS. The environment restricts and guides the developer so that he makes less mistakes.}

Some concepts and design patterns known from object oriented programming can be used in the language definition --- inheritance, interfaces, and so on.
The AST node can be inherited from a parent (abstract) node and it can implement a special interface node.
Most importantly, nodes have child nodes and that is how the AST itself is created.

\todo{Zduraznit ze AST node (object) ma nekolik aspectu: structure, editor, textgen. Definice MPS language je organizovana takto --- pres AST nodes, ktere maji nekolik aspectu (ruzne pohledy).}

\paragraph{Structure.}
The most important aspect of any MPS language is structure. It must be created first for each AST node.
It specifies core attributes such as name of the AST node, inheritance relationships, possible child nodes (types and cardinality), implemented interfaces, references to other nodes, other properties (fields of any type that can hold values), etc.
In Figure~\ref{fig:if_statement_structure}, you can see the structure aspect definition for the if statement.

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.55]{./images/if_statement_structure.png}
	\caption{"If statement" structure aspect definition}
	\label{fig:if_statement_structure}
\end{figure}

\paragraph{Editor.}
Editor aspect is where the user defines what the projectional editor representation of a code fragment (an AST node) looks like.
The editor aspect of an AST node defines its visual representation (of the code) in the projectional editor.
The user usually incorporates all children, references, and properties inside the representation so that future user of the language can insert their value into these placeholders.
All elements can be styled using a~language similar to CSS.

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.5]{./images/if_statement_editor_definition.png}
	\caption{"If statement" editor aspect definition}
	\label{fig:if_editor_definition}
\end{figure}

\todo{From thesis.

The editor is a module dedicated/responsible for the definition of appearance of an AST node (concept).
JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells.
These cells then can be styled to user's liking.
There are many different types of cells, each behaving a little bit different towards its contents.
For example, MPS supports various horizontal and vertical lists, etc.
Extra cells can be added on top --- cells which just specify layout adjustments such as indentation and text color.
In Figure~\ref{fig:if_editor_definition}, you can see what a real editor definition might look like for the if statement of the Java language (as defined in MPS).
}

\paragraph{BaseLanguage}
\todo{Introduce the base language here.}
A program written in an MPS language cannot be executed directly.
MPS uses an approach based on translation into a program in the so-called BaseLanguage.
A plain-text representation of a BaseLanguage program is then passed to a system compiler and the result is executed.
The module responsible for generating the program in the BaseLanguage is called "textgen" (we describe it briefly in the next paragraph).

BaseLanguage is also an MPS language --- it is the official port of Java for MPS (created by JetBrains, authors of MPS).
The default base language of MPS is Java, but others are supported too.

Since BaseLanguage is an MPS language, BaseLanguage code is not a (plain) text source code, but again, an AST built out of nodes belonging to the BaseLanguage language.
This means that generating code inside MPS consists of generating some AST out of BaseLanguage nodes.

Besides representing the default (most common) target of text generation, BaseLanguage is used also for all additional programming inside MPS.

\paragraph{TextGen.}
The TextGen aspect defines how a certain AST node will be translated into a text source code (in the base language).
This will allow us to generate source code in text form out of an AST model and ultimately to generate a text program out of the code user writes in the (imported) MPS language.
The TextGen aspect is basically just one method/procedure definition for each AST node in the language. The method produces string representation of the given AST node and its children.
MPS then calls this method for the root AST node of the program.
TextGen aspect has to be defined (especially its more complex functionality) using the BaseLanguage.
Again, we included an example of the if statement (Figure~\ref{fig:if_statement_textgen}).

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.6]{./images/if_statement_textgen.png}
	\caption{"If statement" TextGen aspect definition}
	\label{fig:if_statement_textgen}
\end{figure}

\todo{From thesis.

After the program's AST is created using the projectional editor, there can be rules defined on how to generate code in plain text out of that tree.
This module is called "text gen".
These defined generators can target any language and platform.

\paragraph{Other aspects.}
MPS supports also other aspects, such as type checking and data-flow analysis, but they are not important for our contribution.
Moreover, information needed for their automated generation is not available/contained inside a grammar file.

}

\subsection{Similar Projects}

Currently, there exists an almost full port of the Java language called BaseLanguage~\cite{ref:BaseLanguage} extended with some MPS specific features.
It was imported manually by JetBrains and it is still undergoing changes as Java itself is evolving.

Within the mbeddr project~\cite{ref:mbeddr}, the C language is also manually tailored for MPS.

Now/here we describe three similar projects (with similar goals) that we are aware of.
We focus on (consider) these three projects: PE4MPS, ANTLR{\_}MPS, and mps-metabnf.

\paragraph{PE4MPS.}
PE4MPS~\cite{ref:PE4MPS} is a project by \todo{Doplnit autora} that is trying to solve the same problem as we do.
It solves the lack of information about code layout in grammars by creating a new grammar notation called PE Grammars~\cite{ref:PE} (the PE abbreviation comes from projectional editing).
It has tw components: PE parser and PE4MPS plugin for MPS.
Their approach is to mimic an existing grammar notation called ANTLRv4~\cite{ANTLR4} and enrich its syntax with custom (their own) constructs.
These extensions tell (guide, hint) the parser what the AST node layout should look like.
Supported extensions (already implemented) include horizontal and vertical lists, and some indentation rules --- however, even these few features make the already not-so-simple syntax of ANTLR v4 much more complex.
The parser then uses this information when generating the projectional editor for am AST node.
Author of this approach/project described the PE syntax using the ANTLRv4 notation~\cite{ANTLR4reference} and then used the ANTLR toolset to automatically generate an ANTLRv4 parser for PE grammar files.
The PE parser reads any PE file and stores the language structure found inside to a custom representation of Java objects.

The PE4MPS plugin/project/tool is built on top of the PE parser.
This plugin uses the PE parser to build the PE file representation (the aforementioned tree-like structure of Java objects) and then creates AST nodes and their aspects inside MPS.
The extended syntax brought by PE describes the layout of each element, e.g. it tells the plugin that one set of child nodes should be displayed horizontally, another set should be vertical with each child on a separate line and with some indentation, and so on.

A limitation of the PE4MPS approach is that it, from what we understand, does not implement full ANTLR syntax.
This means that every grammar might need a non-trivial adjustment before its usage.
One of our goals is to enable import of as many languages out-of-the-box as possible and adopting the full specification.

\paragraph{ANTLR{\_}MPS.}
ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
The author of this project is Fabien Campagne
The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
However, it does not try to generate the projectional editor at all (i.e., it does not deal with this problem), probably because it is in an early stage of development.

The way this import process of ANTLR{\_}MPS works is quite different from what we have seen in the PE4MPS project, and it is quite complex/complicated to use.
It works as follows. We give only a brief high-level overview and omit low-level technical details.
The author created a whole new ANTLRv4 MPS language, which is an MPS port of the grammar notations' syntax.
To import a language, the user utilizes this MPS ANTLR language.
The textual grammar is imported automatically into MPS taking the form of the MPS's ported grammar language (so that the textual grammar is converted into MPS nodes, that means an AST node is created for each grammar rule).
No child-parent relationships in the structure are generated by the tool --- all must be created manually by the user.
There are no editor nor TextGen aspects created neither.

\paragraph{mps-metabnf.}
\todo{Popis na zaklade emailove diskuze (Nizozemci).}

\section{Creating Language Definition in MPS}

From all the available notations in which grammars can be written, we decided to use the ANTLRv4 grammar notation~\cite{ANTLR4reference} as the input format (in which the input programming languages have to be defined) for the following reasons.
\begin{itemize}
	\item It is a very widely used grammar notation.
	\item All popular and widely used programming languages have their syntax captured (represented) using this notation --- see the repository~\cite{ref:ANTLR4grammars}.
	\item There is a whole framework/toolset for generating parsers and editing grammars built around it~\cite{ref:ANTLR4}.
	\item It supports Java (it is written in Java), which is the implementation language of MPS and also the default BaseLanguage.
	\item It uses (is based on) an EBNF (Extended Backus-Naur Form) \footnote{https://en.wikipedia.org/wiki/Extended{\_}Backus\%E2\%80\%93Naur{\_}Form} form that it useful because it meets/respects the object oriented design that is used when describing MPS aspects (definitions of MPS languages).
	\item It has its own syntax captured in its own syntax\footnote{https://github.com/antlr/grammars-v4/tree/master/antlr4} (there is an ANTLRv4 grammar for the ANTLRv4 notation). This is a very important feature as it allows us to create a parser for ANTLRv4 grammars so that we can parse grammars of other languages.
\end{itemize}

We use ANTLRv4 grammars as inputs (and work with them).
	https://github.com/antlr/antlr4/blob/master/doc/index.md
	https://github.com/antlr/grammars-v4

Unlike some of the related projects, we did not add any new features into the grammar notation.
We tried to import grammars as they are.

The generation/import process performed by INGRID consists of three phases (import of structure, usable editor, text generation).
We will talk about all phases of the generation/import process, describing encountered challenges/obstacles using an example language.

To fulfill our goal of maximal possible automation, we import/generate the full language structure and then do a lot of mundane and time-consuming work connected to creating projectional editors.

Now we describe/introduce the main steps of our approach (that INGRID tool performs): parsing the grammar, creating the structure, editor, and textgen.
We are able to create/generate the structure aspect and then very basic editor and some default/basic textgen.
In particular, we did not port the ANTLR notation as an MPS language.
In the following subsections, we provide some details about our approach to parsing the input grammar and creating the main aspects of each element (each type of AST node) of an MPS language (structure, editor, and textgen).

During our work on this project, we also discovered/encountered challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component), whitespace (editor aspect).
There is a challenge/issue that we have discovered during our work on this project (in the process of our own implementation), and that is related to construction of the structure aspect.
Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
We describe/discuss these challenges in more detail (also using examples) and our solutions to the challenges inside the respective following subsections.

Result of the construction of a MPS language from the input grammar (import) has to be polished by the human user.
For example, the aspects not supported by INGRID will be used for adjusting the imported language and improving the usability of the language.

\todo{Briefly discuss how our overall approach differs from similar projects (PE4MPS, etc).}
There are more ways how to approach the problem of a grammar import.

Our overall approach differs from the PE4MPS project (Section X.Y) especially in the level of automation.
While the main goal of (our) the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible, usage of PE does not really spare the user any work because it only shifts the manual work from creating projectional editor to writing PE grammars.
What the user would do inside the MPS UI using the projectional editor is in the case of PE4MPS done using a text editor when editing PE grammars.
MPS UI is specially designed/made/tailored for this purpose, guides the user through this process, and supports aids such as auto-completion. In particular, MPS UI allows the user to perform only changes that result in valid output.
Editing the grammar stored in a text file using a standard editor is an inferior approach with respect to using MPS (projectional editor), because a text editor is not as user-friendly, and, more importantly, it is an error prone environment.

Main drawback of ANTLR{\_}MPS, and main difference from our approach, is very limited automation: it does not create the parent-child relationships between AST nodes in the structure, and does not generate any editor and textgen aspects (for any node) at all.
In addition, there is no attention paid to the problem of generating projectional editor and text generation is neglected too.
We blame all this on the project being in its early stage of development.

In the rest of this section, first we define the language that we will use throughout the paper as the running example.
Then we will talk about parsing grammar files \todo{Tohle je mozna zbytecne}.
Lastly, we will describe our approach to tackling generation of language's aspects (structure, editor, and TextGen).
At some places, we discuss important possible alternatives (as there are several possible approaches).

\subsection{Running Example}

We illustrate the main steps on the example of a simplified XML language, which is defined below (in Figure XY).

\todo{Pouzit jazyk SimpleXML na running exaple.}

We decided to use SimpleXML because it is small enough to fit here, but still can be used to illustrate the main challenges and our solution

Throughout the whole paper, we will be using one example language to illustrate main challenges and our solution.
The language is a simplified version of XML. We call it \emph{SimpleXML}.
We created the grammar of the language by taking the official XML ANTLRv4 grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and stripped it down to eliminate/avoid some not so interesting (advanced) features such as XML entities.
Below in Figure X, you can find the grammar of the SimpleXML language in the ANTLRv4 notation.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\textbf{grammar} \textit{SimpleXML};
	\antlrparserrule{document}     :   \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
	\antlrparserrule{prolog}       :   \antlrliteral{<?xml } \antlrparserrule{attribute}* \antlrliteral{?>} ;
	\antlrparserrule{comment}      :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
	             |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
	             ;
	\antlrparserrule{attribute}    :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
	\antlrparserrule{content}      :   \antlrlexerrule{TEXT}
	             |   \antlrparserrule{element}
	             |   \antlrparserrule{comment}
	             |   \antlrlexerrule{CDATA}
	             ;
	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
	\textbf{fragment}
	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
	\textbf{fragment}
	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
	             |   \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.}
	             |   \antlrlexerrule{DIGIT}
	             ;
	\textbf{fragment}
	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
	\antlrlexerrule{TEXT}         :   \antlrregex{~[<"]*} ;
	\antlrlexerrule{CDATA}        :   \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
\end{antlr}
\caption{SimpleXML grammar}
\end{figure*}

We have colored the grammar so it is easier to find the way around it (to make navigation easier).
The grammar contains following basic elements:

\begin{itemize}
	\item \textbf{ANTLRv4 keywords} -- In black, they are required by the specification.
	\item \antlrparserrule{\textbf{Parser rules}} -- Stated in blue, they are the rules describing language's structure.
		The alternatives (that the element on the left side may break into) are separated by the pipe ($|$) character.
	\item \antlrlexerrule{\textbf{Lexer rules}} -- In yellow, they are rules describing terminal symbols that the parser is matching against the input.
		Terminals are encoded as string values and regular expressions.
		The \textbf{fragment} keyword states that the lexer sub-rule is just a helper rule that brings more clarity into the notation but is not visible in the parser output.
	\item \antlrliteralnoap{\textbf{String values (literals)}} -- In red, always inside of a pair of apostrophes, they are similar to lexer rules.
		They describe a terminal string, not using regular expressions but by an exact string match.
	\item \antlrregex{\textbf{Regular expressions}} -- Colored in green, they describe a string token to be matched using a regular expression with the special ANTLRv4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
	\item \textbf{Rule operators (?+*)} - Left in black, following any element, they quantify the number of occurrences the prepending element can appear in.
		These operators can also be prepended by an asterisk (*), meaning they are non-greedy.
\end{itemize}

\subsection{Parsing (the) Grammar}

\todo{Define some name for the AST representing the grammar, and use it consistently throughout the whole paper. Maybe distinguish between "grammar AST" and "MPS structure AST".}
The first step (stage) of creating the MPS language (of the import process) is parsing the grammar file (written using the ANTLRv4 notation) so that we can get an AST tree representing the grammar.

The abstract syntax tree that comes out of the automatically generated (ANTLR) parser is a little bit complicated and full of information not relevant for our purposes (to us).
That is why we translate the complex ANTLR AST into our own simplified tree made out of our own objects.
We discarded as much irrelevant information as possible and only kept information (data) vital/necessary for construction of MPS languages.

Furthermore, we process the grammar in two iterations.
In the first iteration/pass, we build a tree holding names of other referenced rules in the form of strings.
In the second iteration, we walk this tree and resolve all these references to real pointers to other rule objects.
We end up with a clear and simpler representation that is easy to process in the next steps of the import process.

We also had to post-process (simplify) the representation of lexer rules (tokens).
As a result of the initial parsing, ANTLR parser produces some kind of a tree structure that is representing the lexer rule (meaning/note that lexer rules can also be built from alternatives just like parser rules).
However, lexer rules are, in fact (basically), just regular expressions used for parsing input source code into tokens.
We took each/every such tree that represents the lexer rule and flattened it into the equivalent regular expression.

For illustration, consider the lexer rule \antlrlexerrule{Name} from our/the SimpleXML language (Figure XY), together with all its sub-rules, shown in Figure~\ref{fig:lexerrule}.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
	\textbf{fragment}
	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
	\textbf{fragment}
	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
             |   \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.}
             |   \antlrlexerrule{DIGIT}
             ;
	\textbf{fragment}
	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
\end{antlr}
\caption{Lexer rule from SimpleXML}
\label{fig:lexerrule}
\end{figure*}

Intuitively, we can create a regular expression by gluing elements of each alternative together and then join these alternatives with a classic regex OR ($|$), which is exactly its function in the ANTLRv4 grammar notation.
We have captures this idea in the form of the recursive algorithm in Figure~\ref{fig:algflatten}, which covers the whole process.

% TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
\begin{figure*}[ht]
\centering
\begin{antlr}
	Flatten(\textit{R}):
	1) Define \textit{T} as a list of empty string
	2) For each alternative \textit{A} of the rule \textit{R}:
	3)     Define \textit{R} = \antlrap\antlrap
	4)     For each element \textit{E} of \textit{A}:
	5)         If \textit{E} is not yet flattened:
	6)             Flatten(\textit{E})
	7)         \textit{R}.append(\textit{E})
	8)         \textit{R}.append(\textit{E}.operator)
	9)     \textit{T}.add(\textit{R})
	10) Build a string S out of elements t$_{\text{1}}$, t$_{\text{2}}$, ... t$_{\text{n}}$ of \textit{T}:
	11)     (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
	12) Return \textit{S}
\end{antlr}
\caption{Flatten algorithm}
\label{fig:algflatten}
\end{figure*}

The output of \texttt{Flatten(Name)} (output of the Flatten algorithm applied to the \antlrlexerrule{Name} rule) is this regular expression (in this form):

\begin{center}
	\texttt{([:a-zA-Z])(([:a-zA-Z])|(-)|({\_})|(.)|([0-9]))*}
\end{center}

\subsection{Structure}

Now, as the next step, we need to translate/transform the AST (the types of AST nodes) derived from the input grammar into elements of the structure aspect of the MPS language definition (into types of AST nodes), and linking the nodes appropriately.

The goal of this second step is to translate (transfer) the syntax (structure) of the input language that is captured by the grammar (ANTLR), into the MPS structure aspect.
The structure of the MPS language is somewhat similar to the grammar.
This is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
The process of transferring grammar rules into the structure of a MPS language presents/involves some interesting challenges.
The input grammar (EBNF, ANTLR) serves a little bit different purpose than MPS language, and the way grammar rules are usually written down causes some problems that we had to overcome in order to recreate the language inside MPS.
Note that the challenges and problems are not specific to MPS --- they are common for everyone tackling a similar problem.

We are going to describe two different approaches that we designed --- the straightforward approach (Section~\ref{sect:straightforward_approach}) and the shortcut approach (Section~\ref{sect:shortcut_approach}).
We will describe two possible approaches that we developed in Section X and Section Y, respectively.
\todo{Rict neco jako ze ten druhy, slozitejsi, approach vlastne rozsiruje ten prvni, jednodussi.}

However, before focusing on the two approaches, we must describe the basic principles of translation from the AST representing the grammar to the MPS structure, which are common to both approaches.
The structure is created according to object-oriented principles.
Each node (element) is represented by an object (and can be treated as such) that may have a parent node, child nodes, and properties of any data type (similar to fields), it may implement any number of interfaces, and it may also hold references to other nodes.
For every rule of the grammar (for every node in the grammar AST) that breaks into more/multiple alternatives, we must create a node in the MPS structure AST for each alternative of the grammar (every child node).
Names of the nodes for alternatives are composed from the name of the rule and a number indicating the order of appearance in the grammar.
For example, considering the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlgrammarelement}

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
             \textcolor{gray}{|   \antlrap<\antlrap Name attribute* \antlrap/>\antlrap}
             \textcolor{gray}{;}
\end{antlr}
\caption{Fragment of the SimpleXML grammar}
\label{fig:xmlgrammarelement}
\end{figure*}

Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.75]{./images/element_node_common.png}
	\caption{Resulting Element{\_}1 node}
	\label{fig:element_node_common}
\end{figure}

String literals (e.g., keywords such as \textbf{for}) are omitted/skipped because they will appear only in the corresponding projectional editor.
The node named \mpsconcept{Element{\_}1} contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name} inside the first alternative of the \antlrparserrule{element} rule,
Value of the property will be restricted using the regular expression representing the \antlrlexerrule{Name} rule.
We explain later how the references to parser rules (\antlrparserrule{attribute} and \antlrparserrule{content}) as this is where the approaches differ.
The main/key difference between these two lies in the way we create children fields for parser rules and in the way we are going to link them together using interface concepts.

\subsubsection{The Straightforward Approach}
\label{sect:straightforward_approach}

The key idea of this approach is that, for each rule with more than one alternative, an interface node is created.
Consider the \antlrparserrule{content} rule from our SimpleXML language (Figure~\ref{fig:contentrule}).

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
           |   \antlrparserrule{element}
           |   \antlrparserrule{comment}
           |   \antlrlexerrule{CDATA}
           ;
\end{antlr}
\caption{Content rule}
\label{fig:contentrule}
\end{figure*}

Any of the four nodes corresponding to alternatives could appear anywhere the \antlrparserrule{content} rule is referenced.
So we will create one interface node, and then for each alternative of this rule we create one node that implements the interface.
The resulting fragment of structure will look like the one in Figure~\ref{fig:icontentitf}.

\begin{figure*}[ht]
\centering
\begin{mps}
	\mpsinterface{IContent}   :   \mpsconcept{Content{\_}1}
           |   \mpsconcept{Content{\_}2}
           |   \mpsconcept{Content{\_}3}
           |   \mpsconcept{Content{\_}4}
\end{mps}
\caption{IContent interface}
\label{fig:icontentitf}
\end{figure*}

References to other parser rules are captured by links to children (interfaces or object nodes) in the AST nodes.
For the \antlrparserrule{element} rule that is referencing the \antlrparserrule{content} rule, defined as in Figure~\ref{fig:elementrule}, we get the AST node representing the first alternative that is shown in Figure~\ref{fig:element_node_full}.
There is one \mpsinterface{IElement} interface node that is implemented by two object nodes \mpsconcept{Element{\_}1}, \mpsconcept{Element{\_}2}.
The node \mpsconcept{Element{\_}1} has two child links pointing to \mpsinterface{IAttribute} and \mpsinterface{IContent}.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
           ;
\end{antlr}
\caption{Element rule}
\label{fig:elementrule}
\end{figure*}

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.75]{./images/element_node_full.png}
	\caption{Element{\_}1 node's structure aspect}
	\label{fig:element_node_full}
\end{figure}

A problem with this approach, which we call \emph{the layer problem}, is related to the grammar rules (and corresponding AST nodes) that represent intermediary layers created only to make the grammar more readable.
The intermediary layers are created by authors of the grammars in order to make them more readable and more easily (better) maintainable.
We illustrate this problem (and explain it) on the behavior of auto-completion in MPS.
Imagine there is a freshly inserted \mpsconcept{Element{\_}1} node (representing the full XML element as is stated in the first alternative of the \antlrparserrule{element} parser rule).
Now we would like to insert another nested XML element inside.
MPS auto-completion mechanism would give us four options that are captured in the left part of Figure~\ref{fig:layer_problem}.
This is because there exist four nodes (objects) that implement the \mpsinterface{IContent} interface as per each alternative of the rule.

\begin{figure*}[ht]
	\centering
	\includegraphics[scale=0.5]{./images/layer_problem.png}
	\caption{Layer problem in auto-completion}
	\label{fig:layer_problem}
\end{figure*}

However, in order to correctly insert another nested element inside, a user would have to first insert the \mpsconcept{Content{\_}2} node inside \mpsconcept{Element{\_}1} that has an \mpsinterface{IElement} child inside (and nothing else).
Then, as the second step, a user would invoke/trigger/call for the auto-complete again and insert either \mpsconcept{Element{\_}1} or \mpsconcept{Element{\_}2} inside \mpsconcept{Content{\_}2}.
This mean that a user would have to go through two steps, and in the first stepe either correctly guess (or remember the grammar rule's alternative order), which option/item offered by the auto-complete menu to select.

Similarly, if one would decide to replace the nested \mpsconcept{Element{\_}1} node with, let's say, an XML comment (a \mpsconcept{Comment} node), we need to delete both intermediary layers before we get back to the original \mpsconcept{Content{\_}X} selection (crossroads).
The user cannot really see what is happening, since the intermediary level has no appearance or indicator.
This may lead to confusion on the user's part.

The layer problem is solved/addressed by the second approach, which we describe in the next subsection.

\subsubsection{The Shortcut Approach}
\label{sect:shortcut_approach}

The key idea of this approach is to identify rules and nodes that are at the end of the (derivation, inheritance) chain, and offer them directly to the user through the auto-completion menu.
We call such \emph{end rules} and \emph{end nodes}, respectively, because they cannot transparently break into more rules/nodes.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
           |   \antlrparserrule{element}
           |   \antlrparserrule{comment}
           |   \antlrlexerrule{CDATA}
           ;

	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
           ;

	\antlrparserrule{comment}    :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
\end{antlr}
\caption{Content parser rule with children}
\label{fig:contentrulewithchildren}
\end{figure*}

For example, the \antlrparserrule{content} rule, given in Figure~\ref{fig:contentrulewithchildren} together with its child rules, can ultimately expand into following nodes:

\begin{itemize}
	\itemsep0em
	\item \textbf{Content{\_}1} (TEXT)
	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}1}
	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}2}
	\item Content{\_}3 $\rightarrow$ \textbf{Comment}
	\item \textbf{Content{\_}4} (CDATA)
\end{itemize}

The end nodes are highlighted using the bold font.

Figure XY shows the algorithm (in pseudocode) that finds all end rules for a given parser rule.

% TODO pouzit standardni environment pro algoritmy 
\begin{figure*}[ht]
\centering
\begin{antlr}
	FindPathsToEndNodes($R$):
	1)  Define $L$ as an empty list of list of nodes
	2)  Return FindPathsToEndNodes($R$, $L$)

	FindPathsToEndNodes($R$, $L$):
	3)  Define $Q$ as list of list of nodes
	4)  For each alternative $A$ of rule $R$:
	5)      $L1$ = Clone($L$)
	6)      If $A$ is a parser rule with only one element $E$:
	7)          Let $I$ be interface/concept representing rule $E$
	8)          $L1$.Add($I$)
	9)          $P$ = FindPathsToEndNodes($E$, $L1$)
	10)         $Q$ = Merge($Q$, $P$)
	11)     Else
	12)         $L1$.Add($R$)
	13)         Let $P$ be concept representing $A$
	14)         $L1$.Add($P$)
	15)         $Q$.Add($L1$)
	16) Return $Q$
\end{antlr}
\caption{Algorithm for finding end rules}
\label{fig:shortcut_algorithm}
\end{figure*}

To find end rules for each parser rule, the algorithm recursively scans through the parser tree that was built before.
For each parser rule, it tries to find paths leading to some end rule through its alternatives:
\begin{itemize}
	\item Whenever it finds an alternative that contains only one element, and this element is a reference to another parser rule, it has found an intermediary level that can be transparently hidden from the user of the language.
		The process (algorithm) continues by recursively processing alternatives of this "level" rule (since we are not at the end of the chain yet).
	\item Otherwise, an end rule was found (recursion stops here).
\end{itemize}

By appending the rule that is leading to current element (line 12) and then appending that alternative's element itself (line 14), we get a path that contains the full path and the target end rule as the last element of the chain.
The result of this algorithm, for example for the \antlrparserrule{content} rule, equals the listing of the five paths mentioned above.

We do/repeat this for all parser rules and for each rule we get a list of paths that lead from that particular rule to an end node.
We call these paths as \textbf{shortcuts}, because they provide a shortcut from the rule to the end of the chain.
Now that we have the shortcuts, we will discuss several ways how to use the shortcuts (them) to improve the MPS language under construction (how to make the language better).

The primary use of shortcuts is to generate options for the auto-completion menus such that only the end nodes are offered.
This can be technically done in two ways --- (1) through custom auto-complete menus and deletion handlers or (2) by defining a special interface for each rule such that the interface is implemented only by end nodes for the rule.
Shortcuts have to be considered when inserting nodes but also when deleting them, i.e. the whole chain (shortcut path) consisting of possibly multiple intermediary AST nodes (levels) up to the end node must be added, respectively deleted.
The effect of insertion must be fully reversed when the user wants to delete some end node (the given end node).

\subsection{Editor}

When all types of AST nodes in the language are created, the next step is to define their visual representation in the projectional editor.
As stated before (in Section X (about JetBrains MPS, part of background)), MPS uses a cellular system that allows placing node's properties and children into a table-like arrangement.
Cells of different types are available --- for storing property values, references to child nodes, and keywords (string literals), and also cells that influence layout (e.g., indentation).

The purpose (goals) of this (third) step is to create the editor aspect for each language element.
It should project all attributes of the element --- name, properties, children, and so on --- using the respective cell types.

The main challenge (problem) here is that an input grammar (ANTLR) does not provide/give any help/aid/information regarding the code layout.
As we already said in the introduction, one of the biggest challenges/problems that arise here is that the grammar description of a language does not hold any information about the code layout.
The structural rules in the grammar only tell us, what the syntax tree looks like and how the code is broken into nodes.
In particular, the grammar says nothing about indentation, line breaks, and another formatting.

We tackled this problem by designing an automated learning approach, which generates the layout using several heuristics.
We present our solution to this problem. It involves some intermediary step whose purpose is to generate the necessary information.
Our solution is semi-automated (using some heuristics) --- user's help in an interactive manner is needed for final touches (polishing).
It works as follows.

As input for the algorithm (learning approach), the user has to supply to a set of valid source files written in the imported language, together with the grammar file.
These four steps/operations are then performed.
\begin{enumerate}
	\item Automated generation of an ANTLR parser for the input language using the ANTLR library.
	\item Modify/extend the code of the generated parser in order to record some additional information.
	\item Compile the parser into an executable form.
	\item Use the parser to parse the supplied source code files and extract information about the layout of the code during this process.
\end{enumerate}
This way, the imported/created MPS language would inherit the code style/layout directly from the user, i.e. it would learn directly from the code.
In step 2, the parser is modified such that it remembers/stores the line number and column for each parsed token.
This is very efficient and it enables to create a nice layout.
From the line number, the algorithm can easily determine whether there is a line break between some tokens (statements, language elements).
The column number can be then used to detect indentation.
However, this approach is still heuristic and may give imprecise (suboptimal) results --- mainly because it is not always possible to precisely match language elements (in the MPS structure) with all tokens that belong to it.
Note also that quality of the extracted information always depends on the amount of the source code files provided (available, supplied) and on how representative the set of source code files actually is.
When our algorithm cannot determine some custom layout, it just creates all the cells (content) of the editor aspect and arranges them in a single row.
Figure~\ref{fig:editor_adjustment} XY shows the difference between the default single-row layout (in the upper left corner) and the fully customized layout (in the bottom right corner) for the \mpsconcept{Element{\_}1} node.

% TODO smazat cervenou sipku z toho obrazku a mozna rozdelit na dve casti (vlozit tam caru): default single-row code layout, optimal fully customized layout
\begin{figure*}[ht]
	\centering
	\includegraphics[scale=0.5]{./images/editor_adjustment.png}
	\caption{Projectional editor adjustment of the Element{\_}1 node}
	\label{fig:editor_adjustment}
\end{figure*}

In practice, however, the user still has to tweak the editor aspect --- to adjust the code layout --- because the heuristics are not 100\% reliable.
But it takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language~\cite{ref:javascript} and manually adjusting all editors that needed it, all that in a less than hour time.
More details are provided later in Section X (examples) --- there we describe this issue further.

\subsection{Text Generation}

The final/last, fourth step, is to create the TextGen aspect for each AST node.
The purpose of the TextGen aspect is to define, for a language element (AST node), what code in the BaseLanguage will be generated out of it, once we want to compile and run our MPS code/program.
This is needed to allow/enable users to generate plain-text source code representation out of the AST they built/created inside MPS.
Again, a TextGen aspect must be created for each language element.
Basically, it is a single method in the BaseLanguage (or in Java?) for each language element.
The method generates a string (plain-text) representation of the element, appends the string to the output stream, and then calls TextGen methods on all children of the element.

A very basic example of BaseLanguage code that can be used as the TextGen aspect of an \mpsconcept{Element{\_}1} node, which represents the full XML tag with content, is shown in Figure~\ref{fig:textgen_example}.
It appends all literals, properties, and children in the same order as they appear in the grammar definition.

\begin{figure}[ht]
	\centering
	\includegraphics[width=90mm]{./images/textgen_example.png}
	\caption{Example of a~TextGen aspect}
	\label{fig:textgen_example}
\end{figure}

Like in the case of editor, we discovered some interesting challenges related to the code layout, whitespaces, and more (so on).
We describe our solution to the challenges here in this section.

The reader might have noticed that there is no whitespace handling inside of our SimpleXML grammar (Section XY, running example).
Grammars in the ANTLRv4 notation are usually written in such a manner that whitespace characters are skipped (ignored).
However/Nevertheless, TextGen aspect (generator) should/must know between which children there must be a whitespace and where it is forbidden (i.e., where to put some whitespace characters) so that the produced code is a valid one.
For example, in the case of the \antlrparserrule{element} rule, TextGen must determine that there should be a space in between \antlrparserrule{attribute} and \antlrlexerrule{Name}, while it is not required between \antlrliteral{\textless} and \antlrlexerrule{Name}.
In general, it is sufficient to put just a single space character at places/locations where there has to be some whitespace in order to make the code valid (syntactically correct).

Layout of the generated code is also important. For example, the user may want to produce more readable code by adding indentation or line breaks.

In fact, the problem of TextGen layout is quite similar to the one we have seen with the projectional editor, described in Section XY.
Since we are dealing mostly with text-based languages, their representation in projectional editor must be almost the same as their expected plain-text representation (output).
Therefore, one could use the same approach for editor and TextGen --- adding line breaks and indentation.
Once we would have some information about the layout for generating better projectional editor, we could leverage the same information and use it to improve TextGen.

Nevertheless, in our current approach, the editor aspects have to be adjusted manually, after the import is done, because for now it is the fastest and the most efficient way how to deal with the problem of customizing the editor to capture a nice layout.
Therefore, we used a different approach for TextGen --- a really simple heuristics which gives surprisingly good results.
In the first step, a very basic TextGen is created that inserts spaces in between every two tokens of the language element (AST node).
Then, as the second step, spaces are eliminated (restricted) from positions where they are not really needed.
We describe all the cases here:

\begin{enumerate}
	\item Whenever there is a non-alphabetical literal,	appearing as a plain string token defined in the grammar, that might get recognized by the parser safely without the need for a whitespace separator around, then we omit spaces around it.
		For example, consider \textbf{'\textless'} in SimpleXML.
	\item Another group are strings inserted by the user of the language, whose form is constrained by a regular expression.
		For example, an XML tag's name belongs to this group.
		They are usually values of properties defined in structure aspects.
		A space must be preserved only when a neighboring literal may end with an alphabetical character. Otherwise, we can omit the space.
		In practice, this will eliminate redundant spaces inside of quotes, next to semicolons, and around brackets, but, on the other hand, it will separate language keywords (function, var, in, etc.) from other content by keeping the space character in between them.
	\item Spaces are omitted also in the case of child nodes that are not present (e.g., when they are optional), and in the case of empty lists of child nodes, so that spaces do not accumulate.
	\item On the other hand, when two child elements are next to each other, we always insert space.
	\item Sequences of children are separated with space or with a line break.
\end{enumerate}
These few simple heuristics work surprisingly well and give us some very nice results.
We discuss our experience with XML, JSON, and JavaScript in Section XY (Evaluation).

Figure~\ref{fig:textgen_final} shows an example of a TextGen aspect for the full SimpleXML element.

\begin{figure}[ht]
	\centering
	\includegraphics[width=95mm]{./images/textgen_final.png}
	\caption{Generated TextGen aspect for the SimpleXML element}
	\label{fig:textgen_final}
\end{figure}

The automatically generated TextGen code can be adjusted manually very easily, so that it generates nicely indented XML code.
In the case of the example above (in Figure F), we only need to wrap the \mpsconcept{Content{\_}2} child with indentation and change the sequence separator to a new line character.
The resulting adjusted aspect (code) is shown below in Figure~\ref{fig:textgen_adjusted}.

\begin{figure}[t!]
	\centering
	\includegraphics[width=85mm]{./images/textgen_adjusted.png}
	\caption{Adjusted indentation inside the TextGen aspect}
	\label{fig:textgen_adjusted}
\end{figure}

\subsection{Remarks about Grammars}

As we have already shown in previous sections, it is very easy to write a grammar in a way that will cause problems during import into MPS (or once it is imported).
Problems might occur during the creation of any aspect.
Here we will look at some general problems/issues that grammar import poses (might pose).
We will talk about these problems with respect to the challenges/obstacles we tried to overcome and which we described in previous sections (dedicated to structure, editor, and texgen, respectively).
We also show a few examples of additional possible complications.

\paragraph{Adjusting grammars.}
There are some cases, where altering the input grammar might yield far better MPS language.
We will show two examples how the usability of the resulting MPS language can be improved.
Let us look at the definition of an XML attribute in Figure~\ref{fig:xmlattribute}.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
	\antlrlexerrule{STRING}      :   \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
	            |   \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'}
	            ;
\end{antlr}
\caption{Definition of an XML attribute}
\label{fig:xmlattribute}
\end{figure*}

The original XML grammar has quotes as a part of the value.
For the resulting MPS language, it would mean that there would be a placeholder for the attribute value that would expect us to input the leading and trailing quote together with the value too each time.
It would also be marked red (by the syntax checker) unless we enter both quotes inside the value since the regular expression checking for quotes will not match.
The user might be confused by this and will not be able to tell why his string value is incorrect.

In our SimpleXML language, we adjusted the grammar easily in the manner that we show in Figure~\ref{fig:xmladjustgrammar}.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
	            |   \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'}
	            ;
	\antlrlexerrule{TEXT1}       :   \antlrregex{~["]*} ;
	\antlrlexerrule{TEXT2}       :   \antlrregex{~[']*} ;
\end{antlr}
\caption{Adjusted grammar of SimpleXML}
\label{fig:xmladjustgrammar}
\end{figure*}

We turned quotes into literals, thus ensuring that they will only appear in the projectional editor as fixed constant cells.
We will not have to encapsulate the value in them each time.
The user will only have to choose, which attribute version he wants to use (single or double quotes).

As the second example, we will show the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
Every statement in JavaScript needs to be either followed by a semicolon, newline, file end or end of the block --- see the Figure~\ref{fig:javascriptstmt}.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{eos}            : \antlrlexerrule{SemiColon}
	               | \antlrlexerrule{EOF}
	               | {lineTerminatorAhead()}?
	               | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}?
	               ;
	\textcolor{gray}{// Example reference of the eos rule}
	\antlrparserrule{breakStatement} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos}
	               ;
\end{antlr}
\caption{Statements in JavaScript}
\label{fig:javascriptstmt}
\end{figure*}

Because there are multiple options, our import plugin would create a placeholder at the end of every statement.
Every language element representing a statement would have one child of the \mpsinterface{IEos} interface type.
This placeholder needs to be manually filled in for each statement.

Since the projectional editor has much bigger power over the form of the code, we (a user) might want to have each statement on a separate line.
Since we can differentiate between statements on the AST level, we do not need an explicit separator between them.
This means that we might want to simplify the language and leave the semicolon out, or leave it just as a constant fixed part of the projectional editor, but not as something the user must explicitly fill in.
Then we can just put each statement on a separate line as it is usual for JavaScript code, but we do not need the semicolon anymore.
This small adjustment is very quick when done inside the grammar.
We just change the \antlrparserrule{eos} rule to the form shown in Figure~\ref{fig:eosrule}.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{eos} : \antlrliteral{;} ;
\end{antlr}
\caption{eos parser rule}
\label{fig:eosrule}
\end{figure*}

This way, we can change the grammar in a way that it will not describe the same ECMAScript language as before, but will definitely make our MPS language more usable.
TextGen aspect may be defined in such a way that puts the semicolon back (after each statement) in the generated plain-text representation. 
This would be very important especially when JavaScript is available as another base language (which is actually a plan (future work) of the JetBrains company).

Above, we have shown that adjusting the input grammar might be a very fast mean of tweaking/improving the usability of generated MPS language, and sometimes it may be the only proper solution in some complex situations.
In general, the question we can ask ourselves is whether one aims for a full precise/faithful port of the input language or for an MPS alternative.

\todo{Nahradit slovo "break" v nasledujicich odstavcich (v kontextu "break the parser") za vhodnejsi slovo ktere pusobi vic odborne.}

\paragraph{Breaking grammars and parsers.}
However, we have found that there is one big problem with grammar adjustment that we would like to point out.
The problem is that it is very easy to change the input ANTLR grammar in a way that breaks the ANTLR parser generated out of it.
By breaking we mean that it stops parsing the original language.
We illustrate the problem on our SimpleXML language.

As stated before, when creating the SimpleXML grammar, we have started off with the original XML grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and did some adjustments to it.
We ended up with a grammar that can be processed by our method/plugin and creates a nice usable MPS language.
Nevertheless, we noticed that even though the imported language behaves well enough and mimics the XML language quite nicely, the ANTLR parser generated out of this grammar no longer parses XML successfully.
Some changes we have made, such as the attribute adjustment, broke the grammar down.
More precisely, our changes improved the MPS language while breaking down the parser, and we have not even noticed it because, from the perspective of MPS, the imported language still corresponds to XML.

What we are trying to say (show here) is that it is very easy to perform a grammar adjustment that will, at first, seem harmless and valid inside MPS, but will break the parser.
The problem is that the user performing this change probably will not be aware of breaking it.

The main (underlying) cause of this problem is the way the ANTLR parser is implemented and the quite different purpose we are using the grammar for in our project/tool/method.
There are many ways how various parsers deal with, for example, token matching.
For instance, ANTLR introduces so called \emph{greedy} and \emph{non-greedy} operators.
The greedy way, in which ANTLR matches input on defined tokens and prioritizes their selection, makes some rules very dangerous.
When a rule matches a wide range of input, it might happen during parsing that it is prioritized over other rules and swallows a lot more input than the author of it intended.
Usually, these dangerous rules are bound to some parser context, which makes them behave well.

However, for practical reasons, it is very important that grammar is not broken by our custom adjustments (and that the ANTLR parser of the language works properly), because we might need to automatically generate parsers of the adjusted language.
Imagine this very real scenario (that will be supported in the future):
\begin{enumerate}
	\item We have imported the language inside MPS.
	\item MPS now knows the structure of the language, and we can code in MPS using this language.
	\item We would expect MPS to be able to load an existing source code, written in this language, from a text file and import it inside MPS.
	\item We would like to use MPS to safely edit this code, using all the features of MPS.
	\item We would also like to export the code in a plain-text form again and save it back to the source file.
\end{enumerate}
In order to make this happen, we must be able to generate a correctly-working ANTLR parser out of the adjusted source grammar that we used to (that helped us) import the language.
We must also be able to match nodes of the AST coming out of the ANTLR parser to elements of the MPS language, and we must build the MPS AST out of the ANTLR AST.

\section{Evaluation}

We implemented everything (the proposed approach described above) in the form of an MPS plugin that allows users to carry out the import.

We used a mix of Java and BaseLanguage in our implementation.
We used the MPS API, which can be called from the BaseLanguage programs too, to generate some of the language elements.
BaseLanguage is needed to use/call the MPS (language) API, i.e. to programmatically create new language elements and their aspects.

The plugin uses the ANTLR library to work with ANTLRv4 grammar files (to parse them, etc) and several MPS libraries (that implement the MPS API).

For the first step of automated construction of MPS languages from grammars (i.e., for parsing the grammar), the plugin uses the ANTLR library~\cite{ref:ANTLR4} to generate a Java ANTLR parser of the ANTLRv4 notation, because there exists an ANTLRv4 grammar for the ANTLRv4 notation~\cite{ref:ANTLR4reference}.

\todo{
	We should provide url to the github repository (implementation) and pointer to some package with ANTLR grammars and imported MPS languages (take from Attachments.zip in the master thesis).
}

\todo{Popsat nase zkusenosti s importem slozitych jazyku (XML, JavaScript, Python, a dalsi).}

From introduction: We will also show some example languages imported with INGRID and discuss our experience.

In order to evaluate INGRID, we tried to import several real/mainstream languages:
\begin{itemize}
	\item \textbf{SimpleXML} --- Simplified XML that we used in this thesis for explanatory purposes.
	\item \textbf{JSON} --- Similar to XML, but this time a full port of the specification.
	\item \textbf{ECMAScript 5.1} --- Specification of the language known as JavaScript, dated to the year 2011, which is currently the most frequently adopted version.
\end{itemize}
Results of the import process are included in the package that is available at \todo{url}.
The repository \todo{Add url} contains an example MPS project. There are three languages.
For each example language, you can find its original import, exactly in the form as created by our plugin.

We experimented with import of XML, JavaScript, Python, and maybe some other languages.

These general purpose languages are quite complex when it comes to their structure and the overall syntax variety.

We must emphasize that the automatically imported languages are not ready-to-use full-fledged MPS languages.
We did not expect to achieve that at this stage, as that is a much bigger, if not impossible, challenge.
Imported languages contain full structure found in the respective grammars and some more aspect definitions that will help users of the languages in creating code.
The imported language has to be manually adjusted (tweak, polished) by the end user (developer) and some of its complications have to be resolved manually by a human.

We found out that manual adjusting of the editor/textgen aspect (adding line breaks and indentation) is far more effective and precise than any automated approach we came up with (during our research).
The imported language can be, in our opinion, adjusted very fast into a more readable form.

The package at \todo{url} contains also adjusted versions of all three languages.
Then, for each language, there is an adjusted version.

This adjusted version was created from the original import and the author spent no more than between 20 to 60 minutes of working on it.
The first author spent only a couple of minutes of (manually) adjusting them after the automated import, so that we try to prove our point above.
More specifically, the JSON language was ready for use in no more than 20 minutes of minor adjusting.
We chose this language because it does not require us to implement complex aspects, such as type checking or data flows.
Similarly, the SimpleXML turned out very nice.

Our main goal was to prove that even though our method (plugin) does not create a perfect language, it can be customized very fast into a very usable form.
We manually tweaked/customized the editor and the TextGen aspect only, structure aspect was left in its original form.
This justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical (and the best).

We have also tried to use the INGRID tool/plugin to import the JavaScript language, which is exactly the type of a complicated general purpose language that might be of interest to other people.
There are already some projects, where a manual port of JavaScript is being done.
One of them is the ECMAScript4MPS~\cite{ref:ECMAScript4MPS} from the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section X (similar projects, related work).
The result of import by INGRID is quite good, comparable to the PE4MPS project.
Our language is, of course, missing a lot of more advanced aspects/features that, for example, can help deriving types of expressions and so on.
These aspects improve the usability by a lot, e.g. when the user starts writing numbers, a number literal is inserted, whereas in our import the user has to first insert the concept representing the number literal and then proceed with writing numbers.
These actions are needed in any MPS language and are expected to be implemented after the import step, \todo{Maybe say that INGRID does not generate these actions/aspects/features yet but it is our future work} as it is not possible to generate them automatically.
However, when it comes to language structure, concept aliases, and auto-completion, we think that INGRID (our method/plugin) did a very good job.
The structure aspect is very similar to the ECMAScript4MPS's one, which was created manually over the course of a large number of hours.
INGRID does that (achieves the same result) fully automatically.

To summarize, import languages such as JavaScript, JSON, MySQL and others is giving nice results.

Results of our experiments with languages such as JSON or JavaScript show that the proposed approach (INGRID) yields very good results, comparable to existing manual ports, and spares the user from hours of tedious, error prone and time-consuming work.

Limitation: the project is lacking proper error reporting and cannot handle cyclic rules (which could be found in Python grammar, for example).

\section{Conclusion and Future Work}

\todo{Write something.}

There is lot of further follow-up work to be done. Many open problems remain.

Automated generation of the editor aspect, probably using some kind of machine learning.
Generating nicely designed and user-friendly editors, maybe using artificial intelligence (machine learning) that would evaluate (analyze) existing code in plain text and based on that create the editors.

Follow-up work might look more into the problem of code layout detection, possibly exploring suggested approaches such as the learning approach described in Section XY \todo{Tohle se snad podari aspon castecne implementovat}.
This could lead to better results in editor and TextGen aspect generation which could definitely use some improvement.
Another future endeavors will definitely touch the subject of generating parsers that would enable the user to import existing source code into MPS.

\acks
% TODO
This research was supported by JetBrains. \todo{Pridat jeste nejaky grant a kompletne preformulovat.}

\bibliographystyle{abbrvnat}

% TODO opravit formatovani jednotlivych zaznamu a vhodne seradit
\begin{thebibliography}{}

\bibitem{ref:MPS} JetBrains MPS, \url{https://www.jetbrains.com/mps/}

\bibitem{ref:mbeddr} mbeddr, \url{mbeddr.com/}

\bibitem{ref:BaseLanguage} BaseLanguage, \url{confluence.jetbrains.com/display/MPSD34/Base+Language}

\bibitem{ref:PE4MPS} PE4MPS project, \url{https://github.com/mar9000/pe4mps}

\bibitem{ref:PE} PE project, \url{https://github.com/mar9000/pe}

\bibitem{ref:ANTLR2MPS} ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR{\_}MPS}

\bibitem{ref:ANTLR4} ANTLRv4, \url{https://github.com/antlr/antlr4}

\bibitem{ref:ANTLR4reference} T. Parr. The Definitive ANTLR 4 Reference. The Pragmatic Bookshelf, 2013

\bibitem{ref:ANTLR4grammars} ANTLRv4 grammar repository, \url{https://github.com/antlr/grammars-v4}

\bibitem{ref:ECMAScript4MPS} ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}

\bibitem{ref:javascript} JavaScript ANTLRv4 grammar, \url{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}

\end{thebibliography}

\end{document}

### 2016-09-20 paper.tex
small note
JB comments and changes added to chapter 1 Introduction

@@
 
-Prevailing approach to the definition of valid syntax of programming languages is through/using grammars written, for example, using notations based on (or similar to) EBNF [reference].
-Many popular tools exist for automated generation of parsers from the grammars, such as ANTLR and Bison/Flex.
-Source code of programs in the language is then written in a text editor, which is a part of some IDE.
-
-An alternative emerged recently in the form of projectional editors that are used mainly in special IDEs for domain specific languages, which are called language workbenches.
-A popular tool of this kind is MPS [odkaz] developed by JetBrains.
-In MPS, you can design a custom language, and then use it as an IDE for writing programs in the language and compiling them into executables.
-The key idea behind projectional editors is that a developer directly manipulates the AST representation of the source code, and not a plain text.
-\todo{Kratce zminit vyhody a nevyhody toho postupu.} One advantage of projectional editors is that developers practically cannot make syntax errors (MPS IDE will not allow them to make syntax errors).
-Furthermore, languages can be defined in a modular way (object-oriented, hierarchical) and multiple languages can be easily combined together.
-The mechanisms supported by MPS also give us, for instance, the possibility to extend existing languages very easily, just by inheritance.
-Nice illustration of the abilities of MPS is the mbeddr project [citation] which extends C with domain-specific constructs and data types.
+\JB{Odstavec mirne upraven} The prevailing approach to define valid syntaxes for programming languages is through grammars written typically in notations based on (or similar to) EBNF [reference].
+Many popular tools exist for automated generation of parsers from the grammar definitions, such as ANTLR and Bison/Flex.
+The source code of programs in such languages is written in a text editors, usually embedded into an IDE.
+
+\JB{Odstavec mirne upraven} Projectional (or structural) editing offers an alternative approach.
+While the idea emerged as early as in 1970s, it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
+A recent revival of projectional editing has been observed in the area of language workbenches - IDE-like tools that enable the developers to manipulate the actual language definition.
+The Meta-programming System (MPS) [reference] by JetBrains is an open-source language workbench that focuses of Domain Specific Languages and leverages the concept of projectional editor.
+MPS allows developers to design custom languages and then use use them to write code and compile it into executables.
+MPS provides the whole IDE infrastructure for such languages.
+
+\JB{Odstavec mirne upraven} The key idea behind projectional editors is that a developer directly manipulates the AST representation of the source code instead of plain text.
+To avoid the fate of the early projectional editors MPS has put extra effort into making text manipulation feel like editing real text.
+MPS models, although being persisted as XML or binary files, can be versioned using the industry standard source-control systems (Git, Subversion).
+On-line model browsers have been incorporated into some of the popular public source code repositories (BitBucket) [reference] .
+
+\JB{Odstavec mirne upraven} The absence of parsers brings about numerous benefits, which are in particular useful for DSLs, which are frequently used by domain experts, who may not be professional programmers.
+\begin{itemize}
+	\item Since a projectional editors controls the interaction between the user and the code, the developers practically cannot enter syntactically invalid code.
+	\item Languages can be defined in a modular way and multiple languages can be easily combined together or extend one another, while parsers seriously limit language modularity for traditional parsed languages.
+	\item The languages may offer contextual or non-parseable notations, such as tables, diagrams or positional syntax.
+	\item Since the projection is detached from the physical representation of code (AST), authors can define multiple notations and allow the developers to switch between different representations of the code.
+\end{itemize}
+A motivating illustration of the abilities of MPS is the mbeddr project [citation], which extends C with numerous domain-specific constructs and data types.
 We provide more details about MPS in Section 2.1.
 
+\JB{Navrhuji smazat nasledujici odstavec}
 \todo{From thesis.
@@
 \todo{Pouzivat jen terminy "AST nodes" a "language elements" (to spis). Nepouzivat pojem "concepts" abychom ctenare nematli.}
-In order to use a language inside MPS, it has to be defined using the MPS concepts.
-In particular, one has to create the structure (types of AST nodes), editor (projection of AST), and text generation (to enable creation of plain text files used as input for compilers).
+
+\JB{Odstavec upraven} Projectional editing introduces some new problems we haven't encountered before.
+One of them is the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
+In order to use a language inside MPS, it has to be defined through the MPS infrastructure.
+In particular, one has to create the abstract syntax called Structure (the types of allowed AST nodes), concrete syntax called Editor (projection of the AST on the screen and its interactions with the user), and text generation (to enable creation of plain text files used as input for compilers).
 We provide more details in Section X (Background about MPS).
 
-However, it is a lot of work to create a definition of a programming language in MPS. Therefore, very few mainstream languages are now fully supported by MPS (through some extensions).
+\JB{Odstavec pridan pro vysvetleni motivace prace} MPS is typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
+These DSLs are then transformed on the AST level by a series of model-to-model transformations to an AST model representing the desired semantics in a General-purpose language (GPL), such as C or Java.
+This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
+This implies that the target GPL must also be defined by the MPS means, to allow the DSLs to have their models transformed into a model in the GPL.
+
+\JB{Odstavec doplnen} However, it requires substantial effort to create a definition of a GPL in MPS.
+Therefore, very few mainstream GPLs are now fully supported by MPS.
+Only Java and C have been implemented to date, which limits the DSLs authors to only target these supported target GPLs.
+We believe that an automated process of importing grammars of GPLs into MPS would encourage more GPLs to be migrated to MPS and thus giving the DSL authors more options.
+
 \todo{From thesis.
@@
 
-\todo{From thesis.
-	We will look into the possibility of automatic import of already existing general purpose languages into MPS using a grammar description of their syntax.
-}
+\JB{Odstavec upraven} Since a big part of the language definition process in MPS is quite straightforward and can possibly be automated, we will look into the possibility of automatic import of already existing general purpose languages into MPS using a grammar description of their syntax.
+
 We present INGRID --- a method and a tool for semi-automated construction of MPS definition of programming languages based on their ANTLR grammars.
@@
 Our primary motivation is to enable better support for mainstream and/or general purpose programming languages (XML, C, C++, JavaScript, Python) in MPS, especially to enable writing programs in these languages using the projectional editor and to combine them with domain-specific languages natively created in MPS.
+\todo{Zduraznit take pouziti importovanych jazyku pro cile generovani z DSL --- to plati zejmena u mainstream general-purpose languages jako treba JavaScript.}
 

### 2016-09-21 paper.tex
Initial edits in chapter 2 by JB

@@
 
-JetBrains MPS (Meta Programming System) is a development environment that allows building/creating/defining custom (domain specific) languages or extending existing ones.
-Developers can use the newly created languages inside MPS, and for their programs they can also instruct MPS to generate code in a given target language such as Java that is compiled to make an executable.
-
-We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
-Focus on basic/fundamental and relevant features.
-In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.
-
-MPS differs from mainstream/typical IDEs in one important aspect.
-In MPS, user/developer does not work with the textual representation of the source code directly, but rather with its AST (abstract syntax tree) that is the model of the code.
-The code itself, which is later compiled and run, is built out of this tree.
-When users are creating programs using MPS and projectional editor, they are basically assembling the tree (AST) out of predefined building blocks of the language.
-The definition of the language dictates, which statements or elements can be nested inside each other and what structure the AST can have.
-
-Main building blocks of any MPS language are nodes of the AST.
-Code of a program is built from the AST nodes.
-
+\JB{Odstavec upraven} JetBrains MPS is a complete language workbench - a development environment that allows developers to define their own languages and then use them to develop code.
+The code can then be transformed into a target language, typically a GPL, such as Java or C.
+
+\JB{Navrhuji odstranit} We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
+Focus on basic/fundamental and relevant features. In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.
+
+\JB{odstavec upraven} MPS differs from typical IDEs in one important aspect - the projectional editor.
+The developer does not work with the textual representation of the source code directly, but rather with its AST (abstract syntax tree) that is the model of the code.
+When users are creating programs using the MPS projectional editor, they are basically assembling the tree (AST) out of predefined building blocks of the imported languages.
+The definition of the language dictates, which where in the AST can certain elements be placed or how they can be nested inside other elements.
+In traditional IDEs it is the parser that constructs an AST using the language's elements.
+
+\JB{Odstavec upraven} The building blocks of MPS models are called nodes.
+Code of any program in MPS is built from nodes and these nodes represent instances of elements defined in one of the languages that the program declares to be using.
+These language elements are called concepts.
+In brief, nodes in code are instances of concepts defined in the language.
+
+\JB{Vyhody/nevyhody diskutujeme v kapitole 1. Navrhuji nasledujici odstavec zrusit nebo text z kapitoly 1. presunout sem}
 Keeping the model of the code in the AST form has several advantages.
@@
 
-Another advantage is that every node of the AST can be represented in any visual form and take any shape.
-It is just up to the designer of the language what shape will each node take inside the projectional editor.
-For a better understanding, we will give a~small example.
-Let's assume we have a language with typical if/else structure similar to the one in Java:
+\JB{Odstavec upraven} One of key advantages of projectional editor stems from separation of abstract and concrete syntax.
+While the AST provides an exhaustive representation of the code, the way it is displayed on the screen and the way the user interacts with it are unconstrained.
+The editor can take any visual form and shape.
+The language author can define multiple visualizations and let the developer to choose one that fits best the task at hands - editing, debugging, reviewing, resolving merge conflicts, etc.
+
+For a better understanding, we will give a~small example. Let's assume we have a language with typical if/else structure similar to the one in Java:
 
@@
 
-Its abstract syntax tree representation then might look something like shown on Figure~\ref{fig:if_ast}.
+Its abstract syntax tree representation will look something like shown on Figure~\ref{fig:if_ast}.
 
@@
 
-Alternative representations are possible, and they are not bound to be just the textual representation at all.
-For example, blocks of both branches can be aligned next to each other like shown in Figure~\ref{fig:if_editor}.
+\JB{odstavec upraven} Alternative visual representations are possible and they are not bound to be just the textual representation at all.
+For example, blocks for both of the branches can be aligned next to each other and displayed with differend background color, like shown in Figure~\ref{fig:if_editor}.
 
@@
 
-The definition of a language element/construct (a type of an AST node) consists of several aspects where each aspect represents a different point of view.
-The most important aspects are the following: structure (basic properties, where the node can appear in the AST (tree)), editor (how does it look like, how it is projected/presented), textgen (what code in the underlying base language is generated out of it).
-Editor, textgen, and other aspects of a language (see below) must be defined in order to make the language comfortably usable inside MPS.
-
-The three most important aspects (components) of a language definition in MPS are the following: structure (AST nodes and their possible relationships/hierarchy/tree), editor, text generation.
-They are also most relevant for the contribution of this paper, so we describe them here (and neglect/ignore the other aspects such as type system and data-flow).
+\JB{Odstavec mirne upraven}The definition of a language element (the concept) consists of several aspects and each aspect codifies a different part of the nodes' behavior.
+The essential aspects are the following: structure (abstract syntax), editor (concrete syntax - how the code is projected and edited) and textgen (how to transform the AST into textual source code).
+If, instead of generating text directly, the language is supposed to be generated into another language that is available in MPS, the generator aspect must be used to specify the model-transformation rules.
+The structure, editor and textgen aspects are most relevant for the contribution of this paper, so we describe them here (and neglect/ignore the other aspects such as type system and data-flow).
 

### 2016-09-21 paper.tex
Additional edits in chapter 2 by JB

@@
 The type of nodes can be restricted by the language structure definition.
-For example, we can restrict the condition in the if/else statement to be an expression, statement block to be a list of statements, etc.
+For example, we can restrict the condition in the if/else statement to be an expression, the statement block to be a list of statements, etc.
 
@@
 
-Some concepts and design patterns known from object oriented programming can be used in the language definition --- inheritance, interfaces, and so on.
-The AST node can be inherited from a parent (abstract) node and it can implement a special interface node.
-Most importantly, nodes have child nodes and that is how the AST itself is created.
-
-\todo{Zduraznit ze AST node (object) ma nekolik aspectu: structure, editor, textgen. Definice MPS language je organizovana takto --- pres AST nodes, ktere maji nekolik aspectu (ruzne pohledy).}
+Some concepts and design patterns known from object oriented programming can be used in language definition for MPS --- inheritance between concepts, interfaces, delegation and so on.
+Most importantly, nodes can have child nodes and that is how the AST itself is created.
+
+\JB{Zmineno vyse, navrhuji odstranit} \todo{Zduraznit ze AST node (object) ma nekolik aspectu: structure, editor, textgen.
+Definice MPS language je organizovana takto --- pres AST nodes, ktere maji nekolik aspectu (ruzne pohledy).}
 
 \paragraph{Structure.}
-The most important aspect of any MPS language is structure. It must be created first for each AST node.
-It specifies core attributes such as name of the AST node, inheritance relationships, possible child nodes (types and cardinality), implemented interfaces, references to other nodes, other properties (fields of any type that can hold values), etc.
+\JB{Odstavec mirne upraven} The fundamental aspect of any MPS language is structure.
+It must be created first for each type of the intended AST node.
+Structure specifies core attributes such as name of the concept, inheritance relationships, possible child concepts (their types and cardinalities), the implemented interfaces, references to other nodes, other properties (fields of any type that can hold values), etc.
 In Figure~\ref{fig:if_statement_structure}, you can see the structure aspect definition for the if statement.
@@
 \paragraph{Editor.}
-Editor aspect is where the user defines what the projectional editor representation of a code fragment (an AST node) looks like.
-The editor aspect of an AST node defines its visual representation (of the code) in the projectional editor.
-The user usually incorporates all children, references, and properties inside the representation so that future user of the language can insert their value into these placeholders.
-All elements can be styled using a~language similar to CSS.
+\JB{Odstavec mirne upraven}The editor aspect is where the user defines what the projectional editor representation of a code fragment (an AST node) looks like on the screen and how the user interacts with the code.
+JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells.
+The author usually incorporates all concept's children, references, and properties inside the representation so that the future users of the language can insert the values that the node expects.
+Additionally, all visual cells of the editor can be styled using a~language similar to CSS.
 
@@
 
+\JB{Nasledujici odstavec navrhuji odstranit}
 \todo{From thesis.
@@
 \todo{Introduce the base language here.}
-A program written in an MPS language cannot be executed directly.
+A program written in MPS language cannot be executed directly.
 MPS uses an approach based on translation into a program in the so-called BaseLanguage.
@@
 \paragraph{Other aspects.}
-MPS supports also other aspects, such as type checking and data-flow analysis, but they are not important for our contribution.
-Moreover, information needed for their automated generation is not available/contained inside a grammar file.
+MPS supports also other aspects, such as type checking, model transformations, static and data-flow code analysis, refactorings, language migrations, etc., but these are not important for our contribution.
+Moreover, the information needed for their automated generation is not available/contained inside a grammar file.
 

### 2016-09-21 paper.tex
Final edits in chapter 2 by JB

@@
 \todo{Introduce the base language here.}
-A program written in MPS language cannot be executed directly.
-MPS uses an approach based on translation into a program in the so-called BaseLanguage.
-A plain-text representation of a BaseLanguage program is then passed to a system compiler and the result is executed.
-The module responsible for generating the program in the BaseLanguage is called "textgen" (we describe it briefly in the next paragraph).
-
-BaseLanguage is also an MPS language --- it is the official port of Java for MPS (created by JetBrains, authors of MPS).
-The default base language of MPS is Java, but others are supported too.
-
-Since BaseLanguage is an MPS language, BaseLanguage code is not a (plain) text source code, but again, an AST built out of nodes belonging to the BaseLanguage language.
-This means that generating code inside MPS consists of generating some AST out of BaseLanguage nodes.
-
-Besides representing the default (most common) target of text generation, BaseLanguage is used also for all additional programming inside MPS.
+\JB{Odstavec prepsan}
+In order to implement MPS itself and also to support the basic set of language-definition DSLs, BaseLanguage was developed in the early days of MPS.
+BaseLanguage is a clone of Java implemented using the MPS constructs.
+Although BaseLanguage is syntactically almost identical to Java, it is edited in a projectional editor, just like all languages in MPS.
+The DSLs that language authors use in order to define languages are generated into BaseLanguage.
+Similarly, all customer DSLs that are meant to be generated into Java choose BaseLanguage as their generation target and the conversion to textual Java will be handled by BaseLanguage without any further effort, since BaseLanguage has a TextGen aspect defined, which translates BaseLanguage code into textual Java sources.
 
 \paragraph{TextGen.}
-The TextGen aspect defines how a certain AST node will be translated into a text source code (in the base language).
-This will allow us to generate source code in text form out of an AST model and ultimately to generate a text program out of the code user writes in the (imported) MPS language.
-The TextGen aspect is basically just one method/procedure definition for each AST node in the language. The method produces string representation of the given AST node and its children.
-MPS then calls this method for the root AST node of the program.
+\JB{Odstavec upraven} The TextGen aspect defines how a certain AST node will be translated into text.
+The TextGen aspect is typically only needed for the bottom-line (base) languages.
+DSLs, on the other hand, need to define rules for model-to-model conversion (a generator), since these are rarely converted to text directly.
+After TextGen has generated textual sources from an AST, a compiler for the particular GPL gets invoke in order to compile the generate sources into binary code.
+The TextGen definition follows a very straightforward pattern - each node outputs its textual representation into a textual buffer, calling TextGen of its children nodes at the right moments.
+MPS then calls this method for the root AST nodes of the given program.
 TextGen aspect has to be defined (especially its more complex functionality) using the BaseLanguage.
@@
 
+\JB{Navrhuji odebrat}
 \todo{From thesis.

### 2016-09-23 chapters/background.tex
Split paper into chapter files

--- 2016-09-21 paper.tex
@@
-\documentclass[10pt]{sigplanconf}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[noline]{algorithm2e}
-\usepackage{comment}
-\usepackage{color}
-\usepackage{times}
-\usepackage{multirow}
-\usepackage{balance}
-\usepackage{adjustbox}
-\usepackage{alltt}
-\usepackage{textcomp}
-\usepackage[usenames]{xcolor}
-
-\setcounter{tocdepth}{3}
-
-% configures style of source code listings
-\lstset{
-	basicstyle=\small\sffamily
-	%basicstyle=\scriptsize\ttfamily
-	%numbers=left
-}
-
-\lstdefinelanguage{java}{
-	morekeywords={abstract,continue,for,new,switch,assert,default,goto,package,synchronized,boolean,stream,event,filter,when,do,if,private,this,break,double,implements,protected,throw,byte,else,import,public,throws,case,enum,instanceof,return,transient,catch,extends,int,short,try,char,final,interface,static,void,class,finally,long,strictfp,volatile,const,float,native,super,while},
-	sensitive=false,
-	morestring=[b]",
-	morestring=[b]',
-	morecomment=[l]{//},
-	morecomment=[s]{/*}{*/)},
-	escapeinside=??,
-	moredelim=[is][\textit]{__}{__},
-	moredelim=[is][\textbf]{_*}{*_}
-}
-\lstnewenvironment{displayjava}
-	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b,xleftmargin=1em,xrightmargin=1em}}{}
-\lstnewenvironment{java}
-	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b}}{}
-
-\newcommand{\code}[1]{{\sf \small #1}\xspace}
-
-\newcommand{\mi}[1]{\mathit{#1}}
-\newcommand{\mr}[1]{\mathrm{#1}}
-\newcommand{\mt}[1]{\mathtt{#1}}
-
-\newcommand{\todo}[1]{{\bfseries #1}}
-
-% for writing comments
-\newcommand{\PP}[1]{\textcolor{blue}{\bfseries PP: #1}} % Pavel Parizek
-\newcommand{\PV}[1]{\textcolor{red}{\bfseries PV: #1}} % Premek Vysoky
-\newcommand{\JB}[1]{\textcolor{orange}{\bfseries JB: #1}} % JetBrains
-
-% colors for the ANTLR environment
-\definecolor{antlrbgcolor}{rgb}{.962, .962, .962}
-\definecolor{antlrparserrulecolor}{RGB}{11,83,148}
-\definecolor{antlrlexerrulecolor}{RGB}{191,144,0}
-\definecolor{antlrliteralcolor}{RGB}{133,32,12}
-\definecolor{antlrregexcolor}{RGB}{56,118,29}
-
-% colors for the MPS environment
-\definecolor{mpsbgcolor}{rgb}{.962, .962, .962}
-\definecolor{mpsconceptcolor}{RGB}{42, 140, 0}
-\definecolor{mpsinterfacecolor}{RGB}{170,0,229}
-
-% coloring commands
-\newcommand{\antlrap}{\textquotesingle}
-\newcommand{\antlrparserrule}[1]{\textcolor{antlrparserrulecolor}{#1}}
-\newcommand{\antlrlexerrule}[1]{\textcolor{antlrlexerrulecolor}{#1}}
-\newcommand{\antlrliteralnoap}[1]{\textcolor{antlrliteralcolor}{#1}}
-\newcommand{\antlrliteral}[1]{\antlrap\antlrliteralnoap{#1}\antlrap}
-\newcommand{\antlrregex}[1]{\textcolor{antlrregexcolor}{#1}}
-% TODO mozna zmenit "mpsconcept" na "mpslangelem"
-\newcommand{\mpsconcept}[1]{\textcolor{mpsconceptcolor}{#1}}
-\newcommand{\mpsinterface}[1]{\textcolor{mpsinterfacecolor}{#1}}
-
-% TODO zmenit barvu pozadi na bilou
-% environment for the ANTLR code (grammars)
-\newenvironment{antlr}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=antlrbgcolor,frame,minipage=394pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
-
-% environment for the MPS code
-\newenvironment{mps}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=mpsbgcolor,frame,minipage=394pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
-
-\begin{document}
-
-% TODO
-%\conferenceinfo{OOPSLA'12,} {October 19--26, 2012, Tuscon, Arizona, USA.}
-%\CopyrightYear{2012}
-%\copyrightdata{978-1-4503-1561-6/12/10}
-
-% TODO
-\title{\todo{INGRID: Creating Languages in MPS by Importing ANTLR Grammars}}
-% alternative: "INGRID: Importing Grammars Into MPS"
-
-\authorinfo{P\v{r}emysl Vysok\'y \and Pavel Par\'izek}
-	{Charles University, Faculty of Mathematics and Physics, Prague, Czech Republic}
-	{premek.vysoky@gmail.com, parizek@d3s.mff.cuni.cz}
-
-% TODO pridat nekoho dalsiho z JetBrains
-% TODO upravit affiliation pro JetBrains -> doplnit "s.r.o" nebo radsi "Inc."
-% TODO muzeme taky napsat jejich emaily pokud budou lidi z JetBrains souhlasit
-\authorinfo{V\'aclav Pech}
-	{JetBrains}
-
-\maketitle
-
-\begin{abstract}
-% TODO
-\todo{write later}
-
-\todo{Copy from master thesis.}
-JetBrains MPS is a language workbench focusing on domain-specific languages.
-Unlike many other language workbenches and IDEs, it uses a projectional editor for code.
-The developer directly manipulates the program in its tree form (AST) and not by editing a text source code.
-This brings many advantages, but on the other hand requires time-consuming and complicated MPS language definition.
-The thesis elaborates on the possibility of automating the process of creating MPS language definition from its grammar description.
-It introduces the MPS editor, evaluates approaches of related projects and describes author's efforts to implement an MPS plugin that allows this import.
-The chosen approach and the selection of tools used for implementation are justified in the thesis.
-We point out important problems that any similar project might deal with and we introduce some possible solutions.
-Furthermore, the thesis contains examples of imported languages, showing the potency of the chosen approach.
-The thesis also aims to lay groundwork for future extensions and suggest possible improvements.
-\end{abstract}
-
-% TODO possible keywords: JetBrains MPS, programming language, grammar, projectional editor
-
-\section{Introduction}
-
-\JB{Odstavec mirne upraven} The prevailing approach to define valid syntaxes for programming languages is through grammars written typically in notations based on (or similar to) EBNF [reference].
-Many popular tools exist for automated generation of parsers from the grammar definitions, such as ANTLR and Bison/Flex.
-The source code of programs in such languages is written in a text editors, usually embedded into an IDE.
-
-\JB{Odstavec mirne upraven} Projectional (or structural) editing offers an alternative approach.
-While the idea emerged as early as in 1970s, it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
-A recent revival of projectional editing has been observed in the area of language workbenches - IDE-like tools that enable the developers to manipulate the actual language definition.
-The Meta-programming System (MPS) [reference] by JetBrains is an open-source language workbench that focuses of Domain Specific Languages and leverages the concept of projectional editor.
-MPS allows developers to design custom languages and then use use them to write code and compile it into executables.
-MPS provides the whole IDE infrastructure for such languages.
-
-\JB{Odstavec mirne upraven} The key idea behind projectional editors is that a developer directly manipulates the AST representation of the source code instead of plain text.
-To avoid the fate of the early projectional editors MPS has put extra effort into making text manipulation feel like editing real text.
-MPS models, although being persisted as XML or binary files, can be versioned using the industry standard source-control systems (Git, Subversion).
-On-line model browsers have been incorporated into some of the popular public source code repositories (BitBucket) [reference] .
-
-\JB{Odstavec mirne upraven} The absence of parsers brings about numerous benefits, which are in particular useful for DSLs, which are frequently used by domain experts, who may not be professional programmers.
-\begin{itemize}
-	\item Since a projectional editors controls the interaction between the user and the code, the developers practically cannot enter syntactically invalid code.
-	\item Languages can be defined in a modular way and multiple languages can be easily combined together or extend one another, while parsers seriously limit language modularity for traditional parsed languages.
-	\item The languages may offer contextual or non-parseable notations, such as tables, diagrams or positional syntax.
-	\item Since the projection is detached from the physical representation of code (AST), authors can define multiple notations and allow the developers to switch between different representations of the code.
-\end{itemize}
-A motivating illustration of the abilities of MPS is the mbeddr project [citation], which extends C with numerous domain-specific constructs and data types.
-We provide more details about MPS in Section 2.1.
-
-\JB{Navrhuji smazat nasledujici odstavec}
-\todo{From thesis.
-	One of these tools is the JetBrains MPS~\cite{ref:MPS}, which approaches languages from a~different point of view.
-	MPS doesn't use the textual representation of code as usual, but rather works with the actual abstract syntax tree that holds the code's structure.
-	This has some positive implications as it gives us a~very powerful tool rich on features but also introduces some new problems we haven't encountered before.
-	One of them being the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-	MPS also becomes a~very powerful tool once the language has already been created and you start using it inside MPS.
-	There is a~huge variety of new languages constructed for MPS that usually solve domain specific problems.
-}
-
-\todo{Pouzivat jen terminy "AST nodes" a "language elements" (to spis). Nepouzivat pojem "concepts" abychom ctenare nematli.}
-
-\JB{Odstavec upraven} Projectional editing introduces some new problems we haven't encountered before.
-One of them is the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-In order to use a language inside MPS, it has to be defined through the MPS infrastructure.
-In particular, one has to create the abstract syntax called Structure (the types of allowed AST nodes), concrete syntax called Editor (projection of the AST on the screen and its interactions with the user), and text generation (to enable creation of plain text files used as input for compilers).
-We provide more details in Section X (Background about MPS).
-
-\JB{Odstavec pridan pro vysvetleni motivace prace} MPS is typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
-These DSLs are then transformed on the AST level by a series of model-to-model transformations to an AST model representing the desired semantics in a General-purpose language (GPL), such as C or Java.
-This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
-This implies that the target GPL must also be defined by the MPS means, to allow the DSLs to have their models transformed into a model in the GPL.
-
-\JB{Odstavec doplnen} However, it requires substantial effort to create a definition of a GPL in MPS.
-Therefore, very few mainstream GPLs are now fully supported by MPS.
-Only Java and C have been implemented to date, which limits the DSLs authors to only target these supported target GPLs.
-We believe that an automated process of importing grammars of GPLs into MPS would encourage more GPLs to be migrated to MPS and thus giving the DSL authors more options.
-
-\todo{From thesis.
-	However, building these languages is a~complicated and time-consuming process and a~lot of effort must be put into them before they are ready for use.
-
-	These examples show that recreating a~full language inside MPS is not an easy task and a~lot of time must be spent on implementing all aspects of the language.
-	A big part of this effort is, however, quite straightforward and can be possibly automated, as we show in this paper. Automation speeds up the process of adoption of new languages to a great degree.
-}
-
-\JB{Odstavec upraven} Since a big part of the language definition process in MPS is quite straightforward and can possibly be automated, we will look into the possibility of automatic import of already existing general purpose languages into MPS using a grammar description of their syntax.
-
-We present INGRID --- a method and a tool for semi-automated construction of MPS definition of programming languages based on their ANTLR grammars.
-We also discuss challenges that we encountered during this process and our solution to them.
-Most of these challenges are general, i.e.
-not MPS specific, as they apply to the relation between grammars defined in EBNF-like style (rules and tokens, everything defined in plain text) and structured (object-oriented) approach used in MPS (and other language workbenches).
-
-\todo{Hodne zduraznit automation, tedy ze chceme maximalizovat automatizaci.}
-The main goal of our/the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible.
-
-\todo{From thesis.
-	Automatic import of already existing languages into MPS from the grammar description of their syntax.
-}
-\todo{Mozna prodat INGRID jako automatickou konverzi gramatik do podoby MPS languages, tedy jako nastroj ktery umoznuje portaci jazyku do prostredi MPS a jejich nasledne pouziti v MPS (treba kombinace s dalsimi jazyky).}
-
-Our primary motivation is to enable better support for mainstream and/or general purpose programming languages (XML, C, C++, JavaScript, Python) in MPS, especially to enable writing programs in these languages using the projectional editor and to combine them with domain-specific languages natively created in MPS.
-\todo{Zduraznit take pouziti importovanych jazyku pro cile generovani z DSL --- to plati zejmena u mainstream general-purpose languages jako treba JavaScript.}
-
-\todo{Briefly mention the biggest challenges.}
-The are related to the way grammar encoding/description of languages are usually written.
-They do not hold any information about the code layout, and contain many rules that do not directly correspond to AST nodes (programming language constructs).
-We discuss them in more detail later (Section X).
-
-\todo{From thesis.
-	INGRID allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
-	The biggest benefit is the possibility to code in these mainstream general purpose languages using the projectional editor.
-}
-
-However, we do not target just the very complex general purpose languages such as JavaScript.
-INGRID can be used also for simple languages.
-We believe that, for simple languages, it is less time consuming to create the grammar by hand (e.g., using the ANLTR notation and toolset) and then create the language definition in MPS by using the INGRID tool, than to create it manually using the MPS GUI --- this is another benefit of INGRID.
-
-Structure of the paper.
-In the beginning, we will describe the MPS editor and related/similar projects.
-Then, we describe our approach to creating language definition in MPS from their grammars, and discuss the major challenges together with our solution.
-We will also show some example languages imported with INGRID and discuss our experience.
-
 \section{Background and Related Work}
@@
 \todo{Popis na zaklade emailove diskuze (Nizozemci).}
-
-\section{Creating Language Definition in MPS}
-
-From all the available notations in which grammars can be written, we decided to use the ANTLRv4 grammar notation~\cite{ANTLR4reference} as the input format (in which the input programming languages have to be defined) for the following reasons.
-\begin{itemize}
-	\item It is a very widely used grammar notation.
-	\item All popular and widely used programming languages have their syntax captured (represented) using this notation --- see the repository~\cite{ref:ANTLR4grammars}.
-	\item There is a whole framework/toolset for generating parsers and editing grammars built around it~\cite{ref:ANTLR4}.
-	\item It supports Java (it is written in Java), which is the implementation language of MPS and also the default BaseLanguage.
-	\item It uses (is based on) an EBNF (Extended Backus-Naur Form) \footnote{https://en.wikipedia.org/wiki/Extended{\_}Backus\%E2\%80\%93Naur{\_}Form} form that it useful because it meets/respects the object oriented design that is used when describing MPS aspects (definitions of MPS languages).
-	\item It has its own syntax captured in its own syntax\footnote{https://github.com/antlr/grammars-v4/tree/master/antlr4} (there is an ANTLRv4 grammar for the ANTLRv4 notation).
-This is a very important feature as it allows us to create a parser for ANTLRv4 grammars so that we can parse grammars of other languages.
-\end{itemize}
-
-We use ANTLRv4 grammars as inputs (and work with them).
-	https://github.com/antlr/antlr4/blob/master/doc/index.md
-	https://github.com/antlr/grammars-v4
-
-Unlike some of the related projects, we did not add any new features into the grammar notation.
-We tried to import grammars as they are.
-
-The generation/import process performed by INGRID consists of three phases (import of structure, usable editor, text generation).
-We will talk about all phases of the generation/import process, describing encountered challenges/obstacles using an example language.
-
-To fulfill our goal of maximal possible automation, we import/generate the full language structure and then do a lot of mundane and time-consuming work connected to creating projectional editors.
-
-Now we describe/introduce the main steps of our approach (that INGRID tool performs): parsing the grammar, creating the structure, editor, and textgen.
-We are able to create/generate the structure aspect and then very basic editor and some default/basic textgen.
-In particular, we did not port the ANTLR notation as an MPS language.
-In the following subsections, we provide some details about our approach to parsing the input grammar and creating the main aspects of each element (each type of AST node) of an MPS language (structure, editor, and textgen).
-
-During our work on this project, we also discovered/encountered challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component), whitespace (editor aspect).
-There is a challenge/issue that we have discovered during our work on this project (in the process of our own implementation), and that is related to construction of the structure aspect.
-Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
-However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
-We describe/discuss these challenges in more detail (also using examples) and our solutions to the challenges inside the respective following subsections.
-
-Result of the construction of a MPS language from the input grammar (import) has to be polished by the human user.
-For example, the aspects not supported by INGRID will be used for adjusting the imported language and improving the usability of the language.
-
-\todo{Briefly discuss how our overall approach differs from similar projects (PE4MPS, etc).}
-There are more ways how to approach the problem of a grammar import.
-
-Our overall approach differs from the PE4MPS project (Section X.Y) especially in the level of automation.
-While the main goal of (our) the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible, usage of PE does not really spare the user any work because it only shifts the manual work from creating projectional editor to writing PE grammars.
-What the user would do inside the MPS UI using the projectional editor is in the case of PE4MPS done using a text editor when editing PE grammars.
-MPS UI is specially designed/made/tailored for this purpose, guides the user through this process, and supports aids such as auto-completion.
-In particular, MPS UI allows the user to perform only changes that result in valid output.
-Editing the grammar stored in a text file using a standard editor is an inferior approach with respect to using MPS (projectional editor), because a text editor is not as user-friendly, and, more importantly, it is an error prone environment.
-
-Main drawback of ANTLR{\_}MPS, and main difference from our approach, is very limited automation: it does not create the parent-child relationships between AST nodes in the structure, and does not generate any editor and textgen aspects (for any node) at all.
-In addition, there is no attention paid to the problem of generating projectional editor and text generation is neglected too.
-We blame all this on the project being in its early stage of development.
-
-In the rest of this section, first we define the language that we will use throughout the paper as the running example.
-Then we will talk about parsing grammar files \todo{Tohle je mozna zbytecne}.
-Lastly, we will describe our approach to tackling generation of language's aspects (structure, editor, and TextGen).
-At some places, we discuss important possible alternatives (as there are several possible approaches).
-
-\subsection{Running Example}
-
-We illustrate the main steps on the example of a simplified XML language, which is defined below (in Figure XY).
-
-\todo{Pouzit jazyk SimpleXML na running exaple.}
-
-We decided to use SimpleXML because it is small enough to fit here, but still can be used to illustrate the main challenges and our solution
-
-Throughout the whole paper, we will be using one example language to illustrate main challenges and our solution.
-The language is a simplified version of XML. We call it \emph{SimpleXML}.
-We created the grammar of the language by taking the official XML ANTLRv4 grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and stripped it down to eliminate/avoid some not so interesting (advanced) features such as XML entities.
-Below in Figure X, you can find the grammar of the SimpleXML language in the ANTLRv4 notation.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\textbf{grammar} \textit{SimpleXML};
-	\antlrparserrule{document}     :   \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
-	\antlrparserrule{prolog}       :   \antlrliteral{<?xml } \antlrparserrule{attribute}* \antlrliteral{?>} ;
-	\antlrparserrule{comment}      :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-	             |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-	             ;
-	\antlrparserrule{attribute}    :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
-	\antlrparserrule{content}      :   \antlrlexerrule{TEXT}
-	             |   \antlrparserrule{element}
-	             |   \antlrparserrule{comment}
-	             |   \antlrlexerrule{CDATA}
-	             ;
-	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-	\textbf{fragment}
-	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
-	             |   \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.}
-	             |   \antlrlexerrule{DIGIT}
-	             ;
-	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
-	\antlrlexerrule{TEXT}         :   \antlrregex{~[<"]*} ;
-	\antlrlexerrule{CDATA}        :   \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
-\end{antlr}
-\caption{SimpleXML grammar}
-\end{figure*}
-
-We have colored the grammar so it is easier to find the way around it (to make navigation easier).
-The grammar contains following basic elements:
-
-\begin{itemize}
-	\item \textbf{ANTLRv4 keywords} -- In black, they are required by the specification.
-	\item \antlrparserrule{\textbf{Parser rules}} -- Stated in blue, they are the rules describing language's structure.
-		The alternatives (that the element on the left side may break into) are separated by the pipe ($|$) character.
-	\item \antlrlexerrule{\textbf{Lexer rules}} -- In yellow, they are rules describing terminal symbols that the parser is matching against the input.
-		Terminals are encoded as string values and regular expressions.
-		The \textbf{fragment} keyword states that the lexer sub-rule is just a helper rule that brings more clarity into the notation but is not visible in the parser output.
-	\item \antlrliteralnoap{\textbf{String values (literals)}} -- In red, always inside of a pair of apostrophes, they are similar to lexer rules.
-		They describe a terminal string, not using regular expressions but by an exact string match.
-	\item \antlrregex{\textbf{Regular expressions}} -- Colored in green, they describe a string token to be matched using a regular expression with the special ANTLRv4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
-	\item \textbf{Rule operators (?+*)} - Left in black, following any element, they quantify the number of occurrences the prepending element can appear in.
-		These operators can also be prepended by an asterisk (*), meaning they are non-greedy.
-\end{itemize}
-
-\subsection{Parsing (the) Grammar}
-
-\todo{Define some name for the AST representing the grammar, and use it consistently throughout the whole paper. Maybe distinguish between "grammar AST" and "MPS structure AST".}
-The first step (stage) of creating the MPS language (of the import process) is parsing the grammar file (written using the ANTLRv4 notation) so that we can get an AST tree representing the grammar.
-
-The abstract syntax tree that comes out of the automatically generated (ANTLR) parser is a little bit complicated and full of information not relevant for our purposes (to us).
-That is why we translate the complex ANTLR AST into our own simplified tree made out of our own objects.
-We discarded as much irrelevant information as possible and only kept information (data) vital/necessary for construction of MPS languages.
-
-Furthermore, we process the grammar in two iterations.
-In the first iteration/pass, we build a tree holding names of other referenced rules in the form of strings.
-In the second iteration, we walk this tree and resolve all these references to real pointers to other rule objects.
-We end up with a clear and simpler representation that is easy to process in the next steps of the import process.
-
-We also had to post-process (simplify) the representation of lexer rules (tokens).
-As a result of the initial parsing, ANTLR parser produces some kind of a tree structure that is representing the lexer rule (meaning/note that lexer rules can also be built from alternatives just like parser rules).
-However, lexer rules are, in fact (basically), just regular expressions used for parsing input source code into tokens.
-We took each/every such tree that represents the lexer rule and flattened it into the equivalent regular expression.
-
-For illustration, consider the lexer rule \antlrlexerrule{Name} from our/the SimpleXML language (Figure XY), together with all its sub-rules, shown in Figure~\ref{fig:lexerrule}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-	\textbf{fragment}
-	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
-             |   \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.}
-             |   \antlrlexerrule{DIGIT}
-             ;
-	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
-\end{antlr}
-\caption{Lexer rule from SimpleXML}
-\label{fig:lexerrule}
-\end{figure*}
-
-Intuitively, we can create a regular expression by gluing elements of each alternative together and then join these alternatives with a classic regex OR ($|$), which is exactly its function in the ANTLRv4 grammar notation.
-We have captures this idea in the form of the recursive algorithm in Figure~\ref{fig:algflatten}, which covers the whole process.
-
-% TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	Flatten(\textit{R}):
-	1) Define \textit{T} as a list of empty string
-	2) For each alternative \textit{A} of the rule \textit{R}:
-	3)     Define \textit{R} = \antlrap\antlrap
-	4)     For each element \textit{E} of \textit{A}:
-	5)         If \textit{E} is not yet flattened:
-	6)             Flatten(\textit{E})
-	7)         \textit{R}.append(\textit{E})
-	8)         \textit{R}.append(\textit{E}.operator)
-	9)     \textit{T}.add(\textit{R})
-	10) Build a string S out of elements t$_{\text{1}}$, t$_{\text{2}}$, ... t$_{\text{n}}$ of \textit{T}:
-	11)     (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
-	12) Return \textit{S}
-\end{antlr}
-\caption{Flatten algorithm}
-\label{fig:algflatten}
-\end{figure*}
-
-The output of \texttt{Flatten(Name)} (output of the Flatten algorithm applied to the \antlrlexerrule{Name} rule) is this regular expression (in this form):
-
-\begin{center}
-	\texttt{([:a-zA-Z])(([:a-zA-Z])|(-)|({\_})|(.)|([0-9]))*}
-\end{center}
-
-\subsection{Structure}
-
-Now, as the next step, we need to translate/transform the AST (the types of AST nodes) derived from the input grammar into elements of the structure aspect of the MPS language definition (into types of AST nodes), and linking the nodes appropriately.
-
-The goal of this second step is to translate (transfer) the syntax (structure) of the input language that is captured by the grammar (ANTLR), into the MPS structure aspect.
-The structure of the MPS language is somewhat similar to the grammar.
-This is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
-The process of transferring grammar rules into the structure of a MPS language presents/involves some interesting challenges.
-The input grammar (EBNF, ANTLR) serves a little bit different purpose than MPS language, and the way grammar rules are usually written down causes some problems that we had to overcome in order to recreate the language inside MPS.
-Note that the challenges and problems are not specific to MPS --- they are common for everyone tackling a similar problem.
-
-We are going to describe two different approaches that we designed --- the straightforward approach (Section~\ref{sect:straightforward_approach}) and the shortcut approach (Section~\ref{sect:shortcut_approach}).
-We will describe two possible approaches that we developed in Section X and Section Y, respectively.
-\todo{Rict neco jako ze ten druhy, slozitejsi, approach vlastne rozsiruje ten prvni, jednodussi.}
-
-However, before focusing on the two approaches, we must describe the basic principles of translation from the AST representing the grammar to the MPS structure, which are common to both approaches.
-The structure is created according to object-oriented principles.
-Each node (element) is represented by an object (and can be treated as such) that may have a parent node, child nodes, and properties of any data type (similar to fields), it may implement any number of interfaces, and it may also hold references to other nodes.
-For every rule of the grammar (for every node in the grammar AST) that breaks into more/multiple alternatives, we must create a node in the MPS structure AST for each alternative of the grammar (every child node).
-Names of the nodes for alternatives are composed from the name of the rule and a number indicating the order of appearance in the grammar.
-For example, considering the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlgrammarelement}
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-             \textcolor{gray}{|   \antlrap<\antlrap Name attribute* \antlrap/>\antlrap}
-             \textcolor{gray}{;}
-\end{antlr}
-\caption{Fragment of the SimpleXML grammar}
-\label{fig:xmlgrammarelement}
-\end{figure*}
-
-Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.75]{./images/element_node_common.png}
-	\caption{Resulting Element{\_}1 node}
-	\label{fig:element_node_common}
-\end{figure}
-
-String literals (e.g., keywords such as \textbf{for}) are omitted/skipped because they will appear only in the corresponding projectional editor.
-The node named \mpsconcept{Element{\_}1} contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name} inside the first alternative of the \antlrparserrule{element} rule,
-Value of the property will be restricted using the regular expression representing the \antlrlexerrule{Name} rule.
-We explain later how the references to parser rules (\antlrparserrule{attribute} and \antlrparserrule{content}) as this is where the approaches differ.
-The main/key difference between these two lies in the way we create children fields for parser rules and in the way we are going to link them together using interface concepts.
-
-\subsubsection{The Straightforward Approach}
-\label{sect:straightforward_approach}
-
-The key idea of this approach is that, for each rule with more than one alternative, an interface node is created.
-Consider the \antlrparserrule{content} rule from our SimpleXML language (Figure~\ref{fig:contentrule}).
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
-\end{antlr}
-\caption{Content rule}
-\label{fig:contentrule}
-\end{figure*}
-
-Any of the four nodes corresponding to alternatives could appear anywhere the \antlrparserrule{content} rule is referenced.
-So we will create one interface node, and then for each alternative of this rule we create one node that implements the interface.
-The resulting fragment of structure will look like the one in Figure~\ref{fig:icontentitf}.
-
-\begin{figure*}[ht]
-\centering
-\begin{mps}
-	\mpsinterface{IContent}   :   \mpsconcept{Content{\_}1}
-           |   \mpsconcept{Content{\_}2}
-           |   \mpsconcept{Content{\_}3}
-           |   \mpsconcept{Content{\_}4}
-\end{mps}
-\caption{IContent interface}
-\label{fig:icontentitf}
-\end{figure*}
-
-References to other parser rules are captured by links to children (interfaces or object nodes) in the AST nodes.
-For the \antlrparserrule{element} rule that is referencing the \antlrparserrule{content} rule, defined as in Figure~\ref{fig:elementrule}, we get the AST node representing the first alternative that is shown in Figure~\ref{fig:element_node_full}.
-There is one \mpsinterface{IElement} interface node that is implemented by two object nodes \mpsconcept{Element{\_}1}, \mpsconcept{Element{\_}2}.
-The node \mpsconcept{Element{\_}1} has two child links pointing to \mpsinterface{IAttribute} and \mpsinterface{IContent}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-\end{antlr}
-\caption{Element rule}
-\label{fig:elementrule}
-\end{figure*}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.75]{./images/element_node_full.png}
-	\caption{Element{\_}1 node's structure aspect}
-	\label{fig:element_node_full}
-\end{figure}
-
-A problem with this approach, which we call \emph{the layer problem}, is related to the grammar rules (and corresponding AST nodes) that represent intermediary layers created only to make the grammar more readable.
-The intermediary layers are created by authors of the grammars in order to make them more readable and more easily (better) maintainable.
-We illustrate this problem (and explain it) on the behavior of auto-completion in MPS.
-Imagine there is a freshly inserted \mpsconcept{Element{\_}1} node (representing the full XML element as is stated in the first alternative of the \antlrparserrule{element} parser rule).
-Now we would like to insert another nested XML element inside.
-MPS auto-completion mechanism would give us four options that are captured in the left part of Figure~\ref{fig:layer_problem}.
-This is because there exist four nodes (objects) that implement the \mpsinterface{IContent} interface as per each alternative of the rule.
-
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/layer_problem.png}
-	\caption{Layer problem in auto-completion}
-	\label{fig:layer_problem}
-\end{figure*}
-
-However, in order to correctly insert another nested element inside, a user would have to first insert the \mpsconcept{Content{\_}2} node inside \mpsconcept{Element{\_}1} that has an \mpsinterface{IElement} child inside (and nothing else).
-Then, as the second step, a user would invoke/trigger/call for the auto-complete again and insert either \mpsconcept{Element{\_}1} or \mpsconcept{Element{\_}2} inside \mpsconcept{Content{\_}2}.
-This mean that a user would have to go through two steps, and in the first stepe either correctly guess (or remember the grammar rule's alternative order), which option/item offered by the auto-complete menu to select.
-
-Similarly, if one would decide to replace the nested \mpsconcept{Element{\_}1} node with, let's say, an XML comment (a \mpsconcept{Comment} node), we need to delete both intermediary layers before we get back to the original \mpsconcept{Content{\_}X} selection (crossroads).
-The user cannot really see what is happening, since the intermediary level has no appearance or indicator.
-This may lead to confusion on the user's part.
-
-The layer problem is solved/addressed by the second approach, which we describe in the next subsection.
-
-\subsubsection{The Shortcut Approach}
-\label{sect:shortcut_approach}
-
-The key idea of this approach is to identify rules and nodes that are at the end of the (derivation, inheritance) chain, and offer them directly to the user through the auto-completion menu.
-We call such \emph{end rules} and \emph{end nodes}, respectively, because they cannot transparently break into more rules/nodes.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
-
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-
-	\antlrparserrule{comment}    :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-\end{antlr}
-\caption{Content parser rule with children}
-\label{fig:contentrulewithchildren}
-\end{figure*}
-
-For example, the \antlrparserrule{content} rule, given in Figure~\ref{fig:contentrulewithchildren} together with its child rules, can ultimately expand into following nodes:
-
-\begin{itemize}
-	\itemsep0em
-	\item \textbf{Content{\_}1} (TEXT)
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}1}
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}2}
-	\item Content{\_}3 $\rightarrow$ \textbf{Comment}
-	\item \textbf{Content{\_}4} (CDATA)
-\end{itemize}
-
-The end nodes are highlighted using the bold font.
-
-Figure XY shows the algorithm (in pseudocode) that finds all end rules for a given parser rule.
-
-% TODO pouzit standardni environment pro algoritmy 
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	FindPathsToEndNodes($R$):
-	1)  Define $L$ as an empty list of list of nodes
-	2)  Return FindPathsToEndNodes($R$, $L$)
-
-	FindPathsToEndNodes($R$, $L$):
-	3)  Define $Q$ as list of list of nodes
-	4)  For each alternative $A$ of rule $R$:
-	5)      $L1$ = Clone($L$)
-	6)      If $A$ is a parser rule with only one element $E$:
-	7)          Let $I$ be interface/concept representing rule $E$
-	8)          $L1$.Add($I$)
-	9)          $P$ = FindPathsToEndNodes($E$, $L1$)
-	10)         $Q$ = Merge($Q$, $P$)
-	11)     Else
-	12)         $L1$.Add($R$)
-	13)         Let $P$ be concept representing $A$
-	14)         $L1$.Add($P$)
-	15)         $Q$.Add($L1$)
-	16) Return $Q$
-\end{antlr}
-\caption{Algorithm for finding end rules}
-\label{fig:shortcut_algorithm}
-\end{figure*}
-
-To find end rules for each parser rule, the algorithm recursively scans through the parser tree that was built before.
-For each parser rule, it tries to find paths leading to some end rule through its alternatives:
-\begin{itemize}
-	\item Whenever it finds an alternative that contains only one element, and this element is a reference to another parser rule, it has found an intermediary level that can be transparently hidden from the user of the language.
-		The process (algorithm) continues by recursively processing alternatives of this "level" rule (since we are not at the end of the chain yet).
-	\item Otherwise, an end rule was found (recursion stops here).
-\end{itemize}
-
-By appending the rule that is leading to current element (line 12) and then appending that alternative's element itself (line 14), we get a path that contains the full path and the target end rule as the last element of the chain.
-The result of this algorithm, for example for the \antlrparserrule{content} rule, equals the listing of the five paths mentioned above.
-
-We do/repeat this for all parser rules and for each rule we get a list of paths that lead from that particular rule to an end node.
-We call these paths as \textbf{shortcuts}, because they provide a shortcut from the rule to the end of the chain.
-Now that we have the shortcuts, we will discuss several ways how to use the shortcuts (them) to improve the MPS language under construction (how to make the language better).
-
-The primary use of shortcuts is to generate options for the auto-completion menus such that only the end nodes are offered.
-This can be technically done in two ways --- (1) through custom auto-complete menus and deletion handlers or (2) by defining a special interface for each rule such that the interface is implemented only by end nodes for the rule.
-Shortcuts have to be considered when inserting nodes but also when deleting them, i.e.
-the whole chain (shortcut path) consisting of possibly multiple intermediary AST nodes (levels) up to the end node must be added, respectively deleted.
-The effect of insertion must be fully reversed when the user wants to delete some end node (the given end node).
-
-\subsection{Editor}
-
-When all types of AST nodes in the language are created, the next step is to define their visual representation in the projectional editor.
-As stated before (in Section X (about JetBrains MPS, part of background)), MPS uses a cellular system that allows placing node's properties and children into a table-like arrangement.
-Cells of different types are available --- for storing property values, references to child nodes, and keywords (string literals), and also cells that influence layout (e.g., indentation).
-
-The purpose (goals) of this (third) step is to create the editor aspect for each language element.
-It should project all attributes of the element --- name, properties, children, and so on --- using the respective cell types.
-
-The main challenge (problem) here is that an input grammar (ANTLR) does not provide/give any help/aid/information regarding the code layout.
-As we already said in the introduction, one of the biggest challenges/problems that arise here is that the grammar description of a language does not hold any information about the code layout.
-The structural rules in the grammar only tell us, what the syntax tree looks like and how the code is broken into nodes.
-In particular, the grammar says nothing about indentation, line breaks, and another formatting.
-
-We tackled this problem by designing an automated learning approach, which generates the layout using several heuristics.
-We present our solution to this problem. It involves some intermediary step whose purpose is to generate the necessary information.
-Our solution is semi-automated (using some heuristics) --- user's help in an interactive manner is needed for final touches (polishing).
-It works as follows.
-
-As input for the algorithm (learning approach), the user has to supply to a set of valid source files written in the imported language, together with the grammar file.
-These four steps/operations are then performed.
-\begin{enumerate}
-	\item Automated generation of an ANTLR parser for the input language using the ANTLR library.
-	\item Modify/extend the code of the generated parser in order to record some additional information.
-	\item Compile the parser into an executable form.
-	\item Use the parser to parse the supplied source code files and extract information about the layout of the code during this process.
-\end{enumerate}
-This way, the imported/created MPS language would inherit the code style/layout directly from the user, i.e. it would learn directly from the code.
-In step 2, the parser is modified such that it remembers/stores the line number and column for each parsed token.
-This is very efficient and it enables to create a nice layout.
-From the line number, the algorithm can easily determine whether there is a line break between some tokens (statements, language elements).
-The column number can be then used to detect indentation.
-However, this approach is still heuristic and may give imprecise (suboptimal) results --- mainly because it is not always possible to precisely match language elements (in the MPS structure) with all tokens that belong to it.
-Note also that quality of the extracted information always depends on the amount of the source code files provided (available, supplied) and on how representative the set of source code files actually is.
-When our algorithm cannot determine some custom layout, it just creates all the cells (content) of the editor aspect and arranges them in a single row.
-Figure~\ref{fig:editor_adjustment} XY shows the difference between the default single-row layout (in the upper left corner) and the fully customized layout (in the bottom right corner) for the \mpsconcept{Element{\_}1} node.
-
-% TODO smazat cervenou sipku z toho obrazku a mozna rozdelit na dve casti (vlozit tam caru): default single-row code layout, optimal fully customized layout
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/editor_adjustment.png}
-	\caption{Projectional editor adjustment of the Element{\_}1 node}
-	\label{fig:editor_adjustment}
-\end{figure*}
-
-In practice, however, the user still has to tweak the editor aspect --- to adjust the code layout --- because the heuristics are not 100\% reliable.
-But it takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language~\cite{ref:javascript} and manually adjusting all editors that needed it, all that in a less than hour time.
-More details are provided later in Section X (examples) --- there we describe this issue further.
-
-\subsection{Text Generation}
-
-The final/last, fourth step, is to create the TextGen aspect for each AST node.
-The purpose of the TextGen aspect is to define, for a language element (AST node), what code in the BaseLanguage will be generated out of it, once we want to compile and run our MPS code/program.
-This is needed to allow/enable users to generate plain-text source code representation out of the AST they built/created inside MPS.
-Again, a TextGen aspect must be created for each language element.
-Basically, it is a single method in the BaseLanguage (or in Java?) for each language element.
-The method generates a string (plain-text) representation of the element, appends the string to the output stream, and then calls TextGen methods on all children of the element.
-
-A very basic example of BaseLanguage code that can be used as the TextGen aspect of an \mpsconcept{Element{\_}1} node, which represents the full XML tag with content, is shown in Figure~\ref{fig:textgen_example}.
-It appends all literals, properties, and children in the same order as they appear in the grammar definition.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=90mm]{./images/textgen_example.png}
-	\caption{Example of a~TextGen aspect}
-	\label{fig:textgen_example}
-\end{figure}
-
-Like in the case of editor, we discovered some interesting challenges related to the code layout, whitespaces, and more (so on).
-We describe our solution to the challenges here in this section.
-
-The reader might have noticed that there is no whitespace handling inside of our SimpleXML grammar (Section XY, running example).
-Grammars in the ANTLRv4 notation are usually written in such a manner that whitespace characters are skipped (ignored).
-However/Nevertheless, TextGen aspect (generator) should/must know between which children there must be a whitespace and where it is forbidden (i.e., where to put some whitespace characters) so that the produced code is a valid one.
-For example, in the case of the \antlrparserrule{element} rule, TextGen must determine that there should be a space in between \antlrparserrule{attribute} and \antlrlexerrule{Name}, while it is not required between \antlrliteral{\textless} and \antlrlexerrule{Name}.
-In general, it is sufficient to put just a single space character at places/locations where there has to be some whitespace in order to make the code valid (syntactically correct).
-
-Layout of the generated code is also important. For example, the user may want to produce more readable code by adding indentation or line breaks.
-
-In fact, the problem of TextGen layout is quite similar to the one we have seen with the projectional editor, described in Section XY.
-Since we are dealing mostly with text-based languages, their representation in projectional editor must be almost the same as their expected plain-text representation (output).
-Therefore, one could use the same approach for editor and TextGen --- adding line breaks and indentation.
-Once we would have some information about the layout for generating better projectional editor, we could leverage the same information and use it to improve TextGen.
-
-Nevertheless, in our current approach, the editor aspects have to be adjusted manually, after the import is done, because for now it is the fastest and the most efficient way how to deal with the problem of customizing the editor to capture a nice layout.
-Therefore, we used a different approach for TextGen --- a really simple heuristics which gives surprisingly good results.
-In the first step, a very basic TextGen is created that inserts spaces in between every two tokens of the language element (AST node).
-Then, as the second step, spaces are eliminated (restricted) from positions where they are not really needed.
-We describe all the cases here:
-
-\begin{enumerate}
-	\item Whenever there is a non-alphabetical literal,	appearing as a plain string token defined in the grammar, that might get recognized by the parser safely without the need for a whitespace separator around, then we omit spaces around it.
-		For example, consider \textbf{'\textless'} in SimpleXML.
-	\item Another group are strings inserted by the user of the language, whose form is constrained by a regular expression.
-		For example, an XML tag's name belongs to this group.
-		They are usually values of properties defined in structure aspects.
-		A space must be preserved only when a neighboring literal may end with an alphabetical character. Otherwise, we can omit the space.
-		In practice, this will eliminate redundant spaces inside of quotes, next to semicolons, and around brackets, but, on the other hand, it will separate language keywords (function, var, in, etc.) from other content by keeping the space character in between them.
-	\item Spaces are omitted also in the case of child nodes that are not present (e.g., when they are optional), and in the case of empty lists of child nodes, so that spaces do not accumulate.
-	\item On the other hand, when two child elements are next to each other, we always insert space.
-	\item Sequences of children are separated with space or with a line break.
-\end{enumerate}
-These few simple heuristics work surprisingly well and give us some very nice results.
-We discuss our experience with XML, JSON, and JavaScript in Section XY (Evaluation).
-
-Figure~\ref{fig:textgen_final} shows an example of a TextGen aspect for the full SimpleXML element.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=95mm]{./images/textgen_final.png}
-	\caption{Generated TextGen aspect for the SimpleXML element}
-	\label{fig:textgen_final}
-\end{figure}
-
-The automatically generated TextGen code can be adjusted manually very easily, so that it generates nicely indented XML code.
-In the case of the example above (in Figure F), we only need to wrap the \mpsconcept{Content{\_}2} child with indentation and change the sequence separator to a new line character.
-The resulting adjusted aspect (code) is shown below in Figure~\ref{fig:textgen_adjusted}.
-
-\begin{figure}[t!]
-	\centering
-	\includegraphics[width=85mm]{./images/textgen_adjusted.png}
-	\caption{Adjusted indentation inside the TextGen aspect}
-	\label{fig:textgen_adjusted}
-\end{figure}
-
-\subsection{Remarks about Grammars}
-
-As we have already shown in previous sections, it is very easy to write a grammar in a way that will cause problems during import into MPS (or once it is imported).
-Problems might occur during the creation of any aspect.
-Here we will look at some general problems/issues that grammar import poses (might pose).
-We will talk about these problems with respect to the challenges/obstacles we tried to overcome and which we described in previous sections (dedicated to structure, editor, and texgen, respectively).
-We also show a few examples of additional possible complications.
-
-\paragraph{Adjusting grammars.}
-There are some cases, where altering the input grammar might yield far better MPS language.
-We will show two examples how the usability of the resulting MPS language can be improved.
-Let us look at the definition of an XML attribute in Figure~\ref{fig:xmlattribute}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
-	\antlrlexerrule{STRING}      :   \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
-	            |   \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'}
-	            ;
-\end{antlr}
-\caption{Definition of an XML attribute}
-\label{fig:xmlattribute}
-\end{figure*}
-
-The original XML grammar has quotes as a part of the value.
-For the resulting MPS language, it would mean that there would be a placeholder for the attribute value that would expect us to input the leading and trailing quote together with the value too each time.
-It would also be marked red (by the syntax checker) unless we enter both quotes inside the value since the regular expression checking for quotes will not match.
-The user might be confused by this and will not be able to tell why his string value is incorrect.
-
-In our SimpleXML language, we adjusted the grammar easily in the manner that we show in Figure~\ref{fig:xmladjustgrammar}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
-	            |   \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'}
-	            ;
-	\antlrlexerrule{TEXT1}       :   \antlrregex{~["]*} ;
-	\antlrlexerrule{TEXT2}       :   \antlrregex{~[']*} ;
-\end{antlr}
-\caption{Adjusted grammar of SimpleXML}
-\label{fig:xmladjustgrammar}
-\end{figure*}
-
-We turned quotes into literals, thus ensuring that they will only appear in the projectional editor as fixed constant cells.
-We will not have to encapsulate the value in them each time.
-The user will only have to choose, which attribute version he wants to use (single or double quotes).
-
-As the second example, we will show the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
-Every statement in JavaScript needs to be either followed by a semicolon, newline, file end or end of the block --- see the Figure~\ref{fig:javascriptstmt}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{eos}            : \antlrlexerrule{SemiColon}
-	               | \antlrlexerrule{EOF}
-	               | {lineTerminatorAhead()}?
-	               | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}?
-	               ;
-	\textcolor{gray}{// Example reference of the eos rule}
-	\antlrparserrule{breakStatement} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos}
-	               ;
-\end{antlr}
-\caption{Statements in JavaScript}
-\label{fig:javascriptstmt}
-\end{figure*}
-
-Because there are multiple options, our import plugin would create a placeholder at the end of every statement.
-Every language element representing a statement would have one child of the \mpsinterface{IEos} interface type.
-This placeholder needs to be manually filled in for each statement.
-
-Since the projectional editor has much bigger power over the form of the code, we (a user) might want to have each statement on a separate line.
-Since we can differentiate between statements on the AST level, we do not need an explicit separator between them.
-This means that we might want to simplify the language and leave the semicolon out, or leave it just as a constant fixed part of the projectional editor, but not as something the user must explicitly fill in.
-Then we can just put each statement on a separate line as it is usual for JavaScript code, but we do not need the semicolon anymore.
-This small adjustment is very quick when done inside the grammar.
-We just change the \antlrparserrule{eos} rule to the form shown in Figure~\ref{fig:eosrule}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{eos} : \antlrliteral{;} ;
-\end{antlr}
-\caption{eos parser rule}
-\label{fig:eosrule}
-\end{figure*}
-
-This way, we can change the grammar in a way that it will not describe the same ECMAScript language as before, but will definitely make our MPS language more usable.
-TextGen aspect may be defined in such a way that puts the semicolon back (after each statement) in the generated plain-text representation. 
-This would be very important especially when JavaScript is available as another base language (which is actually a plan (future work) of the JetBrains company).
-
-Above, we have shown that adjusting the input grammar might be a very fast mean of tweaking/improving the usability of generated MPS language, and sometimes it may be the only proper solution in some complex situations.
-In general, the question we can ask ourselves is whether one aims for a full precise/faithful port of the input language or for an MPS alternative.
-
-\todo{Nahradit slovo "break" v nasledujicich odstavcich (v kontextu "break the parser") za vhodnejsi slovo ktere pusobi vic odborne.}
-
-\paragraph{Breaking grammars and parsers.}
-However, we have found that there is one big problem with grammar adjustment that we would like to point out.
-The problem is that it is very easy to change the input ANTLR grammar in a way that breaks the ANTLR parser generated out of it.
-By breaking we mean that it stops parsing the original language.
-We illustrate the problem on our SimpleXML language.
-
-As stated before, when creating the SimpleXML grammar, we have started off with the original XML grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and did some adjustments to it.
-We ended up with a grammar that can be processed by our method/plugin and creates a nice usable MPS language.
-Nevertheless, we noticed that even though the imported language behaves well enough and mimics the XML language quite nicely, the ANTLR parser generated out of this grammar no longer parses XML successfully.
-Some changes we have made, such as the attribute adjustment, broke the grammar down.
-More precisely, our changes improved the MPS language while breaking down the parser, and we have not even noticed it because, from the perspective of MPS, the imported language still corresponds to XML.
-
-What we are trying to say (show here) is that it is very easy to perform a grammar adjustment that will, at first, seem harmless and valid inside MPS, but will break the parser.
-The problem is that the user performing this change probably will not be aware of breaking it.
-
-The main (underlying) cause of this problem is the way the ANTLR parser is implemented and the quite different purpose we are using the grammar for in our project/tool/method.
-There are many ways how various parsers deal with, for example, token matching.
-For instance, ANTLR introduces so called \emph{greedy} and \emph{non-greedy} operators.
-The greedy way, in which ANTLR matches input on defined tokens and prioritizes their selection, makes some rules very dangerous.
-When a rule matches a wide range of input, it might happen during parsing that it is prioritized over other rules and swallows a lot more input than the author of it intended.
-Usually, these dangerous rules are bound to some parser context, which makes them behave well.
-
-However, for practical reasons, it is very important that grammar is not broken by our custom adjustments (and that the ANTLR parser of the language works properly), because we might need to automatically generate parsers of the adjusted language.
-Imagine this very real scenario (that will be supported in the future):
-\begin{enumerate}
-	\item We have imported the language inside MPS.
-	\item MPS now knows the structure of the language, and we can code in MPS using this language.
-	\item We would expect MPS to be able to load an existing source code, written in this language, from a text file and import it inside MPS.
-	\item We would like to use MPS to safely edit this code, using all the features of MPS.
-	\item We would also like to export the code in a plain-text form again and save it back to the source file.
-\end{enumerate}
-In order to make this happen, we must be able to generate a correctly-working ANTLR parser out of the adjusted source grammar that we used to (that helped us) import the language.
-We must also be able to match nodes of the AST coming out of the ANTLR parser to elements of the MPS language, and we must build the MPS AST out of the ANTLR AST.
-
-\section{Evaluation}
-
-We implemented everything (the proposed approach described above) in the form of an MPS plugin that allows users to carry out the import.
-
-We used a mix of Java and BaseLanguage in our implementation.
-We used the MPS API, which can be called from the BaseLanguage programs too, to generate some of the language elements.
-BaseLanguage is needed to use/call the MPS (language) API, i.e. to programmatically create new language elements and their aspects.
-
-The plugin uses the ANTLR library to work with ANTLRv4 grammar files (to parse them, etc) and several MPS libraries (that implement the MPS API).
-
-For the first step of automated construction of MPS languages from grammars (i.e., for parsing the grammar), the plugin uses the ANTLR library~\cite{ref:ANTLR4} to generate a Java ANTLR parser of the ANTLRv4 notation, because there exists an ANTLRv4 grammar for the ANTLRv4 notation~\cite{ref:ANTLR4reference}.
-
-\todo{
-	We should provide url to the github repository (implementation) and pointer to some package with ANTLR grammars and imported MPS languages (take from Attachments.zip in the master thesis).
-}
-
-\todo{Popsat nase zkusenosti s importem slozitych jazyku (XML, JavaScript, Python, a dalsi).}
-
-From introduction: We will also show some example languages imported with INGRID and discuss our experience.
-
-In order to evaluate INGRID, we tried to import several real/mainstream languages:
-\begin{itemize}
-	\item \textbf{SimpleXML} --- Simplified XML that we used in this thesis for explanatory purposes.
-	\item \textbf{JSON} --- Similar to XML, but this time a full port of the specification.
-	\item \textbf{ECMAScript 5.1} --- Specification of the language known as JavaScript, dated to the year 2011, which is currently the most frequently adopted version.
-\end{itemize}
-Results of the import process are included in the package that is available at \todo{url}.
-The repository \todo{Add url} contains an example MPS project. There are three languages.
-For each example language, you can find its original import, exactly in the form as created by our plugin.
-
-We experimented with import of XML, JavaScript, Python, and maybe some other languages.
-
-These general purpose languages are quite complex when it comes to their structure and the overall syntax variety.
-
-We must emphasize that the automatically imported languages are not ready-to-use full-fledged MPS languages.
-We did not expect to achieve that at this stage, as that is a much bigger, if not impossible, challenge.
-Imported languages contain full structure found in the respective grammars and some more aspect definitions that will help users of the languages in creating code.
-The imported language has to be manually adjusted (tweak, polished) by the end user (developer) and some of its complications have to be resolved manually by a human.
-
-We found out that manual adjusting of the editor/textgen aspect (adding line breaks and indentation) is far more effective and precise than any automated approach we came up with (during our research).
-The imported language can be, in our opinion, adjusted very fast into a more readable form.
-
-The package at \todo{url} contains also adjusted versions of all three languages.
-Then, for each language, there is an adjusted version.
-
-This adjusted version was created from the original import and the author spent no more than between 20 to 60 minutes of working on it.
-The first author spent only a couple of minutes of (manually) adjusting them after the automated import, so that we try to prove our point above.
-More specifically, the JSON language was ready for use in no more than 20 minutes of minor adjusting.
-We chose this language because it does not require us to implement complex aspects, such as type checking or data flows.
-Similarly, the SimpleXML turned out very nice.
-
-Our main goal was to prove that even though our method (plugin) does not create a perfect language, it can be customized very fast into a very usable form.
-We manually tweaked/customized the editor and the TextGen aspect only, structure aspect was left in its original form.
-This justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical (and the best).
-
-We have also tried to use the INGRID tool/plugin to import the JavaScript language, which is exactly the type of a complicated general purpose language that might be of interest to other people.
-There are already some projects, where a manual port of JavaScript is being done.
-One of them is the ECMAScript4MPS~\cite{ref:ECMAScript4MPS} from the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section X (similar projects, related work).
-The result of import by INGRID is quite good, comparable to the PE4MPS project.
-Our language is, of course, missing a lot of more advanced aspects/features that, for example, can help deriving types of expressions and so on.
-These aspects improve the usability by a lot, e.g.
-when the user starts writing numbers, a number literal is inserted, whereas in our import the user has to first insert the concept representing the number literal and then proceed with writing numbers.
-These actions are needed in any MPS language and are expected to be implemented after the import step, \todo{Maybe say that INGRID does not generate these actions/aspects/features yet but it is our future work} as it is not possible to generate them automatically.
-However, when it comes to language structure, concept aliases, and auto-completion, we think that INGRID (our method/plugin) did a very good job.
-The structure aspect is very similar to the ECMAScript4MPS's one, which was created manually over the course of a large number of hours.
-INGRID does that (achieves the same result) fully automatically.
-
-To summarize, import languages such as JavaScript, JSON, MySQL and others is giving nice results.
-
-Results of our experiments with languages such as JSON or JavaScript show that the proposed approach (INGRID) yields very good results, comparable to existing manual ports, and spares the user from hours of tedious, error prone and time-consuming work.
-
-Limitation: the project is lacking proper error reporting and cannot handle cyclic rules (which could be found in Python grammar, for example).
-
-\section{Conclusion and Future Work}
-
-\todo{Write something.}
-
-There is lot of further follow-up work to be done. Many open problems remain.
-
-Automated generation of the editor aspect, probably using some kind of machine learning.
-Generating nicely designed and user-friendly editors, maybe using artificial intelligence (machine learning) that would evaluate (analyze) existing code in plain text and based on that create the editors.
-
-Follow-up work might look more into the problem of code layout detection, possibly exploring suggested approaches such as the learning approach described in Section XY \todo{Tohle se snad podari aspon castecne implementovat}.
-This could lead to better results in editor and TextGen aspect generation which could definitely use some improvement.
-Another future endeavors will definitely touch the subject of generating parsers that would enable the user to import existing source code into MPS.
-
-\acks
-% TODO
-This research was supported by JetBrains. \todo{Pridat jeste nejaky grant a kompletne preformulovat.}
-
-\bibliographystyle{abbrvnat}
-
-% TODO opravit formatovani jednotlivych zaznamu a vhodne seradit
-\begin{thebibliography}{}
-
-\bibitem{ref:MPS} JetBrains MPS, \url{https://www.jetbrains.com/mps/}
-
-\bibitem{ref:mbeddr} mbeddr, \url{mbeddr.com/}
-
-\bibitem{ref:BaseLanguage} BaseLanguage, \url{confluence.jetbrains.com/display/MPSD34/Base+Language}
-
-\bibitem{ref:PE4MPS} PE4MPS project, \url{https://github.com/mar9000/pe4mps}
-
-\bibitem{ref:PE} PE project, \url{https://github.com/mar9000/pe}
-
-\bibitem{ref:ANTLR2MPS} ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR{\_}MPS}
-
-\bibitem{ref:ANTLR4} ANTLRv4, \url{https://github.com/antlr/antlr4}
-
-\bibitem{ref:ANTLR4reference} T. Parr. The Definitive ANTLR 4 Reference. The Pragmatic Bookshelf, 2013
-
-\bibitem{ref:ANTLR4grammars} ANTLRv4 grammar repository, \url{https://github.com/antlr/grammars-v4}
-
-\bibitem{ref:ECMAScript4MPS} ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}
-
-\bibitem{ref:javascript} JavaScript ANTLRv4 grammar, \url{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}
-
-\end{thebibliography}
-
-\end{document}

### 2016-09-23 chapters/conclusion.tex
Split paper into chapter files

--- 2016-09-21 paper.tex
@@
-\documentclass[10pt]{sigplanconf}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[noline]{algorithm2e}
-\usepackage{comment}
-\usepackage{color}
-\usepackage{times}
-\usepackage{multirow}
-\usepackage{balance}
-\usepackage{adjustbox}
-\usepackage{alltt}
-\usepackage{textcomp}
-\usepackage[usenames]{xcolor}
-
-\setcounter{tocdepth}{3}
-
-% configures style of source code listings
-\lstset{
-	basicstyle=\small\sffamily
-	%basicstyle=\scriptsize\ttfamily
-	%numbers=left
-}
-
-\lstdefinelanguage{java}{
-	morekeywords={abstract,continue,for,new,switch,assert,default,goto,package,synchronized,boolean,stream,event,filter,when,do,if,private,this,break,double,implements,protected,throw,byte,else,import,public,throws,case,enum,instanceof,return,transient,catch,extends,int,short,try,char,final,interface,static,void,class,finally,long,strictfp,volatile,const,float,native,super,while},
-	sensitive=false,
-	morestring=[b]",
-	morestring=[b]',
-	morecomment=[l]{//},
-	morecomment=[s]{/*}{*/)},
-	escapeinside=??,
-	moredelim=[is][\textit]{__}{__},
-	moredelim=[is][\textbf]{_*}{*_}
-}
-\lstnewenvironment{displayjava}
-	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b,xleftmargin=1em,xrightmargin=1em}}{}
-\lstnewenvironment{java}
-	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b}}{}
-
-\newcommand{\code}[1]{{\sf \small #1}\xspace}
-
-\newcommand{\mi}[1]{\mathit{#1}}
-\newcommand{\mr}[1]{\mathrm{#1}}
-\newcommand{\mt}[1]{\mathtt{#1}}
-
-\newcommand{\todo}[1]{{\bfseries #1}}
-
-% for writing comments
-\newcommand{\PP}[1]{\textcolor{blue}{\bfseries PP: #1}} % Pavel Parizek
-\newcommand{\PV}[1]{\textcolor{red}{\bfseries PV: #1}} % Premek Vysoky
-\newcommand{\JB}[1]{\textcolor{orange}{\bfseries JB: #1}} % JetBrains
-
-% colors for the ANTLR environment
-\definecolor{antlrbgcolor}{rgb}{.962, .962, .962}
-\definecolor{antlrparserrulecolor}{RGB}{11,83,148}
-\definecolor{antlrlexerrulecolor}{RGB}{191,144,0}
-\definecolor{antlrliteralcolor}{RGB}{133,32,12}
-\definecolor{antlrregexcolor}{RGB}{56,118,29}
-
-% colors for the MPS environment
-\definecolor{mpsbgcolor}{rgb}{.962, .962, .962}
-\definecolor{mpsconceptcolor}{RGB}{42, 140, 0}
-\definecolor{mpsinterfacecolor}{RGB}{170,0,229}
-
-% coloring commands
-\newcommand{\antlrap}{\textquotesingle}
-\newcommand{\antlrparserrule}[1]{\textcolor{antlrparserrulecolor}{#1}}
-\newcommand{\antlrlexerrule}[1]{\textcolor{antlrlexerrulecolor}{#1}}
-\newcommand{\antlrliteralnoap}[1]{\textcolor{antlrliteralcolor}{#1}}
-\newcommand{\antlrliteral}[1]{\antlrap\antlrliteralnoap{#1}\antlrap}
-\newcommand{\antlrregex}[1]{\textcolor{antlrregexcolor}{#1}}
-% TODO mozna zmenit "mpsconcept" na "mpslangelem"
-\newcommand{\mpsconcept}[1]{\textcolor{mpsconceptcolor}{#1}}
-\newcommand{\mpsinterface}[1]{\textcolor{mpsinterfacecolor}{#1}}
-
-% TODO zmenit barvu pozadi na bilou
-% environment for the ANTLR code (grammars)
-\newenvironment{antlr}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=antlrbgcolor,frame,minipage=394pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
-
-% environment for the MPS code
-\newenvironment{mps}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=mpsbgcolor,frame,minipage=394pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
-
-\begin{document}
-
-% TODO
-%\conferenceinfo{OOPSLA'12,} {October 19--26, 2012, Tuscon, Arizona, USA.}
-%\CopyrightYear{2012}
-%\copyrightdata{978-1-4503-1561-6/12/10}
-
-% TODO
-\title{\todo{INGRID: Creating Languages in MPS by Importing ANTLR Grammars}}
-% alternative: "INGRID: Importing Grammars Into MPS"
-
-\authorinfo{P\v{r}emysl Vysok\'y \and Pavel Par\'izek}
-	{Charles University, Faculty of Mathematics and Physics, Prague, Czech Republic}
-	{premek.vysoky@gmail.com, parizek@d3s.mff.cuni.cz}
-
-% TODO pridat nekoho dalsiho z JetBrains
-% TODO upravit affiliation pro JetBrains -> doplnit "s.r.o" nebo radsi "Inc."
-% TODO muzeme taky napsat jejich emaily pokud budou lidi z JetBrains souhlasit
-\authorinfo{V\'aclav Pech}
-	{JetBrains}
-
-\maketitle
-
-\begin{abstract}
-% TODO
-\todo{write later}
-
-\todo{Copy from master thesis.}
-JetBrains MPS is a language workbench focusing on domain-specific languages.
-Unlike many other language workbenches and IDEs, it uses a projectional editor for code.
-The developer directly manipulates the program in its tree form (AST) and not by editing a text source code.
-This brings many advantages, but on the other hand requires time-consuming and complicated MPS language definition.
-The thesis elaborates on the possibility of automating the process of creating MPS language definition from its grammar description.
-It introduces the MPS editor, evaluates approaches of related projects and describes author's efforts to implement an MPS plugin that allows this import.
-The chosen approach and the selection of tools used for implementation are justified in the thesis.
-We point out important problems that any similar project might deal with and we introduce some possible solutions.
-Furthermore, the thesis contains examples of imported languages, showing the potency of the chosen approach.
-The thesis also aims to lay groundwork for future extensions and suggest possible improvements.
-\end{abstract}
-
-% TODO possible keywords: JetBrains MPS, programming language, grammar, projectional editor
-
-\section{Introduction}
-
-\JB{Odstavec mirne upraven} The prevailing approach to define valid syntaxes for programming languages is through grammars written typically in notations based on (or similar to) EBNF [reference].
-Many popular tools exist for automated generation of parsers from the grammar definitions, such as ANTLR and Bison/Flex.
-The source code of programs in such languages is written in a text editors, usually embedded into an IDE.
-
-\JB{Odstavec mirne upraven} Projectional (or structural) editing offers an alternative approach.
-While the idea emerged as early as in 1970s, it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
-A recent revival of projectional editing has been observed in the area of language workbenches - IDE-like tools that enable the developers to manipulate the actual language definition.
-The Meta-programming System (MPS) [reference] by JetBrains is an open-source language workbench that focuses of Domain Specific Languages and leverages the concept of projectional editor.
-MPS allows developers to design custom languages and then use use them to write code and compile it into executables.
-MPS provides the whole IDE infrastructure for such languages.
-
-\JB{Odstavec mirne upraven} The key idea behind projectional editors is that a developer directly manipulates the AST representation of the source code instead of plain text.
-To avoid the fate of the early projectional editors MPS has put extra effort into making text manipulation feel like editing real text.
-MPS models, although being persisted as XML or binary files, can be versioned using the industry standard source-control systems (Git, Subversion).
-On-line model browsers have been incorporated into some of the popular public source code repositories (BitBucket) [reference] .
-
-\JB{Odstavec mirne upraven} The absence of parsers brings about numerous benefits, which are in particular useful for DSLs, which are frequently used by domain experts, who may not be professional programmers.
-\begin{itemize}
-	\item Since a projectional editors controls the interaction between the user and the code, the developers practically cannot enter syntactically invalid code.
-	\item Languages can be defined in a modular way and multiple languages can be easily combined together or extend one another, while parsers seriously limit language modularity for traditional parsed languages.
-	\item The languages may offer contextual or non-parseable notations, such as tables, diagrams or positional syntax.
-	\item Since the projection is detached from the physical representation of code (AST), authors can define multiple notations and allow the developers to switch between different representations of the code.
-\end{itemize}
-A motivating illustration of the abilities of MPS is the mbeddr project [citation], which extends C with numerous domain-specific constructs and data types.
-We provide more details about MPS in Section 2.1.
-
-\JB{Navrhuji smazat nasledujici odstavec}
-\todo{From thesis.
-	One of these tools is the JetBrains MPS~\cite{ref:MPS}, which approaches languages from a~different point of view.
-	MPS doesn't use the textual representation of code as usual, but rather works with the actual abstract syntax tree that holds the code's structure.
-	This has some positive implications as it gives us a~very powerful tool rich on features but also introduces some new problems we haven't encountered before.
-	One of them being the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-	MPS also becomes a~very powerful tool once the language has already been created and you start using it inside MPS.
-	There is a~huge variety of new languages constructed for MPS that usually solve domain specific problems.
-}
-
-\todo{Pouzivat jen terminy "AST nodes" a "language elements" (to spis). Nepouzivat pojem "concepts" abychom ctenare nematli.}
-
-\JB{Odstavec upraven} Projectional editing introduces some new problems we haven't encountered before.
-One of them is the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-In order to use a language inside MPS, it has to be defined through the MPS infrastructure.
-In particular, one has to create the abstract syntax called Structure (the types of allowed AST nodes), concrete syntax called Editor (projection of the AST on the screen and its interactions with the user), and text generation (to enable creation of plain text files used as input for compilers).
-We provide more details in Section X (Background about MPS).
-
-\JB{Odstavec pridan pro vysvetleni motivace prace} MPS is typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
-These DSLs are then transformed on the AST level by a series of model-to-model transformations to an AST model representing the desired semantics in a General-purpose language (GPL), such as C or Java.
-This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
-This implies that the target GPL must also be defined by the MPS means, to allow the DSLs to have their models transformed into a model in the GPL.
-
-\JB{Odstavec doplnen} However, it requires substantial effort to create a definition of a GPL in MPS.
-Therefore, very few mainstream GPLs are now fully supported by MPS.
-Only Java and C have been implemented to date, which limits the DSLs authors to only target these supported target GPLs.
-We believe that an automated process of importing grammars of GPLs into MPS would encourage more GPLs to be migrated to MPS and thus giving the DSL authors more options.
-
-\todo{From thesis.
-	However, building these languages is a~complicated and time-consuming process and a~lot of effort must be put into them before they are ready for use.
-
-	These examples show that recreating a~full language inside MPS is not an easy task and a~lot of time must be spent on implementing all aspects of the language.
-	A big part of this effort is, however, quite straightforward and can be possibly automated, as we show in this paper. Automation speeds up the process of adoption of new languages to a great degree.
-}
-
-\JB{Odstavec upraven} Since a big part of the language definition process in MPS is quite straightforward and can possibly be automated, we will look into the possibility of automatic import of already existing general purpose languages into MPS using a grammar description of their syntax.
-
-We present INGRID --- a method and a tool for semi-automated construction of MPS definition of programming languages based on their ANTLR grammars.
-We also discuss challenges that we encountered during this process and our solution to them.
-Most of these challenges are general, i.e.
-not MPS specific, as they apply to the relation between grammars defined in EBNF-like style (rules and tokens, everything defined in plain text) and structured (object-oriented) approach used in MPS (and other language workbenches).
-
-\todo{Hodne zduraznit automation, tedy ze chceme maximalizovat automatizaci.}
-The main goal of our/the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible.
-
-\todo{From thesis.
-	Automatic import of already existing languages into MPS from the grammar description of their syntax.
-}
-\todo{Mozna prodat INGRID jako automatickou konverzi gramatik do podoby MPS languages, tedy jako nastroj ktery umoznuje portaci jazyku do prostredi MPS a jejich nasledne pouziti v MPS (treba kombinace s dalsimi jazyky).}
-
-Our primary motivation is to enable better support for mainstream and/or general purpose programming languages (XML, C, C++, JavaScript, Python) in MPS, especially to enable writing programs in these languages using the projectional editor and to combine them with domain-specific languages natively created in MPS.
-\todo{Zduraznit take pouziti importovanych jazyku pro cile generovani z DSL --- to plati zejmena u mainstream general-purpose languages jako treba JavaScript.}
-
-\todo{Briefly mention the biggest challenges.}
-The are related to the way grammar encoding/description of languages are usually written.
-They do not hold any information about the code layout, and contain many rules that do not directly correspond to AST nodes (programming language constructs).
-We discuss them in more detail later (Section X).
-
-\todo{From thesis.
-	INGRID allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
-	The biggest benefit is the possibility to code in these mainstream general purpose languages using the projectional editor.
-}
-
-However, we do not target just the very complex general purpose languages such as JavaScript.
-INGRID can be used also for simple languages.
-We believe that, for simple languages, it is less time consuming to create the grammar by hand (e.g., using the ANLTR notation and toolset) and then create the language definition in MPS by using the INGRID tool, than to create it manually using the MPS GUI --- this is another benefit of INGRID.
-
-Structure of the paper.
-In the beginning, we will describe the MPS editor and related/similar projects.
-Then, we describe our approach to creating language definition in MPS from their grammars, and discuss the major challenges together with our solution.
-We will also show some example languages imported with INGRID and discuss our experience.
-
-\section{Background and Related Work}
-
-We will briefly introduce fundamental basics of the MPS environment.
-We will analyze existing related projects that are trying to accomplish similar goals, and then we will weigh advantages and disadvantages of these approaches.
-
-\subsection{JetBrains MPS}
-
-\JB{Odstavec upraven} JetBrains MPS is a complete language workbench - a development environment that allows developers to define their own languages and then use them to develop code.
-The code can then be transformed into a target language, typically a GPL, such as Java or C.
-
-\JB{Navrhuji odstranit} We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
-Focus on basic/fundamental and relevant features. In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.
-
-\JB{odstavec upraven} MPS differs from typical IDEs in one important aspect - the projectional editor.
-The developer does not work with the textual representation of the source code directly, but rather with its AST (abstract syntax tree) that is the model of the code.
-When users are creating programs using the MPS projectional editor, they are basically assembling the tree (AST) out of predefined building blocks of the imported languages.
-The definition of the language dictates, which where in the AST can certain elements be placed or how they can be nested inside other elements.
-In traditional IDEs it is the parser that constructs an AST using the language's elements.
-
-\JB{Odstavec upraven} The building blocks of MPS models are called nodes.
-Code of any program in MPS is built from nodes and these nodes represent instances of elements defined in one of the languages that the program declares to be using.
-These language elements are called concepts.
-In brief, nodes in code are instances of concepts defined in the language.
-
-\JB{Vyhody/nevyhody diskutujeme v kapitole 1. Navrhuji nasledujici odstavec zrusit nebo text z kapitoly 1. presunout sem}
-Keeping the model of the code in the AST form has several advantages.
-One of them is that MPS then only allows composing the program strictly following the syntax of the language, which results in an inability to actually write syntactically invalid code.
-The code is generated out of the always-valid AST beyond user's reach.
-
-\JB{Odstavec upraven} One of key advantages of projectional editor stems from separation of abstract and concrete syntax.
-While the AST provides an exhaustive representation of the code, the way it is displayed on the screen and the way the user interacts with it are unconstrained.
-The editor can take any visual form and shape.
-The language author can define multiple visualizations and let the developer to choose one that fits best the task at hands - editing, debugging, reviewing, resolving merge conflicts, etc.
-
-For a better understanding, we will give a~small example. Let's assume we have a language with typical if/else structure similar to the one in Java:
-
-\begin{center}
-	\begin{minipage}{.38\textwidth}
-		\begin{alltt}
-			if (\textit{condition})
-			    \textit{then statement block}
-			else
-			    \textit{else statement block}
-		\end{alltt}
-	\end{minipage}
-\end{center}
-
-Its abstract syntax tree representation will look something like shown on Figure~\ref{fig:if_ast}.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[scale=0.6]{./images/if_statement_ast.png}
-  \caption{"If statement" abstract syntax tree}
-  \label{fig:if_ast}
-\end{figure}
-
-The underlying child nodes have their own children and so on.
-The type of nodes can be restricted by the language structure definition.
-For example, we can restrict the condition in the if/else statement to be an expression, statement block to be a list of statements, etc.
-
-\JB{odstavec upraven} Alternative visual representations are possible and they are not bound to be just the textual representation at all.
-For example, blocks for both of the branches can be aligned next to each other and displayed with differend background color, like shown in Figure~\ref{fig:if_editor}.
-
-\begin{figure}[ht]
-	\centering
-	\hspace{-4mm}
-	\includegraphics[scale=0.75]{./images/if_statement_editor.png}
-	\caption{"If statement" projectional editor example}
-	\label{fig:if_editor}
-\end{figure}
-
-\JB{Odstavec mirne upraven}The definition of a language element (the concept) consists of several aspects and each aspect codifies a different part of the nodes' behavior.
-The essential aspects are the following: structure (abstract syntax), editor (concrete syntax - how the code is projected and edited) and textgen (how to transform the AST into textual source code).
-If, instead of generating text directly, the language is supposed to be generated into another language that is available in MPS, the generator aspect must be used to specify the model-transformation rules.
-The structure, editor and textgen aspects are most relevant for the contribution of this paper, so we describe them here (and neglect/ignore the other aspects such as type system and data-flow).
-
-\todo{Highlight the object-oriented hierarchical nature of language definitions in MPS. The environment restricts and guides the developer so that he makes less mistakes.}
-
-Some concepts and design patterns known from object oriented programming can be used in the language definition --- inheritance, interfaces, and so on.
-The AST node can be inherited from a parent (abstract) node and it can implement a special interface node.
-Most importantly, nodes have child nodes and that is how the AST itself is created.
-
-\todo{Zduraznit ze AST node (object) ma nekolik aspectu: structure, editor, textgen. Definice MPS language je organizovana takto --- pres AST nodes, ktere maji nekolik aspectu (ruzne pohledy).}
-
-\paragraph{Structure.}
-The most important aspect of any MPS language is structure. It must be created first for each AST node.
-It specifies core attributes such as name of the AST node, inheritance relationships, possible child nodes (types and cardinality), implemented interfaces, references to other nodes, other properties (fields of any type that can hold values), etc.
-In Figure~\ref{fig:if_statement_structure}, you can see the structure aspect definition for the if statement.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.55]{./images/if_statement_structure.png}
-	\caption{"If statement" structure aspect definition}
-	\label{fig:if_statement_structure}
-\end{figure}
-
-\paragraph{Editor.}
-Editor aspect is where the user defines what the projectional editor representation of a code fragment (an AST node) looks like.
-The editor aspect of an AST node defines its visual representation (of the code) in the projectional editor.
-The user usually incorporates all children, references, and properties inside the representation so that future user of the language can insert their value into these placeholders.
-All elements can be styled using a~language similar to CSS.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/if_statement_editor_definition.png}
-	\caption{"If statement" editor aspect definition}
-	\label{fig:if_editor_definition}
-\end{figure}
-
-\todo{From thesis.
-
-The editor is a module dedicated/responsible for the definition of appearance of an AST node (concept).
-JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells.
-These cells then can be styled to user's liking.
-There are many different types of cells, each behaving a little bit different towards its contents.
-For example, MPS supports various horizontal and vertical lists, etc.
-Extra cells can be added on top --- cells which just specify layout adjustments such as indentation and text color.
-In Figure~\ref{fig:if_editor_definition}, you can see what a real editor definition might look like for the if statement of the Java language (as defined in MPS).
-}
-
-\paragraph{BaseLanguage}
-\todo{Introduce the base language here.}
-A program written in an MPS language cannot be executed directly.
-MPS uses an approach based on translation into a program in the so-called BaseLanguage.
-A plain-text representation of a BaseLanguage program is then passed to a system compiler and the result is executed.
-The module responsible for generating the program in the BaseLanguage is called "textgen" (we describe it briefly in the next paragraph).
-
-BaseLanguage is also an MPS language --- it is the official port of Java for MPS (created by JetBrains, authors of MPS).
-The default base language of MPS is Java, but others are supported too.
-
-Since BaseLanguage is an MPS language, BaseLanguage code is not a (plain) text source code, but again, an AST built out of nodes belonging to the BaseLanguage language.
-This means that generating code inside MPS consists of generating some AST out of BaseLanguage nodes.
-
-Besides representing the default (most common) target of text generation, BaseLanguage is used also for all additional programming inside MPS.
-
-\paragraph{TextGen.}
-The TextGen aspect defines how a certain AST node will be translated into a text source code (in the base language).
-This will allow us to generate source code in text form out of an AST model and ultimately to generate a text program out of the code user writes in the (imported) MPS language.
-The TextGen aspect is basically just one method/procedure definition for each AST node in the language. The method produces string representation of the given AST node and its children.
-MPS then calls this method for the root AST node of the program.
-TextGen aspect has to be defined (especially its more complex functionality) using the BaseLanguage.
-Again, we included an example of the if statement (Figure~\ref{fig:if_statement_textgen}).
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.6]{./images/if_statement_textgen.png}
-	\caption{"If statement" TextGen aspect definition}
-	\label{fig:if_statement_textgen}
-\end{figure}
-
-\todo{From thesis.
-
-After the program's AST is created using the projectional editor, there can be rules defined on how to generate code in plain text out of that tree.
-This module is called "text gen".
-These defined generators can target any language and platform.
-
-\paragraph{Other aspects.}
-MPS supports also other aspects, such as type checking and data-flow analysis, but they are not important for our contribution.
-Moreover, information needed for their automated generation is not available/contained inside a grammar file.
-
-}
-
-\subsection{Similar Projects}
-
-Currently, there exists an almost full port of the Java language called BaseLanguage~\cite{ref:BaseLanguage} extended with some MPS specific features.
-It was imported manually by JetBrains and it is still undergoing changes as Java itself is evolving.
-
-Within the mbeddr project~\cite{ref:mbeddr}, the C language is also manually tailored for MPS.
-
-Now/here we describe three similar projects (with similar goals) that we are aware of.
-We focus on (consider) these three projects: PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
-
-\paragraph{PE4MPS.}
-PE4MPS~\cite{ref:PE4MPS} is a project by \todo{Doplnit autora} that is trying to solve the same problem as we do.
-It solves the lack of information about code layout in grammars by creating a new grammar notation called PE Grammars~\cite{ref:PE} (the PE abbreviation comes from projectional editing).
-It has tw components: PE parser and PE4MPS plugin for MPS.
-Their approach is to mimic an existing grammar notation called ANTLRv4~\cite{ANTLR4} and enrich its syntax with custom (their own) constructs.
-These extensions tell (guide, hint) the parser what the AST node layout should look like.
-Supported extensions (already implemented) include horizontal and vertical lists, and some indentation rules --- however, even these few features make the already not-so-simple syntax of ANTLR v4 much more complex.
-The parser then uses this information when generating the projectional editor for am AST node.
-Author of this approach/project described the PE syntax using the ANTLRv4 notation~\cite{ANTLR4reference} and then used the ANTLR toolset to automatically generate an ANTLRv4 parser for PE grammar files.
-The PE parser reads any PE file and stores the language structure found inside to a custom representation of Java objects.
-
-The PE4MPS plugin/project/tool is built on top of the PE parser.
-This plugin uses the PE parser to build the PE file representation (the aforementioned tree-like structure of Java objects) and then creates AST nodes and their aspects inside MPS.
-The extended syntax brought by PE describes the layout of each element, e.g.
-it tells the plugin that one set of child nodes should be displayed horizontally, another set should be vertical with each child on a separate line and with some indentation, and so on.
-
-A limitation of the PE4MPS approach is that it, from what we understand, does not implement full ANTLR syntax.
-This means that every grammar might need a non-trivial adjustment before its usage.
-One of our goals is to enable import of as many languages out-of-the-box as possible and adopting the full specification.
-
-\paragraph{ANTLR{\_}MPS.}
-ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
-The author of this project is Fabien Campagne
-The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
-However, it does not try to generate the projectional editor at all (i.e., it does not deal with this problem), probably because it is in an early stage of development.
-
-The way this import process of ANTLR{\_}MPS works is quite different from what we have seen in the PE4MPS project, and it is quite complex/complicated to use.
-It works as follows. We give only a brief high-level overview and omit low-level technical details.
-The author created a whole new ANTLRv4 MPS language, which is an MPS port of the grammar notations' syntax.
-To import a language, the user utilizes this MPS ANTLR language.
-The textual grammar is imported automatically into MPS taking the form of the MPS's ported grammar language (so that the textual grammar is converted into MPS nodes, that means an AST node is created for each grammar rule).
-No child-parent relationships in the structure are generated by the tool --- all must be created manually by the user.
-There are no editor nor TextGen aspects created neither.
-
-\paragraph{mps-metabnf.}
-\todo{Popis na zaklade emailove diskuze (Nizozemci).}
-
-\section{Creating Language Definition in MPS}
-
-From all the available notations in which grammars can be written, we decided to use the ANTLRv4 grammar notation~\cite{ANTLR4reference} as the input format (in which the input programming languages have to be defined) for the following reasons.
-\begin{itemize}
-	\item It is a very widely used grammar notation.
-	\item All popular and widely used programming languages have their syntax captured (represented) using this notation --- see the repository~\cite{ref:ANTLR4grammars}.
-	\item There is a whole framework/toolset for generating parsers and editing grammars built around it~\cite{ref:ANTLR4}.
-	\item It supports Java (it is written in Java), which is the implementation language of MPS and also the default BaseLanguage.
-	\item It uses (is based on) an EBNF (Extended Backus-Naur Form) \footnote{https://en.wikipedia.org/wiki/Extended{\_}Backus\%E2\%80\%93Naur{\_}Form} form that it useful because it meets/respects the object oriented design that is used when describing MPS aspects (definitions of MPS languages).
-	\item It has its own syntax captured in its own syntax\footnote{https://github.com/antlr/grammars-v4/tree/master/antlr4} (there is an ANTLRv4 grammar for the ANTLRv4 notation).
-This is a very important feature as it allows us to create a parser for ANTLRv4 grammars so that we can parse grammars of other languages.
-\end{itemize}
-
-We use ANTLRv4 grammars as inputs (and work with them).
-	https://github.com/antlr/antlr4/blob/master/doc/index.md
-	https://github.com/antlr/grammars-v4
-
-Unlike some of the related projects, we did not add any new features into the grammar notation.
-We tried to import grammars as they are.
-
-The generation/import process performed by INGRID consists of three phases (import of structure, usable editor, text generation).
-We will talk about all phases of the generation/import process, describing encountered challenges/obstacles using an example language.
-
-To fulfill our goal of maximal possible automation, we import/generate the full language structure and then do a lot of mundane and time-consuming work connected to creating projectional editors.
-
-Now we describe/introduce the main steps of our approach (that INGRID tool performs): parsing the grammar, creating the structure, editor, and textgen.
-We are able to create/generate the structure aspect and then very basic editor and some default/basic textgen.
-In particular, we did not port the ANTLR notation as an MPS language.
-In the following subsections, we provide some details about our approach to parsing the input grammar and creating the main aspects of each element (each type of AST node) of an MPS language (structure, editor, and textgen).
-
-During our work on this project, we also discovered/encountered challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component), whitespace (editor aspect).
-There is a challenge/issue that we have discovered during our work on this project (in the process of our own implementation), and that is related to construction of the structure aspect.
-Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
-However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
-We describe/discuss these challenges in more detail (also using examples) and our solutions to the challenges inside the respective following subsections.
-
-Result of the construction of a MPS language from the input grammar (import) has to be polished by the human user.
-For example, the aspects not supported by INGRID will be used for adjusting the imported language and improving the usability of the language.
-
-\todo{Briefly discuss how our overall approach differs from similar projects (PE4MPS, etc).}
-There are more ways how to approach the problem of a grammar import.
-
-Our overall approach differs from the PE4MPS project (Section X.Y) especially in the level of automation.
-While the main goal of (our) the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible, usage of PE does not really spare the user any work because it only shifts the manual work from creating projectional editor to writing PE grammars.
-What the user would do inside the MPS UI using the projectional editor is in the case of PE4MPS done using a text editor when editing PE grammars.
-MPS UI is specially designed/made/tailored for this purpose, guides the user through this process, and supports aids such as auto-completion.
-In particular, MPS UI allows the user to perform only changes that result in valid output.
-Editing the grammar stored in a text file using a standard editor is an inferior approach with respect to using MPS (projectional editor), because a text editor is not as user-friendly, and, more importantly, it is an error prone environment.
-
-Main drawback of ANTLR{\_}MPS, and main difference from our approach, is very limited automation: it does not create the parent-child relationships between AST nodes in the structure, and does not generate any editor and textgen aspects (for any node) at all.
-In addition, there is no attention paid to the problem of generating projectional editor and text generation is neglected too.
-We blame all this on the project being in its early stage of development.
-
-In the rest of this section, first we define the language that we will use throughout the paper as the running example.
-Then we will talk about parsing grammar files \todo{Tohle je mozna zbytecne}.
-Lastly, we will describe our approach to tackling generation of language's aspects (structure, editor, and TextGen).
-At some places, we discuss important possible alternatives (as there are several possible approaches).
-
-\subsection{Running Example}
-
-We illustrate the main steps on the example of a simplified XML language, which is defined below (in Figure XY).
-
-\todo{Pouzit jazyk SimpleXML na running exaple.}
-
-We decided to use SimpleXML because it is small enough to fit here, but still can be used to illustrate the main challenges and our solution
-
-Throughout the whole paper, we will be using one example language to illustrate main challenges and our solution.
-The language is a simplified version of XML. We call it \emph{SimpleXML}.
-We created the grammar of the language by taking the official XML ANTLRv4 grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and stripped it down to eliminate/avoid some not so interesting (advanced) features such as XML entities.
-Below in Figure X, you can find the grammar of the SimpleXML language in the ANTLRv4 notation.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\textbf{grammar} \textit{SimpleXML};
-	\antlrparserrule{document}     :   \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
-	\antlrparserrule{prolog}       :   \antlrliteral{<?xml } \antlrparserrule{attribute}* \antlrliteral{?>} ;
-	\antlrparserrule{comment}      :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-	             |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-	             ;
-	\antlrparserrule{attribute}    :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
-	\antlrparserrule{content}      :   \antlrlexerrule{TEXT}
-	             |   \antlrparserrule{element}
-	             |   \antlrparserrule{comment}
-	             |   \antlrlexerrule{CDATA}
-	             ;
-	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-	\textbf{fragment}
-	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
-	             |   \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.}
-	             |   \antlrlexerrule{DIGIT}
-	             ;
-	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
-	\antlrlexerrule{TEXT}         :   \antlrregex{~[<"]*} ;
-	\antlrlexerrule{CDATA}        :   \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
-\end{antlr}
-\caption{SimpleXML grammar}
-\end{figure*}
-
-We have colored the grammar so it is easier to find the way around it (to make navigation easier).
-The grammar contains following basic elements:
-
-\begin{itemize}
-	\item \textbf{ANTLRv4 keywords} -- In black, they are required by the specification.
-	\item \antlrparserrule{\textbf{Parser rules}} -- Stated in blue, they are the rules describing language's structure.
-		The alternatives (that the element on the left side may break into) are separated by the pipe ($|$) character.
-	\item \antlrlexerrule{\textbf{Lexer rules}} -- In yellow, they are rules describing terminal symbols that the parser is matching against the input.
-		Terminals are encoded as string values and regular expressions.
-		The \textbf{fragment} keyword states that the lexer sub-rule is just a helper rule that brings more clarity into the notation but is not visible in the parser output.
-	\item \antlrliteralnoap{\textbf{String values (literals)}} -- In red, always inside of a pair of apostrophes, they are similar to lexer rules.
-		They describe a terminal string, not using regular expressions but by an exact string match.
-	\item \antlrregex{\textbf{Regular expressions}} -- Colored in green, they describe a string token to be matched using a regular expression with the special ANTLRv4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
-	\item \textbf{Rule operators (?+*)} - Left in black, following any element, they quantify the number of occurrences the prepending element can appear in.
-		These operators can also be prepended by an asterisk (*), meaning they are non-greedy.
-\end{itemize}
-
-\subsection{Parsing (the) Grammar}
-
-\todo{Define some name for the AST representing the grammar, and use it consistently throughout the whole paper. Maybe distinguish between "grammar AST" and "MPS structure AST".}
-The first step (stage) of creating the MPS language (of the import process) is parsing the grammar file (written using the ANTLRv4 notation) so that we can get an AST tree representing the grammar.
-
-The abstract syntax tree that comes out of the automatically generated (ANTLR) parser is a little bit complicated and full of information not relevant for our purposes (to us).
-That is why we translate the complex ANTLR AST into our own simplified tree made out of our own objects.
-We discarded as much irrelevant information as possible and only kept information (data) vital/necessary for construction of MPS languages.
-
-Furthermore, we process the grammar in two iterations.
-In the first iteration/pass, we build a tree holding names of other referenced rules in the form of strings.
-In the second iteration, we walk this tree and resolve all these references to real pointers to other rule objects.
-We end up with a clear and simpler representation that is easy to process in the next steps of the import process.
-
-We also had to post-process (simplify) the representation of lexer rules (tokens).
-As a result of the initial parsing, ANTLR parser produces some kind of a tree structure that is representing the lexer rule (meaning/note that lexer rules can also be built from alternatives just like parser rules).
-However, lexer rules are, in fact (basically), just regular expressions used for parsing input source code into tokens.
-We took each/every such tree that represents the lexer rule and flattened it into the equivalent regular expression.
-
-For illustration, consider the lexer rule \antlrlexerrule{Name} from our/the SimpleXML language (Figure XY), together with all its sub-rules, shown in Figure~\ref{fig:lexerrule}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-	\textbf{fragment}
-	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
-             |   \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.}
-             |   \antlrlexerrule{DIGIT}
-             ;
-	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
-\end{antlr}
-\caption{Lexer rule from SimpleXML}
-\label{fig:lexerrule}
-\end{figure*}
-
-Intuitively, we can create a regular expression by gluing elements of each alternative together and then join these alternatives with a classic regex OR ($|$), which is exactly its function in the ANTLRv4 grammar notation.
-We have captures this idea in the form of the recursive algorithm in Figure~\ref{fig:algflatten}, which covers the whole process.
-
-% TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	Flatten(\textit{R}):
-	1) Define \textit{T} as a list of empty string
-	2) For each alternative \textit{A} of the rule \textit{R}:
-	3)     Define \textit{R} = \antlrap\antlrap
-	4)     For each element \textit{E} of \textit{A}:
-	5)         If \textit{E} is not yet flattened:
-	6)             Flatten(\textit{E})
-	7)         \textit{R}.append(\textit{E})
-	8)         \textit{R}.append(\textit{E}.operator)
-	9)     \textit{T}.add(\textit{R})
-	10) Build a string S out of elements t$_{\text{1}}$, t$_{\text{2}}$, ... t$_{\text{n}}$ of \textit{T}:
-	11)     (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
-	12) Return \textit{S}
-\end{antlr}
-\caption{Flatten algorithm}
-\label{fig:algflatten}
-\end{figure*}
-
-The output of \texttt{Flatten(Name)} (output of the Flatten algorithm applied to the \antlrlexerrule{Name} rule) is this regular expression (in this form):
-
-\begin{center}
-	\texttt{([:a-zA-Z])(([:a-zA-Z])|(-)|({\_})|(.)|([0-9]))*}
-\end{center}
-
-\subsection{Structure}
-
-Now, as the next step, we need to translate/transform the AST (the types of AST nodes) derived from the input grammar into elements of the structure aspect of the MPS language definition (into types of AST nodes), and linking the nodes appropriately.
-
-The goal of this second step is to translate (transfer) the syntax (structure) of the input language that is captured by the grammar (ANTLR), into the MPS structure aspect.
-The structure of the MPS language is somewhat similar to the grammar.
-This is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
-The process of transferring grammar rules into the structure of a MPS language presents/involves some interesting challenges.
-The input grammar (EBNF, ANTLR) serves a little bit different purpose than MPS language, and the way grammar rules are usually written down causes some problems that we had to overcome in order to recreate the language inside MPS.
-Note that the challenges and problems are not specific to MPS --- they are common for everyone tackling a similar problem.
-
-We are going to describe two different approaches that we designed --- the straightforward approach (Section~\ref{sect:straightforward_approach}) and the shortcut approach (Section~\ref{sect:shortcut_approach}).
-We will describe two possible approaches that we developed in Section X and Section Y, respectively.
-\todo{Rict neco jako ze ten druhy, slozitejsi, approach vlastne rozsiruje ten prvni, jednodussi.}
-
-However, before focusing on the two approaches, we must describe the basic principles of translation from the AST representing the grammar to the MPS structure, which are common to both approaches.
-The structure is created according to object-oriented principles.
-Each node (element) is represented by an object (and can be treated as such) that may have a parent node, child nodes, and properties of any data type (similar to fields), it may implement any number of interfaces, and it may also hold references to other nodes.
-For every rule of the grammar (for every node in the grammar AST) that breaks into more/multiple alternatives, we must create a node in the MPS structure AST for each alternative of the grammar (every child node).
-Names of the nodes for alternatives are composed from the name of the rule and a number indicating the order of appearance in the grammar.
-For example, considering the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlgrammarelement}
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-             \textcolor{gray}{|   \antlrap<\antlrap Name attribute* \antlrap/>\antlrap}
-             \textcolor{gray}{;}
-\end{antlr}
-\caption{Fragment of the SimpleXML grammar}
-\label{fig:xmlgrammarelement}
-\end{figure*}
-
-Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.75]{./images/element_node_common.png}
-	\caption{Resulting Element{\_}1 node}
-	\label{fig:element_node_common}
-\end{figure}
-
-String literals (e.g., keywords such as \textbf{for}) are omitted/skipped because they will appear only in the corresponding projectional editor.
-The node named \mpsconcept{Element{\_}1} contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name} inside the first alternative of the \antlrparserrule{element} rule,
-Value of the property will be restricted using the regular expression representing the \antlrlexerrule{Name} rule.
-We explain later how the references to parser rules (\antlrparserrule{attribute} and \antlrparserrule{content}) as this is where the approaches differ.
-The main/key difference between these two lies in the way we create children fields for parser rules and in the way we are going to link them together using interface concepts.
-
-\subsubsection{The Straightforward Approach}
-\label{sect:straightforward_approach}
-
-The key idea of this approach is that, for each rule with more than one alternative, an interface node is created.
-Consider the \antlrparserrule{content} rule from our SimpleXML language (Figure~\ref{fig:contentrule}).
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
-\end{antlr}
-\caption{Content rule}
-\label{fig:contentrule}
-\end{figure*}
-
-Any of the four nodes corresponding to alternatives could appear anywhere the \antlrparserrule{content} rule is referenced.
-So we will create one interface node, and then for each alternative of this rule we create one node that implements the interface.
-The resulting fragment of structure will look like the one in Figure~\ref{fig:icontentitf}.
-
-\begin{figure*}[ht]
-\centering
-\begin{mps}
-	\mpsinterface{IContent}   :   \mpsconcept{Content{\_}1}
-           |   \mpsconcept{Content{\_}2}
-           |   \mpsconcept{Content{\_}3}
-           |   \mpsconcept{Content{\_}4}
-\end{mps}
-\caption{IContent interface}
-\label{fig:icontentitf}
-\end{figure*}
-
-References to other parser rules are captured by links to children (interfaces or object nodes) in the AST nodes.
-For the \antlrparserrule{element} rule that is referencing the \antlrparserrule{content} rule, defined as in Figure~\ref{fig:elementrule}, we get the AST node representing the first alternative that is shown in Figure~\ref{fig:element_node_full}.
-There is one \mpsinterface{IElement} interface node that is implemented by two object nodes \mpsconcept{Element{\_}1}, \mpsconcept{Element{\_}2}.
-The node \mpsconcept{Element{\_}1} has two child links pointing to \mpsinterface{IAttribute} and \mpsinterface{IContent}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-\end{antlr}
-\caption{Element rule}
-\label{fig:elementrule}
-\end{figure*}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.75]{./images/element_node_full.png}
-	\caption{Element{\_}1 node's structure aspect}
-	\label{fig:element_node_full}
-\end{figure}
-
-A problem with this approach, which we call \emph{the layer problem}, is related to the grammar rules (and corresponding AST nodes) that represent intermediary layers created only to make the grammar more readable.
-The intermediary layers are created by authors of the grammars in order to make them more readable and more easily (better) maintainable.
-We illustrate this problem (and explain it) on the behavior of auto-completion in MPS.
-Imagine there is a freshly inserted \mpsconcept{Element{\_}1} node (representing the full XML element as is stated in the first alternative of the \antlrparserrule{element} parser rule).
-Now we would like to insert another nested XML element inside.
-MPS auto-completion mechanism would give us four options that are captured in the left part of Figure~\ref{fig:layer_problem}.
-This is because there exist four nodes (objects) that implement the \mpsinterface{IContent} interface as per each alternative of the rule.
-
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/layer_problem.png}
-	\caption{Layer problem in auto-completion}
-	\label{fig:layer_problem}
-\end{figure*}
-
-However, in order to correctly insert another nested element inside, a user would have to first insert the \mpsconcept{Content{\_}2} node inside \mpsconcept{Element{\_}1} that has an \mpsinterface{IElement} child inside (and nothing else).
-Then, as the second step, a user would invoke/trigger/call for the auto-complete again and insert either \mpsconcept{Element{\_}1} or \mpsconcept{Element{\_}2} inside \mpsconcept{Content{\_}2}.
-This mean that a user would have to go through two steps, and in the first stepe either correctly guess (or remember the grammar rule's alternative order), which option/item offered by the auto-complete menu to select.
-
-Similarly, if one would decide to replace the nested \mpsconcept{Element{\_}1} node with, let's say, an XML comment (a \mpsconcept{Comment} node), we need to delete both intermediary layers before we get back to the original \mpsconcept{Content{\_}X} selection (crossroads).
-The user cannot really see what is happening, since the intermediary level has no appearance or indicator.
-This may lead to confusion on the user's part.
-
-The layer problem is solved/addressed by the second approach, which we describe in the next subsection.
-
-\subsubsection{The Shortcut Approach}
-\label{sect:shortcut_approach}
-
-The key idea of this approach is to identify rules and nodes that are at the end of the (derivation, inheritance) chain, and offer them directly to the user through the auto-completion menu.
-We call such \emph{end rules} and \emph{end nodes}, respectively, because they cannot transparently break into more rules/nodes.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
-
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-
-	\antlrparserrule{comment}    :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-\end{antlr}
-\caption{Content parser rule with children}
-\label{fig:contentrulewithchildren}
-\end{figure*}
-
-For example, the \antlrparserrule{content} rule, given in Figure~\ref{fig:contentrulewithchildren} together with its child rules, can ultimately expand into following nodes:
-
-\begin{itemize}
-	\itemsep0em
-	\item \textbf{Content{\_}1} (TEXT)
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}1}
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}2}
-	\item Content{\_}3 $\rightarrow$ \textbf{Comment}
-	\item \textbf{Content{\_}4} (CDATA)
-\end{itemize}
-
-The end nodes are highlighted using the bold font.
-
-Figure XY shows the algorithm (in pseudocode) that finds all end rules for a given parser rule.
-
-% TODO pouzit standardni environment pro algoritmy 
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	FindPathsToEndNodes($R$):
-	1)  Define $L$ as an empty list of list of nodes
-	2)  Return FindPathsToEndNodes($R$, $L$)
-
-	FindPathsToEndNodes($R$, $L$):
-	3)  Define $Q$ as list of list of nodes
-	4)  For each alternative $A$ of rule $R$:
-	5)      $L1$ = Clone($L$)
-	6)      If $A$ is a parser rule with only one element $E$:
-	7)          Let $I$ be interface/concept representing rule $E$
-	8)          $L1$.Add($I$)
-	9)          $P$ = FindPathsToEndNodes($E$, $L1$)
-	10)         $Q$ = Merge($Q$, $P$)
-	11)     Else
-	12)         $L1$.Add($R$)
-	13)         Let $P$ be concept representing $A$
-	14)         $L1$.Add($P$)
-	15)         $Q$.Add($L1$)
-	16) Return $Q$
-\end{antlr}
-\caption{Algorithm for finding end rules}
-\label{fig:shortcut_algorithm}
-\end{figure*}
-
-To find end rules for each parser rule, the algorithm recursively scans through the parser tree that was built before.
-For each parser rule, it tries to find paths leading to some end rule through its alternatives:
-\begin{itemize}
-	\item Whenever it finds an alternative that contains only one element, and this element is a reference to another parser rule, it has found an intermediary level that can be transparently hidden from the user of the language.
-		The process (algorithm) continues by recursively processing alternatives of this "level" rule (since we are not at the end of the chain yet).
-	\item Otherwise, an end rule was found (recursion stops here).
-\end{itemize}
-
-By appending the rule that is leading to current element (line 12) and then appending that alternative's element itself (line 14), we get a path that contains the full path and the target end rule as the last element of the chain.
-The result of this algorithm, for example for the \antlrparserrule{content} rule, equals the listing of the five paths mentioned above.
-
-We do/repeat this for all parser rules and for each rule we get a list of paths that lead from that particular rule to an end node.
-We call these paths as \textbf{shortcuts}, because they provide a shortcut from the rule to the end of the chain.
-Now that we have the shortcuts, we will discuss several ways how to use the shortcuts (them) to improve the MPS language under construction (how to make the language better).
-
-The primary use of shortcuts is to generate options for the auto-completion menus such that only the end nodes are offered.
-This can be technically done in two ways --- (1) through custom auto-complete menus and deletion handlers or (2) by defining a special interface for each rule such that the interface is implemented only by end nodes for the rule.
-Shortcuts have to be considered when inserting nodes but also when deleting them, i.e.
-the whole chain (shortcut path) consisting of possibly multiple intermediary AST nodes (levels) up to the end node must be added, respectively deleted.
-The effect of insertion must be fully reversed when the user wants to delete some end node (the given end node).
-
-\subsection{Editor}
-
-When all types of AST nodes in the language are created, the next step is to define their visual representation in the projectional editor.
-As stated before (in Section X (about JetBrains MPS, part of background)), MPS uses a cellular system that allows placing node's properties and children into a table-like arrangement.
-Cells of different types are available --- for storing property values, references to child nodes, and keywords (string literals), and also cells that influence layout (e.g., indentation).
-
-The purpose (goals) of this (third) step is to create the editor aspect for each language element.
-It should project all attributes of the element --- name, properties, children, and so on --- using the respective cell types.
-
-The main challenge (problem) here is that an input grammar (ANTLR) does not provide/give any help/aid/information regarding the code layout.
-As we already said in the introduction, one of the biggest challenges/problems that arise here is that the grammar description of a language does not hold any information about the code layout.
-The structural rules in the grammar only tell us, what the syntax tree looks like and how the code is broken into nodes.
-In particular, the grammar says nothing about indentation, line breaks, and another formatting.
-
-We tackled this problem by designing an automated learning approach, which generates the layout using several heuristics.
-We present our solution to this problem. It involves some intermediary step whose purpose is to generate the necessary information.
-Our solution is semi-automated (using some heuristics) --- user's help in an interactive manner is needed for final touches (polishing).
-It works as follows.
-
-As input for the algorithm (learning approach), the user has to supply to a set of valid source files written in the imported language, together with the grammar file.
-These four steps/operations are then performed.
-\begin{enumerate}
-	\item Automated generation of an ANTLR parser for the input language using the ANTLR library.
-	\item Modify/extend the code of the generated parser in order to record some additional information.
-	\item Compile the parser into an executable form.
-	\item Use the parser to parse the supplied source code files and extract information about the layout of the code during this process.
-\end{enumerate}
-This way, the imported/created MPS language would inherit the code style/layout directly from the user, i.e. it would learn directly from the code.
-In step 2, the parser is modified such that it remembers/stores the line number and column for each parsed token.
-This is very efficient and it enables to create a nice layout.
-From the line number, the algorithm can easily determine whether there is a line break between some tokens (statements, language elements).
-The column number can be then used to detect indentation.
-However, this approach is still heuristic and may give imprecise (suboptimal) results --- mainly because it is not always possible to precisely match language elements (in the MPS structure) with all tokens that belong to it.
-Note also that quality of the extracted information always depends on the amount of the source code files provided (available, supplied) and on how representative the set of source code files actually is.
-When our algorithm cannot determine some custom layout, it just creates all the cells (content) of the editor aspect and arranges them in a single row.
-Figure~\ref{fig:editor_adjustment} XY shows the difference between the default single-row layout (in the upper left corner) and the fully customized layout (in the bottom right corner) for the \mpsconcept{Element{\_}1} node.
-
-% TODO smazat cervenou sipku z toho obrazku a mozna rozdelit na dve casti (vlozit tam caru): default single-row code layout, optimal fully customized layout
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/editor_adjustment.png}
-	\caption{Projectional editor adjustment of the Element{\_}1 node}
-	\label{fig:editor_adjustment}
-\end{figure*}
-
-In practice, however, the user still has to tweak the editor aspect --- to adjust the code layout --- because the heuristics are not 100\% reliable.
-But it takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language~\cite{ref:javascript} and manually adjusting all editors that needed it, all that in a less than hour time.
-More details are provided later in Section X (examples) --- there we describe this issue further.
-
-\subsection{Text Generation}
-
-The final/last, fourth step, is to create the TextGen aspect for each AST node.
-The purpose of the TextGen aspect is to define, for a language element (AST node), what code in the BaseLanguage will be generated out of it, once we want to compile and run our MPS code/program.
-This is needed to allow/enable users to generate plain-text source code representation out of the AST they built/created inside MPS.
-Again, a TextGen aspect must be created for each language element.
-Basically, it is a single method in the BaseLanguage (or in Java?) for each language element.
-The method generates a string (plain-text) representation of the element, appends the string to the output stream, and then calls TextGen methods on all children of the element.
-
-A very basic example of BaseLanguage code that can be used as the TextGen aspect of an \mpsconcept{Element{\_}1} node, which represents the full XML tag with content, is shown in Figure~\ref{fig:textgen_example}.
-It appends all literals, properties, and children in the same order as they appear in the grammar definition.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=90mm]{./images/textgen_example.png}
-	\caption{Example of a~TextGen aspect}
-	\label{fig:textgen_example}
-\end{figure}
-
-Like in the case of editor, we discovered some interesting challenges related to the code layout, whitespaces, and more (so on).
-We describe our solution to the challenges here in this section.
-
-The reader might have noticed that there is no whitespace handling inside of our SimpleXML grammar (Section XY, running example).
-Grammars in the ANTLRv4 notation are usually written in such a manner that whitespace characters are skipped (ignored).
-However/Nevertheless, TextGen aspect (generator) should/must know between which children there must be a whitespace and where it is forbidden (i.e., where to put some whitespace characters) so that the produced code is a valid one.
-For example, in the case of the \antlrparserrule{element} rule, TextGen must determine that there should be a space in between \antlrparserrule{attribute} and \antlrlexerrule{Name}, while it is not required between \antlrliteral{\textless} and \antlrlexerrule{Name}.
-In general, it is sufficient to put just a single space character at places/locations where there has to be some whitespace in order to make the code valid (syntactically correct).
-
-Layout of the generated code is also important. For example, the user may want to produce more readable code by adding indentation or line breaks.
-
-In fact, the problem of TextGen layout is quite similar to the one we have seen with the projectional editor, described in Section XY.
-Since we are dealing mostly with text-based languages, their representation in projectional editor must be almost the same as their expected plain-text representation (output).
-Therefore, one could use the same approach for editor and TextGen --- adding line breaks and indentation.
-Once we would have some information about the layout for generating better projectional editor, we could leverage the same information and use it to improve TextGen.
-
-Nevertheless, in our current approach, the editor aspects have to be adjusted manually, after the import is done, because for now it is the fastest and the most efficient way how to deal with the problem of customizing the editor to capture a nice layout.
-Therefore, we used a different approach for TextGen --- a really simple heuristics which gives surprisingly good results.
-In the first step, a very basic TextGen is created that inserts spaces in between every two tokens of the language element (AST node).
-Then, as the second step, spaces are eliminated (restricted) from positions where they are not really needed.
-We describe all the cases here:
-
-\begin{enumerate}
-	\item Whenever there is a non-alphabetical literal,	appearing as a plain string token defined in the grammar, that might get recognized by the parser safely without the need for a whitespace separator around, then we omit spaces around it.
-		For example, consider \textbf{'\textless'} in SimpleXML.
-	\item Another group are strings inserted by the user of the language, whose form is constrained by a regular expression.
-		For example, an XML tag's name belongs to this group.
-		They are usually values of properties defined in structure aspects.
-		A space must be preserved only when a neighboring literal may end with an alphabetical character. Otherwise, we can omit the space.
-		In practice, this will eliminate redundant spaces inside of quotes, next to semicolons, and around brackets, but, on the other hand, it will separate language keywords (function, var, in, etc.) from other content by keeping the space character in between them.
-	\item Spaces are omitted also in the case of child nodes that are not present (e.g., when they are optional), and in the case of empty lists of child nodes, so that spaces do not accumulate.
-	\item On the other hand, when two child elements are next to each other, we always insert space.
-	\item Sequences of children are separated with space or with a line break.
-\end{enumerate}
-These few simple heuristics work surprisingly well and give us some very nice results.
-We discuss our experience with XML, JSON, and JavaScript in Section XY (Evaluation).
-
-Figure~\ref{fig:textgen_final} shows an example of a TextGen aspect for the full SimpleXML element.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=95mm]{./images/textgen_final.png}
-	\caption{Generated TextGen aspect for the SimpleXML element}
-	\label{fig:textgen_final}
-\end{figure}
-
-The automatically generated TextGen code can be adjusted manually very easily, so that it generates nicely indented XML code.
-In the case of the example above (in Figure F), we only need to wrap the \mpsconcept{Content{\_}2} child with indentation and change the sequence separator to a new line character.
-The resulting adjusted aspect (code) is shown below in Figure~\ref{fig:textgen_adjusted}.
-
-\begin{figure}[t!]
-	\centering
-	\includegraphics[width=85mm]{./images/textgen_adjusted.png}
-	\caption{Adjusted indentation inside the TextGen aspect}
-	\label{fig:textgen_adjusted}
-\end{figure}
-
-\subsection{Remarks about Grammars}
-
-As we have already shown in previous sections, it is very easy to write a grammar in a way that will cause problems during import into MPS (or once it is imported).
-Problems might occur during the creation of any aspect.
-Here we will look at some general problems/issues that grammar import poses (might pose).
-We will talk about these problems with respect to the challenges/obstacles we tried to overcome and which we described in previous sections (dedicated to structure, editor, and texgen, respectively).
-We also show a few examples of additional possible complications.
-
-\paragraph{Adjusting grammars.}
-There are some cases, where altering the input grammar might yield far better MPS language.
-We will show two examples how the usability of the resulting MPS language can be improved.
-Let us look at the definition of an XML attribute in Figure~\ref{fig:xmlattribute}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
-	\antlrlexerrule{STRING}      :   \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
-	            |   \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'}
-	            ;
-\end{antlr}
-\caption{Definition of an XML attribute}
-\label{fig:xmlattribute}
-\end{figure*}
-
-The original XML grammar has quotes as a part of the value.
-For the resulting MPS language, it would mean that there would be a placeholder for the attribute value that would expect us to input the leading and trailing quote together with the value too each time.
-It would also be marked red (by the syntax checker) unless we enter both quotes inside the value since the regular expression checking for quotes will not match.
-The user might be confused by this and will not be able to tell why his string value is incorrect.
-
-In our SimpleXML language, we adjusted the grammar easily in the manner that we show in Figure~\ref{fig:xmladjustgrammar}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
-	            |   \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'}
-	            ;
-	\antlrlexerrule{TEXT1}       :   \antlrregex{~["]*} ;
-	\antlrlexerrule{TEXT2}       :   \antlrregex{~[']*} ;
-\end{antlr}
-\caption{Adjusted grammar of SimpleXML}
-\label{fig:xmladjustgrammar}
-\end{figure*}
-
-We turned quotes into literals, thus ensuring that they will only appear in the projectional editor as fixed constant cells.
-We will not have to encapsulate the value in them each time.
-The user will only have to choose, which attribute version he wants to use (single or double quotes).
-
-As the second example, we will show the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
-Every statement in JavaScript needs to be either followed by a semicolon, newline, file end or end of the block --- see the Figure~\ref{fig:javascriptstmt}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{eos}            : \antlrlexerrule{SemiColon}
-	               | \antlrlexerrule{EOF}
-	               | {lineTerminatorAhead()}?
-	               | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}?
-	               ;
-	\textcolor{gray}{// Example reference of the eos rule}
-	\antlrparserrule{breakStatement} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos}
-	               ;
-\end{antlr}
-\caption{Statements in JavaScript}
-\label{fig:javascriptstmt}
-\end{figure*}
-
-Because there are multiple options, our import plugin would create a placeholder at the end of every statement.
-Every language element representing a statement would have one child of the \mpsinterface{IEos} interface type.
-This placeholder needs to be manually filled in for each statement.
-
-Since the projectional editor has much bigger power over the form of the code, we (a user) might want to have each statement on a separate line.
-Since we can differentiate between statements on the AST level, we do not need an explicit separator between them.
-This means that we might want to simplify the language and leave the semicolon out, or leave it just as a constant fixed part of the projectional editor, but not as something the user must explicitly fill in.
-Then we can just put each statement on a separate line as it is usual for JavaScript code, but we do not need the semicolon anymore.
-This small adjustment is very quick when done inside the grammar.
-We just change the \antlrparserrule{eos} rule to the form shown in Figure~\ref{fig:eosrule}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{eos} : \antlrliteral{;} ;
-\end{antlr}
-\caption{eos parser rule}
-\label{fig:eosrule}
-\end{figure*}
-
-This way, we can change the grammar in a way that it will not describe the same ECMAScript language as before, but will definitely make our MPS language more usable.
-TextGen aspect may be defined in such a way that puts the semicolon back (after each statement) in the generated plain-text representation. 
-This would be very important especially when JavaScript is available as another base language (which is actually a plan (future work) of the JetBrains company).
-
-Above, we have shown that adjusting the input grammar might be a very fast mean of tweaking/improving the usability of generated MPS language, and sometimes it may be the only proper solution in some complex situations.
-In general, the question we can ask ourselves is whether one aims for a full precise/faithful port of the input language or for an MPS alternative.
-
-\todo{Nahradit slovo "break" v nasledujicich odstavcich (v kontextu "break the parser") za vhodnejsi slovo ktere pusobi vic odborne.}
-
-\paragraph{Breaking grammars and parsers.}
-However, we have found that there is one big problem with grammar adjustment that we would like to point out.
-The problem is that it is very easy to change the input ANTLR grammar in a way that breaks the ANTLR parser generated out of it.
-By breaking we mean that it stops parsing the original language.
-We illustrate the problem on our SimpleXML language.
-
-As stated before, when creating the SimpleXML grammar, we have started off with the original XML grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and did some adjustments to it.
-We ended up with a grammar that can be processed by our method/plugin and creates a nice usable MPS language.
-Nevertheless, we noticed that even though the imported language behaves well enough and mimics the XML language quite nicely, the ANTLR parser generated out of this grammar no longer parses XML successfully.
-Some changes we have made, such as the attribute adjustment, broke the grammar down.
-More precisely, our changes improved the MPS language while breaking down the parser, and we have not even noticed it because, from the perspective of MPS, the imported language still corresponds to XML.
-
-What we are trying to say (show here) is that it is very easy to perform a grammar adjustment that will, at first, seem harmless and valid inside MPS, but will break the parser.
-The problem is that the user performing this change probably will not be aware of breaking it.
-
-The main (underlying) cause of this problem is the way the ANTLR parser is implemented and the quite different purpose we are using the grammar for in our project/tool/method.
-There are many ways how various parsers deal with, for example, token matching.
-For instance, ANTLR introduces so called \emph{greedy} and \emph{non-greedy} operators.
-The greedy way, in which ANTLR matches input on defined tokens and prioritizes their selection, makes some rules very dangerous.
-When a rule matches a wide range of input, it might happen during parsing that it is prioritized over other rules and swallows a lot more input than the author of it intended.
-Usually, these dangerous rules are bound to some parser context, which makes them behave well.
-
-However, for practical reasons, it is very important that grammar is not broken by our custom adjustments (and that the ANTLR parser of the language works properly), because we might need to automatically generate parsers of the adjusted language.
-Imagine this very real scenario (that will be supported in the future):
-\begin{enumerate}
-	\item We have imported the language inside MPS.
-	\item MPS now knows the structure of the language, and we can code in MPS using this language.
-	\item We would expect MPS to be able to load an existing source code, written in this language, from a text file and import it inside MPS.
-	\item We would like to use MPS to safely edit this code, using all the features of MPS.
-	\item We would also like to export the code in a plain-text form again and save it back to the source file.
-\end{enumerate}
-In order to make this happen, we must be able to generate a correctly-working ANTLR parser out of the adjusted source grammar that we used to (that helped us) import the language.
-We must also be able to match nodes of the AST coming out of the ANTLR parser to elements of the MPS language, and we must build the MPS AST out of the ANTLR AST.
-
-\section{Evaluation}
-
-We implemented everything (the proposed approach described above) in the form of an MPS plugin that allows users to carry out the import.
-
-We used a mix of Java and BaseLanguage in our implementation.
-We used the MPS API, which can be called from the BaseLanguage programs too, to generate some of the language elements.
-BaseLanguage is needed to use/call the MPS (language) API, i.e. to programmatically create new language elements and their aspects.
-
-The plugin uses the ANTLR library to work with ANTLRv4 grammar files (to parse them, etc) and several MPS libraries (that implement the MPS API).
-
-For the first step of automated construction of MPS languages from grammars (i.e., for parsing the grammar), the plugin uses the ANTLR library~\cite{ref:ANTLR4} to generate a Java ANTLR parser of the ANTLRv4 notation, because there exists an ANTLRv4 grammar for the ANTLRv4 notation~\cite{ref:ANTLR4reference}.
-
-\todo{
-	We should provide url to the github repository (implementation) and pointer to some package with ANTLR grammars and imported MPS languages (take from Attachments.zip in the master thesis).
-}
-
-\todo{Popsat nase zkusenosti s importem slozitych jazyku (XML, JavaScript, Python, a dalsi).}
-
-From introduction: We will also show some example languages imported with INGRID and discuss our experience.
-
-In order to evaluate INGRID, we tried to import several real/mainstream languages:
-\begin{itemize}
-	\item \textbf{SimpleXML} --- Simplified XML that we used in this thesis for explanatory purposes.
-	\item \textbf{JSON} --- Similar to XML, but this time a full port of the specification.
-	\item \textbf{ECMAScript 5.1} --- Specification of the language known as JavaScript, dated to the year 2011, which is currently the most frequently adopted version.
-\end{itemize}
-Results of the import process are included in the package that is available at \todo{url}.
-The repository \todo{Add url} contains an example MPS project. There are three languages.
-For each example language, you can find its original import, exactly in the form as created by our plugin.
-
-We experimented with import of XML, JavaScript, Python, and maybe some other languages.
-
-These general purpose languages are quite complex when it comes to their structure and the overall syntax variety.
-
-We must emphasize that the automatically imported languages are not ready-to-use full-fledged MPS languages.
-We did not expect to achieve that at this stage, as that is a much bigger, if not impossible, challenge.
-Imported languages contain full structure found in the respective grammars and some more aspect definitions that will help users of the languages in creating code.
-The imported language has to be manually adjusted (tweak, polished) by the end user (developer) and some of its complications have to be resolved manually by a human.
-
-We found out that manual adjusting of the editor/textgen aspect (adding line breaks and indentation) is far more effective and precise than any automated approach we came up with (during our research).
-The imported language can be, in our opinion, adjusted very fast into a more readable form.
-
-The package at \todo{url} contains also adjusted versions of all three languages.
-Then, for each language, there is an adjusted version.
-
-This adjusted version was created from the original import and the author spent no more than between 20 to 60 minutes of working on it.
-The first author spent only a couple of minutes of (manually) adjusting them after the automated import, so that we try to prove our point above.
-More specifically, the JSON language was ready for use in no more than 20 minutes of minor adjusting.
-We chose this language because it does not require us to implement complex aspects, such as type checking or data flows.
-Similarly, the SimpleXML turned out very nice.
-
-Our main goal was to prove that even though our method (plugin) does not create a perfect language, it can be customized very fast into a very usable form.
-We manually tweaked/customized the editor and the TextGen aspect only, structure aspect was left in its original form.
-This justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical (and the best).
-
-We have also tried to use the INGRID tool/plugin to import the JavaScript language, which is exactly the type of a complicated general purpose language that might be of interest to other people.
-There are already some projects, where a manual port of JavaScript is being done.
-One of them is the ECMAScript4MPS~\cite{ref:ECMAScript4MPS} from the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section X (similar projects, related work).
-The result of import by INGRID is quite good, comparable to the PE4MPS project.
-Our language is, of course, missing a lot of more advanced aspects/features that, for example, can help deriving types of expressions and so on.
-These aspects improve the usability by a lot, e.g.
-when the user starts writing numbers, a number literal is inserted, whereas in our import the user has to first insert the concept representing the number literal and then proceed with writing numbers.
-These actions are needed in any MPS language and are expected to be implemented after the import step, \todo{Maybe say that INGRID does not generate these actions/aspects/features yet but it is our future work} as it is not possible to generate them automatically.
-However, when it comes to language structure, concept aliases, and auto-completion, we think that INGRID (our method/plugin) did a very good job.
-The structure aspect is very similar to the ECMAScript4MPS's one, which was created manually over the course of a large number of hours.
-INGRID does that (achieves the same result) fully automatically.
-
-To summarize, import languages such as JavaScript, JSON, MySQL and others is giving nice results.
-
-Results of our experiments with languages such as JSON or JavaScript show that the proposed approach (INGRID) yields very good results, comparable to existing manual ports, and spares the user from hours of tedious, error prone and time-consuming work.
-
-Limitation: the project is lacking proper error reporting and cannot handle cyclic rules (which could be found in Python grammar, for example).
-
 \section{Conclusion and Future Work}
@@
 Another future endeavors will definitely touch the subject of generating parsers that would enable the user to import existing source code into MPS.
-
-\acks
-% TODO
-This research was supported by JetBrains. \todo{Pridat jeste nejaky grant a kompletne preformulovat.}
-
-\bibliographystyle{abbrvnat}
-
-% TODO opravit formatovani jednotlivych zaznamu a vhodne seradit
-\begin{thebibliography}{}
-
-\bibitem{ref:MPS} JetBrains MPS, \url{https://www.jetbrains.com/mps/}
-
-\bibitem{ref:mbeddr} mbeddr, \url{mbeddr.com/}
-
-\bibitem{ref:BaseLanguage} BaseLanguage, \url{confluence.jetbrains.com/display/MPSD34/Base+Language}
-
-\bibitem{ref:PE4MPS} PE4MPS project, \url{https://github.com/mar9000/pe4mps}
-
-\bibitem{ref:PE} PE project, \url{https://github.com/mar9000/pe}
-
-\bibitem{ref:ANTLR2MPS} ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR{\_}MPS}
-
-\bibitem{ref:ANTLR4} ANTLRv4, \url{https://github.com/antlr/antlr4}
-
-\bibitem{ref:ANTLR4reference} T. Parr. The Definitive ANTLR 4 Reference. The Pragmatic Bookshelf, 2013
-
-\bibitem{ref:ANTLR4grammars} ANTLRv4 grammar repository, \url{https://github.com/antlr/grammars-v4}
-
-\bibitem{ref:ECMAScript4MPS} ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}
-
-\bibitem{ref:javascript} JavaScript ANTLRv4 grammar, \url{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}
-
-\end{thebibliography}
-
-\end{document}

### 2016-09-23 chapters/evaluation.tex
Split paper into chapter files

--- 2016-09-21 paper.tex
@@
-\documentclass[10pt]{sigplanconf}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[noline]{algorithm2e}
-\usepackage{comment}
-\usepackage{color}
-\usepackage{times}
-\usepackage{multirow}
-\usepackage{balance}
-\usepackage{adjustbox}
-\usepackage{alltt}
-\usepackage{textcomp}
-\usepackage[usenames]{xcolor}
-
-\setcounter{tocdepth}{3}
-
-% configures style of source code listings
-\lstset{
-	basicstyle=\small\sffamily
-	%basicstyle=\scriptsize\ttfamily
-	%numbers=left
-}
-
-\lstdefinelanguage{java}{
-	morekeywords={abstract,continue,for,new,switch,assert,default,goto,package,synchronized,boolean,stream,event,filter,when,do,if,private,this,break,double,implements,protected,throw,byte,else,import,public,throws,case,enum,instanceof,return,transient,catch,extends,int,short,try,char,final,interface,static,void,class,finally,long,strictfp,volatile,const,float,native,super,while},
-	sensitive=false,
-	morestring=[b]",
-	morestring=[b]',
-	morecomment=[l]{//},
-	morecomment=[s]{/*}{*/)},
-	escapeinside=??,
-	moredelim=[is][\textit]{__}{__},
-	moredelim=[is][\textbf]{_*}{*_}
-}
-\lstnewenvironment{displayjava}
-	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b,xleftmargin=1em,xrightmargin=1em}}{}
-\lstnewenvironment{java}
-	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b}}{}
-
-\newcommand{\code}[1]{{\sf \small #1}\xspace}
-
-\newcommand{\mi}[1]{\mathit{#1}}
-\newcommand{\mr}[1]{\mathrm{#1}}
-\newcommand{\mt}[1]{\mathtt{#1}}
-
-\newcommand{\todo}[1]{{\bfseries #1}}
-
-% for writing comments
-\newcommand{\PP}[1]{\textcolor{blue}{\bfseries PP: #1}} % Pavel Parizek
-\newcommand{\PV}[1]{\textcolor{red}{\bfseries PV: #1}} % Premek Vysoky
-\newcommand{\JB}[1]{\textcolor{orange}{\bfseries JB: #1}} % JetBrains
-
-% colors for the ANTLR environment
-\definecolor{antlrbgcolor}{rgb}{.962, .962, .962}
-\definecolor{antlrparserrulecolor}{RGB}{11,83,148}
-\definecolor{antlrlexerrulecolor}{RGB}{191,144,0}
-\definecolor{antlrliteralcolor}{RGB}{133,32,12}
-\definecolor{antlrregexcolor}{RGB}{56,118,29}
-
-% colors for the MPS environment
-\definecolor{mpsbgcolor}{rgb}{.962, .962, .962}
-\definecolor{mpsconceptcolor}{RGB}{42, 140, 0}
-\definecolor{mpsinterfacecolor}{RGB}{170,0,229}
-
-% coloring commands
-\newcommand{\antlrap}{\textquotesingle}
-\newcommand{\antlrparserrule}[1]{\textcolor{antlrparserrulecolor}{#1}}
-\newcommand{\antlrlexerrule}[1]{\textcolor{antlrlexerrulecolor}{#1}}
-\newcommand{\antlrliteralnoap}[1]{\textcolor{antlrliteralcolor}{#1}}
-\newcommand{\antlrliteral}[1]{\antlrap\antlrliteralnoap{#1}\antlrap}
-\newcommand{\antlrregex}[1]{\textcolor{antlrregexcolor}{#1}}
-% TODO mozna zmenit "mpsconcept" na "mpslangelem"
-\newcommand{\mpsconcept}[1]{\textcolor{mpsconceptcolor}{#1}}
-\newcommand{\mpsinterface}[1]{\textcolor{mpsinterfacecolor}{#1}}
-
-% TODO zmenit barvu pozadi na bilou
-% environment for the ANTLR code (grammars)
-\newenvironment{antlr}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=antlrbgcolor,frame,minipage=394pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
-
-% environment for the MPS code
-\newenvironment{mps}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=mpsbgcolor,frame,minipage=394pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
-
-\begin{document}
-
-% TODO
-%\conferenceinfo{OOPSLA'12,} {October 19--26, 2012, Tuscon, Arizona, USA.}
-%\CopyrightYear{2012}
-%\copyrightdata{978-1-4503-1561-6/12/10}
-
-% TODO
-\title{\todo{INGRID: Creating Languages in MPS by Importing ANTLR Grammars}}
-% alternative: "INGRID: Importing Grammars Into MPS"
-
-\authorinfo{P\v{r}emysl Vysok\'y \and Pavel Par\'izek}
-	{Charles University, Faculty of Mathematics and Physics, Prague, Czech Republic}
-	{premek.vysoky@gmail.com, parizek@d3s.mff.cuni.cz}
-
-% TODO pridat nekoho dalsiho z JetBrains
-% TODO upravit affiliation pro JetBrains -> doplnit "s.r.o" nebo radsi "Inc."
-% TODO muzeme taky napsat jejich emaily pokud budou lidi z JetBrains souhlasit
-\authorinfo{V\'aclav Pech}
-	{JetBrains}
-
-\maketitle
-
-\begin{abstract}
-% TODO
-\todo{write later}
-
-\todo{Copy from master thesis.}
-JetBrains MPS is a language workbench focusing on domain-specific languages.
-Unlike many other language workbenches and IDEs, it uses a projectional editor for code.
-The developer directly manipulates the program in its tree form (AST) and not by editing a text source code.
-This brings many advantages, but on the other hand requires time-consuming and complicated MPS language definition.
-The thesis elaborates on the possibility of automating the process of creating MPS language definition from its grammar description.
-It introduces the MPS editor, evaluates approaches of related projects and describes author's efforts to implement an MPS plugin that allows this import.
-The chosen approach and the selection of tools used for implementation are justified in the thesis.
-We point out important problems that any similar project might deal with and we introduce some possible solutions.
-Furthermore, the thesis contains examples of imported languages, showing the potency of the chosen approach.
-The thesis also aims to lay groundwork for future extensions and suggest possible improvements.
-\end{abstract}
-
-% TODO possible keywords: JetBrains MPS, programming language, grammar, projectional editor
-
-\section{Introduction}
-
-\JB{Odstavec mirne upraven} The prevailing approach to define valid syntaxes for programming languages is through grammars written typically in notations based on (or similar to) EBNF [reference].
-Many popular tools exist for automated generation of parsers from the grammar definitions, such as ANTLR and Bison/Flex.
-The source code of programs in such languages is written in a text editors, usually embedded into an IDE.
-
-\JB{Odstavec mirne upraven} Projectional (or structural) editing offers an alternative approach.
-While the idea emerged as early as in 1970s, it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
-A recent revival of projectional editing has been observed in the area of language workbenches - IDE-like tools that enable the developers to manipulate the actual language definition.
-The Meta-programming System (MPS) [reference] by JetBrains is an open-source language workbench that focuses of Domain Specific Languages and leverages the concept of projectional editor.
-MPS allows developers to design custom languages and then use use them to write code and compile it into executables.
-MPS provides the whole IDE infrastructure for such languages.
-
-\JB{Odstavec mirne upraven} The key idea behind projectional editors is that a developer directly manipulates the AST representation of the source code instead of plain text.
-To avoid the fate of the early projectional editors MPS has put extra effort into making text manipulation feel like editing real text.
-MPS models, although being persisted as XML or binary files, can be versioned using the industry standard source-control systems (Git, Subversion).
-On-line model browsers have been incorporated into some of the popular public source code repositories (BitBucket) [reference] .
-
-\JB{Odstavec mirne upraven} The absence of parsers brings about numerous benefits, which are in particular useful for DSLs, which are frequently used by domain experts, who may not be professional programmers.
-\begin{itemize}
-	\item Since a projectional editors controls the interaction between the user and the code, the developers practically cannot enter syntactically invalid code.
-	\item Languages can be defined in a modular way and multiple languages can be easily combined together or extend one another, while parsers seriously limit language modularity for traditional parsed languages.
-	\item The languages may offer contextual or non-parseable notations, such as tables, diagrams or positional syntax.
-	\item Since the projection is detached from the physical representation of code (AST), authors can define multiple notations and allow the developers to switch between different representations of the code.
-\end{itemize}
-A motivating illustration of the abilities of MPS is the mbeddr project [citation], which extends C with numerous domain-specific constructs and data types.
-We provide more details about MPS in Section 2.1.
-
-\JB{Navrhuji smazat nasledujici odstavec}
-\todo{From thesis.
-	One of these tools is the JetBrains MPS~\cite{ref:MPS}, which approaches languages from a~different point of view.
-	MPS doesn't use the textual representation of code as usual, but rather works with the actual abstract syntax tree that holds the code's structure.
-	This has some positive implications as it gives us a~very powerful tool rich on features but also introduces some new problems we haven't encountered before.
-	One of them being the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-	MPS also becomes a~very powerful tool once the language has already been created and you start using it inside MPS.
-	There is a~huge variety of new languages constructed for MPS that usually solve domain specific problems.
-}
-
-\todo{Pouzivat jen terminy "AST nodes" a "language elements" (to spis). Nepouzivat pojem "concepts" abychom ctenare nematli.}
-
-\JB{Odstavec upraven} Projectional editing introduces some new problems we haven't encountered before.
-One of them is the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-In order to use a language inside MPS, it has to be defined through the MPS infrastructure.
-In particular, one has to create the abstract syntax called Structure (the types of allowed AST nodes), concrete syntax called Editor (projection of the AST on the screen and its interactions with the user), and text generation (to enable creation of plain text files used as input for compilers).
-We provide more details in Section X (Background about MPS).
-
-\JB{Odstavec pridan pro vysvetleni motivace prace} MPS is typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
-These DSLs are then transformed on the AST level by a series of model-to-model transformations to an AST model representing the desired semantics in a General-purpose language (GPL), such as C or Java.
-This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
-This implies that the target GPL must also be defined by the MPS means, to allow the DSLs to have their models transformed into a model in the GPL.
-
-\JB{Odstavec doplnen} However, it requires substantial effort to create a definition of a GPL in MPS.
-Therefore, very few mainstream GPLs are now fully supported by MPS.
-Only Java and C have been implemented to date, which limits the DSLs authors to only target these supported target GPLs.
-We believe that an automated process of importing grammars of GPLs into MPS would encourage more GPLs to be migrated to MPS and thus giving the DSL authors more options.
-
-\todo{From thesis.
-	However, building these languages is a~complicated and time-consuming process and a~lot of effort must be put into them before they are ready for use.
-
-	These examples show that recreating a~full language inside MPS is not an easy task and a~lot of time must be spent on implementing all aspects of the language.
-	A big part of this effort is, however, quite straightforward and can be possibly automated, as we show in this paper. Automation speeds up the process of adoption of new languages to a great degree.
-}
-
-\JB{Odstavec upraven} Since a big part of the language definition process in MPS is quite straightforward and can possibly be automated, we will look into the possibility of automatic import of already existing general purpose languages into MPS using a grammar description of their syntax.
-
-We present INGRID --- a method and a tool for semi-automated construction of MPS definition of programming languages based on their ANTLR grammars.
-We also discuss challenges that we encountered during this process and our solution to them.
-Most of these challenges are general, i.e.
-not MPS specific, as they apply to the relation between grammars defined in EBNF-like style (rules and tokens, everything defined in plain text) and structured (object-oriented) approach used in MPS (and other language workbenches).
-
-\todo{Hodne zduraznit automation, tedy ze chceme maximalizovat automatizaci.}
-The main goal of our/the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible.
-
-\todo{From thesis.
-	Automatic import of already existing languages into MPS from the grammar description of their syntax.
-}
-\todo{Mozna prodat INGRID jako automatickou konverzi gramatik do podoby MPS languages, tedy jako nastroj ktery umoznuje portaci jazyku do prostredi MPS a jejich nasledne pouziti v MPS (treba kombinace s dalsimi jazyky).}
-
-Our primary motivation is to enable better support for mainstream and/or general purpose programming languages (XML, C, C++, JavaScript, Python) in MPS, especially to enable writing programs in these languages using the projectional editor and to combine them with domain-specific languages natively created in MPS.
-\todo{Zduraznit take pouziti importovanych jazyku pro cile generovani z DSL --- to plati zejmena u mainstream general-purpose languages jako treba JavaScript.}
-
-\todo{Briefly mention the biggest challenges.}
-The are related to the way grammar encoding/description of languages are usually written.
-They do not hold any information about the code layout, and contain many rules that do not directly correspond to AST nodes (programming language constructs).
-We discuss them in more detail later (Section X).
-
-\todo{From thesis.
-	INGRID allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
-	The biggest benefit is the possibility to code in these mainstream general purpose languages using the projectional editor.
-}
-
-However, we do not target just the very complex general purpose languages such as JavaScript.
-INGRID can be used also for simple languages.
-We believe that, for simple languages, it is less time consuming to create the grammar by hand (e.g., using the ANLTR notation and toolset) and then create the language definition in MPS by using the INGRID tool, than to create it manually using the MPS GUI --- this is another benefit of INGRID.
-
-Structure of the paper.
-In the beginning, we will describe the MPS editor and related/similar projects.
-Then, we describe our approach to creating language definition in MPS from their grammars, and discuss the major challenges together with our solution.
-We will also show some example languages imported with INGRID and discuss our experience.
-
-\section{Background and Related Work}
-
-We will briefly introduce fundamental basics of the MPS environment.
-We will analyze existing related projects that are trying to accomplish similar goals, and then we will weigh advantages and disadvantages of these approaches.
-
-\subsection{JetBrains MPS}
-
-\JB{Odstavec upraven} JetBrains MPS is a complete language workbench - a development environment that allows developers to define their own languages and then use them to develop code.
-The code can then be transformed into a target language, typically a GPL, such as Java or C.
-
-\JB{Navrhuji odstranit} We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
-Focus on basic/fundamental and relevant features. In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.
-
-\JB{odstavec upraven} MPS differs from typical IDEs in one important aspect - the projectional editor.
-The developer does not work with the textual representation of the source code directly, but rather with its AST (abstract syntax tree) that is the model of the code.
-When users are creating programs using the MPS projectional editor, they are basically assembling the tree (AST) out of predefined building blocks of the imported languages.
-The definition of the language dictates, which where in the AST can certain elements be placed or how they can be nested inside other elements.
-In traditional IDEs it is the parser that constructs an AST using the language's elements.
-
-\JB{Odstavec upraven} The building blocks of MPS models are called nodes.
-Code of any program in MPS is built from nodes and these nodes represent instances of elements defined in one of the languages that the program declares to be using.
-These language elements are called concepts.
-In brief, nodes in code are instances of concepts defined in the language.
-
-\JB{Vyhody/nevyhody diskutujeme v kapitole 1. Navrhuji nasledujici odstavec zrusit nebo text z kapitoly 1. presunout sem}
-Keeping the model of the code in the AST form has several advantages.
-One of them is that MPS then only allows composing the program strictly following the syntax of the language, which results in an inability to actually write syntactically invalid code.
-The code is generated out of the always-valid AST beyond user's reach.
-
-\JB{Odstavec upraven} One of key advantages of projectional editor stems from separation of abstract and concrete syntax.
-While the AST provides an exhaustive representation of the code, the way it is displayed on the screen and the way the user interacts with it are unconstrained.
-The editor can take any visual form and shape.
-The language author can define multiple visualizations and let the developer to choose one that fits best the task at hands - editing, debugging, reviewing, resolving merge conflicts, etc.
-
-For a better understanding, we will give a~small example. Let's assume we have a language with typical if/else structure similar to the one in Java:
-
-\begin{center}
-	\begin{minipage}{.38\textwidth}
-		\begin{alltt}
-			if (\textit{condition})
-			    \textit{then statement block}
-			else
-			    \textit{else statement block}
-		\end{alltt}
-	\end{minipage}
-\end{center}
-
-Its abstract syntax tree representation will look something like shown on Figure~\ref{fig:if_ast}.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[scale=0.6]{./images/if_statement_ast.png}
-  \caption{"If statement" abstract syntax tree}
-  \label{fig:if_ast}
-\end{figure}
-
-The underlying child nodes have their own children and so on.
-The type of nodes can be restricted by the language structure definition.
-For example, we can restrict the condition in the if/else statement to be an expression, statement block to be a list of statements, etc.
-
-\JB{odstavec upraven} Alternative visual representations are possible and they are not bound to be just the textual representation at all.
-For example, blocks for both of the branches can be aligned next to each other and displayed with differend background color, like shown in Figure~\ref{fig:if_editor}.
-
-\begin{figure}[ht]
-	\centering
-	\hspace{-4mm}
-	\includegraphics[scale=0.75]{./images/if_statement_editor.png}
-	\caption{"If statement" projectional editor example}
-	\label{fig:if_editor}
-\end{figure}
-
-\JB{Odstavec mirne upraven}The definition of a language element (the concept) consists of several aspects and each aspect codifies a different part of the nodes' behavior.
-The essential aspects are the following: structure (abstract syntax), editor (concrete syntax - how the code is projected and edited) and textgen (how to transform the AST into textual source code).
-If, instead of generating text directly, the language is supposed to be generated into another language that is available in MPS, the generator aspect must be used to specify the model-transformation rules.
-The structure, editor and textgen aspects are most relevant for the contribution of this paper, so we describe them here (and neglect/ignore the other aspects such as type system and data-flow).
-
-\todo{Highlight the object-oriented hierarchical nature of language definitions in MPS. The environment restricts and guides the developer so that he makes less mistakes.}
-
-Some concepts and design patterns known from object oriented programming can be used in the language definition --- inheritance, interfaces, and so on.
-The AST node can be inherited from a parent (abstract) node and it can implement a special interface node.
-Most importantly, nodes have child nodes and that is how the AST itself is created.
-
-\todo{Zduraznit ze AST node (object) ma nekolik aspectu: structure, editor, textgen. Definice MPS language je organizovana takto --- pres AST nodes, ktere maji nekolik aspectu (ruzne pohledy).}
-
-\paragraph{Structure.}
-The most important aspect of any MPS language is structure. It must be created first for each AST node.
-It specifies core attributes such as name of the AST node, inheritance relationships, possible child nodes (types and cardinality), implemented interfaces, references to other nodes, other properties (fields of any type that can hold values), etc.
-In Figure~\ref{fig:if_statement_structure}, you can see the structure aspect definition for the if statement.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.55]{./images/if_statement_structure.png}
-	\caption{"If statement" structure aspect definition}
-	\label{fig:if_statement_structure}
-\end{figure}
-
-\paragraph{Editor.}
-Editor aspect is where the user defines what the projectional editor representation of a code fragment (an AST node) looks like.
-The editor aspect of an AST node defines its visual representation (of the code) in the projectional editor.
-The user usually incorporates all children, references, and properties inside the representation so that future user of the language can insert their value into these placeholders.
-All elements can be styled using a~language similar to CSS.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/if_statement_editor_definition.png}
-	\caption{"If statement" editor aspect definition}
-	\label{fig:if_editor_definition}
-\end{figure}
-
-\todo{From thesis.
-
-The editor is a module dedicated/responsible for the definition of appearance of an AST node (concept).
-JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells.
-These cells then can be styled to user's liking.
-There are many different types of cells, each behaving a little bit different towards its contents.
-For example, MPS supports various horizontal and vertical lists, etc.
-Extra cells can be added on top --- cells which just specify layout adjustments such as indentation and text color.
-In Figure~\ref{fig:if_editor_definition}, you can see what a real editor definition might look like for the if statement of the Java language (as defined in MPS).
-}
-
-\paragraph{BaseLanguage}
-\todo{Introduce the base language here.}
-A program written in an MPS language cannot be executed directly.
-MPS uses an approach based on translation into a program in the so-called BaseLanguage.
-A plain-text representation of a BaseLanguage program is then passed to a system compiler and the result is executed.
-The module responsible for generating the program in the BaseLanguage is called "textgen" (we describe it briefly in the next paragraph).
-
-BaseLanguage is also an MPS language --- it is the official port of Java for MPS (created by JetBrains, authors of MPS).
-The default base language of MPS is Java, but others are supported too.
-
-Since BaseLanguage is an MPS language, BaseLanguage code is not a (plain) text source code, but again, an AST built out of nodes belonging to the BaseLanguage language.
-This means that generating code inside MPS consists of generating some AST out of BaseLanguage nodes.
-
-Besides representing the default (most common) target of text generation, BaseLanguage is used also for all additional programming inside MPS.
-
-\paragraph{TextGen.}
-The TextGen aspect defines how a certain AST node will be translated into a text source code (in the base language).
-This will allow us to generate source code in text form out of an AST model and ultimately to generate a text program out of the code user writes in the (imported) MPS language.
-The TextGen aspect is basically just one method/procedure definition for each AST node in the language. The method produces string representation of the given AST node and its children.
-MPS then calls this method for the root AST node of the program.
-TextGen aspect has to be defined (especially its more complex functionality) using the BaseLanguage.
-Again, we included an example of the if statement (Figure~\ref{fig:if_statement_textgen}).
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.6]{./images/if_statement_textgen.png}
-	\caption{"If statement" TextGen aspect definition}
-	\label{fig:if_statement_textgen}
-\end{figure}
-
-\todo{From thesis.
-
-After the program's AST is created using the projectional editor, there can be rules defined on how to generate code in plain text out of that tree.
-This module is called "text gen".
-These defined generators can target any language and platform.
-
-\paragraph{Other aspects.}
-MPS supports also other aspects, such as type checking and data-flow analysis, but they are not important for our contribution.
-Moreover, information needed for their automated generation is not available/contained inside a grammar file.
-
-}
-
-\subsection{Similar Projects}
-
-Currently, there exists an almost full port of the Java language called BaseLanguage~\cite{ref:BaseLanguage} extended with some MPS specific features.
-It was imported manually by JetBrains and it is still undergoing changes as Java itself is evolving.
-
-Within the mbeddr project~\cite{ref:mbeddr}, the C language is also manually tailored for MPS.
-
-Now/here we describe three similar projects (with similar goals) that we are aware of.
-We focus on (consider) these three projects: PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
-
-\paragraph{PE4MPS.}
-PE4MPS~\cite{ref:PE4MPS} is a project by \todo{Doplnit autora} that is trying to solve the same problem as we do.
-It solves the lack of information about code layout in grammars by creating a new grammar notation called PE Grammars~\cite{ref:PE} (the PE abbreviation comes from projectional editing).
-It has tw components: PE parser and PE4MPS plugin for MPS.
-Their approach is to mimic an existing grammar notation called ANTLRv4~\cite{ANTLR4} and enrich its syntax with custom (their own) constructs.
-These extensions tell (guide, hint) the parser what the AST node layout should look like.
-Supported extensions (already implemented) include horizontal and vertical lists, and some indentation rules --- however, even these few features make the already not-so-simple syntax of ANTLR v4 much more complex.
-The parser then uses this information when generating the projectional editor for am AST node.
-Author of this approach/project described the PE syntax using the ANTLRv4 notation~\cite{ANTLR4reference} and then used the ANTLR toolset to automatically generate an ANTLRv4 parser for PE grammar files.
-The PE parser reads any PE file and stores the language structure found inside to a custom representation of Java objects.
-
-The PE4MPS plugin/project/tool is built on top of the PE parser.
-This plugin uses the PE parser to build the PE file representation (the aforementioned tree-like structure of Java objects) and then creates AST nodes and their aspects inside MPS.
-The extended syntax brought by PE describes the layout of each element, e.g.
-it tells the plugin that one set of child nodes should be displayed horizontally, another set should be vertical with each child on a separate line and with some indentation, and so on.
-
-A limitation of the PE4MPS approach is that it, from what we understand, does not implement full ANTLR syntax.
-This means that every grammar might need a non-trivial adjustment before its usage.
-One of our goals is to enable import of as many languages out-of-the-box as possible and adopting the full specification.
-
-\paragraph{ANTLR{\_}MPS.}
-ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
-The author of this project is Fabien Campagne
-The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
-However, it does not try to generate the projectional editor at all (i.e., it does not deal with this problem), probably because it is in an early stage of development.
-
-The way this import process of ANTLR{\_}MPS works is quite different from what we have seen in the PE4MPS project, and it is quite complex/complicated to use.
-It works as follows. We give only a brief high-level overview and omit low-level technical details.
-The author created a whole new ANTLRv4 MPS language, which is an MPS port of the grammar notations' syntax.
-To import a language, the user utilizes this MPS ANTLR language.
-The textual grammar is imported automatically into MPS taking the form of the MPS's ported grammar language (so that the textual grammar is converted into MPS nodes, that means an AST node is created for each grammar rule).
-No child-parent relationships in the structure are generated by the tool --- all must be created manually by the user.
-There are no editor nor TextGen aspects created neither.
-
-\paragraph{mps-metabnf.}
-\todo{Popis na zaklade emailove diskuze (Nizozemci).}
-
-\section{Creating Language Definition in MPS}
-
-From all the available notations in which grammars can be written, we decided to use the ANTLRv4 grammar notation~\cite{ANTLR4reference} as the input format (in which the input programming languages have to be defined) for the following reasons.
-\begin{itemize}
-	\item It is a very widely used grammar notation.
-	\item All popular and widely used programming languages have their syntax captured (represented) using this notation --- see the repository~\cite{ref:ANTLR4grammars}.
-	\item There is a whole framework/toolset for generating parsers and editing grammars built around it~\cite{ref:ANTLR4}.
-	\item It supports Java (it is written in Java), which is the implementation language of MPS and also the default BaseLanguage.
-	\item It uses (is based on) an EBNF (Extended Backus-Naur Form) \footnote{https://en.wikipedia.org/wiki/Extended{\_}Backus\%E2\%80\%93Naur{\_}Form} form that it useful because it meets/respects the object oriented design that is used when describing MPS aspects (definitions of MPS languages).
-	\item It has its own syntax captured in its own syntax\footnote{https://github.com/antlr/grammars-v4/tree/master/antlr4} (there is an ANTLRv4 grammar for the ANTLRv4 notation).
-This is a very important feature as it allows us to create a parser for ANTLRv4 grammars so that we can parse grammars of other languages.
-\end{itemize}
-
-We use ANTLRv4 grammars as inputs (and work with them).
-	https://github.com/antlr/antlr4/blob/master/doc/index.md
-	https://github.com/antlr/grammars-v4
-
-Unlike some of the related projects, we did not add any new features into the grammar notation.
-We tried to import grammars as they are.
-
-The generation/import process performed by INGRID consists of three phases (import of structure, usable editor, text generation).
-We will talk about all phases of the generation/import process, describing encountered challenges/obstacles using an example language.
-
-To fulfill our goal of maximal possible automation, we import/generate the full language structure and then do a lot of mundane and time-consuming work connected to creating projectional editors.
-
-Now we describe/introduce the main steps of our approach (that INGRID tool performs): parsing the grammar, creating the structure, editor, and textgen.
-We are able to create/generate the structure aspect and then very basic editor and some default/basic textgen.
-In particular, we did not port the ANTLR notation as an MPS language.
-In the following subsections, we provide some details about our approach to parsing the input grammar and creating the main aspects of each element (each type of AST node) of an MPS language (structure, editor, and textgen).
-
-During our work on this project, we also discovered/encountered challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component), whitespace (editor aspect).
-There is a challenge/issue that we have discovered during our work on this project (in the process of our own implementation), and that is related to construction of the structure aspect.
-Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
-However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
-We describe/discuss these challenges in more detail (also using examples) and our solutions to the challenges inside the respective following subsections.
-
-Result of the construction of a MPS language from the input grammar (import) has to be polished by the human user.
-For example, the aspects not supported by INGRID will be used for adjusting the imported language and improving the usability of the language.
-
-\todo{Briefly discuss how our overall approach differs from similar projects (PE4MPS, etc).}
-There are more ways how to approach the problem of a grammar import.
-
-Our overall approach differs from the PE4MPS project (Section X.Y) especially in the level of automation.
-While the main goal of (our) the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible, usage of PE does not really spare the user any work because it only shifts the manual work from creating projectional editor to writing PE grammars.
-What the user would do inside the MPS UI using the projectional editor is in the case of PE4MPS done using a text editor when editing PE grammars.
-MPS UI is specially designed/made/tailored for this purpose, guides the user through this process, and supports aids such as auto-completion.
-In particular, MPS UI allows the user to perform only changes that result in valid output.
-Editing the grammar stored in a text file using a standard editor is an inferior approach with respect to using MPS (projectional editor), because a text editor is not as user-friendly, and, more importantly, it is an error prone environment.
-
-Main drawback of ANTLR{\_}MPS, and main difference from our approach, is very limited automation: it does not create the parent-child relationships between AST nodes in the structure, and does not generate any editor and textgen aspects (for any node) at all.
-In addition, there is no attention paid to the problem of generating projectional editor and text generation is neglected too.
-We blame all this on the project being in its early stage of development.
-
-In the rest of this section, first we define the language that we will use throughout the paper as the running example.
-Then we will talk about parsing grammar files \todo{Tohle je mozna zbytecne}.
-Lastly, we will describe our approach to tackling generation of language's aspects (structure, editor, and TextGen).
-At some places, we discuss important possible alternatives (as there are several possible approaches).
-
-\subsection{Running Example}
-
-We illustrate the main steps on the example of a simplified XML language, which is defined below (in Figure XY).
-
-\todo{Pouzit jazyk SimpleXML na running exaple.}
-
-We decided to use SimpleXML because it is small enough to fit here, but still can be used to illustrate the main challenges and our solution
-
-Throughout the whole paper, we will be using one example language to illustrate main challenges and our solution.
-The language is a simplified version of XML. We call it \emph{SimpleXML}.
-We created the grammar of the language by taking the official XML ANTLRv4 grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and stripped it down to eliminate/avoid some not so interesting (advanced) features such as XML entities.
-Below in Figure X, you can find the grammar of the SimpleXML language in the ANTLRv4 notation.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\textbf{grammar} \textit{SimpleXML};
-	\antlrparserrule{document}     :   \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
-	\antlrparserrule{prolog}       :   \antlrliteral{<?xml } \antlrparserrule{attribute}* \antlrliteral{?>} ;
-	\antlrparserrule{comment}      :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-	             |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-	             ;
-	\antlrparserrule{attribute}    :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
-	\antlrparserrule{content}      :   \antlrlexerrule{TEXT}
-	             |   \antlrparserrule{element}
-	             |   \antlrparserrule{comment}
-	             |   \antlrlexerrule{CDATA}
-	             ;
-	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-	\textbf{fragment}
-	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
-	             |   \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.}
-	             |   \antlrlexerrule{DIGIT}
-	             ;
-	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
-	\antlrlexerrule{TEXT}         :   \antlrregex{~[<"]*} ;
-	\antlrlexerrule{CDATA}        :   \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
-\end{antlr}
-\caption{SimpleXML grammar}
-\end{figure*}
-
-We have colored the grammar so it is easier to find the way around it (to make navigation easier).
-The grammar contains following basic elements:
-
-\begin{itemize}
-	\item \textbf{ANTLRv4 keywords} -- In black, they are required by the specification.
-	\item \antlrparserrule{\textbf{Parser rules}} -- Stated in blue, they are the rules describing language's structure.
-		The alternatives (that the element on the left side may break into) are separated by the pipe ($|$) character.
-	\item \antlrlexerrule{\textbf{Lexer rules}} -- In yellow, they are rules describing terminal symbols that the parser is matching against the input.
-		Terminals are encoded as string values and regular expressions.
-		The \textbf{fragment} keyword states that the lexer sub-rule is just a helper rule that brings more clarity into the notation but is not visible in the parser output.
-	\item \antlrliteralnoap{\textbf{String values (literals)}} -- In red, always inside of a pair of apostrophes, they are similar to lexer rules.
-		They describe a terminal string, not using regular expressions but by an exact string match.
-	\item \antlrregex{\textbf{Regular expressions}} -- Colored in green, they describe a string token to be matched using a regular expression with the special ANTLRv4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
-	\item \textbf{Rule operators (?+*)} - Left in black, following any element, they quantify the number of occurrences the prepending element can appear in.
-		These operators can also be prepended by an asterisk (*), meaning they are non-greedy.
-\end{itemize}
-
-\subsection{Parsing (the) Grammar}
-
-\todo{Define some name for the AST representing the grammar, and use it consistently throughout the whole paper. Maybe distinguish between "grammar AST" and "MPS structure AST".}
-The first step (stage) of creating the MPS language (of the import process) is parsing the grammar file (written using the ANTLRv4 notation) so that we can get an AST tree representing the grammar.
-
-The abstract syntax tree that comes out of the automatically generated (ANTLR) parser is a little bit complicated and full of information not relevant for our purposes (to us).
-That is why we translate the complex ANTLR AST into our own simplified tree made out of our own objects.
-We discarded as much irrelevant information as possible and only kept information (data) vital/necessary for construction of MPS languages.
-
-Furthermore, we process the grammar in two iterations.
-In the first iteration/pass, we build a tree holding names of other referenced rules in the form of strings.
-In the second iteration, we walk this tree and resolve all these references to real pointers to other rule objects.
-We end up with a clear and simpler representation that is easy to process in the next steps of the import process.
-
-We also had to post-process (simplify) the representation of lexer rules (tokens).
-As a result of the initial parsing, ANTLR parser produces some kind of a tree structure that is representing the lexer rule (meaning/note that lexer rules can also be built from alternatives just like parser rules).
-However, lexer rules are, in fact (basically), just regular expressions used for parsing input source code into tokens.
-We took each/every such tree that represents the lexer rule and flattened it into the equivalent regular expression.
-
-For illustration, consider the lexer rule \antlrlexerrule{Name} from our/the SimpleXML language (Figure XY), together with all its sub-rules, shown in Figure~\ref{fig:lexerrule}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-	\textbf{fragment}
-	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
-             |   \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.}
-             |   \antlrlexerrule{DIGIT}
-             ;
-	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
-\end{antlr}
-\caption{Lexer rule from SimpleXML}
-\label{fig:lexerrule}
-\end{figure*}
-
-Intuitively, we can create a regular expression by gluing elements of each alternative together and then join these alternatives with a classic regex OR ($|$), which is exactly its function in the ANTLRv4 grammar notation.
-We have captures this idea in the form of the recursive algorithm in Figure~\ref{fig:algflatten}, which covers the whole process.
-
-% TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	Flatten(\textit{R}):
-	1) Define \textit{T} as a list of empty string
-	2) For each alternative \textit{A} of the rule \textit{R}:
-	3)     Define \textit{R} = \antlrap\antlrap
-	4)     For each element \textit{E} of \textit{A}:
-	5)         If \textit{E} is not yet flattened:
-	6)             Flatten(\textit{E})
-	7)         \textit{R}.append(\textit{E})
-	8)         \textit{R}.append(\textit{E}.operator)
-	9)     \textit{T}.add(\textit{R})
-	10) Build a string S out of elements t$_{\text{1}}$, t$_{\text{2}}$, ... t$_{\text{n}}$ of \textit{T}:
-	11)     (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
-	12) Return \textit{S}
-\end{antlr}
-\caption{Flatten algorithm}
-\label{fig:algflatten}
-\end{figure*}
-
-The output of \texttt{Flatten(Name)} (output of the Flatten algorithm applied to the \antlrlexerrule{Name} rule) is this regular expression (in this form):
-
-\begin{center}
-	\texttt{([:a-zA-Z])(([:a-zA-Z])|(-)|({\_})|(.)|([0-9]))*}
-\end{center}
-
-\subsection{Structure}
-
-Now, as the next step, we need to translate/transform the AST (the types of AST nodes) derived from the input grammar into elements of the structure aspect of the MPS language definition (into types of AST nodes), and linking the nodes appropriately.
-
-The goal of this second step is to translate (transfer) the syntax (structure) of the input language that is captured by the grammar (ANTLR), into the MPS structure aspect.
-The structure of the MPS language is somewhat similar to the grammar.
-This is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
-The process of transferring grammar rules into the structure of a MPS language presents/involves some interesting challenges.
-The input grammar (EBNF, ANTLR) serves a little bit different purpose than MPS language, and the way grammar rules are usually written down causes some problems that we had to overcome in order to recreate the language inside MPS.
-Note that the challenges and problems are not specific to MPS --- they are common for everyone tackling a similar problem.
-
-We are going to describe two different approaches that we designed --- the straightforward approach (Section~\ref{sect:straightforward_approach}) and the shortcut approach (Section~\ref{sect:shortcut_approach}).
-We will describe two possible approaches that we developed in Section X and Section Y, respectively.
-\todo{Rict neco jako ze ten druhy, slozitejsi, approach vlastne rozsiruje ten prvni, jednodussi.}
-
-However, before focusing on the two approaches, we must describe the basic principles of translation from the AST representing the grammar to the MPS structure, which are common to both approaches.
-The structure is created according to object-oriented principles.
-Each node (element) is represented by an object (and can be treated as such) that may have a parent node, child nodes, and properties of any data type (similar to fields), it may implement any number of interfaces, and it may also hold references to other nodes.
-For every rule of the grammar (for every node in the grammar AST) that breaks into more/multiple alternatives, we must create a node in the MPS structure AST for each alternative of the grammar (every child node).
-Names of the nodes for alternatives are composed from the name of the rule and a number indicating the order of appearance in the grammar.
-For example, considering the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlgrammarelement}
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-             \textcolor{gray}{|   \antlrap<\antlrap Name attribute* \antlrap/>\antlrap}
-             \textcolor{gray}{;}
-\end{antlr}
-\caption{Fragment of the SimpleXML grammar}
-\label{fig:xmlgrammarelement}
-\end{figure*}
-
-Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.75]{./images/element_node_common.png}
-	\caption{Resulting Element{\_}1 node}
-	\label{fig:element_node_common}
-\end{figure}
-
-String literals (e.g., keywords such as \textbf{for}) are omitted/skipped because they will appear only in the corresponding projectional editor.
-The node named \mpsconcept{Element{\_}1} contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name} inside the first alternative of the \antlrparserrule{element} rule,
-Value of the property will be restricted using the regular expression representing the \antlrlexerrule{Name} rule.
-We explain later how the references to parser rules (\antlrparserrule{attribute} and \antlrparserrule{content}) as this is where the approaches differ.
-The main/key difference between these two lies in the way we create children fields for parser rules and in the way we are going to link them together using interface concepts.
-
-\subsubsection{The Straightforward Approach}
-\label{sect:straightforward_approach}
-
-The key idea of this approach is that, for each rule with more than one alternative, an interface node is created.
-Consider the \antlrparserrule{content} rule from our SimpleXML language (Figure~\ref{fig:contentrule}).
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
-\end{antlr}
-\caption{Content rule}
-\label{fig:contentrule}
-\end{figure*}
-
-Any of the four nodes corresponding to alternatives could appear anywhere the \antlrparserrule{content} rule is referenced.
-So we will create one interface node, and then for each alternative of this rule we create one node that implements the interface.
-The resulting fragment of structure will look like the one in Figure~\ref{fig:icontentitf}.
-
-\begin{figure*}[ht]
-\centering
-\begin{mps}
-	\mpsinterface{IContent}   :   \mpsconcept{Content{\_}1}
-           |   \mpsconcept{Content{\_}2}
-           |   \mpsconcept{Content{\_}3}
-           |   \mpsconcept{Content{\_}4}
-\end{mps}
-\caption{IContent interface}
-\label{fig:icontentitf}
-\end{figure*}
-
-References to other parser rules are captured by links to children (interfaces or object nodes) in the AST nodes.
-For the \antlrparserrule{element} rule that is referencing the \antlrparserrule{content} rule, defined as in Figure~\ref{fig:elementrule}, we get the AST node representing the first alternative that is shown in Figure~\ref{fig:element_node_full}.
-There is one \mpsinterface{IElement} interface node that is implemented by two object nodes \mpsconcept{Element{\_}1}, \mpsconcept{Element{\_}2}.
-The node \mpsconcept{Element{\_}1} has two child links pointing to \mpsinterface{IAttribute} and \mpsinterface{IContent}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-\end{antlr}
-\caption{Element rule}
-\label{fig:elementrule}
-\end{figure*}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.75]{./images/element_node_full.png}
-	\caption{Element{\_}1 node's structure aspect}
-	\label{fig:element_node_full}
-\end{figure}
-
-A problem with this approach, which we call \emph{the layer problem}, is related to the grammar rules (and corresponding AST nodes) that represent intermediary layers created only to make the grammar more readable.
-The intermediary layers are created by authors of the grammars in order to make them more readable and more easily (better) maintainable.
-We illustrate this problem (and explain it) on the behavior of auto-completion in MPS.
-Imagine there is a freshly inserted \mpsconcept{Element{\_}1} node (representing the full XML element as is stated in the first alternative of the \antlrparserrule{element} parser rule).
-Now we would like to insert another nested XML element inside.
-MPS auto-completion mechanism would give us four options that are captured in the left part of Figure~\ref{fig:layer_problem}.
-This is because there exist four nodes (objects) that implement the \mpsinterface{IContent} interface as per each alternative of the rule.
-
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/layer_problem.png}
-	\caption{Layer problem in auto-completion}
-	\label{fig:layer_problem}
-\end{figure*}
-
-However, in order to correctly insert another nested element inside, a user would have to first insert the \mpsconcept{Content{\_}2} node inside \mpsconcept{Element{\_}1} that has an \mpsinterface{IElement} child inside (and nothing else).
-Then, as the second step, a user would invoke/trigger/call for the auto-complete again and insert either \mpsconcept{Element{\_}1} or \mpsconcept{Element{\_}2} inside \mpsconcept{Content{\_}2}.
-This mean that a user would have to go through two steps, and in the first stepe either correctly guess (or remember the grammar rule's alternative order), which option/item offered by the auto-complete menu to select.
-
-Similarly, if one would decide to replace the nested \mpsconcept{Element{\_}1} node with, let's say, an XML comment (a \mpsconcept{Comment} node), we need to delete both intermediary layers before we get back to the original \mpsconcept{Content{\_}X} selection (crossroads).
-The user cannot really see what is happening, since the intermediary level has no appearance or indicator.
-This may lead to confusion on the user's part.
-
-The layer problem is solved/addressed by the second approach, which we describe in the next subsection.
-
-\subsubsection{The Shortcut Approach}
-\label{sect:shortcut_approach}
-
-The key idea of this approach is to identify rules and nodes that are at the end of the (derivation, inheritance) chain, and offer them directly to the user through the auto-completion menu.
-We call such \emph{end rules} and \emph{end nodes}, respectively, because they cannot transparently break into more rules/nodes.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
-
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-
-	\antlrparserrule{comment}    :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-\end{antlr}
-\caption{Content parser rule with children}
-\label{fig:contentrulewithchildren}
-\end{figure*}
-
-For example, the \antlrparserrule{content} rule, given in Figure~\ref{fig:contentrulewithchildren} together with its child rules, can ultimately expand into following nodes:
-
-\begin{itemize}
-	\itemsep0em
-	\item \textbf{Content{\_}1} (TEXT)
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}1}
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}2}
-	\item Content{\_}3 $\rightarrow$ \textbf{Comment}
-	\item \textbf{Content{\_}4} (CDATA)
-\end{itemize}
-
-The end nodes are highlighted using the bold font.
-
-Figure XY shows the algorithm (in pseudocode) that finds all end rules for a given parser rule.
-
-% TODO pouzit standardni environment pro algoritmy 
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	FindPathsToEndNodes($R$):
-	1)  Define $L$ as an empty list of list of nodes
-	2)  Return FindPathsToEndNodes($R$, $L$)
-
-	FindPathsToEndNodes($R$, $L$):
-	3)  Define $Q$ as list of list of nodes
-	4)  For each alternative $A$ of rule $R$:
-	5)      $L1$ = Clone($L$)
-	6)      If $A$ is a parser rule with only one element $E$:
-	7)          Let $I$ be interface/concept representing rule $E$
-	8)          $L1$.Add($I$)
-	9)          $P$ = FindPathsToEndNodes($E$, $L1$)
-	10)         $Q$ = Merge($Q$, $P$)
-	11)     Else
-	12)         $L1$.Add($R$)
-	13)         Let $P$ be concept representing $A$
-	14)         $L1$.Add($P$)
-	15)         $Q$.Add($L1$)
-	16) Return $Q$
-\end{antlr}
-\caption{Algorithm for finding end rules}
-\label{fig:shortcut_algorithm}
-\end{figure*}
-
-To find end rules for each parser rule, the algorithm recursively scans through the parser tree that was built before.
-For each parser rule, it tries to find paths leading to some end rule through its alternatives:
-\begin{itemize}
-	\item Whenever it finds an alternative that contains only one element, and this element is a reference to another parser rule, it has found an intermediary level that can be transparently hidden from the user of the language.
-		The process (algorithm) continues by recursively processing alternatives of this "level" rule (since we are not at the end of the chain yet).
-	\item Otherwise, an end rule was found (recursion stops here).
-\end{itemize}
-
-By appending the rule that is leading to current element (line 12) and then appending that alternative's element itself (line 14), we get a path that contains the full path and the target end rule as the last element of the chain.
-The result of this algorithm, for example for the \antlrparserrule{content} rule, equals the listing of the five paths mentioned above.
-
-We do/repeat this for all parser rules and for each rule we get a list of paths that lead from that particular rule to an end node.
-We call these paths as \textbf{shortcuts}, because they provide a shortcut from the rule to the end of the chain.
-Now that we have the shortcuts, we will discuss several ways how to use the shortcuts (them) to improve the MPS language under construction (how to make the language better).
-
-The primary use of shortcuts is to generate options for the auto-completion menus such that only the end nodes are offered.
-This can be technically done in two ways --- (1) through custom auto-complete menus and deletion handlers or (2) by defining a special interface for each rule such that the interface is implemented only by end nodes for the rule.
-Shortcuts have to be considered when inserting nodes but also when deleting them, i.e.
-the whole chain (shortcut path) consisting of possibly multiple intermediary AST nodes (levels) up to the end node must be added, respectively deleted.
-The effect of insertion must be fully reversed when the user wants to delete some end node (the given end node).
-
-\subsection{Editor}
-
-When all types of AST nodes in the language are created, the next step is to define their visual representation in the projectional editor.
-As stated before (in Section X (about JetBrains MPS, part of background)), MPS uses a cellular system that allows placing node's properties and children into a table-like arrangement.
-Cells of different types are available --- for storing property values, references to child nodes, and keywords (string literals), and also cells that influence layout (e.g., indentation).
-
-The purpose (goals) of this (third) step is to create the editor aspect for each language element.
-It should project all attributes of the element --- name, properties, children, and so on --- using the respective cell types.
-
-The main challenge (problem) here is that an input grammar (ANTLR) does not provide/give any help/aid/information regarding the code layout.
-As we already said in the introduction, one of the biggest challenges/problems that arise here is that the grammar description of a language does not hold any information about the code layout.
-The structural rules in the grammar only tell us, what the syntax tree looks like and how the code is broken into nodes.
-In particular, the grammar says nothing about indentation, line breaks, and another formatting.
-
-We tackled this problem by designing an automated learning approach, which generates the layout using several heuristics.
-We present our solution to this problem. It involves some intermediary step whose purpose is to generate the necessary information.
-Our solution is semi-automated (using some heuristics) --- user's help in an interactive manner is needed for final touches (polishing).
-It works as follows.
-
-As input for the algorithm (learning approach), the user has to supply to a set of valid source files written in the imported language, together with the grammar file.
-These four steps/operations are then performed.
-\begin{enumerate}
-	\item Automated generation of an ANTLR parser for the input language using the ANTLR library.
-	\item Modify/extend the code of the generated parser in order to record some additional information.
-	\item Compile the parser into an executable form.
-	\item Use the parser to parse the supplied source code files and extract information about the layout of the code during this process.
-\end{enumerate}
-This way, the imported/created MPS language would inherit the code style/layout directly from the user, i.e. it would learn directly from the code.
-In step 2, the parser is modified such that it remembers/stores the line number and column for each parsed token.
-This is very efficient and it enables to create a nice layout.
-From the line number, the algorithm can easily determine whether there is a line break between some tokens (statements, language elements).
-The column number can be then used to detect indentation.
-However, this approach is still heuristic and may give imprecise (suboptimal) results --- mainly because it is not always possible to precisely match language elements (in the MPS structure) with all tokens that belong to it.
-Note also that quality of the extracted information always depends on the amount of the source code files provided (available, supplied) and on how representative the set of source code files actually is.
-When our algorithm cannot determine some custom layout, it just creates all the cells (content) of the editor aspect and arranges them in a single row.
-Figure~\ref{fig:editor_adjustment} XY shows the difference between the default single-row layout (in the upper left corner) and the fully customized layout (in the bottom right corner) for the \mpsconcept{Element{\_}1} node.
-
-% TODO smazat cervenou sipku z toho obrazku a mozna rozdelit na dve casti (vlozit tam caru): default single-row code layout, optimal fully customized layout
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/editor_adjustment.png}
-	\caption{Projectional editor adjustment of the Element{\_}1 node}
-	\label{fig:editor_adjustment}
-\end{figure*}
-
-In practice, however, the user still has to tweak the editor aspect --- to adjust the code layout --- because the heuristics are not 100\% reliable.
-But it takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language~\cite{ref:javascript} and manually adjusting all editors that needed it, all that in a less than hour time.
-More details are provided later in Section X (examples) --- there we describe this issue further.
-
-\subsection{Text Generation}
-
-The final/last, fourth step, is to create the TextGen aspect for each AST node.
-The purpose of the TextGen aspect is to define, for a language element (AST node), what code in the BaseLanguage will be generated out of it, once we want to compile and run our MPS code/program.
-This is needed to allow/enable users to generate plain-text source code representation out of the AST they built/created inside MPS.
-Again, a TextGen aspect must be created for each language element.
-Basically, it is a single method in the BaseLanguage (or in Java?) for each language element.
-The method generates a string (plain-text) representation of the element, appends the string to the output stream, and then calls TextGen methods on all children of the element.
-
-A very basic example of BaseLanguage code that can be used as the TextGen aspect of an \mpsconcept{Element{\_}1} node, which represents the full XML tag with content, is shown in Figure~\ref{fig:textgen_example}.
-It appends all literals, properties, and children in the same order as they appear in the grammar definition.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=90mm]{./images/textgen_example.png}
-	\caption{Example of a~TextGen aspect}
-	\label{fig:textgen_example}
-\end{figure}
-
-Like in the case of editor, we discovered some interesting challenges related to the code layout, whitespaces, and more (so on).
-We describe our solution to the challenges here in this section.
-
-The reader might have noticed that there is no whitespace handling inside of our SimpleXML grammar (Section XY, running example).
-Grammars in the ANTLRv4 notation are usually written in such a manner that whitespace characters are skipped (ignored).
-However/Nevertheless, TextGen aspect (generator) should/must know between which children there must be a whitespace and where it is forbidden (i.e., where to put some whitespace characters) so that the produced code is a valid one.
-For example, in the case of the \antlrparserrule{element} rule, TextGen must determine that there should be a space in between \antlrparserrule{attribute} and \antlrlexerrule{Name}, while it is not required between \antlrliteral{\textless} and \antlrlexerrule{Name}.
-In general, it is sufficient to put just a single space character at places/locations where there has to be some whitespace in order to make the code valid (syntactically correct).
-
-Layout of the generated code is also important. For example, the user may want to produce more readable code by adding indentation or line breaks.
-
-In fact, the problem of TextGen layout is quite similar to the one we have seen with the projectional editor, described in Section XY.
-Since we are dealing mostly with text-based languages, their representation in projectional editor must be almost the same as their expected plain-text representation (output).
-Therefore, one could use the same approach for editor and TextGen --- adding line breaks and indentation.
-Once we would have some information about the layout for generating better projectional editor, we could leverage the same information and use it to improve TextGen.
-
-Nevertheless, in our current approach, the editor aspects have to be adjusted manually, after the import is done, because for now it is the fastest and the most efficient way how to deal with the problem of customizing the editor to capture a nice layout.
-Therefore, we used a different approach for TextGen --- a really simple heuristics which gives surprisingly good results.
-In the first step, a very basic TextGen is created that inserts spaces in between every two tokens of the language element (AST node).
-Then, as the second step, spaces are eliminated (restricted) from positions where they are not really needed.
-We describe all the cases here:
-
-\begin{enumerate}
-	\item Whenever there is a non-alphabetical literal,	appearing as a plain string token defined in the grammar, that might get recognized by the parser safely without the need for a whitespace separator around, then we omit spaces around it.
-		For example, consider \textbf{'\textless'} in SimpleXML.
-	\item Another group are strings inserted by the user of the language, whose form is constrained by a regular expression.
-		For example, an XML tag's name belongs to this group.
-		They are usually values of properties defined in structure aspects.
-		A space must be preserved only when a neighboring literal may end with an alphabetical character. Otherwise, we can omit the space.
-		In practice, this will eliminate redundant spaces inside of quotes, next to semicolons, and around brackets, but, on the other hand, it will separate language keywords (function, var, in, etc.) from other content by keeping the space character in between them.
-	\item Spaces are omitted also in the case of child nodes that are not present (e.g., when they are optional), and in the case of empty lists of child nodes, so that spaces do not accumulate.
-	\item On the other hand, when two child elements are next to each other, we always insert space.
-	\item Sequences of children are separated with space or with a line break.
-\end{enumerate}
-These few simple heuristics work surprisingly well and give us some very nice results.
-We discuss our experience with XML, JSON, and JavaScript in Section XY (Evaluation).
-
-Figure~\ref{fig:textgen_final} shows an example of a TextGen aspect for the full SimpleXML element.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=95mm]{./images/textgen_final.png}
-	\caption{Generated TextGen aspect for the SimpleXML element}
-	\label{fig:textgen_final}
-\end{figure}
-
-The automatically generated TextGen code can be adjusted manually very easily, so that it generates nicely indented XML code.
-In the case of the example above (in Figure F), we only need to wrap the \mpsconcept{Content{\_}2} child with indentation and change the sequence separator to a new line character.
-The resulting adjusted aspect (code) is shown below in Figure~\ref{fig:textgen_adjusted}.
-
-\begin{figure}[t!]
-	\centering
-	\includegraphics[width=85mm]{./images/textgen_adjusted.png}
-	\caption{Adjusted indentation inside the TextGen aspect}
-	\label{fig:textgen_adjusted}
-\end{figure}
-
-\subsection{Remarks about Grammars}
-
-As we have already shown in previous sections, it is very easy to write a grammar in a way that will cause problems during import into MPS (or once it is imported).
-Problems might occur during the creation of any aspect.
-Here we will look at some general problems/issues that grammar import poses (might pose).
-We will talk about these problems with respect to the challenges/obstacles we tried to overcome and which we described in previous sections (dedicated to structure, editor, and texgen, respectively).
-We also show a few examples of additional possible complications.
-
-\paragraph{Adjusting grammars.}
-There are some cases, where altering the input grammar might yield far better MPS language.
-We will show two examples how the usability of the resulting MPS language can be improved.
-Let us look at the definition of an XML attribute in Figure~\ref{fig:xmlattribute}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
-	\antlrlexerrule{STRING}      :   \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
-	            |   \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'}
-	            ;
-\end{antlr}
-\caption{Definition of an XML attribute}
-\label{fig:xmlattribute}
-\end{figure*}
-
-The original XML grammar has quotes as a part of the value.
-For the resulting MPS language, it would mean that there would be a placeholder for the attribute value that would expect us to input the leading and trailing quote together with the value too each time.
-It would also be marked red (by the syntax checker) unless we enter both quotes inside the value since the regular expression checking for quotes will not match.
-The user might be confused by this and will not be able to tell why his string value is incorrect.
-
-In our SimpleXML language, we adjusted the grammar easily in the manner that we show in Figure~\ref{fig:xmladjustgrammar}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
-	            |   \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'}
-	            ;
-	\antlrlexerrule{TEXT1}       :   \antlrregex{~["]*} ;
-	\antlrlexerrule{TEXT2}       :   \antlrregex{~[']*} ;
-\end{antlr}
-\caption{Adjusted grammar of SimpleXML}
-\label{fig:xmladjustgrammar}
-\end{figure*}
-
-We turned quotes into literals, thus ensuring that they will only appear in the projectional editor as fixed constant cells.
-We will not have to encapsulate the value in them each time.
-The user will only have to choose, which attribute version he wants to use (single or double quotes).
-
-As the second example, we will show the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
-Every statement in JavaScript needs to be either followed by a semicolon, newline, file end or end of the block --- see the Figure~\ref{fig:javascriptstmt}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{eos}            : \antlrlexerrule{SemiColon}
-	               | \antlrlexerrule{EOF}
-	               | {lineTerminatorAhead()}?
-	               | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}?
-	               ;
-	\textcolor{gray}{// Example reference of the eos rule}
-	\antlrparserrule{breakStatement} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos}
-	               ;
-\end{antlr}
-\caption{Statements in JavaScript}
-\label{fig:javascriptstmt}
-\end{figure*}
-
-Because there are multiple options, our import plugin would create a placeholder at the end of every statement.
-Every language element representing a statement would have one child of the \mpsinterface{IEos} interface type.
-This placeholder needs to be manually filled in for each statement.
-
-Since the projectional editor has much bigger power over the form of the code, we (a user) might want to have each statement on a separate line.
-Since we can differentiate between statements on the AST level, we do not need an explicit separator between them.
-This means that we might want to simplify the language and leave the semicolon out, or leave it just as a constant fixed part of the projectional editor, but not as something the user must explicitly fill in.
-Then we can just put each statement on a separate line as it is usual for JavaScript code, but we do not need the semicolon anymore.
-This small adjustment is very quick when done inside the grammar.
-We just change the \antlrparserrule{eos} rule to the form shown in Figure~\ref{fig:eosrule}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{eos} : \antlrliteral{;} ;
-\end{antlr}
-\caption{eos parser rule}
-\label{fig:eosrule}
-\end{figure*}
-
-This way, we can change the grammar in a way that it will not describe the same ECMAScript language as before, but will definitely make our MPS language more usable.
-TextGen aspect may be defined in such a way that puts the semicolon back (after each statement) in the generated plain-text representation. 
-This would be very important especially when JavaScript is available as another base language (which is actually a plan (future work) of the JetBrains company).
-
-Above, we have shown that adjusting the input grammar might be a very fast mean of tweaking/improving the usability of generated MPS language, and sometimes it may be the only proper solution in some complex situations.
-In general, the question we can ask ourselves is whether one aims for a full precise/faithful port of the input language or for an MPS alternative.
-
-\todo{Nahradit slovo "break" v nasledujicich odstavcich (v kontextu "break the parser") za vhodnejsi slovo ktere pusobi vic odborne.}
-
-\paragraph{Breaking grammars and parsers.}
-However, we have found that there is one big problem with grammar adjustment that we would like to point out.
-The problem is that it is very easy to change the input ANTLR grammar in a way that breaks the ANTLR parser generated out of it.
-By breaking we mean that it stops parsing the original language.
-We illustrate the problem on our SimpleXML language.
-
-As stated before, when creating the SimpleXML grammar, we have started off with the original XML grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and did some adjustments to it.
-We ended up with a grammar that can be processed by our method/plugin and creates a nice usable MPS language.
-Nevertheless, we noticed that even though the imported language behaves well enough and mimics the XML language quite nicely, the ANTLR parser generated out of this grammar no longer parses XML successfully.
-Some changes we have made, such as the attribute adjustment, broke the grammar down.
-More precisely, our changes improved the MPS language while breaking down the parser, and we have not even noticed it because, from the perspective of MPS, the imported language still corresponds to XML.
-
-What we are trying to say (show here) is that it is very easy to perform a grammar adjustment that will, at first, seem harmless and valid inside MPS, but will break the parser.
-The problem is that the user performing this change probably will not be aware of breaking it.
-
-The main (underlying) cause of this problem is the way the ANTLR parser is implemented and the quite different purpose we are using the grammar for in our project/tool/method.
-There are many ways how various parsers deal with, for example, token matching.
-For instance, ANTLR introduces so called \emph{greedy} and \emph{non-greedy} operators.
-The greedy way, in which ANTLR matches input on defined tokens and prioritizes their selection, makes some rules very dangerous.
-When a rule matches a wide range of input, it might happen during parsing that it is prioritized over other rules and swallows a lot more input than the author of it intended.
-Usually, these dangerous rules are bound to some parser context, which makes them behave well.
-
-However, for practical reasons, it is very important that grammar is not broken by our custom adjustments (and that the ANTLR parser of the language works properly), because we might need to automatically generate parsers of the adjusted language.
-Imagine this very real scenario (that will be supported in the future):
-\begin{enumerate}
-	\item We have imported the language inside MPS.
-	\item MPS now knows the structure of the language, and we can code in MPS using this language.
-	\item We would expect MPS to be able to load an existing source code, written in this language, from a text file and import it inside MPS.
-	\item We would like to use MPS to safely edit this code, using all the features of MPS.
-	\item We would also like to export the code in a plain-text form again and save it back to the source file.
-\end{enumerate}
-In order to make this happen, we must be able to generate a correctly-working ANTLR parser out of the adjusted source grammar that we used to (that helped us) import the language.
-We must also be able to match nodes of the AST coming out of the ANTLR parser to elements of the MPS language, and we must build the MPS AST out of the ANTLR AST.
-
 \section{Evaluation}
@@
 Limitation: the project is lacking proper error reporting and cannot handle cyclic rules (which could be found in Python grammar, for example).
-
-\section{Conclusion and Future Work}
-
-\todo{Write something.}
-
-There is lot of further follow-up work to be done. Many open problems remain.
-
-Automated generation of the editor aspect, probably using some kind of machine learning.
-Generating nicely designed and user-friendly editors, maybe using artificial intelligence (machine learning) that would evaluate (analyze) existing code in plain text and based on that create the editors.
-
-Follow-up work might look more into the problem of code layout detection, possibly exploring suggested approaches such as the learning approach described in Section XY \todo{Tohle se snad podari aspon castecne implementovat}.
-This could lead to better results in editor and TextGen aspect generation which could definitely use some improvement.
-Another future endeavors will definitely touch the subject of generating parsers that would enable the user to import existing source code into MPS.
-
-\acks
-% TODO
-This research was supported by JetBrains. \todo{Pridat jeste nejaky grant a kompletne preformulovat.}
-
-\bibliographystyle{abbrvnat}
-
-% TODO opravit formatovani jednotlivych zaznamu a vhodne seradit
-\begin{thebibliography}{}
-
-\bibitem{ref:MPS} JetBrains MPS, \url{https://www.jetbrains.com/mps/}
-
-\bibitem{ref:mbeddr} mbeddr, \url{mbeddr.com/}
-
-\bibitem{ref:BaseLanguage} BaseLanguage, \url{confluence.jetbrains.com/display/MPSD34/Base+Language}
-
-\bibitem{ref:PE4MPS} PE4MPS project, \url{https://github.com/mar9000/pe4mps}
-
-\bibitem{ref:PE} PE project, \url{https://github.com/mar9000/pe}
-
-\bibitem{ref:ANTLR2MPS} ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR{\_}MPS}
-
-\bibitem{ref:ANTLR4} ANTLRv4, \url{https://github.com/antlr/antlr4}
-
-\bibitem{ref:ANTLR4reference} T. Parr. The Definitive ANTLR 4 Reference. The Pragmatic Bookshelf, 2013
-
-\bibitem{ref:ANTLR4grammars} ANTLRv4 grammar repository, \url{https://github.com/antlr/grammars-v4}
-
-\bibitem{ref:ECMAScript4MPS} ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}
-
-\bibitem{ref:javascript} JavaScript ANTLRv4 grammar, \url{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}
-
-\end{thebibliography}
-
-\end{document}

### 2016-09-23 chapters/introduction.tex
Split paper into chapter files

--- 2016-09-21 paper.tex
@@
-\documentclass[10pt]{sigplanconf}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[noline]{algorithm2e}
-\usepackage{comment}
-\usepackage{color}
-\usepackage{times}
-\usepackage{multirow}
-\usepackage{balance}
-\usepackage{adjustbox}
-\usepackage{alltt}
-\usepackage{textcomp}
-\usepackage[usenames]{xcolor}
-
-\setcounter{tocdepth}{3}
-
-% configures style of source code listings
-\lstset{
-	basicstyle=\small\sffamily
-	%basicstyle=\scriptsize\ttfamily
-	%numbers=left
-}
-
-\lstdefinelanguage{java}{
-	morekeywords={abstract,continue,for,new,switch,assert,default,goto,package,synchronized,boolean,stream,event,filter,when,do,if,private,this,break,double,implements,protected,throw,byte,else,import,public,throws,case,enum,instanceof,return,transient,catch,extends,int,short,try,char,final,interface,static,void,class,finally,long,strictfp,volatile,const,float,native,super,while},
-	sensitive=false,
-	morestring=[b]",
-	morestring=[b]',
-	morecomment=[l]{//},
-	morecomment=[s]{/*}{*/)},
-	escapeinside=??,
-	moredelim=[is][\textit]{__}{__},
-	moredelim=[is][\textbf]{_*}{*_}
-}
-\lstnewenvironment{displayjava}
-	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b,xleftmargin=1em,xrightmargin=1em}}{}
-\lstnewenvironment{java}
-	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b}}{}
-
-\newcommand{\code}[1]{{\sf \small #1}\xspace}
-
-\newcommand{\mi}[1]{\mathit{#1}}
-\newcommand{\mr}[1]{\mathrm{#1}}
-\newcommand{\mt}[1]{\mathtt{#1}}
-
-\newcommand{\todo}[1]{{\bfseries #1}}
-
-% for writing comments
-\newcommand{\PP}[1]{\textcolor{blue}{\bfseries PP: #1}} % Pavel Parizek
-\newcommand{\PV}[1]{\textcolor{red}{\bfseries PV: #1}} % Premek Vysoky
-\newcommand{\JB}[1]{\textcolor{orange}{\bfseries JB: #1}} % JetBrains
-
-% colors for the ANTLR environment
-\definecolor{antlrbgcolor}{rgb}{.962, .962, .962}
-\definecolor{antlrparserrulecolor}{RGB}{11,83,148}
-\definecolor{antlrlexerrulecolor}{RGB}{191,144,0}
-\definecolor{antlrliteralcolor}{RGB}{133,32,12}
-\definecolor{antlrregexcolor}{RGB}{56,118,29}
-
-% colors for the MPS environment
-\definecolor{mpsbgcolor}{rgb}{.962, .962, .962}
-\definecolor{mpsconceptcolor}{RGB}{42, 140, 0}
-\definecolor{mpsinterfacecolor}{RGB}{170,0,229}
-
-% coloring commands
-\newcommand{\antlrap}{\textquotesingle}
-\newcommand{\antlrparserrule}[1]{\textcolor{antlrparserrulecolor}{#1}}
-\newcommand{\antlrlexerrule}[1]{\textcolor{antlrlexerrulecolor}{#1}}
-\newcommand{\antlrliteralnoap}[1]{\textcolor{antlrliteralcolor}{#1}}
-\newcommand{\antlrliteral}[1]{\antlrap\antlrliteralnoap{#1}\antlrap}
-\newcommand{\antlrregex}[1]{\textcolor{antlrregexcolor}{#1}}
-% TODO mozna zmenit "mpsconcept" na "mpslangelem"
-\newcommand{\mpsconcept}[1]{\textcolor{mpsconceptcolor}{#1}}
-\newcommand{\mpsinterface}[1]{\textcolor{mpsinterfacecolor}{#1}}
-
-% TODO zmenit barvu pozadi na bilou
-% environment for the ANTLR code (grammars)
-\newenvironment{antlr}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=antlrbgcolor,frame,minipage=394pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
-
-% environment for the MPS code
-\newenvironment{mps}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=mpsbgcolor,frame,minipage=394pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
-
-\begin{document}
-
-% TODO
-%\conferenceinfo{OOPSLA'12,} {October 19--26, 2012, Tuscon, Arizona, USA.}
-%\CopyrightYear{2012}
-%\copyrightdata{978-1-4503-1561-6/12/10}
-
-% TODO
-\title{\todo{INGRID: Creating Languages in MPS by Importing ANTLR Grammars}}
-% alternative: "INGRID: Importing Grammars Into MPS"
-
-\authorinfo{P\v{r}emysl Vysok\'y \and Pavel Par\'izek}
-	{Charles University, Faculty of Mathematics and Physics, Prague, Czech Republic}
-	{premek.vysoky@gmail.com, parizek@d3s.mff.cuni.cz}
-
-% TODO pridat nekoho dalsiho z JetBrains
-% TODO upravit affiliation pro JetBrains -> doplnit "s.r.o" nebo radsi "Inc."
-% TODO muzeme taky napsat jejich emaily pokud budou lidi z JetBrains souhlasit
-\authorinfo{V\'aclav Pech}
-	{JetBrains}
-
-\maketitle
-
-\begin{abstract}
-% TODO
-\todo{write later}
-
-\todo{Copy from master thesis.}
-JetBrains MPS is a language workbench focusing on domain-specific languages.
-Unlike many other language workbenches and IDEs, it uses a projectional editor for code.
-The developer directly manipulates the program in its tree form (AST) and not by editing a text source code.
-This brings many advantages, but on the other hand requires time-consuming and complicated MPS language definition.
-The thesis elaborates on the possibility of automating the process of creating MPS language definition from its grammar description.
-It introduces the MPS editor, evaluates approaches of related projects and describes author's efforts to implement an MPS plugin that allows this import.
-The chosen approach and the selection of tools used for implementation are justified in the thesis.
-We point out important problems that any similar project might deal with and we introduce some possible solutions.
-Furthermore, the thesis contains examples of imported languages, showing the potency of the chosen approach.
-The thesis also aims to lay groundwork for future extensions and suggest possible improvements.
-\end{abstract}
-
-% TODO possible keywords: JetBrains MPS, programming language, grammar, projectional editor
-
 \section{Introduction}
@@
 We will also show some example languages imported with INGRID and discuss our experience.
-
-\section{Background and Related Work}
-
-We will briefly introduce fundamental basics of the MPS environment.
-We will analyze existing related projects that are trying to accomplish similar goals, and then we will weigh advantages and disadvantages of these approaches.
-
-\subsection{JetBrains MPS}
-
-\JB{Odstavec upraven} JetBrains MPS is a complete language workbench - a development environment that allows developers to define their own languages and then use them to develop code.
-The code can then be transformed into a target language, typically a GPL, such as Java or C.
-
-\JB{Navrhuji odstranit} We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
-Focus on basic/fundamental and relevant features. In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.
-
-\JB{odstavec upraven} MPS differs from typical IDEs in one important aspect - the projectional editor.
-The developer does not work with the textual representation of the source code directly, but rather with its AST (abstract syntax tree) that is the model of the code.
-When users are creating programs using the MPS projectional editor, they are basically assembling the tree (AST) out of predefined building blocks of the imported languages.
-The definition of the language dictates, which where in the AST can certain elements be placed or how they can be nested inside other elements.
-In traditional IDEs it is the parser that constructs an AST using the language's elements.
-
-\JB{Odstavec upraven} The building blocks of MPS models are called nodes.
-Code of any program in MPS is built from nodes and these nodes represent instances of elements defined in one of the languages that the program declares to be using.
-These language elements are called concepts.
-In brief, nodes in code are instances of concepts defined in the language.
-
-\JB{Vyhody/nevyhody diskutujeme v kapitole 1. Navrhuji nasledujici odstavec zrusit nebo text z kapitoly 1. presunout sem}
-Keeping the model of the code in the AST form has several advantages.
-One of them is that MPS then only allows composing the program strictly following the syntax of the language, which results in an inability to actually write syntactically invalid code.
-The code is generated out of the always-valid AST beyond user's reach.
-
-\JB{Odstavec upraven} One of key advantages of projectional editor stems from separation of abstract and concrete syntax.
-While the AST provides an exhaustive representation of the code, the way it is displayed on the screen and the way the user interacts with it are unconstrained.
-The editor can take any visual form and shape.
-The language author can define multiple visualizations and let the developer to choose one that fits best the task at hands - editing, debugging, reviewing, resolving merge conflicts, etc.
-
-For a better understanding, we will give a~small example. Let's assume we have a language with typical if/else structure similar to the one in Java:
-
-\begin{center}
-	\begin{minipage}{.38\textwidth}
-		\begin{alltt}
-			if (\textit{condition})
-			    \textit{then statement block}
-			else
-			    \textit{else statement block}
-		\end{alltt}
-	\end{minipage}
-\end{center}
-
-Its abstract syntax tree representation will look something like shown on Figure~\ref{fig:if_ast}.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[scale=0.6]{./images/if_statement_ast.png}
-  \caption{"If statement" abstract syntax tree}
-  \label{fig:if_ast}
-\end{figure}
-
-The underlying child nodes have their own children and so on.
-The type of nodes can be restricted by the language structure definition.
-For example, we can restrict the condition in the if/else statement to be an expression, statement block to be a list of statements, etc.
-
-\JB{odstavec upraven} Alternative visual representations are possible and they are not bound to be just the textual representation at all.
-For example, blocks for both of the branches can be aligned next to each other and displayed with differend background color, like shown in Figure~\ref{fig:if_editor}.
-
-\begin{figure}[ht]
-	\centering
-	\hspace{-4mm}
-	\includegraphics[scale=0.75]{./images/if_statement_editor.png}
-	\caption{"If statement" projectional editor example}
-	\label{fig:if_editor}
-\end{figure}
-
-\JB{Odstavec mirne upraven}The definition of a language element (the concept) consists of several aspects and each aspect codifies a different part of the nodes' behavior.
-The essential aspects are the following: structure (abstract syntax), editor (concrete syntax - how the code is projected and edited) and textgen (how to transform the AST into textual source code).
-If, instead of generating text directly, the language is supposed to be generated into another language that is available in MPS, the generator aspect must be used to specify the model-transformation rules.
-The structure, editor and textgen aspects are most relevant for the contribution of this paper, so we describe them here (and neglect/ignore the other aspects such as type system and data-flow).
-
-\todo{Highlight the object-oriented hierarchical nature of language definitions in MPS. The environment restricts and guides the developer so that he makes less mistakes.}
-
-Some concepts and design patterns known from object oriented programming can be used in the language definition --- inheritance, interfaces, and so on.
-The AST node can be inherited from a parent (abstract) node and it can implement a special interface node.
-Most importantly, nodes have child nodes and that is how the AST itself is created.
-
-\todo{Zduraznit ze AST node (object) ma nekolik aspectu: structure, editor, textgen. Definice MPS language je organizovana takto --- pres AST nodes, ktere maji nekolik aspectu (ruzne pohledy).}
-
-\paragraph{Structure.}
-The most important aspect of any MPS language is structure. It must be created first for each AST node.
-It specifies core attributes such as name of the AST node, inheritance relationships, possible child nodes (types and cardinality), implemented interfaces, references to other nodes, other properties (fields of any type that can hold values), etc.
-In Figure~\ref{fig:if_statement_structure}, you can see the structure aspect definition for the if statement.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.55]{./images/if_statement_structure.png}
-	\caption{"If statement" structure aspect definition}
-	\label{fig:if_statement_structure}
-\end{figure}
-
-\paragraph{Editor.}
-Editor aspect is where the user defines what the projectional editor representation of a code fragment (an AST node) looks like.
-The editor aspect of an AST node defines its visual representation (of the code) in the projectional editor.
-The user usually incorporates all children, references, and properties inside the representation so that future user of the language can insert their value into these placeholders.
-All elements can be styled using a~language similar to CSS.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/if_statement_editor_definition.png}
-	\caption{"If statement" editor aspect definition}
-	\label{fig:if_editor_definition}
-\end{figure}
-
-\todo{From thesis.
-
-The editor is a module dedicated/responsible for the definition of appearance of an AST node (concept).
-JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells.
-These cells then can be styled to user's liking.
-There are many different types of cells, each behaving a little bit different towards its contents.
-For example, MPS supports various horizontal and vertical lists, etc.
-Extra cells can be added on top --- cells which just specify layout adjustments such as indentation and text color.
-In Figure~\ref{fig:if_editor_definition}, you can see what a real editor definition might look like for the if statement of the Java language (as defined in MPS).
-}
-
-\paragraph{BaseLanguage}
-\todo{Introduce the base language here.}
-A program written in an MPS language cannot be executed directly.
-MPS uses an approach based on translation into a program in the so-called BaseLanguage.
-A plain-text representation of a BaseLanguage program is then passed to a system compiler and the result is executed.
-The module responsible for generating the program in the BaseLanguage is called "textgen" (we describe it briefly in the next paragraph).
-
-BaseLanguage is also an MPS language --- it is the official port of Java for MPS (created by JetBrains, authors of MPS).
-The default base language of MPS is Java, but others are supported too.
-
-Since BaseLanguage is an MPS language, BaseLanguage code is not a (plain) text source code, but again, an AST built out of nodes belonging to the BaseLanguage language.
-This means that generating code inside MPS consists of generating some AST out of BaseLanguage nodes.
-
-Besides representing the default (most common) target of text generation, BaseLanguage is used also for all additional programming inside MPS.
-
-\paragraph{TextGen.}
-The TextGen aspect defines how a certain AST node will be translated into a text source code (in the base language).
-This will allow us to generate source code in text form out of an AST model and ultimately to generate a text program out of the code user writes in the (imported) MPS language.
-The TextGen aspect is basically just one method/procedure definition for each AST node in the language. The method produces string representation of the given AST node and its children.
-MPS then calls this method for the root AST node of the program.
-TextGen aspect has to be defined (especially its more complex functionality) using the BaseLanguage.
-Again, we included an example of the if statement (Figure~\ref{fig:if_statement_textgen}).
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.6]{./images/if_statement_textgen.png}
-	\caption{"If statement" TextGen aspect definition}
-	\label{fig:if_statement_textgen}
-\end{figure}
-
-\todo{From thesis.
-
-After the program's AST is created using the projectional editor, there can be rules defined on how to generate code in plain text out of that tree.
-This module is called "text gen".
-These defined generators can target any language and platform.
-
-\paragraph{Other aspects.}
-MPS supports also other aspects, such as type checking and data-flow analysis, but they are not important for our contribution.
-Moreover, information needed for their automated generation is not available/contained inside a grammar file.
-
-}
-
-\subsection{Similar Projects}
-
-Currently, there exists an almost full port of the Java language called BaseLanguage~\cite{ref:BaseLanguage} extended with some MPS specific features.
-It was imported manually by JetBrains and it is still undergoing changes as Java itself is evolving.
-
-Within the mbeddr project~\cite{ref:mbeddr}, the C language is also manually tailored for MPS.
-
-Now/here we describe three similar projects (with similar goals) that we are aware of.
-We focus on (consider) these three projects: PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
-
-\paragraph{PE4MPS.}
-PE4MPS~\cite{ref:PE4MPS} is a project by \todo{Doplnit autora} that is trying to solve the same problem as we do.
-It solves the lack of information about code layout in grammars by creating a new grammar notation called PE Grammars~\cite{ref:PE} (the PE abbreviation comes from projectional editing).
-It has tw components: PE parser and PE4MPS plugin for MPS.
-Their approach is to mimic an existing grammar notation called ANTLRv4~\cite{ANTLR4} and enrich its syntax with custom (their own) constructs.
-These extensions tell (guide, hint) the parser what the AST node layout should look like.
-Supported extensions (already implemented) include horizontal and vertical lists, and some indentation rules --- however, even these few features make the already not-so-simple syntax of ANTLR v4 much more complex.
-The parser then uses this information when generating the projectional editor for am AST node.
-Author of this approach/project described the PE syntax using the ANTLRv4 notation~\cite{ANTLR4reference} and then used the ANTLR toolset to automatically generate an ANTLRv4 parser for PE grammar files.
-The PE parser reads any PE file and stores the language structure found inside to a custom representation of Java objects.
-
-The PE4MPS plugin/project/tool is built on top of the PE parser.
-This plugin uses the PE parser to build the PE file representation (the aforementioned tree-like structure of Java objects) and then creates AST nodes and their aspects inside MPS.
-The extended syntax brought by PE describes the layout of each element, e.g.
-it tells the plugin that one set of child nodes should be displayed horizontally, another set should be vertical with each child on a separate line and with some indentation, and so on.
-
-A limitation of the PE4MPS approach is that it, from what we understand, does not implement full ANTLR syntax.
-This means that every grammar might need a non-trivial adjustment before its usage.
-One of our goals is to enable import of as many languages out-of-the-box as possible and adopting the full specification.
-
-\paragraph{ANTLR{\_}MPS.}
-ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
-The author of this project is Fabien Campagne
-The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
-However, it does not try to generate the projectional editor at all (i.e., it does not deal with this problem), probably because it is in an early stage of development.
-
-The way this import process of ANTLR{\_}MPS works is quite different from what we have seen in the PE4MPS project, and it is quite complex/complicated to use.
-It works as follows. We give only a brief high-level overview and omit low-level technical details.
-The author created a whole new ANTLRv4 MPS language, which is an MPS port of the grammar notations' syntax.
-To import a language, the user utilizes this MPS ANTLR language.
-The textual grammar is imported automatically into MPS taking the form of the MPS's ported grammar language (so that the textual grammar is converted into MPS nodes, that means an AST node is created for each grammar rule).
-No child-parent relationships in the structure are generated by the tool --- all must be created manually by the user.
-There are no editor nor TextGen aspects created neither.
-
-\paragraph{mps-metabnf.}
-\todo{Popis na zaklade emailove diskuze (Nizozemci).}
-
-\section{Creating Language Definition in MPS}
-
-From all the available notations in which grammars can be written, we decided to use the ANTLRv4 grammar notation~\cite{ANTLR4reference} as the input format (in which the input programming languages have to be defined) for the following reasons.
-\begin{itemize}
-	\item It is a very widely used grammar notation.
-	\item All popular and widely used programming languages have their syntax captured (represented) using this notation --- see the repository~\cite{ref:ANTLR4grammars}.
-	\item There is a whole framework/toolset for generating parsers and editing grammars built around it~\cite{ref:ANTLR4}.
-	\item It supports Java (it is written in Java), which is the implementation language of MPS and also the default BaseLanguage.
-	\item It uses (is based on) an EBNF (Extended Backus-Naur Form) \footnote{https://en.wikipedia.org/wiki/Extended{\_}Backus\%E2\%80\%93Naur{\_}Form} form that it useful because it meets/respects the object oriented design that is used when describing MPS aspects (definitions of MPS languages).
-	\item It has its own syntax captured in its own syntax\footnote{https://github.com/antlr/grammars-v4/tree/master/antlr4} (there is an ANTLRv4 grammar for the ANTLRv4 notation).
-This is a very important feature as it allows us to create a parser for ANTLRv4 grammars so that we can parse grammars of other languages.
-\end{itemize}
-
-We use ANTLRv4 grammars as inputs (and work with them).
-	https://github.com/antlr/antlr4/blob/master/doc/index.md
-	https://github.com/antlr/grammars-v4
-
-Unlike some of the related projects, we did not add any new features into the grammar notation.
-We tried to import grammars as they are.
-
-The generation/import process performed by INGRID consists of three phases (import of structure, usable editor, text generation).
-We will talk about all phases of the generation/import process, describing encountered challenges/obstacles using an example language.
-
-To fulfill our goal of maximal possible automation, we import/generate the full language structure and then do a lot of mundane and time-consuming work connected to creating projectional editors.
-
-Now we describe/introduce the main steps of our approach (that INGRID tool performs): parsing the grammar, creating the structure, editor, and textgen.
-We are able to create/generate the structure aspect and then very basic editor and some default/basic textgen.
-In particular, we did not port the ANTLR notation as an MPS language.
-In the following subsections, we provide some details about our approach to parsing the input grammar and creating the main aspects of each element (each type of AST node) of an MPS language (structure, editor, and textgen).
-
-During our work on this project, we also discovered/encountered challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component), whitespace (editor aspect).
-There is a challenge/issue that we have discovered during our work on this project (in the process of our own implementation), and that is related to construction of the structure aspect.
-Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
-However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
-We describe/discuss these challenges in more detail (also using examples) and our solutions to the challenges inside the respective following subsections.
-
-Result of the construction of a MPS language from the input grammar (import) has to be polished by the human user.
-For example, the aspects not supported by INGRID will be used for adjusting the imported language and improving the usability of the language.
-
-\todo{Briefly discuss how our overall approach differs from similar projects (PE4MPS, etc).}
-There are more ways how to approach the problem of a grammar import.
-
-Our overall approach differs from the PE4MPS project (Section X.Y) especially in the level of automation.
-While the main goal of (our) the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible, usage of PE does not really spare the user any work because it only shifts the manual work from creating projectional editor to writing PE grammars.
-What the user would do inside the MPS UI using the projectional editor is in the case of PE4MPS done using a text editor when editing PE grammars.
-MPS UI is specially designed/made/tailored for this purpose, guides the user through this process, and supports aids such as auto-completion.
-In particular, MPS UI allows the user to perform only changes that result in valid output.
-Editing the grammar stored in a text file using a standard editor is an inferior approach with respect to using MPS (projectional editor), because a text editor is not as user-friendly, and, more importantly, it is an error prone environment.
-
-Main drawback of ANTLR{\_}MPS, and main difference from our approach, is very limited automation: it does not create the parent-child relationships between AST nodes in the structure, and does not generate any editor and textgen aspects (for any node) at all.
-In addition, there is no attention paid to the problem of generating projectional editor and text generation is neglected too.
-We blame all this on the project being in its early stage of development.
-
-In the rest of this section, first we define the language that we will use throughout the paper as the running example.
-Then we will talk about parsing grammar files \todo{Tohle je mozna zbytecne}.
-Lastly, we will describe our approach to tackling generation of language's aspects (structure, editor, and TextGen).
-At some places, we discuss important possible alternatives (as there are several possible approaches).
-
-\subsection{Running Example}
-
-We illustrate the main steps on the example of a simplified XML language, which is defined below (in Figure XY).
-
-\todo{Pouzit jazyk SimpleXML na running exaple.}
-
-We decided to use SimpleXML because it is small enough to fit here, but still can be used to illustrate the main challenges and our solution
-
-Throughout the whole paper, we will be using one example language to illustrate main challenges and our solution.
-The language is a simplified version of XML. We call it \emph{SimpleXML}.
-We created the grammar of the language by taking the official XML ANTLRv4 grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and stripped it down to eliminate/avoid some not so interesting (advanced) features such as XML entities.
-Below in Figure X, you can find the grammar of the SimpleXML language in the ANTLRv4 notation.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\textbf{grammar} \textit{SimpleXML};
-	\antlrparserrule{document}     :   \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
-	\antlrparserrule{prolog}       :   \antlrliteral{<?xml } \antlrparserrule{attribute}* \antlrliteral{?>} ;
-	\antlrparserrule{comment}      :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-	             |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-	             ;
-	\antlrparserrule{attribute}    :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
-	\antlrparserrule{content}      :   \antlrlexerrule{TEXT}
-	             |   \antlrparserrule{element}
-	             |   \antlrparserrule{comment}
-	             |   \antlrlexerrule{CDATA}
-	             ;
-	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-	\textbf{fragment}
-	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
-	             |   \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.}
-	             |   \antlrlexerrule{DIGIT}
-	             ;
-	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
-	\antlrlexerrule{TEXT}         :   \antlrregex{~[<"]*} ;
-	\antlrlexerrule{CDATA}        :   \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
-\end{antlr}
-\caption{SimpleXML grammar}
-\end{figure*}
-
-We have colored the grammar so it is easier to find the way around it (to make navigation easier).
-The grammar contains following basic elements:
-
-\begin{itemize}
-	\item \textbf{ANTLRv4 keywords} -- In black, they are required by the specification.
-	\item \antlrparserrule{\textbf{Parser rules}} -- Stated in blue, they are the rules describing language's structure.
-		The alternatives (that the element on the left side may break into) are separated by the pipe ($|$) character.
-	\item \antlrlexerrule{\textbf{Lexer rules}} -- In yellow, they are rules describing terminal symbols that the parser is matching against the input.
-		Terminals are encoded as string values and regular expressions.
-		The \textbf{fragment} keyword states that the lexer sub-rule is just a helper rule that brings more clarity into the notation but is not visible in the parser output.
-	\item \antlrliteralnoap{\textbf{String values (literals)}} -- In red, always inside of a pair of apostrophes, they are similar to lexer rules.
-		They describe a terminal string, not using regular expressions but by an exact string match.
-	\item \antlrregex{\textbf{Regular expressions}} -- Colored in green, they describe a string token to be matched using a regular expression with the special ANTLRv4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
-	\item \textbf{Rule operators (?+*)} - Left in black, following any element, they quantify the number of occurrences the prepending element can appear in.
-		These operators can also be prepended by an asterisk (*), meaning they are non-greedy.
-\end{itemize}
-
-\subsection{Parsing (the) Grammar}
-
-\todo{Define some name for the AST representing the grammar, and use it consistently throughout the whole paper. Maybe distinguish between "grammar AST" and "MPS structure AST".}
-The first step (stage) of creating the MPS language (of the import process) is parsing the grammar file (written using the ANTLRv4 notation) so that we can get an AST tree representing the grammar.
-
-The abstract syntax tree that comes out of the automatically generated (ANTLR) parser is a little bit complicated and full of information not relevant for our purposes (to us).
-That is why we translate the complex ANTLR AST into our own simplified tree made out of our own objects.
-We discarded as much irrelevant information as possible and only kept information (data) vital/necessary for construction of MPS languages.
-
-Furthermore, we process the grammar in two iterations.
-In the first iteration/pass, we build a tree holding names of other referenced rules in the form of strings.
-In the second iteration, we walk this tree and resolve all these references to real pointers to other rule objects.
-We end up with a clear and simpler representation that is easy to process in the next steps of the import process.
-
-We also had to post-process (simplify) the representation of lexer rules (tokens).
-As a result of the initial parsing, ANTLR parser produces some kind of a tree structure that is representing the lexer rule (meaning/note that lexer rules can also be built from alternatives just like parser rules).
-However, lexer rules are, in fact (basically), just regular expressions used for parsing input source code into tokens.
-We took each/every such tree that represents the lexer rule and flattened it into the equivalent regular expression.
-
-For illustration, consider the lexer rule \antlrlexerrule{Name} from our/the SimpleXML language (Figure XY), together with all its sub-rules, shown in Figure~\ref{fig:lexerrule}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-	\textbf{fragment}
-	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
-             |   \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.}
-             |   \antlrlexerrule{DIGIT}
-             ;
-	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
-\end{antlr}
-\caption{Lexer rule from SimpleXML}
-\label{fig:lexerrule}
-\end{figure*}
-
-Intuitively, we can create a regular expression by gluing elements of each alternative together and then join these alternatives with a classic regex OR ($|$), which is exactly its function in the ANTLRv4 grammar notation.
-We have captures this idea in the form of the recursive algorithm in Figure~\ref{fig:algflatten}, which covers the whole process.
-
-% TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	Flatten(\textit{R}):
-	1) Define \textit{T} as a list of empty string
-	2) For each alternative \textit{A} of the rule \textit{R}:
-	3)     Define \textit{R} = \antlrap\antlrap
-	4)     For each element \textit{E} of \textit{A}:
-	5)         If \textit{E} is not yet flattened:
-	6)             Flatten(\textit{E})
-	7)         \textit{R}.append(\textit{E})
-	8)         \textit{R}.append(\textit{E}.operator)
-	9)     \textit{T}.add(\textit{R})
-	10) Build a string S out of elements t$_{\text{1}}$, t$_{\text{2}}$, ... t$_{\text{n}}$ of \textit{T}:
-	11)     (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
-	12) Return \textit{S}
-\end{antlr}
-\caption{Flatten algorithm}
-\label{fig:algflatten}
-\end{figure*}
-
-The output of \texttt{Flatten(Name)} (output of the Flatten algorithm applied to the \antlrlexerrule{Name} rule) is this regular expression (in this form):
-
-\begin{center}
-	\texttt{([:a-zA-Z])(([:a-zA-Z])|(-)|({\_})|(.)|([0-9]))*}
-\end{center}
-
-\subsection{Structure}
-
-Now, as the next step, we need to translate/transform the AST (the types of AST nodes) derived from the input grammar into elements of the structure aspect of the MPS language definition (into types of AST nodes), and linking the nodes appropriately.
-
-The goal of this second step is to translate (transfer) the syntax (structure) of the input language that is captured by the grammar (ANTLR), into the MPS structure aspect.
-The structure of the MPS language is somewhat similar to the grammar.
-This is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
-The process of transferring grammar rules into the structure of a MPS language presents/involves some interesting challenges.
-The input grammar (EBNF, ANTLR) serves a little bit different purpose than MPS language, and the way grammar rules are usually written down causes some problems that we had to overcome in order to recreate the language inside MPS.
-Note that the challenges and problems are not specific to MPS --- they are common for everyone tackling a similar problem.
-
-We are going to describe two different approaches that we designed --- the straightforward approach (Section~\ref{sect:straightforward_approach}) and the shortcut approach (Section~\ref{sect:shortcut_approach}).
-We will describe two possible approaches that we developed in Section X and Section Y, respectively.
-\todo{Rict neco jako ze ten druhy, slozitejsi, approach vlastne rozsiruje ten prvni, jednodussi.}
-
-However, before focusing on the two approaches, we must describe the basic principles of translation from the AST representing the grammar to the MPS structure, which are common to both approaches.
-The structure is created according to object-oriented principles.
-Each node (element) is represented by an object (and can be treated as such) that may have a parent node, child nodes, and properties of any data type (similar to fields), it may implement any number of interfaces, and it may also hold references to other nodes.
-For every rule of the grammar (for every node in the grammar AST) that breaks into more/multiple alternatives, we must create a node in the MPS structure AST for each alternative of the grammar (every child node).
-Names of the nodes for alternatives are composed from the name of the rule and a number indicating the order of appearance in the grammar.
-For example, considering the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlgrammarelement}
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-             \textcolor{gray}{|   \antlrap<\antlrap Name attribute* \antlrap/>\antlrap}
-             \textcolor{gray}{;}
-\end{antlr}
-\caption{Fragment of the SimpleXML grammar}
-\label{fig:xmlgrammarelement}
-\end{figure*}
-
-Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.75]{./images/element_node_common.png}
-	\caption{Resulting Element{\_}1 node}
-	\label{fig:element_node_common}
-\end{figure}
-
-String literals (e.g., keywords such as \textbf{for}) are omitted/skipped because they will appear only in the corresponding projectional editor.
-The node named \mpsconcept{Element{\_}1} contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name} inside the first alternative of the \antlrparserrule{element} rule,
-Value of the property will be restricted using the regular expression representing the \antlrlexerrule{Name} rule.
-We explain later how the references to parser rules (\antlrparserrule{attribute} and \antlrparserrule{content}) as this is where the approaches differ.
-The main/key difference between these two lies in the way we create children fields for parser rules and in the way we are going to link them together using interface concepts.
-
-\subsubsection{The Straightforward Approach}
-\label{sect:straightforward_approach}
-
-The key idea of this approach is that, for each rule with more than one alternative, an interface node is created.
-Consider the \antlrparserrule{content} rule from our SimpleXML language (Figure~\ref{fig:contentrule}).
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
-\end{antlr}
-\caption{Content rule}
-\label{fig:contentrule}
-\end{figure*}
-
-Any of the four nodes corresponding to alternatives could appear anywhere the \antlrparserrule{content} rule is referenced.
-So we will create one interface node, and then for each alternative of this rule we create one node that implements the interface.
-The resulting fragment of structure will look like the one in Figure~\ref{fig:icontentitf}.
-
-\begin{figure*}[ht]
-\centering
-\begin{mps}
-	\mpsinterface{IContent}   :   \mpsconcept{Content{\_}1}
-           |   \mpsconcept{Content{\_}2}
-           |   \mpsconcept{Content{\_}3}
-           |   \mpsconcept{Content{\_}4}
-\end{mps}
-\caption{IContent interface}
-\label{fig:icontentitf}
-\end{figure*}
-
-References to other parser rules are captured by links to children (interfaces or object nodes) in the AST nodes.
-For the \antlrparserrule{element} rule that is referencing the \antlrparserrule{content} rule, defined as in Figure~\ref{fig:elementrule}, we get the AST node representing the first alternative that is shown in Figure~\ref{fig:element_node_full}.
-There is one \mpsinterface{IElement} interface node that is implemented by two object nodes \mpsconcept{Element{\_}1}, \mpsconcept{Element{\_}2}.
-The node \mpsconcept{Element{\_}1} has two child links pointing to \mpsinterface{IAttribute} and \mpsinterface{IContent}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-\end{antlr}
-\caption{Element rule}
-\label{fig:elementrule}
-\end{figure*}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.75]{./images/element_node_full.png}
-	\caption{Element{\_}1 node's structure aspect}
-	\label{fig:element_node_full}
-\end{figure}
-
-A problem with this approach, which we call \emph{the layer problem}, is related to the grammar rules (and corresponding AST nodes) that represent intermediary layers created only to make the grammar more readable.
-The intermediary layers are created by authors of the grammars in order to make them more readable and more easily (better) maintainable.
-We illustrate this problem (and explain it) on the behavior of auto-completion in MPS.
-Imagine there is a freshly inserted \mpsconcept{Element{\_}1} node (representing the full XML element as is stated in the first alternative of the \antlrparserrule{element} parser rule).
-Now we would like to insert another nested XML element inside.
-MPS auto-completion mechanism would give us four options that are captured in the left part of Figure~\ref{fig:layer_problem}.
-This is because there exist four nodes (objects) that implement the \mpsinterface{IContent} interface as per each alternative of the rule.
-
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/layer_problem.png}
-	\caption{Layer problem in auto-completion}
-	\label{fig:layer_problem}
-\end{figure*}
-
-However, in order to correctly insert another nested element inside, a user would have to first insert the \mpsconcept{Content{\_}2} node inside \mpsconcept{Element{\_}1} that has an \mpsinterface{IElement} child inside (and nothing else).
-Then, as the second step, a user would invoke/trigger/call for the auto-complete again and insert either \mpsconcept{Element{\_}1} or \mpsconcept{Element{\_}2} inside \mpsconcept{Content{\_}2}.
-This mean that a user would have to go through two steps, and in the first stepe either correctly guess (or remember the grammar rule's alternative order), which option/item offered by the auto-complete menu to select.
-
-Similarly, if one would decide to replace the nested \mpsconcept{Element{\_}1} node with, let's say, an XML comment (a \mpsconcept{Comment} node), we need to delete both intermediary layers before we get back to the original \mpsconcept{Content{\_}X} selection (crossroads).
-The user cannot really see what is happening, since the intermediary level has no appearance or indicator.
-This may lead to confusion on the user's part.
-
-The layer problem is solved/addressed by the second approach, which we describe in the next subsection.
-
-\subsubsection{The Shortcut Approach}
-\label{sect:shortcut_approach}
-
-The key idea of this approach is to identify rules and nodes that are at the end of the (derivation, inheritance) chain, and offer them directly to the user through the auto-completion menu.
-We call such \emph{end rules} and \emph{end nodes}, respectively, because they cannot transparently break into more rules/nodes.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
-
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-
-	\antlrparserrule{comment}    :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-\end{antlr}
-\caption{Content parser rule with children}
-\label{fig:contentrulewithchildren}
-\end{figure*}
-
-For example, the \antlrparserrule{content} rule, given in Figure~\ref{fig:contentrulewithchildren} together with its child rules, can ultimately expand into following nodes:
-
-\begin{itemize}
-	\itemsep0em
-	\item \textbf{Content{\_}1} (TEXT)
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}1}
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}2}
-	\item Content{\_}3 $\rightarrow$ \textbf{Comment}
-	\item \textbf{Content{\_}4} (CDATA)
-\end{itemize}
-
-The end nodes are highlighted using the bold font.
-
-Figure XY shows the algorithm (in pseudocode) that finds all end rules for a given parser rule.
-
-% TODO pouzit standardni environment pro algoritmy 
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	FindPathsToEndNodes($R$):
-	1)  Define $L$ as an empty list of list of nodes
-	2)  Return FindPathsToEndNodes($R$, $L$)
-
-	FindPathsToEndNodes($R$, $L$):
-	3)  Define $Q$ as list of list of nodes
-	4)  For each alternative $A$ of rule $R$:
-	5)      $L1$ = Clone($L$)
-	6)      If $A$ is a parser rule with only one element $E$:
-	7)          Let $I$ be interface/concept representing rule $E$
-	8)          $L1$.Add($I$)
-	9)          $P$ = FindPathsToEndNodes($E$, $L1$)
-	10)         $Q$ = Merge($Q$, $P$)
-	11)     Else
-	12)         $L1$.Add($R$)
-	13)         Let $P$ be concept representing $A$
-	14)         $L1$.Add($P$)
-	15)         $Q$.Add($L1$)
-	16) Return $Q$
-\end{antlr}
-\caption{Algorithm for finding end rules}
-\label{fig:shortcut_algorithm}
-\end{figure*}
-
-To find end rules for each parser rule, the algorithm recursively scans through the parser tree that was built before.
-For each parser rule, it tries to find paths leading to some end rule through its alternatives:
-\begin{itemize}
-	\item Whenever it finds an alternative that contains only one element, and this element is a reference to another parser rule, it has found an intermediary level that can be transparently hidden from the user of the language.
-		The process (algorithm) continues by recursively processing alternatives of this "level" rule (since we are not at the end of the chain yet).
-	\item Otherwise, an end rule was found (recursion stops here).
-\end{itemize}
-
-By appending the rule that is leading to current element (line 12) and then appending that alternative's element itself (line 14), we get a path that contains the full path and the target end rule as the last element of the chain.
-The result of this algorithm, for example for the \antlrparserrule{content} rule, equals the listing of the five paths mentioned above.
-
-We do/repeat this for all parser rules and for each rule we get a list of paths that lead from that particular rule to an end node.
-We call these paths as \textbf{shortcuts}, because they provide a shortcut from the rule to the end of the chain.
-Now that we have the shortcuts, we will discuss several ways how to use the shortcuts (them) to improve the MPS language under construction (how to make the language better).
-
-The primary use of shortcuts is to generate options for the auto-completion menus such that only the end nodes are offered.
-This can be technically done in two ways --- (1) through custom auto-complete menus and deletion handlers or (2) by defining a special interface for each rule such that the interface is implemented only by end nodes for the rule.
-Shortcuts have to be considered when inserting nodes but also when deleting them, i.e.
-the whole chain (shortcut path) consisting of possibly multiple intermediary AST nodes (levels) up to the end node must be added, respectively deleted.
-The effect of insertion must be fully reversed when the user wants to delete some end node (the given end node).
-
-\subsection{Editor}
-
-When all types of AST nodes in the language are created, the next step is to define their visual representation in the projectional editor.
-As stated before (in Section X (about JetBrains MPS, part of background)), MPS uses a cellular system that allows placing node's properties and children into a table-like arrangement.
-Cells of different types are available --- for storing property values, references to child nodes, and keywords (string literals), and also cells that influence layout (e.g., indentation).
-
-The purpose (goals) of this (third) step is to create the editor aspect for each language element.
-It should project all attributes of the element --- name, properties, children, and so on --- using the respective cell types.
-
-The main challenge (problem) here is that an input grammar (ANTLR) does not provide/give any help/aid/information regarding the code layout.
-As we already said in the introduction, one of the biggest challenges/problems that arise here is that the grammar description of a language does not hold any information about the code layout.
-The structural rules in the grammar only tell us, what the syntax tree looks like and how the code is broken into nodes.
-In particular, the grammar says nothing about indentation, line breaks, and another formatting.
-
-We tackled this problem by designing an automated learning approach, which generates the layout using several heuristics.
-We present our solution to this problem. It involves some intermediary step whose purpose is to generate the necessary information.
-Our solution is semi-automated (using some heuristics) --- user's help in an interactive manner is needed for final touches (polishing).
-It works as follows.
-
-As input for the algorithm (learning approach), the user has to supply to a set of valid source files written in the imported language, together with the grammar file.
-These four steps/operations are then performed.
-\begin{enumerate}
-	\item Automated generation of an ANTLR parser for the input language using the ANTLR library.
-	\item Modify/extend the code of the generated parser in order to record some additional information.
-	\item Compile the parser into an executable form.
-	\item Use the parser to parse the supplied source code files and extract information about the layout of the code during this process.
-\end{enumerate}
-This way, the imported/created MPS language would inherit the code style/layout directly from the user, i.e. it would learn directly from the code.
-In step 2, the parser is modified such that it remembers/stores the line number and column for each parsed token.
-This is very efficient and it enables to create a nice layout.
-From the line number, the algorithm can easily determine whether there is a line break between some tokens (statements, language elements).
-The column number can be then used to detect indentation.
-However, this approach is still heuristic and may give imprecise (suboptimal) results --- mainly because it is not always possible to precisely match language elements (in the MPS structure) with all tokens that belong to it.
-Note also that quality of the extracted information always depends on the amount of the source code files provided (available, supplied) and on how representative the set of source code files actually is.
-When our algorithm cannot determine some custom layout, it just creates all the cells (content) of the editor aspect and arranges them in a single row.
-Figure~\ref{fig:editor_adjustment} XY shows the difference between the default single-row layout (in the upper left corner) and the fully customized layout (in the bottom right corner) for the \mpsconcept{Element{\_}1} node.
-
-% TODO smazat cervenou sipku z toho obrazku a mozna rozdelit na dve casti (vlozit tam caru): default single-row code layout, optimal fully customized layout
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/editor_adjustment.png}
-	\caption{Projectional editor adjustment of the Element{\_}1 node}
-	\label{fig:editor_adjustment}
-\end{figure*}
-
-In practice, however, the user still has to tweak the editor aspect --- to adjust the code layout --- because the heuristics are not 100\% reliable.
-But it takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language~\cite{ref:javascript} and manually adjusting all editors that needed it, all that in a less than hour time.
-More details are provided later in Section X (examples) --- there we describe this issue further.
-
-\subsection{Text Generation}
-
-The final/last, fourth step, is to create the TextGen aspect for each AST node.
-The purpose of the TextGen aspect is to define, for a language element (AST node), what code in the BaseLanguage will be generated out of it, once we want to compile and run our MPS code/program.
-This is needed to allow/enable users to generate plain-text source code representation out of the AST they built/created inside MPS.
-Again, a TextGen aspect must be created for each language element.
-Basically, it is a single method in the BaseLanguage (or in Java?) for each language element.
-The method generates a string (plain-text) representation of the element, appends the string to the output stream, and then calls TextGen methods on all children of the element.
-
-A very basic example of BaseLanguage code that can be used as the TextGen aspect of an \mpsconcept{Element{\_}1} node, which represents the full XML tag with content, is shown in Figure~\ref{fig:textgen_example}.
-It appends all literals, properties, and children in the same order as they appear in the grammar definition.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=90mm]{./images/textgen_example.png}
-	\caption{Example of a~TextGen aspect}
-	\label{fig:textgen_example}
-\end{figure}
-
-Like in the case of editor, we discovered some interesting challenges related to the code layout, whitespaces, and more (so on).
-We describe our solution to the challenges here in this section.
-
-The reader might have noticed that there is no whitespace handling inside of our SimpleXML grammar (Section XY, running example).
-Grammars in the ANTLRv4 notation are usually written in such a manner that whitespace characters are skipped (ignored).
-However/Nevertheless, TextGen aspect (generator) should/must know between which children there must be a whitespace and where it is forbidden (i.e., where to put some whitespace characters) so that the produced code is a valid one.
-For example, in the case of the \antlrparserrule{element} rule, TextGen must determine that there should be a space in between \antlrparserrule{attribute} and \antlrlexerrule{Name}, while it is not required between \antlrliteral{\textless} and \antlrlexerrule{Name}.
-In general, it is sufficient to put just a single space character at places/locations where there has to be some whitespace in order to make the code valid (syntactically correct).
-
-Layout of the generated code is also important. For example, the user may want to produce more readable code by adding indentation or line breaks.
-
-In fact, the problem of TextGen layout is quite similar to the one we have seen with the projectional editor, described in Section XY.
-Since we are dealing mostly with text-based languages, their representation in projectional editor must be almost the same as their expected plain-text representation (output).
-Therefore, one could use the same approach for editor and TextGen --- adding line breaks and indentation.
-Once we would have some information about the layout for generating better projectional editor, we could leverage the same information and use it to improve TextGen.
-
-Nevertheless, in our current approach, the editor aspects have to be adjusted manually, after the import is done, because for now it is the fastest and the most efficient way how to deal with the problem of customizing the editor to capture a nice layout.
-Therefore, we used a different approach for TextGen --- a really simple heuristics which gives surprisingly good results.
-In the first step, a very basic TextGen is created that inserts spaces in between every two tokens of the language element (AST node).
-Then, as the second step, spaces are eliminated (restricted) from positions where they are not really needed.
-We describe all the cases here:
-
-\begin{enumerate}
-	\item Whenever there is a non-alphabetical literal,	appearing as a plain string token defined in the grammar, that might get recognized by the parser safely without the need for a whitespace separator around, then we omit spaces around it.
-		For example, consider \textbf{'\textless'} in SimpleXML.
-	\item Another group are strings inserted by the user of the language, whose form is constrained by a regular expression.
-		For example, an XML tag's name belongs to this group.
-		They are usually values of properties defined in structure aspects.
-		A space must be preserved only when a neighboring literal may end with an alphabetical character. Otherwise, we can omit the space.
-		In practice, this will eliminate redundant spaces inside of quotes, next to semicolons, and around brackets, but, on the other hand, it will separate language keywords (function, var, in, etc.) from other content by keeping the space character in between them.
-	\item Spaces are omitted also in the case of child nodes that are not present (e.g., when they are optional), and in the case of empty lists of child nodes, so that spaces do not accumulate.
-	\item On the other hand, when two child elements are next to each other, we always insert space.
-	\item Sequences of children are separated with space or with a line break.
-\end{enumerate}
-These few simple heuristics work surprisingly well and give us some very nice results.
-We discuss our experience with XML, JSON, and JavaScript in Section XY (Evaluation).
-
-Figure~\ref{fig:textgen_final} shows an example of a TextGen aspect for the full SimpleXML element.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=95mm]{./images/textgen_final.png}
-	\caption{Generated TextGen aspect for the SimpleXML element}
-	\label{fig:textgen_final}
-\end{figure}
-
-The automatically generated TextGen code can be adjusted manually very easily, so that it generates nicely indented XML code.
-In the case of the example above (in Figure F), we only need to wrap the \mpsconcept{Content{\_}2} child with indentation and change the sequence separator to a new line character.
-The resulting adjusted aspect (code) is shown below in Figure~\ref{fig:textgen_adjusted}.
-
-\begin{figure}[t!]
-	\centering
-	\includegraphics[width=85mm]{./images/textgen_adjusted.png}
-	\caption{Adjusted indentation inside the TextGen aspect}
-	\label{fig:textgen_adjusted}
-\end{figure}
-
-\subsection{Remarks about Grammars}
-
-As we have already shown in previous sections, it is very easy to write a grammar in a way that will cause problems during import into MPS (or once it is imported).
-Problems might occur during the creation of any aspect.
-Here we will look at some general problems/issues that grammar import poses (might pose).
-We will talk about these problems with respect to the challenges/obstacles we tried to overcome and which we described in previous sections (dedicated to structure, editor, and texgen, respectively).
-We also show a few examples of additional possible complications.
-
-\paragraph{Adjusting grammars.}
-There are some cases, where altering the input grammar might yield far better MPS language.
-We will show two examples how the usability of the resulting MPS language can be improved.
-Let us look at the definition of an XML attribute in Figure~\ref{fig:xmlattribute}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
-	\antlrlexerrule{STRING}      :   \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
-	            |   \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'}
-	            ;
-\end{antlr}
-\caption{Definition of an XML attribute}
-\label{fig:xmlattribute}
-\end{figure*}
-
-The original XML grammar has quotes as a part of the value.
-For the resulting MPS language, it would mean that there would be a placeholder for the attribute value that would expect us to input the leading and trailing quote together with the value too each time.
-It would also be marked red (by the syntax checker) unless we enter both quotes inside the value since the regular expression checking for quotes will not match.
-The user might be confused by this and will not be able to tell why his string value is incorrect.
-
-In our SimpleXML language, we adjusted the grammar easily in the manner that we show in Figure~\ref{fig:xmladjustgrammar}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
-	            |   \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'}
-	            ;
-	\antlrlexerrule{TEXT1}       :   \antlrregex{~["]*} ;
-	\antlrlexerrule{TEXT2}       :   \antlrregex{~[']*} ;
-\end{antlr}
-\caption{Adjusted grammar of SimpleXML}
-\label{fig:xmladjustgrammar}
-\end{figure*}
-
-We turned quotes into literals, thus ensuring that they will only appear in the projectional editor as fixed constant cells.
-We will not have to encapsulate the value in them each time.
-The user will only have to choose, which attribute version he wants to use (single or double quotes).
-
-As the second example, we will show the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
-Every statement in JavaScript needs to be either followed by a semicolon, newline, file end or end of the block --- see the Figure~\ref{fig:javascriptstmt}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{eos}            : \antlrlexerrule{SemiColon}
-	               | \antlrlexerrule{EOF}
-	               | {lineTerminatorAhead()}?
-	               | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}?
-	               ;
-	\textcolor{gray}{// Example reference of the eos rule}
-	\antlrparserrule{breakStatement} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos}
-	               ;
-\end{antlr}
-\caption{Statements in JavaScript}
-\label{fig:javascriptstmt}
-\end{figure*}
-
-Because there are multiple options, our import plugin would create a placeholder at the end of every statement.
-Every language element representing a statement would have one child of the \mpsinterface{IEos} interface type.
-This placeholder needs to be manually filled in for each statement.
-
-Since the projectional editor has much bigger power over the form of the code, we (a user) might want to have each statement on a separate line.
-Since we can differentiate between statements on the AST level, we do not need an explicit separator between them.
-This means that we might want to simplify the language and leave the semicolon out, or leave it just as a constant fixed part of the projectional editor, but not as something the user must explicitly fill in.
-Then we can just put each statement on a separate line as it is usual for JavaScript code, but we do not need the semicolon anymore.
-This small adjustment is very quick when done inside the grammar.
-We just change the \antlrparserrule{eos} rule to the form shown in Figure~\ref{fig:eosrule}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{eos} : \antlrliteral{;} ;
-\end{antlr}
-\caption{eos parser rule}
-\label{fig:eosrule}
-\end{figure*}
-
-This way, we can change the grammar in a way that it will not describe the same ECMAScript language as before, but will definitely make our MPS language more usable.
-TextGen aspect may be defined in such a way that puts the semicolon back (after each statement) in the generated plain-text representation. 
-This would be very important especially when JavaScript is available as another base language (which is actually a plan (future work) of the JetBrains company).
-
-Above, we have shown that adjusting the input grammar might be a very fast mean of tweaking/improving the usability of generated MPS language, and sometimes it may be the only proper solution in some complex situations.
-In general, the question we can ask ourselves is whether one aims for a full precise/faithful port of the input language or for an MPS alternative.
-
-\todo{Nahradit slovo "break" v nasledujicich odstavcich (v kontextu "break the parser") za vhodnejsi slovo ktere pusobi vic odborne.}
-
-\paragraph{Breaking grammars and parsers.}
-However, we have found that there is one big problem with grammar adjustment that we would like to point out.
-The problem is that it is very easy to change the input ANTLR grammar in a way that breaks the ANTLR parser generated out of it.
-By breaking we mean that it stops parsing the original language.
-We illustrate the problem on our SimpleXML language.
-
-As stated before, when creating the SimpleXML grammar, we have started off with the original XML grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and did some adjustments to it.
-We ended up with a grammar that can be processed by our method/plugin and creates a nice usable MPS language.
-Nevertheless, we noticed that even though the imported language behaves well enough and mimics the XML language quite nicely, the ANTLR parser generated out of this grammar no longer parses XML successfully.
-Some changes we have made, such as the attribute adjustment, broke the grammar down.
-More precisely, our changes improved the MPS language while breaking down the parser, and we have not even noticed it because, from the perspective of MPS, the imported language still corresponds to XML.
-
-What we are trying to say (show here) is that it is very easy to perform a grammar adjustment that will, at first, seem harmless and valid inside MPS, but will break the parser.
-The problem is that the user performing this change probably will not be aware of breaking it.
-
-The main (underlying) cause of this problem is the way the ANTLR parser is implemented and the quite different purpose we are using the grammar for in our project/tool/method.
-There are many ways how various parsers deal with, for example, token matching.
-For instance, ANTLR introduces so called \emph{greedy} and \emph{non-greedy} operators.
-The greedy way, in which ANTLR matches input on defined tokens and prioritizes their selection, makes some rules very dangerous.
-When a rule matches a wide range of input, it might happen during parsing that it is prioritized over other rules and swallows a lot more input than the author of it intended.
-Usually, these dangerous rules are bound to some parser context, which makes them behave well.
-
-However, for practical reasons, it is very important that grammar is not broken by our custom adjustments (and that the ANTLR parser of the language works properly), because we might need to automatically generate parsers of the adjusted language.
-Imagine this very real scenario (that will be supported in the future):
-\begin{enumerate}
-	\item We have imported the language inside MPS.
-	\item MPS now knows the structure of the language, and we can code in MPS using this language.
-	\item We would expect MPS to be able to load an existing source code, written in this language, from a text file and import it inside MPS.
-	\item We would like to use MPS to safely edit this code, using all the features of MPS.
-	\item We would also like to export the code in a plain-text form again and save it back to the source file.
-\end{enumerate}
-In order to make this happen, we must be able to generate a correctly-working ANTLR parser out of the adjusted source grammar that we used to (that helped us) import the language.
-We must also be able to match nodes of the AST coming out of the ANTLR parser to elements of the MPS language, and we must build the MPS AST out of the ANTLR AST.
-
-\section{Evaluation}
-
-We implemented everything (the proposed approach described above) in the form of an MPS plugin that allows users to carry out the import.
-
-We used a mix of Java and BaseLanguage in our implementation.
-We used the MPS API, which can be called from the BaseLanguage programs too, to generate some of the language elements.
-BaseLanguage is needed to use/call the MPS (language) API, i.e. to programmatically create new language elements and their aspects.
-
-The plugin uses the ANTLR library to work with ANTLRv4 grammar files (to parse them, etc) and several MPS libraries (that implement the MPS API).
-
-For the first step of automated construction of MPS languages from grammars (i.e., for parsing the grammar), the plugin uses the ANTLR library~\cite{ref:ANTLR4} to generate a Java ANTLR parser of the ANTLRv4 notation, because there exists an ANTLRv4 grammar for the ANTLRv4 notation~\cite{ref:ANTLR4reference}.
-
-\todo{
-	We should provide url to the github repository (implementation) and pointer to some package with ANTLR grammars and imported MPS languages (take from Attachments.zip in the master thesis).
-}
-
-\todo{Popsat nase zkusenosti s importem slozitych jazyku (XML, JavaScript, Python, a dalsi).}
-
-From introduction: We will also show some example languages imported with INGRID and discuss our experience.
-
-In order to evaluate INGRID, we tried to import several real/mainstream languages:
-\begin{itemize}
-	\item \textbf{SimpleXML} --- Simplified XML that we used in this thesis for explanatory purposes.
-	\item \textbf{JSON} --- Similar to XML, but this time a full port of the specification.
-	\item \textbf{ECMAScript 5.1} --- Specification of the language known as JavaScript, dated to the year 2011, which is currently the most frequently adopted version.
-\end{itemize}
-Results of the import process are included in the package that is available at \todo{url}.
-The repository \todo{Add url} contains an example MPS project. There are three languages.
-For each example language, you can find its original import, exactly in the form as created by our plugin.
-
-We experimented with import of XML, JavaScript, Python, and maybe some other languages.
-
-These general purpose languages are quite complex when it comes to their structure and the overall syntax variety.
-
-We must emphasize that the automatically imported languages are not ready-to-use full-fledged MPS languages.
-We did not expect to achieve that at this stage, as that is a much bigger, if not impossible, challenge.
-Imported languages contain full structure found in the respective grammars and some more aspect definitions that will help users of the languages in creating code.
-The imported language has to be manually adjusted (tweak, polished) by the end user (developer) and some of its complications have to be resolved manually by a human.
-
-We found out that manual adjusting of the editor/textgen aspect (adding line breaks and indentation) is far more effective and precise than any automated approach we came up with (during our research).
-The imported language can be, in our opinion, adjusted very fast into a more readable form.
-
-The package at \todo{url} contains also adjusted versions of all three languages.
-Then, for each language, there is an adjusted version.
-
-This adjusted version was created from the original import and the author spent no more than between 20 to 60 minutes of working on it.
-The first author spent only a couple of minutes of (manually) adjusting them after the automated import, so that we try to prove our point above.
-More specifically, the JSON language was ready for use in no more than 20 minutes of minor adjusting.
-We chose this language because it does not require us to implement complex aspects, such as type checking or data flows.
-Similarly, the SimpleXML turned out very nice.
-
-Our main goal was to prove that even though our method (plugin) does not create a perfect language, it can be customized very fast into a very usable form.
-We manually tweaked/customized the editor and the TextGen aspect only, structure aspect was left in its original form.
-This justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical (and the best).
-
-We have also tried to use the INGRID tool/plugin to import the JavaScript language, which is exactly the type of a complicated general purpose language that might be of interest to other people.
-There are already some projects, where a manual port of JavaScript is being done.
-One of them is the ECMAScript4MPS~\cite{ref:ECMAScript4MPS} from the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section X (similar projects, related work).
-The result of import by INGRID is quite good, comparable to the PE4MPS project.
-Our language is, of course, missing a lot of more advanced aspects/features that, for example, can help deriving types of expressions and so on.
-These aspects improve the usability by a lot, e.g.
-when the user starts writing numbers, a number literal is inserted, whereas in our import the user has to first insert the concept representing the number literal and then proceed with writing numbers.
-These actions are needed in any MPS language and are expected to be implemented after the import step, \todo{Maybe say that INGRID does not generate these actions/aspects/features yet but it is our future work} as it is not possible to generate them automatically.
-However, when it comes to language structure, concept aliases, and auto-completion, we think that INGRID (our method/plugin) did a very good job.
-The structure aspect is very similar to the ECMAScript4MPS's one, which was created manually over the course of a large number of hours.
-INGRID does that (achieves the same result) fully automatically.
-
-To summarize, import languages such as JavaScript, JSON, MySQL and others is giving nice results.
-
-Results of our experiments with languages such as JSON or JavaScript show that the proposed approach (INGRID) yields very good results, comparable to existing manual ports, and spares the user from hours of tedious, error prone and time-consuming work.
-
-Limitation: the project is lacking proper error reporting and cannot handle cyclic rules (which could be found in Python grammar, for example).
-
-\section{Conclusion and Future Work}
-
-\todo{Write something.}
-
-There is lot of further follow-up work to be done. Many open problems remain.
-
-Automated generation of the editor aspect, probably using some kind of machine learning.
-Generating nicely designed and user-friendly editors, maybe using artificial intelligence (machine learning) that would evaluate (analyze) existing code in plain text and based on that create the editors.
-
-Follow-up work might look more into the problem of code layout detection, possibly exploring suggested approaches such as the learning approach described in Section XY \todo{Tohle se snad podari aspon castecne implementovat}.
-This could lead to better results in editor and TextGen aspect generation which could definitely use some improvement.
-Another future endeavors will definitely touch the subject of generating parsers that would enable the user to import existing source code into MPS.
-
-\acks
-% TODO
-This research was supported by JetBrains. \todo{Pridat jeste nejaky grant a kompletne preformulovat.}
-
-\bibliographystyle{abbrvnat}
-
-% TODO opravit formatovani jednotlivych zaznamu a vhodne seradit
-\begin{thebibliography}{}
-
-\bibitem{ref:MPS} JetBrains MPS, \url{https://www.jetbrains.com/mps/}
-
-\bibitem{ref:mbeddr} mbeddr, \url{mbeddr.com/}
-
-\bibitem{ref:BaseLanguage} BaseLanguage, \url{confluence.jetbrains.com/display/MPSD34/Base+Language}
-
-\bibitem{ref:PE4MPS} PE4MPS project, \url{https://github.com/mar9000/pe4mps}
-
-\bibitem{ref:PE} PE project, \url{https://github.com/mar9000/pe}
-
-\bibitem{ref:ANTLR2MPS} ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR{\_}MPS}
-
-\bibitem{ref:ANTLR4} ANTLRv4, \url{https://github.com/antlr/antlr4}
-
-\bibitem{ref:ANTLR4reference} T. Parr. The Definitive ANTLR 4 Reference. The Pragmatic Bookshelf, 2013
-
-\bibitem{ref:ANTLR4grammars} ANTLRv4 grammar repository, \url{https://github.com/antlr/grammars-v4}
-
-\bibitem{ref:ECMAScript4MPS} ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}
-
-\bibitem{ref:javascript} JavaScript ANTLRv4 grammar, \url{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}
-
-\end{thebibliography}
-
-\end{document}

### 2016-09-23 chapters/language_definition.tex
Split paper into chapter files

--- 2016-09-21 paper.tex
@@
-\documentclass[10pt]{sigplanconf}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[noline]{algorithm2e}
-\usepackage{comment}
-\usepackage{color}
-\usepackage{times}
-\usepackage{multirow}
-\usepackage{balance}
-\usepackage{adjustbox}
-\usepackage{alltt}
-\usepackage{textcomp}
-\usepackage[usenames]{xcolor}
-
-\setcounter{tocdepth}{3}
-
-% configures style of source code listings
-\lstset{
-	basicstyle=\small\sffamily
-	%basicstyle=\scriptsize\ttfamily
-	%numbers=left
-}
-
-\lstdefinelanguage{java}{
-	morekeywords={abstract,continue,for,new,switch,assert,default,goto,package,synchronized,boolean,stream,event,filter,when,do,if,private,this,break,double,implements,protected,throw,byte,else,import,public,throws,case,enum,instanceof,return,transient,catch,extends,int,short,try,char,final,interface,static,void,class,finally,long,strictfp,volatile,const,float,native,super,while},
-	sensitive=false,
-	morestring=[b]",
-	morestring=[b]',
-	morecomment=[l]{//},
-	morecomment=[s]{/*}{*/)},
-	escapeinside=??,
-	moredelim=[is][\textit]{__}{__},
-	moredelim=[is][\textbf]{_*}{*_}
-}
-\lstnewenvironment{displayjava}
-	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b,xleftmargin=1em,xrightmargin=1em}}{}
-\lstnewenvironment{java}
-	{\lstset{language=java,basicstyle=\small\sffamily,tabsize=2,columns=fullflexible,captionpos=b}}{}
-
-\newcommand{\code}[1]{{\sf \small #1}\xspace}
-
-\newcommand{\mi}[1]{\mathit{#1}}
-\newcommand{\mr}[1]{\mathrm{#1}}
-\newcommand{\mt}[1]{\mathtt{#1}}
-
-\newcommand{\todo}[1]{{\bfseries #1}}
-
-% for writing comments
-\newcommand{\PP}[1]{\textcolor{blue}{\bfseries PP: #1}} % Pavel Parizek
-\newcommand{\PV}[1]{\textcolor{red}{\bfseries PV: #1}} % Premek Vysoky
-\newcommand{\JB}[1]{\textcolor{orange}{\bfseries JB: #1}} % JetBrains
-
-% colors for the ANTLR environment
-\definecolor{antlrbgcolor}{rgb}{.962, .962, .962}
-\definecolor{antlrparserrulecolor}{RGB}{11,83,148}
-\definecolor{antlrlexerrulecolor}{RGB}{191,144,0}
-\definecolor{antlrliteralcolor}{RGB}{133,32,12}
-\definecolor{antlrregexcolor}{RGB}{56,118,29}
-
-% colors for the MPS environment
-\definecolor{mpsbgcolor}{rgb}{.962, .962, .962}
-\definecolor{mpsconceptcolor}{RGB}{42, 140, 0}
-\definecolor{mpsinterfacecolor}{RGB}{170,0,229}
-
-% coloring commands
-\newcommand{\antlrap}{\textquotesingle}
-\newcommand{\antlrparserrule}[1]{\textcolor{antlrparserrulecolor}{#1}}
-\newcommand{\antlrlexerrule}[1]{\textcolor{antlrlexerrulecolor}{#1}}
-\newcommand{\antlrliteralnoap}[1]{\textcolor{antlrliteralcolor}{#1}}
-\newcommand{\antlrliteral}[1]{\antlrap\antlrliteralnoap{#1}\antlrap}
-\newcommand{\antlrregex}[1]{\textcolor{antlrregexcolor}{#1}}
-% TODO mozna zmenit "mpsconcept" na "mpslangelem"
-\newcommand{\mpsconcept}[1]{\textcolor{mpsconceptcolor}{#1}}
-\newcommand{\mpsinterface}[1]{\textcolor{mpsinterfacecolor}{#1}}
-
-% TODO zmenit barvu pozadi na bilou
-% environment for the ANTLR code (grammars)
-\newenvironment{antlr}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=antlrbgcolor,frame,minipage=394pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
-
-% environment for the MPS code
-\newenvironment{mps}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=mpsbgcolor,frame,minipage=394pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
-
-\begin{document}
-
-% TODO
-%\conferenceinfo{OOPSLA'12,} {October 19--26, 2012, Tuscon, Arizona, USA.}
-%\CopyrightYear{2012}
-%\copyrightdata{978-1-4503-1561-6/12/10}
-
-% TODO
-\title{\todo{INGRID: Creating Languages in MPS by Importing ANTLR Grammars}}
-% alternative: "INGRID: Importing Grammars Into MPS"
-
-\authorinfo{P\v{r}emysl Vysok\'y \and Pavel Par\'izek}
-	{Charles University, Faculty of Mathematics and Physics, Prague, Czech Republic}
-	{premek.vysoky@gmail.com, parizek@d3s.mff.cuni.cz}
-
-% TODO pridat nekoho dalsiho z JetBrains
-% TODO upravit affiliation pro JetBrains -> doplnit "s.r.o" nebo radsi "Inc."
-% TODO muzeme taky napsat jejich emaily pokud budou lidi z JetBrains souhlasit
-\authorinfo{V\'aclav Pech}
-	{JetBrains}
-
-\maketitle
-
-\begin{abstract}
-% TODO
-\todo{write later}
-
-\todo{Copy from master thesis.}
-JetBrains MPS is a language workbench focusing on domain-specific languages.
-Unlike many other language workbenches and IDEs, it uses a projectional editor for code.
-The developer directly manipulates the program in its tree form (AST) and not by editing a text source code.
-This brings many advantages, but on the other hand requires time-consuming and complicated MPS language definition.
-The thesis elaborates on the possibility of automating the process of creating MPS language definition from its grammar description.
-It introduces the MPS editor, evaluates approaches of related projects and describes author's efforts to implement an MPS plugin that allows this import.
-The chosen approach and the selection of tools used for implementation are justified in the thesis.
-We point out important problems that any similar project might deal with and we introduce some possible solutions.
-Furthermore, the thesis contains examples of imported languages, showing the potency of the chosen approach.
-The thesis also aims to lay groundwork for future extensions and suggest possible improvements.
-\end{abstract}
-
-% TODO possible keywords: JetBrains MPS, programming language, grammar, projectional editor
-
-\section{Introduction}
-
-\JB{Odstavec mirne upraven} The prevailing approach to define valid syntaxes for programming languages is through grammars written typically in notations based on (or similar to) EBNF [reference].
-Many popular tools exist for automated generation of parsers from the grammar definitions, such as ANTLR and Bison/Flex.
-The source code of programs in such languages is written in a text editors, usually embedded into an IDE.
-
-\JB{Odstavec mirne upraven} Projectional (or structural) editing offers an alternative approach.
-While the idea emerged as early as in 1970s, it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
-A recent revival of projectional editing has been observed in the area of language workbenches - IDE-like tools that enable the developers to manipulate the actual language definition.
-The Meta-programming System (MPS) [reference] by JetBrains is an open-source language workbench that focuses of Domain Specific Languages and leverages the concept of projectional editor.
-MPS allows developers to design custom languages and then use use them to write code and compile it into executables.
-MPS provides the whole IDE infrastructure for such languages.
-
-\JB{Odstavec mirne upraven} The key idea behind projectional editors is that a developer directly manipulates the AST representation of the source code instead of plain text.
-To avoid the fate of the early projectional editors MPS has put extra effort into making text manipulation feel like editing real text.
-MPS models, although being persisted as XML or binary files, can be versioned using the industry standard source-control systems (Git, Subversion).
-On-line model browsers have been incorporated into some of the popular public source code repositories (BitBucket) [reference] .
-
-\JB{Odstavec mirne upraven} The absence of parsers brings about numerous benefits, which are in particular useful for DSLs, which are frequently used by domain experts, who may not be professional programmers.
-\begin{itemize}
-	\item Since a projectional editors controls the interaction between the user and the code, the developers practically cannot enter syntactically invalid code.
-	\item Languages can be defined in a modular way and multiple languages can be easily combined together or extend one another, while parsers seriously limit language modularity for traditional parsed languages.
-	\item The languages may offer contextual or non-parseable notations, such as tables, diagrams or positional syntax.
-	\item Since the projection is detached from the physical representation of code (AST), authors can define multiple notations and allow the developers to switch between different representations of the code.
-\end{itemize}
-A motivating illustration of the abilities of MPS is the mbeddr project [citation], which extends C with numerous domain-specific constructs and data types.
-We provide more details about MPS in Section 2.1.
-
-\JB{Navrhuji smazat nasledujici odstavec}
-\todo{From thesis.
-	One of these tools is the JetBrains MPS~\cite{ref:MPS}, which approaches languages from a~different point of view.
-	MPS doesn't use the textual representation of code as usual, but rather works with the actual abstract syntax tree that holds the code's structure.
-	This has some positive implications as it gives us a~very powerful tool rich on features but also introduces some new problems we haven't encountered before.
-	One of them being the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-	MPS also becomes a~very powerful tool once the language has already been created and you start using it inside MPS.
-	There is a~huge variety of new languages constructed for MPS that usually solve domain specific problems.
-}
-
-\todo{Pouzivat jen terminy "AST nodes" a "language elements" (to spis). Nepouzivat pojem "concepts" abychom ctenare nematli.}
-
-\JB{Odstavec upraven} Projectional editing introduces some new problems we haven't encountered before.
-One of them is the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-In order to use a language inside MPS, it has to be defined through the MPS infrastructure.
-In particular, one has to create the abstract syntax called Structure (the types of allowed AST nodes), concrete syntax called Editor (projection of the AST on the screen and its interactions with the user), and text generation (to enable creation of plain text files used as input for compilers).
-We provide more details in Section X (Background about MPS).
-
-\JB{Odstavec pridan pro vysvetleni motivace prace} MPS is typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
-These DSLs are then transformed on the AST level by a series of model-to-model transformations to an AST model representing the desired semantics in a General-purpose language (GPL), such as C or Java.
-This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
-This implies that the target GPL must also be defined by the MPS means, to allow the DSLs to have their models transformed into a model in the GPL.
-
-\JB{Odstavec doplnen} However, it requires substantial effort to create a definition of a GPL in MPS.
-Therefore, very few mainstream GPLs are now fully supported by MPS.
-Only Java and C have been implemented to date, which limits the DSLs authors to only target these supported target GPLs.
-We believe that an automated process of importing grammars of GPLs into MPS would encourage more GPLs to be migrated to MPS and thus giving the DSL authors more options.
-
-\todo{From thesis.
-	However, building these languages is a~complicated and time-consuming process and a~lot of effort must be put into them before they are ready for use.
-
-	These examples show that recreating a~full language inside MPS is not an easy task and a~lot of time must be spent on implementing all aspects of the language.
-	A big part of this effort is, however, quite straightforward and can be possibly automated, as we show in this paper. Automation speeds up the process of adoption of new languages to a great degree.
-}
-
-\JB{Odstavec upraven} Since a big part of the language definition process in MPS is quite straightforward and can possibly be automated, we will look into the possibility of automatic import of already existing general purpose languages into MPS using a grammar description of their syntax.
-
-We present INGRID --- a method and a tool for semi-automated construction of MPS definition of programming languages based on their ANTLR grammars.
-We also discuss challenges that we encountered during this process and our solution to them.
-Most of these challenges are general, i.e.
-not MPS specific, as they apply to the relation between grammars defined in EBNF-like style (rules and tokens, everything defined in plain text) and structured (object-oriented) approach used in MPS (and other language workbenches).
-
-\todo{Hodne zduraznit automation, tedy ze chceme maximalizovat automatizaci.}
-The main goal of our/the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible.
-
-\todo{From thesis.
-	Automatic import of already existing languages into MPS from the grammar description of their syntax.
-}
-\todo{Mozna prodat INGRID jako automatickou konverzi gramatik do podoby MPS languages, tedy jako nastroj ktery umoznuje portaci jazyku do prostredi MPS a jejich nasledne pouziti v MPS (treba kombinace s dalsimi jazyky).}
-
-Our primary motivation is to enable better support for mainstream and/or general purpose programming languages (XML, C, C++, JavaScript, Python) in MPS, especially to enable writing programs in these languages using the projectional editor and to combine them with domain-specific languages natively created in MPS.
-\todo{Zduraznit take pouziti importovanych jazyku pro cile generovani z DSL --- to plati zejmena u mainstream general-purpose languages jako treba JavaScript.}
-
-\todo{Briefly mention the biggest challenges.}
-The are related to the way grammar encoding/description of languages are usually written.
-They do not hold any information about the code layout, and contain many rules that do not directly correspond to AST nodes (programming language constructs).
-We discuss them in more detail later (Section X).
-
-\todo{From thesis.
-	INGRID allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
-	The biggest benefit is the possibility to code in these mainstream general purpose languages using the projectional editor.
-}
-
-However, we do not target just the very complex general purpose languages such as JavaScript.
-INGRID can be used also for simple languages.
-We believe that, for simple languages, it is less time consuming to create the grammar by hand (e.g., using the ANLTR notation and toolset) and then create the language definition in MPS by using the INGRID tool, than to create it manually using the MPS GUI --- this is another benefit of INGRID.
-
-Structure of the paper.
-In the beginning, we will describe the MPS editor and related/similar projects.
-Then, we describe our approach to creating language definition in MPS from their grammars, and discuss the major challenges together with our solution.
-We will also show some example languages imported with INGRID and discuss our experience.
-
-\section{Background and Related Work}
-
-We will briefly introduce fundamental basics of the MPS environment.
-We will analyze existing related projects that are trying to accomplish similar goals, and then we will weigh advantages and disadvantages of these approaches.
-
-\subsection{JetBrains MPS}
-
-\JB{Odstavec upraven} JetBrains MPS is a complete language workbench - a development environment that allows developers to define their own languages and then use them to develop code.
-The code can then be transformed into a target language, typically a GPL, such as Java or C.
-
-\JB{Navrhuji odstranit} We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
-Focus on basic/fundamental and relevant features. In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.
-
-\JB{odstavec upraven} MPS differs from typical IDEs in one important aspect - the projectional editor.
-The developer does not work with the textual representation of the source code directly, but rather with its AST (abstract syntax tree) that is the model of the code.
-When users are creating programs using the MPS projectional editor, they are basically assembling the tree (AST) out of predefined building blocks of the imported languages.
-The definition of the language dictates, which where in the AST can certain elements be placed or how they can be nested inside other elements.
-In traditional IDEs it is the parser that constructs an AST using the language's elements.
-
-\JB{Odstavec upraven} The building blocks of MPS models are called nodes.
-Code of any program in MPS is built from nodes and these nodes represent instances of elements defined in one of the languages that the program declares to be using.
-These language elements are called concepts.
-In brief, nodes in code are instances of concepts defined in the language.
-
-\JB{Vyhody/nevyhody diskutujeme v kapitole 1. Navrhuji nasledujici odstavec zrusit nebo text z kapitoly 1. presunout sem}
-Keeping the model of the code in the AST form has several advantages.
-One of them is that MPS then only allows composing the program strictly following the syntax of the language, which results in an inability to actually write syntactically invalid code.
-The code is generated out of the always-valid AST beyond user's reach.
-
-\JB{Odstavec upraven} One of key advantages of projectional editor stems from separation of abstract and concrete syntax.
-While the AST provides an exhaustive representation of the code, the way it is displayed on the screen and the way the user interacts with it are unconstrained.
-The editor can take any visual form and shape.
-The language author can define multiple visualizations and let the developer to choose one that fits best the task at hands - editing, debugging, reviewing, resolving merge conflicts, etc.
-
-For a better understanding, we will give a~small example. Let's assume we have a language with typical if/else structure similar to the one in Java:
-
-\begin{center}
-	\begin{minipage}{.38\textwidth}
-		\begin{alltt}
-			if (\textit{condition})
-			    \textit{then statement block}
-			else
-			    \textit{else statement block}
-		\end{alltt}
-	\end{minipage}
-\end{center}
-
-Its abstract syntax tree representation will look something like shown on Figure~\ref{fig:if_ast}.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[scale=0.6]{./images/if_statement_ast.png}
-  \caption{"If statement" abstract syntax tree}
-  \label{fig:if_ast}
-\end{figure}
-
-The underlying child nodes have their own children and so on.
-The type of nodes can be restricted by the language structure definition.
-For example, we can restrict the condition in the if/else statement to be an expression, statement block to be a list of statements, etc.
-
-\JB{odstavec upraven} Alternative visual representations are possible and they are not bound to be just the textual representation at all.
-For example, blocks for both of the branches can be aligned next to each other and displayed with differend background color, like shown in Figure~\ref{fig:if_editor}.
-
-\begin{figure}[ht]
-	\centering
-	\hspace{-4mm}
-	\includegraphics[scale=0.75]{./images/if_statement_editor.png}
-	\caption{"If statement" projectional editor example}
-	\label{fig:if_editor}
-\end{figure}
-
-\JB{Odstavec mirne upraven}The definition of a language element (the concept) consists of several aspects and each aspect codifies a different part of the nodes' behavior.
-The essential aspects are the following: structure (abstract syntax), editor (concrete syntax - how the code is projected and edited) and textgen (how to transform the AST into textual source code).
-If, instead of generating text directly, the language is supposed to be generated into another language that is available in MPS, the generator aspect must be used to specify the model-transformation rules.
-The structure, editor and textgen aspects are most relevant for the contribution of this paper, so we describe them here (and neglect/ignore the other aspects such as type system and data-flow).
-
-\todo{Highlight the object-oriented hierarchical nature of language definitions in MPS. The environment restricts and guides the developer so that he makes less mistakes.}
-
-Some concepts and design patterns known from object oriented programming can be used in the language definition --- inheritance, interfaces, and so on.
-The AST node can be inherited from a parent (abstract) node and it can implement a special interface node.
-Most importantly, nodes have child nodes and that is how the AST itself is created.
-
-\todo{Zduraznit ze AST node (object) ma nekolik aspectu: structure, editor, textgen. Definice MPS language je organizovana takto --- pres AST nodes, ktere maji nekolik aspectu (ruzne pohledy).}
-
-\paragraph{Structure.}
-The most important aspect of any MPS language is structure. It must be created first for each AST node.
-It specifies core attributes such as name of the AST node, inheritance relationships, possible child nodes (types and cardinality), implemented interfaces, references to other nodes, other properties (fields of any type that can hold values), etc.
-In Figure~\ref{fig:if_statement_structure}, you can see the structure aspect definition for the if statement.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.55]{./images/if_statement_structure.png}
-	\caption{"If statement" structure aspect definition}
-	\label{fig:if_statement_structure}
-\end{figure}
-
-\paragraph{Editor.}
-Editor aspect is where the user defines what the projectional editor representation of a code fragment (an AST node) looks like.
-The editor aspect of an AST node defines its visual representation (of the code) in the projectional editor.
-The user usually incorporates all children, references, and properties inside the representation so that future user of the language can insert their value into these placeholders.
-All elements can be styled using a~language similar to CSS.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/if_statement_editor_definition.png}
-	\caption{"If statement" editor aspect definition}
-	\label{fig:if_editor_definition}
-\end{figure}
-
-\todo{From thesis.
-
-The editor is a module dedicated/responsible for the definition of appearance of an AST node (concept).
-JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells.
-These cells then can be styled to user's liking.
-There are many different types of cells, each behaving a little bit different towards its contents.
-For example, MPS supports various horizontal and vertical lists, etc.
-Extra cells can be added on top --- cells which just specify layout adjustments such as indentation and text color.
-In Figure~\ref{fig:if_editor_definition}, you can see what a real editor definition might look like for the if statement of the Java language (as defined in MPS).
-}
-
-\paragraph{BaseLanguage}
-\todo{Introduce the base language here.}
-A program written in an MPS language cannot be executed directly.
-MPS uses an approach based on translation into a program in the so-called BaseLanguage.
-A plain-text representation of a BaseLanguage program is then passed to a system compiler and the result is executed.
-The module responsible for generating the program in the BaseLanguage is called "textgen" (we describe it briefly in the next paragraph).
-
-BaseLanguage is also an MPS language --- it is the official port of Java for MPS (created by JetBrains, authors of MPS).
-The default base language of MPS is Java, but others are supported too.
-
-Since BaseLanguage is an MPS language, BaseLanguage code is not a (plain) text source code, but again, an AST built out of nodes belonging to the BaseLanguage language.
-This means that generating code inside MPS consists of generating some AST out of BaseLanguage nodes.
-
-Besides representing the default (most common) target of text generation, BaseLanguage is used also for all additional programming inside MPS.
-
-\paragraph{TextGen.}
-The TextGen aspect defines how a certain AST node will be translated into a text source code (in the base language).
-This will allow us to generate source code in text form out of an AST model and ultimately to generate a text program out of the code user writes in the (imported) MPS language.
-The TextGen aspect is basically just one method/procedure definition for each AST node in the language. The method produces string representation of the given AST node and its children.
-MPS then calls this method for the root AST node of the program.
-TextGen aspect has to be defined (especially its more complex functionality) using the BaseLanguage.
-Again, we included an example of the if statement (Figure~\ref{fig:if_statement_textgen}).
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.6]{./images/if_statement_textgen.png}
-	\caption{"If statement" TextGen aspect definition}
-	\label{fig:if_statement_textgen}
-\end{figure}
-
-\todo{From thesis.
-
-After the program's AST is created using the projectional editor, there can be rules defined on how to generate code in plain text out of that tree.
-This module is called "text gen".
-These defined generators can target any language and platform.
-
-\paragraph{Other aspects.}
-MPS supports also other aspects, such as type checking and data-flow analysis, but they are not important for our contribution.
-Moreover, information needed for their automated generation is not available/contained inside a grammar file.
-
-}
-
-\subsection{Similar Projects}
-
-Currently, there exists an almost full port of the Java language called BaseLanguage~\cite{ref:BaseLanguage} extended with some MPS specific features.
-It was imported manually by JetBrains and it is still undergoing changes as Java itself is evolving.
-
-Within the mbeddr project~\cite{ref:mbeddr}, the C language is also manually tailored for MPS.
-
-Now/here we describe three similar projects (with similar goals) that we are aware of.
-We focus on (consider) these three projects: PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
-
-\paragraph{PE4MPS.}
-PE4MPS~\cite{ref:PE4MPS} is a project by \todo{Doplnit autora} that is trying to solve the same problem as we do.
-It solves the lack of information about code layout in grammars by creating a new grammar notation called PE Grammars~\cite{ref:PE} (the PE abbreviation comes from projectional editing).
-It has tw components: PE parser and PE4MPS plugin for MPS.
-Their approach is to mimic an existing grammar notation called ANTLRv4~\cite{ANTLR4} and enrich its syntax with custom (their own) constructs.
-These extensions tell (guide, hint) the parser what the AST node layout should look like.
-Supported extensions (already implemented) include horizontal and vertical lists, and some indentation rules --- however, even these few features make the already not-so-simple syntax of ANTLR v4 much more complex.
-The parser then uses this information when generating the projectional editor for am AST node.
-Author of this approach/project described the PE syntax using the ANTLRv4 notation~\cite{ANTLR4reference} and then used the ANTLR toolset to automatically generate an ANTLRv4 parser for PE grammar files.
-The PE parser reads any PE file and stores the language structure found inside to a custom representation of Java objects.
-
-The PE4MPS plugin/project/tool is built on top of the PE parser.
-This plugin uses the PE parser to build the PE file representation (the aforementioned tree-like structure of Java objects) and then creates AST nodes and their aspects inside MPS.
-The extended syntax brought by PE describes the layout of each element, e.g.
-it tells the plugin that one set of child nodes should be displayed horizontally, another set should be vertical with each child on a separate line and with some indentation, and so on.
-
-A limitation of the PE4MPS approach is that it, from what we understand, does not implement full ANTLR syntax.
-This means that every grammar might need a non-trivial adjustment before its usage.
-One of our goals is to enable import of as many languages out-of-the-box as possible and adopting the full specification.
-
-\paragraph{ANTLR{\_}MPS.}
-ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
-The author of this project is Fabien Campagne
-The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
-However, it does not try to generate the projectional editor at all (i.e., it does not deal with this problem), probably because it is in an early stage of development.
-
-The way this import process of ANTLR{\_}MPS works is quite different from what we have seen in the PE4MPS project, and it is quite complex/complicated to use.
-It works as follows. We give only a brief high-level overview and omit low-level technical details.
-The author created a whole new ANTLRv4 MPS language, which is an MPS port of the grammar notations' syntax.
-To import a language, the user utilizes this MPS ANTLR language.
-The textual grammar is imported automatically into MPS taking the form of the MPS's ported grammar language (so that the textual grammar is converted into MPS nodes, that means an AST node is created for each grammar rule).
-No child-parent relationships in the structure are generated by the tool --- all must be created manually by the user.
-There are no editor nor TextGen aspects created neither.
-
-\paragraph{mps-metabnf.}
-\todo{Popis na zaklade emailove diskuze (Nizozemci).}
-
 \section{Creating Language Definition in MPS}
@@
 We must also be able to match nodes of the AST coming out of the ANTLR parser to elements of the MPS language, and we must build the MPS AST out of the ANTLR AST.
-
-\section{Evaluation}
-
-We implemented everything (the proposed approach described above) in the form of an MPS plugin that allows users to carry out the import.
-
-We used a mix of Java and BaseLanguage in our implementation.
-We used the MPS API, which can be called from the BaseLanguage programs too, to generate some of the language elements.
-BaseLanguage is needed to use/call the MPS (language) API, i.e. to programmatically create new language elements and their aspects.
-
-The plugin uses the ANTLR library to work with ANTLRv4 grammar files (to parse them, etc) and several MPS libraries (that implement the MPS API).
-
-For the first step of automated construction of MPS languages from grammars (i.e., for parsing the grammar), the plugin uses the ANTLR library~\cite{ref:ANTLR4} to generate a Java ANTLR parser of the ANTLRv4 notation, because there exists an ANTLRv4 grammar for the ANTLRv4 notation~\cite{ref:ANTLR4reference}.
-
-\todo{
-	We should provide url to the github repository (implementation) and pointer to some package with ANTLR grammars and imported MPS languages (take from Attachments.zip in the master thesis).
-}
-
-\todo{Popsat nase zkusenosti s importem slozitych jazyku (XML, JavaScript, Python, a dalsi).}
-
-From introduction: We will also show some example languages imported with INGRID and discuss our experience.
-
-In order to evaluate INGRID, we tried to import several real/mainstream languages:
-\begin{itemize}
-	\item \textbf{SimpleXML} --- Simplified XML that we used in this thesis for explanatory purposes.
-	\item \textbf{JSON} --- Similar to XML, but this time a full port of the specification.
-	\item \textbf{ECMAScript 5.1} --- Specification of the language known as JavaScript, dated to the year 2011, which is currently the most frequently adopted version.
-\end{itemize}
-Results of the import process are included in the package that is available at \todo{url}.
-The repository \todo{Add url} contains an example MPS project. There are three languages.
-For each example language, you can find its original import, exactly in the form as created by our plugin.
-
-We experimented with import of XML, JavaScript, Python, and maybe some other languages.
-
-These general purpose languages are quite complex when it comes to their structure and the overall syntax variety.
-
-We must emphasize that the automatically imported languages are not ready-to-use full-fledged MPS languages.
-We did not expect to achieve that at this stage, as that is a much bigger, if not impossible, challenge.
-Imported languages contain full structure found in the respective grammars and some more aspect definitions that will help users of the languages in creating code.
-The imported language has to be manually adjusted (tweak, polished) by the end user (developer) and some of its complications have to be resolved manually by a human.
-
-We found out that manual adjusting of the editor/textgen aspect (adding line breaks and indentation) is far more effective and precise than any automated approach we came up with (during our research).
-The imported language can be, in our opinion, adjusted very fast into a more readable form.
-
-The package at \todo{url} contains also adjusted versions of all three languages.
-Then, for each language, there is an adjusted version.
-
-This adjusted version was created from the original import and the author spent no more than between 20 to 60 minutes of working on it.
-The first author spent only a couple of minutes of (manually) adjusting them after the automated import, so that we try to prove our point above.
-More specifically, the JSON language was ready for use in no more than 20 minutes of minor adjusting.
-We chose this language because it does not require us to implement complex aspects, such as type checking or data flows.
-Similarly, the SimpleXML turned out very nice.
-
-Our main goal was to prove that even though our method (plugin) does not create a perfect language, it can be customized very fast into a very usable form.
-We manually tweaked/customized the editor and the TextGen aspect only, structure aspect was left in its original form.
-This justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical (and the best).
-
-We have also tried to use the INGRID tool/plugin to import the JavaScript language, which is exactly the type of a complicated general purpose language that might be of interest to other people.
-There are already some projects, where a manual port of JavaScript is being done.
-One of them is the ECMAScript4MPS~\cite{ref:ECMAScript4MPS} from the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section X (similar projects, related work).
-The result of import by INGRID is quite good, comparable to the PE4MPS project.
-Our language is, of course, missing a lot of more advanced aspects/features that, for example, can help deriving types of expressions and so on.
-These aspects improve the usability by a lot, e.g.
-when the user starts writing numbers, a number literal is inserted, whereas in our import the user has to first insert the concept representing the number literal and then proceed with writing numbers.
-These actions are needed in any MPS language and are expected to be implemented after the import step, \todo{Maybe say that INGRID does not generate these actions/aspects/features yet but it is our future work} as it is not possible to generate them automatically.
-However, when it comes to language structure, concept aliases, and auto-completion, we think that INGRID (our method/plugin) did a very good job.
-The structure aspect is very similar to the ECMAScript4MPS's one, which was created manually over the course of a large number of hours.
-INGRID does that (achieves the same result) fully automatically.
-
-To summarize, import languages such as JavaScript, JSON, MySQL and others is giving nice results.
-
-Results of our experiments with languages such as JSON or JavaScript show that the proposed approach (INGRID) yields very good results, comparable to existing manual ports, and spares the user from hours of tedious, error prone and time-consuming work.
-
-Limitation: the project is lacking proper error reporting and cannot handle cyclic rules (which could be found in Python grammar, for example).
-
-\section{Conclusion and Future Work}
-
-\todo{Write something.}
-
-There is lot of further follow-up work to be done. Many open problems remain.
-
-Automated generation of the editor aspect, probably using some kind of machine learning.
-Generating nicely designed and user-friendly editors, maybe using artificial intelligence (machine learning) that would evaluate (analyze) existing code in plain text and based on that create the editors.
-
-Follow-up work might look more into the problem of code layout detection, possibly exploring suggested approaches such as the learning approach described in Section XY \todo{Tohle se snad podari aspon castecne implementovat}.
-This could lead to better results in editor and TextGen aspect generation which could definitely use some improvement.
-Another future endeavors will definitely touch the subject of generating parsers that would enable the user to import existing source code into MPS.
-
-\acks
-% TODO
-This research was supported by JetBrains. \todo{Pridat jeste nejaky grant a kompletne preformulovat.}
-
-\bibliographystyle{abbrvnat}
-
-% TODO opravit formatovani jednotlivych zaznamu a vhodne seradit
-\begin{thebibliography}{}
-
-\bibitem{ref:MPS} JetBrains MPS, \url{https://www.jetbrains.com/mps/}
-
-\bibitem{ref:mbeddr} mbeddr, \url{mbeddr.com/}
-
-\bibitem{ref:BaseLanguage} BaseLanguage, \url{confluence.jetbrains.com/display/MPSD34/Base+Language}
-
-\bibitem{ref:PE4MPS} PE4MPS project, \url{https://github.com/mar9000/pe4mps}
-
-\bibitem{ref:PE} PE project, \url{https://github.com/mar9000/pe}
-
-\bibitem{ref:ANTLR2MPS} ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR{\_}MPS}
-
-\bibitem{ref:ANTLR4} ANTLRv4, \url{https://github.com/antlr/antlr4}
-
-\bibitem{ref:ANTLR4reference} T. Parr. The Definitive ANTLR 4 Reference. The Pragmatic Bookshelf, 2013
-
-\bibitem{ref:ANTLR4grammars} ANTLRv4 grammar repository, \url{https://github.com/antlr/grammars-v4}
-
-\bibitem{ref:ECMAScript4MPS} ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}
-
-\bibitem{ref:javascript} JavaScript ANTLRv4 grammar, \url{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}
-
-\end{thebibliography}
-
-\end{document}

### 2016-09-23 paper.tex
Split paper into chapter files

--- 2016-09-21 paper.tex
@@
 
-% TODO pridat nekoho dalsiho z JetBrains
+% TODO pridat nekoho dalsiho z JetBrains (Alex Shatalin?)
 % TODO upravit affiliation pro JetBrains -> doplnit "s.r.o" nebo radsi "Inc."
@@
 
-\section{Introduction}
-
-\JB{Odstavec mirne upraven} The prevailing approach to define valid syntaxes for programming languages is through grammars written typically in notations based on (or similar to) EBNF [reference].
-Many popular tools exist for automated generation of parsers from the grammar definitions, such as ANTLR and Bison/Flex.
-The source code of programs in such languages is written in a text editors, usually embedded into an IDE.
-
-\JB{Odstavec mirne upraven} Projectional (or structural) editing offers an alternative approach.
-While the idea emerged as early as in 1970s, it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
-A recent revival of projectional editing has been observed in the area of language workbenches - IDE-like tools that enable the developers to manipulate the actual language definition.
-The Meta-programming System (MPS) [reference] by JetBrains is an open-source language workbench that focuses of Domain Specific Languages and leverages the concept of projectional editor.
-MPS allows developers to design custom languages and then use use them to write code and compile it into executables.
-MPS provides the whole IDE infrastructure for such languages.
-
-\JB{Odstavec mirne upraven} The key idea behind projectional editors is that a developer directly manipulates the AST representation of the source code instead of plain text.
-To avoid the fate of the early projectional editors MPS has put extra effort into making text manipulation feel like editing real text.
-MPS models, although being persisted as XML or binary files, can be versioned using the industry standard source-control systems (Git, Subversion).
-On-line model browsers have been incorporated into some of the popular public source code repositories (BitBucket) [reference] .
-
-\JB{Odstavec mirne upraven} The absence of parsers brings about numerous benefits, which are in particular useful for DSLs, which are frequently used by domain experts, who may not be professional programmers.
-\begin{itemize}
-	\item Since a projectional editors controls the interaction between the user and the code, the developers practically cannot enter syntactically invalid code.
-	\item Languages can be defined in a modular way and multiple languages can be easily combined together or extend one another, while parsers seriously limit language modularity for traditional parsed languages.
-	\item The languages may offer contextual or non-parseable notations, such as tables, diagrams or positional syntax.
-	\item Since the projection is detached from the physical representation of code (AST), authors can define multiple notations and allow the developers to switch between different representations of the code.
-\end{itemize}
-A motivating illustration of the abilities of MPS is the mbeddr project [citation], which extends C with numerous domain-specific constructs and data types.
-We provide more details about MPS in Section 2.1.
-
-\JB{Navrhuji smazat nasledujici odstavec}
-\todo{From thesis.
-	One of these tools is the JetBrains MPS~\cite{ref:MPS}, which approaches languages from a~different point of view.
-	MPS doesn't use the textual representation of code as usual, but rather works with the actual abstract syntax tree that holds the code's structure.
-	This has some positive implications as it gives us a~very powerful tool rich on features but also introduces some new problems we haven't encountered before.
-	One of them being the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-	MPS also becomes a~very powerful tool once the language has already been created and you start using it inside MPS.
-	There is a~huge variety of new languages constructed for MPS that usually solve domain specific problems.
-}
-
-\todo{Pouzivat jen terminy "AST nodes" a "language elements" (to spis). Nepouzivat pojem "concepts" abychom ctenare nematli.}
-
-\JB{Odstavec upraven} Projectional editing introduces some new problems we haven't encountered before.
-One of them is the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-In order to use a language inside MPS, it has to be defined through the MPS infrastructure.
-In particular, one has to create the abstract syntax called Structure (the types of allowed AST nodes), concrete syntax called Editor (projection of the AST on the screen and its interactions with the user), and text generation (to enable creation of plain text files used as input for compilers).
-We provide more details in Section X (Background about MPS).
-
-\JB{Odstavec pridan pro vysvetleni motivace prace} MPS is typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
-These DSLs are then transformed on the AST level by a series of model-to-model transformations to an AST model representing the desired semantics in a General-purpose language (GPL), such as C or Java.
-This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
-This implies that the target GPL must also be defined by the MPS means, to allow the DSLs to have their models transformed into a model in the GPL.
-
-\JB{Odstavec doplnen} However, it requires substantial effort to create a definition of a GPL in MPS.
-Therefore, very few mainstream GPLs are now fully supported by MPS.
-Only Java and C have been implemented to date, which limits the DSLs authors to only target these supported target GPLs.
-We believe that an automated process of importing grammars of GPLs into MPS would encourage more GPLs to be migrated to MPS and thus giving the DSL authors more options.
-
-\todo{From thesis.
-	However, building these languages is a~complicated and time-consuming process and a~lot of effort must be put into them before they are ready for use.
-
-	These examples show that recreating a~full language inside MPS is not an easy task and a~lot of time must be spent on implementing all aspects of the language.
-	A big part of this effort is, however, quite straightforward and can be possibly automated, as we show in this paper. Automation speeds up the process of adoption of new languages to a great degree.
-}
-
-\JB{Odstavec upraven} Since a big part of the language definition process in MPS is quite straightforward and can possibly be automated, we will look into the possibility of automatic import of already existing general purpose languages into MPS using a grammar description of their syntax.
-
-We present INGRID --- a method and a tool for semi-automated construction of MPS definition of programming languages based on their ANTLR grammars.
-We also discuss challenges that we encountered during this process and our solution to them.
-Most of these challenges are general, i.e.
-not MPS specific, as they apply to the relation between grammars defined in EBNF-like style (rules and tokens, everything defined in plain text) and structured (object-oriented) approach used in MPS (and other language workbenches).
-
-\todo{Hodne zduraznit automation, tedy ze chceme maximalizovat automatizaci.}
-The main goal of our/the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible.
-
-\todo{From thesis.
-	Automatic import of already existing languages into MPS from the grammar description of their syntax.
-}
-\todo{Mozna prodat INGRID jako automatickou konverzi gramatik do podoby MPS languages, tedy jako nastroj ktery umoznuje portaci jazyku do prostredi MPS a jejich nasledne pouziti v MPS (treba kombinace s dalsimi jazyky).}
-
-Our primary motivation is to enable better support for mainstream and/or general purpose programming languages (XML, C, C++, JavaScript, Python) in MPS, especially to enable writing programs in these languages using the projectional editor and to combine them with domain-specific languages natively created in MPS.
-\todo{Zduraznit take pouziti importovanych jazyku pro cile generovani z DSL --- to plati zejmena u mainstream general-purpose languages jako treba JavaScript.}
-
-\todo{Briefly mention the biggest challenges.}
-The are related to the way grammar encoding/description of languages are usually written.
-They do not hold any information about the code layout, and contain many rules that do not directly correspond to AST nodes (programming language constructs).
-We discuss them in more detail later (Section X).
-
-\todo{From thesis.
-	INGRID allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
-	The biggest benefit is the possibility to code in these mainstream general purpose languages using the projectional editor.
-}
-
-However, we do not target just the very complex general purpose languages such as JavaScript.
-INGRID can be used also for simple languages.
-We believe that, for simple languages, it is less time consuming to create the grammar by hand (e.g., using the ANLTR notation and toolset) and then create the language definition in MPS by using the INGRID tool, than to create it manually using the MPS GUI --- this is another benefit of INGRID.
-
-Structure of the paper.
-In the beginning, we will describe the MPS editor and related/similar projects.
-Then, we describe our approach to creating language definition in MPS from their grammars, and discuss the major challenges together with our solution.
-We will also show some example languages imported with INGRID and discuss our experience.
-
-\section{Background and Related Work}
-
-We will briefly introduce fundamental basics of the MPS environment.
-We will analyze existing related projects that are trying to accomplish similar goals, and then we will weigh advantages and disadvantages of these approaches.
-
-\subsection{JetBrains MPS}
-
-\JB{Odstavec upraven} JetBrains MPS is a complete language workbench - a development environment that allows developers to define their own languages and then use them to develop code.
-The code can then be transformed into a target language, typically a GPL, such as Java or C.
-
-\JB{Navrhuji odstranit} We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
-Focus on basic/fundamental and relevant features. In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.
-
-\JB{odstavec upraven} MPS differs from typical IDEs in one important aspect - the projectional editor.
-The developer does not work with the textual representation of the source code directly, but rather with its AST (abstract syntax tree) that is the model of the code.
-When users are creating programs using the MPS projectional editor, they are basically assembling the tree (AST) out of predefined building blocks of the imported languages.
-The definition of the language dictates, which where in the AST can certain elements be placed or how they can be nested inside other elements.
-In traditional IDEs it is the parser that constructs an AST using the language's elements.
-
-\JB{Odstavec upraven} The building blocks of MPS models are called nodes.
-Code of any program in MPS is built from nodes and these nodes represent instances of elements defined in one of the languages that the program declares to be using.
-These language elements are called concepts.
-In brief, nodes in code are instances of concepts defined in the language.
-
-\JB{Vyhody/nevyhody diskutujeme v kapitole 1. Navrhuji nasledujici odstavec zrusit nebo text z kapitoly 1. presunout sem}
-Keeping the model of the code in the AST form has several advantages.
-One of them is that MPS then only allows composing the program strictly following the syntax of the language, which results in an inability to actually write syntactically invalid code.
-The code is generated out of the always-valid AST beyond user's reach.
-
-\JB{Odstavec upraven} One of key advantages of projectional editor stems from separation of abstract and concrete syntax.
-While the AST provides an exhaustive representation of the code, the way it is displayed on the screen and the way the user interacts with it are unconstrained.
-The editor can take any visual form and shape.
-The language author can define multiple visualizations and let the developer to choose one that fits best the task at hands - editing, debugging, reviewing, resolving merge conflicts, etc.
-
-For a better understanding, we will give a~small example. Let's assume we have a language with typical if/else structure similar to the one in Java:
-
-\begin{center}
-	\begin{minipage}{.38\textwidth}
-		\begin{alltt}
-			if (\textit{condition})
-			    \textit{then statement block}
-			else
-			    \textit{else statement block}
-		\end{alltt}
-	\end{minipage}
-\end{center}
-
-Its abstract syntax tree representation will look something like shown on Figure~\ref{fig:if_ast}.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[scale=0.6]{./images/if_statement_ast.png}
-  \caption{"If statement" abstract syntax tree}
-  \label{fig:if_ast}
-\end{figure}
-
-The underlying child nodes have their own children and so on.
-The type of nodes can be restricted by the language structure definition.
-For example, we can restrict the condition in the if/else statement to be an expression, statement block to be a list of statements, etc.
-
-\JB{odstavec upraven} Alternative visual representations are possible and they are not bound to be just the textual representation at all.
-For example, blocks for both of the branches can be aligned next to each other and displayed with differend background color, like shown in Figure~\ref{fig:if_editor}.
-
-\begin{figure}[ht]
-	\centering
-	\hspace{-4mm}
-	\includegraphics[scale=0.75]{./images/if_statement_editor.png}
-	\caption{"If statement" projectional editor example}
-	\label{fig:if_editor}
-\end{figure}
-
-\JB{Odstavec mirne upraven}The definition of a language element (the concept) consists of several aspects and each aspect codifies a different part of the nodes' behavior.
-The essential aspects are the following: structure (abstract syntax), editor (concrete syntax - how the code is projected and edited) and textgen (how to transform the AST into textual source code).
-If, instead of generating text directly, the language is supposed to be generated into another language that is available in MPS, the generator aspect must be used to specify the model-transformation rules.
-The structure, editor and textgen aspects are most relevant for the contribution of this paper, so we describe them here (and neglect/ignore the other aspects such as type system and data-flow).
-
-\todo{Highlight the object-oriented hierarchical nature of language definitions in MPS. The environment restricts and guides the developer so that he makes less mistakes.}
-
-Some concepts and design patterns known from object oriented programming can be used in the language definition --- inheritance, interfaces, and so on.
-The AST node can be inherited from a parent (abstract) node and it can implement a special interface node.
-Most importantly, nodes have child nodes and that is how the AST itself is created.
-
-\todo{Zduraznit ze AST node (object) ma nekolik aspectu: structure, editor, textgen. Definice MPS language je organizovana takto --- pres AST nodes, ktere maji nekolik aspectu (ruzne pohledy).}
-
-\paragraph{Structure.}
-The most important aspect of any MPS language is structure. It must be created first for each AST node.
-It specifies core attributes such as name of the AST node, inheritance relationships, possible child nodes (types and cardinality), implemented interfaces, references to other nodes, other properties (fields of any type that can hold values), etc.
-In Figure~\ref{fig:if_statement_structure}, you can see the structure aspect definition for the if statement.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.55]{./images/if_statement_structure.png}
-	\caption{"If statement" structure aspect definition}
-	\label{fig:if_statement_structure}
-\end{figure}
-
-\paragraph{Editor.}
-Editor aspect is where the user defines what the projectional editor representation of a code fragment (an AST node) looks like.
-The editor aspect of an AST node defines its visual representation (of the code) in the projectional editor.
-The user usually incorporates all children, references, and properties inside the representation so that future user of the language can insert their value into these placeholders.
-All elements can be styled using a~language similar to CSS.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/if_statement_editor_definition.png}
-	\caption{"If statement" editor aspect definition}
-	\label{fig:if_editor_definition}
-\end{figure}
-
-\todo{From thesis.
-
-The editor is a module dedicated/responsible for the definition of appearance of an AST node (concept).
-JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells.
-These cells then can be styled to user's liking.
-There are many different types of cells, each behaving a little bit different towards its contents.
-For example, MPS supports various horizontal and vertical lists, etc.
-Extra cells can be added on top --- cells which just specify layout adjustments such as indentation and text color.
-In Figure~\ref{fig:if_editor_definition}, you can see what a real editor definition might look like for the if statement of the Java language (as defined in MPS).
-}
-
-\paragraph{BaseLanguage}
-\todo{Introduce the base language here.}
-A program written in an MPS language cannot be executed directly.
-MPS uses an approach based on translation into a program in the so-called BaseLanguage.
-A plain-text representation of a BaseLanguage program is then passed to a system compiler and the result is executed.
-The module responsible for generating the program in the BaseLanguage is called "textgen" (we describe it briefly in the next paragraph).
-
-BaseLanguage is also an MPS language --- it is the official port of Java for MPS (created by JetBrains, authors of MPS).
-The default base language of MPS is Java, but others are supported too.
-
-Since BaseLanguage is an MPS language, BaseLanguage code is not a (plain) text source code, but again, an AST built out of nodes belonging to the BaseLanguage language.
-This means that generating code inside MPS consists of generating some AST out of BaseLanguage nodes.
-
-Besides representing the default (most common) target of text generation, BaseLanguage is used also for all additional programming inside MPS.
-
-\paragraph{TextGen.}
-The TextGen aspect defines how a certain AST node will be translated into a text source code (in the base language).
-This will allow us to generate source code in text form out of an AST model and ultimately to generate a text program out of the code user writes in the (imported) MPS language.
-The TextGen aspect is basically just one method/procedure definition for each AST node in the language. The method produces string representation of the given AST node and its children.
-MPS then calls this method for the root AST node of the program.
-TextGen aspect has to be defined (especially its more complex functionality) using the BaseLanguage.
-Again, we included an example of the if statement (Figure~\ref{fig:if_statement_textgen}).
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.6]{./images/if_statement_textgen.png}
-	\caption{"If statement" TextGen aspect definition}
-	\label{fig:if_statement_textgen}
-\end{figure}
-
-\todo{From thesis.
-
-After the program's AST is created using the projectional editor, there can be rules defined on how to generate code in plain text out of that tree.
-This module is called "text gen".
-These defined generators can target any language and platform.
-
-\paragraph{Other aspects.}
-MPS supports also other aspects, such as type checking and data-flow analysis, but they are not important for our contribution.
-Moreover, information needed for their automated generation is not available/contained inside a grammar file.
-
-}
-
-\subsection{Similar Projects}
-
-Currently, there exists an almost full port of the Java language called BaseLanguage~\cite{ref:BaseLanguage} extended with some MPS specific features.
-It was imported manually by JetBrains and it is still undergoing changes as Java itself is evolving.
-
-Within the mbeddr project~\cite{ref:mbeddr}, the C language is also manually tailored for MPS.
-
-Now/here we describe three similar projects (with similar goals) that we are aware of.
-We focus on (consider) these three projects: PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
-
-\paragraph{PE4MPS.}
-PE4MPS~\cite{ref:PE4MPS} is a project by \todo{Doplnit autora} that is trying to solve the same problem as we do.
-It solves the lack of information about code layout in grammars by creating a new grammar notation called PE Grammars~\cite{ref:PE} (the PE abbreviation comes from projectional editing).
-It has tw components: PE parser and PE4MPS plugin for MPS.
-Their approach is to mimic an existing grammar notation called ANTLRv4~\cite{ANTLR4} and enrich its syntax with custom (their own) constructs.
-These extensions tell (guide, hint) the parser what the AST node layout should look like.
-Supported extensions (already implemented) include horizontal and vertical lists, and some indentation rules --- however, even these few features make the already not-so-simple syntax of ANTLR v4 much more complex.
-The parser then uses this information when generating the projectional editor for am AST node.
-Author of this approach/project described the PE syntax using the ANTLRv4 notation~\cite{ANTLR4reference} and then used the ANTLR toolset to automatically generate an ANTLRv4 parser for PE grammar files.
-The PE parser reads any PE file and stores the language structure found inside to a custom representation of Java objects.
-
-The PE4MPS plugin/project/tool is built on top of the PE parser.
-This plugin uses the PE parser to build the PE file representation (the aforementioned tree-like structure of Java objects) and then creates AST nodes and their aspects inside MPS.
-The extended syntax brought by PE describes the layout of each element, e.g.
-it tells the plugin that one set of child nodes should be displayed horizontally, another set should be vertical with each child on a separate line and with some indentation, and so on.
-
-A limitation of the PE4MPS approach is that it, from what we understand, does not implement full ANTLR syntax.
-This means that every grammar might need a non-trivial adjustment before its usage.
-One of our goals is to enable import of as many languages out-of-the-box as possible and adopting the full specification.
-
-\paragraph{ANTLR{\_}MPS.}
-ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
-The author of this project is Fabien Campagne
-The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
-However, it does not try to generate the projectional editor at all (i.e., it does not deal with this problem), probably because it is in an early stage of development.
-
-The way this import process of ANTLR{\_}MPS works is quite different from what we have seen in the PE4MPS project, and it is quite complex/complicated to use.
-It works as follows. We give only a brief high-level overview and omit low-level technical details.
-The author created a whole new ANTLRv4 MPS language, which is an MPS port of the grammar notations' syntax.
-To import a language, the user utilizes this MPS ANTLR language.
-The textual grammar is imported automatically into MPS taking the form of the MPS's ported grammar language (so that the textual grammar is converted into MPS nodes, that means an AST node is created for each grammar rule).
-No child-parent relationships in the structure are generated by the tool --- all must be created manually by the user.
-There are no editor nor TextGen aspects created neither.
-
-\paragraph{mps-metabnf.}
-\todo{Popis na zaklade emailove diskuze (Nizozemci).}
-
-\section{Creating Language Definition in MPS}
-
-From all the available notations in which grammars can be written, we decided to use the ANTLRv4 grammar notation~\cite{ANTLR4reference} as the input format (in which the input programming languages have to be defined) for the following reasons.
-\begin{itemize}
-	\item It is a very widely used grammar notation.
-	\item All popular and widely used programming languages have their syntax captured (represented) using this notation --- see the repository~\cite{ref:ANTLR4grammars}.
-	\item There is a whole framework/toolset for generating parsers and editing grammars built around it~\cite{ref:ANTLR4}.
-	\item It supports Java (it is written in Java), which is the implementation language of MPS and also the default BaseLanguage.
-	\item It uses (is based on) an EBNF (Extended Backus-Naur Form) \footnote{https://en.wikipedia.org/wiki/Extended{\_}Backus\%E2\%80\%93Naur{\_}Form} form that it useful because it meets/respects the object oriented design that is used when describing MPS aspects (definitions of MPS languages).
-	\item It has its own syntax captured in its own syntax\footnote{https://github.com/antlr/grammars-v4/tree/master/antlr4} (there is an ANTLRv4 grammar for the ANTLRv4 notation).
-This is a very important feature as it allows us to create a parser for ANTLRv4 grammars so that we can parse grammars of other languages.
-\end{itemize}
-
-We use ANTLRv4 grammars as inputs (and work with them).
-	https://github.com/antlr/antlr4/blob/master/doc/index.md
-	https://github.com/antlr/grammars-v4
-
-Unlike some of the related projects, we did not add any new features into the grammar notation.
-We tried to import grammars as they are.
-
-The generation/import process performed by INGRID consists of three phases (import of structure, usable editor, text generation).
-We will talk about all phases of the generation/import process, describing encountered challenges/obstacles using an example language.
-
-To fulfill our goal of maximal possible automation, we import/generate the full language structure and then do a lot of mundane and time-consuming work connected to creating projectional editors.
-
-Now we describe/introduce the main steps of our approach (that INGRID tool performs): parsing the grammar, creating the structure, editor, and textgen.
-We are able to create/generate the structure aspect and then very basic editor and some default/basic textgen.
-In particular, we did not port the ANTLR notation as an MPS language.
-In the following subsections, we provide some details about our approach to parsing the input grammar and creating the main aspects of each element (each type of AST node) of an MPS language (structure, editor, and textgen).
-
-During our work on this project, we also discovered/encountered challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component), whitespace (editor aspect).
-There is a challenge/issue that we have discovered during our work on this project (in the process of our own implementation), and that is related to construction of the structure aspect.
-Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
-However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
-We describe/discuss these challenges in more detail (also using examples) and our solutions to the challenges inside the respective following subsections.
-
-Result of the construction of a MPS language from the input grammar (import) has to be polished by the human user.
-For example, the aspects not supported by INGRID will be used for adjusting the imported language and improving the usability of the language.
-
-\todo{Briefly discuss how our overall approach differs from similar projects (PE4MPS, etc).}
-There are more ways how to approach the problem of a grammar import.
-
-Our overall approach differs from the PE4MPS project (Section X.Y) especially in the level of automation.
-While the main goal of (our) the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible, usage of PE does not really spare the user any work because it only shifts the manual work from creating projectional editor to writing PE grammars.
-What the user would do inside the MPS UI using the projectional editor is in the case of PE4MPS done using a text editor when editing PE grammars.
-MPS UI is specially designed/made/tailored for this purpose, guides the user through this process, and supports aids such as auto-completion.
-In particular, MPS UI allows the user to perform only changes that result in valid output.
-Editing the grammar stored in a text file using a standard editor is an inferior approach with respect to using MPS (projectional editor), because a text editor is not as user-friendly, and, more importantly, it is an error prone environment.
-
-Main drawback of ANTLR{\_}MPS, and main difference from our approach, is very limited automation: it does not create the parent-child relationships between AST nodes in the structure, and does not generate any editor and textgen aspects (for any node) at all.
-In addition, there is no attention paid to the problem of generating projectional editor and text generation is neglected too.
-We blame all this on the project being in its early stage of development.
-
-In the rest of this section, first we define the language that we will use throughout the paper as the running example.
-Then we will talk about parsing grammar files \todo{Tohle je mozna zbytecne}.
-Lastly, we will describe our approach to tackling generation of language's aspects (structure, editor, and TextGen).
-At some places, we discuss important possible alternatives (as there are several possible approaches).
-
-\subsection{Running Example}
-
-We illustrate the main steps on the example of a simplified XML language, which is defined below (in Figure XY).
-
-\todo{Pouzit jazyk SimpleXML na running exaple.}
-
-We decided to use SimpleXML because it is small enough to fit here, but still can be used to illustrate the main challenges and our solution
-
-Throughout the whole paper, we will be using one example language to illustrate main challenges and our solution.
-The language is a simplified version of XML. We call it \emph{SimpleXML}.
-We created the grammar of the language by taking the official XML ANTLRv4 grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and stripped it down to eliminate/avoid some not so interesting (advanced) features such as XML entities.
-Below in Figure X, you can find the grammar of the SimpleXML language in the ANTLRv4 notation.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\textbf{grammar} \textit{SimpleXML};
-	\antlrparserrule{document}     :   \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
-	\antlrparserrule{prolog}       :   \antlrliteral{<?xml } \antlrparserrule{attribute}* \antlrliteral{?>} ;
-	\antlrparserrule{comment}      :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-	             |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-	             ;
-	\antlrparserrule{attribute}    :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
-	\antlrparserrule{content}      :   \antlrlexerrule{TEXT}
-	             |   \antlrparserrule{element}
-	             |   \antlrparserrule{comment}
-	             |   \antlrlexerrule{CDATA}
-	             ;
-	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-	\textbf{fragment}
-	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
-	             |   \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.}
-	             |   \antlrlexerrule{DIGIT}
-	             ;
-	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
-	\antlrlexerrule{TEXT}         :   \antlrregex{~[<"]*} ;
-	\antlrlexerrule{CDATA}        :   \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
-\end{antlr}
-\caption{SimpleXML grammar}
-\end{figure*}
-
-We have colored the grammar so it is easier to find the way around it (to make navigation easier).
-The grammar contains following basic elements:
-
-\begin{itemize}
-	\item \textbf{ANTLRv4 keywords} -- In black, they are required by the specification.
-	\item \antlrparserrule{\textbf{Parser rules}} -- Stated in blue, they are the rules describing language's structure.
-		The alternatives (that the element on the left side may break into) are separated by the pipe ($|$) character.
-	\item \antlrlexerrule{\textbf{Lexer rules}} -- In yellow, they are rules describing terminal symbols that the parser is matching against the input.
-		Terminals are encoded as string values and regular expressions.
-		The \textbf{fragment} keyword states that the lexer sub-rule is just a helper rule that brings more clarity into the notation but is not visible in the parser output.
-	\item \antlrliteralnoap{\textbf{String values (literals)}} -- In red, always inside of a pair of apostrophes, they are similar to lexer rules.
-		They describe a terminal string, not using regular expressions but by an exact string match.
-	\item \antlrregex{\textbf{Regular expressions}} -- Colored in green, they describe a string token to be matched using a regular expression with the special ANTLRv4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
-	\item \textbf{Rule operators (?+*)} - Left in black, following any element, they quantify the number of occurrences the prepending element can appear in.
-		These operators can also be prepended by an asterisk (*), meaning they are non-greedy.
-\end{itemize}
-
-\subsection{Parsing (the) Grammar}
-
-\todo{Define some name for the AST representing the grammar, and use it consistently throughout the whole paper. Maybe distinguish between "grammar AST" and "MPS structure AST".}
-The first step (stage) of creating the MPS language (of the import process) is parsing the grammar file (written using the ANTLRv4 notation) so that we can get an AST tree representing the grammar.
-
-The abstract syntax tree that comes out of the automatically generated (ANTLR) parser is a little bit complicated and full of information not relevant for our purposes (to us).
-That is why we translate the complex ANTLR AST into our own simplified tree made out of our own objects.
-We discarded as much irrelevant information as possible and only kept information (data) vital/necessary for construction of MPS languages.
-
-Furthermore, we process the grammar in two iterations.
-In the first iteration/pass, we build a tree holding names of other referenced rules in the form of strings.
-In the second iteration, we walk this tree and resolve all these references to real pointers to other rule objects.
-We end up with a clear and simpler representation that is easy to process in the next steps of the import process.
-
-We also had to post-process (simplify) the representation of lexer rules (tokens).
-As a result of the initial parsing, ANTLR parser produces some kind of a tree structure that is representing the lexer rule (meaning/note that lexer rules can also be built from alternatives just like parser rules).
-However, lexer rules are, in fact (basically), just regular expressions used for parsing input source code into tokens.
-We took each/every such tree that represents the lexer rule and flattened it into the equivalent regular expression.
-
-For illustration, consider the lexer rule \antlrlexerrule{Name} from our/the SimpleXML language (Figure XY), together with all its sub-rules, shown in Figure~\ref{fig:lexerrule}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-	\textbf{fragment}
-	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
-             |   \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.}
-             |   \antlrlexerrule{DIGIT}
-             ;
-	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
-\end{antlr}
-\caption{Lexer rule from SimpleXML}
-\label{fig:lexerrule}
-\end{figure*}
-
-Intuitively, we can create a regular expression by gluing elements of each alternative together and then join these alternatives with a classic regex OR ($|$), which is exactly its function in the ANTLRv4 grammar notation.
-We have captures this idea in the form of the recursive algorithm in Figure~\ref{fig:algflatten}, which covers the whole process.
-
-% TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	Flatten(\textit{R}):
-	1) Define \textit{T} as a list of empty string
-	2) For each alternative \textit{A} of the rule \textit{R}:
-	3)     Define \textit{R} = \antlrap\antlrap
-	4)     For each element \textit{E} of \textit{A}:
-	5)         If \textit{E} is not yet flattened:
-	6)             Flatten(\textit{E})
-	7)         \textit{R}.append(\textit{E})
-	8)         \textit{R}.append(\textit{E}.operator)
-	9)     \textit{T}.add(\textit{R})
-	10) Build a string S out of elements t$_{\text{1}}$, t$_{\text{2}}$, ... t$_{\text{n}}$ of \textit{T}:
-	11)     (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
-	12) Return \textit{S}
-\end{antlr}
-\caption{Flatten algorithm}
-\label{fig:algflatten}
-\end{figure*}
-
-The output of \texttt{Flatten(Name)} (output of the Flatten algorithm applied to the \antlrlexerrule{Name} rule) is this regular expression (in this form):
-
-\begin{center}
-	\texttt{([:a-zA-Z])(([:a-zA-Z])|(-)|({\_})|(.)|([0-9]))*}
-\end{center}
-
-\subsection{Structure}
-
-Now, as the next step, we need to translate/transform the AST (the types of AST nodes) derived from the input grammar into elements of the structure aspect of the MPS language definition (into types of AST nodes), and linking the nodes appropriately.
-
-The goal of this second step is to translate (transfer) the syntax (structure) of the input language that is captured by the grammar (ANTLR), into the MPS structure aspect.
-The structure of the MPS language is somewhat similar to the grammar.
-This is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
-The process of transferring grammar rules into the structure of a MPS language presents/involves some interesting challenges.
-The input grammar (EBNF, ANTLR) serves a little bit different purpose than MPS language, and the way grammar rules are usually written down causes some problems that we had to overcome in order to recreate the language inside MPS.
-Note that the challenges and problems are not specific to MPS --- they are common for everyone tackling a similar problem.
-
-We are going to describe two different approaches that we designed --- the straightforward approach (Section~\ref{sect:straightforward_approach}) and the shortcut approach (Section~\ref{sect:shortcut_approach}).
-We will describe two possible approaches that we developed in Section X and Section Y, respectively.
-\todo{Rict neco jako ze ten druhy, slozitejsi, approach vlastne rozsiruje ten prvni, jednodussi.}
-
-However, before focusing on the two approaches, we must describe the basic principles of translation from the AST representing the grammar to the MPS structure, which are common to both approaches.
-The structure is created according to object-oriented principles.
-Each node (element) is represented by an object (and can be treated as such) that may have a parent node, child nodes, and properties of any data type (similar to fields), it may implement any number of interfaces, and it may also hold references to other nodes.
-For every rule of the grammar (for every node in the grammar AST) that breaks into more/multiple alternatives, we must create a node in the MPS structure AST for each alternative of the grammar (every child node).
-Names of the nodes for alternatives are composed from the name of the rule and a number indicating the order of appearance in the grammar.
-For example, considering the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlgrammarelement}
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-             \textcolor{gray}{|   \antlrap<\antlrap Name attribute* \antlrap/>\antlrap}
-             \textcolor{gray}{;}
-\end{antlr}
-\caption{Fragment of the SimpleXML grammar}
-\label{fig:xmlgrammarelement}
-\end{figure*}
-
-Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.75]{./images/element_node_common.png}
-	\caption{Resulting Element{\_}1 node}
-	\label{fig:element_node_common}
-\end{figure}
-
-String literals (e.g., keywords such as \textbf{for}) are omitted/skipped because they will appear only in the corresponding projectional editor.
-The node named \mpsconcept{Element{\_}1} contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name} inside the first alternative of the \antlrparserrule{element} rule,
-Value of the property will be restricted using the regular expression representing the \antlrlexerrule{Name} rule.
-We explain later how the references to parser rules (\antlrparserrule{attribute} and \antlrparserrule{content}) as this is where the approaches differ.
-The main/key difference between these two lies in the way we create children fields for parser rules and in the way we are going to link them together using interface concepts.
-
-\subsubsection{The Straightforward Approach}
-\label{sect:straightforward_approach}
-
-The key idea of this approach is that, for each rule with more than one alternative, an interface node is created.
-Consider the \antlrparserrule{content} rule from our SimpleXML language (Figure~\ref{fig:contentrule}).
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
-\end{antlr}
-\caption{Content rule}
-\label{fig:contentrule}
-\end{figure*}
-
-Any of the four nodes corresponding to alternatives could appear anywhere the \antlrparserrule{content} rule is referenced.
-So we will create one interface node, and then for each alternative of this rule we create one node that implements the interface.
-The resulting fragment of structure will look like the one in Figure~\ref{fig:icontentitf}.
-
-\begin{figure*}[ht]
-\centering
-\begin{mps}
-	\mpsinterface{IContent}   :   \mpsconcept{Content{\_}1}
-           |   \mpsconcept{Content{\_}2}
-           |   \mpsconcept{Content{\_}3}
-           |   \mpsconcept{Content{\_}4}
-\end{mps}
-\caption{IContent interface}
-\label{fig:icontentitf}
-\end{figure*}
-
-References to other parser rules are captured by links to children (interfaces or object nodes) in the AST nodes.
-For the \antlrparserrule{element} rule that is referencing the \antlrparserrule{content} rule, defined as in Figure~\ref{fig:elementrule}, we get the AST node representing the first alternative that is shown in Figure~\ref{fig:element_node_full}.
-There is one \mpsinterface{IElement} interface node that is implemented by two object nodes \mpsconcept{Element{\_}1}, \mpsconcept{Element{\_}2}.
-The node \mpsconcept{Element{\_}1} has two child links pointing to \mpsinterface{IAttribute} and \mpsinterface{IContent}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-\end{antlr}
-\caption{Element rule}
-\label{fig:elementrule}
-\end{figure*}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.75]{./images/element_node_full.png}
-	\caption{Element{\_}1 node's structure aspect}
-	\label{fig:element_node_full}
-\end{figure}
-
-A problem with this approach, which we call \emph{the layer problem}, is related to the grammar rules (and corresponding AST nodes) that represent intermediary layers created only to make the grammar more readable.
-The intermediary layers are created by authors of the grammars in order to make them more readable and more easily (better) maintainable.
-We illustrate this problem (and explain it) on the behavior of auto-completion in MPS.
-Imagine there is a freshly inserted \mpsconcept{Element{\_}1} node (representing the full XML element as is stated in the first alternative of the \antlrparserrule{element} parser rule).
-Now we would like to insert another nested XML element inside.
-MPS auto-completion mechanism would give us four options that are captured in the left part of Figure~\ref{fig:layer_problem}.
-This is because there exist four nodes (objects) that implement the \mpsinterface{IContent} interface as per each alternative of the rule.
-
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/layer_problem.png}
-	\caption{Layer problem in auto-completion}
-	\label{fig:layer_problem}
-\end{figure*}
-
-However, in order to correctly insert another nested element inside, a user would have to first insert the \mpsconcept{Content{\_}2} node inside \mpsconcept{Element{\_}1} that has an \mpsinterface{IElement} child inside (and nothing else).
-Then, as the second step, a user would invoke/trigger/call for the auto-complete again and insert either \mpsconcept{Element{\_}1} or \mpsconcept{Element{\_}2} inside \mpsconcept{Content{\_}2}.
-This mean that a user would have to go through two steps, and in the first stepe either correctly guess (or remember the grammar rule's alternative order), which option/item offered by the auto-complete menu to select.
-
-Similarly, if one would decide to replace the nested \mpsconcept{Element{\_}1} node with, let's say, an XML comment (a \mpsconcept{Comment} node), we need to delete both intermediary layers before we get back to the original \mpsconcept{Content{\_}X} selection (crossroads).
-The user cannot really see what is happening, since the intermediary level has no appearance or indicator.
-This may lead to confusion on the user's part.
-
-The layer problem is solved/addressed by the second approach, which we describe in the next subsection.
-
-\subsubsection{The Shortcut Approach}
-\label{sect:shortcut_approach}
-
-The key idea of this approach is to identify rules and nodes that are at the end of the (derivation, inheritance) chain, and offer them directly to the user through the auto-completion menu.
-We call such \emph{end rules} and \emph{end nodes}, respectively, because they cannot transparently break into more rules/nodes.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
-
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-
-	\antlrparserrule{comment}    :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-\end{antlr}
-\caption{Content parser rule with children}
-\label{fig:contentrulewithchildren}
-\end{figure*}
-
-For example, the \antlrparserrule{content} rule, given in Figure~\ref{fig:contentrulewithchildren} together with its child rules, can ultimately expand into following nodes:
-
-\begin{itemize}
-	\itemsep0em
-	\item \textbf{Content{\_}1} (TEXT)
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}1}
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}2}
-	\item Content{\_}3 $\rightarrow$ \textbf{Comment}
-	\item \textbf{Content{\_}4} (CDATA)
-\end{itemize}
-
-The end nodes are highlighted using the bold font.
-
-Figure XY shows the algorithm (in pseudocode) that finds all end rules for a given parser rule.
-
-% TODO pouzit standardni environment pro algoritmy 
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	FindPathsToEndNodes($R$):
-	1)  Define $L$ as an empty list of list of nodes
-	2)  Return FindPathsToEndNodes($R$, $L$)
-
-	FindPathsToEndNodes($R$, $L$):
-	3)  Define $Q$ as list of list of nodes
-	4)  For each alternative $A$ of rule $R$:
-	5)      $L1$ = Clone($L$)
-	6)      If $A$ is a parser rule with only one element $E$:
-	7)          Let $I$ be interface/concept representing rule $E$
-	8)          $L1$.Add($I$)
-	9)          $P$ = FindPathsToEndNodes($E$, $L1$)
-	10)         $Q$ = Merge($Q$, $P$)
-	11)     Else
-	12)         $L1$.Add($R$)
-	13)         Let $P$ be concept representing $A$
-	14)         $L1$.Add($P$)
-	15)         $Q$.Add($L1$)
-	16) Return $Q$
-\end{antlr}
-\caption{Algorithm for finding end rules}
-\label{fig:shortcut_algorithm}
-\end{figure*}
-
-To find end rules for each parser rule, the algorithm recursively scans through the parser tree that was built before.
-For each parser rule, it tries to find paths leading to some end rule through its alternatives:
-\begin{itemize}
-	\item Whenever it finds an alternative that contains only one element, and this element is a reference to another parser rule, it has found an intermediary level that can be transparently hidden from the user of the language.
-		The process (algorithm) continues by recursively processing alternatives of this "level" rule (since we are not at the end of the chain yet).
-	\item Otherwise, an end rule was found (recursion stops here).
-\end{itemize}
-
-By appending the rule that is leading to current element (line 12) and then appending that alternative's element itself (line 14), we get a path that contains the full path and the target end rule as the last element of the chain.
-The result of this algorithm, for example for the \antlrparserrule{content} rule, equals the listing of the five paths mentioned above.
-
-We do/repeat this for all parser rules and for each rule we get a list of paths that lead from that particular rule to an end node.
-We call these paths as \textbf{shortcuts}, because they provide a shortcut from the rule to the end of the chain.
-Now that we have the shortcuts, we will discuss several ways how to use the shortcuts (them) to improve the MPS language under construction (how to make the language better).
-
-The primary use of shortcuts is to generate options for the auto-completion menus such that only the end nodes are offered.
-This can be technically done in two ways --- (1) through custom auto-complete menus and deletion handlers or (2) by defining a special interface for each rule such that the interface is implemented only by end nodes for the rule.
-Shortcuts have to be considered when inserting nodes but also when deleting them, i.e.
-the whole chain (shortcut path) consisting of possibly multiple intermediary AST nodes (levels) up to the end node must be added, respectively deleted.
-The effect of insertion must be fully reversed when the user wants to delete some end node (the given end node).
-
-\subsection{Editor}
-
-When all types of AST nodes in the language are created, the next step is to define their visual representation in the projectional editor.
-As stated before (in Section X (about JetBrains MPS, part of background)), MPS uses a cellular system that allows placing node's properties and children into a table-like arrangement.
-Cells of different types are available --- for storing property values, references to child nodes, and keywords (string literals), and also cells that influence layout (e.g., indentation).
-
-The purpose (goals) of this (third) step is to create the editor aspect for each language element.
-It should project all attributes of the element --- name, properties, children, and so on --- using the respective cell types.
-
-The main challenge (problem) here is that an input grammar (ANTLR) does not provide/give any help/aid/information regarding the code layout.
-As we already said in the introduction, one of the biggest challenges/problems that arise here is that the grammar description of a language does not hold any information about the code layout.
-The structural rules in the grammar only tell us, what the syntax tree looks like and how the code is broken into nodes.
-In particular, the grammar says nothing about indentation, line breaks, and another formatting.
-
-We tackled this problem by designing an automated learning approach, which generates the layout using several heuristics.
-We present our solution to this problem. It involves some intermediary step whose purpose is to generate the necessary information.
-Our solution is semi-automated (using some heuristics) --- user's help in an interactive manner is needed for final touches (polishing).
-It works as follows.
-
-As input for the algorithm (learning approach), the user has to supply to a set of valid source files written in the imported language, together with the grammar file.
-These four steps/operations are then performed.
-\begin{enumerate}
-	\item Automated generation of an ANTLR parser for the input language using the ANTLR library.
-	\item Modify/extend the code of the generated parser in order to record some additional information.
-	\item Compile the parser into an executable form.
-	\item Use the parser to parse the supplied source code files and extract information about the layout of the code during this process.
-\end{enumerate}
-This way, the imported/created MPS language would inherit the code style/layout directly from the user, i.e. it would learn directly from the code.
-In step 2, the parser is modified such that it remembers/stores the line number and column for each parsed token.
-This is very efficient and it enables to create a nice layout.
-From the line number, the algorithm can easily determine whether there is a line break between some tokens (statements, language elements).
-The column number can be then used to detect indentation.
-However, this approach is still heuristic and may give imprecise (suboptimal) results --- mainly because it is not always possible to precisely match language elements (in the MPS structure) with all tokens that belong to it.
-Note also that quality of the extracted information always depends on the amount of the source code files provided (available, supplied) and on how representative the set of source code files actually is.
-When our algorithm cannot determine some custom layout, it just creates all the cells (content) of the editor aspect and arranges them in a single row.
-Figure~\ref{fig:editor_adjustment} XY shows the difference between the default single-row layout (in the upper left corner) and the fully customized layout (in the bottom right corner) for the \mpsconcept{Element{\_}1} node.
-
-% TODO smazat cervenou sipku z toho obrazku a mozna rozdelit na dve casti (vlozit tam caru): default single-row code layout, optimal fully customized layout
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[scale=0.5]{./images/editor_adjustment.png}
-	\caption{Projectional editor adjustment of the Element{\_}1 node}
-	\label{fig:editor_adjustment}
-\end{figure*}
-
-In practice, however, the user still has to tweak the editor aspect --- to adjust the code layout --- because the heuristics are not 100\% reliable.
-But it takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language~\cite{ref:javascript} and manually adjusting all editors that needed it, all that in a less than hour time.
-More details are provided later in Section X (examples) --- there we describe this issue further.
-
-\subsection{Text Generation}
-
-The final/last, fourth step, is to create the TextGen aspect for each AST node.
-The purpose of the TextGen aspect is to define, for a language element (AST node), what code in the BaseLanguage will be generated out of it, once we want to compile and run our MPS code/program.
-This is needed to allow/enable users to generate plain-text source code representation out of the AST they built/created inside MPS.
-Again, a TextGen aspect must be created for each language element.
-Basically, it is a single method in the BaseLanguage (or in Java?) for each language element.
-The method generates a string (plain-text) representation of the element, appends the string to the output stream, and then calls TextGen methods on all children of the element.
-
-A very basic example of BaseLanguage code that can be used as the TextGen aspect of an \mpsconcept{Element{\_}1} node, which represents the full XML tag with content, is shown in Figure~\ref{fig:textgen_example}.
-It appends all literals, properties, and children in the same order as they appear in the grammar definition.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=90mm]{./images/textgen_example.png}
-	\caption{Example of a~TextGen aspect}
-	\label{fig:textgen_example}
-\end{figure}
-
-Like in the case of editor, we discovered some interesting challenges related to the code layout, whitespaces, and more (so on).
-We describe our solution to the challenges here in this section.
-
-The reader might have noticed that there is no whitespace handling inside of our SimpleXML grammar (Section XY, running example).
-Grammars in the ANTLRv4 notation are usually written in such a manner that whitespace characters are skipped (ignored).
-However/Nevertheless, TextGen aspect (generator) should/must know between which children there must be a whitespace and where it is forbidden (i.e., where to put some whitespace characters) so that the produced code is a valid one.
-For example, in the case of the \antlrparserrule{element} rule, TextGen must determine that there should be a space in between \antlrparserrule{attribute} and \antlrlexerrule{Name}, while it is not required between \antlrliteral{\textless} and \antlrlexerrule{Name}.
-In general, it is sufficient to put just a single space character at places/locations where there has to be some whitespace in order to make the code valid (syntactically correct).
-
-Layout of the generated code is also important. For example, the user may want to produce more readable code by adding indentation or line breaks.
-
-In fact, the problem of TextGen layout is quite similar to the one we have seen with the projectional editor, described in Section XY.
-Since we are dealing mostly with text-based languages, their representation in projectional editor must be almost the same as their expected plain-text representation (output).
-Therefore, one could use the same approach for editor and TextGen --- adding line breaks and indentation.
-Once we would have some information about the layout for generating better projectional editor, we could leverage the same information and use it to improve TextGen.
-
-Nevertheless, in our current approach, the editor aspects have to be adjusted manually, after the import is done, because for now it is the fastest and the most efficient way how to deal with the problem of customizing the editor to capture a nice layout.
-Therefore, we used a different approach for TextGen --- a really simple heuristics which gives surprisingly good results.
-In the first step, a very basic TextGen is created that inserts spaces in between every two tokens of the language element (AST node).
-Then, as the second step, spaces are eliminated (restricted) from positions where they are not really needed.
-We describe all the cases here:
-
-\begin{enumerate}
-	\item Whenever there is a non-alphabetical literal,	appearing as a plain string token defined in the grammar, that might get recognized by the parser safely without the need for a whitespace separator around, then we omit spaces around it.
-		For example, consider \textbf{'\textless'} in SimpleXML.
-	\item Another group are strings inserted by the user of the language, whose form is constrained by a regular expression.
-		For example, an XML tag's name belongs to this group.
-		They are usually values of properties defined in structure aspects.
-		A space must be preserved only when a neighboring literal may end with an alphabetical character. Otherwise, we can omit the space.
-		In practice, this will eliminate redundant spaces inside of quotes, next to semicolons, and around brackets, but, on the other hand, it will separate language keywords (function, var, in, etc.) from other content by keeping the space character in between them.
-	\item Spaces are omitted also in the case of child nodes that are not present (e.g., when they are optional), and in the case of empty lists of child nodes, so that spaces do not accumulate.
-	\item On the other hand, when two child elements are next to each other, we always insert space.
-	\item Sequences of children are separated with space or with a line break.
-\end{enumerate}
-These few simple heuristics work surprisingly well and give us some very nice results.
-We discuss our experience with XML, JSON, and JavaScript in Section XY (Evaluation).
-
-Figure~\ref{fig:textgen_final} shows an example of a TextGen aspect for the full SimpleXML element.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=95mm]{./images/textgen_final.png}
-	\caption{Generated TextGen aspect for the SimpleXML element}
-	\label{fig:textgen_final}
-\end{figure}
-
-The automatically generated TextGen code can be adjusted manually very easily, so that it generates nicely indented XML code.
-In the case of the example above (in Figure F), we only need to wrap the \mpsconcept{Content{\_}2} child with indentation and change the sequence separator to a new line character.
-The resulting adjusted aspect (code) is shown below in Figure~\ref{fig:textgen_adjusted}.
-
-\begin{figure}[t!]
-	\centering
-	\includegraphics[width=85mm]{./images/textgen_adjusted.png}
-	\caption{Adjusted indentation inside the TextGen aspect}
-	\label{fig:textgen_adjusted}
-\end{figure}
-
-\subsection{Remarks about Grammars}
-
-As we have already shown in previous sections, it is very easy to write a grammar in a way that will cause problems during import into MPS (or once it is imported).
-Problems might occur during the creation of any aspect.
-Here we will look at some general problems/issues that grammar import poses (might pose).
-We will talk about these problems with respect to the challenges/obstacles we tried to overcome and which we described in previous sections (dedicated to structure, editor, and texgen, respectively).
-We also show a few examples of additional possible complications.
-
-\paragraph{Adjusting grammars.}
-There are some cases, where altering the input grammar might yield far better MPS language.
-We will show two examples how the usability of the resulting MPS language can be improved.
-Let us look at the definition of an XML attribute in Figure~\ref{fig:xmlattribute}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
-	\antlrlexerrule{STRING}      :   \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
-	            |   \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'}
-	            ;
-\end{antlr}
-\caption{Definition of an XML attribute}
-\label{fig:xmlattribute}
-\end{figure*}
-
-The original XML grammar has quotes as a part of the value.
-For the resulting MPS language, it would mean that there would be a placeholder for the attribute value that would expect us to input the leading and trailing quote together with the value too each time.
-It would also be marked red (by the syntax checker) unless we enter both quotes inside the value since the regular expression checking for quotes will not match.
-The user might be confused by this and will not be able to tell why his string value is incorrect.
-
-In our SimpleXML language, we adjusted the grammar easily in the manner that we show in Figure~\ref{fig:xmladjustgrammar}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
-	            |   \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'}
-	            ;
-	\antlrlexerrule{TEXT1}       :   \antlrregex{~["]*} ;
-	\antlrlexerrule{TEXT2}       :   \antlrregex{~[']*} ;
-\end{antlr}
-\caption{Adjusted grammar of SimpleXML}
-\label{fig:xmladjustgrammar}
-\end{figure*}
-
-We turned quotes into literals, thus ensuring that they will only appear in the projectional editor as fixed constant cells.
-We will not have to encapsulate the value in them each time.
-The user will only have to choose, which attribute version he wants to use (single or double quotes).
-
-As the second example, we will show the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
-Every statement in JavaScript needs to be either followed by a semicolon, newline, file end or end of the block --- see the Figure~\ref{fig:javascriptstmt}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{eos}            : \antlrlexerrule{SemiColon}
-	               | \antlrlexerrule{EOF}
-	               | {lineTerminatorAhead()}?
-	               | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}?
-	               ;
-	\textcolor{gray}{// Example reference of the eos rule}
-	\antlrparserrule{breakStatement} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos}
-	               ;
-\end{antlr}
-\caption{Statements in JavaScript}
-\label{fig:javascriptstmt}
-\end{figure*}
-
-Because there are multiple options, our import plugin would create a placeholder at the end of every statement.
-Every language element representing a statement would have one child of the \mpsinterface{IEos} interface type.
-This placeholder needs to be manually filled in for each statement.
-
-Since the projectional editor has much bigger power over the form of the code, we (a user) might want to have each statement on a separate line.
-Since we can differentiate between statements on the AST level, we do not need an explicit separator between them.
-This means that we might want to simplify the language and leave the semicolon out, or leave it just as a constant fixed part of the projectional editor, but not as something the user must explicitly fill in.
-Then we can just put each statement on a separate line as it is usual for JavaScript code, but we do not need the semicolon anymore.
-This small adjustment is very quick when done inside the grammar.
-We just change the \antlrparserrule{eos} rule to the form shown in Figure~\ref{fig:eosrule}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{eos} : \antlrliteral{;} ;
-\end{antlr}
-\caption{eos parser rule}
-\label{fig:eosrule}
-\end{figure*}
-
-This way, we can change the grammar in a way that it will not describe the same ECMAScript language as before, but will definitely make our MPS language more usable.
-TextGen aspect may be defined in such a way that puts the semicolon back (after each statement) in the generated plain-text representation. 
-This would be very important especially when JavaScript is available as another base language (which is actually a plan (future work) of the JetBrains company).
-
-Above, we have shown that adjusting the input grammar might be a very fast mean of tweaking/improving the usability of generated MPS language, and sometimes it may be the only proper solution in some complex situations.
-In general, the question we can ask ourselves is whether one aims for a full precise/faithful port of the input language or for an MPS alternative.
-
-\todo{Nahradit slovo "break" v nasledujicich odstavcich (v kontextu "break the parser") za vhodnejsi slovo ktere pusobi vic odborne.}
-
-\paragraph{Breaking grammars and parsers.}
-However, we have found that there is one big problem with grammar adjustment that we would like to point out.
-The problem is that it is very easy to change the input ANTLR grammar in a way that breaks the ANTLR parser generated out of it.
-By breaking we mean that it stops parsing the original language.
-We illustrate the problem on our SimpleXML language.
-
-As stated before, when creating the SimpleXML grammar, we have started off with the original XML grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and did some adjustments to it.
-We ended up with a grammar that can be processed by our method/plugin and creates a nice usable MPS language.
-Nevertheless, we noticed that even though the imported language behaves well enough and mimics the XML language quite nicely, the ANTLR parser generated out of this grammar no longer parses XML successfully.
-Some changes we have made, such as the attribute adjustment, broke the grammar down.
-More precisely, our changes improved the MPS language while breaking down the parser, and we have not even noticed it because, from the perspective of MPS, the imported language still corresponds to XML.
-
-What we are trying to say (show here) is that it is very easy to perform a grammar adjustment that will, at first, seem harmless and valid inside MPS, but will break the parser.
-The problem is that the user performing this change probably will not be aware of breaking it.
-
-The main (underlying) cause of this problem is the way the ANTLR parser is implemented and the quite different purpose we are using the grammar for in our project/tool/method.
-There are many ways how various parsers deal with, for example, token matching.
-For instance, ANTLR introduces so called \emph{greedy} and \emph{non-greedy} operators.
-The greedy way, in which ANTLR matches input on defined tokens and prioritizes their selection, makes some rules very dangerous.
-When a rule matches a wide range of input, it might happen during parsing that it is prioritized over other rules and swallows a lot more input than the author of it intended.
-Usually, these dangerous rules are bound to some parser context, which makes them behave well.
-
-However, for practical reasons, it is very important that grammar is not broken by our custom adjustments (and that the ANTLR parser of the language works properly), because we might need to automatically generate parsers of the adjusted language.
-Imagine this very real scenario (that will be supported in the future):
-\begin{enumerate}
-	\item We have imported the language inside MPS.
-	\item MPS now knows the structure of the language, and we can code in MPS using this language.
-	\item We would expect MPS to be able to load an existing source code, written in this language, from a text file and import it inside MPS.
-	\item We would like to use MPS to safely edit this code, using all the features of MPS.
-	\item We would also like to export the code in a plain-text form again and save it back to the source file.
-\end{enumerate}
-In order to make this happen, we must be able to generate a correctly-working ANTLR parser out of the adjusted source grammar that we used to (that helped us) import the language.
-We must also be able to match nodes of the AST coming out of the ANTLR parser to elements of the MPS language, and we must build the MPS AST out of the ANTLR AST.
-
-\section{Evaluation}
-
-We implemented everything (the proposed approach described above) in the form of an MPS plugin that allows users to carry out the import.
-
-We used a mix of Java and BaseLanguage in our implementation.
-We used the MPS API, which can be called from the BaseLanguage programs too, to generate some of the language elements.
-BaseLanguage is needed to use/call the MPS (language) API, i.e. to programmatically create new language elements and their aspects.
-
-The plugin uses the ANTLR library to work with ANTLRv4 grammar files (to parse them, etc) and several MPS libraries (that implement the MPS API).
-
-For the first step of automated construction of MPS languages from grammars (i.e., for parsing the grammar), the plugin uses the ANTLR library~\cite{ref:ANTLR4} to generate a Java ANTLR parser of the ANTLRv4 notation, because there exists an ANTLRv4 grammar for the ANTLRv4 notation~\cite{ref:ANTLR4reference}.
-
-\todo{
-	We should provide url to the github repository (implementation) and pointer to some package with ANTLR grammars and imported MPS languages (take from Attachments.zip in the master thesis).
-}
-
-\todo{Popsat nase zkusenosti s importem slozitych jazyku (XML, JavaScript, Python, a dalsi).}
-
-From introduction: We will also show some example languages imported with INGRID and discuss our experience.
-
-In order to evaluate INGRID, we tried to import several real/mainstream languages:
-\begin{itemize}
-	\item \textbf{SimpleXML} --- Simplified XML that we used in this thesis for explanatory purposes.
-	\item \textbf{JSON} --- Similar to XML, but this time a full port of the specification.
-	\item \textbf{ECMAScript 5.1} --- Specification of the language known as JavaScript, dated to the year 2011, which is currently the most frequently adopted version.
-\end{itemize}
-Results of the import process are included in the package that is available at \todo{url}.
-The repository \todo{Add url} contains an example MPS project. There are three languages.
-For each example language, you can find its original import, exactly in the form as created by our plugin.
-
-We experimented with import of XML, JavaScript, Python, and maybe some other languages.
-
-These general purpose languages are quite complex when it comes to their structure and the overall syntax variety.
-
-We must emphasize that the automatically imported languages are not ready-to-use full-fledged MPS languages.
-We did not expect to achieve that at this stage, as that is a much bigger, if not impossible, challenge.
-Imported languages contain full structure found in the respective grammars and some more aspect definitions that will help users of the languages in creating code.
-The imported language has to be manually adjusted (tweak, polished) by the end user (developer) and some of its complications have to be resolved manually by a human.
-
-We found out that manual adjusting of the editor/textgen aspect (adding line breaks and indentation) is far more effective and precise than any automated approach we came up with (during our research).
-The imported language can be, in our opinion, adjusted very fast into a more readable form.
-
-The package at \todo{url} contains also adjusted versions of all three languages.
-Then, for each language, there is an adjusted version.
-
-This adjusted version was created from the original import and the author spent no more than between 20 to 60 minutes of working on it.
-The first author spent only a couple of minutes of (manually) adjusting them after the automated import, so that we try to prove our point above.
-More specifically, the JSON language was ready for use in no more than 20 minutes of minor adjusting.
-We chose this language because it does not require us to implement complex aspects, such as type checking or data flows.
-Similarly, the SimpleXML turned out very nice.
-
-Our main goal was to prove that even though our method (plugin) does not create a perfect language, it can be customized very fast into a very usable form.
-We manually tweaked/customized the editor and the TextGen aspect only, structure aspect was left in its original form.
-This justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical (and the best).
-
-We have also tried to use the INGRID tool/plugin to import the JavaScript language, which is exactly the type of a complicated general purpose language that might be of interest to other people.
-There are already some projects, where a manual port of JavaScript is being done.
-One of them is the ECMAScript4MPS~\cite{ref:ECMAScript4MPS} from the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section X (similar projects, related work).
-The result of import by INGRID is quite good, comparable to the PE4MPS project.
-Our language is, of course, missing a lot of more advanced aspects/features that, for example, can help deriving types of expressions and so on.
-These aspects improve the usability by a lot, e.g.
-when the user starts writing numbers, a number literal is inserted, whereas in our import the user has to first insert the concept representing the number literal and then proceed with writing numbers.
-These actions are needed in any MPS language and are expected to be implemented after the import step, \todo{Maybe say that INGRID does not generate these actions/aspects/features yet but it is our future work} as it is not possible to generate them automatically.
-However, when it comes to language structure, concept aliases, and auto-completion, we think that INGRID (our method/plugin) did a very good job.
-The structure aspect is very similar to the ECMAScript4MPS's one, which was created manually over the course of a large number of hours.
-INGRID does that (achieves the same result) fully automatically.
-
-To summarize, import languages such as JavaScript, JSON, MySQL and others is giving nice results.
-
-Results of our experiments with languages such as JSON or JavaScript show that the proposed approach (INGRID) yields very good results, comparable to existing manual ports, and spares the user from hours of tedious, error prone and time-consuming work.
-
-Limitation: the project is lacking proper error reporting and cannot handle cyclic rules (which could be found in Python grammar, for example).
-
-\section{Conclusion and Future Work}
-
-\todo{Write something.}
-
-There is lot of further follow-up work to be done. Many open problems remain.
-
-Automated generation of the editor aspect, probably using some kind of machine learning.
-Generating nicely designed and user-friendly editors, maybe using artificial intelligence (machine learning) that would evaluate (analyze) existing code in plain text and based on that create the editors.
-
-Follow-up work might look more into the problem of code layout detection, possibly exploring suggested approaches such as the learning approach described in Section XY \todo{Tohle se snad podari aspon castecne implementovat}.
-This could lead to better results in editor and TextGen aspect generation which could definitely use some improvement.
-Another future endeavors will definitely touch the subject of generating parsers that would enable the user to import existing source code into MPS.
+\input{chapters/introduction}
+\input{chapters/background}
+\input{chapters/language_definition}
+\input{chapters/evaluation}
+\input{chapters/conclusion}
 

### 2016-09-23 chapters/background.tex
meta-bnf project description

@@
 
-\JB{Navrhuji odstranit} We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
+\JB{Navrhuji odstranit}
+\PV{Souhlasim}
+We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
 Focus on basic/fundamental and relevant features. In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.
@@
 \JB{Vyhody/nevyhody diskutujeme v kapitole 1. Navrhuji nasledujici odstavec zrusit nebo text z kapitoly 1. presunout sem}
+\PV{Spis bych zrusil tady, kapitola 1 mi prijde vhodnejsi pro zmineni motivace}
 Keeping the model of the code in the AST form has several advantages.
@@
 \paragraph{PE4MPS.}
-PE4MPS~\cite{ref:PE4MPS} is a project by \todo{Doplnit autora} that is trying to solve the same problem as we do.
+PE4MPS~\cite{ref:PE4MPS} is a project by Marco Lombardo that is trying to solve the same problem as we do.
 It solves the lack of information about code layout in grammars by creating a new grammar notation called PE Grammars~\cite{ref:PE} (the PE abbreviation comes from projectional editing).
@@
 \paragraph{ANTLR{\_}MPS.}
+\PV{Tenhle projekt byl z tech tri nejmene uzitecny a podle me i nejmene funkcni, tak mozna by se dal v ramci zkracovani vypustit. V dalsim projektu je na nej reference, tak treba ji vyhodit.}
 ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
-The author of this project is Fabien Campagne
+The author of this project is Fabien Campagne.
 The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
@@
 \paragraph{mps-metabnf.}
-\todo{Popis na zaklade emailove diskuze (Nizozemci).}
+\PV{Pridal jsem popis projektu}
+% TODO - citace na https://github.com/DSLFoundry/mps-metabnf
+The mps-metabnf project comes from the DSLFoundry group and also takes on grammar importing.
+Even though it is in a very early stage of development, it holds some interesting ideas.
+Similarly to ANTLR{\_}MPS, authors of mps-metabnf decided to create an MPS language describing the grammar.
+When grammar is being imported into MPS, it is just transformed into the terms of this MPS language.
+User is then able to adjust this grammar.
+After this, a second step of the import process is needed, that generates the final MPS language out of this adjusted grammar definition.
+
+The key takeaways from the mps-metabnf project are that the intermediate grammar MPS representation, when finished, will offer some powerful means of adjustment.
+Using the projectional editor, the user can easily specify information about the code layout, export the target MPS language and in case further adjustments are needed, come back to the grammar definition and fix any problems.
+The MPS language of the grammar definition can be easily extended with needed features (indenting, line breaks...), which is superior to our approach.
+
+It also has some downsides.
+Because MPS doesn't offer any tools that would allow easy transformation from the MPS grammar to the MPS language, the second step of the import process hasn't been implemented yet and will be probably very problematic.
+Another downside might be the need for a double transformation implementation -- first one being the import of the text grammar to the MPS grammar definition.

### 2016-09-23 chapters/background.tex
Notes to evaluation

@@
 The key takeaways from the mps-metabnf project are that the intermediate grammar MPS representation, when finished, will offer some powerful means of adjustment.
-Using the projectional editor, the user can easily specify information about the code layout, export the target MPS language and in case further adjustments are needed, come back to the grammar definition and fix any problems.
-The MPS language of the grammar definition can be easily extended with needed features (indenting, line breaks...), which is superior to our approach.
+Using the projectional editor, the user can easily specify information about the code layout, export the target MPS language and, in case further adjustments are needed, come back to the grammar definition, fix any problems and regenerate the language.
+The MPS language of the grammar definition can be easily extended with needed features (indenting, line breaks...), which is superior to our approach, because they can be represented using many visual means.
 
-It also has some downsides.
+This approach also has some downsides.
 Because MPS doesn't offer any tools that would allow easy transformation from the MPS grammar to the MPS language, the second step of the import process hasn't been implemented yet and will be probably very problematic.
 Another downside might be the need for a double transformation implementation -- first one being the import of the text grammar to the MPS grammar definition.
+
+From another point of view -- for each additional feature, that we decide to add to the grammar notation, we need to implement a corresponding transformation.
+This subsequently leads to duplicating the projectional editor that is already present in MPS.
+This means that there needs to be a line drawn between which features are added and which will be left for the user to add to the target language using MPS.

### 2016-09-23 chapters/evaluation.tex
Notes to evaluation

@@
 Results of the import process are included in the package that is available at \todo{url}.
-The repository \todo{Add url} contains an example MPS project. There are three languages.
+The repository \todo{Add url} contains an example MPS project with three languages.
 For each example language, you can find its original import, exactly in the form as created by our plugin.
 
-We experimented with import of XML, JavaScript, Python, and maybe some other languages.
+We experimented with import of XML, JavaScript, Python, Ruby, C#, Swift, MySQL and some other languages.
 
-These general purpose languages are quite complex when it comes to their structure and the overall syntax variety.
+Most of these languages are quite complex when it comes to their structure and the overall syntax variety.
 
@@
 Imported languages contain full structure found in the respective grammars and some more aspect definitions that will help users of the languages in creating code.
+\PV{The imported language has to be manually tweaked by the end user.}
 The imported language has to be manually adjusted (tweak, polished) by the end user (developer) and some of its complications have to be resolved manually by a human.
@@
 We manually tweaked/customized the editor and the TextGen aspect only, structure aspect was left in its original form.
-This justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical (and the best).
+This justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical (and the best\PV{?}).
 
@@
 
-To summarize, import languages such as JavaScript, JSON, MySQL and others is giving nice results.
+\PV{Veta se opakuje nize, odstranil bych tuhle}
+To summarize, import languages such as JavaScript or JSON and others are giving nice results.
 
@@
 
+\PV{Tohle uz neni pravda, error reporting i cyklicka pravidla jsem pred odevzdanim DP vyladil a kod celkove dodelal.
+Cyklicka pravidla to tedy neumi, jen na ne upozorni (a rekne, ktere pravidlo je obsahuje).}
 Limitation: the project is lacking proper error reporting and cannot handle cyclic rules (which could be found in Python grammar, for example).

### 2016-10-02 chapters/introduction.tex
writing introduction

@@
 
-\JB{Odstavec mirne upraven} The prevailing approach to define valid syntaxes for programming languages is through grammars written typically in notations based on (or similar to) EBNF [reference].
-Many popular tools exist for automated generation of parsers from the grammar definitions, such as ANTLR and Bison/Flex.
-The source code of programs in such languages is written in a text editors, usually embedded into an IDE.
+The prevailing approach to define valid syntaxes for programming languages is through grammars, which are typically written in notations based on the Extended Backus-Naur Form (EBNF).
+Many popular tools exist for automated generation of parsers from the grammar definitions --- for example, ANTLR~\cite{ref:ANTLR} and Bison/Flex~\cite{ref:BISONFLEX}.
+The source code of programs is written in a text editor, usually embedded into an IDE. Parsers are used to create an abstract syntax tree (AST) representation of the code, e.g. inside the compiler.
 
-\JB{Odstavec mirne upraven} Projectional (or structural) editing offers an alternative approach.
-While the idea emerged as early as in 1970s, it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
-A recent revival of projectional editing has been observed in the area of language workbenches - IDE-like tools that enable the developers to manipulate the actual language definition.
-The Meta-programming System (MPS) [reference] by JetBrains is an open-source language workbench that focuses of Domain Specific Languages and leverages the concept of projectional editor.
-MPS allows developers to design custom languages and then use use them to write code and compile it into executables.
-MPS provides the whole IDE infrastructure for such languages.
+An alternative approach is projectional (or structural) editing, where a developer directly manipulates the AST representation of the source code instead of plain text.
+This idea emerged as early as in 1970s, but it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
+A recent revival of projectional editing has been observed in the area of language workbenches --- IDE-like tools that enable the developers to manipulate the actual language definition.
+The Meta Programming System (MPS)~\cite{ref:MPS} by JetBrains is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
+MPS provides the whole IDE infrastructure that enables developers to design custom languages, use them to write program code, and compile the code into executables.
+We provide more technical details about MPS in Section~\ref{sect:MPS}.
+In the rest of the paper, we will use MPS to illustrate the general concepts and issues that apply also to other language workbenches.
 
-\JB{Odstavec mirne upraven} The key idea behind projectional editors is that a developer directly manipulates the AST representation of the source code instead of plain text.
-To avoid the fate of the early projectional editors MPS has put extra effort into making text manipulation feel like editing real text.
-MPS models, although being persisted as XML or binary files, can be versioned using the industry standard source-control systems (Git, Subversion).
-On-line model browsers have been incorporated into some of the popular public source code repositories (BitBucket) [reference] .
+%MPS has put extra effort into making text manipulation feel like editing real text.
+%MPS models, although being persisted as XML or binary files, can be versioned using the industry standard source-control systems (Git, Subversion).
+%On-line model browsers have been incorporated into some of the popular public source code repositories (BitBucket) [reference] .
 
-\JB{Odstavec mirne upraven} The absence of parsers brings about numerous benefits, which are in particular useful for DSLs, which are frequently used by domain experts, who may not be professional programmers.
+Projectional editing, and in particular the absence of parsers, brings several benefits.
 \begin{itemize}
-	\item Since a projectional editors controls the interaction between the user and the code, the developers practically cannot enter syntactically invalid code.
-	\item Languages can be defined in a modular way and multiple languages can be easily combined together or extend one another, while parsers seriously limit language modularity for traditional parsed languages.
-	\item The languages may offer contextual or non-parseable notations, such as tables, diagrams or positional syntax.
-	\item Since the projection is detached from the physical representation of code (AST), authors can define multiple notations and allow the developers to switch between different representations of the code.
+	\item The developers practically cannot enter syntactically invalid code, because a projectional editor controls the interaction between the user and the code.
+	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together or one can extend another, while parsers seriously limit language modularity for traditional parsed languages.
+	\item The languages may support contextual or non-parseable notations, such as tables, diagrams and positional syntax.
+	\item Since the projection is detached from the physical representation of code (AST), authors of languages can define multiple notations and allow the developers to switch between different representations of the code.
 \end{itemize}
-A motivating illustration of the abilities of MPS is the mbeddr project [citation], which extends C with numerous domain-specific constructs and data types.
-We provide more details about MPS in Section 2.1.
+All of this is useful especially for DSLs, which are frequently used by domain experts who may not be professional software developers.
+The mbeddr project~\cite{ref:MBEDDR}, which extends C with numerous domain-specific constructs and data types, illustrates the abilities of language workbenches in general and of MPS in particular.
 
-\JB{Navrhuji smazat nasledujici odstavec}
-\todo{From thesis.
-	One of these tools is the JetBrains MPS~\cite{ref:MPS}, which approaches languages from a~different point of view.
-	MPS doesn't use the textual representation of code as usual, but rather works with the actual abstract syntax tree that holds the code's structure.
-	This has some positive implications as it gives us a~very powerful tool rich on features but also introduces some new problems we haven't encountered before.
-	One of them being the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-	MPS also becomes a~very powerful tool once the language has already been created and you start using it inside MPS.
-	There is a~huge variety of new languages constructed for MPS that usually solve domain specific problems.
-}
+On the other hand, usage of projectional editors introduces some new problems.
+Before a language can be used inside MPS, it has to be defined in a special way through the MPS infrastructure, so that the IDE can understand the language and work with it.
+Specifically, an author of a language has to create:
+\begin{enumerate}
+	\item the abstract syntax called \emph{Structure}, which defines the types of allowed AST nodes,
+	\item the concrete syntax called \emph{Editor}, i.e. projection of the AST on the screen and its interactions with the user, and
+	\item \emph{text generation} to enable creation of plain text files used as input for compilers.
+\end{enumerate}	
+Again, we provide additional technical details in Section~\ref{sect:MPS}.
 
-\todo{Pouzivat jen terminy "AST nodes" a "language elements" (to spis). Nepouzivat pojem "concepts" abychom ctenare nematli.}
-
-\JB{Odstavec upraven} Projectional editing introduces some new problems we haven't encountered before.
-One of them is the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
-In order to use a language inside MPS, it has to be defined through the MPS infrastructure.
-In particular, one has to create the abstract syntax called Structure (the types of allowed AST nodes), concrete syntax called Editor (projection of the AST on the screen and its interactions with the user), and text generation (to enable creation of plain text files used as input for compilers).
-We provide more details in Section X (Background about MPS).
+% TODO continue from here
 
@@
 We will also show some example languages imported with INGRID and discuss our experience.
+
+Although we will explain/describe the challenges/problems and our solutions in the context of MPS and ANTLR, we believe that they apply (are relevant) more generally - i.e., also for other language workbenches (to authors of other language workbenches) and parser generators.

### 2016-10-02 paper.tex
meta-bnf project description
writing introduction

@@
 
-\newcommand{\todo}[1]{{\bfseries #1}}
+\newcommand{\todo}[1]{{\bfseries TODO #1}}
 
@@
 % colors for the ANTLR environment
-\definecolor{antlrbgcolor}{rgb}{.962, .962, .962}
+\definecolor{antlrbgcolor}{rgb}{1,1,1}
 \definecolor{antlrparserrulecolor}{RGB}{11,83,148}
@@
 % colors for the MPS environment
-\definecolor{mpsbgcolor}{rgb}{.962, .962, .962}
-\definecolor{mpsconceptcolor}{RGB}{42, 140, 0}
+\definecolor{mpsbgcolor}{rgb}{1,1,1}
+\definecolor{mpsconceptcolor}{RGB}{42,140,0}
 \definecolor{mpsinterfacecolor}{RGB}{170,0,229}
@@
 \newcommand{\antlrregex}[1]{\textcolor{antlrregexcolor}{#1}}
-% TODO mozna zmenit "mpsconcept" na "mpslangelem"
 \newcommand{\mpsconcept}[1]{\textcolor{mpsconceptcolor}{#1}}
@@
 
-% TODO zmenit barvu pozadi na bilou
 % environment for the ANTLR code (grammars)
@@
 	\noindent
-	\adjustbox{margin=3mm,bgcolor=antlrbgcolor,frame,minipage=394pt,inner=\linewidth}
+	\adjustbox{margin=3mm,bgcolor=antlrbgcolor,frame,minipage=485pt,inner=\linewidth}
 	\bgroup
@@
 % TODO
-%\conferenceinfo{OOPSLA'12,} {October 19--26, 2012, Tuscon, Arizona, USA.}
-%\CopyrightYear{2012}
-%\copyrightdata{978-1-4503-1561-6/12/10}
+%\conferenceinfo{PLDI'17,} {June 19--26, 2017, Barcelona, Spain.}
+%\CopyrightYear{2017}
+%\copyrightdata{TODO}
 
-% TODO
-\title{\todo{INGRID: Creating Languages in MPS by Importing ANTLR Grammars}}
-% alternative: "INGRID: Importing Grammars Into MPS"
+\title{\todo{INGRID: Creating Languages in MPS from ANTLR Grammars}}
 
@@
 
-% TODO opravit formatovani jednotlivych zaznamu a vhodne seradit
 \begin{thebibliography}{}
 
+\bibitem{ref:MBEDDR} M. Voelter, D. Ratiu, B. Schaetz, and B. Kolb. mbeddr: an Extensible C-based Programming Language and IDE for Embedded Systems. SPLASH 2012, ACM.
+
+\bibitem{ref:ANTLR} ANTLR parser generator, \url{http://www.antlr.org}
+
+\bibitem{ref:BISONFLEX} J. Levine. Flex \& Bison: Text Processing Tools. O'Reilly Media, 2009.
+
 \bibitem{ref:MPS} JetBrains MPS, \url{https://www.jetbrains.com/mps/}
+
+% TODO opravit formatovani nasledujicich polozek a vhodne zaradit
 

### 2016-10-02 chapters/introduction.tex
finished introduction

@@
 	\item the concrete syntax called \emph{Editor}, i.e. projection of the AST on the screen and its interactions with the user, and
-	\item \emph{text generation} to enable creation of plain text files used as input for compilers.
+	\item text generation scripts, also called \emph{TextGen}, to enable creation of plain text files used as input for compilers.
 \end{enumerate}	
@@
 
-% TODO continue from here
+MPS is typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
+Before a program written in such a DSL can be executed, it is transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose language (GPL), such as C or Java.
+This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
+Therefore, the target GPL must also be defined by the MPS means in order to allow the DSLs to have their models transformed into a model in the GPL.
 
-\JB{Odstavec pridan pro vysvetleni motivace prace} MPS is typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
-These DSLs are then transformed on the AST level by a series of model-to-model transformations to an AST model representing the desired semantics in a General-purpose language (GPL), such as C or Java.
-This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
-This implies that the target GPL must also be defined by the MPS means, to allow the DSLs to have their models transformed into a model in the GPL.
+However, very few mainstream GPLs are now fully supported by MPS, because it requires substantial effort to manually create a full definition of a GPL in MPS.
+Only Java and C have been implemented to date, which limits the authors of DSLs to only target these supported GPLs.
+The overall goal of our project is to automatize the process of language definition in MPS as much as possible, exploiting the fact that a big part of this process is quite straightforward.
+We believe that such an automated process would encourage and speed-up the migration of more GPLs into MPS, thus giving the authors of DSLs more options.
+Other possible applications include the support for writing programs in GPLs using the projectional editor and combination of GPLs with domain-specific languages natively created in MPS.
 
-\JB{Odstavec doplnen} However, it requires substantial effort to create a definition of a GPL in MPS.
-Therefore, very few mainstream GPLs are now fully supported by MPS.
-Only Java and C have been implemented to date, which limits the DSLs authors to only target these supported target GPLs.
-We believe that an automated process of importing grammars of GPLs into MPS would encourage more GPLs to be migrated to MPS and thus giving the DSL authors more options.
+In this paper we present INGRID, a method and a tool for semi-automated construction of the definition of a programming language in MPS that uses an ANTLR grammar of the language as a description of its syntax.
+We discuss the main challenges that we encountered during the development of INGRID and our solution to them.
+Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges are general. 
+They apply to the relation between two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
+Specifically, the description of a language in the form of a grammar does not hold any information about the code layout, and the grammar typically contains many rules that do not directly correspond to AST nodes and programming language constructs.
+We provide more details in Section~\ref{sect:LANGDEF}.
 
-\todo{From thesis.
-	However, building these languages is a~complicated and time-consuming process and a~lot of effort must be put into them before they are ready for use.
+The INGRID tool can help also to make the construction of DSLs more efficient.
+Based on our experience, we believe that, especially for simple languages, it is less time consuming to write their grammar by hand and then create the language definition in MPS using INGRID, than to create the language manually using the MPS GUI.
 
-	These examples show that recreating a~full language inside MPS is not an easy task and a~lot of time must be spent on implementing all aspects of the language.
-	A big part of this effort is, however, quite straightforward and can be possibly automated, as we show in this paper. Automation speeds up the process of adoption of new languages to a great degree.
-}
+%INGRID allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
 
-\JB{Odstavec upraven} Since a big part of the language definition process in MPS is quite straightforward and can possibly be automated, we will look into the possibility of automatic import of already existing general purpose languages into MPS using a grammar description of their syntax.
-
-We present INGRID --- a method and a tool for semi-automated construction of MPS definition of programming languages based on their ANTLR grammars.
-We also discuss challenges that we encountered during this process and our solution to them.
-Most of these challenges are general, i.e.
-not MPS specific, as they apply to the relation between grammars defined in EBNF-like style (rules and tokens, everything defined in plain text) and structured (object-oriented) approach used in MPS (and other language workbenches).
-
-\todo{Hodne zduraznit automation, tedy ze chceme maximalizovat automatizaci.}
-The main goal of our/the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible.
-
-\todo{From thesis.
-	Automatic import of already existing languages into MPS from the grammar description of their syntax.
-}
-\todo{Mozna prodat INGRID jako automatickou konverzi gramatik do podoby MPS languages, tedy jako nastroj ktery umoznuje portaci jazyku do prostredi MPS a jejich nasledne pouziti v MPS (treba kombinace s dalsimi jazyky).}
-
-Our primary motivation is to enable better support for mainstream and/or general purpose programming languages (XML, C, C++, JavaScript, Python) in MPS, especially to enable writing programs in these languages using the projectional editor and to combine them with domain-specific languages natively created in MPS.
-\todo{Zduraznit take pouziti importovanych jazyku pro cile generovani z DSL --- to plati zejmena u mainstream general-purpose languages jako treba JavaScript.}
-
-\todo{Briefly mention the biggest challenges.}
-The are related to the way grammar encoding/description of languages are usually written.
-They do not hold any information about the code layout, and contain many rules that do not directly correspond to AST nodes (programming language constructs).
-We discuss them in more detail later (Section X).
-
-\todo{From thesis.
-	INGRID allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
-	The biggest benefit is the possibility to code in these mainstream general purpose languages using the projectional editor.
-}
-
-However, we do not target just the very complex general purpose languages such as JavaScript.
-INGRID can be used also for simple languages.
-We believe that, for simple languages, it is less time consuming to create the grammar by hand (e.g., using the ANLTR notation and toolset) and then create the language definition in MPS by using the INGRID tool, than to create it manually using the MPS GUI --- this is another benefit of INGRID.
-
-Structure of the paper.
-In the beginning, we will describe the MPS editor and related/similar projects.
-Then, we describe our approach to creating language definition in MPS from their grammars, and discuss the major challenges together with our solution.
-We will also show some example languages imported with INGRID and discuss our experience.
-
-Although we will explain/describe the challenges/problems and our solutions in the context of MPS and ANTLR, we believe that they apply (are relevant) more generally - i.e., also for other language workbenches (to authors of other language workbenches) and parser generators.
+The rest of the paper has the following structure.
+In the next section, we provide necessary information about MPS and a brief overview of projects with similar goals as INGRID.
+Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
+We also discuss our experience with usage of INGRID on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.

### 2016-10-04 chapters/language_definition.tex
finished introduction
tweak code listing

@@
 \section{Creating Language Definition in MPS}
+\label{sect:LANGDEF}
 
@@
 \centering
-\begin{antlr}
+\begin{framed}
+\begin{alltt}
 	\textbf{grammar} \textit{SimpleXML};
 	\antlrparserrule{document}  : \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
-	\antlrparserrule{prolog}       :   \antlrliteral{<?xml } \antlrparserrule{attribute}* \antlrliteral{?>} ;
+	\antlrparserrule{prolog}    : \antlrliteral{<?xml } \antlrparserrule{attrib}* \antlrliteral{?>} ;
 	\antlrparserrule{comment}   : \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-	             |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-	             ;
-	\antlrparserrule{attribute}    :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
-	\antlrparserrule{content}      :   \antlrlexerrule{TEXT}
-	             |   \antlrparserrule{element}
-	             |   \antlrparserrule{comment}
-	             |   \antlrlexerrule{CDATA}
-	             ;
+	\antlrparserrule{element}   : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{/>} ;
+	\antlrparserrule{attrib}    : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
+	\antlrparserrule{content}   : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
 	\antlrlexerrule{Name}      : \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
@@
 	\textbf{fragment}
-	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
-	             |   \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.}
-	             |   \antlrlexerrule{DIGIT}
-	             ;
+	\antlrlexerrule{NameChar}  : \antlrlexerrule{NameStartChar} | \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
 	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
+	\antlrlexerrule{NameStartChar} : \antlrregex{[:a-zA-Z]} ;
 	\antlrlexerrule{TEXT}      : \antlrregex{~[<"]*} ;
 	\antlrlexerrule{CDATA}     : \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
-\end{antlr}
+\end{alltt}
+\end{framed}
 \caption{SimpleXML grammar}
@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
 \centering
-\begin{antlr}
+\begin{framed}
+\begin{alltt}
 	\antlrlexerrule{Name}     : \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
@@
 	\antlrlexerrule{NameChar} : \antlrlexerrule{NameStartChar} 
-             |   \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.}
-             |   \antlrlexerrule{DIGIT}
-             ;
+	         | \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
 	\textbf{fragment}
-	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
-\end{antlr}
+	\antlrlexerrule{NameStartChar} : \antlrregex{[:a-zA-Z]} ;
+\end{alltt}
+\end{framed}
 \caption{Lexer rule from SimpleXML}
 \label{fig:lexerrule}
-\end{figure*}
+\end{figure}
 
@@
 % TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
-\begin{figure*}[ht]
+\begin{figure}[ht]
 \centering
-\begin{antlr}
+\begin{framed}
+\begin{alltt}
 Flatten(\textit{R}):
-	1) Define \textit{T} as a list of empty string
-	2) For each alternative \textit{A} of the rule \textit{R}:
-	3)     Define \textit{R} = \antlrap\antlrap
-	4)     For each element \textit{E} of \textit{A}:
-	5)         If \textit{E} is not yet flattened:
-	6)             Flatten(\textit{E})
-	7)         \textit{R}.append(\textit{E})
-	8)         \textit{R}.append(\textit{E}.operator)
-	9)     \textit{T}.add(\textit{R})
-	10) Build a string S out of elements t$_{\text{1}}$, t$_{\text{2}}$, ... t$_{\text{n}}$ of \textit{T}:
-	11)     (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
-	12) Return \textit{S}
-\end{antlr}
+  Define \textit{T} as a list of empty string
+  For each alternative \textit{A} of the rule \textit{R}:
+    Define \textit{R} = \antlrap\antlrap
+    For each element \textit{E} of \textit{A}:
+	  If \textit{E} is not yet flattened:
+	    Flatten(\textit{E})
+      \textit{R}.append(\textit{E})
+      \textit{R}.append(\textit{E}.operator)
+    \textit{T}.add(\textit{R})
+  Build a string S from elements of \textit{T}:
+    (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
+  Return \textit{S}
+\end{alltt}
+\end{framed}
 \caption{Flatten algorithm}
 \label{fig:algflatten}
-\end{figure*}
+\end{figure}
 
@@
 Names of the nodes for alternatives are composed from the name of the rule and a number indicating the order of appearance in the grammar.
-For example, considering the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlgrammarelement}
+For example, consider the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlelementrule}.
+Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.
 
@@
 \centering
-\begin{antlr}
-	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-             \textcolor{gray}{|   \antlrap<\antlrap Name attribute* \antlrap/>\antlrap}
-             \textcolor{gray}{;}
-\end{antlr}
-\caption{Fragment of the SimpleXML grammar}
-\label{fig:xmlgrammarelement}
+\begin{framed}
+\begin{alltt}
+	\antlrparserrule{element} : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrap<\antlrap Name attrib* \antlrap/>\antlrap ;
+\end{alltt}
+\end{framed}
+\caption{Element rule from the SimpleXML grammar}
+\label{fig:xmlelementrule}
 \end{figure*}
-
-Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.
 
@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
 \centering
-\begin{antlr}
+\begin{framed}
+\begin{alltt}
 	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
@@
            ;
-\end{antlr}
+\end{alltt}
+\end{framed}
 \caption{Content rule}
 \label{fig:contentrule}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
 \centering
-\begin{mps}
+\begin{framed}
+\begin{alltt}
 	\mpsinterface{IContent}   :   \mpsconcept{Content{\_}1}
@@
            |   \mpsconcept{Content{\_}4}
-\end{mps}
+\end{alltt}
+\end{framed}
 \caption{IContent interface}
 \label{fig:icontentitf}
-\end{figure*}
+\end{figure}
 
 References to other parser rules are captured by links to children (interfaces or object nodes) in the AST nodes.
-For the \antlrparserrule{element} rule that is referencing the \antlrparserrule{content} rule, defined as in Figure~\ref{fig:elementrule}, we get the AST node representing the first alternative that is shown in Figure~\ref{fig:element_node_full}.
+For the \antlrparserrule{element} rule that is referencing the \antlrparserrule{content} rule, defined as in Figure~\ref{fig:xmlelementrule}, we get the AST node representing the first alternative that is shown in Figure~\ref{fig:element_node_full}.
 There is one \mpsinterface{IElement} interface node that is implemented by two object nodes \mpsconcept{Element{\_}1}, \mpsconcept{Element{\_}2}.
 The node \mpsconcept{Element{\_}1} has two child links pointing to \mpsinterface{IAttribute} and \mpsinterface{IContent}.
-
-\begin{figure*}[ht]
-\centering
-\begin{antlr}
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-\end{antlr}
-\caption{Element rule}
-\label{fig:elementrule}
-\end{figure*}
 
@@
 \centering
-\begin{antlr}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
-
-	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
-           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
-           ;
-
+\begin{framed}
+\begin{alltt}
+	\antlrparserrule{content} : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
+	\antlrparserrule{element} : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{/>} ;
 	\antlrparserrule{comment} : \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-\end{antlr}
+\end{alltt}
+\end{framed}
 \caption{Content parser rule with children}
@@
 \centering
-\begin{antlr}
+\begin{framed}
+\begin{alltt}
 	FindPathsToEndNodes($R$):
-	1)  Define $L$ as an empty list of list of nodes
-	2)  Return FindPathsToEndNodes($R$, $L$)
+	  Define $L$ as an empty list of list of nodes
+	  Return FindPathsToEndNodes($R$, $L$)
 
 	FindPathsToEndNodes($R$, $L$):
-	3)  Define $Q$ as list of list of nodes
-	4)  For each alternative $A$ of rule $R$:
-	5)      $L1$ = Clone($L$)
-	6)      If $A$ is a parser rule with only one element $E$:
-	7)          Let $I$ be interface/concept representing rule $E$
-	8)          $L1$.Add($I$)
-	9)          $P$ = FindPathsToEndNodes($E$, $L1$)
-	10)         $Q$ = Merge($Q$, $P$)
-	11)     Else
-	12)         $L1$.Add($R$)
-	13)         Let $P$ be concept representing $A$
-	14)         $L1$.Add($P$)
-	15)         $Q$.Add($L1$)
-	16) Return $Q$
-\end{antlr}
+	  Define $Q$ as list of list of nodes
+	  For each alternative $A$ of rule $R$:
+	    $L1$ = Clone($L$)
+	    If $A$ is a parser rule with only one element $E$:
+	      Let $I$ be interface/concept representing rule $E$
+	      $L1$.Add($I$)
+	      $P$ = FindPathsToEndNodes($E$, $L1$)
+	      $Q$ = Merge($Q$, $P$)
+	    Else
+	      $L1$.Add($R$)
+	      Let $P$ be concept representing $A$
+	      $L1$.Add($P$)
+	      $Q$.Add($L1$)
+	  Return $Q$
+\end{alltt}
+\end{framed}
 \caption{Algorithm for finding end rules}
@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
 \centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
+\begin{framed}
+\begin{alltt}
+	\antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
 	\antlrlexerrule{STRING} : \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
-	            |   \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'}
-	            ;
-\end{antlr}
+	       | \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'} ;
+\end{alltt}
+\end{framed}
 \caption{Definition of an XML attribute}
 \label{fig:xmlattribute}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
 \centering
-\begin{antlr}
-	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
-	            |   \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'}
-	            ;
+\begin{framed}
+\begin{alltt}
+	\antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
+	       | \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'} ;
 	\antlrlexerrule{TEXT1} : \antlrregex{~["]*} ;
 	\antlrlexerrule{TEXT2} : \antlrregex{~[']*} ;
-\end{antlr}
+\end{alltt}
+\end{framed}
 \caption{Adjusted grammar of SimpleXML}
 \label{fig:xmladjustgrammar}
-\end{figure*}
+\end{figure}
 
@@
 
-As the second example, we will show the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
+As the second example, we will use the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
 Every statement in JavaScript needs to be either followed by a semicolon, newline, file end or end of the block --- see the Figure~\ref{fig:javascriptstmt}.
@@
 \centering
-\begin{antlr}
-	\antlrparserrule{eos}            : \antlrlexerrule{SemiColon}
-	               | \antlrlexerrule{EOF}
-	               | {lineTerminatorAhead()}?
-	               | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}?
-	               ;
+\begin{framed}
+\begin{alltt}
+	\antlrparserrule{eos} : \antlrlexerrule{SemiColon} | \antlrlexerrule{EOF} | {lineTerminatorAhead()}? 
+	    | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}? ;
 	\textcolor{gray}{// Example reference of the eos rule}
-	\antlrparserrule{breakStatement} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos}
-	               ;
-\end{antlr}
+	\antlrparserrule{breakStmt} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos} ;
+\end{alltt}
+\end{framed}
 \caption{Statements in JavaScript}
@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
 \centering
-\begin{antlr}
+\begin{framed}
+\begin{alltt}
 	\antlrparserrule{eos} : \antlrliteral{;} ;
-\end{antlr}
+\end{alltt}
+\end{framed}
 \caption{eos parser rule}
 \label{fig:eosrule}
-\end{figure*}
+\end{figure}
 

### 2016-10-06 chapters/background.tex
writing introduction
finished introduction
tweak code listing
working on background

@@
 \section{Background and Related Work}
+\label{sect:BACKGRELWORK}
 
-We will briefly introduce fundamental basics of the MPS environment.
-We will analyze existing related projects that are trying to accomplish similar goals, and then we will weigh advantages and disadvantages of these approaches.
+In this section, first we introduce basic features of the MPS platform, and then we discuss other existing projects that aim to add support for general purpose languages into MPS.
 
 \subsection{JetBrains MPS}
+\label{sect:MPS}
 
-\JB{Odstavec upraven} JetBrains MPS is a complete language workbench - a development environment that allows developers to define their own languages and then use them to develop code.
-The code can then be transformed into a target language, typically a GPL, such as Java or C.
+JetBrains MPS is a complete language workbench --- an integrated development environment that allows developers to create their own languages and use them to write code.
+The code can then be transformed into a target language, typically a GPL, such as Java or C, and eventually compiled into executable programs.
 
-\JB{Navrhuji odstranit}
-\PV{Souhlasim}
-We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
-Focus on basic/fundamental and relevant features. In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.
+As we already indicated, MPS differs from typical IDEs in one important aspect --- the projectional editor.
+The developer does not work with the textual representation of the source code directly, but rather with its AST that is the model of the code.
+Basically, when using the MPS projectional editor, programs are created by assembling the tree (AST) out of predefined building blocks of selected languages.
+The definition of a language in MPS dictates, where in the AST can certain elements be placed and how they can be nested inside other elements.
+On the other hand, in traditional IDEs it is the parser that constructs an AST of a program using the language's elements.
 
-\JB{odstavec upraven} MPS differs from typical IDEs in one important aspect - the projectional editor.
-The developer does not work with the textual representation of the source code directly, but rather with its AST (abstract syntax tree) that is the model of the code.
-When users are creating programs using the MPS projectional editor, they are basically assembling the tree (AST) out of predefined building blocks of the imported languages.
-The definition of the language dictates, which where in the AST can certain elements be placed or how they can be nested inside other elements.
-In traditional IDEs it is the parser that constructs an AST using the language's elements.
+The building blocks of MPS models are called nodes.
+Code of any program in MPS is built from nodes, which represent instances of concepts defined in one of the languages that the program declares to be using.
+In the context of MPS, a concept is, in fact, a building block of a language definition.
+We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.
 
-\JB{Odstavec upraven} The building blocks of MPS models are called nodes.
-Code of any program in MPS is built from nodes and these nodes represent instances of elements defined in one of the languages that the program declares to be using.
-These language elements are called concepts.
-In brief, nodes in code are instances of concepts defined in the language.
+One of the key advantages of projectional editing stems from the separation of abstract and concrete syntax.
+While AST provides a complete and precise representation of the code, the way it is displayed on the screen and the way the user interacts with it are unconstrained.
+The editor can take any visual form and shape.
+The language author can define multiple alternative visualizations and let the developer to choose one that fits best the task at hand --- for example, editing, debugging, reviewing, resolving merge conflicts, etc.
+In particular, the visual representations are not bound to be just textual at all.
+For example, blocks corresponding to branches of an \code{if-then-else} statement can be aligned next to each other and displayed with different background colors.
 
-\JB{Vyhody/nevyhody diskutujeme v kapitole 1. Navrhuji nasledujici odstavec zrusit nebo text z kapitoly 1. presunout sem}
-\PV{Spis bych zrusil tady, kapitola 1 mi prijde vhodnejsi pro zmineni motivace}
-Keeping the model of the code in the AST form has several advantages.
-One of them is that MPS then only allows composing the program strictly following the syntax of the language, which results in an inability to actually write syntactically invalid code.
-The code is generated out of the always-valid AST beyond user's reach.
+The definition of an MPS concept (language element) consists of several aspects, where each aspect codifies a different part of the AST nodes' behavior.
+The essential aspects are the following: \emph{Structure} (abstract syntax), \emph{Editor} (concrete syntax) and \emph{TextGen}.
+Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
+If, instead of generating text directly, programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
+Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we describe these three below in more detail, and neglect other aspects such as type system and data-flow.
 
-\JB{Odstavec upraven} One of key advantages of projectional editor stems from separation of abstract and concrete syntax.
-While the AST provides an exhaustive representation of the code, the way it is displayed on the screen and the way the user interacts with it are unconstrained.
-The editor can take any visual form and shape.
-The language author can define multiple visualizations and let the developer to choose one that fits best the task at hands - editing, debugging, reviewing, resolving merge conflicts, etc.
+Languages are built from the concepts using techniques known from object-oriented programming --- containment, inheritance, interfaces, and so on.
+Therefore, a definition of a whole language in MPS typically has an object-oriented and hierarchical nature.
 
-For a better understanding, we will give a~small example. Let's assume we have a language with typical if/else structure similar to the one in Java:
-
-\begin{center}
-	\begin{minipage}{.38\textwidth}
-		\begin{alltt}
-			if (\textit{condition})
-			    \textit{then statement block}
-			else
-			    \textit{else statement block}
-		\end{alltt}
-	\end{minipage}
-\end{center}
-
-Its abstract syntax tree representation will look something like shown on Figure~\ref{fig:if_ast}.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[scale=0.6]{./images/if_statement_ast.png}
-  \caption{"If statement" abstract syntax tree}
-  \label{fig:if_ast}
-\end{figure}
-
-The underlying child nodes have their own children and so on.
-The type of nodes can be restricted by the language structure definition.
-For example, we can restrict the condition in the if/else statement to be an expression, the statement block to be a list of statements, etc.
-
-\JB{odstavec upraven} Alternative visual representations are possible and they are not bound to be just the textual representation at all.
-For example, blocks for both of the branches can be aligned next to each other and displayed with differend background color, like shown in Figure~\ref{fig:if_editor}.
-
-\begin{figure}[ht]
-	\centering
-	\hspace{-4mm}
-	\includegraphics[scale=0.75]{./images/if_statement_editor.png}
-	\caption{"If statement" projectional editor example}
-	\label{fig:if_editor}
-\end{figure}
-
-\JB{Odstavec mirne upraven}The definition of a language element (the concept) consists of several aspects and each aspect codifies a different part of the nodes' behavior.
-The essential aspects are the following: structure (abstract syntax), editor (concrete syntax - how the code is projected and edited) and textgen (how to transform the AST into textual source code).
-If, instead of generating text directly, the language is supposed to be generated into another language that is available in MPS, the generator aspect must be used to specify the model-transformation rules.
-The structure, editor and textgen aspects are most relevant for the contribution of this paper, so we describe them here (and neglect/ignore the other aspects such as type system and data-flow).
-
-\todo{Highlight the object-oriented hierarchical nature of language definitions in MPS. The environment restricts and guides the developer so that he makes less mistakes.}
-
-Some concepts and design patterns known from object oriented programming can be used in language definition for MPS --- inheritance between concepts, interfaces, delegation and so on.
-Most importantly, nodes can have child nodes and that is how the AST itself is created.
-
-\JB{Zmineno vyse, navrhuji odstranit} \todo{Zduraznit ze AST node (object) ma nekolik aspectu: structure, editor, textgen.
-Definice MPS language je organizovana takto --- pres AST nodes, ktere maji nekolik aspectu (ruzne pohledy).}
+% TODO continue from here
 
@@
 In Figure~\ref{fig:if_statement_structure}, you can see the structure aspect definition for the if statement.
+
+The language structure definition restricts the type of nodes (concepts) that can appear at a particular place in AST.
+For example, we can restrict the condition in the if-then-else statement to be an expression, the then-block to be a list of statements, etc.
 
@@
 
+% TODO mozna zmenit nazev teto sekce
 \subsection{Similar Projects}
+
+We are aware of several projects that provide certain support for GPLs in the context of the MPS platform.
+We will analyze existing related projects that are trying to accomplish similar goals, and then we will weigh advantages and disadvantages of these approaches.
 

### 2016-10-07 chapters/background.tex
structure

@@
 
-% TODO continue from here
+\paragraph{Structure.}
+The fundamental aspect of any MPS language is Structure.
+It must be created first for each intended element (concept) of the language.
+Structure specifies core attributes of an MPS concept such as the name, inheritance relationships, possible child concepts (including their types and cardinalities), the implemented interfaces, and references to other AST nodes.
+% other properties (fields of any type that can hold values), etc.
+Figure~\ref{fig:if_statement_structure} shows definition of the Structure aspect for the \code{if} statement.
 
-\paragraph{Structure.}
-\JB{Odstavec mirne upraven} The fundamental aspect of any MPS language is structure.
-It must be created first for each type of the intended AST node.
-Structure specifies core attributes such as name of the concept, inheritance relationships, possible child concepts (their types and cardinalities), the implemented interfaces, references to other nodes, other properties (fields of any type that can hold values), etc.
-In Figure~\ref{fig:if_statement_structure}, you can see the structure aspect definition for the if statement.
-
-The language structure definition restricts the type of nodes (concepts) that can appear at a particular place in AST.
-For example, we can restrict the condition in the if-then-else statement to be an expression, the then-block to be a list of statements, etc.
+The definition of Structure restricts the type of AST nodes that can appear at a particular place in the tree.
+For example, one can restrict the condition in the \code{if-then-else} statement to be a boolean expression, and the \code{then}-block to be a list of statements.
 
@@
 \centering
-	\includegraphics[scale=0.55]{./images/if_statement_structure.png}
-	\caption{"If statement" structure aspect definition}
+\begin{alltt}
+\small
+\mpskeyword{concept} IfStatement \mpskeyword{extends} Statement
+        \mpskeyword{implements} IContainsStatementList
+                   IDontSubstituteByDefault
+                   IConditional
+
+  \mpskeyword{instance can be root:} false
+  \mpskeyword{alias:} \mpsalias{if}
+  \mpskeyword{short description:} \mpsplaceholder{<no short description>}
+
+  \mpskeyword{properties:}
+  \mpsproperty{forceOneLine}   : boolean
+  \mpsproperty{forceMultiLine} : boolean
+
+  \mpskeyword{children:}
+  \mpsproperty{condition}        : Expression[\mpscardinality{1}]
+  \mpsproperty{ifFalseStatement} : Statement[\mpscardinality{0..1}]
+  \mpsproperty{ifTrue}           : StatementList[\mpscardinality{1}]
+  \mpsproperty{elsifClauses}     : ElsifClause[\mpscardinality{0..n}]
+
+  \mpskeyword{references:}
+  \mpsplaceholder{<< ... >>}
+\end{alltt}
+\caption{Structure aspect of the \code{if} statement}
 \label{fig:if_statement_structure}
 \end{figure}
+
+% TODO continue from here
 

### 2016-10-10 chapters/background.tex
editor and textgen

@@
 % other properties (fields of any type that can hold values), etc.
-Figure~\ref{fig:if_statement_structure} shows definition of the Structure aspect for the \code{if} statement.
+Figure~\ref{fig:if_statement_structure} shows definition of the Structure aspect for the \code{if-else} statement.
+\PP{Mozna trochu vic popsat obsah figure --- co znamenaji jednotlive prvky definice.}
 
@@
 \small
-\mpskeyword{concept} IfStatement \mpskeyword{extends} Statement
-        \mpskeyword{implements} IContainsStatementList
+\mpsstkeyword{concept} IfStatement \mpsstkeyword{extends} Statement
+        \mpsstkeyword{implements} IContainsStatementList
                    IDontSubstituteByDefault
@@
 
-  \mpskeyword{instance can be root:} false
-  \mpskeyword{alias:} \mpsalias{if}
-  \mpskeyword{short description:} \mpsplaceholder{<no short description>}
+  \mpsstkeyword{instance can be root:} false
+  \mpsstkeyword{alias:} \mpsstalias{if}
+  \mpsstkeyword{short description:} \mpsstplaceholder{<no short description>}
 
-  \mpskeyword{properties:}
-  \mpsproperty{forceOneLine}   : boolean
-  \mpsproperty{forceMultiLine} : boolean
+  \mpsstkeyword{properties:}
+  \mpsstproperty{forceOneLine}   : boolean
+  \mpsstproperty{forceMultiLine} : boolean
 
-  \mpskeyword{children:}
-  \mpsproperty{condition}        : Expression[\mpscardinality{1}]
-  \mpsproperty{ifFalseStatement} : Statement[\mpscardinality{0..1}]
-  \mpsproperty{ifTrue}           : StatementList[\mpscardinality{1}]
-  \mpsproperty{elsifClauses}     : ElsifClause[\mpscardinality{0..n}]
+  \mpsstkeyword{children:}
+  \mpsstproperty{condition}        : Expression[\mpsstcardinality{1}]
+  \mpsstproperty{ifFalseStatement} : Statement[\mpsstcardinality{0..1}]
+  \mpsstproperty{ifTrue}           : StatementList[\mpsstcardinality{1}]
+  \mpsstproperty{elsifClauses}     : ElsifClause[\mpsstcardinality{0..n}]
 
-  \mpskeyword{references:}
-  \mpsplaceholder{<< ... >>}
+  \mpsstkeyword{references:}
+  \mpsstplaceholder{<< ... >>}
 \end{alltt}
-\caption{Structure aspect of the \code{if} statement}
+\caption{Structure aspect of the \code{if-else} statement}
 \label{fig:if_statement_structure}
@@
 
-% TODO continue from here
+\paragraph{Editor.}
+The Editor aspect is where the language designer specifies what the projectional editor representation of a code fragment (an AST) looks like on the screen and how the user interacts with the code.
+JetBrains have developed a cellular system that enables placing properties and children of a node (concept) into different cells.
+The author usually incorporates all of the node's children, references, and properties inside the representation, so that the future users of the language can insert all values that the node expects.
+Additionally, cells of the editor can be styled using a language similar to CSS.
+The supported visual characteristics include text color and indentation.
 
-\paragraph{Editor.}
-\JB{Odstavec mirne upraven}The editor aspect is where the user defines what the projectional editor representation of a code fragment (an AST node) looks like on the screen and how the user interacts with the code.
-JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells.
-The author usually incorporates all concept's children, references, and properties inside the representation so that the future users of the language can insert the values that the node expects.
-Additionally, all visual cells of the editor can be styled using a~language similar to CSS.
+Figure~\ref{fig:if_editor_definition} shows an example of what the definition of the Editor aspect for the \code{if-else} statement might look like.
+While here we indicate positions of cell borders on each line by spaces (just for the purpose of illustration), MPS GUI actually uses a graphically much more appealing way of displaying the Editor aspects, which involves vertical and horizontal lines of different colors and also background colors other than white for some cells.
+\PP{Vysvetlit symboly "[-", "-]" \%, a dalsi ktere nejsou hned zrejme.}
 
@@
 \centering
-	\includegraphics[scale=0.5]{./images/if_statement_editor_definition.png}
-	\caption{"If statement" editor aspect definition}
+\begin{alltt}
+\small
+\mpsedannotation{<default>} \mpsedkeyword{editor for concept} \mpsedtarget{IfStatement}
+  \mpsedkeyword{node cell layout:}
+    [-
+      \mpsedkeyword{if} \textbf{(} \% conditions \% \textbf{)} [-
+      \textbf{\{}
+      [- \% ifTrue % -]
+      \textbf{\}}
+    -]
+    ?[- \mpsedkeyword{else} \% ifFalseStatement \% -]
+    -]
+\end{alltt}
+\caption{Editor aspect for the \code{if-else} statement}
 \label{fig:if_editor_definition}
@@
 
-\JB{Nasledujici odstavec navrhuji odstranit}
-\todo{From thesis.
-
-The editor is a module dedicated/responsible for the definition of appearance of an AST node (concept).
-JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells.
-These cells then can be styled to user's liking.
-There are many different types of cells, each behaving a little bit different towards its contents.
-For example, MPS supports various horizontal and vertical lists, etc.
-Extra cells can be added on top --- cells which just specify layout adjustments such as indentation and text color.
-In Figure~\ref{fig:if_editor_definition}, you can see what a real editor definition might look like for the if statement of the Java language (as defined in MPS).
-}
-
-\paragraph{BaseLanguage}
-\todo{Introduce the base language here.}
-\JB{Odstavec prepsan}
-In order to implement MPS itself and also to support the basic set of language-definition DSLs, BaseLanguage was developed in the early days of MPS.
-BaseLanguage is a clone of Java implemented using the MPS constructs.
-Although BaseLanguage is syntactically almost identical to Java, it is edited in a projectional editor, just like all languages in MPS.
-The DSLs that language authors use in order to define languages are generated into BaseLanguage.
-Similarly, all customer DSLs that are meant to be generated into Java choose BaseLanguage as their generation target and the conversion to textual Java will be handled by BaseLanguage without any further effort, since BaseLanguage has a TextGen aspect defined, which translates BaseLanguage code into textual Java sources.
+\paragraph{BaseLanguage.}
+Another important feature of MPS that we need to describe in more detail is \emph{BaseLanguage}.
+It is a clone of Java implemented using the MPS constructs.
+BaseLanguage was developed in the early days of MPS in order to implement MPS itself and also to support the basic set of language-definition DSLs,
+Although BaseLanguage is syntactically almost identical to Java, it is edited in a projectional editor, just like all the languages in MPS.
+The language-definition DSLs, used to define custom languages by their authors, are generated into the BaseLanguage.
+Similarly, all the custom DSLs that are meant to be generated into Java choose BaseLanguage as their generation target.
+The conversion to textual Java is handled by BaseLanguage without any further manual effort, since BaseLanguage has a TextGen aspect defined, which translates BaseLanguage code into textual Java sources.
 
 \paragraph{TextGen.}
-\JB{Odstavec upraven} The TextGen aspect defines how a certain AST node will be translated into text.
-The TextGen aspect is typically only needed for the bottom-line (base) languages.
-DSLs, on the other hand, need to define rules for model-to-model conversion (a generator), since these are rarely converted to text directly.
-After TextGen has generated textual sources from an AST, a compiler for the particular GPL gets invoke in order to compile the generate sources into binary code.
-The TextGen definition follows a very straightforward pattern - each node outputs its textual representation into a textual buffer, calling TextGen of its children nodes at the right moments.
-MPS then calls this method for the root AST nodes of the given program.
-TextGen aspect has to be defined (especially its more complex functionality) using the BaseLanguage.
-Again, we included an example of the if statement (Figure~\ref{fig:if_statement_textgen}).
+The TextGen aspect specifies how a given AST node will be translated into plain text representation.
+It is typically needed only for the bottom-line base languages.
+DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since these are rarely converted to text directly.
+After TextGen has generated textual sources from an AST, a compiler for the particular GPL is invoked to compile the textual sources into binary code.
+The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, calling TextGen of its children nodes at the right moments.
+MPS calls the corresponding method on the root AST nodes of the given program.
+
+TextGen aspect for each AST node (concept of the language) has to be defined using the BaseLanguage.
+Again, we include an example for the \code{if-else} statement (Figure~\ref{fig:if_statement_textgen}).
 
@@
 \centering
-	\includegraphics[scale=0.6]{./images/if_statement_textgen.png}
-	\caption{"If statement" TextGen aspect definition}
+\begin{alltt}
+\small
+\mpstgkeyword{text gen component for concept} \mpstgtarget{IfStatement} \{
+  \mpstgparam{(context, buffer, node)->void} \{
+    \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n;}}
+    \mpstgaction{indent buffer;}
+    \mpstgaction{append} \{\mpstgliteral{if (}\} \$\{\mpstgparam{node}.\mpstgnodeprop{condition}\} \{\mpstgliteral{) \{}\};
+    \mpstgkeyword{with indent} \{
+      \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{ifTrue}\};
+    \}
+    \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n}} \{\mpstgliteral{\}}\} \$list\{\mpstgparam{node}.\mpstgnodeprop{elsifClauses}\};
+    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{ifFalseStatement}.\textbf{isNotNull}) \{
+      \mpstgaction{append} \{ \mpstgliteral{else}\} \$\{\mpstgparam{node}.\mpstgnodeprop{isFalseStatement}\};
+    \}
+  \}
+\}
+\end{alltt}
+\caption{TextGen aspect definition for the \code{if-else} statement}
 \label{fig:if_statement_textgen}
@@
 
-\JB{Navrhuji odebrat}
-\todo{From thesis.
-
-After the program's AST is created using the projectional editor, there can be rules defined on how to generate code in plain text out of that tree.
-This module is called "text gen".
-These defined generators can target any language and platform.
-
-\paragraph{Other aspects.}
-MPS supports also other aspects, such as type checking, model transformations, static and data-flow code analysis, refactorings, language migrations, etc., but these are not important for our contribution.
-Moreover, the information needed for their automated generation is not available/contained inside a grammar file.
-
-}
+% TODO continue from here
 

### 2016-10-12 chapters/background.tex
minor changes

@@
 Figure~\ref{fig:if_statement_structure} shows definition of the Structure aspect for the \code{if-else} statement.
-\PP{Mozna trochu vic popsat obsah figure --- co znamenaji jednotlive prvky definice.}
 
@@
 While here we indicate positions of cell borders on each line by spaces (just for the purpose of illustration), MPS GUI actually uses a graphically much more appealing way of displaying the Editor aspects, which involves vertical and horizontal lines of different colors and also background colors other than white for some cells.
-\PP{Vysvetlit symboly "[-", "-]" \%, a dalsi ktere nejsou hned zrejme.}
+The symbols \verb|[-| and \verb|-]| represent layout information (cells), such as indentation.
 
@@
 This approach also has some downsides.
-Because MPS doesn't offer any tools that would allow easy transformation from the MPS grammar to the MPS language, the second step of the import process hasn't been implemented yet and will be probably very problematic.
+Because MPS doesn't offer any tools that would allow easy transformation from the MPS grammar \PP{To je konkretni gramatika zadefinovana pomoci MPS jazyka pro ANTLR} to the MPS language, the second step of the import process hasn't been implemented yet and will be probably very problematic.
 Another downside might be the need for a double transformation implementation -- first one being the import of the text grammar to the MPS grammar definition.

### 2016-10-13 chapters/background.tex
writing about related projects

@@
 \paragraph{BaseLanguage.}
-Another important feature of MPS that we need to describe in more detail is \emph{BaseLanguage}.
+Another important feature of MPS that we need to describe in more detail is \emph{BaseLanguage}~\cite{ref:BaseLanguage}.
 It is a clone of Java implemented using the MPS constructs.
@@
 MPS calls the corresponding method on the root AST nodes of the given program.
-
-TextGen aspect for each AST node (concept of the language) has to be defined using the BaseLanguage.
-Again, we include an example for the \code{if-else} statement (Figure~\ref{fig:if_statement_textgen}).
 
@@
 
+TextGen aspect for each AST node (concept of the language) has to be defined using the BaseLanguage.
+Again, we include an example for the \code{if-else} statement (Figure~\ref{fig:if_statement_textgen}).
+
+\subsection{Related Projects}
+
+The BaseLanguage~\cite{ref:BaseLanguage} is an almost full port of the Java language, extended with MPS-specific features.
+It was created manually by JetBrains, and it is still undergoing changes as Java itself is evolving.
+
+The C language has also been manually tailored for MPS within the mbeddr project~\cite{ref:MBEDDR}.
+
+Besides that, we are aware of three other projects that provide certain support for GPLs in the context of MPS --- namely PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
+We describe their main features and limitations in this section.
+
 % TODO continue from here
-
-% TODO mozna zmenit nazev teto sekce
-\subsection{Similar Projects}
-
-We are aware of several projects that provide certain support for GPLs in the context of the MPS platform.
-We will analyze existing related projects that are trying to accomplish similar goals, and then we will weigh advantages and disadvantages of these approaches.
-
-Currently, there exists an almost full port of the Java language called BaseLanguage~\cite{ref:BaseLanguage} extended with some MPS specific features.
-It was imported manually by JetBrains and it is still undergoing changes as Java itself is evolving.
-
-Within the mbeddr project~\cite{ref:mbeddr}, the C language is also manually tailored for MPS.
-
-Now/here we describe three similar projects (with similar goals) that we are aware of.
-We focus on (consider) these three projects: PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
 

### 2016-10-13 paper.tex
finished introduction
tweak code listing
structure
editor and textgen
writing about related projects

@@
 \usepackage{balance}
-\usepackage{adjustbox}
+\usepackage{framed}
 \usepackage{alltt}
 \usepackage{textcomp}
-\usepackage[usenames]{xcolor}
+\usepackage[dvipsnames]{xcolor}
 
@@
 
-% colors for the ANTLR environment
-\definecolor{antlrbgcolor}{rgb}{1,1,1}
-\definecolor{antlrparserrulecolor}{RGB}{11,83,148}
-\definecolor{antlrlexerrulecolor}{RGB}{191,144,0}
-\definecolor{antlrliteralcolor}{RGB}{133,32,12}
-\definecolor{antlrregexcolor}{RGB}{56,118,29}
-
-% colors for the MPS environment
-\definecolor{mpsbgcolor}{rgb}{1,1,1}
-\definecolor{mpsconceptcolor}{RGB}{42,140,0}
-\definecolor{mpsinterfacecolor}{RGB}{170,0,229}
-
-% coloring commands
+% coloring and font style commands
+  % ANTLR
 \newcommand{\antlrap}{\textquotesingle}
-\newcommand{\antlrparserrule}[1]{\textcolor{antlrparserrulecolor}{#1}}
-\newcommand{\antlrlexerrule}[1]{\textcolor{antlrlexerrulecolor}{#1}}
-\newcommand{\antlrliteralnoap}[1]{\textcolor{antlrliteralcolor}{#1}}
+\newcommand{\antlrparserrule}[1]{\textcolor{Blue}{#1}}
+\newcommand{\antlrlexerrule}[1]{\textcolor{RedOrange}{#1}}
+\newcommand{\antlrliteralnoap}[1]{\textcolor{Maroon}{#1}}
 \newcommand{\antlrliteral}[1]{\antlrap\antlrliteralnoap{#1}\antlrap}
-\newcommand{\antlrregex}[1]{\textcolor{antlrregexcolor}{#1}}
-\newcommand{\mpsconcept}[1]{\textcolor{mpsconceptcolor}{#1}}
-\newcommand{\mpsinterface}[1]{\textcolor{mpsinterfacecolor}{#1}}
-
-% environment for the ANTLR code (grammars)
-\newenvironment{antlr}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=antlrbgcolor,frame,minipage=485pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
-
-% environment for the MPS code
-\newenvironment{mps}
-{
-	\vspace{3mm}
-	\par
-	\noindent
-	\adjustbox{margin=3mm,bgcolor=mpsbgcolor,frame,minipage=394pt,inner=\linewidth}
-	\bgroup
-	\alltt
-}
-{
-	\endalltt
-	\egroup
-	\\
-	\vspace{2mm}
-}
+\newcommand{\antlrregex}[1]{\textcolor{PineGreen}{#1}}
+  % MPS in general
+\newcommand{\mpsconcept}[1]{\textcolor{OliveGreen}{#1}}
+\newcommand{\mpsinterface}[1]{\textcolor{Plum}{#1}}
+  % MPS structure
+\newcommand{\mpsstkeyword}[1]{\textcolor{Blue}{\textbf{#1}}}
+\newcommand{\mpsstalias}[1]{\textcolor{Green}{\textbf{#1}}}
+\newcommand{\mpsstproperty}[1]{\textcolor{Fuchsia}{\textbf{#1}}}
+\newcommand{\mpsstplaceholder}[1]{\textcolor{Gray}{\textbf{#1}}}
+\newcommand{\mpsstcardinality}[1]{\textcolor{RoyalBlue}{#1}}
+  % MPS editor
+\newcommand{\mpsedkeyword}[1]{\textcolor{Black}{\textbf{#1}}}
+\newcommand{\mpsedtarget}[1]{\textcolor{Fuchsia}{#1}}
+\newcommand{\mpsedannotation}[1]{\textcolor{Gray}{\textbf{#1}}}
+  % MPS textgen
+\newcommand{\mpstgkeyword}[1]{\textcolor{Blue}{\textbf{#1}}}
+\newcommand{\mpstgtarget}[1]{\textcolor{Fuchsia}{#1}}
+\newcommand{\mpstgparam}[1]{\textcolor{Violet}{#1}}
+\newcommand{\mpstgaction}[1]{\textcolor{Fuchsia}{\textbf{#1}}}
+\newcommand{\mpstgliteral}[1]{\textcolor{Green}{\textbf{#1}}}
+\newcommand{\mpstgnodeprop}[1]{\textcolor{darkgray}{#1}}
 
@@
 
-\title{\todo{INGRID: Creating Languages in MPS from ANTLR Grammars}}
+\title{INGRID: Creating Languages in MPS from ANTLR Grammars}
 
@@
 
+\bibitem{ref:BaseLanguage} MPS BaseLanguage, \url{https://confluence.jetbrains.com/display/MPSD34/Base+Language}
+
 % TODO opravit formatovani nasledujicich polozek a vhodne zaradit
-
-\bibitem{ref:mbeddr} mbeddr, \url{mbeddr.com/}
-
-\bibitem{ref:BaseLanguage} BaseLanguage, \url{confluence.jetbrains.com/display/MPSD34/Base+Language}
 
@@
 \bibitem{ref:ANTLR2MPS} ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR{\_}MPS}
+
+\bibitem{ref:mpsmetabnf} mps-metabnf \url{https://github.com/DSLFoundry/mps-metabnf}
 

### 2016-10-13 chapters/introduction.tex
working on background
Minor typo fixes in chapter 1

@@
 
-Projectional editing, and in particular the absence of parsers, brings several benefits.
+Projectional editing, keeping the code in the AST form, and in particular the absence of parsers, brings along several benefits.
 \begin{itemize}
 	\item The developers practically cannot enter syntactically invalid code, because a projectional editor controls the interaction between the user and the code.
+An AST created by the editor can represent only a syntactically valid code.
 	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together or one can extend another, while parsers seriously limit language modularity for traditional parsed languages.
 	\item The languages may support contextual or non-parseable notations, such as tables, diagrams and positional syntax.
-	\item Since the projection is detached from the physical representation of code (AST), authors of languages can define multiple notations and allow the developers to switch between different representations of the code.
+	\item Since the projection is detached from the physical representation of code (AST), authors of languages can define multiple notations and allow the developers to switch between different representations of the code on the screen.
 \end{itemize}
@@
 
-On the other hand, usage of projectional editors introduces some new problems.
+On the other hand, the usage of projectional editors introduces some new problems.
 Before a language can be used inside MPS, it has to be defined in a special way through the MPS infrastructure, so that the IDE can understand the language and work with it.
@@
 Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges are general. 
-They apply to the relation between two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
+They apply to the relation between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
 Specifically, the description of a language in the form of a grammar does not hold any information about the code layout, and the grammar typically contains many rules that do not directly correspond to AST nodes and programming language constructs.

### 2016-10-13 chapters/language_definition.tex
Fixes & comments for section 3

@@
 There is a challenge/issue that we have discovered during our work on this project (in the process of our own implementation), and that is related to construction of the structure aspect.
+
 Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
 However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
+\PV{Dalsi veta asi zbytecna}
 We describe/discuss these challenges in more detail (also using examples) and our solutions to the challenges inside the respective following subsections.
@@
 \todo{Pouzit jazyk SimpleXML na running exaple.}
+\PV{Dalsi veta asi prijde niz?}
 
@@
 	\textbf{fragment}
-	\antlrlexerrule{NameChar}  : \antlrlexerrule{NameStartChar} | \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
+	\antlrlexerrule{NameChar}  : \antlrlexerrule{NameStartChar} | \antlrliteral{-} | \antlrliteral{\_} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
 	\textbf{fragment}
@@
 \begin{center}
-	\texttt{([:a-zA-Z])(([:a-zA-Z])|(-)|({\_})|(.)|([0-9]))*}
+	\texttt{[:a-zA-Z]([:a-zA-Z]|\textbackslash-|{\_}|\textbackslash.|[0-9])*}
 \end{center}

### 2016-10-13 chapters/background.tex
Minor polish to section 2.1

@@
 The editor can take any visual form and shape.
-The language author can define multiple alternative visualizations and let the developer to choose one that fits best the task at hand --- for example, editing, debugging, reviewing, resolving merge conflicts, etc.
+The language author can define multiple alternative visualizations and let the developer choose one that fits best the task at hand --- for example, editing, debugging, reviewing, resolving merge conflicts, etc.
 In particular, the visual representations are not bound to be just textual at all.
@@
 While here we indicate positions of cell borders on each line by spaces (just for the purpose of illustration), MPS GUI actually uses a graphically much more appealing way of displaying the Editor aspects, which involves vertical and horizontal lines of different colors and also background colors other than white for some cells.
-The symbols \verb|[-| and \verb|-]| represent layout information (cells), such as indentation.
+The symbols \verb|[-| and \verb|-]| represent layout information (cells), which define rules for mutual positioning for the contained cells (vertical, horizontal, indentation).
 
@@
 After TextGen has generated textual sources from an AST, a compiler for the particular GPL is invoked to compile the textual sources into binary code.
-The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, calling TextGen of its children nodes at the right moments.
+The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes at the right moments.
 MPS calls the corresponding method on the root AST nodes of the given program.

### 2016-10-13 chapters/background.tex
text about pe4mps

@@
 While here we indicate positions of cell borders on each line by spaces (just for the purpose of illustration), MPS GUI actually uses a graphically much more appealing way of displaying the Editor aspects, which involves vertical and horizontal lines of different colors and also background colors other than white for some cells.
-The symbols \verb|[-| and \verb|-]| represent layout information (cells), which define rules for mutual positioning for the contained cells (vertical, horizontal, indentation).
+The symbols \verb|[-| and \verb|-]| represent layout information (cells), which define rules for mutual positioning of the contained cells (vertical, horizontal, indentation).
 
@@
 
+\paragraph{PE4MPS.}
+PE4MPS~\cite{ref:PE4MPS} is a project that addresses the lack of information about code layout in grammars by using so called PE grammars~\cite{ref:PE}, which is a new grammar notation proposed by the same author.
+The abbreviation PE stands for projectional editing.
+
+PE grammars extend the syntax of the ANTLR v4 notation by custom constructs that provide information about the possible layout of AST nodes.
+The current version, as of November 2016, supports just horizontal lists, vertical lists, and some indentation rules.
+However, even these few features make the already complex syntax of ANTLR v4 much more complicated.
+
+An MPS language is created by the PE4MPS tool in two steps.
+First, the PE parser is used to load an input PE file and build its intermediate representation, which has the form of a tree-like structure consisting of Java objects.
+Then, all the concepts (AST nodes) and their aspects are created inside MPS.
+The information about the code layout, extracted from the PE file, is used when generating the projectional editor for every AST node.
+
+Another limitation of PE4MPS is that it does not implement the full ANTLR syntax.
+Every grammar might require non-trivial adjustments before it can be processed using the PE4MPS tool.
+To compare, the goals behind our INGRID approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.
+
 % TODO continue from here
-
-\paragraph{PE4MPS.}
-PE4MPS~\cite{ref:PE4MPS} is a project by Marco Lombardo that is trying to solve the same problem as we do.
-It solves the lack of information about code layout in grammars by creating a new grammar notation called PE Grammars~\cite{ref:PE} (the PE abbreviation comes from projectional editing).
-It has tw components: PE parser and PE4MPS plugin for MPS.
-Their approach is to mimic an existing grammar notation called ANTLRv4~\cite{ANTLR4} and enrich its syntax with custom (their own) constructs.
-These extensions tell (guide, hint) the parser what the AST node layout should look like.
-Supported extensions (already implemented) include horizontal and vertical lists, and some indentation rules --- however, even these few features make the already not-so-simple syntax of ANTLR v4 much more complex.
-The parser then uses this information when generating the projectional editor for am AST node.
-Author of this approach/project described the PE syntax using the ANTLRv4 notation~\cite{ANTLR4reference} and then used the ANTLR toolset to automatically generate an ANTLRv4 parser for PE grammar files.
-The PE parser reads any PE file and stores the language structure found inside to a custom representation of Java objects.
-
-The PE4MPS plugin/project/tool is built on top of the PE parser.
-This plugin uses the PE parser to build the PE file representation (the aforementioned tree-like structure of Java objects) and then creates AST nodes and their aspects inside MPS.
-The extended syntax brought by PE describes the layout of each element, e.g.
-it tells the plugin that one set of child nodes should be displayed horizontally, another set should be vertical with each child on a separate line and with some indentation, and so on.
-
-A limitation of the PE4MPS approach is that it, from what we understand, does not implement full ANTLR syntax.
-This means that every grammar might need a non-trivial adjustment before its usage.
-One of our goals is to enable import of as many languages out-of-the-box as possible and adopting the full specification.
 

### 2016-10-13 chapters/introduction.tex
text about pe4mps

@@
 The prevailing approach to define valid syntaxes for programming languages is through grammars, which are typically written in notations based on the Extended Backus-Naur Form (EBNF).
-Many popular tools exist for automated generation of parsers from the grammar definitions --- for example, ANTLR~\cite{ref:ANTLR} and Bison/Flex~\cite{ref:BISONFLEX}.
+Many popular tools exist for automated generation of parsers from the grammar definitions --- for example, ANTLR~\cite{ref:ANTLRBOOK,ref:ANTLR} and Bison/Flex~\cite{ref:BISONFLEX}.
 The source code of programs is written in a text editor, usually embedded into an IDE. Parsers are used to create an abstract syntax tree (AST) representation of the code, e.g. inside the compiler.

### 2016-10-13 chapters/background.tex
related projects

@@
 \paragraph{BaseLanguage.}
-Another important feature of MPS that we need to describe in more detail is \emph{BaseLanguage}~\cite{ref:BaseLanguage}.
+Another important feature of MPS that we need to describe in more detail is \emph{BaseLanguage}~\cite{ref:BASELANG}.
 It is a clone of Java implemented using the MPS constructs.
@@
 
-The BaseLanguage~\cite{ref:BaseLanguage} is an almost full port of the Java language, extended with MPS-specific features.
+The BaseLanguage~\cite{ref:BASELANG} is an almost full port of the Java language, extended with MPS-specific features.
 It was created manually by JetBrains, and it is still undergoing changes as Java itself is evolving.
@@
 
-% TODO continue from here
+\paragraph{ANTLR{\_}MPS.}
+The ANTLR{\_}MPS project~\cite{ref:ANTLR2MPS} also uses the ANTLR v4 grammar notation, but otherwise works quite differently from PE4MPS.
+We provide only a brief mention here because this project is in an early stage of development.
+Its author created an ANTLRv4 MPS language, which captures the syntax of ANTLRv4 notation using MPS concepts.
+Given some language as input, the textual grammar of the language is imported automatically into MPS, taking the form of the ANTLRv4 MPS language.
+The next step would be to create a new MPS language from the grammar definition in MPS, but it is not implemented yet.
 
-\paragraph{ANTLR{\_}MPS.}
-\PV{Tenhle projekt byl z tech tri nejmene uzitecny a podle me i nejmene funkcni, tak mozna by se dal v ramci zkracovani vypustit. V dalsim projektu je na nej reference, tak treba ji vyhodit.}
-ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
-The author of this project is Fabien Campagne.
-The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
-However, it does not try to generate the projectional editor at all (i.e., it does not deal with this problem), probably because it is in an early stage of development.
-
-The way this import process of ANTLR{\_}MPS works is quite different from what we have seen in the PE4MPS project, and it is quite complex/complicated to use.
-It works as follows. We give only a brief high-level overview and omit low-level technical details.
-The author created a whole new ANTLRv4 MPS language, which is an MPS port of the grammar notations' syntax.
-To import a language, the user utilizes this MPS ANTLR language.
-The textual grammar is imported automatically into MPS taking the form of the MPS's ported grammar language (so that the textual grammar is converted into MPS nodes, that means an AST node is created for each grammar rule).
-No child-parent relationships in the structure are generated by the tool --- all must be created manually by the user.
-There are no editor nor TextGen aspects created neither.
+However, the project has also other important limitations.
+The tool does not generate any parent-child relationships in the Structure aspect, and it completely neglects the Editor and TextGen aspects.
+In particular, it does not support creating of the projectional editor at all.
 
 \paragraph{mps-metabnf.}
-\PV{Pridal jsem popis projektu}
-% TODO - citace na https://github.com/DSLFoundry/mps-metabnf
-The mps-metabnf project comes from the DSLFoundry group and also takes on grammar importing.
-Even though it is in a very early stage of development, it holds some interesting ideas.
-Similarly to ANTLR{\_}MPS, authors of mps-metabnf decided to create an MPS language describing the grammar.
-When grammar is being imported into MPS, it is just transformed into the terms of this MPS language.
-User is then able to adjust this grammar.
-After this, a second step of the import process is needed, that generates the final MPS language out of this adjusted grammar definition.
+The mps-metabnf tool~\cite{ref:MPSMETABNF}, implemented by members of the DSLFoundry group, builds upon similar ideas as the ANTLR{\_}MPS project described above.
+Authors of this project, too, created an MPS language that can be used to capture ANTLR grammars.
+Every imported grammar is represented using this ANTLRv4 MPS language.
+A user is then able to adjust the grammar as he wishes, specifying the code layout in the projectional editor.
+In the last step, the target MPS language is derived from the adjusted grammar definition.
+Many iterations of this process, involving adjustments of the grammar followed by recreation of the target language, can be performed.
 
-The key takeaways from the mps-metabnf project are that the intermediate grammar MPS representation, when finished, will offer some powerful means of adjustment.
-Using the projectional editor, the user can easily specify information about the code layout, export the target MPS language and, in case further adjustments are needed, come back to the grammar definition, fix any problems and regenerate the language.
-The MPS language of the grammar definition can be easily extended with needed features (indenting, line breaks...), which is superior to our approach, because they can be represented using many visual means.
-
-This approach also has some downsides.
-Because MPS doesn't offer any tools that would allow easy transformation from the MPS grammar \PP{To je konkretni gramatika zadefinovana pomoci MPS jazyka pro ANTLR} to the MPS language, the second step of the import process hasn't been implemented yet and will be probably very problematic.
-Another downside might be the need for a double transformation implementation -- first one being the import of the text grammar to the MPS grammar definition.
-
-From another point of view -- for each additional feature, that we decide to add to the grammar notation, we need to implement a corresponding transformation.
-This subsequently leads to duplicating the projectional editor that is already present in MPS.
-This means that there needs to be a line drawn between which features are added and which will be left for the user to add to the target language using MPS.
+The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLRv4 MPS language into the actual target language.

### 2016-10-13 paper.tex
text about pe4mps
related projects

@@
 
+\bibitem{ref:BISONFLEX} J. Levine. Flex \& Bison: Text Processing Tools. O'Reilly Media, 2009.
+
+\bibitem{ref:ANTLRBOOK} T. Parr. The Definitive ANTLR 4 Reference. Pragmatic Bookshelf, 2013.
+
 \bibitem{ref:MBEDDR} M. Voelter, D. Ratiu, B. Schaetz, and B. Kolb. mbeddr: an Extensible C-based Programming Language and IDE for Embedded Systems. SPLASH 2012, ACM.
@@
 
-\bibitem{ref:BISONFLEX} J. Levine. Flex \& Bison: Text Processing Tools. O'Reilly Media, 2009.
+\bibitem{ref:ANTLR2MPS} F. Campagne. The ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR_MPS}
+
+\bibitem{ref:MPSMETABNF} DSLFoundry. The mps-metabnf project, \url{https://github.com/DSLFoundry/mps-metabnf}
 
@@
 
-\bibitem{ref:BaseLanguage} MPS BaseLanguage, \url{https://confluence.jetbrains.com/display/MPSD34/Base+Language}
+\bibitem{ref:PE4MPS} M. Lombardo. The PE4MPS project, \url{https://github.com/mar9000/pe4mps}
+
+\bibitem{ref:PE} M. Lombardo. PE grammars, \url{https://github.com/mar9000/pe}
+
+\bibitem{ref:BASELANG} MPS BaseLanguage, \url{https://confluence.jetbrains.com/display/MPSD34/Base+Language}
 
 % TODO opravit formatovani nasledujicich polozek a vhodne zaradit
-
-\bibitem{ref:PE4MPS} PE4MPS project, \url{https://github.com/mar9000/pe4mps}
-
-\bibitem{ref:PE} PE project, \url{https://github.com/mar9000/pe}
-
-\bibitem{ref:ANTLR2MPS} ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR{\_}MPS}
-
-\bibitem{ref:mpsmetabnf} mps-metabnf \url{https://github.com/DSLFoundry/mps-metabnf}
 

### 2016-10-22 chapters/background.tex
writing section 3

@@
 \subsection{Related Projects}
+\label{sect:RELATED}
 
@@
 
+The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
+Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
 Another limitation of PE4MPS is that it does not implement the full ANTLR syntax.

### 2016-10-22 chapters/evaluation.tex
fix to make the paper compile
finished introduction
writing section 3

@@
 \section{Evaluation}
+\label{sect:EVAL}
+
+\PP{Zhodnotit jak se podarilo dosahnout automatizace. How much we fulfilled our goal of maximal possible automation.}
 
@@
 
-We experimented with import of XML, JavaScript, Python, Ruby, C#, Swift, MySQL and some other languages.
+We experimented with import of XML, JavaScript, Python, Ruby, C\#, Swift, MySQL and some other languages.
 

### 2016-10-22 chapters/language_definition.tex
related projects
writing section 3

@@
-\section{Creating Language Definition in MPS}
+\section{Constructing Languages in MPS}
 \label{sect:LANGDEF}
 
-From all the available notations in which grammars can be written, we decided to use the ANTLRv4 grammar notation~\cite{ANTLR4reference} as the input format (in which the input programming languages have to be defined) for the following reasons.
-\begin{itemize}
-	\item It is a very widely used grammar notation.
-	\item All popular and widely used programming languages have their syntax captured (represented) using this notation --- see the repository~\cite{ref:ANTLR4grammars}.
-	\item There is a whole framework/toolset for generating parsers and editing grammars built around it~\cite{ref:ANTLR4}.
-	\item It supports Java (it is written in Java), which is the implementation language of MPS and also the default BaseLanguage.
-	\item It uses (is based on) an EBNF (Extended Backus-Naur Form) \footnote{https://en.wikipedia.org/wiki/Extended{\_}Backus\%E2\%80\%93Naur{\_}Form} form that it useful because it meets/respects the object oriented design that is used when describing MPS aspects (definitions of MPS languages).
-	\item It has its own syntax captured in its own syntax\footnote{https://github.com/antlr/grammars-v4/tree/master/antlr4} (there is an ANTLRv4 grammar for the ANTLRv4 notation).
-This is a very important feature as it allows us to create a parser for ANTLRv4 grammars so that we can parse grammars of other languages.
-\end{itemize}
+The proposed INGRID method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
+All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
+Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.
 
-We use ANTLRv4 grammars as inputs (and work with them).
-	https://github.com/antlr/antlr4/blob/master/doc/index.md
-	https://github.com/antlr/grammars-v4
+The process of MPS language construction, as performed by INGRID, consists of four steps --- parsing of the input grammar, followed by definition of the essential aspects (Structure, Editor, and TextGen).
+Each of the steps 2-4 is exclusively and fully responsible for one aspect.
 
-Unlike some of the related projects, we did not add any new features into the grammar notation.
-We tried to import grammars as they are.
+INGRID is currently able to create (i) a full Structure aspect for each element (type of an AST node) of the given language, (ii) a very basic Editor, and (iii) a basic TextGen aspect.
+Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the human users must also define the remaining aspects not yet supported by INGRID.
 
-The generation/import process performed by INGRID consists of three phases (import of structure, usable editor, text generation).
-We will talk about all phases of the generation/import process, describing encountered challenges/obstacles using an example language.
+Our approach differs from the related projects (Section~\ref{sect:RELATED}) especially in the level of automation. It also has much better support for the definition of Editor and TextGen.
 
-To fulfill our goal of maximal possible automation, we import/generate the full language structure and then do a lot of mundane and time-consuming work connected to creating projectional editors.
-
-Now we describe/introduce the main steps of our approach (that INGRID tool performs): parsing the grammar, creating the structure, editor, and textgen.
-We are able to create/generate the structure aspect and then very basic editor and some default/basic textgen.
-In particular, we did not port the ANTLR notation as an MPS language.
-In the following subsections, we provide some details about our approach to parsing the input grammar and creating the main aspects of each element (each type of AST node) of an MPS language (structure, editor, and textgen).
-
-During our work on this project, we also discovered/encountered challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component), whitespace (editor aspect).
-There is a challenge/issue that we have discovered during our work on this project (in the process of our own implementation), and that is related to construction of the structure aspect.
-
-Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
-However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
-\PV{Dalsi veta asi zbytecna}
-We describe/discuss these challenges in more detail (also using examples) and our solutions to the challenges inside the respective following subsections.
-
-Result of the construction of a MPS language from the input grammar (import) has to be polished by the human user.
-For example, the aspects not supported by INGRID will be used for adjusting the imported language and improving the usability of the language.
-
-\todo{Briefly discuss how our overall approach differs from similar projects (PE4MPS, etc).}
-There are more ways how to approach the problem of a grammar import.
-
-Our overall approach differs from the PE4MPS project (Section X.Y) especially in the level of automation.
-While the main goal of (our) the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible, usage of PE does not really spare the user any work because it only shifts the manual work from creating projectional editor to writing PE grammars.
-What the user would do inside the MPS UI using the projectional editor is in the case of PE4MPS done using a text editor when editing PE grammars.
-MPS UI is specially designed/made/tailored for this purpose, guides the user through this process, and supports aids such as auto-completion.
-In particular, MPS UI allows the user to perform only changes that result in valid output.
-Editing the grammar stored in a text file using a standard editor is an inferior approach with respect to using MPS (projectional editor), because a text editor is not as user-friendly, and, more importantly, it is an error prone environment.
-
-Main drawback of ANTLR{\_}MPS, and main difference from our approach, is very limited automation: it does not create the parent-child relationships between AST nodes in the structure, and does not generate any editor and textgen aspects (for any node) at all.
-In addition, there is no attention paid to the problem of generating projectional editor and text generation is neglected too.
-We blame all this on the project being in its early stage of development.
-
-In the rest of this section, first we define the language that we will use throughout the paper as the running example.
-Then we will talk about parsing grammar files \todo{Tohle je mozna zbytecne}.
-Lastly, we will describe our approach to tackling generation of language's aspects (structure, editor, and TextGen).
-At some places, we discuss important possible alternatives (as there are several possible approaches).
+% TODO continue from here
 
@@
 
-We illustrate the main steps on the example of a simplified XML language, which is defined below (in Figure XY).
+We illustrate the main steps of the INGRID method on the example of a simplified XML language, which is defined below (in Figure XY).
 
@@
 
+\todo{We import/generate the complete/full language structure fully automatically.}
+
 Now, as the next step, we need to translate/transform the AST (the types of AST nodes) derived from the input grammar into elements of the structure aspect of the MPS language definition (into types of AST nodes), and linking the nodes appropriately.
@@
 The process of transferring grammar rules into the structure of a MPS language presents/involves some interesting challenges.
+
+As we already indicated in the introduction, the main problem/difficulty/challenge related to Structure (creating the structure) is that a grammar typically contains many rules that do not directly correspond to AST nodes and programming language constructs.
+Challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component).
+There is a challenge/issue that is related to construction of the structure aspect.
+Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
+However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
 The input grammar (EBNF, ANTLR) serves a little bit different purpose than MPS language, and the way grammar rules are usually written down causes some problems that we had to overcome in order to recreate the language inside MPS.
 Note that the challenges and problems are not specific to MPS --- they are common for everyone tackling a similar problem.
+In the rest of this section, we describe our solution to this problem/challenge.
 
@@
 
+\todo{Definition of projectional editors is only partially automated. We do a lot of mundane and time-consuming work connected to creating projectional editors.}
+
 When all types of AST nodes in the language are created, the next step is to define their visual representation in the projectional editor.
@@
 
+The main problems/challenges in the process of definition of projectional editors, also/again related to input grammars, are code layout is whitespace (editor aspect).
+As we already said in the introduction, one of the biggest challenges/problems that arise here is that the grammar description of a language (the description of a language in the form of a grammar) does not hold any information about the code layout.
 The main challenge (problem) here is that an input grammar (ANTLR) does not provide/give any help/aid/information regarding the code layout.
-As we already said in the introduction, one of the biggest challenges/problems that arise here is that the grammar description of a language does not hold any information about the code layout.
 The structural rules in the grammar only tell us, what the syntax tree looks like and how the code is broken into nodes.

### 2016-10-22 chapters/language_definition.tex
running example

@@
-\section{Constructing Languages in MPS}
+\section{Creating Language Definition in MPS}
 \label{sect:LANGDEF}
@@
 
-% TODO continue from here
-
 \subsection{Running Example}
 
-We illustrate the main steps of the INGRID method on the example of a simplified XML language, which is defined below (in Figure XY).
-
-\todo{Pouzit jazyk SimpleXML na running exaple.}
-\PV{Dalsi veta asi prijde niz?}
-
-We decided to use SimpleXML because it is small enough to fit here, but still can be used to illustrate the main challenges and our solution
-
-Throughout the whole paper, we will be using one example language to illustrate main challenges and our solution.
-The language is a simplified version of XML. We call it \emph{SimpleXML}.
-We created the grammar of the language by taking the official XML ANTLRv4 grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and stripped it down to eliminate/avoid some not so interesting (advanced) features such as XML entities.
-Below in Figure X, you can find the grammar of the SimpleXML language in the ANTLRv4 notation.
+We will describe the INGRID method using the example of a simplified XML language that is defined in Figure~\ref{fig:SIMPLEXML}.
+The \emph{SimpleXML} language is small but complex enough to be used for illustration of the main challenges and behavior of the proposed algorithms.
 
@@
 \end{framed}
-\caption{SimpleXML grammar}
+\caption{Grammar of the SimpleXML language in the ANTLR v4 notation}
+\label{fig:SIMPLEXML}
 \end{figure*}
 
-We have colored the grammar so it is easier to find the way around it (to make navigation easier).
-The grammar contains following basic elements:
+The grammar of SimpleXML contains elements of the kinds that are listed below.
+Each color in Figure~\ref{fig:SIMPLEXML} corresponds to a different kind in a way that is indicated by the list item headers.
+\begin{itemize}
+	\item \textbf{ANTLR v4 keywords} are required by the notation.
+	\item \antlrparserrule{\textbf{Parser rules}} describe the structure of a language.
+		Alternatives on the right side of a rule are separated by the pipe character ($|$).
+	\item \antlrlexerrule{\textbf{Lexer rules}} describe terminal symbols that the parser matches against the input string.
+		A terminal symbol can be encoded as a string value or using a regular expression.
+		The \textbf{fragment} keyword states that the lexer rule just brings more clarity into the notation but it is not visible in the parser output.
+	\item \antlrliteralnoap{\textbf{String literals}}, always written inside of a pair of single quote marks, also represent terminal symbols, but by exact match to a string constant.
+	\item \antlrregex{\textbf{Regular expressions}} describe string tokens to be matched against regular expressions with a special ANTLR v4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
+\end{itemize}
+Elements on the right side of a rule can be annotated with standard EBNF operators (?+*) that specify the allowed number of occurrences.
 
-\begin{itemize}
-	\item \textbf{ANTLRv4 keywords} -- In black, they are required by the specification.
-	\item \antlrparserrule{\textbf{Parser rules}} -- Stated in blue, they are the rules describing language's structure.
-		The alternatives (that the element on the left side may break into) are separated by the pipe ($|$) character.
-	\item \antlrlexerrule{\textbf{Lexer rules}} -- In yellow, they are rules describing terminal symbols that the parser is matching against the input.
-		Terminals are encoded as string values and regular expressions.
-		The \textbf{fragment} keyword states that the lexer sub-rule is just a helper rule that brings more clarity into the notation but is not visible in the parser output.
-	\item \antlrliteralnoap{\textbf{String values (literals)}} -- In red, always inside of a pair of apostrophes, they are similar to lexer rules.
-		They describe a terminal string, not using regular expressions but by an exact string match.
-	\item \antlrregex{\textbf{Regular expressions}} -- Colored in green, they describe a string token to be matched using a regular expression with the special ANTLRv4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
-	\item \textbf{Rule operators (?+*)} - Left in black, following any element, they quantify the number of occurrences the prepending element can appear in.
-		These operators can also be prepended by an asterisk (*), meaning they are non-greedy.
-\end{itemize}
+% TODO continue from here
 

### 2016-10-24 chapters/language_definition.tex
parsing

@@
 
-The process of MPS language construction, as performed by INGRID, consists of four steps --- parsing of the input grammar, followed by definition of the essential aspects (Structure, Editor, and TextGen).
-Each of the steps 2-4 is exclusively and fully responsible for one aspect.
+The process of MPS language construction, as performed by INGRID, consists of four phases --- the first is parsing of the input grammar, which is followed by definition of the essential aspects (Structure, Editor, and TextGen).
+Each of the phases 2-4 is exclusively and fully responsible for one aspect.
 
@@
 \end{itemize}
-Elements on the right side of a rule can be annotated with standard EBNF operators (?+*) that specify the allowed number of occurrences.
+Elements on the right side of a rule can be annotated with standard EBNF operators (\code{?}, \code{+}, \code{*}) that specify the allowed number of occurrences.
 
-% TODO continue from here
+\subsection{Parsing Input Grammar}
 
-\subsection{Parsing (the) Grammar}
+The first phase in the process of MPS language construction is parsing of the input grammar that is defined using the ANTLR v4 notation.
+An output of this phase is an AST representing the grammar.
+Nodes of the AST correspond to rules and other grammar elements.
 
-\todo{Define some name for the AST representing the grammar, and use it consistently throughout the whole paper. Maybe distinguish between "grammar AST" and "MPS structure AST".}
-The first step (stage) of creating the MPS language (of the import process) is parsing the grammar file (written using the ANTLRv4 notation) so that we can get an AST tree representing the grammar.
+Nevertheless, the full AST, which comes out of the automatically generated parser, is quite complex and contains information not relevant for the INGRID algorithm.
+In order to get a simple representation that is easy process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST have to be performed in this phase.
 
-The abstract syntax tree that comes out of the automatically generated (ANTLR) parser is a little bit complicated and full of information not relevant for our purposes (to us).
-That is why we translate the complex ANTLR AST into our own simplified tree made out of our own objects.
-We discarded as much irrelevant information as possible and only kept information (data) vital/necessary for construction of MPS languages.
+A simplified tree of objects is derived from the AST in two passes over the grammar.
+The result of the first pass is a tree where each node contains names of other referenced rules in the form of strings.
+In the second pass, the whole tree is traversed and all the references are resolved to have a form of real pointers to other rule objects.
 
-Furthermore, we process the grammar in two iterations.
-In the first iteration/pass, we build a tree holding names of other referenced rules in the form of strings.
-In the second iteration, we walk this tree and resolve all these references to real pointers to other rule objects.
-We end up with a clear and simpler representation that is easy to process in the next steps of the import process.
+The representation of lexer rules (tokens) in the AST has to be simplified too.
+Note that, in the ANTLR notation, lexer rules can be built from alternatives just like the parser rules --- see, for example, the lexer rule \antlrlexerrule{Name} from our SimpleXML language (Figure~\ref{fig:SIMPLEXML}).
+For each lexer rule, the parser produces a tree that captures the structure of the rule.
+However, the lexer rules are, in fact, just regular expressions used to recognize tokens in the input string.
 
-We also had to post-process (simplify) the representation of lexer rules (tokens).
-As a result of the initial parsing, ANTLR parser produces some kind of a tree structure that is representing the lexer rule (meaning/note that lexer rules can also be built from alternatives just like parser rules).
-However, lexer rules are, in fact (basically), just regular expressions used for parsing input source code into tokens.
-We took each/every such tree that represents the lexer rule and flattened it into the equivalent regular expression.
-
-For illustration, consider the lexer rule \antlrlexerrule{Name} from our/the SimpleXML language (Figure XY), together with all its sub-rules, shown in Figure~\ref{fig:lexerrule}.
+Every tree that represents a lexer rule is flattened into the equivalent regular expression by the recursive algorithm specified in Figure~\ref{fig:ALGFLATTEN}.
+A distinct sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.
 
@@
 \begin{alltt}
-	\antlrlexerrule{Name}     : \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-	\textbf{fragment}
-	\antlrlexerrule{DIGIT}    : \antlrregex{[0-9]} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameChar} : \antlrlexerrule{NameStartChar} 
-	         | \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
-	\textbf{fragment}
-	\antlrlexerrule{NameStartChar} : \antlrregex{[:a-zA-Z]} ;
-\end{alltt}
-\end{framed}
-\caption{Lexer rule from SimpleXML}
-\label{fig:lexerrule}
-\end{figure}
-
-Intuitively, we can create a regular expression by gluing elements of each alternative together and then join these alternatives with a classic regex OR ($|$), which is exactly its function in the ANTLRv4 grammar notation.
-We have captures this idea in the form of the recursive algorithm in Figure~\ref{fig:algflatten}, which covers the whole process.
-
-% TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
-\begin{figure}[ht]
-\centering
-\begin{framed}
-\begin{alltt}
 Flatten(\textit{R}):
-  Define \textit{T} as a list of empty string
-  For each alternative \textit{A} of the rule \textit{R}:
-    Define \textit{R} = \antlrap\antlrap
-    For each element \textit{E} of \textit{A}:
-	  If \textit{E} is not yet flattened:
+  \textit{T} = empty list
+  \textbf{for each} alternative \textit{A} \textbf{in} rule \textit{R}:
+    \textit{R} = \antlrap\antlrap
+    \textbf{for each} element \textit{E} \textbf{of} \textit{A}:
+      \textbf{if} \textit{E} is not yet flattened \textbf{then}:
         Flatten(\textit{E})
@@
     \textit{T}.add(\textit{R})
-  Build a string S from elements of \textit{T}:
-    (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
-  Return \textit{S}
+  build string S from elements of \textit{T}:
+    \textit{S} = \(t\sb{1}\) | \(t\sb{2}\) | \(\ldots\) | \(t\sb{n}\)
+  \textbf{return} \textit{S}
 \end{alltt}
 \end{framed}
-\caption{Flatten algorithm}
-\label{fig:algflatten}
+\caption{Flattening algorithm}
+\label{fig:ALGFLATTEN}
 \end{figure}
 
-The output of \texttt{Flatten(Name)} (output of the Flatten algorithm applied to the \antlrlexerrule{Name} rule) is this regular expression (in this form):
+The output of \texttt{Flatten(Name)}, i.e. application of the algorithm to the \antlrlexerrule{Name} rule, is this regular expression:
 
@@
 \end{center}
+It defines the syntactically valid identifiers of elements and attributes in SimpleXML.
+Each identifier must begin with a letter, followed by any combination of letters, digits, underscore, dash, or a dot.
+
+% TODO continue from here
 

### 2016-10-25 chapters/language_definition.tex
structure - in progress

@@
 
-The process of MPS language construction, as performed by INGRID, consists of four phases --- the first is parsing of the input grammar, which is followed by definition of the essential aspects (Structure, Editor, and TextGen).
+The process of MPS language construction, as performed by INGRID, consists of four phases --- the first is parsing of the input grammar, which is followed by definition of the essential aspects of the language (Structure, Editor, and TextGen).
 Each of the phases 2-4 is exclusively and fully responsible for one aspect.
@@
 
-\subsection{Parsing Input Grammar}
+\subsection{Phase 1: Parsing Input Grammar}
 
@@
 
+\subsection{Phase 2: Structure}
+
+In the next phase, the complete structure of the MPS language is automatically generated from the AST that represents syntax of the input language.
+Elements of the Structure aspect are derived from the AST nodes, and then linked appropriately.
+Therefore, structure of the MPS language is usually similar to the original ANTLR grammar.
+
+When designing the process for transforming AST nodes (i.e., grammar rules) into the MPS language structure, we faced several challenges.
+The main challenge related to the structure, as we already mentioned in the introduction, is that a grammar typically contains rules that do not directly correspond to programming language constructs.
+Such rules exist at the intermediate layers of a syntax hierarchy, and their main purpose is to enable easier understanding of the grammar by humans.
+We show examples illustrating this problem later, in the subsections where we describe our solution.
+Nevertheless, presence of the intermediate layers significantly complicates usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
+
+We have designed two approaches to eliminate the unnecessary layers during the process of creating the Structure aspect --- the \emph{straightforward} approach (Section~\ref{sect:STRAIGHTFWD}), and the \emph{shortcut} approach (Section~\ref{sect:SHORTCUT}) that is more complex and builds upon the first one.
+
 % TODO continue from here
-
-\subsection{Structure}
-
-\todo{We import/generate the complete/full language structure fully automatically.}
-
-Now, as the next step, we need to translate/transform the AST (the types of AST nodes) derived from the input grammar into elements of the structure aspect of the MPS language definition (into types of AST nodes), and linking the nodes appropriately.
-
-The goal of this second step is to translate (transfer) the syntax (structure) of the input language that is captured by the grammar (ANTLR), into the MPS structure aspect.
-The structure of the MPS language is somewhat similar to the grammar.
-This is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
-The process of transferring grammar rules into the structure of a MPS language presents/involves some interesting challenges.
-
-As we already indicated in the introduction, the main problem/difficulty/challenge related to Structure (creating the structure) is that a grammar typically contains many rules that do not directly correspond to AST nodes and programming language constructs.
-Challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component).
-There is a challenge/issue that is related to construction of the structure aspect.
-Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
-However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
-The input grammar (EBNF, ANTLR) serves a little bit different purpose than MPS language, and the way grammar rules are usually written down causes some problems that we had to overcome in order to recreate the language inside MPS.
-Note that the challenges and problems are not specific to MPS --- they are common for everyone tackling a similar problem.
-In the rest of this section, we describe our solution to this problem/challenge.
-
-We are going to describe two different approaches that we designed --- the straightforward approach (Section~\ref{sect:straightforward_approach}) and the shortcut approach (Section~\ref{sect:shortcut_approach}).
-We will describe two possible approaches that we developed in Section X and Section Y, respectively.
-\todo{Rict neco jako ze ten druhy, slozitejsi, approach vlastne rozsiruje ten prvni, jednodussi.}
 
@@
 \subsubsection{The Straightforward Approach}
-\label{sect:straightforward_approach}
+\label{sect:STRAIGHTFWD}
 
@@
 \subsubsection{The Shortcut Approach}
-\label{sect:shortcut_approach}
+\label{sect:SHORTCUT}
 

### 2016-10-26 chapters/language_definition.tex
straightforward approach

@@
 
-When designing the process for transforming AST nodes (i.e., grammar rules) into the MPS language structure, we faced several challenges.
+When designing the process for translating AST nodes (i.e., grammar rules) into the MPS language structure, we faced several challenges.
 The main challenge related to the structure, as we already mentioned in the introduction, is that a grammar typically contains rules that do not directly correspond to programming language constructs.
-Such rules exist at the intermediate layers of a syntax hierarchy, and their main purpose is to enable easier understanding of the grammar by humans.
-We show examples illustrating this problem later, in the subsections where we describe our solution.
-Nevertheless, presence of the intermediate layers significantly complicates usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
+Such rules exist at the intermediate layers of a syntax hierarchy.
+Their main purpose is to enable easier understanding and maintenance of the grammar by humans.
+However, presence of the intermediate layers significantly complicates usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
+We show examples illustrating this problem, which we call a \emph{layer problem}, later in the subsection where we describe our solution.
 
-We have designed two approaches to eliminate the unnecessary layers during the process of creating the Structure aspect --- the \emph{straightforward} approach (Section~\ref{sect:STRAIGHTFWD}), and the \emph{shortcut} approach (Section~\ref{sect:SHORTCUT}) that is more complex and builds upon the first one.
+As a part of the INGRID method, we have designed an approach to eliminate the unnecessary layers during construction of the Structure aspect --- we call it the \emph{shortcut} approach (Section~\ref{sect:SHORTCUT}).
+However, before focusing on the intermediate layers and the shortcut approach, we describe the basic principles of translation from the AST into the language structure.
 
-% TODO continue from here
-
-However, before focusing on the two approaches, we must describe the basic principles of translation from the AST representing the grammar to the MPS structure, which are common to both approaches.
-The structure is created according to object-oriented principles.
-Each node (element) is represented by an object (and can be treated as such) that may have a parent node, child nodes, and properties of any data type (similar to fields), it may implement any number of interfaces, and it may also hold references to other nodes.
-For every rule of the grammar (for every node in the grammar AST) that breaks into more/multiple alternatives, we must create a node in the MPS structure AST for each alternative of the grammar (every child node).
-Names of the nodes for alternatives are composed from the name of the rule and a number indicating the order of appearance in the grammar.
-For example, consider the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlelementrule}.
-Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.
-
-\begin{figure*}[ht]
-\centering
-\begin{framed}
-\begin{alltt}
-	\antlrparserrule{element} : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrap<\antlrap Name attrib* \antlrap/>\antlrap ;
-\end{alltt}
-\end{framed}
-\caption{Element rule from the SimpleXML grammar}
-\label{fig:xmlelementrule}
-\end{figure*}
+Each concept of the language (i.e., every AST node) is represented by an object that may have a parent, some children, and properties of any data type.
+In addition, the object may implement any number of interfaces, and it may also contain references to objects representing other AST nodes.
+The parent-child relationships between objects that make the Structure aspect are derived from rules of the grammar (i.e., from the structure of the AST).
+For each rule, the object corresponding to the left-hand side is in the parent-child relationships with sets of objects that represent language elements referenced by the right-hand side.
+If a rule has multiple alternatives, then a distinct object (MPS concept) has to be created in the structure for each alternative.
+The name of a concept (object) in MPS is composed from (1) the name of the AST node, which is equivalent to the string encoding of the left-hand side of the corresponding rule, and (2) the number indicating the position of the respective alternative on the right-hand side of the rule.
 
@@
 \centering
-	\includegraphics[scale=0.75]{./images/element_node_common.png}
-	\caption{Resulting Element{\_}1 node}
-	\label{fig:element_node_common}
+\begin{alltt}
+\small
+\mpsstkeyword{concept} Element\_1 \mpsstkeyword{extends} BaseConcept
+        \mpsstkeyword{implements} IContent, IElement
+
+  \mpsstkeyword{instance can be root:} false
+  \mpsstkeyword{alias:} \mpsstalias{< > </ >}
+  \mpsstkeyword{short description:} \mpsstalias{Element}
+
+  \mpsstkeyword{properties:}
+  \mpsstproperty{Name\_1} : Name
+  \mpsstproperty{Name\_2} : Name
+
+  \mpsstkeyword{children:}
+  \mpsstproperty{Attribute\_1} : Attribute[\mpsstcardinality{0..n}]
+  \mpsstproperty{Content\_2}   : IContent[\mpsstcardinality{0..n}]
+\end{alltt}
+\caption{Structure aspect of the \code{Element{\_}1} concept}
+\label{fig:ELEMENTSTRUCT}
 \end{figure}
 
-String literals (e.g., keywords such as \textbf{for}) are omitted/skipped because they will appear only in the corresponding projectional editor.
-The node named \mpsconcept{Element{\_}1} contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name} inside the first alternative of the \antlrparserrule{element} rule,
-Value of the property will be restricted using the regular expression representing the \antlrlexerrule{Name} rule.
-We explain later how the references to parser rules (\antlrparserrule{attribute} and \antlrparserrule{content}) as this is where the approaches differ.
-The main/key difference between these two lies in the way we create children fields for parser rules and in the way we are going to link them together using interface concepts.
+Consider the parser rule \antlrparserrule{element} from the grammar in Figure~\ref{fig:SIMPLEXML}.
+Its first alternative represents the full XML element with content.
+Figure~\ref{fig:ELEMENTSTRUCT} shows the Structure aspect for the MPS concept (object) named \code{Element{\_}1} that corresponds to the alternative.
+The object contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name}.
+Values of these properties should be restricted using the regular expression that corresponds to the \antlrlexerrule{Name} rule, but this is left to the human user.
+String literals, such as the keyword \textbf{for}, are omitted because they will be defined only in the Editor aspect for this concept.
 
-\subsubsection{The Straightforward Approach}
-\label{sect:STRAIGHTFWD}
-
-The key idea of this approach is that, for each rule with more than one alternative, an interface node is created.
-Consider the \antlrparserrule{content} rule from our SimpleXML language (Figure~\ref{fig:contentrule}).
+References to other parser rules are captured by pointers to child objects.
+The types of child objects, such as \code{IContent} in the \code{Element{\_}1} concept, are determined as follows.
+Consider the \antlrparserrule{content} rule from the SimpleXML language.
+We show just the relevant fragment of the grammar again in Figure~\ref{fig:CONTENTRULE}.
+An object corresponding to any one of the four alternatives could be the actual value anywhere the \antlrparserrule{content} rule is referenced.
 
@@
 \centering
-\begin{framed}
 \begin{alltt}
-	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
-           |   \antlrparserrule{element}
-           |   \antlrparserrule{comment}
-           |   \antlrlexerrule{CDATA}
-           ;
+\small
+\antlrparserrule{content} : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
 \end{alltt}
-\end{framed}
-\caption{Content rule}
-\label{fig:contentrule}
+\caption{Parser rule \antlrparserrule{content}}
+\label{fig:CONTENTRULE}
 \end{figure}
 
-Any of the four nodes corresponding to alternatives could appear anywhere the \antlrparserrule{content} rule is referenced.
-So we will create one interface node, and then for each alternative of this rule we create one node that implements the interface.
-The resulting fragment of structure will look like the one in Figure~\ref{fig:icontentitf}.
+Our solution is to use interface concepts.
+For each rule with more than one alternative on the right side, i.e.
+for each AST node with more than one child, first an interface concept is defined in the MPS language structure, and then one object that implements the given interface is created for each alternative.
+The resulting fragment of the language structure looks like the one in Figure~\ref{fig:ICONTENTITF}.
+It contains one interface \mpsinterface{IContent} that is implemented by four object concepts \mpsconcept{Content{\_}1}, $\ldots$, \mpsconcept{Content{\_}4}.
 
@@
 \centering
-\begin{framed}
 \begin{alltt}
-	\mpsinterface{IContent}   :   \mpsconcept{Content{\_}1}
-           |   \mpsconcept{Content{\_}2}
-           |   \mpsconcept{Content{\_}3}
-           |   \mpsconcept{Content{\_}4}
+\small
+\mpsinterface{IContent} : \mpsconcept{Content{\_}1} | \mpsconcept{Content{\_}2} |
+           \mpsconcept{Content{\_}3} | \mpsconcept{Content{\_}4}
 \end{alltt}
-\end{framed}
-\caption{IContent interface}
-\label{fig:icontentitf}
+\caption{MPS interface \code{IContent} and the types of implementing objects}
+\label{fig:ICONTENTITF}
 \end{figure}
 
-References to other parser rules are captured by links to children (interfaces or object nodes) in the AST nodes.
-For the \antlrparserrule{element} rule that is referencing the \antlrparserrule{content} rule, defined as in Figure~\ref{fig:xmlelementrule}, we get the AST node representing the first alternative that is shown in Figure~\ref{fig:element_node_full}.
-There is one \mpsinterface{IElement} interface node that is implemented by two object nodes \mpsconcept{Element{\_}1}, \mpsconcept{Element{\_}2}.
-The node \mpsconcept{Element{\_}1} has two child links pointing to \mpsinterface{IAttribute} and \mpsinterface{IContent}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[scale=0.75]{./images/element_node_full.png}
-	\caption{Element{\_}1 node's structure aspect}
-	\label{fig:element_node_full}
-\end{figure}
-
-A problem with this approach, which we call \emph{the layer problem}, is related to the grammar rules (and corresponding AST nodes) that represent intermediary layers created only to make the grammar more readable.
-The intermediary layers are created by authors of the grammars in order to make them more readable and more easily (better) maintainable.
-We illustrate this problem (and explain it) on the behavior of auto-completion in MPS.
-Imagine there is a freshly inserted \mpsconcept{Element{\_}1} node (representing the full XML element as is stated in the first alternative of the \antlrparserrule{element} parser rule).
-Now we would like to insert another nested XML element inside.
-MPS auto-completion mechanism would give us four options that are captured in the left part of Figure~\ref{fig:layer_problem}.
-This is because there exist four nodes (objects) that implement the \mpsinterface{IContent} interface as per each alternative of the rule.
+Now we can illustrate the layer problem on the behavior of auto-completion in MPS.
+Suppose that a user is creating a new document in the SimpleXML language, just inserted a fresh node of the type \mpsconcept{Element{\_1}} (see above), and would like to insert another nested XML element inside.
+The auto-completion mechanism of MPS offers four options that are displayed in the left part of Figure~\ref{fig:LAYERPROBLEM}.
+Each option represents one of the MPS concepts that implement the interface \mpsinterface{IContent}, and therefore also one alternative of the \antlrparserrule{content} rule.
 
@@
 	\caption{Layer problem in auto-completion}
-	\label{fig:layer_problem}
+	\label{fig:LAYERPROBLEM}
 \end{figure*}
 
-However, in order to correctly insert another nested element inside, a user would have to first insert the \mpsconcept{Content{\_}2} node inside \mpsconcept{Element{\_}1} that has an \mpsinterface{IElement} child inside (and nothing else).
-Then, as the second step, a user would invoke/trigger/call for the auto-complete again and insert either \mpsconcept{Element{\_}1} or \mpsconcept{Element{\_}2} inside \mpsconcept{Content{\_}2}.
-This mean that a user would have to go through two steps, and in the first stepe either correctly guess (or remember the grammar rule's alternative order), which option/item offered by the auto-complete menu to select.
+However, in order to correctly insert another nested element, a user has to perform two steps:
+\begin{enumerate}
+	\item Insert an object (node) of the type \mpsconcept{Content{\_}2} inside the \mpsconcept{Element{\_}1} node.
+The \mpsconcept{Content{\_}2} object has only a single child node of the interface type \mpsinterface{IElement}.
+	\item Then, the user must trigger the auto-complete again and insert either an \mpsconcept{Element{\_}1} node or an \mpsconcept{Element{\_}2} node into the \mpsconcept{Content{\_}2} node.
+See the right part of Figure~\ref{fig:LAYERPROBLEM}.
+\end{enumerate}
+The main difficulty here is that, in the first step, the user has to (i) either correctly guess what option offered by the auto-complete menu to select, or (2) to remember the order of alternatives in the grammar rule.
 
-Similarly, if one would decide to replace the nested \mpsconcept{Element{\_}1} node with, let's say, an XML comment (a \mpsconcept{Comment} node), we need to delete both intermediary layers before we get back to the original \mpsconcept{Content{\_}X} selection (crossroads).
-The user cannot really see what is happening, since the intermediary level has no appearance or indicator.
-This may lead to confusion on the user's part.
+Similarly, if the user would like to replace a nested \mpsconcept{Element{\_}1} node with, for example, an XML comment (represented by the \mpsconcept{Comment} node), then both intermediary layers have to be deleted before she gets back to the original selection among the concepts \mpsconcept{Content{\_}1}, $\ldots$, \mpsconcept{Content{\_}4}.
 
-The layer problem is solved/addressed by the second approach, which we describe in the next subsection.
+Intermediary layers have no visual appearance, and therefore it is very difficult for users to see what is actually happening and they may get confused very easily.
+The layer problem is addressed by the shortcut approach, which we describe in the next subsection.
+
+% TODO continue from here
 
@@
 \label{sect:SHORTCUT}
+
+\todo{Shortcut approach differs in the way references to children (parser rules) are created.}
+\todo{The main/key difference between these two lies in the way we create children fields for parser rules and in the way we are going to link them together using interface concepts.}
 
@@
 
-% TODO pouzit standardni environment pro algoritmy
 \begin{figure*}[ht]
@@
 
-\subsection{Editor}
+\subsection{Phase 3: Editor}
 
@@
 
-\subsection{Text Generation}
+\subsection{Phase 4: Text Generation}
 

### 2016-10-27 chapters/language_definition.tex
shortcuts - in progress

@@
 
-% TODO continue from here
-
 \subsubsection{The Shortcut Approach}
 \label{sect:SHORTCUT}
+
+The key idea of this approach is to avoid all the intermediary layers (nodes) in the syntax tree by considering just the leaf nodes.
+
+% TODO put here the example from below
+
+When given a particular AST node $n$ (or a rule of the grammar) as input, the parser tree (AST) built in the earlier phases is systematically traversed in order to identify each AST node that may appear at the end of a derivation chain from the input node $n$.
+
+% TODO continue from here
 
@@
 
-The key idea of this approach is to identify rules and nodes that are at the end of the (derivation, inheritance) chain, and offer them directly to the user through the auto-completion menu.
+The key idea of this approach is to identify rules and nodes that are at the end of the (derivation, inheritance) chain
+
 We call such \emph{end rules} and \emph{end nodes}, respectively, because they cannot transparently break into more rules/nodes.
+
+End nodes and rules are then directly offered to the user through the auto-completion menu.
 

### 2016-10-27 chapters/language_definition.tex
little more progress

@@
 
-The key idea of this approach is to avoid all the intermediary layers (nodes) in the syntax tree by considering just the leaf nodes.
+The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree and consider just the leaf nodes.
 
@@
 % TODO continue from here
-
-\todo{Shortcut approach differs in the way references to children (parser rules) are created.}
-\todo{The main/key difference between these two lies in the way we create children fields for parser rules and in the way we are going to link them together using interface concepts.}
-
-The key idea of this approach is to identify rules and nodes that are at the end of the (derivation, inheritance) chain
-
-We call such \emph{end rules} and \emph{end nodes}, respectively, because they cannot transparently break into more rules/nodes.
-
-End nodes and rules are then directly offered to the user through the auto-completion menu.
 
@@
 The end nodes are highlighted using the bold font.
+
+We call the leaf nodes as \emph{end nodes} and \emph{end rules}, respectively, because they cannot transparently break into more rules/nodes.
+
+End nodes and rules are then directly offered to the user through the auto-completion menu.
 

### 2016-10-28 chapters/language_definition.tex
structure - done

--- 2016-10-26 chapters/language_definition.tex
@@
 
-% TODO continue from here
-
 \subsubsection{The Shortcut Approach}
@@
 
-\todo{Shortcut approach differs in the way references to children (parser rules) are created.}
-\todo{The main/key difference between these two lies in the way we create children fields for parser rules and in the way we are going to link them together using interface concepts.}
-
-The key idea of this approach is to identify rules and nodes that are at the end of the (derivation, inheritance) chain, and offer them directly to the user through the auto-completion menu.
-We call such \emph{end rules} and \emph{end nodes}, respectively, because they cannot transparently break into more rules/nodes.
+The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree and consider just the leaf nodes, e.g., to be directly offered to the user through the auto-completion menu.
+Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, which is given in Figure~\ref{fig:CONTENTRULE} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the leaf nodes that are highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.
 
@@
 \end{framed}
-\caption{Content parser rule with children}
-\label{fig:contentrulewithchildren}
+\caption{The parser rule \antlrparserrule{content} with other rules that determine the relevant fragment of the syntax hierarchy}
+\label{fig:CONTENTRULE}
 \end{figure*}
 
-For example, the \antlrparserrule{content} rule, given in Figure~\ref{fig:contentrulewithchildren} together with its child rules, can ultimately expand into following nodes:
+\begin{figure}[ht]
+\begin{framed}
+\textbf{Content{\_}1} (TEXT) \\
+\ \ \ Content{\_}2 $\rightarrow$ \textbf{Element{\_}1} \\
+\ \ \ Content{\_}2 $\rightarrow$ \textbf{Element{\_}2} \\
+\ \ \ Content{\_}3 $\rightarrow$ \textbf{Comment} \\
+\ \ \ \textbf{Content{\_}4} (CDATA)
+\end{framed}
+\caption{Leaf nodes of the parser tree fragment that has the \antlrparserrule{content} rule as its root}
+\label{fig:CONTENTEXPAND}
+\end{figure}
 
-\begin{itemize}
-	\itemsep0em
-	\item \textbf{Content{\_}1} (TEXT)
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}1}
-	\item Content{\_}2 $\rightarrow$ \textbf{Element{\_}2}
-	\item Content{\_}3 $\rightarrow$ \textbf{Comment}
-	\item \textbf{Content{\_}4} (CDATA)
-\end{itemize}
-
-The end nodes are highlighted using the bold font.
-
-Figure XY shows the algorithm (in pseudocode) that finds all end rules for a given parser rule.
+For the input that consists of a particular AST node $N$ and the grammar rule $R$ that expands $N$, the procedure implementing the shortcut approach systematically traverses the parser tree (AST) built in the earlier phases in order to identify each AST node that may appear at the end of some derivation chain starting by the rule $R$ from the node $N$.
+Such nodes cannot expand further based on any rule in the grammar, and for that reason we call them \emph{end nodes}.
 
 \begin{figure*}[ht]
-\centering
 \begin{framed}
 \begin{alltt}
-	FindPathsToEndNodes($R$):
-	  Define $L$ as an empty list of list of nodes
-	  Return FindPathsToEndNodes($R$, $L$)
-
-	FindPathsToEndNodes($R$, $L$):
-	  Define $Q$ as list of list of nodes
-	  For each alternative $A$ of rule $R$:
-	    $L1$ = Clone($L$)
-	    If $A$ is a parser rule with only one element $E$:
-	      Let $I$ be interface/concept representing rule $E$
-	      $L1$.Add($I$)
-	      $P$ = FindPathsToEndNodes($E$, $L1$)
-	      $Q$ = Merge($Q$, $P$)
-	    Else
-	      $L1$.Add($R$)
-	      Let $P$ be concept representing $A$
-	      $L1$.Add($P$)
-	      $Q$.Add($L1$)
-	  Return $Q$
+ 1 FindAllPathsToEndNodes(\textit{R}):
+ 2   \textit{CurPath} = empty list of rules and nodes
+ 3   \textbf{return} FindPaths(\textit{R}, \textit{CurPath})
+ 4
+ 5 FindPaths(\textit{R}, \textit{CurPath}):
+ 6  \textit{Paths} = empty list
+ 7  \textbf{for each} alternative \textit{A} \textbf{in} rule \textit{R}:
+ 8    \textit{NewCurPath} = Clone(\textit{CurPath})
+ 9    \textbf{if} \textit{A} contains only a single element \textit{E}:
+10      \textit{NewCurPath}.Add(\(N\sb{E}\)) \textbf{where} \(N\sb{E}\) is the node representing \textit{E}
+11      \textit{P} = FindPaths(\(R\sb{E}\), \textit{NewCurPath}) \textbf{where} \(R\sb{E}\) is the rule that expands \textit{E}
+12      \textit{Paths} = Merge(\textit{Paths}, \textit{P})
+13    \textbf{else}:
+14      \textit{NewCurPath}.Add(\textit{R})
+15      \textit{NewCurPath}.Add(\(N\sb{A}\)) \textbf{where} \(N\sb{A}\) is the node representing \textit{A}
+16      \textit{Paths}.Add(\textit{NewCurPath})
+17  \textbf{return} \textit{Paths}
 \end{alltt}
 \end{framed}
-\caption{Algorithm for finding end rules}
-\label{fig:shortcut_algorithm}
+\caption{Algorithm to find all paths to end nodes for a parser rule}
+\label{fig:SHORTCUTALG}
 \end{figure*}
 
-To find end rules for each parser rule, the algorithm recursively scans through the parser tree that was built before.
-For each parser rule, it tries to find paths leading to some end rule through its alternatives:
+Figure~\ref{fig:SHORTCUTALG} shows the algorithm that finds all paths to some end node from a given parser rule $R$.
+The algorithm is based on recursive traversal of the parser tree.
+At each level of recursion, it gathers all paths that lead from the current parser rule $R$ to some end node through its alternatives (line~7).
+Two cases may occur:
 \begin{itemize}
-	\item Whenever it finds an alternative that contains only one element, and this element is a reference to another parser rule, it has found an intermediary level that can be transparently hidden from the user of the language.
-		The process (algorithm) continues by recursively processing alternatives of this "level" rule (since we are not at the end of the chain yet).
-	\item Otherwise, an end rule was found (recursion stops here).
+	\item When a particular alternative $A$ of the rule $R$ contains only a single element $E$, and the element is a reference to another parser rule (line 9), the alternative $A$ is an intermediary layer that can be transparently hidden from the user of the MPS language.
+		  A run of the algorithm continues, at line 11, by recursively processing alternatives of the rule corresponding to $E$, which lies at the next level of the parser tree.
+	\item Otherwise, an end node of a derivation chain was found and the recursion stops (lines 13-16).
 \end{itemize}
+By appending the node corresponding to the current alternative (lines 10 and 15) and the rule that leads to the alternative (line 14), the algorithm creates a full path that contains the target end node as the last element of the chain.
 
-By appending the rule that is leading to current element (line 12) and then appending that alternative's element itself (line 14), we get a path that contains the full path and the target end rule as the last element of the chain.
-The result of this algorithm, for example for the \antlrparserrule{content} rule, equals the listing of the five paths mentioned above.
+We use the name \emph{shortcut approach} for this algorithm, because the paths collected by the algorithm provide shortcuts from the given rule to end nodes, by the virtue of hiding all intermediary layers.
+For example, the result of this algorithm for the \antlrparserrule{content} rule (Figure~\ref{fig:CONTENTRULE}) is the list of five items already shown in Figure~\ref{fig:CONTENTEXPAND}.
 
-We do/repeat this for all parser rules and for each rule we get a list of paths that lead from that particular rule to an end node.
-We call these paths as \textbf{shortcuts}, because they provide a shortcut from the rule to the end of the chain.
-Now that we have the shortcuts, we will discuss several ways how to use the shortcuts (them) to improve the MPS language under construction (how to make the language better).
+The primary use case for shortcuts is to generate options for the auto-completion menus, such that only the end nodes are offered.
+Shortcuts have to be considered not only when nodes are inserted, but also when they are deleted.
+In each case, the whole chain including possibly multiple intermediary AST nodes (up to the end node) must be added, respectively deleted.
+When the user wants to delete some end node from the AST of a program or document written in the MPS language, the effect of an insertion must be fully reversed.
 
-The primary use of shortcuts is to generate options for the auto-completion menus such that only the end nodes are offered.
-This can be technically done in two ways --- (1) through custom auto-complete menus and deletion handlers or (2) by defining a special interface for each rule such that the interface is implemented only by end nodes for the rule.
-Shortcuts have to be considered when inserting nodes but also when deleting them, i.e.
-the whole chain (shortcut path) consisting of possibly multiple intermediary AST nodes (levels) up to the end node must be added, respectively deleted.
-The effect of insertion must be fully reversed when the user wants to delete some end node (the given end node).
+% TODO continue from here
 

### 2016-10-29 chapters/conclusion.tex
structure - in progress
editor - done

@@
 \section{Conclusion and Future Work}
+
+\todo{Highlight again:} The challenges and problems discussed in this paper, and our solution, are not specific to MPS --- they are common for everyone tackling a similar problem (in the field of DSL and language workbenches).
 
@@
 
-Automated generation of the editor aspect, probably using some kind of machine learning.
+Besides using some kind of automated (machine) learning to generate the Editor aspect, we would also like to do XYZ.
 Generating nicely designed and user-friendly editors, maybe using artificial intelligence (machine learning) that would evaluate (analyze) existing code in plain text and based on that create the editors.

### 2016-10-29 chapters/evaluation.tex
structure - in progress
editor - done

@@
 }
+
+\todo{Discuss where INGRID helps the most.}
+
+Automated generation of the Structure aspect is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
 
@@
 We have also tried to use the INGRID tool/plugin to import the JavaScript language, which is exactly the type of a complicated general purpose language that might be of interest to other people.
+We have confirmed this assumption (that code layout in the Editor aspect can be adjusted very fast, from the basic default single-row layout) by importing the JavaScript language [12] and manually adjusting all editors, that needed it, in a less than hour time.
+It takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language~\cite{ref:javascript} and manually adjusting all editors that needed it, all that in a less than hour time.
 There are already some projects, where a manual port of JavaScript is being done.

### 2016-10-29 chapters/language_definition.tex
editor - done

@@
 
-% TODO continue from here
-
 \subsection{Phase 3: Editor}
 
-\todo{Definition of projectional editors is only partially automated. We do a lot of mundane and time-consuming work connected to creating projectional editors.}
+Having the complete structure of the new MPS language, the next phase is to define the visual representation of all concepts (AST nodes) in the projectional editor.
+As we said in Section~\ref{sect:MPS}, MPS uses a cellular system that enables the language developer to arrange the children and properties of an AST node in a table-like manner.
+Cells of different types are supported by MPS --- for storing property values, references to child nodes, and keywords (string literals), and also cells that influence layout (e.g., indentation).
+The specific goal of this phase is to create the Editor aspect for each language concept (element), such that all attributes of the concept --- name, properties, children --- are projected using the respective cell types.
 
-When all types of AST nodes in the language are created, the next step is to define their visual representation in the projectional editor.
-As stated before (in Section X (about JetBrains MPS, part of background)), MPS uses a cellular system that allows placing node's properties and children into a table-like arrangement.
-Cells of different types are available --- for storing property values, references to child nodes, and keywords (string literals), and also cells that influence layout (e.g., indentation).
+The main problem that we had to address is the absence of information about the code layout and whitespace in the ANTLR grammar of an input language.
+Rules forming the grammar only tell what the syntax tree looks like and how the program code is decomposed into AST nodes.
+Here we present a solution that is only partially automated for reasons explained below.
 
-The purpose (goals) of this (third) step is to create the editor aspect for each language element.
-It should project all attributes of the element --- name, properties, children, and so on --- using the respective cell types.
+We experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results.
+However, we observed that the most tedious and error-prone step in the manual definition of the Editor aspect is the creation of cells for all literals (keywords), properties, children, and other fields of a given concept that should appear in its visual representation.
+This step can be very easily automated.
 
-The main problems/challenges in the process of definition of projectional editors, also/again related to input grammars, are code layout is whitespace (editor aspect).
-As we already said in the introduction, one of the biggest challenges/problems that arise here is that the grammar description of a language (the description of a language in the form of a grammar) does not hold any information about the code layout.
-The main challenge (problem) here is that an input grammar (ANTLR) does not provide/give any help/aid/information regarding the code layout.
-The structural rules in the grammar only tell us, what the syntax tree looks like and how the code is broken into nodes.
-In particular, the grammar says nothing about indentation, line breaks, and another formatting.
+Our solution that we implemented in the current version of the INGRID method is to create all the cells and place them in a single row.
+The resulting basic layout is illustrated on the example of the \mpsconcept{Element{\_}1} concept in the upper left corner of Figure~\ref{fig:EDITORADJUST}.
+Further adjustments of the layout, such as indentation and line breaks, can be done very efficiently in the MPS IDE.
+The bottom right part of Figure~\ref{fig:EDITORADJUST} shows a fully customized layout for the \mpsconcept{Element{\_}1} concept.
 
-We tackled this problem by designing an automated learning approach, which generates the layout using several heuristics.
-We present our solution to this problem. It involves some intermediary step whose purpose is to generate the necessary information.
-Our solution is semi-automated (using some heuristics) --- user's help in an interactive manner is needed for final touches (polishing).
-It works as follows.
-
-As input for the algorithm (learning approach), the user has to supply to a set of valid source files written in the imported language, together with the grammar file.
-These four steps/operations are then performed.
-\begin{enumerate}
-	\item Automated generation of an ANTLR parser for the input language using the ANTLR library.
-	\item Modify/extend the code of the generated parser in order to record some additional information.
-	\item Compile the parser into an executable form.
-	\item Use the parser to parse the supplied source code files and extract information about the layout of the code during this process.
-\end{enumerate}
-This way, the imported/created MPS language would inherit the code style/layout directly from the user, i.e. it would learn directly from the code.
-In step 2, the parser is modified such that it remembers/stores the line number and column for each parsed token.
-This is very efficient and it enables to create a nice layout.
-From the line number, the algorithm can easily determine whether there is a line break between some tokens (statements, language elements).
-The column number can be then used to detect indentation.
-However, this approach is still heuristic and may give imprecise (suboptimal) results --- mainly because it is not always possible to precisely match language elements (in the MPS structure) with all tokens that belong to it.
-Note also that quality of the extracted information always depends on the amount of the source code files provided (available, supplied) and on how representative the set of source code files actually is.
-When our algorithm cannot determine some custom layout, it just creates all the cells (content) of the editor aspect and arranges them in a single row.
-Figure~\ref{fig:editor_adjustment} XY shows the difference between the default single-row layout (in the upper left corner) and the fully customized layout (in the bottom right corner) for the \mpsconcept{Element{\_}1} node.
-
-% TODO smazat cervenou sipku z toho obrazku a mozna rozdelit na dve casti (vlozit tam caru): default single-row code layout, optimal fully customized layout
 \begin{figure*}[ht]
 	\centering
-	\includegraphics[scale=0.5]{./images/editor_adjustment.png}
-	\caption{Projectional editor adjustment of the Element{\_}1 node}
-	\label{fig:editor_adjustment}
+	\includegraphics[scale=0.55]{./images/editor_adjustment.png}
+	\caption{Editor aspect of the \mpsconcept{Element{\_}1} concept}
+	\label{fig:EDITORADJUST}
 \end{figure*}
 
-In practice, however, the user still has to tweak the editor aspect --- to adjust the code layout --- because the heuristics are not 100\% reliable.
-But it takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language~\cite{ref:javascript} and manually adjusting all editors that needed it, all that in a less than hour time.
-More details are provided later in Section X (examples) --- there we describe this issue further.
+Based on our experience, it takes a very short time to manually adjust the layout into a form much better than any fully automated heuristic could achieve.
+We expect that a typical user of the INGRID method will be able to use the projectional editor in MPS quite efficiently.
+More details are provided in Section~\ref{sect:EVAL}, including our experience with several mainstream programming languages.
+
+We plan to work on a more automated approach to the definition of Editor aspects in the future, probably using some techniques of machine learning.
+The key idea would be to derive the code layout from a set of valid input source files.
+We believe that the user would have to adjust the editor manually a little bit even in this case, because the results of a machine learning-based approach would not be perfect.
+
+% TODO continue from here
 

### 2016-10-29 chapters/language_definition.tex
textgen - in progress

@@
 
-% TODO continue from here
-
 \subsection{Phase 4: Text Generation}
 
-The final/last, fourth step, is to create the TextGen aspect for each AST node.
-The purpose of the TextGen aspect is to define, for a language element (AST node), what code in the BaseLanguage will be generated out of it, once we want to compile and run our MPS code/program.
-This is needed to allow/enable users to generate plain-text source code representation out of the AST they built/created inside MPS.
-Again, a TextGen aspect must be created for each language element.
-Basically, it is a single method in the BaseLanguage (or in Java?) for each language element.
-The method generates a string (plain-text) representation of the element, appends the string to the output stream, and then calls TextGen methods on all children of the element.
-
-A very basic example of BaseLanguage code that can be used as the TextGen aspect of an \mpsconcept{Element{\_}1} node, which represents the full XML tag with content, is shown in Figure~\ref{fig:textgen_example}.
-It appends all literals, properties, and children in the same order as they appear in the grammar definition.
+The purpose of the last phase of the MPS language construction is to generate the TextGen aspect for each concept.
+We use the \mpsconcept{Element{\_}1} concept again for illustration.
+Figure~\ref{fig:TEXTGENEX} shows a very basic example of BaseLanguage code that can be used as its TextGen aspect.
+The code appends all literals, properties, and children of \mpsconcept{Element{\_}1} to the output buffer, and does that in the same order as they appear in the corresponding grammar rule.
 
 \begin{figure}[ht]
-	\centering
-	\includegraphics[width=90mm]{./images/textgen_example.png}
-	\caption{Example of a~TextGen aspect}
-	\label{fig:textgen_example}
+\begin{alltt}
+\small
+\mpstgkeyword{text gen component for concept} \mpstgtarget{Element{\_}1} \{
+  \mpstgparam{(context, buffer, node)->void} \{
+    \mpstgaction{append} \{\mpstgliteral{<}\};
+    \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{1}\};
+    \mpstgaction{append} \{\ \};
+    \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Attribute}}{\_}\textit{1}\};
+    \mpstgaction{append} \{\mpstgliteral{>}\};
+    \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2}\};
+    \mpstgaction{append} \{\mpstgliteral{</}\};
+    \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{2}\};
+    \mpstgaction{append} \{\mpstgliteral{>}\};
+  \}
+\}
+\end{alltt}
+\caption{Basic TextGen aspect for the \mpsconcept{Element{\_}1} concept}
+\label{fig:TEXTGENEX}
 \end{figure}
+
+% TODO continue from here
 

### 2016-10-30 chapters/language_definition.tex
textgen - done

@@
 
-We plan to work on a more automated approach to the definition of Editor aspects in the future, probably using some techniques of machine learning.
+We have found that the combination of two steps, (1) automated placing of cells into a single row and (2) subsequent manual adjustment of the layout, is a very fast and efficient way of creating a nice code layout.
+Nevertheless, we plan to work on a more automated approach to the definition of Editor aspects in the future, probably using some techniques of machine learning.
 The key idea would be to derive the code layout from a set of valid input source files.
@@
 We use the \mpsconcept{Element{\_}1} concept again for illustration.
-Figure~\ref{fig:TEXTGENEX} shows a very basic example of BaseLanguage code that can be used as its TextGen aspect.
+Figure~\ref{fig:TEXTGENBASIC} shows a very basic example of BaseLanguage code that can be used as its TextGen aspect.
 The code appends all literals, properties, and children of \mpsconcept{Element{\_}1} to the output buffer, and does that in the same order as they appear in the corresponding grammar rule.
@@
 \caption{Basic TextGen aspect for the \mpsconcept{Element{\_}1} concept}
-\label{fig:TEXTGENEX}
+\label{fig:TEXTGENBASIC}
 \end{figure}
 
-% TODO continue from here
+Like in the case of a projectional editor, the main challenge associated with TextGen is to produce valid code with a reasonable layout.
+An implementation of the TextGen aspect must determine properly where to put line breaks, indentation, and other whitespace characters.
+For example, in the case of the concept representing an XML element, there must be a space between the element's name and the first attribute, while it is not required between the opening bracket \antlrliteral{\textless} and the name (Figure~\ref{fig:TEXTGENBASIC}).
 
-Like in the case of editor, we discovered some interesting challenges related to the code layout, whitespaces, and more (so on).
-We describe our solution to the challenges here in this section.
+Our INGRID method targets mostly text-based languages, for whose concepts the visual representation in a projectional editor must be almost equivalent to their plain-text representation.
+An obvious choice would therefore be to use the same approach for Editor and TextGen.
+Nevertheless, since the Editor aspect has to be adjusted manually, we decided to use a different approach for TextGen.
+We designed a procedure based on a simple fully automated heuristic that provides surprisingly good results.
 
-The reader might have noticed that there is no whitespace handling inside of our SimpleXML grammar (Section XY, running example).
-Grammars in the ANTLRv4 notation are usually written in such a manner that whitespace characters are skipped (ignored).
-However/Nevertheless, TextGen aspect (generator) should/must know between which children there must be a whitespace and where it is forbidden (i.e., where to put some whitespace characters) so that the produced code is a valid one.
-For example, in the case of the \antlrparserrule{element} rule, TextGen must determine that there should be a space in between \antlrparserrule{attribute} and \antlrlexerrule{Name}, while it is not required between \antlrliteral{\textless} and \antlrlexerrule{Name}.
-In general, it is sufficient to put just a single space character at places/locations where there has to be some whitespace in order to make the code valid (syntactically correct).
-
-Layout of the generated code is also important. For example, the user may want to produce more readable code by adding indentation or line breaks.
-
-In fact, the problem of TextGen layout is quite similar to the one we have seen with the projectional editor, described in Section XY.
-Since we are dealing mostly with text-based languages, their representation in projectional editor must be almost the same as their expected plain-text representation (output).
-Therefore, one could use the same approach for editor and TextGen --- adding line breaks and indentation.
-Once we would have some information about the layout for generating better projectional editor, we could leverage the same information and use it to improve TextGen.
-
-Nevertheless, in our current approach, the editor aspects have to be adjusted manually, after the import is done, because for now it is the fastest and the most efficient way how to deal with the problem of customizing the editor to capture a nice layout.
-Therefore, we used a different approach for TextGen --- a really simple heuristics which gives surprisingly good results.
-In the first step, a very basic TextGen is created that inserts spaces in between every two tokens of the language element (AST node).
-Then, as the second step, spaces are eliminated (restricted) from positions where they are not really needed.
-We describe all the cases here:
-
-\begin{enumerate}
-	\item Whenever there is a non-alphabetical literal,	appearing as a plain string token defined in the grammar, that might get recognized by the parser safely without the need for a whitespace separator around, then we omit spaces around it.
-		For example, consider \textbf{'\textless'} in SimpleXML.
-	\item Another group are strings inserted by the user of the language, whose form is constrained by a regular expression.
-		For example, an XML tag's name belongs to this group.
-		They are usually values of properties defined in structure aspects.
-		A space must be preserved only when a neighboring literal may end with an alphabetical character. Otherwise, we can omit the space.
-		In practice, this will eliminate redundant spaces inside of quotes, next to semicolons, and around brackets, but, on the other hand, it will separate language keywords (function, var, in, etc.) from other content by keeping the space character in between them.
-	\item Spaces are omitted also in the case of child nodes that are not present (e.g., when they are optional), and in the case of empty lists of child nodes, so that spaces do not accumulate.
-	\item On the other hand, when two child elements are next to each other, we always insert space.
-	\item Sequences of children are separated with space or with a line break.
-\end{enumerate}
-These few simple heuristics work surprisingly well and give us some very nice results.
-We discuss our experience with XML, JSON, and JavaScript in Section XY (Evaluation).
-
-Figure~\ref{fig:textgen_final} shows an example of a TextGen aspect for the full SimpleXML element.
+The procedure creates the TextGen aspect for a given language concept (AST node) in two steps.
+First, it generates a basic variant that inserts spaces in between every two tokens of the textual representation of the concept.
+In the second step, which is the core of our heuristic, spaces are eliminated from places where they are not really needed.
+The main criterion is whether the generated plain-text output can be safely processed by a parser of the original language (using the ANTLR grammar).
+We discuss all the cases here:
+\begin{itemize}
+	\item When there is a non-alphabetical literal that is used as a token in the grammar, and that might get recognized by the parser without the need for whitespace separators around it, then we can omit the spaces.
+An example of such literal is '\textless' in SimpleXML.
+	\item In the case of an arbitrary string, the whitespace may be omitted when the adjacent literal ends, respectively begins, with a non-alphabetical character.
+		This applies especially to the values of properties defined in Structure aspects, such as the name of an XML element.
+		Based on this heuristic, redundant spaces inside of quotes will be eliminated, as well as spaces next to semicolons and around brackets.
+		On the other hand, it will distinguish language keywords (e.g., \code{function}, \code{var}, \code{in}) from other literals by preserving the space character in between them.
+	\item Spaces can be safely omitted also when specific optional child nodes are not present, and in the case of empty lists of child nodes, so that whitespace does not accumulate.
+\end{itemize}
+A space must be alway inserted when two child nodes are next to each other.
+Elements of a sequence of child nodes are separated with space or a line break.
 
 \begin{figure}[ht]
-	\centering
-	\includegraphics[width=95mm]{./images/textgen_final.png}
-	\caption{Generated TextGen aspect for the SimpleXML element}
-	\label{fig:textgen_final}
+\begin{alltt}
+\small
+\mpstgkeyword{text gen component for concept} \mpstgtarget{Element{\_}1} \{
+  \mpstgparam{(context, buffer, node)->void} \{
+    \mpstgaction{append} \{\mpstgliteral{<}\};
+    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{1}.\textbf{isNotEmpty}) \{
+      \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{1}\};
+    \}
+    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Attribute}}{\_}\textit{1}.\textbf{size} > 0) \{
+      \mpstgaction{append} \{\ \};
+      \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Attribute}}{\_}\textit{1} with  \};
+    \}
+    \mpstgaction{append} \{\mpstgliteral{>}\};
+    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2}.\textbf{size} > 0) \{
+      \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2} with  \};
+    \}
+    \mpstgaction{append} \{\mpstgliteral{</}\};
+    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{2}.\textbf{isNotEmpty}) \{
+      \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{2}\};
+    \}
+    \mpstgaction{append} \{\mpstgliteral{>}\};
+  \}
+\}
+\end{alltt}
+\caption{Full TextGen aspect for the \mpsconcept{Element{\_}1} concept}
+\label{fig:TEXTGENFINAL}
 \end{figure}
 
-The automatically generated TextGen code can be adjusted manually very easily, so that it generates nicely indented XML code.
-In the case of the example above (in Figure F), we only need to wrap the \mpsconcept{Content{\_}2} child with indentation and change the sequence separator to a new line character.
-The resulting adjusted aspect (code) is shown below in Figure~\ref{fig:textgen_adjusted}.
+\begin{figure}[ht]
+\begin{alltt}
+\small
+\mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2}.\textbf{size} > 0) \{
+  \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n;}}
+  \mpstgaction{indent buffer;}
+  \mpstgkeyword{with indent} \{
+    \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2} with  \};
+  \}
+  \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n;}}
+\} 
+\end{alltt}
+\caption{Fragment of the TextGen aspect of \mpsconcept{Element{\_}1} with adjusted indentation}
+\label{fig:TEXTGENADJUSTED}
+\end{figure}
 
-\begin{figure}[t!]
-	\centering
-	\includegraphics[width=85mm]{./images/textgen_adjusted.png}
-	\caption{Adjusted indentation inside the TextGen aspect}
-	\label{fig:textgen_adjusted}
-\end{figure}
+Figure~\ref{fig:TEXTGENFINAL} contains the complete, automatically generated, TextGen aspect for the \mpsconcept{Element{\_}1} concept, which represents the full SimpleXML element.
+Such code may be adjusted manually very easily in order to produce nicely indented XML documents.
+We only need to wrap the \mpsconcept{Content{\_}2} child node with indentation and change the sequence separator to a new line character.
+A fragment of the resulting adjusted aspect (code) is in Figure~\ref{fig:TEXTGENADJUSTED}.
+
+% TODO continue from here
 

### 2016-10-30 chapters/language_definition.tex
remarks - first part

@@
 
-% TODO continue from here
-
 \subsection{Remarks about Grammars}
 
-As we have already shown in previous sections, it is very easy to write a grammar in a way that will cause problems during import into MPS (or once it is imported).
-Problems might occur during the creation of any aspect.
-Here we will look at some general problems/issues that grammar import poses (might pose).
-We will talk about these problems with respect to the challenges/obstacles we tried to overcome and which we described in previous sections (dedicated to structure, editor, and texgen, respectively).
-We also show a few examples of additional possible complications.
+During our work on the INGRID method, we have also observed that practical usability of a resulting MPS language depends on the specific manner in which the input ANTLR grammar is defined.
+We discuss several issues and our workarounds in this section.
 
 \paragraph{Adjusting grammars.}
-There are some cases, where altering the input grammar might yield far better MPS language.
-We will show two examples how the usability of the resulting MPS language can be improved.
-Let us look at the definition of an XML attribute in Figure~\ref{fig:xmlattribute}.
+In some cases, a small adjustment of the input grammar before the run of INGRID might yield a better and more useful MPS language.
+Here we illustrate this on two simple examples.
+
+For the first example, we use the definition of an XML attribute in Figure~\ref{fig:XMLATTRIB}, which is taken from the original XML grammar.
+The lexer rule \antlrlexerrule{STRING} actually says that quotes make a part of the attribute's value.
+If the MPS language would faithfully match the grammar, the user would have to always input the leading and trailing quote together with the value in the projectional editor.
 
 \begin{figure}[ht]
-\centering
 \begin{framed}
@@
 \caption{Definition of an XML attribute}
-\label{fig:xmlattribute}
+\label{fig:XMLATTRIB}
 \end{figure}
 
-The original XML grammar has quotes as a part of the value.
-For the resulting MPS language, it would mean that there would be a placeholder for the attribute value that would expect us to input the leading and trailing quote together with the value too each time.
-It would also be marked red (by the syntax checker) unless we enter both quotes inside the value since the regular expression checking for quotes will not match.
-The user might be confused by this and will not be able to tell why his string value is incorrect.
-
-In our SimpleXML language, we adjusted the grammar easily in the manner that we show in Figure~\ref{fig:xmladjustgrammar}.
+In our SimpleXML language, we adjusted the grammar easily in a way that we show in Figure~\ref{fig:XMLADJUST}.
+We turned quotes into literals, ensuring that they will only appear in the projectional editor as string constant cells.
 
 \begin{figure}[ht]
-\centering
 \begin{framed}
@@
 \end{framed}
-\caption{Adjusted grammar of SimpleXML}
-\label{fig:xmladjustgrammar}
+\caption{Adjusted fragment of the SimpleXML grammar}
+\label{fig:XMLADJUST}
 \end{figure}
 
-We turned quotes into literals, thus ensuring that they will only appear in the projectional editor as fixed constant cells.
-We will not have to encapsulate the value in them each time.
-The user will only have to choose, which attribute version he wants to use (single or double quotes).
+The second example relates to a fragment of the JavaScript language, also known as ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}.
+Every statement in JavaScript needs to be terminated by a semicolon, newline, end of the file, or end of the block --- see the Figure~\ref{fig:JAVASCRIPTSTMT}, which contains the corresponding fragment of the ECMAScript grammar.
+Since there are multiple options, a user would have to select one of them for each statement in the editor.
+Technically, every language concept representing a statement would contain one child node of the interface type \mpsinterface{IEos}, which has to be assigned a proper object (i.e., an AST node corresponding to one of the options listed above).
 
-As the second example, we will use the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
-Every statement in JavaScript needs to be either followed by a semicolon, newline, file end or end of the block --- see the Figure~\ref{fig:javascriptstmt}.
-
-\begin{figure*}[ht]
-\centering
+\begin{figure}[ht]
 \begin{framed}
 \begin{alltt}
-	\antlrparserrule{eos} : \antlrlexerrule{SemiColon} | \antlrlexerrule{EOF} | {lineTerminatorAhead()}? 
-	    | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}? ;
-	\textcolor{gray}{// Example reference of the eos rule}
+\small
+\antlrparserrule{eos} : \antlrlexerrule{SemiColon} | \antlrlexerrule{EOF} | {lineBreakAhead()}? 
+    | {{\_}input.LT(1).type() == \antlrlexerrule{CloseBrace}}? ;
+\textcolor{gray}{// example reference to the eos rule}
 \antlrparserrule{breakStmt} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos} ;
@@
 \caption{Statements in JavaScript}
-\label{fig:javascriptstmt}
-\end{figure*}
-
-Because there are multiple options, our import plugin would create a placeholder at the end of every statement.
-Every language element representing a statement would have one child of the \mpsinterface{IEos} interface type.
-This placeholder needs to be manually filled in for each statement.
-
-Since the projectional editor has much bigger power over the form of the code, we (a user) might want to have each statement on a separate line.
-Since we can differentiate between statements on the AST level, we do not need an explicit separator between them.
-This means that we might want to simplify the language and leave the semicolon out, or leave it just as a constant fixed part of the projectional editor, but not as something the user must explicitly fill in.
-Then we can just put each statement on a separate line as it is usual for JavaScript code, but we do not need the semicolon anymore.
-This small adjustment is very quick when done inside the grammar.
-We just change the \antlrparserrule{eos} rule to the form shown in Figure~\ref{fig:eosrule}.
-
-\begin{figure}[ht]
-\centering
-\begin{framed}
-\begin{alltt}
-	\antlrparserrule{eos} : \antlrliteral{;} ;
-\end{alltt}
-\end{framed}
-\caption{eos parser rule}
-\label{fig:eosrule}
+\label{fig:JAVASCRIPTSTMT}
 \end{figure}
 
-This way, we can change the grammar in a way that it will not describe the same ECMAScript language as before, but will definitely make our MPS language more usable.
-TextGen aspect may be defined in such a way that puts the semicolon back (after each statement) in the generated plain-text representation.
-This would be very important especially when JavaScript is available as another base language (which is actually a plan (future work) of the JetBrains company).
+Since MPS can differentiate between statements on the AST level, there is no need for an explicit separator.
+For example, we can put each statement on a distinct line, as is usual for JavaScript code, and keep just the semicolon as a fixed literal in the projectional editor.
+The \antlrparserrule{eos} rule must be changed to this form: \antlrparserrule{eos} : \antlrliteral{;}.
+A small adjustment of the grammar, like this one, is a very quick solution and definitely makes the generated language more usable in MPS.
+On the other hand, this adjustment of the original ANTLR grammar, performed just for the purpose of creating the MPS language, changes the grammar in such a way that it does not precisely describe the standard JavaScript language.
+To ensure compatibility with JavaScript, the TextGen aspect in the MPS language should put the semicolon after each statement in the plain-text representation of a program code.
 
-Above, we have shown that adjusting the input grammar might be a very fast mean of tweaking/improving the usability of generated MPS language, and sometimes it may be the only proper solution in some complex situations.
-In general, the question we can ask ourselves is whether one aims for a full precise/faithful port of the input language or for an MPS alternative.
+% TODO continue from here
 

### 2016-10-30 chapters/language_definition.tex
remarks - done

@@
 
-% TODO continue from here
+\paragraph{Breaking original grammars and parsers.}
+We also want to point out a general problem with grammar adjustments that we encountered, and which all potential users of the INGRID method should be aware of.
+The original ANTLR grammar for an input language can be very easily changed in a way that, at first, seems harmless and valid inside MPS, but the parser generated out of the adjusted grammar stops accepting the original language.
+In particular, adjustments needed to improve the resulting MPS language can break down the parser.
 
-\todo{Nahradit slovo "break" v nasledujicich odstavcich (v kontextu "break the parser") za vhodnejsi slovo ktere pusobi vic odborne.}
-
-\paragraph{Breaking grammars and parsers.}
-However, we have found that there is one big problem with grammar adjustment that we would like to point out.
-The problem is that it is very easy to change the input ANTLR grammar in a way that breaks the ANTLR parser generated out of it.
-By breaking we mean that it stops parsing the original language.
-We illustrate the problem on our SimpleXML language.
-
-As stated before, when creating the SimpleXML grammar, we have started off with the original XML grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and did some adjustments to it.
-We ended up with a grammar that can be processed by our method/plugin and creates a nice usable MPS language.
-Nevertheless, we noticed that even though the imported language behaves well enough and mimics the XML language quite nicely, the ANTLR parser generated out of this grammar no longer parses XML successfully.
-Some changes we have made, such as the attribute adjustment, broke the grammar down.
-More precisely, our changes improved the MPS language while breaking down the parser, and we have not even noticed it because, from the perspective of MPS, the imported language still corresponds to XML.
-
-What we are trying to say (show here) is that it is very easy to perform a grammar adjustment that will, at first, seem harmless and valid inside MPS, but will break the parser.
-The problem is that the user performing this change probably will not be aware of breaking it.
-
-The main (underlying) cause of this problem is the way the ANTLR parser is implemented and the quite different purpose we are using the grammar for in our project/tool/method.
-There are many ways how various parsers deal with, for example, token matching.
-For instance, ANTLR introduces so called \emph{greedy} and \emph{non-greedy} operators.
-The greedy way, in which ANTLR matches input on defined tokens and prioritizes their selection, makes some rules very dangerous.
-When a rule matches a wide range of input, it might happen during parsing that it is prioritized over other rules and swallows a lot more input than the author of it intended.
-Usually, these dangerous rules are bound to some parser context, which makes them behave well.
-
-However, for practical reasons, it is very important that grammar is not broken by our custom adjustments (and that the ANTLR parser of the language works properly), because we might need to automatically generate parsers of the adjusted language.
-Imagine this very real scenario (that will be supported in the future):
-\begin{enumerate}
-	\item We have imported the language inside MPS.
-	\item MPS now knows the structure of the language, and we can code in MPS using this language.
-	\item We would expect MPS to be able to load an existing source code, written in this language, from a text file and import it inside MPS.
-	\item We would like to use MPS to safely edit this code, using all the features of MPS.
-	\item We would also like to export the code in a plain-text form again and save it back to the source file.
-\end{enumerate}
-In order to make this happen, we must be able to generate a correctly-working ANTLR parser out of the adjusted source grammar that we used to (that helped us) import the language.
-We must also be able to match nodes of the AST coming out of the ANTLR parser to elements of the MPS language, and we must build the MPS AST out of the ANTLR AST.
+This problem has two main causes: (1) low-level implementation details of token matching in the ANTLR parser and (2) usage of ANTLR grammars for a different purpose (that was not expected) in our project.
+Since, for practical reasons, it is very important that a parser works even for documents (programs) written according to the original grammar, adjustments have to be performed very carefully by the users of INGRID.

### 2016-10-31 chapters/background.tex
Related projects

@@
 
-An MPS language is created by the PE4MPS tool in two steps.
-First, the PE parser is used to load an input PE file and build its intermediate representation, which has the form of a tree-like structure consisting of Java objects.
-Then, all the concepts (AST nodes) and their aspects are created inside MPS.
-The information about the code layout, extracted from the PE file, is used when generating the projectional editor for every AST node.
-
 The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
+User is forced to input the layout information by extending the grammar manually.
 Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
@@
 
-However, the project has also other important limitations.
-The tool does not generate any parent-child relationships in the Structure aspect, and it completely neglects the Editor and TextGen aspects.
-In particular, it does not support creating of the projectional editor at all.
+The major take away from this project is the idea of an intermediate grammar representation inside of MPS which we have also considered.
+The advantage of this could be the possibility of extending the ANTLRv4 notation using some graphic tools that MPS has to offer. 
+Using the projectional editor, a user could add the layout information in an effective way.
+The disadvantage is the dual implementation that is needed.
+First to transform the ANTLRv4 text file into the MPS grammar language and secondly to transform this grammar into the target MPS language.
 
@@
 Many iterations of this process, involving adjustments of the grammar followed by recreation of the target language, can be performed.
-
-The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLRv4 MPS language into the actual target language.

### 2016-10-31 chapters/background.tex
Rollback related project changes, added comments instead

--- 2016-10-13 chapters/background.tex
@@
-\section{Background and Related Work}
-\label{sect:BACKGRELWORK}
-
-In this section, first we introduce basic features of the MPS platform, and then we discuss other existing projects that aim to add support for general purpose languages into MPS.
-
-\subsection{JetBrains MPS}
-\label{sect:MPS}
-
-JetBrains MPS is a complete language workbench --- an integrated development environment that allows developers to create their own languages and use them to write code.
-The code can then be transformed into a target language, typically a GPL, such as Java or C, and eventually compiled into executable programs.
-
-As we already indicated, MPS differs from typical IDEs in one important aspect --- the projectional editor.
-The developer does not work with the textual representation of the source code directly, but rather with its AST that is the model of the code.
-Basically, when using the MPS projectional editor, programs are created by assembling the tree (AST) out of predefined building blocks of selected languages.
-The definition of a language in MPS dictates, where in the AST can certain elements be placed and how they can be nested inside other elements.
-On the other hand, in traditional IDEs it is the parser that constructs an AST of a program using the language's elements.
-
-The building blocks of MPS models are called nodes.
-Code of any program in MPS is built from nodes, which represent instances of concepts defined in one of the languages that the program declares to be using.
-In the context of MPS, a concept is, in fact, a building block of a language definition.
-We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.
-
-One of the key advantages of projectional editing stems from the separation of abstract and concrete syntax.
-While AST provides a complete and precise representation of the code, the way it is displayed on the screen and the way the user interacts with it are unconstrained.
-The editor can take any visual form and shape.
-The language author can define multiple alternative visualizations and let the developer choose one that fits best the task at hand --- for example, editing, debugging, reviewing, resolving merge conflicts, etc.
-In particular, the visual representations are not bound to be just textual at all.
-For example, blocks corresponding to branches of an \code{if-then-else} statement can be aligned next to each other and displayed with different background colors.
-
-The definition of an MPS concept (language element) consists of several aspects, where each aspect codifies a different part of the AST nodes' behavior.
-The essential aspects are the following: \emph{Structure} (abstract syntax), \emph{Editor} (concrete syntax) and \emph{TextGen}.
-Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
-If, instead of generating text directly, programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
-Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we describe these three below in more detail, and neglect other aspects such as type system and data-flow.
-
-Languages are built from the concepts using techniques known from object-oriented programming --- containment, inheritance, interfaces, and so on.
-Therefore, a definition of a whole language in MPS typically has an object-oriented and hierarchical nature.
-
-\paragraph{Structure.}
-The fundamental aspect of any MPS language is Structure.
-It must be created first for each intended element (concept) of the language.
-Structure specifies core attributes of an MPS concept such as the name, inheritance relationships, possible child concepts (including their types and cardinalities), the implemented interfaces, and references to other AST nodes.
-% other properties (fields of any type that can hold values), etc.
-Figure~\ref{fig:if_statement_structure} shows definition of the Structure aspect for the \code{if-else} statement.
-
-The definition of Structure restricts the type of AST nodes that can appear at a particular place in the tree.
-For example, one can restrict the condition in the \code{if-then-else} statement to be a boolean expression, and the \code{then}-block to be a list of statements.
-
-\begin{figure}[ht]
-\centering
-\begin{alltt}
-\small
-\mpsstkeyword{concept} IfStatement \mpsstkeyword{extends} Statement
-        \mpsstkeyword{implements} IContainsStatementList
-                   IDontSubstituteByDefault
-                   IConditional
-
-  \mpsstkeyword{instance can be root:} false
-  \mpsstkeyword{alias:} \mpsstalias{if}
-  \mpsstkeyword{short description:} \mpsstplaceholder{<no short description>}
-
-  \mpsstkeyword{properties:}
-  \mpsstproperty{forceOneLine}   : boolean
-  \mpsstproperty{forceMultiLine} : boolean
-
-  \mpsstkeyword{children:}
-  \mpsstproperty{condition}        : Expression[\mpsstcardinality{1}]
-  \mpsstproperty{ifFalseStatement} : Statement[\mpsstcardinality{0..1}]
-  \mpsstproperty{ifTrue}           : StatementList[\mpsstcardinality{1}]
-  \mpsstproperty{elsifClauses}     : ElsifClause[\mpsstcardinality{0..n}]
-
-  \mpsstkeyword{references:}
-  \mpsstplaceholder{<< ... >>}
-\end{alltt}
-\caption{Structure aspect of the \code{if-else} statement}
-\label{fig:if_statement_structure}
-\end{figure}
-
-\paragraph{Editor.}
-The Editor aspect is where the language designer specifies what the projectional editor representation of a code fragment (an AST) looks like on the screen and how the user interacts with the code.
-JetBrains have developed a cellular system that enables placing properties and children of a node (concept) into different cells.
-The author usually incorporates all of the node's children, references, and properties inside the representation, so that the future users of the language can insert all values that the node expects.
-Additionally, cells of the editor can be styled using a language similar to CSS.
-The supported visual characteristics include text color and indentation.
-
-Figure~\ref{fig:if_editor_definition} shows an example of what the definition of the Editor aspect for the \code{if-else} statement might look like.
-While here we indicate positions of cell borders on each line by spaces (just for the purpose of illustration), MPS GUI actually uses a graphically much more appealing way of displaying the Editor aspects, which involves vertical and horizontal lines of different colors and also background colors other than white for some cells.
-The symbols \verb|[-| and \verb|-]| represent layout information (cells), which define rules for mutual positioning of the contained cells (vertical, horizontal, indentation).
-
-\begin{figure}[ht]
-\centering
-\begin{alltt}
-\small
-\mpsedannotation{<default>} \mpsedkeyword{editor for concept} \mpsedtarget{IfStatement}
-  \mpsedkeyword{node cell layout:}
-    [-
-      \mpsedkeyword{if} \textbf{(} \% conditions \% \textbf{)} [-
-      \textbf{\{}
-      [- \% ifTrue % -]
-      \textbf{\}}
-    -]
-    ?[- \mpsedkeyword{else} \% ifFalseStatement \% -]
-    -]
-\end{alltt}
-\caption{Editor aspect for the \code{if-else} statement}
-\label{fig:if_editor_definition}
-\end{figure}
-
-\paragraph{BaseLanguage.}
-Another important feature of MPS that we need to describe in more detail is \emph{BaseLanguage}~\cite{ref:BASELANG}.
-It is a clone of Java implemented using the MPS constructs.
-BaseLanguage was developed in the early days of MPS in order to implement MPS itself and also to support the basic set of language-definition DSLs,
-Although BaseLanguage is syntactically almost identical to Java, it is edited in a projectional editor, just like all the languages in MPS.
-The language-definition DSLs, used to define custom languages by their authors, are generated into the BaseLanguage.
-Similarly, all the custom DSLs that are meant to be generated into Java choose BaseLanguage as their generation target.
-The conversion to textual Java is handled by BaseLanguage without any further manual effort, since BaseLanguage has a TextGen aspect defined, which translates BaseLanguage code into textual Java sources.
-
-\paragraph{TextGen.}
-The TextGen aspect specifies how a given AST node will be translated into plain text representation.
-It is typically needed only for the bottom-line base languages.
-DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since these are rarely converted to text directly.
-After TextGen has generated textual sources from an AST, a compiler for the particular GPL is invoked to compile the textual sources into binary code.
-The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes at the right moments.
-MPS calls the corresponding method on the root AST nodes of the given program.
-
-\begin{figure}[ht]
-\centering
-\begin{alltt}
-\small
-\mpstgkeyword{text gen component for concept} \mpstgtarget{IfStatement} \{
-  \mpstgparam{(context, buffer, node)->void} \{
-    \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n;}}
-    \mpstgaction{indent buffer;}
-    \mpstgaction{append} \{\mpstgliteral{if (}\} \$\{\mpstgparam{node}.\mpstgnodeprop{condition}\} \{\mpstgliteral{) \{}\};
-    \mpstgkeyword{with indent} \{
-      \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{ifTrue}\};
-    \}
-    \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n}} \{\mpstgliteral{\}}\} \$list\{\mpstgparam{node}.\mpstgnodeprop{elsifClauses}\};
-    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{ifFalseStatement}.\textbf{isNotNull}) \{
-      \mpstgaction{append} \{ \mpstgliteral{else}\} \$\{\mpstgparam{node}.\mpstgnodeprop{isFalseStatement}\};
-    \}
-  \}
-\}
-\end{alltt}
-\caption{TextGen aspect definition for the \code{if-else} statement}
-\label{fig:if_statement_textgen}
-\end{figure}
-
-TextGen aspect for each AST node (concept of the language) has to be defined using the BaseLanguage.
-Again, we include an example for the \code{if-else} statement (Figure~\ref{fig:if_statement_textgen}).
-
-\subsection{Related Projects}
-
-The BaseLanguage~\cite{ref:BASELANG} is an almost full port of the Java language, extended with MPS-specific features.
-It was created manually by JetBrains, and it is still undergoing changes as Java itself is evolving.
-
-The C language has also been manually tailored for MPS within the mbeddr project~\cite{ref:MBEDDR}.
-
-Besides that, we are aware of three other projects that provide certain support for GPLs in the context of MPS --- namely PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
-We describe their main features and limitations in this section.
-
 \paragraph{PE4MPS.}
@@
 
+\PV{Nasledujici dve vety byly vzaty z popisu INGRID, tak funguje INGRID, ne PE4MPS.}
 An MPS language is created by the PE4MPS tool in two steps.
@@
 
+The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
+\PV{Navrhuji pridat (aby to bylo vice jasne?): User is forced to input the layout information by extending the grammar manually.}
+Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
 Another limitation of PE4MPS is that it does not implement the full ANTLR syntax.
@@
 The tool does not generate any parent-child relationships in the Structure aspect, and it completely neglects the Editor and TextGen aspects.
+\PV{Dalsi veta mozna zbytecna?}
 In particular, it does not support creating of the projectional editor at all.
+
+\PV{Navrhuji pridat neco jako (ale mozna uz o tom mluvime jinde, jen si ted nejsem jisty):
+The major take away from this project is the idea of an intermediate grammar representation inside of MPS which we have also considered.
+The advantage of this could be the possibility of extending the ANTLRv4 notation using some graphic tools that MPS has to offer. 
+Using the projectional editor, a user could add the layout information in an effective way.
+The disadvantage is the dual implementation that is needed.
+First to transform the ANTLRv4 text file into the MPS grammar language and secondly to transform this grammar into the target MPS language.}
 
@@
 In the last step, the target MPS language is derived from the adjusted grammar definition.
+\PV{Navrhuji odsud dolu smazat.}
 Many iterations of this process, involving adjustments of the grammar followed by recreation of the target language, can be performed.
@@
 The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLRv4 MPS language into the actual target language.
+\PV{Smazat az sem.}
+\PV{A pridat: This step has not yet been implemented, because MPS doesn't offer any mechanisms for automatic transformation.
+This means that implementation would require traversing the AST of the grammar and recreating the definition from that.}

### 2016-10-31 chapters/background.tex
reverting delete of large part of background text

--- 2016-10-22 chapters/background.tex
@@
 The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
+\PV{Navrhuji pridat (aby to bylo vice jasne?): User is forced to input the layout information by extending the grammar manually.}
 Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.

### 2016-10-31 chapters/language_definition.tex
Comments to "Creating language definition"

@@
 All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
-Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.
+Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation \PV{Mozna vysvetlit proc (protoze je to dvoji prace) nebo odkazat tam, kde je to vysvetleny?}.
 
@@
 \antlrlexerrule{Name}      : \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
-\textbf{fragment}
 \antlrlexerrule{DIGIT}     : \antlrregex{[0-9]} ;
-\textbf{fragment}
 \antlrlexerrule{NameChar}  : \antlrlexerrule{NameStartChar} | \antlrliteral{-} | \antlrliteral{\_} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
-\textbf{fragment}
 \antlrlexerrule{NameStartChar} : \antlrregex{[:a-zA-Z]} ;
@@
 		A terminal symbol can be encoded as a string value or using a regular expression.
-		The \textbf{fragment} keyword states that the lexer rule just brings more clarity into the notation but it is not visible in the parser output.
 	\item \antlrliteralnoap{\textbf{String literals}}, always written inside of a pair of single quote marks, also represent terminal symbols, but by exact match to a string constant.
@@
 
-The first phase in the process of MPS language construction is parsing of the input grammar that is defined using the ANTLR v4 notation.
-An output of this phase is an AST representing the grammar.
-Nodes of the AST correspond to rules and other grammar elements.
+\PV{Nasledujici odstavec jsem lehce vylepsil, asi je to ted vysvetleno lepe.}
+The first phase in the process of MPS language construction is parsing of the input ANTLRv4 grammar.
+This is done using an automatically generated ANTLRv4 parser that was created from the grammar of the ANTLRv4 notation itself.
+The full tree, which comes out of the parser is, however, quite complex and contains information not relevant for the INGRID algorithm.
+In order to get a simple representation that is easy process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST have to be performed in this phase.
+An output of this phase is a custom tree-like structure representing the grammar.
 
-Nevertheless, the full AST, which comes out of the automatically generated parser, is quite complex and contains information not relevant for the INGRID algorithm.
-In order to get a simple representation that is easy process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST have to be performed in this phase.
-
-A simplified tree of objects is derived from the AST in two passes over the grammar.
+\PV{Tenhle dalsi odstavec bych klidne celej odebral, neprinasi zas tolik informaci, vzhledem k tomu, ze nejde do hloubky.}
+A simplified tree of objects is derived in two passes over the grammar.
 The result of the first pass is a tree where each node contains names of other referenced rules in the form of strings.
 In the second pass, the whole tree is traversed and all the references are resolved to have a form of real pointers to other rule objects.
+\PV{Az sem}
 
@@
 
+\PV{Tohle se mi libi, jak si prepsal, asi neni co vylepsit.}
 In the next phase, the complete structure of the MPS language is automatically generated from the AST that represents syntax of the input language.
@@
 
+\PV{Tady bych na zacatek vety pridal neco jako "In MPS...". Neni jasne, jestli mluvime o necem v ingrid nebo o konceptech MPS.}
 Each concept of the language (i.e., every AST node) is represented by an object that may have a parent, some children, and properties of any data type.
@@
 The object contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name}.
+\PV{Dalsi vetu bych nahradil: }
+\PV{Values of these properties will be restricted using the regular expression that corresponds to the \antlrlexerrule{Name} lexer rule.}
+\PV{Protoze }
 Values of these properties should be restricted using the regular expression that corresponds to the \antlrlexerrule{Name} rule, but this is left to the human user.
+\PV{Mozna "for" nahradit necim z XML, treba zavorkami </>, najednou totiz se preskoci na for, ktery se vzal odnikud.}
 String literals, such as the keyword \textbf{for}, are omitted because they will be defined only in the Editor aspect for this concept.
@@
 Consider the \antlrparserrule{content} rule from the SimpleXML language.
+\PV{Label CONTENTRULE defined multiple times! Pak to ukazuje spatny cislo Figure}
 We show just the relevant fragment of the grammar again in Figure~\ref{fig:CONTENTRULE}.
@@
 Our solution is to use interface concepts.
+\PV{Odstranil bych ", i.e. for each AST node with more than one child"}
 For each rule with more than one alternative on the right side, i.e.
@@
 
+\PV{K Phase 3 nemam vyhrady, to je prepsano super.}
 Having the complete structure of the new MPS language, the next phase is to define the visual representation of all concepts (AST nodes) in the projectional editor.
@@
 For example, in the case of the concept representing an XML element, there must be a space between the element's name and the first attribute, while it is not required between the opening bracket \antlrliteral{\textless} and the name (Figure~\ref{fig:TEXTGENBASIC}).
+
+\PV{Nejsem si jisty, jestli to nebylo az pozdeji, ale byl tu nekde odstavec, ktery se zminoval o tom, ze whitespace je silent-polykan parserem a proto pravidla nemusi obsahovat explicitne informaci o tom, kde vsude muze byt mezera \\\texttt{'<' WS* Name WS* attribute* WS* '>' WS*...}.\\To podle me dobre ilustrovalo, ze se tam ztraci ta informace.
+Prislo mi to jako zajimave poukazani, ale samozrejme to tu byt nemusi, pokud se ti zda, ze to az tak zajimave neni.}
 

### 2016-10-31 chapters/language_definition.tex
accepting some comments - others left for later

@@
 All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
-Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation \PV{Mozna vysvetlit proc (protoze je to dvoji prace) nebo odkazat tam, kde je to vysvetleny?}.
+Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.
+\PV{Mozna vysvetlit proc (protoze je to dvoji prace) nebo odkazat tam, kde je to vysvetleny?}
 
@@
 
-\PV{Nasledujici odstavec jsem lehce vylepsil, asi je to ted vysvetleno lepe.}
-The first phase in the process of MPS language construction is parsing of the input ANTLRv4 grammar.
-This is done using an automatically generated ANTLRv4 parser that was created from the grammar of the ANTLRv4 notation itself.
-The full tree, which comes out of the parser is, however, quite complex and contains information not relevant for the INGRID algorithm.
+The first phase in the process of MPS language construction is parsing of the input ANTLR v4 grammar.
+It is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
+Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the INGRID algorithm.
 In order to get a simple representation that is easy process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST have to be performed in this phase.
-An output of this phase is a custom tree-like structure representing the grammar.
+An output of this phase is a custom AST-like structure representing the grammar.
 
@@
 
-\PV{Tohle se mi libi, jak si prepsal, asi neni co vylepsit.}
 In the next phase, the complete structure of the MPS language is automatically generated from the AST that represents syntax of the input language.
@@
 
-\PV{K Phase 3 nemam vyhrady, to je prepsano super.}
 Having the complete structure of the new MPS language, the next phase is to define the visual representation of all concepts (AST nodes) in the projectional editor.

### 2016-11-01 chapters/language_definition.tex
Fix a few typos

@@
 Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the INGRID algorithm.
-In order to get a simple representation that is easy process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST have to be performed in this phase.
+In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST have to be performed in this phase.
 An output of this phase is a custom AST-like structure representing the grammar.
@@
 
-Our INGRID method targets mostly text-based languages, for whose concepts the visual representation in a projectional editor must be almost equivalent to their plain-text representation.
+Our INGRID method targets mostly text-based languages, for the concepts of which the visual representation in a projectional editor must be almost equivalent to their plain-text representation.
 An obvious choice would therefore be to use the same approach for Editor and TextGen.
@@
 \end{itemize}
-A space must be alway inserted when two child nodes are next to each other.
-Elements of a sequence of child nodes are separated with space or a line break.
+A space must be always inserted when two child nodes are next to each other.
+Elements from a sequence of child nodes are separated with spaces or line breaks.
 

### 2016-11-03 chapters/background.tex
evaluation - in progress

@@
 
-\PV{Nasledujici dve vety byly vzaty z popisu INGRID, tak funguje INGRID, ne PE4MPS.}
+\PV{Veta ktera zminuje dve faze s Java objekty se tyka INGRID (popisuje fungovani INGRID).}
+\PP{PE4MPS funguje takhle:
+Je tam specialni jazyk, kde uzivatel zada informace k importu.
+Je tam specialni koncept z nejakeho Import.process jazyka, kde se vyplni jeho atributy: 1) Nazev ciloveho jazyka, 2) Cesta k souboru, 3) A pak se klikne na "Import" v Tools menu.
+Ten jazyk uz musi existovat (byt zalozeny uvnitr MPS) a neni tam primo vyber souboru, ale musi se rucne vyplnovat cesta.
+Primo z PE souboru to pak nageneruje ten cilovy jazyk. Nejsou tam zadne dve faze.
+}
 An MPS language is created by the PE4MPS tool in two steps.

### 2016-11-03 chapters/conclusion.tex
textgen - done
evaluation - in progress

@@
 \section{Conclusion and Future Work}
+
+The proposed INGRID method is only partially-automated, but nevertheless it reduces the amount of necessary manual work/effort (that the user has to perform) to a great degree.
 
@@
 Generating nicely designed and user-friendly editors, maybe using artificial intelligence (machine learning) that would evaluate (analyze) existing code in plain text and based on that create the editors.
+Moreover, once we would have some information about the layout for generating better projectional editor, we could leverage the same information and use it to improve TextGen.
 

### 2016-11-03 chapters/evaluation.tex
textgen - done
remarks - done
evaluation - in progress

--- 2016-09-23 chapters/evaluation.tex
@@
 \section{Evaluation}
+\label{sect:EVAL}
 
-We implemented everything (the proposed approach described above) in the form of an MPS plugin that allows users to carry out the import.
+% TODO continue from here
 
+We implemented the proposed INGRID method as an (in the form of) an MPS plugin that allows users to run (perform, carry out) the construction of an MPS language from an ANTLR v4 grammar.
 We used a mix of Java and BaseLanguage in our implementation.
 We used the MPS API, which can be called from the BaseLanguage programs too, to generate some of the language elements.
-BaseLanguage is needed to use/call the MPS (language) API, i.e. to programmatically create new language elements and their aspects.
+BaseLanguage is needed to use/call the MPS API, e.g. in order to programmatically create new language elements and their aspects.
+The plugin uses the ANTLR library~\cite{ref:ANTLR4} to work with ANTLR v4 grammar files (to parse them, etc) and several MPS libraries (that implement the MPS API).
+Our complete implementation is available at \todo{url of github repository}.
 
-The plugin uses the ANTLR library to work with ANTLRv4 grammar files (to parse them, etc) and several MPS libraries (that implement the MPS API).
+For the purpose of evaluation, we applied INGRID to several complex mainstream languages, including JSON, JavaScript, and C\#.
+We used the version of JavaScript that is standardized as ECMAScript 5.1 --- the specification was released/finalized in 2011 and it is currently the most frequently adopted version.
+MPS projects that contain definitions of all these three programming languages are available in the public repository at \todo{url}
+In the rest of this section, we discuss our experience with application of INGRID to the selected languages, and then we highlight general observations.
 
-For the first step of automated construction of MPS languages from grammars (i.e., for parsing the grammar), the plugin uses the ANTLR library~\cite{ref:ANTLR4} to generate a Java ANTLR parser of the ANTLRv4 notation, because there exists an ANTLRv4 grammar for the ANTLRv4 notation~\cite{ref:ANTLR4reference}.
+However, before giving details on the experience, we must emphasize that the MPS languages produced by INGRID (the automatically imported languages) are not ready-to-use full-fledged MPS counterparts of the original input languages.
+Imported languages contain full structure found in the respective ANTLR grammars and some more aspect definitions that will help users of the languages in creating code.
+Every imported language has to be manually tweaked/adjusted by the end user (e.g., in particular/especially its Editor aspect).
+The imported language has to be manually adjusted (tweaked, polished) by the end user (developer) and some of its complications have to be resolved manually by a human.
+One of our goals was to show that even though our method (plugin) does not create a perfect MPS language, it can be customized very fast into a very usable form.
+We manually tweaked/customized the editor and the TextGen aspect only, structure aspect was left in its original form.
 
-\todo{
-	We should provide url to the github repository (implementation) and pointer to some package with ANTLR grammars and imported MPS languages (take from Attachments.zip in the master thesis).
-}
+For each of the languages, we provide the result (MPS language) exactly in the form as created by our implementation (plugin), and also the adjusted grammars used as input for INGRID, both/all in the repository with implementation.
 
-\todo{Popsat nase zkusenosti s importem slozitych jazyku (XML, JavaScript, Python, a dalsi).}
-
-From introduction: We will also show some example languages imported with INGRID and discuss our experience.
-
-In order to evaluate INGRID, we tried to import several real/mainstream languages:
-\begin{itemize}
-	\item \textbf{SimpleXML} --- Simplified XML that we used in this thesis for explanatory purposes.
-	\item \textbf{JSON} --- Similar to XML, but this time a full port of the specification.
-	\item \textbf{ECMAScript 5.1} --- Specification of the language known as JavaScript, dated to the year 2011, which is currently the most frequently adopted version.
-\end{itemize}
-Results of the import process are included in the package that is available at \todo{url}.
-The repository \todo{Add url} contains an example MPS project with three languages.
-For each example language, you can find its original import, exactly in the form as created by our plugin.
-
-We experimented with import of XML, JavaScript, Python, Ruby, C#, Swift, MySQL and some other languages.
-
-Most of these languages are quite complex when it comes to their structure and the overall syntax variety.
-
-We must emphasize that the automatically imported languages are not ready-to-use full-fledged MPS languages.
-We did not expect to achieve that at this stage, as that is a much bigger, if not impossible, challenge.
-Imported languages contain full structure found in the respective grammars and some more aspect definitions that will help users of the languages in creating code.
-\PV{The imported language has to be manually tweaked by the end user.}
-The imported language has to be manually adjusted (tweak, polished) by the end user (developer) and some of its complications have to be resolved manually by a human.
-
-We found out that manual adjusting of the editor/textgen aspect (adding line breaks and indentation) is far more effective and precise than any automated approach we came up with (during our research).
-The imported language can be, in our opinion, adjusted very fast into a more readable form.
-
-The package at \todo{url} contains also adjusted versions of all three languages.
-Then, for each language, there is an adjusted version.
-
-This adjusted version was created from the original import and the author spent no more than between 20 to 60 minutes of working on it.
-The first author spent only a couple of minutes of (manually) adjusting them after the automated import, so that we try to prove our point above.
+\paragraph{JSON.}
+The first language that we tried was JSON \todo{nejaka reference na specifikaci nebo pripadne gramatiku}.
+The adjusted version of the JSON language in MPS was created from the original by the first author in no more than 20 minutes of working on it.
 More specifically, the JSON language was ready for use in no more than 20 minutes of minor adjusting.
 We chose this language because it does not require us to implement complex aspects, such as type checking or data flows.
-Similarly, the SimpleXML turned out very nice.
 
-Our main goal was to prove that even though our method (plugin) does not create a perfect language, it can be customized very fast into a very usable form.
-We manually tweaked/customized the editor and the TextGen aspect only, structure aspect was left in its original form.
-This justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical (and the best\PV{?}).
+\paragraph{JavaScript.}
+We have also tried to use the INGRID tool/plugin to import the JavaScript language, which is an example/instance of a widely-used complicated general purpose language.
+Importing the JavaScript language~\cite{ref:javascript} and manually adjusting all editors, that needed it, in a less than hour time.
 
-We have also tried to use the INGRID tool/plugin to import the JavaScript language, which is exactly the type of a complicated general purpose language that might be of interest to other people.
 There are already some projects, where a manual port of JavaScript is being done.
@@
 The structure aspect is very similar to the ECMAScript4MPS's one, which was created manually over the course of a large number of hours.
-INGRID does that (achieves the same result) fully automatically.
+INGRID does that (achieves the same result) fully automatically, and yields very good/nice results.
 
-\PV{Veta se opakuje nize, odstranil bych tuhle}
-To summarize, import languages such as JavaScript or JSON and others are giving nice results.
+\paragraph{C\#}.
+Pred generovanim MPS jazyka pro C\# bylo nutne upravit gramatiku (coz zabralo asi hodinu prace) aby INGRID vyprodukoval pomerne rozumny vysledek.
+The definition of C\# in MPS is quite large and complex.
+Obsahuje asi vic nez 800 konceptu (zkontrolovat).
+Import of the adjusted C\# language funguje docela hezky (ale je hodne velky a dlouho se kompiluje).
+Ale na skutecne dobre pouzitelny jazyk C\# uvnitr MPS by to chtelo jeste vic manualnich uprav gramatiky a take urcitych aspektu vysledneho MPS jazyka (editor, textgen, apod).
 
-Results of our experiments with languages such as JSON or JavaScript show that the proposed approach (INGRID) yields very good results, comparable to existing manual ports, and spares the user from hours of tedious, error prone and time-consuming work.
+\paragraph{Other languages.}
+We also experimented with import of other languages, such as Python and Ruby.
+Import jazyku Python a Ruby moc nefunguje protoze gramatika je napsana stylem ktery neni vhodny pro INGRID.
+INGRID cannot handle cyclic rules (which could be found in Python grammar, for example).
+Cyklicka pravidla to tedy neumi, jen na ne upozorni (a rekne, ktere pravidlo je obsahuje).
+Museli bychom gramatiku rucne upravit (bude to ale nase future work).
 
-\PV{Tohle uz neni pravda, error reporting i cyklicka pravidla jsem pred odevzdanim DP vyladil a kod celkove dodelal.
-Cyklicka pravidla to tedy neumi, jen na ne upozorni (a rekne, ktere pravidlo je obsahuje).}
-Limitation: the project is lacking proper error reporting and cannot handle cyclic rules (which could be found in Python grammar, for example).
+\paragraph{General observations.}
+Most of the languages (that we tried to define in MPS, use/import into MPS) are quite complex when it comes to their structure and the overall syntax variety, so defining them manually would be a very tedious, error-prone and time-consuming process.
+\todo{Discuss where INGRID helps the most based on the experience.}
+(Our experience/experiments show that) Automated generation of the Structure aspect is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
+On the other hand, we found out that manual tweaking/adjusting of the Editor and Textgen aspects (adding line breaks and indentation) is very effective, and takes only a relatively little/short time (much less than we expected).
+The imported language can be, in our opinion, adjusted very fast into a more readable form.
+In fact/total, the adjusted version of each language was created from the original import by the first author, who spent no more than between 20 to 60 minutes of working on each of them.
+The first author spent only a couple of minutes of (manually) adjusting the languages after the automated import.
+Our experience with all the languages confirms our hypothesis (that code layout in the Editor aspect can be adjusted very fast, from the basic default single-row layout), and justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical.
+It takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language and manually adjusting all editors that needed it, all that in a less than hour time.
+\todo{Mozna rict:} However, during our work on this project, we also tried to design some/various automated heuristics, but all that we came up with so far give rather suboptimal/bad results especially when compared to results produced by a human user.

### 2016-11-04 chapters/evaluation.tex
evaluation - first part

@@
 
-% TODO continue from here
+We implemented the proposed INGRID method as an MPS plugin that allows users to generate languages from ANTLR v4 grammars.
+While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to call the MPS API, e.g. in order to programmatically create new language concepts and their aspects.
+The plugin uses the ANTLR library~\cite{ref:ANTLR} for parsing of ANTLR v4 grammar files, and several MPS libraries that implement the MPS API.
+Our complete implementation is available at \url{https://github.com/premun/ingrid}.
 
-We implemented the proposed INGRID method as an (in the form of) an MPS plugin that allows users to run (perform, carry out) the construction of an MPS language from an ANTLR v4 grammar.
-We used a mix of Java and BaseLanguage in our implementation.
-We used the MPS API, which can be called from the BaseLanguage programs too, to generate some of the language elements.
-BaseLanguage is needed to use/call the MPS API, e.g. in order to programmatically create new language elements and their aspects.
-The plugin uses the ANTLR library~\cite{ref:ANTLR4} to work with ANTLR v4 grammar files (to parse them, etc) and several MPS libraries (that implement the MPS API).
-Our complete implementation is available at \todo{url of github repository}.
+For the purpose of evaluation, we applied INGRID to several well-established and widely used languages, including JSON, JavaScript, and C\#.
+We used the version of JavaScript that is standardized as ECMAScript 5.1~\cite{ref:ECMASCRIPT51} --- the specification was finalized in June 2011, and it is currently the most frequently adopted version.
+MPS projects that contain definitions of all three languages are also released in the public repository (\url{https://github.com/premun/ingrid}).
+In the rest of this section, we discuss our experience with application of INGRID to these languages, and then we highlight few general observations.
 
-For the purpose of evaluation, we applied INGRID to several complex mainstream languages, including JSON, JavaScript, and C\#.
-We used the version of JavaScript that is standardized as ECMAScript 5.1 --- the specification was released/finalized in 2011 and it is currently the most frequently adopted version.
-MPS projects that contain definitions of all these three programming languages are available in the public repository at \todo{url}
-In the rest of this section, we discuss our experience with application of INGRID to the selected languages, and then we highlight general observations.
-
-However, before giving details on the experience, we must emphasize that the MPS languages produced by INGRID (the automatically imported languages) are not ready-to-use full-fledged MPS counterparts of the original input languages.
-Imported languages contain full structure found in the respective ANTLR grammars and some more aspect definitions that will help users of the languages in creating code.
-Every imported language has to be manually tweaked/adjusted by the end user (e.g., in particular/especially its Editor aspect).
-The imported language has to be manually adjusted (tweaked, polished) by the end user (developer) and some of its complications have to be resolved manually by a human.
-One of our goals was to show that even though our method (plugin) does not create a perfect MPS language, it can be customized very fast into a very usable form.
-We manually tweaked/customized the editor and the TextGen aspect only, structure aspect was left in its original form.
-
-For each of the languages, we provide the result (MPS language) exactly in the form as created by our implementation (plugin), and also the adjusted grammars used as input for INGRID, both/all in the repository with implementation.
+However, first we must emphasize that MPS languages automatically produced by the INGRID method, as defined in this paper, are not ready-to-use full-fledged MPS counterparts of the original input languages.
+The structure of such a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects (e.g., the Editor) would have to be manually tweaked (or defined from scratch) by the end user.
+INGRID does not yet support advanced features of MPS, such as the aspect responsible for type checking and inference.
+For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition exactly in the form created by INGRID and (2) the adjusted ANTLR v4 grammar used as input for INGRID, both in the repository that contains also our implementation.
 
 \paragraph{JSON.}
-The first language that we tried was JSON \todo{nejaka reference na specifikaci nebo pripadne gramatiku}.
-The adjusted version of the JSON language in MPS was created from the original by the first author in no more than 20 minutes of working on it.
-More specifically, the JSON language was ready for use in no more than 20 minutes of minor adjusting.
-We chose this language because it does not require us to implement complex aspects, such as type checking or data flows.
+The least amount of manual adjusting after the import into MPS was needed in case of the JSON language~\cite{ref:JSON}, because it is the simplest language from all that we used for our experiments.
+Specifically, the first author spent less than 20 minutes in order to get a language that is ready to use.
 
 \paragraph{JavaScript.}
-We have also tried to use the INGRID tool/plugin to import the JavaScript language, which is an example/instance of a widely-used complicated general purpose language.
-Importing the JavaScript language~\cite{ref:javascript} and manually adjusting all editors, that needed it, in a less than hour time.
+In the case of JavaScript, which is an example of a widely-used complex general purpose language, automated generating of the language definition in MPS from the ANTLR grammar\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} and subsequent manual adjusting was done in less than one hour.
 
-There are already some projects, where a manual port of JavaScript is being done.
-One of them is the ECMAScript4MPS~\cite{ref:ECMAScript4MPS} from the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section X (similar projects, related work).
-The result of import by INGRID is quite good, comparable to the PE4MPS project.
-Our language is, of course, missing a lot of more advanced aspects/features that, for example, can help deriving types of expressions and so on.
-These aspects improve the usability by a lot, e.g.
-when the user starts writing numbers, a number literal is inserted, whereas in our import the user has to first insert the concept representing the number literal and then proceed with writing numbers.
-These actions are needed in any MPS language and are expected to be implemented after the import step, \todo{Maybe say that INGRID does not generate these actions/aspects/features yet but it is our future work} as it is not possible to generate them automatically.
-However, when it comes to language structure, concept aliases, and auto-completion, we think that INGRID (our method/plugin) did a very good job.
-The structure aspect is very similar to the ECMAScript4MPS's one, which was created manually over the course of a large number of hours.
-INGRID does that (achieves the same result) fully automatically, and yields very good/nice results.
+We are aware of other projects that aim to create a manual port of JavaScript into MPS.
+For example, there is ECMAScript4MPS~\cite{ref:ECMA4MPS} developed by the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section~\ref{sect:RELATED}.
+
+The main advantage of INGRID over ECMAScript4MPS is that, despite its current limitations, INGRID fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful.
+INGRID achieved a very good result especially in the case of language structure, concept aliases, and support for auto-completion.
+The Structure aspect generated by INGRID is very similar to that of the ECMAScript4MPS project, which was created manually in a large number of hours.
+
+% TODO continue from here
 
@@
 The imported language can be, in our opinion, adjusted very fast into a more readable form.
+The generated MPS language can be customized very fast into a very usable form.
 In fact/total, the adjusted version of each language was created from the original import by the first author, who spent no more than between 20 to 60 minutes of working on each of them.

### 2016-11-04 paper.tex
writing section 3
evaluation - first part

@@
 
+\bibitem{ref:ECMASCRIPT51} ECMAScript Language Specification, Edition 5.1 (June 2011), \url{http://www.ecma-international.org/ecma-262/5.1/Ecma-262.pdf}
+
 \bibitem{ref:MPS} JetBrains MPS, \url{https://www.jetbrains.com/mps/}
+
+\bibitem{ref:JSON} The JSON Data Interchange Format, 1st Edition (October 2013), \url{http://www.ecma-international.org/publications/files/ECMA-ST/ECMA-404.pdf}
+
+\bibitem{ref:ECMA4MPS} M. Lombardo. The ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}
 
@@
 
-% TODO opravit formatovani nasledujicich polozek a vhodne zaradit
-
-\bibitem{ref:ANTLR4} ANTLRv4, \url{https://github.com/antlr/antlr4}
-
-\bibitem{ref:ANTLR4reference} T. Parr. The Definitive ANTLR 4 Reference. The Pragmatic Bookshelf, 2013
-
-\bibitem{ref:ANTLR4grammars} ANTLRv4 grammar repository, \url{https://github.com/antlr/grammars-v4}
-
-\bibitem{ref:ECMAScript4MPS} ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}
-
-\bibitem{ref:javascript} JavaScript ANTLRv4 grammar, \url{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}
-
 \end{thebibliography}

### 2016-11-05 chapters/evaluation.tex
evaluation - done

@@
 
-% TODO continue from here
+\paragraph{C\#.}
+Another very complex programming language that we used for experiments is C\#.
+Before we could run INGRID, we had to manually adjust the ANTLR grammar of C\# in order to ensure that INGRID produces a reasonably useful MPS language.
+Roughly one hour of manual effort was needed for this step.
+Nevertheless, additional modifications of the grammar and selected aspects of the MPS language (Editor, TextGen) are still needed to get an optimal result.
 
-\paragraph{C\#}.
-Pred generovanim MPS jazyka pro C\# bylo nutne upravit gramatiku (coz zabralo asi hodinu prace) aby INGRID vyprodukoval pomerne rozumny vysledek.
-The definition of C\# in MPS is quite large and complex.
-Obsahuje asi vic nez 800 konceptu (zkontrolovat).
-Import of the adjusted C\# language funguje docela hezky (ale je hodne velky a dlouho se kompiluje).
-Ale na skutecne dobre pouzitelny jazyk C\# uvnitr MPS by to chtelo jeste vic manualnich uprav gramatiky a take urcitych aspektu vysledneho MPS jazyka (editor, textgen, apod).
+The generated definition of C\$ in MPS is quite large, involving more than 800 concepts, and therefore building of the language takes a rather long time.
 
 \paragraph{Other languages.}
-We also experimented with import of other languages, such as Python and Ruby.
-Import jazyku Python a Ruby moc nefunguje protoze gramatika je napsana stylem ktery neni vhodny pro INGRID.
-INGRID cannot handle cyclic rules (which could be found in Python grammar, for example).
-Cyklicka pravidla to tedy neumi, jen na ne upozorni (a rekne, ktere pravidlo je obsahuje).
-Museli bychom gramatiku rucne upravit (bude to ale nase future work).
+We also tried to apply INGRID on few other languages, such as Python and Ruby.
+Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with INGRID.
+For example, INGRID cannot handle cyclic rules (which appear in the Python grammar) and just prints a warning on the screen.
+Significant modifications of the grammar would be necessary.
+We plan to address this limitation in the future.
 
 \paragraph{General observations.}
-Most of the languages (that we tried to define in MPS, use/import into MPS) are quite complex when it comes to their structure and the overall syntax variety, so defining them manually would be a very tedious, error-prone and time-consuming process.
-\todo{Discuss where INGRID helps the most based on the experience.}
-(Our experience/experiments show that) Automated generation of the Structure aspect is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
-On the other hand, we found out that manual tweaking/adjusting of the Editor and Textgen aspects (adding line breaks and indentation) is very effective, and takes only a relatively little/short time (much less than we expected).
-The imported language can be, in our opinion, adjusted very fast into a more readable form.
-The generated MPS language can be customized very fast into a very usable form.
-In fact/total, the adjusted version of each language was created from the original import by the first author, who spent no more than between 20 to 60 minutes of working on each of them.
-The first author spent only a couple of minutes of (manually) adjusting the languages after the automated import.
-Our experience with all the languages confirms our hypothesis (that code layout in the Editor aspect can be adjusted very fast, from the basic default single-row layout), and justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical.
-It takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language and manually adjusting all editors that needed it, all that in a less than hour time.
-\todo{Mozna rict:} However, during our work on this project, we also tried to design some/various automated heuristics, but all that we came up with so far give rather suboptimal/bad results especially when compared to results produced by a human user.
+The main overall benefit of the INGRID method is partial automation.
+Most of the languages discussed above, which we tried to create in MPS by using INGRID, are quite complex regarding their structure and syntax variety.
+Therefore, completely manual definition would be a very time-consuming and error-prone process.
+Fully automated generation of the Structure aspect is where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
+
+On the other hand, our experiments with complex languages show that, in the case of the Editor and Textgen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, much less time than we expected.
+The first author spent between 20 and 60 minutes of work on each language, when adjusting the result of automated generation into a more useful and readable form.
+MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing users to reach optimal results very fast.
+All of this justifies some of our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
+Despite that, we also tried to design some automated heuristics during our work on the INGRID project, but so far all yield rather average results especially when compared to what human users can achieve efficiently.

### 2016-11-06 chapters/background.tex
addressing comments

--- 2016-10-31 chapters/background.tex
@@
 
-\PV{Nasledujici dve vety byly vzaty z popisu INGRID, tak funguje INGRID, ne PE4MPS.}
-An MPS language is created by the PE4MPS tool in two steps.
-First, the PE parser is used to load an input PE file and build its intermediate representation, which has the form of a tree-like structure consisting of Java objects.
-Then, all the concepts (AST nodes) and their aspects are created inside MPS.
-The information about the code layout, extracted from the PE file, is used when generating the projectional editor for every AST node.
+The PE4MPS tool creates an MPS language in a single atomic step.
+An input PE file is loaded and processed by the PE parser, and then all concepts (AST nodes) and their aspects are directly generated inside MPS.
+Information about the code layout, extracted from the PE file, is used when generating the projectional editor for every AST node.
 
 The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
-\PV{Navrhuji pridat (aby to bylo vice jasne?): User is forced to input the layout information by extending the grammar manually.}
+In particular, language developers are forced to specify the code layout manually in PE grammars, through the corresponding extensions.
 Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
@@
 However, the project has also other important limitations.
-The tool does not generate any parent-child relationships in the Structure aspect, and it completely neglects the Editor and TextGen aspects.
-\PV{Dalsi veta mozna zbytecna?}
-In particular, it does not support creating of the projectional editor at all.
-
-\PV{Navrhuji pridat neco jako (ale mozna uz o tom mluvime jinde, jen si ted nejsem jisty):
-The major take away from this project is the idea of an intermediate grammar representation inside of MPS which we have also considered.
-The advantage of this could be the possibility of extending the ANTLRv4 notation using some graphic tools that MPS has to offer. 
-Using the projectional editor, a user could add the layout information in an effective way.
-The disadvantage is the dual implementation that is needed.
-First to transform the ANTLRv4 text file into the MPS grammar language and secondly to transform this grammar into the target MPS language.}
+The tool does not generate any parent-child relationships in the Structure aspect, and it provides no support for the projectional editor at all (by completely neglecting the Editor and TextGen aspects).
 
@@
 In the last step, the target MPS language is derived from the adjusted grammar definition.
-\PV{Navrhuji odsud dolu smazat.}
-Many iterations of this process, involving adjustments of the grammar followed by recreation of the target language, can be performed.
 
-The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLRv4 MPS language into the actual target language.
-\PV{Smazat az sem.}
-\PV{A pridat: This step has not yet been implemented, because MPS doesn't offer any mechanisms for automatic transformation.
-This means that implementation would require traversing the AST of the grammar and recreating the definition from that.}
+The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLRv4 MPS language into the definition of the actual target MPS language.

### 2016-11-06 chapters/language_definition.tex
evaluation - first part
evaluation - done
addressing comments

@@
 Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.
-\PV{Mozna vysvetlit proc (protoze je to dvoji prace) nebo odkazat tam, kde je to vysvetleny?}
 
@@
 Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the INGRID algorithm.
-In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST have to be performed in this phase.
-An output of this phase is a custom AST-like structure representing the grammar.
-
-\PV{Tenhle dalsi odstavec bych klidne celej odebral, neprinasi zas tolik informaci, vzhledem k tomu, ze nejde do hloubky.}
-A simplified tree of objects is derived in two passes over the grammar.
-The result of the first pass is a tree where each node contains names of other referenced rules in the form of strings.
-In the second pass, the whole tree is traversed and all the references are resolved to have a form of real pointers to other rule objects.
-\PV{Az sem}
+In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST are performed in this phase.
+An output is a custom AST-like structure representing the grammar, especially the hierarchy of parser rules.
 
@@
 
-\PV{Tady bych na zacatek vety pridal neco jako "In MPS...". Neni jasne, jestli mluvime o necem v ingrid nebo o konceptech MPS.}
-Each concept of the language (i.e., every AST node) is represented by an object that may have a parent, some children, and properties of any data type.
-In addition, the object may implement any number of interfaces, and it may also contain references to objects representing other AST nodes.
+In MPS, each concept of the language (i.e., every AST node) is represented by an object that may have a parent, some children, and properties of any data type.
+Additionally, the object may implement any number of interfaces, and it may also contain references to objects representing other AST nodes.
 The parent-child relationships between objects that make the Structure aspect are derived from rules of the grammar (i.e., from the structure of the AST).
@@
 The object contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name}.
-\PV{Dalsi vetu bych nahradil: }
-\PV{Values of these properties will be restricted using the regular expression that corresponds to the \antlrlexerrule{Name} lexer rule.}
-\PV{Protoze }
-Values of these properties should be restricted using the regular expression that corresponds to the \antlrlexerrule{Name} rule, but this is left to the human user.
-\PV{Mozna "for" nahradit necim z XML, treba zavorkami </>, najednou totiz se preskoci na for, ktery se vzal odnikud.}
-String literals, such as the keyword \textbf{for}, are omitted because they will be defined only in the Editor aspect for this concept.
+Values of these properties will be restricted using the regular expression that corresponds to the lexer rule \antlrlexerrule{Name}.
+String literals, such as the opening and closing brackets (\antlrliteral{\textless}, \antlrliteral{/\textgreater}) in XML, are omitted because they will be defined only in the Editor aspect for this concept.
 
@@
 Consider the \antlrparserrule{content} rule from the SimpleXML language.
-\PV{Label CONTENTRULE defined multiple times! Pak to ukazuje spatny cislo Figure}
 We show just the relevant fragment of the grammar again in Figure~\ref{fig:CONTENTRULE}.
@@
 Our solution is to use interface concepts.
-\PV{Odstranil bych ", i.e. for each AST node with more than one child"}
-For each rule with more than one alternative on the right side, i.e.
-for each AST node with more than one child, first an interface concept is defined in the MPS language structure, and then one object that implements the given interface is created for each alternative.
+For each rule with more than one alternative on the right side, first an interface concept is defined in the MPS language structure, and then one object that implements the given interface is created for each alternative.
 The resulting fragment of the language structure looks like the one in Figure~\ref{fig:ICONTENTITF}.
@@
 The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree and consider just the leaf nodes, e.g., to be directly offered to the user through the auto-completion menu.
-Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, which is given in Figure~\ref{fig:CONTENTRULE} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the leaf nodes that are highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.
+Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, which is given in Figure~\ref{fig:CONTENTHIERARCHY} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the leaf nodes that are highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.
 
@@
 \caption{The parser rule \antlrparserrule{content} with other rules that determine the relevant fragment of the syntax hierarchy}
-\label{fig:CONTENTRULE}
+\label{fig:CONTENTHIERARCHY}
 \end{figure*}
@@
 We use the name \emph{shortcut approach} for this algorithm, because the paths collected by the algorithm provide shortcuts from the given rule to end nodes, by the virtue of hiding all intermediary layers.
-For example, the result of this algorithm for the \antlrparserrule{content} rule (Figure~\ref{fig:CONTENTRULE}) is the list of five items already shown in Figure~\ref{fig:CONTENTEXPAND}.
+For example, the result of this algorithm for the \antlrparserrule{content} rule (Figure~\ref{fig:CONTENTHIERARCHY}) is the list of five items already shown in Figure~\ref{fig:CONTENTEXPAND}.
 
@@
 \subsection{Phase 3: Editor}
+\label{sect:EDITORDEF}
 
@@
 \subsection{Phase 4: Text Generation}
+\label{sect:TEXTGENDEF}
 
@@
 For example, in the case of the concept representing an XML element, there must be a space between the element's name and the first attribute, while it is not required between the opening bracket \antlrliteral{\textless} and the name (Figure~\ref{fig:TEXTGENBASIC}).
-
-\PV{Nejsem si jisty, jestli to nebylo az pozdeji, ale byl tu nekde odstavec, ktery se zminoval o tom, ze whitespace je silent-polykan parserem a proto pravidla nemusi obsahovat explicitne informaci o tom, kde vsude muze byt mezera \\\texttt{'<' WS* Name WS* attribute* WS* '>' WS*...}.\\To podle me dobre ilustrovalo, ze se tam ztraci ta informace.
-Prislo mi to jako zajimave poukazani, ale samozrejme to tu byt nemusi, pokud se ti zda, ze to az tak zajimave neni.}
 
@@
 
-The second example relates to a fragment of the JavaScript language, also known as ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}.
+The second example relates to a fragment of the JavaScript language, also known as ECMAScript.
+%\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}.
 Every statement in JavaScript needs to be terminated by a semicolon, newline, end of the file, or end of the block --- see the Figure~\ref{fig:JAVASCRIPTSTMT}, which contains the corresponding fragment of the ECMAScript grammar.

### 2016-11-07 chapters/evaluation.tex
Poznamky k evaluation

@@
 We implemented the proposed INGRID method as an MPS plugin that allows users to generate languages from ANTLR v4 grammars.
-While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to call the MPS API, e.g. in order to programmatically create new language concepts and their aspects.
+While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to call \PV{nebo "to bind with"} the MPS API,  \PV{nebo "which was used to programatically generate elements of the new language} e.g.
+in order to programmatically create new language concepts and their aspects.
 The plugin uses the ANTLR library~\cite{ref:ANTLR} for parsing of ANTLR v4 grammar files, and several MPS libraries that implement the MPS API.
@@
 For the purpose of evaluation, we applied INGRID to several well-established and widely used languages, including JSON, JavaScript, and C\#.
+\PV{Dalsi vetu bych odebral a treba jen v predchozi vete dal JavaScript (ECMAScript 5.1). Usetri to hodne mista a informace zustane stejna.}
 We used the version of JavaScript that is standardized as ECMAScript 5.1~\cite{ref:ECMASCRIPT51} --- the specification was finalized in June 2011, and it is currently the most frequently adopted version.
@@
 The structure of such a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects (e.g., the Editor) would have to be manually tweaked (or defined from scratch) by the end user.
+\PV{Nejsem si jisty, jestli je to vubec mozne.
+To, ze nejake pravidlo "ma typ", je asi az moc high-level informace.
+To, ze jedno pravidlo je definice metody a jine je volani metody, take z gramatiky nikdy ziskat nepujde.
+Bud vetu odstranit nebo preformulovat (bez yet atd.)}
 INGRID does not yet support advanced features of MPS, such as the aspect responsible for type checking and inference.
+\PV{Toto take neni pravda - editor je v kazdem jazyku lehce upraveny (to je tech 20 min / 1 hodina ...). Structure aspect je stejny, textgen je take krome Csharpu upraveny.}
 For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition exactly in the form created by INGRID and (2) the adjusted ANTLR v4 grammar used as input for INGRID, both in the repository that contains also our implementation.
@@
 \paragraph{JavaScript.}
+\PV{Poznamka vyse o ECMAScript 5.1 by sla take prenest sem.}
 In the case of JavaScript, which is an example of a widely-used complex general purpose language, automated generating of the language definition in MPS from the ANTLR grammar\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} and subsequent manual adjusting was done in less than one hour.
@@
 INGRID achieved a very good result especially in the case of language structure, concept aliases, and support for auto-completion.
-The Structure aspect generated by INGRID is very similar to that of the ECMAScript4MPS project, which was created manually in a large number of hours.
+The Structure aspect generated by INGRID is very similar to that of the ECMAScript4MPS project, which was created manually \PV{over misto in?} in a large number of hours.
 
@@
 Before we could run INGRID, we had to manually adjust the ANTLR grammar of C\# in order to ensure that INGRID produces a reasonably useful MPS language.
+\PV{Na gramatiku asi stacilo 20 minut, kdyz clovek vi, co dela. 40 minut pak na castecnou upravu editoru.}
 Roughly one hour of manual effort was needed for this step.
@@
 
-The generated definition of C\$ in MPS is quite large, involving more than 800 concepts, and therefore building of the language takes a rather long time.
+\PV{Tato veta asi neni az tak dulezita.
+Spis mi tu chybi neco ve smyslu, ze Csharp je opravdu flexibilni jazyk, diky cemuz je import nejdale od pouzitelneho MPS jazyka (prakticky vsechny statementy lze rozvadet pomoci generiky, pretypovani atd atd).}
+The generated definition of C\# in MPS is quite large, involving more than 800 concepts, and therefore building of the language takes a rather long time.
 
@@
 Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with INGRID.
+\PV{Tim not compatible je spis myslena hiearchizace gramatiky a styl rozdeleni pravidel.
+Byl trochu jiny nez jinde a dochazelo tam k dost zmatenym strukturam (co se MPS tyce).
+Cyklicka pravidla jsou asi nejmensi problem, daji se odstranit prakticky hned a vetsinou celkem bez ujmy.}
 For example, INGRID cannot handle cyclic rules (which appear in the Python grammar) and just prints a warning on the screen.
 Significant modifications of the grammar would be necessary.
+\PV{Mozna to ani neni jak adresovat, proste se bude muset upravit gramatika.}
 We plan to address this limitation in the future.
@@
 All of this justifies some of our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
-Despite that, we also tried to design some automated heuristics during our work on the INGRID project, but so far all yield rather average results especially when compared to what human users can achieve efficiently.
+Despite that, we also tried to design some automated heuristics during our work on the INGRID project, but so far all yield rather average results especially when compared to what human users can achieve efficiently \PV{instead}.

### 2016-11-08 chapters/conclusion.tex
conclusion

--- 2016-09-23 chapters/conclusion.tex
@@
 
-\todo{Write something.}
+Our contribution is the INGRID method for constructing language definitions in MPS from their grammars in the ANTLR notation.
+The method is only partially automated, but nevertheless it greatly reduces the amount of manual work that developers of MPS languages have to perform.
+Although we discussed all the challenges, details of our solution and general observations just in the context of ANTLR grammars and JetBrains MPS, we believe they are more general --- and in particular relevant to everyone tackling a similar problem in the field of DSLs and language workbenches.
 
-There is lot of further follow-up work to be done. Many open problems remain.
-
-Automated generation of the editor aspect, probably using some kind of machine learning.
-Generating nicely designed and user-friendly editors, maybe using artificial intelligence (machine learning) that would evaluate (analyze) existing code in plain text and based on that create the editors.
-
-Follow-up work might look more into the problem of code layout detection, possibly exploring suggested approaches such as the learning approach described in Section XY \todo{Tohle se snad podari aspon castecne implementovat}.
-This could lead to better results in editor and TextGen aspect generation which could definitely use some improvement.
-Another future endeavors will definitely touch the subject of generating parsers that would enable the user to import existing source code into MPS.
+In the future, we would like to address the remaining open problems (there are many of them) and increase the level of automation.
+Our primary research goal is to investigate the usage of machine learning for the purpose of generating editors that support nice and readable code layout.
+The information about code layout, once it is available, will be then leveraged and used also to improve TextGen.
+Another subject for future work is generating parsers that would allow users to import existing source code into MPS.

### 2016-11-09 chapters/background.tex
small changes

@@
 Every grammar might require non-trivial adjustments before it can be processed using the PE4MPS tool.
-To compare, the goals behind our INGRID approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.
+To compare, the goals behind our \textsc{Ingrid} approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.
 

### 2016-11-09 chapters/evaluation.tex
small changes

@@
 
-We implemented the proposed INGRID method as an MPS plugin that allows users to generate languages from ANTLR v4 grammars.
+We implemented the proposed \textsc{Ingrid} method as an MPS plugin that allows users to generate languages from ANTLR v4 grammars.
 While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to call \PV{nebo "to bind with"} the MPS API,  \PV{nebo "which was used to programatically generate elements of the new language} e.g.
@@
 
-For the purpose of evaluation, we applied INGRID to several well-established and widely used languages, including JSON, JavaScript, and C\#.
+For the purpose of evaluation, we applied \textsc{Ingrid} to several well-established and widely used languages, including JSON, JavaScript, and C\#.
 \PV{Dalsi vetu bych odebral a treba jen v predchozi vete dal JavaScript (ECMAScript 5.1). Usetri to hodne mista a informace zustane stejna.}
@@
 MPS projects that contain definitions of all three languages are also released in the public repository (\url{https://github.com/premun/ingrid}).
-In the rest of this section, we discuss our experience with application of INGRID to these languages, and then we highlight few general observations.
+In the rest of this section, we discuss our experience with application of \textsc{Ingrid} to these languages, and then we highlight few general observations.
 
-However, first we must emphasize that MPS languages automatically produced by the INGRID method, as defined in this paper, are not ready-to-use full-fledged MPS counterparts of the original input languages.
+However, first we must emphasize that MPS languages automatically produced by the \textsc{Ingrid} method, as defined in this paper, are not ready-to-use full-fledged MPS counterparts of the original input languages.
 The structure of such a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects (e.g., the Editor) would have to be manually tweaked (or defined from scratch) by the end user.
@@
 Bud vetu odstranit nebo preformulovat (bez yet atd.)}
-INGRID does not yet support advanced features of MPS, such as the aspect responsible for type checking and inference.
+\textsc{Ingrid} does not yet support advanced features of MPS, such as the aspect responsible for type checking and inference.
 \PV{Toto take neni pravda - editor je v kazdem jazyku lehce upraveny (to je tech 20 min / 1 hodina ...). Structure aspect je stejny, textgen je take krome Csharpu upraveny.}
-For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition exactly in the form created by INGRID and (2) the adjusted ANTLR v4 grammar used as input for INGRID, both in the repository that contains also our implementation.
+For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition exactly in the form created by \textsc{Ingrid} and (2) the adjusted ANTLR v4 grammar used as input for \textsc{Ingrid}, both in the repository that contains also our implementation.
 
@@
 
-The main advantage of INGRID over ECMAScript4MPS is that, despite its current limitations, INGRID fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful.
-INGRID achieved a very good result especially in the case of language structure, concept aliases, and support for auto-completion.
-The Structure aspect generated by INGRID is very similar to that of the ECMAScript4MPS project, which was created manually \PV{over misto in?} in a large number of hours.
+The main advantage of \textsc{Ingrid} over ECMAScript4MPS is that, despite its current limitations, \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful.
+\textsc{Ingrid} achieved a very good result especially in the case of language structure, concept aliases, and support for auto-completion.
+The Structure aspect generated by \textsc{Ingrid} is very similar to that of the ECMAScript4MPS project, which was created manually \PV{over misto in?} in a large number of hours.
 
@@
 Another very complex programming language that we used for experiments is C\#.
-Before we could run INGRID, we had to manually adjust the ANTLR grammar of C\# in order to ensure that INGRID produces a reasonably useful MPS language.
+Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces a reasonably useful MPS language.
 \PV{Na gramatiku asi stacilo 20 minut, kdyz clovek vi, co dela. 40 minut pak na castecnou upravu editoru.}
@@
 \paragraph{Other languages.}
-We also tried to apply INGRID on few other languages, such as Python and Ruby.
-Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with INGRID.
+We also tried to apply \textsc{Ingrid} on few other languages, such as Python and Ruby.
+Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with \textsc{Ingrid}.
 \PV{Tim not compatible je spis myslena hiearchizace gramatiky a styl rozdeleni pravidel.
@@
 Cyklicka pravidla jsou asi nejmensi problem, daji se odstranit prakticky hned a vetsinou celkem bez ujmy.}
-For example, INGRID cannot handle cyclic rules (which appear in the Python grammar) and just prints a warning on the screen.
+For example, \textsc{Ingrid} cannot handle cyclic rules (which appear in the Python grammar) and just prints a warning on the screen.
 Significant modifications of the grammar would be necessary.
@@
 \paragraph{General observations.}
-The main overall benefit of the INGRID method is partial automation.
-Most of the languages discussed above, which we tried to create in MPS by using INGRID, are quite complex regarding their structure and syntax variety.
+The main overall benefit of the \textsc{Ingrid} method is partial automation.
+Most of the languages discussed above, which we tried to create in MPS by using \textsc{Ingrid}, are quite complex regarding their structure and syntax variety.
 Therefore, completely manual definition would be a very time-consuming and error-prone process.
-Fully automated generation of the Structure aspect is where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
+Fully automated generation of the Structure aspect is where \textsc{Ingrid} spares the user from many hours of tedious and sometimes quite challenging work.
 
@@
 All of this justifies some of our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
-Despite that, we also tried to design some automated heuristics during our work on the INGRID project, but so far all yield rather average results especially when compared to what human users can achieve efficiently \PV{instead}.
+Despite that, we also tried to design some automated heuristics during our work on the \textsc{Ingrid} project, but so far all yield rather average results especially when compared to what human users can achieve efficiently \PV{instead}.

### 2016-11-09 chapters/introduction.tex
small changes

@@
 
-In this paper we present INGRID, a method and a tool for semi-automated construction of the definition of a programming language in MPS that uses an ANTLR grammar of the language as a description of its syntax.
-We discuss the main challenges that we encountered during the development of INGRID and our solution to them.
+In this paper we present \textsc{Ingrid}, a method and a tool for semi-automated construction of the definition of a programming language in MPS that uses an ANTLR grammar of the language as a description of its syntax.
+We discuss the main challenges that we encountered during the development of \textsc{Ingrid} and our solution to them.
 Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges are general. 
@@
 
-The INGRID tool can help also to make the construction of DSLs more efficient.
-Based on our experience, we believe that, especially for simple languages, it is less time consuming to write their grammar by hand and then create the language definition in MPS using INGRID, than to create the language manually using the MPS GUI.
+The \textsc{Ingrid} tool can help also to make the construction of DSLs more efficient.
+Based on our experience, we believe that, especially for simple languages, it is less time consuming to write their grammar by hand and then create the language definition in MPS using \textsc{Ingrid}, than to create the language manually using the MPS GUI.
 
-%INGRID allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
+%\textsc{Ingrid} allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
 
 The rest of the paper has the following structure.
-In the next section, we provide necessary information about MPS and a brief overview of projects with similar goals as INGRID.
+In the next section, we provide necessary information about MPS and a brief overview of projects with similar goals as \textsc{Ingrid}.
 Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
-We also discuss our experience with usage of INGRID on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.
+We also discuss our experience with usage of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.

### 2016-11-09 chapters/language_definition.tex
small changes

@@
 
-The proposed INGRID method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
+The proposed \textsc{Ingrid} method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
 All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
@@
 
-The process of MPS language construction, as performed by INGRID, consists of four phases --- the first is parsing of the input grammar, which is followed by definition of the essential aspects of the language (Structure, Editor, and TextGen).
+The process of MPS language construction, as performed by \textsc{Ingrid}, consists of four phases --- the first is parsing of the input grammar, which is followed by definition of the essential aspects of the language (Structure, Editor, and TextGen).
 Each of the phases 2-4 is exclusively and fully responsible for one aspect.
 
-INGRID is currently able to create (i) a full Structure aspect for each element (type of an AST node) of the given language, (ii) a very basic Editor, and (iii) a basic TextGen aspect.
-Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the human users must also define the remaining aspects not yet supported by INGRID.
+\textsc{Ingrid} is currently able to create (i) a full Structure aspect for each element (type of an AST node) of the given language, (ii) a very basic Editor, and (iii) a basic TextGen aspect.
+Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the human users must also define the remaining aspects not yet supported by \textsc{Ingrid}.
 
@@
 
-We will describe the INGRID method using the example of a simplified XML language that is defined in Figure~\ref{fig:SIMPLEXML}.
+We will describe the \textsc{Ingrid} method using the example of a simplified XML language that is defined in Figure~\ref{fig:SIMPLEXML}.
 The \emph{SimpleXML} language is small but complex enough to be used for illustration of the main challenges and behavior of the proposed algorithms.
@@
 It is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
-Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the INGRID algorithm.
+Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the \textsc{Ingrid} algorithm.
 In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST are performed in this phase.
@@
 
-As a part of the INGRID method, we have designed an approach to eliminate the unnecessary layers during construction of the Structure aspect --- we call it the \emph{shortcut} approach (Section~\ref{sect:SHORTCUT}).
+As a part of the \textsc{Ingrid} method, we have designed an approach to eliminate the unnecessary layers during construction of the Structure aspect --- we call it the \emph{shortcut} approach (Section~\ref{sect:SHORTCUT}).
 However, before focusing on the intermediate layers and the shortcut approach, we describe the basic principles of translation from the AST into the language structure.
@@
 
-Our solution that we implemented in the current version of the INGRID method is to create all the cells and place them in a single row.
+Our solution that we implemented in the current version of the \textsc{Ingrid} method is to create all the cells and place them in a single row.
 The resulting basic layout is illustrated on the example of the \mpsconcept{Element{\_}1} concept in the upper left corner of Figure~\ref{fig:EDITORADJUST}.
@@
 Based on our experience, it takes a very short time to manually adjust the layout into a form much better than any fully automated heuristic could achieve.
-We expect that a typical user of the INGRID method will be able to use the projectional editor in MPS quite efficiently.
+We expect that a typical user of the \textsc{Ingrid} method will be able to use the projectional editor in MPS quite efficiently.
 More details are provided in Section~\ref{sect:EVAL}, including our experience with several mainstream programming languages.
@@
 
-Our INGRID method targets mostly text-based languages, for the concepts of which the visual representation in a projectional editor must be almost equivalent to their plain-text representation.
+Our \textsc{Ingrid} method targets mostly text-based languages, for the concepts of which the visual representation in a projectional editor must be almost equivalent to their plain-text representation.
 An obvious choice would therefore be to use the same approach for Editor and TextGen.
@@
 
-During our work on the INGRID method, we have also observed that practical usability of a resulting MPS language depends on the specific manner in which the input ANTLR grammar is defined.
+During our work on the \textsc{Ingrid} method, we have also observed that practical usability of a resulting MPS language depends on the specific manner in which the input ANTLR grammar is defined.
 We discuss several issues and our workarounds in this section.
@@
 \paragraph{Adjusting grammars.}
-In some cases, a small adjustment of the input grammar before the run of INGRID might yield a better and more useful MPS language.
+In some cases, a small adjustment of the input grammar before the run of \textsc{Ingrid} might yield a better and more useful MPS language.
 Here we illustrate this on two simple examples.
@@
 \paragraph{Breaking original grammars and parsers.}
-We also want to point out a general problem with grammar adjustments that we encountered, and which all potential users of the INGRID method should be aware of.
+We also want to point out a general problem with grammar adjustments that we encountered, and which all potential users of the \textsc{Ingrid} method should be aware of.
 The original ANTLR grammar for an input language can be very easily changed in a way that, at first, seems harmless and valid inside MPS, but the parser generated out of the adjusted grammar stops accepting the original language.
@@
 This problem has two main causes: (1) low-level implementation details of token matching in the ANTLR parser and (2) usage of ANTLR grammars for a different purpose (that was not expected) in our project.
-Since, for practical reasons, it is very important that a parser works even for documents (programs) written according to the original grammar, adjustments have to be performed very carefully by the users of INGRID.
+Since, for practical reasons, it is very important that a parser works even for documents (programs) written according to the original grammar, adjustments have to be performed very carefully by the users of \textsc{Ingrid}.

### 2016-11-09 paper.tex
new latex class
small changes
abstract

@@
-\documentclass[10pt]{sigplanconf}
+\documentclass[preprint,10pt,numbers]{sigplanconf}
 
@@
 % TODO
-%\conferenceinfo{PLDI'17,} {June 19--26, 2017, Barcelona, Spain.}
-%\CopyrightYear{2017}
-%\copyrightdata{TODO}
+%\conferenceinfo{PLDI '17}{June 19--26, 2017, Barcelona, Spain}
+%\copyrightyear{20yy}
+%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
+%\copyrightdoi{nnnnnnn.nnnnnnn}
+
+% TODO
+%\publicationrights{licensed}
 
@@
 
-% TODO pridat nekoho dalsiho z JetBrains (Alex Shatalin?)
 % TODO upravit affiliation pro JetBrains -> doplnit "s.r.o" nebo radsi "Inc."
@@
 \begin{abstract}
-% TODO
-\todo{write later}
+JetBrains MPS is a language workbench, an IDE that allows developers both to write code and manipulate language definitions.
+It leverages the concept of projectional editing, where the developer directly manipulates the AST representation of program code.
+While MPS focuses on domain-specific languages (DSL), it needs to support also general-purpose programming languages (GPLs) --- for example, to compile and run programs written in some DSL, or to enable combination of GPLs with languages created in MPS.
 
-\todo{Copy from master thesis.}
-JetBrains MPS is a language workbench focusing on domain-specific languages.
-Unlike many other language workbenches and IDEs, it uses a projectional editor for code.
-The developer directly manipulates the program in its tree form (AST) and not by editing a text source code.
-This brings many advantages, but on the other hand requires time-consuming and complicated MPS language definition.
-The thesis elaborates on the possibility of automating the process of creating MPS language definition from its grammar description.
-It introduces the MPS editor, evaluates approaches of related projects and describes author's efforts to implement an MPS plugin that allows this import.
-The chosen approach and the selection of tools used for implementation are justified in the thesis.
-We point out important problems that any similar project might deal with and we introduce some possible solutions.
-Furthermore, the thesis contains examples of imported languages, showing the potency of the chosen approach.
-The thesis also aims to lay groundwork for future extensions and suggest possible improvements.
+We present \textsc{Ingrid}, a new method for construction of a language definition in MPS based on its ANTLR grammar.
+The structure of a language is generated automatically, but its other aspects --- including the projectional editor --- must be adjusted manually by the user.
+During the development of \textsc{Ingrid}, we encountered several challenges related to the mapping between (i) grammars written in an EBNF-like notation and (ii) the language definitions in MPS based on object-oriented principles.
+Another difficulty is that grammars do not hold any information about code layout.
+
+We implemented \textsc{Ingrid} as a plugin for MPS and evaluated it on several full-fledged programming languages, such as JavaScript and C\#.
+Results show that our approach is practical.
+Even though it is only partially automated, it saves the user from many hours of tedious and error-prone work.
+Necessary manual adjustments take only a short time.
 \end{abstract}
 
-% TODO possible keywords: JetBrains MPS, programming language, grammar, projectional editor
+\category{D.3.2}{Language Classifications}{Extensible languages}
+\category{D.3.4}{Processors}{Parsing}
+\category{D.3.4}{Processors}{Code generation}
+\category{D.2.3}{Coding Tools and Techniques}{Program editors}
+
+\terms
+Languages
+
+\keywords
+language workbench, projectional editor, grammar, domain-specific language, ANTLR, JetBrains MPS
 

### 2016-11-10 paper.tex
Poznamky k abstraktu

@@
 \begin{abstract}
+\PV{Manipulate language definitions mozna uplne neevokuje tvorbu jazyku. Slo by to rict vic polopaticky? Tady mozna pomuze vic Vaclav.}
 JetBrains MPS is a language workbench, an IDE that allows developers both to write code and manipulate language definitions.

### 2016-11-11 paper.tex
JB review of abstract and conclusion

@@
 \PV{Manipulate language definitions mozna uplne neevokuje tvorbu jazyku. Slo by to rict vic polopaticky? Tady mozna pomuze vic Vaclav.}
+\JB{"Manipulate language definitions" - navrhuji "create or modify language definitions"}
 JetBrains MPS is a language workbench, an IDE that allows developers both to write code and manipulate language definitions.
 It leverages the concept of projectional editing, where the developer directly manipulates the AST representation of program code.
-While MPS focuses on domain-specific languages (DSL), it needs to support also general-purpose programming languages (GPLs) --- for example, to compile and run programs written in some DSL, or to enable combination of GPLs with languages created in MPS.
+While MPS focuses on domain-specific languages (DSL), it needs to support also general-purpose programming languages (GPLs) --- for example, to compile and run programs written in some DSL, or to enable combination of GPLs with \JB{Navrhuji nahradit "languages" za "DSLs"} languages created in MPS.
 

### 2016-11-11 chapters/evaluation.tex
addressing comments

@@
 We implemented the proposed \textsc{Ingrid} method as an MPS plugin that allows users to generate languages from ANTLR v4 grammars.
-While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to call \PV{nebo "to bind with"} the MPS API,  \PV{nebo "which was used to programatically generate elements of the new language} e.g.
-in order to programmatically create new language concepts and their aspects.
+While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to bind with the MPS API, which is used to programatically generate elements of the new language and their aspects.
 The plugin uses the ANTLR library~\cite{ref:ANTLR} for parsing of ANTLR v4 grammar files, and several MPS libraries that implement the MPS API.
@@
 
-For the purpose of evaluation, we applied \textsc{Ingrid} to several well-established and widely used languages, including JSON, JavaScript, and C\#.
-\PV{Dalsi vetu bych odebral a treba jen v predchozi vete dal JavaScript (ECMAScript 5.1). Usetri to hodne mista a informace zustane stejna.}
-We used the version of JavaScript that is standardized as ECMAScript 5.1~\cite{ref:ECMASCRIPT51} --- the specification was finalized in June 2011, and it is currently the most frequently adopted version.
+For the purpose of evaluation, we applied \textsc{Ingrid} to several well-established and widely used languages, including JSON, JavaScript (ECMAScript 5.1~\cite{ref:ECMASCRIPT51}), and C\#.
 MPS projects that contain definitions of all three languages are also released in the public repository (\url{https://github.com/premun/ingrid}).
@@
 However, first we must emphasize that MPS languages automatically produced by the \textsc{Ingrid} method, as defined in this paper, are not ready-to-use full-fledged MPS counterparts of the original input languages.
-The structure of such a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects (e.g., the Editor) would have to be manually tweaked (or defined from scratch) by the end user.
-\PV{Nejsem si jisty, jestli je to vubec mozne.
-To, ze nejake pravidlo "ma typ", je asi az moc high-level informace.
-To, ze jedno pravidlo je definice metody a jine je volani metody, take z gramatiky nikdy ziskat nepujde.
-Bud vetu odstranit nebo preformulovat (bez yet atd.)}
-\textsc{Ingrid} does not yet support advanced features of MPS, such as the aspect responsible for type checking and inference.
-\PV{Toto take neni pravda - editor je v kazdem jazyku lehce upraveny (to je tech 20 min / 1 hodina ...). Structure aspect je stejny, textgen je take krome Csharpu upraveny.}
-For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition exactly in the form created by \textsc{Ingrid} and (2) the adjusted ANTLR v4 grammar used as input for \textsc{Ingrid}, both in the repository that contains also our implementation.
+The structure of such a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects (e.g., the Editor) have to be manually tweaked (or defined from scratch) by the end user.
+\textsc{Ingrid} also does not support advanced features of MPS, such as type checking.
+
+For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition in the form that incorporates all tweaks performed by authors of this paper and (2) the adjusted ANTLR v4 grammar used as input for \textsc{Ingrid}, both in the repository that contains also our implementation.
 
@@
 \paragraph{JavaScript.}
-\PV{Poznamka vyse o ECMAScript 5.1 by sla take prenest sem.}
 In the case of JavaScript, which is an example of a widely-used complex general purpose language, automated generating of the language definition in MPS from the ANTLR grammar\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} and subsequent manual adjusting was done in less than one hour.
@@
 \textsc{Ingrid} achieved a very good result especially in the case of language structure, concept aliases, and support for auto-completion.
-The Structure aspect generated by \textsc{Ingrid} is very similar to that of the ECMAScript4MPS project, which was created manually \PV{over misto in?} in a large number of hours.
+The Structure aspect generated by \textsc{Ingrid} is very similar to that of the ECMAScript4MPS project, which was created manually over a large number of hours.
 
@@
 Another very complex programming language that we used for experiments is C\#.
-Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces a reasonably useful MPS language.
-\PV{Na gramatiku asi stacilo 20 minut, kdyz clovek vi, co dela. 40 minut pak na castecnou upravu editoru.}
-Roughly one hour of manual effort was needed for this step.
+Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces at least a reasonable structure of the MPS language.
+The Editor aspect was modified afterwards, but only for some of the concepts generated by \textsc{Ingrid}. 
+Roughly one hour of manual effort was needed in total.
 Nevertheless, additional modifications of the grammar and selected aspects of the MPS language (Editor, TextGen) are still needed to get an optimal result.
-
-\PV{Tato veta asi neni az tak dulezita.
-Spis mi tu chybi neco ve smyslu, ze Csharp je opravdu flexibilni jazyk, diky cemuz je import nejdale od pouzitelneho MPS jazyka (prakticky vsechny statementy lze rozvadet pomoci generiky, pretypovani atd atd).}
-The generated definition of C\# in MPS is quite large, involving more than 800 concepts, and therefore building of the language takes a rather long time.
+A great flexibility and complexity of the C\# language is the main reason behind all the necessary changes to the language definition in MPS and to the ANTLR grammar.
+The definition is quite large, involving more than 800 concepts, and therefore building of the language takes a rather long time.
 
@@
 Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with \textsc{Ingrid}.
-\PV{Tim not compatible je spis myslena hiearchizace gramatiky a styl rozdeleni pravidel.
-Byl trochu jiny nez jinde a dochazelo tam k dost zmatenym strukturam (co se MPS tyce).
-Cyklicka pravidla jsou asi nejmensi problem, daji se odstranit prakticky hned a vetsinou celkem bez ujmy.}
-For example, \textsc{Ingrid} cannot handle cyclic rules (which appear in the Python grammar) and just prints a warning on the screen.
-Significant modifications of the grammar would be necessary.
-\PV{Mozna to ani neni jak adresovat, proste se bude muset upravit gramatika.}
+For example, the structure and hierarchy of rules in the Python grammar are quite different from JavaScript or JSON, and consequently the language definition created by \textsc{Ingrid} is rather badly organized.
 We plan to address this limitation in the future.
+However, significant modifications of the grammar would probably be necessary anyway.
 
@@
 All of this justifies some of our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
-Despite that, we also tried to design some automated heuristics during our work on the \textsc{Ingrid} project, but so far all yield rather average results especially when compared to what human users can achieve efficiently \PV{instead}.
+Despite that, we also tried to design some automated heuristics during our work on the \textsc{Ingrid} project, but so far all yield rather average results when compared to what human users can achieve efficiently instead.

### 2016-11-11 paper.tex
addressing comments

--- 2016-11-09 paper.tex
@@
 \begin{abstract}
-JetBrains MPS is a language workbench, an IDE that allows developers both to write code and manipulate language definitions.
+JetBrains MPS is a language workbench, an IDE that allows developers both to write code and create or modify language definitions.
 It leverages the concept of projectional editing, where the developer directly manipulates the AST representation of program code.
-While MPS focuses on domain-specific languages (DSL), it needs to support also general-purpose programming languages (GPLs) --- for example, to compile and run programs written in some DSL, or to enable combination of GPLs with languages created in MPS.
+While MPS focuses on domain-specific languages (DSL), it needs to support also general-purpose programming languages (GPLs) --- for example, to compile and run programs written in some DSL, or to enable combination of GPLs with DSLs created in MPS.
 

### 2016-11-11 chapters/introduction.tex
references

@@
 
-An alternative approach is projectional (or structural) editing, where a developer directly manipulates the AST representation of the source code instead of plain text.
+An alternative approach is projectional (or structural) editing~\cite{ref:VWK15}, where a developer directly manipulates the AST representation of the source code instead of plain text.
 This idea emerged as early as in 1970s, but it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
 A recent revival of projectional editing has been observed in the area of language workbenches --- IDE-like tools that enable the developers to manipulate the actual language definition.
-The Meta Programming System (MPS)~\cite{ref:MPS} by JetBrains is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
+The Meta Programming System (MPS)~\cite{ref:MPS,ref:MPSBOOK} by JetBrains is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
 MPS provides the whole IDE infrastructure that enables developers to design custom languages, use them to write program code, and compile the code into executables.

### 2016-11-12 chapters/background.tex
polishing

@@
 
-JetBrains MPS is a complete language workbench --- an integrated development environment that allows developers to create their own languages and use them to write code.
+JetBrains MPS is a language workbench --- an IDE that allows developers to create their own languages and use them to write code.
 The code can then be transformed into a target language, typically a GPL, such as Java or C, and eventually compiled into executable programs.
@@
 The fundamental aspect of any MPS language is Structure.
-It must be created first for each intended element (concept) of the language.
-Structure specifies core attributes of an MPS concept such as the name, inheritance relationships, possible child concepts (including their types and cardinalities), the implemented interfaces, and references to other AST nodes.
+It must be created first for each intended language element (concept).
+Structure specifies core attributes of an MPS concept such as the name, inheritance relationships, child concepts (including their types and cardinalities), the implemented interfaces, and references to other AST nodes.
 % other properties (fields of any type that can hold values), etc.
-Figure~\ref{fig:if_statement_structure} shows definition of the Structure aspect for the \code{if-else} statement.
+Figure~\ref{fig:if_statement_structure} shows definition of the Structure aspect for the \code{if-then-else} statement.
 
-The definition of Structure restricts the type of AST nodes that can appear at a particular place in the tree.
-For example, one can restrict the condition in the \code{if-then-else} statement to be a boolean expression, and the \code{then}-block to be a list of statements.
+Structure also restricts the type of AST nodes that can appear at a particular place in the tree.
+For example, one can restrict the condition in \code{if-then-else} to be a boolean expression, and the \code{then}-block to be a list of statements.
 
@@
 \end{alltt}
-\caption{Structure aspect of the \code{if-else} statement}
+\caption{Structure aspect of the \code{if-then-else} statement}
 \label{fig:if_statement_structure}
@@
 
-Figure~\ref{fig:if_editor_definition} shows an example of what the definition of the Editor aspect for the \code{if-else} statement might look like.
+Figure~\ref{fig:if_editor_definition} shows an example of what the definition of the Editor aspect for the \code{if-then-else} statement might look like.
 While here we indicate positions of cell borders on each line by spaces (just for the purpose of illustration), MPS GUI actually uses a graphically much more appealing way of displaying the Editor aspects, which involves vertical and horizontal lines of different colors and also background colors other than white for some cells.
@@
 
-The BaseLanguage~\cite{ref:BASELANG} is an almost full port of the Java language, extended with MPS-specific features.
-It was created manually by JetBrains, and it is still undergoing changes as Java itself is evolving.
-
+The BaseLanguage~\cite{ref:BASELANG} is an almost full port of Java, which was created manually by JetBrains and extended with MPS-specific features.
+% and it is still undergoing changes as Java itself is evolving.
 The C language has also been manually tailored for MPS within the mbeddr project~\cite{ref:MBEDDR}.
@@
 
-PE grammars extend the syntax of the ANTLR v4 notation by custom constructs that provide information about the possible layout of AST nodes.
+PE grammars extend the syntax of the ANTLR v4 notation by constructs that provide information about the possible layout of AST nodes.
 The current version, as of November 2016, supports just horizontal lists, vertical lists, and some indentation rules.
-However, even these few features make the already complex syntax of ANTLR v4 much more complicated.
+However, even these few features make the syntax much more complicated.
 
@@
 We provide only a brief mention here because this project is in an early stage of development.
-Its author created an ANTLRv4 MPS language, which captures the syntax of ANTLRv4 notation using MPS concepts.
-Given some language as input, the textual grammar of the language is imported automatically into MPS, taking the form of the ANTLRv4 MPS language.
+Its author created an ANTLR v4 MPS language, which captures the syntax of ANTLR v4 notation using MPS concepts.
+Given some language as input, the textual grammar of the language is imported automatically into MPS, taking the form of the ANTLR v4 MPS language.
 The next step would be to create a new MPS language from the grammar definition in MPS, but it is not implemented yet.
@@
 The mps-metabnf tool~\cite{ref:MPSMETABNF}, implemented by members of the DSLFoundry group, builds upon similar ideas as the ANTLR{\_}MPS project described above.
-Authors of this project, too, created an MPS language that can be used to capture ANTLR grammars.
-Every imported grammar is represented using this ANTLRv4 MPS language.
+Authors of this project, too, created an MPS language to capture ANTLR grammars.
+Every imported grammar is represented using this ANTLR v4 MPS language.
 A user is then able to adjust the grammar as he wishes, specifying the code layout in the projectional editor.
@@
 
-The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLRv4 MPS language into the definition of the actual target MPS language.
+The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLR v4 MPS language into the definition of the actual target MPS language.

### 2016-11-12 chapters/evaluation.tex
references
polishing

@@
 
-For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition in the form that incorporates all tweaks performed by authors of this paper and (2) the adjusted ANTLR v4 grammar used as input for \textsc{Ingrid}, both in the repository that contains also our implementation.
+For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition in the form that incorporates all tweaks performed manually by authors of this paper and (2) the adjusted ANTLR v4 grammar used as input for \textsc{Ingrid}, both in the repository that contains also our implementation.
 
@@
 \paragraph{C\#.}
-Another very complex programming language that we used for experiments is C\#.
+Another very complex programming language that we used for experiments is C\#~\cite{ref:CSHARP}.
 Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces at least a reasonable structure of the MPS language.
@@
 
-On the other hand, our experiments with complex languages show that, in the case of the Editor and Textgen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, much less time than we expected.
+On the other hand, our experiments with complex languages show that, in the case of the Editor and TextGen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, much less time than we expected.
 The first author spent between 20 and 60 minutes of work on each language, when adjusting the result of automated generation into a more useful and readable form.

### 2016-11-12 chapters/introduction.tex
polishing

@@
 The prevailing approach to define valid syntaxes for programming languages is through grammars, which are typically written in notations based on the Extended Backus-Naur Form (EBNF).
-Many popular tools exist for automated generation of parsers from the grammar definitions --- for example, ANTLR~\cite{ref:ANTLRBOOK,ref:ANTLR} and Bison/Flex~\cite{ref:BISONFLEX}.
+Many tools exist for automated generation of parsers from the grammar definitions --- for example, ANTLR~\cite{ref:ANTLRBOOK,ref:ANTLR} and Bison/Flex~\cite{ref:BISONFLEX}.
 The source code of programs is written in a text editor, usually embedded into an IDE. Parsers are used to create an abstract syntax tree (AST) representation of the code, e.g. inside the compiler.
@@
 
-Projectional editing, keeping the code in the AST form, and in particular the absence of parsers, brings along several benefits.
+Projectional editing, keeping the code in the AST form, and the absence of parsers, brings along several benefits.
 \begin{itemize}
 	\item The developers practically cannot enter syntactically invalid code, because a projectional editor controls the interaction between the user and the code.
-An AST created by the editor can represent only a syntactically valid code.
+		%An AST created by the editor can represent only a syntactically valid code.
 	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together or one can extend another, while parsers seriously limit language modularity for traditional parsed languages.
@@
 On the other hand, the usage of projectional editors introduces some new problems.
-Before a language can be used inside MPS, it has to be defined in a special way through the MPS infrastructure, so that the IDE can understand the language and work with it.
+Before a language can be used inside MPS, it has to be defined through the MPS infrastructure, so that the IDE can understand the language and work with it.
 Specifically, an author of a language has to create:

### 2016-11-12 chapters/language_definition.tex
polishing

@@
 
-The process of MPS language construction, as performed by \textsc{Ingrid}, consists of four phases --- the first is parsing of the input grammar, which is followed by definition of the essential aspects of the language (Structure, Editor, and TextGen).
-Each of the phases 2-4 is exclusively and fully responsible for one aspect.
+The process of MPS language construction, as performed by \textsc{Ingrid}, consists of four phases --- the first is parsing of the input grammar, which is followed by definition of the essential aspects of the language (Structure, Editor, and TextGen, respectively).
+%Each of the phases 2-4 is exclusively and fully responsible for one aspect.
 
@@
 \begin{alltt}
+\small
 \textbf{grammar} \textit{SimpleXML};
@@
 
-Every tree that represents a lexer rule is flattened into the equivalent regular expression by the recursive algorithm specified in Figure~\ref{fig:ALGFLATTEN}.
-A distinct sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.
+Every tree that represents a lexer rule is flattened into the equivalent regular expression by the recursive algorithm in Figure~\ref{fig:ALGFLATTEN}.
+A sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.
 
@@
 \begin{alltt}
+\small
 Flatten(\textit{R}):
@@
 As a part of the \textsc{Ingrid} method, we have designed an approach to eliminate the unnecessary layers during construction of the Structure aspect --- we call it the \emph{shortcut} approach (Section~\ref{sect:SHORTCUT}).
-However, before focusing on the intermediate layers and the shortcut approach, we describe the basic principles of translation from the AST into the language structure.
+However, before focusing on the intermediate layers, we describe the basic principles of translation from the AST into the language structure.
 
@@
 The object contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name}.
-Values of these properties will be restricted using the regular expression that corresponds to the lexer rule \antlrlexerrule{Name}.
+Values of these properties will be restricted using the regular expression that corresponds to the rule.
 String literals, such as the opening and closing brackets (\antlrliteral{\textless}, \antlrliteral{/\textgreater}) in XML, are omitted because they will be defined only in the Editor aspect for this concept.
@@
 \begin{alltt}
+\small
   \antlrparserrule{content} : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
@@
 \begin{framed}
+\small
 \textbf{Content{\_}1} (TEXT) \\
@@
 \begin{alltt}
+\small
  1 FindAllPathsToEndNodes(\textit{R}):
@@
 \begin{itemize}
-	\item When a particular alternative $A$ of the rule $R$ contains only a single element $E$, and the element is a reference to another parser rule (line 9), the alternative $A$ is an intermediary layer that can be transparently hidden from the user of the MPS language.
+	\item When a particular alternative $A$ of the rule $R$ contains only a single element $E$ that is a reference to another rule (line 9), the alternative $A$ is an intermediary layer that can be hidden from the user of the MPS language.
 		  A run of the algorithm continues, at line 11, by recursively processing alternatives of the rule corresponding to $E$, which lies at the next level of the parser tree.
@@
 \begin{alltt}
+\small
   \antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
@@
 \begin{alltt}
+\small
   \antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
@@
 \paragraph{Breaking original grammars and parsers.}
-We also want to point out a general problem with grammar adjustments that we encountered, and which all potential users of the \textsc{Ingrid} method should be aware of.
+We also want to point out a general problem with grammar adjustments, which all potential users of the \textsc{Ingrid} method should be aware of.
 The original ANTLR grammar for an input language can be very easily changed in a way that, at first, seems harmless and valid inside MPS, but the parser generated out of the adjusted grammar stops accepting the original language.
@@
 This problem has two main causes: (1) low-level implementation details of token matching in the ANTLR parser and (2) usage of ANTLR grammars for a different purpose (that was not expected) in our project.
-Since, for practical reasons, it is very important that a parser works even for documents (programs) written according to the original grammar, adjustments have to be performed very carefully by the users of \textsc{Ingrid}.
+Since, for practical reasons, it is very important that a parser works for programs written according to the original grammar, adjustments have to be performed carefully by the users of \textsc{Ingrid}.

### 2016-11-12 paper.tex
references
polishing

@@
 \newcommand{\antlrparserrule}[1]{\textcolor{Blue}{#1}}
-\newcommand{\antlrlexerrule}[1]{\textcolor{RedOrange}{#1}}
+\newcommand{\antlrlexerrule}[1]{\textcolor{BrickRed}{#1}}
 \newcommand{\antlrliteralnoap}[1]{\textcolor{Maroon}{#1}}
 \newcommand{\antlrliteral}[1]{\antlrap\antlrliteralnoap{#1}\antlrap}
-\newcommand{\antlrregex}[1]{\textcolor{PineGreen}{#1}}
+\newcommand{\antlrregex}[1]{\textcolor{OliveGreen}{#1}}
   % MPS in general
@@
 \newcommand{\mpsstplaceholder}[1]{\textcolor{Gray}{\textbf{#1}}}
-\newcommand{\mpsstcardinality}[1]{\textcolor{RoyalBlue}{#1}}
+\newcommand{\mpsstcardinality}[1]{\textcolor{RoyalBlue}{\textbf{#1}}}
   % MPS editor
 \newcommand{\mpsedkeyword}[1]{\textcolor{Black}{\textbf{#1}}}
-\newcommand{\mpsedtarget}[1]{\textcolor{Fuchsia}{#1}}
+\newcommand{\mpsedtarget}[1]{\textcolor{RubineRed}{\textbf{#1}}}
 \newcommand{\mpsedannotation}[1]{\textcolor{Gray}{\textbf{#1}}}
@@
 \newcommand{\mpstgkeyword}[1]{\textcolor{Blue}{\textbf{#1}}}
-\newcommand{\mpstgtarget}[1]{\textcolor{Fuchsia}{#1}}
-\newcommand{\mpstgparam}[1]{\textcolor{Violet}{#1}}
+\newcommand{\mpstgtarget}[1]{\textcolor{RubineRed}{\textbf{#1}}}
+\newcommand{\mpstgparam}[1]{\textcolor{RoyalPurple}{\textbf{#1}}}
 \newcommand{\mpstgaction}[1]{\textcolor{Fuchsia}{\textbf{#1}}}
 \newcommand{\mpstgliteral}[1]{\textcolor{Green}{\textbf{#1}}}
-\newcommand{\mpstgnodeprop}[1]{\textcolor{darkgray}{#1}}
+\newcommand{\mpstgnodeprop}[1]{\textcolor{darkgray}{\textbf{#1}}}
 
@@
 
-% TODO upravit affiliation pro JetBrains -> doplnit "s.r.o" nebo radsi "Inc."
-% TODO muzeme taky napsat jejich emaily pokud budou lidi z JetBrains souhlasit
 \authorinfo{V\'aclav Pech}
 	{JetBrains}
+	{vaclav@jetbrains.com}
 
@@
 
-We present \textsc{Ingrid}, a new method for construction of a language definition in MPS based on its ANTLR grammar.
-The structure of a language is generated automatically, but its other aspects --- including the projectional editor --- must be adjusted manually by the user.
-During the development of \textsc{Ingrid}, we encountered several challenges related to the mapping between (i) grammars written in an EBNF-like notation and (ii) the language definitions in MPS based on object-oriented principles.
+We present \textsc{Ingrid}, a method for construction of a language definition in MPS based on its ANTLR grammar.
+The structure of a language is generated automatically, but its projectional editor and other aspects must be adjusted manually by the user.
+During the development of \textsc{Ingrid}, we encountered several challenges related to the mapping between grammars and language definitions in MPS that are based on object-oriented principles.
 Another difficulty is that grammars do not hold any information about code layout.
 
-We implemented \textsc{Ingrid} as a plugin for MPS and evaluated it on several full-fledged programming languages, such as JavaScript and C\#.
-Results show that our approach is practical.
-Even though it is only partially automated, it saves the user from many hours of tedious and error-prone work.
+We implemented \textsc{Ingrid} as a plugin for MPS and evaluated it on several mainstream GPLs, such as JavaScript and C\#.
+Results show that our approach is practical, and it saves the user from many hours of tedious and error-prone work.
 Necessary manual adjustments take only a short time.
@@
 \acks
-% TODO
-This research was supported by JetBrains. \todo{Pridat jeste nejaky grant a kompletne preformulovat.}
+% TODO mozna pridat jeste nejaky dalsi grant (az clanek nekam vezmou)
+This work was partially supported by JetBrains.
 
@@
 \begin{thebibliography}{}
+
+\bibitem{ref:MPSBOOK} F. Campagne. The MPS Language Workbench, Volume I. CreateSpace Independent Publishing Platform, 2014.
 
@@
 
+\bibitem{ref:VWK15} M. Voelter, J. Warmer, and B. Kolb. Projecting a Modular Future. IEEE Software, 32(5), 2015.
+
+\bibitem{ref:CSHARP} C\# Language Specification, Version 5.0 (2012), Microsoft Corporation, \url{https://msdn.microsoft.com/en-us/library/ms228593.aspx}
+
+\bibitem{ref:ECMASCRIPT51} ECMAScript Language Specification, Edition 5.1 (June 2011), \url{http://www.ecma-international.org/ecma-262/5.1/Ecma-262.pdf}
+
+\bibitem{ref:JSON} The JSON Data Interchange Format, 1st Edition (October 2013), \url{http://www.ecma-international.org/publications/files/ECMA-ST/ECMA-404.pdf}
+
+\bibitem{ref:MPS} JetBrains MPS, \url{https://www.jetbrains.com/mps/}
+
 \bibitem{ref:ANTLR} ANTLR parser generator, \url{http://www.antlr.org}
@@
 \bibitem{ref:MPSMETABNF} DSLFoundry. The mps-metabnf project, \url{https://github.com/DSLFoundry/mps-metabnf}
-
-\bibitem{ref:ECMASCRIPT51} ECMAScript Language Specification, Edition 5.1 (June 2011), \url{http://www.ecma-international.org/ecma-262/5.1/Ecma-262.pdf}
-
-\bibitem{ref:MPS} JetBrains MPS, \url{https://www.jetbrains.com/mps/}
-
-\bibitem{ref:JSON} The JSON Data Interchange Format, 1st Edition (October 2013), \url{http://www.ecma-international.org/publications/files/ECMA-ST/ECMA-404.pdf}
 

### 2017-06-05 chapters/background.tex
todo

@@
 \label{sect:MPS}
+
+\todo{For each MPS term (Structure, Editor, TextGen, concept, and so on) say which generally/widely known terms (from related academic work) are similar/analogous.
+Relate the MPS-specific terms to the large body of academic work.}
 
@@
 Code of any program in MPS is built from nodes, which represent instances of concepts defined in one of the languages that the program declares to be using.
-In the context of MPS, a concept is, in fact, a building block of a language definition.
+\todo{zvyraznit slovo "concept"} In the context of MPS, a concept is, in fact, a building block of a language definition.
 We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.
@@
 The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLR v4 MPS language into the definition of the actual target MPS language.
+
+\todo{
+S. Erdweg, T. Rendel, C. K\"{a}stner, and K. Ostermann. Layout-Sensitive Generalized Parsing. In Proceedings of SLE 2012, LNCS 7745.
+	Source code layout specified declaratively through constraints on shape and relative positions of syntax trees (AST).
+	Layout constraints are represented/encoded using annotations on production rules of context-free grammars.
+	Useful for layout-sensitive programming language (e.g., Python).
+	Their proposed approach: grammar rule applies only if the input satisfied the constraint (on shape and positioning of source code fragments).
+		Parsing algorithm that accepts only inputs with valid layout (that satisfies the constraints attached to grammar rules).
+	Constraints support (can express): indentation, number of line breaks, alignment.
+	Hand-coded approach used in Python (how the annotated grammar is processed) --- mentioned for comparison.
+		Special tokens that indicate layout aspects (indentation, alignment) are added to stream of tokens by preprocessor.
+		Layout handled by algorithm that works with syntax trees (ASTs) and token streams.
+	The main difference from our approach (INGRID): layout inferred automatically (via automated learning) from several input files.
+		No layout annotations in the grammar are used.
+		Currently, simple heuristics are used to create editor and textgen.
+}

### 2017-06-05 chapters/conclusion.tex
small changes
todo

@@
 
-Our contribution is the INGRID method for constructing language definitions in MPS from their grammars in the ANTLR notation.
+Our contribution is the \textsc{Ingrid} method for constructing language definitions in MPS from their grammars in the ANTLR notation.
 The method is only partially automated, but nevertheless it greatly reduces the amount of manual work that developers of MPS languages have to perform.
@@
 The information about code layout, once it is available, will be then leveraged and used also to improve TextGen.
-Another subject for future work is generating parsers that would allow users to import existing source code into MPS.
+Another subject for future work is generating parsers that would allow users to import existing source code into MPS \todo{Vysvetlit podrobneji ze chceme automaticky generovat parsery textovych zdrojaku ktere budou vytvaret interni reprezentace (ASTs) postavene na zaklade MPS concepts pro dany jazyk}.

### 2017-06-05 chapters/introduction.tex
todo

@@
 We discuss the main challenges that we encountered during the development of \textsc{Ingrid} and our solution to them.
-Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges are general. 
+\todo{Zmenit: we explain our tool using examples from MPS and ANTLR} Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges are general.
 They apply to the relation between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.

### 2017-06-05 chapters/language_definition.tex
todo

@@
 The proposed \textsc{Ingrid} method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
+\todo{Zmenit nasledujici vetu: There exist ANTLR v4 grammars for all widely used programming languages ...}
 All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
@@
 A sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.
+\todo{Algorithm seems to fail if the grammar has right-recursion (according to one reviewer) --- check that and describe how we handle that case.}
 
@@
 However, presence of the intermediate layers significantly complicates usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
-We show examples illustrating this problem, which we call a \emph{layer problem}, later in the subsection where we describe our solution.
+We show examples illustrating this problem, which we call a \emph{layer problem}, later in the subsection where we describe our solution \todo{Pridat odkaz na konkretni section a opravit gramatickou chybu}.
 
@@
 Such nodes cannot expand further based on any rule in the grammar, and for that reason we call them \emph{end nodes}.
+\todo{Srovnat pojem "end nodes" vzhledem ke terminal symbols from grammars (jako analogie).
+Mozna existuje dalsi vhodny termin.}
 
@@
 
-We experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results.
+We experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results \todo{Give examples or quantify this result}.
 However, we observed that the most tedious and error-prone step in the manual definition of the Editor aspect is the creation of cells for all literals (keywords), properties, children, and other fields of a given concept that should appear in its visual representation.
@@
 We have found that the combination of two steps, (1) automated placing of cells into a single row and (2) subsequent manual adjustment of the layout, is a very fast and efficient way of creating a nice code layout.
-Nevertheless, we plan to work on a more automated approach to the definition of Editor aspects in the future, probably using some techniques of machine learning.
+Nevertheless, we plan to work on a more automated approach to the definition of Editor aspects in the future, using automated learning.
 The key idea would be to derive the code layout from a set of valid input source files.
@@
 Nevertheless, since the Editor aspect has to be adjusted manually, we decided to use a different approach for TextGen.
-We designed a procedure based on a simple fully automated heuristic that provides surprisingly good results.
+We designed a procedure based on a simple fully automated heuristic that provides surprisingly good results \todo{Quantify this}.
 
@@
 
+\todo{Figure 15: whitespace is just blank space instead of quoted string, mozna jsem neco smazal omylem (za "with ").}
 Figure~\ref{fig:TEXTGENFINAL} contains the complete, automatically generated, TextGen aspect for the \mpsconcept{Element{\_}1} concept, which represents the full SimpleXML element.

### 2017-06-06 chapters/conclusion.tex
more todos

@@
 Our contribution is the \textsc{Ingrid} method for constructing language definitions in MPS from their grammars in the ANTLR notation.
-The method is only partially automated, but nevertheless it greatly reduces the amount of manual work that developers of MPS languages have to perform.
-Although we discussed all the challenges, details of our solution and general observations just in the context of ANTLR grammars and JetBrains MPS, we believe they are more general --- and in particular relevant to everyone tackling a similar problem in the field of DSLs and language workbenches.
+The method is only partially automated, but nevertheless it greatly reduces the amount of manual work \todo{Jak to vime?} that developers of MPS languages have to perform.
+Although we discussed all the challenges, details of our solution and general observations just in the context of ANTLR grammars and JetBrains MPS, we believe they are more general \todo{Discuss this more thoroughly (it would be valuable)} --- and in particular relevant to everyone tackling a similar problem in the field of DSLs and language workbenches.
 
-In the future, we would like to address the remaining open problems (there are many of them) and increase the level of automation.
-Our primary research goal is to investigate the usage of machine learning for the purpose of generating editors that support nice and readable code layout.
+In the future, we would like to address the remaining open problems (there are many of them) \todo{Provide more concrete details here} and increase the level of automation.
+Our primary research goal is to investigate the usage of machine learning for the purpose of generating editors that support nice and readable code layout \todo{User study would be valuable here, because such a thing may not exist among multiple users}.
 The information about code layout, once it is available, will be then leveraged and used also to improve TextGen.

### 2017-06-06 chapters/evaluation.tex
more todos

@@
 
-The main advantage of \textsc{Ingrid} over ECMAScript4MPS is that, despite its current limitations, \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful.
-\textsc{Ingrid} achieved a very good result especially in the case of language structure, concept aliases, and support for auto-completion.
+The main advantage of \textsc{Ingrid} over ECMAScript4MPS is that, despite its current limitations, \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful \todo{Describe what is the adjustment}.
+\textsc{Ingrid} achieved a very good result \todo{Compared to what} especially in the case of language structure, concept aliases, and support for auto-completion.
 The Structure aspect generated by \textsc{Ingrid} is very similar to that of the ECMAScript4MPS project, which was created manually over a large number of hours.
@@
 Another very complex programming language that we used for experiments is C\#~\cite{ref:CSHARP}.
-Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces at least a reasonable structure of the MPS language.
+Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces at least a reasonable structure \todo{Zpresnit, say what is reasonable (or, what is reasonable)} of the MPS language.
 The Editor aspect was modified afterwards, but only for some of the concepts generated by \textsc{Ingrid}. 
 Roughly one hour of manual effort was needed in total.
-Nevertheless, additional modifications of the grammar and selected aspects of the MPS language (Editor, TextGen) are still needed to get an optimal result.
+Nevertheless, additional modifications of the grammar and selected aspects of the MPS language (Editor, TextGen) are still needed to get an optimal result \todo{Say what would be optimal, try to formally define this criteria}.
 A great flexibility and complexity of the C\# language is the main reason behind all the necessary changes to the language definition in MPS and to the ANTLR grammar.
-The definition is quite large, involving more than 800 concepts, and therefore building of the language takes a rather long time.
+The definition is quite large, involving more than 800 concepts, and therefore building of the language takes a rather long time \todo{Lepe vysvetlit, neni jasne "what this is getting at" --- compilation of the language in MPS}.
 
@@
 Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with \textsc{Ingrid}.
+\todo{Explain what style is this.
+Discussing it would be useful for the community, and it could also lead to improvements in ANTLR grammars.}
 For example, the structure and hierarchy of rules in the Python grammar are quite different from JavaScript or JSON, and consequently the language definition created by \textsc{Ingrid} is rather badly organized.
 We plan to address this limitation in the future.
-However, significant modifications of the grammar would probably be necessary anyway.
+However, significant modifications of the grammar would probably be necessary anyway \todo{Justify this --- it seems pessimistic}.
 
@@
 
-On the other hand, our experiments with complex languages show that, in the case of the Editor and TextGen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, much less time than we expected.
+On the other hand, our experiments with complex languages show that, in the case of the Editor and TextGen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, much less time than we expected \todo{Say how long it was (i.e., what time we expected)}.
 The first author spent between 20 and 60 minutes of work on each language, when adjusting the result of automated generation into a more useful and readable form.
-MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing users to reach optimal results very fast.
-All of this justifies some of our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
-Despite that, we also tried to design some automated heuristics during our work on the \textsc{Ingrid} project, but so far all yield rather average results when compared to what human users can achieve efficiently instead.
+MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing users to reach optimal results very fast \todo{Say what is optimal}.
+\todo{The next sentence is vague.
+Say explicitly which decisions are justified by this.
+Other decisions are justified by other means.} All of this justifies some of our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
+Despite that, we also tried to design some automated heuristics during our work on the \textsc{Ingrid} project, but so far all yield rather average results \todo{Average among what} when compared to what human users can achieve efficiently instead.

### 2017-06-07 paper.tex
new format

@@
-\documentclass[preprint,10pt,numbers]{sigplanconf}
+\documentclass[sigplan,10pt,preprint]{acmart}
 
@@
 \usepackage{color}
-\usepackage{times}
 \usepackage{multirow}
@@
 \usepackage{alltt}
-\usepackage{textcomp}
-\usepackage[dvipsnames]{xcolor}
+\usepackage{xcolor}
 
 \setcounter{tocdepth}{3}
+
+\settopmatter{printfolios=true}
 
@@
 \newcommand{\antlrap}{\textquotesingle}
-\newcommand{\antlrparserrule}[1]{\textcolor{Blue}{#1}}
-\newcommand{\antlrlexerrule}[1]{\textcolor{BrickRed}{#1}}
-\newcommand{\antlrliteralnoap}[1]{\textcolor{Maroon}{#1}}
+\newcommand{\antlrparserrule}[1]{\textcolor{blue}{#1}}
+\newcommand{\antlrlexerrule}[1]{\textcolor{red}{#1}}
+\newcommand{\antlrliteralnoap}[1]{\textcolor{orange}{#1}}
 \newcommand{\antlrliteral}[1]{\antlrap\antlrliteralnoap{#1}\antlrap}
-\newcommand{\antlrregex}[1]{\textcolor{OliveGreen}{#1}}
+\newcommand{\antlrregex}[1]{\textcolor{olive}{#1}}
   % MPS in general
-\newcommand{\mpsconcept}[1]{\textcolor{OliveGreen}{#1}}
-\newcommand{\mpsinterface}[1]{\textcolor{Plum}{#1}}
+\newcommand{\mpsconcept}[1]{\textcolor{olive}{#1}}
+\newcommand{\mpsinterface}[1]{\textcolor{violet}{#1}}
   % MPS structure
-\newcommand{\mpsstkeyword}[1]{\textcolor{Blue}{\textbf{#1}}}
-\newcommand{\mpsstalias}[1]{\textcolor{Green}{\textbf{#1}}}
-\newcommand{\mpsstproperty}[1]{\textcolor{Fuchsia}{\textbf{#1}}}
-\newcommand{\mpsstplaceholder}[1]{\textcolor{Gray}{\textbf{#1}}}
-\newcommand{\mpsstcardinality}[1]{\textcolor{RoyalBlue}{\textbf{#1}}}
+\newcommand{\mpsstkeyword}[1]{\textcolor{blue}{\textbf{#1}}}
+\newcommand{\mpsstalias}[1]{\textcolor{olive}{\textbf{#1}}}
+\newcommand{\mpsstproperty}[1]{\textcolor{violet}{\textbf{#1}}}
+\newcommand{\mpsstplaceholder}[1]{\textcolor{gray}{\textbf{#1}}}
+\newcommand{\mpsstcardinality}[1]{\textcolor{cyan}{\textbf{#1}}}
   % MPS editor
-\newcommand{\mpsedkeyword}[1]{\textcolor{Black}{\textbf{#1}}}
-\newcommand{\mpsedtarget}[1]{\textcolor{RubineRed}{\textbf{#1}}}
-\newcommand{\mpsedannotation}[1]{\textcolor{Gray}{\textbf{#1}}}
+\newcommand{\mpsedkeyword}[1]{\textcolor{black}{\textbf{#1}}}
+\newcommand{\mpsedtarget}[1]{\textcolor{magenta}{\textbf{#1}}}
+\newcommand{\mpsedannotation}[1]{\textcolor{gray}{\textbf{#1}}}
   % MPS textgen
-\newcommand{\mpstgkeyword}[1]{\textcolor{Blue}{\textbf{#1}}}
-\newcommand{\mpstgtarget}[1]{\textcolor{RubineRed}{\textbf{#1}}}
-\newcommand{\mpstgparam}[1]{\textcolor{RoyalPurple}{\textbf{#1}}}
-\newcommand{\mpstgaction}[1]{\textcolor{Fuchsia}{\textbf{#1}}}
-\newcommand{\mpstgliteral}[1]{\textcolor{Green}{\textbf{#1}}}
+\newcommand{\mpstgkeyword}[1]{\textcolor{blue}{\textbf{#1}}}
+\newcommand{\mpstgtarget}[1]{\textcolor{magenta}{\textbf{#1}}}
+\newcommand{\mpstgparam}[1]{\textcolor{violet}{\textbf{#1}}}
+\newcommand{\mpstgaction}[1]{\textcolor{violet}{\textbf{#1}}}
+\newcommand{\mpstgliteral}[1]{\textcolor{olive}{\textbf{#1}}}
 \newcommand{\mpstgnodeprop}[1]{\textcolor{darkgray}{\textbf{#1}}}
+
+% TODO
+%\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
+%\acmYear{2017}
+%\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+%\acmDOI{10.1145/nnnnnnn.nnnnnnn}
+
+% TODO
+%\setcopyright{acmlicensed}
 
@@
 
-% TODO
-%\conferenceinfo{PLDI '17}{June 19--26, 2017, Barcelona, Spain}
-%\copyrightyear{20yy}
-%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
-%\copyrightdoi{nnnnnnn.nnnnnnn}
-
-% TODO
-%\publicationrights{licensed}
-
 \title{INGRID: Creating Languages in MPS from ANTLR Grammars}
 
-\authorinfo{P\v{r}emysl Vysok\'y \and Pavel Par\'izek}
-	{Charles University, Faculty of Mathematics and Physics, Prague, Czech Republic}
-	{premek.vysoky@gmail.com, parizek@d3s.mff.cuni.cz}
+\author{P\v{r}emysl Vysok\'y}
+\affiliation{
+  \institution{Charles University, Faculty of Mathematics and Physics}
+  \city{Prague}
+  \country{Czech Republic}
+}
+\email{premek.vysoky@gmail.com}
 
-\authorinfo{V\'aclav Pech}
-	{JetBrains}
-	{vaclav@jetbrains.com}
+\author{Pavel Par\'izek}
+\affiliation{
+  \institution{Charles University, Faculty of Mathematics and Physics}
+  \city{Prague}
+  \country{Czech Republic}
+}
+\email{parizek@d3s.mff.cuni.cz}
 
-\maketitle
+\author{V\'aclav Pech}
+\affiliation{
+  \institution{JetBrains}
+}
+\email{vaclav@jetbrains.com}
 
@@
 
-\category{D.3.2}{Language Classifications}{Extensible languages}
-\category{D.3.4}{Processors}{Parsing}
-\category{D.3.4}{Processors}{Code generation}
-\category{D.2.3}{Coding Tools and Techniques}{Program editors}
+ \begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007.10011006.10011008.10011009.10011019</concept_id>
+<concept_desc>Software and its engineering~Extensible languages</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011041.10011047</concept_id>
+<concept_desc>Software and its engineering~Source code generation</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011066.10011069</concept_id>
+<concept_desc>Software and its engineering~Integrated and visual development environments</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011041.10011688</concept_id>
+<concept_desc>Software and its engineering~Parsers</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
 
-\terms
-Languages
+\ccsdesc[500]{Software and its engineering~Extensible languages}
+\ccsdesc[500]{Software and its engineering~Source code generation}
+\ccsdesc[500]{Software and its engineering~Integrated and visual development environments}
+\ccsdesc[300]{Software and its engineering~Parsers}
 
-\keywords
-language workbench, projectional editor, grammar, domain-specific language, ANTLR, JetBrains MPS
+\keywords{language workbench, projectional editor, grammar, domain-specific language, ANTLR, JetBrains MPS}
+
+\maketitle
 
@@
 
-\acks
+\begin{acks}
 % TODO mozna pridat jeste nejaky dalsi grant (az clanek nekam vezmou)
 This work was partially supported by JetBrains.
+\end{acks}
 

### 2017-06-07 chapters/background.tex
new format
notes about projectional editors

@@
 Relate the MPS-specific terms to the large body of academic work.}
+
+\todo{Napsat sem nebo pripadne do Introduction (ale radsi sem): MPS platform, and tools built upon it (e.g., mbeddr) addresses to a large degree known limitations of projectional editors, such as usability of editors for languages based on textual notation (e.g., non-linear typing) and support for version control [citovat "M.
+Voelter, J.
+Siegmund, T.
+Berger, and B.
+Kolb.
+Towards User-Friendly Projectional Editors.
+SLE 2014, LNCS 8706"].}
+
+\todo{Mozna pouzit termin "language concept" misto samotneho "concept".}
 
@@
   \mpstgparam{(context, buffer, node)->void} \{
-    \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n;}}
+    \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n;}}
     \mpstgaction{indent buffer;}
@@
     \}
-    \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n}} \{\mpstgliteral{\}}\} \$list\{\mpstgparam{node}.\mpstgnodeprop{elsifClauses}\};
+    \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n}} \{\mpstgliteral{\}}\} \$list\{\mpstgparam{node}.\mpstgnodeprop{elsifClauses}\};
     \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{ifFalseStatement}.\textbf{isNotNull}) \{

### 2017-06-07 chapters/introduction.tex
notes about projectional editors

@@
 The mbeddr project~\cite{ref:MBEDDR}, which extends C with numerous domain-specific constructs and data types, illustrates the abilities of language workbenches in general and of MPS in particular.
+\todo{Popsat dalsi benefits of projectional editors/IDEs (and of MPS in particular): Projectional editors never have to parse the concrete syntax, so they can support/use more complicated syntax/layout.
+Support for non-linear notations (mathematical expressions, graphical or tabular layout), notations that use two-dimensional layout.
+Projectional editors mix different notations/styles.}
+\todo{Nekam pridat text: Benefits and limitations of projectional editors are thoroughly discussed in [M.
+Voelter, J.
+Siegmund, T.
+Berger, and B.
+Kolb.
+Towards User-Friendly Projectional Editors.
+SLE 2014, LNCS 8706].}
 

### 2017-06-08 paper.tex
fixed abstract

@@
 It leverages the concept of projectional editing, where the developer directly manipulates the AST representation of program code.
-While MPS focuses on domain-specific languages (DSL), it needs to support also general-purpose programming languages (GPLs) --- for example, to compile and run programs written in some DSL, or to enable combination of GPLs with DSLs created in MPS.
+While MPS focuses on domain-specific languages (DSL), it needs to support also general-purpose programming languages (GPPLs) --- for example, to compile and run programs written in some DSL, or to enable combination of GPPLs with DSLs created in MPS.
 
@@
 
-We implemented \textsc{Ingrid} as a plugin for MPS and evaluated it on several mainstream GPLs, such as JavaScript and C\#.
+We implemented \textsc{Ingrid} as a plugin for MPS and evaluated it on several mainstream GPPLs, such as JavaScript and C\#.
 Results show that our approach is practical, and it saves the user from many hours of tedious and error-prone work.
@@
 <concept>
+<concept_id>10011007.10011006.10011041.10011688</concept_id>
+<concept_desc>Software and its engineering~Parsers</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
 <concept_id>10011007.10011006.10011066.10011069</concept_id>
@@
 <concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10011007.10011006.10011041.10011688</concept_id>
-<concept_desc>Software and its engineering~Parsers</concept_desc>
-<concept_significance>300</concept_significance>
 </concept>
@@
 \ccsdesc[500]{Software and its engineering~Source code generation}
+\ccsdesc[500]{Software and its engineering~Parsers}
 \ccsdesc[500]{Software and its engineering~Integrated and visual development environments}
-\ccsdesc[300]{Software and its engineering~Parsers}
 

### 2017-06-09 chapters/background.tex
notes about spoofax

@@
 }
+
+\todo{
+Mention other state-of-the-art language workbenches: Spoofax.
+	Reference: L.C.L. Kats and E. Visser. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. OOPSLA 2010
+	Eclipse-based IDE support for SDF-based languages.
+	Parser-based language workbench.
+		Syntax essentially restricted to linear text.
+		But it supports language extension and composition.
+}

### 2017-06-10 chapters/background.tex
added notes to intro

@@
 
-In this section, first we introduce basic features of the MPS platform, and then we discuss other existing projects that aim to add support for general purpose languages into MPS.
+In this section, first we introduce basic features of the MPS platform, and then we discuss related work, including other existing projects that aim to add support for general purpose languages into MPS \todo{IDE based on projectional editor}.
 

### 2017-06-10 chapters/introduction.tex
added notes to intro

@@
 
-An alternative approach is projectional (or structural) editing~\cite{ref:VWK15}, where a developer directly manipulates the AST representation of the source code instead of plain text.
+An alternative approach is projectional (or structural) editing~\cite{ref:VWK15} \todo{Pridat citaci: nedavno precteny clanek o srovnani projectional editors}, where a developer directly manipulates the AST representation of the source code instead of plain text.
 This idea emerged as early as in 1970s, but it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
 A recent revival of projectional editing has been observed in the area of language workbenches --- IDE-like tools that enable the developers to manipulate the actual language definition.
-The Meta Programming System (MPS)~\cite{ref:MPS,ref:MPSBOOK} by JetBrains is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
+Popular IDEs (examples) include JetBrains MPS~\cite{ref:MPS,ref:MPSBOOK} and Spoofax \todo{Pridat referenci}.
+In particular, JetBrains MPS (Meta Programming System) is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
 MPS provides the whole IDE infrastructure that enables developers to design custom languages, use them to write program code, and compile the code into executables.
 We provide more technical details about MPS in Section~\ref{sect:MPS}.
-In the rest of the paper, we will use MPS to illustrate the general concepts and issues that apply also to other language workbenches.
+In the rest of the paper, we will often use \todo{examples from} MPS to illustrate the key concepts of our approach and related issues.
 
@@
 	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together or one can extend another, while parsers seriously limit language modularity for traditional parsed languages.
-	\item The languages may support contextual or non-parseable notations, such as tables, diagrams and positional syntax.
+	\item The languages may support contextual or non-parseable notations, such as tables, diagrams and positional syntax. \todo{Zminit layout.}
 	\item Since the projection is detached from the physical representation of code (AST), authors of languages can define multiple notations and allow the developers to switch between different representations of the code on the screen.
@@
 All of this is useful especially for DSLs, which are frequently used by domain experts who may not be professional software developers.
-The mbeddr project~\cite{ref:MBEDDR}, which extends C with numerous domain-specific constructs and data types, illustrates the abilities of language workbenches in general and of MPS in particular.
+The mbeddr project~\cite{ref:MBEDDR}, which extends C with numerous domain-specific constructs and data types, illustrates the abilities of language workbenches in general and of JetBrains MPS in particular.
 \todo{Popsat dalsi benefits of projectional editors/IDEs (and of MPS in particular): Projectional editors never have to parse the concrete syntax, so they can support/use more complicated syntax/layout.
@@
 On the other hand, the usage of projectional editors introduces some new problems.
-Before a language can be used inside MPS, it has to be defined through the MPS infrastructure, so that the IDE can understand the language and work with it.
+Before a language can be used inside editors\todo{/IDEs} such as MPS, it has to be defined through the \todo{corresponding/specific} infrastructure, so that the IDE can understand the language and work with it.
 Specifically, an author of a language has to create:
 \begin{enumerate}
-	\item the abstract syntax called \emph{Structure}, which defines the types of allowed AST nodes,
-	\item the concrete syntax called \emph{Editor}, i.e. projection of the AST on the screen and its interactions with the user, and
-	\item text generation scripts, also called \emph{TextGen}, to enable creation of plain text files used as input for compilers.
+	\item the abstract syntax (structure of the language), which defines the types of allowed AST nodes,
+	\item the concrete syntax used for editing, i.e. projection of the AST on the screen and its interactions with the user, and
+	\item text generation scripts to enable creation of plain text files used as input for compilers.
 \end{enumerate}
-Again, we provide additional technical details in Section~\ref{sect:MPS}.
 
-MPS is typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
-Before a program written in such a DSL can be executed, it is transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose language (GPL), such as C or Java.
+%Again, we provide additional technical details in Section~\ref{sect:MPS}.
+
+\todo{IDE systems based on projectional editors are} typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
+Before a program written in a rich DSL can be executed, it is \todo{has to be} transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose language (GPL), such as C or Java.
 This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
@@
 
-However, very few mainstream GPLs are now fully supported by MPS, because it requires substantial effort to manually create a full definition of a GPL in MPS.
+However, very few mainstream GPLs are now fully supported by MPS \todo{Zjistit jako to maji dalsi language workbenches (Spoofax, apod)}, because it requires substantial effort to manually create a full definition of a GPL in MPS.
 Only Java and C have been implemented to date, which limits the authors of DSLs to only target these supported GPLs.
@@
 We discuss the main challenges that we encountered during the development of \textsc{Ingrid} and our solution to them.
-\todo{Zmenit: we explain our tool using examples from MPS and ANTLR} Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges are general.
 They apply to the relation between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
@@
 We provide more details in Section~\ref{sect:LANGDEF}.
+
+\todo{
+Napsat ze input ANTLR grammar has to be tweaked before INGRID can process them.
+Za tim bude "We evaluated INGRID on complex mainstream languages, such as JavaScript and C\#. Time needed for adjusting of the mainstream languages is short ... (less than one hour).
+}
 
@@
 
+\todo{Although we define the INGRID method in the context of MPS, it can be easily adapted for/towards construction of language definitions usable in (suitable for) other IDEs and language workbenches based on projectional editing.}
+\todo{Zmenit: we explain our tool using examples from MPS and ANTLR} Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges and principles behind our approach are general and they apply also to other language workbenches.
+\todo{Although we present INGRID on MPS, we believe it can be easily adapted to other language workbenches.}
+
+\todo{Zduraznit (vyzdvihnout) general research contribution (applicable to MPS and other platforms).
+	Mozna rict neco jako "INGRID efektivne resi problem XY".
+	We summarize/highlight our contribution:
+	\begin{itemize}
+		\item Metoda pro konstrukci language definitions usable/suitable for IDEs based on projectional editing, in particular for JetBrains MPS, from ANTLR grammars, ktera je pouzitelna pro (1) import of general-purpose programming languages a (2) efficient construction of DSLs.
+To our belief, the method can be easily adapted to other language workbenches (e.g., Spoofax).
+		\item Discussion of general challenges that apply to differences between (1) the definition of a programming language that uses grammars in an EBNF-like notation and (2) the structured object-oriented approach used in projectional editors.
+	\end{itemize}
+}
+
 The rest of the paper has the following structure.
-In the next section, we provide necessary information about MPS and a brief overview of projects with similar goals as \textsc{Ingrid}.
+In the next section, we provide necessary information about MPS and a brief overview of related work, including projects with similar goals as \textsc{Ingrid}.
 Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.

### 2017-06-12 chapters/background.tex
adding my todo notes

@@
 Relate the MPS-specific terms to the large body of academic work.}
+\todo{Mozna rict jak se zakladni aspekty (Structure, Editor, TextGen) nazyvaji v jinych IDE (Spoofax, apod).}
 
@@
 
-\todo{Mozna pouzit termin "language concept" misto samotneho "concept".}
+\todo{Mozna pouzit termin "language concept" misto samotneho "concept" - ale jen pri definici.}
 
@@
 The building blocks of MPS models are called nodes.
-Code of any program in MPS is built from nodes, which represent instances of concepts defined in one of the languages that the program declares to be using.
-\todo{zvyraznit slovo "concept"} In the context of MPS, a concept is, in fact, a building block of a language definition.
+Code of any program in MPS is built from nodes, which represent instances of concepts from the languages that the program declares to be using.
+\todo{zvyraznit slovo "concept"} In the context of MPS, a concept is, in fact, \todo{a language element or} a building block of a language definition.
 We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.
@@
 The definition of an MPS concept (language element) consists of several aspects, where each aspect codifies a different part of the AST nodes' behavior.
-The essential aspects are the following: \emph{Structure} (abstract syntax), \emph{Editor} (concrete syntax) and \emph{TextGen}.
+The essential aspects are the following: \emph{Structure} (abstract syntax), \emph{Editor} and \emph{TextGen}.
 Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
@@
 It is typically needed only for the bottom-line base languages.
-DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since these are rarely converted to text directly.
+DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since \todo{programs in these languages} are rarely converted to text directly.
 After TextGen has generated textual sources from an AST, a compiler for the particular GPL is invoked to compile the textual sources into binary code.
@@
 
-\subsection{Related Projects}
+\subsection{Related Work}
 \label{sect:RELATED}
+
+\todo{We are aware of only few attempts to create ports of general-purpose languages into (IDEs based on) projectional editors/editing. All of them were conducted manually.}
 
@@
 We describe their main features and limitations in this section.
+\todo{Check what has been done for Spoofax.}
+
+\todo{Pro vsechny nasledujici projekty (PE4MPS, ANTLR{\_}MPS, mps{\_}metabnf) mention GPPLs for which there projects have been used (e.g., JavaScript).
+We discuss application of PE4MPS on JavaScript in Section 4.}
 
@@
 PE grammars extend the syntax of the ANTLR v4 notation by constructs that provide information about the possible layout of AST nodes.
-The current version, as of November 2016, supports just horizontal lists, vertical lists, and some indentation rules.
+\todo{Check if there is a new (more recent) version with more features.} The current version, as of November 2016, supports just horizontal lists, vertical lists, and some indentation rules.
 However, even these few features make the syntax much more complicated.
@@
 
+\todo{Najit dalsi podobne systemy (treba Spoofax) a kratce popsat existujici reseni podobnych uloh (automated import of general-purpose languages) v jejich kontextu.}
+
 \todo{
@@
 		Layout handled by algorithm that works with syntax trees (ASTs) and token streams.
-	The main difference from our approach (INGRID): layout inferred automatically (via automated learning) from several input files.
-		No layout annotations in the grammar are used.
-		Currently, simple heuristics are used to create editor and textgen.
+	The main difference from our approach (INGRID): No layout annotations in the grammar are used. Currently, simple heuristics are used to create editor and textgen (i.e., to determine layout).
 }

### 2017-06-12 chapters/introduction.tex
adding my todo notes

@@
 Za tim bude "We evaluated INGRID on complex mainstream languages, such as JavaScript and C\#. Time needed for adjusting of the mainstream languages is short ... (less than one hour).
+Rict nejak zkracene nasledujici text/vetu: Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the human users must also define the remaining aspects not yet supported by \textsc{Ingrid}.
 }
@@
 In the next section, we provide necessary information about MPS and a brief overview of related work, including projects with similar goals as \textsc{Ingrid}.
+\todo{Specifically, in Section~\ref{sect:MPS} we provide all information necessary to understand the examples from MPS presented in the paper.}
 Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.

### 2017-06-12 chapters/language_definition.tex
new format
adding my todo notes

@@
 
-Our approach differs from the related projects (Section~\ref{sect:RELATED}) especially in the level of automation. It also has much better support for the definition of Editor and TextGen.
+Our approach differs from the \todo{existing} related projects (Section~\ref{sect:RELATED}) especially in the level of automation.
+It also has much better support for the definition of Editor and TextGen.
 
@@
 		A terminal symbol can be encoded as a string value or using a regular expression.
-	\item \antlrliteralnoap{\textbf{String literals}}, always written inside of a pair of single quote marks, also represent terminal symbols, but by exact match to a string constant.
-	\item \antlrregex{\textbf{Regular expressions}} describe string tokens to be matched against regular expressions with a special ANTLR v4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
+	\item \antlrliteralnoap{\textbf{String literals}} \todo{tohle urcite ztmavit}, always written inside of a pair of single quote marks, also represent terminal symbols, but by exact match to a string constant.
+	\item \antlrregex{\textbf{Regular expressions}} \todo{tohle mozna ztmavit} describe string tokens to be matched against regular expressions with a special ANTLR v4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
 \end{itemize}
@@
 \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2}.\textbf{size} > 0) \{
-  \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n;}}
+  \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n;}}
   \mpstgaction{indent buffer;}
@@
   \}
-  \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n;}}
+  \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n;}}
 \} 

### 2017-06-13 chapters/conclusion.tex
notes based on reading papers

@@
 In the future, we would like to address the remaining open problems (there are many of them) \todo{Provide more concrete details here} and increase the level of automation.
-Our primary research goal is to investigate the usage of machine learning for the purpose of generating editors that support nice and readable code layout \todo{User study would be valuable here, because such a thing may not exist among multiple users}.
+Our primary research goal is to investigate the usage of automated (machine) learning for the purpose of generating editors that support nice and readable code layout \todo{User study would be valuable here, because such a thing may not exist among multiple users}.
 The information about code layout, once it is available, will be then leveraged and used also to improve TextGen.

### 2017-06-13 chapters/evaluation.tex
notes based on reading papers

@@
 The main advantage of \textsc{Ingrid} over ECMAScript4MPS is that, despite its current limitations, \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful \todo{Describe what is the adjustment}.
-\textsc{Ingrid} achieved a very good result \todo{Compared to what} especially in the case of language structure, concept aliases, and support for auto-completion.
+\textsc{Ingrid} achieved a very good result \todo{Compared to what: PE4MPS} especially in the case of language structure, concept aliases, and support for auto-completion.
 The Structure aspect generated by \textsc{Ingrid} is very similar to that of the ECMAScript4MPS project, which was created manually over a large number of hours.
@@
 Other decisions are justified by other means.} All of this justifies some of our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
-Despite that, we also tried to design some automated heuristics during our work on the \textsc{Ingrid} project, but so far all yield rather average results \todo{Average among what} when compared to what human users can achieve efficiently instead.
+Despite that, we also tried to design some automated heuristics during our work on the \textsc{Ingrid} project, but so far all yield rather average results \todo{Average among what: mozna jsem chtel rict "mediocre"} when compared to what human users can achieve efficiently instead.

### 2017-06-13 chapters/introduction.tex
notes based on reading papers

@@
 The mbeddr project~\cite{ref:MBEDDR}, which extends C with numerous domain-specific constructs and data types, illustrates the abilities of language workbenches in general and of JetBrains MPS in particular.
+\todo{Benefits of projectional editing: language composition (using multiple languages in a single program), diverse notations and layouts can be used within one program.}
 \todo{Popsat dalsi benefits of projectional editors/IDEs (and of MPS in particular): Projectional editors never have to parse the concrete syntax, so they can support/use more complicated syntax/layout.
@@
 		\item Discussion of general challenges that apply to differences between (1) the definition of a programming language that uses grammars in an EBNF-like notation and (2) the structured object-oriented approach used in projectional editors.
+		\item We discuss how practical usability of generated MPS languages depends on the specific definition of ANTLR grammar and its manual adjustments that improve usability (details in Section 3.6).
 	\end{itemize}

### 2017-06-13 chapters/language_definition.tex
notes based on reading papers

@@
 Now we can illustrate the layer problem on the behavior of auto-completion in MPS.
-Suppose that a user is creating a new document in the SimpleXML language, just inserted a fresh node of the type \mpsconcept{Element{\_1}} (see above), and would like to insert another nested XML element inside.
+Suppose that a user is creating a new document in the SimpleXML language, just inserted a fresh node of the type \todo{ztmavit} \mpsconcept{Element{\_1}} (see above), and would like to insert another nested XML element inside.
 The auto-completion mechanism of MPS offers four options that are displayed in the left part of Figure~\ref{fig:LAYERPROBLEM}.
@@
 \begin{enumerate}
-	\item Insert an object (node) of the type \mpsconcept{Content{\_}2} inside the \mpsconcept{Element{\_}1} node.
+	\item Insert an object (node) of the type \todo{ztmavit} \mpsconcept{Content{\_}2} inside the \mpsconcept{Element{\_}1} node.
 The \mpsconcept{Content{\_}2} object has only a single child node of the interface type \mpsinterface{IElement}.
@@
 Such nodes cannot expand further based on any rule in the grammar, and for that reason we call them \emph{end nodes}.
-\todo{Srovnat pojem "end nodes" vzhledem ke terminal symbols from grammars (jako analogie).
-Mozna existuje dalsi vhodny termin.}
+\todo{Srovnat pojem "end nodes" vzhledem ke terminal symbols from grammars (jako analogie). Nebo pouzit jiny termin (mozna existuje dalsi vhodny).}
 
@@
 
-We experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results \todo{Give examples or quantify this result}.
+We experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results \todo{Give examples or quantify this result --- spis examples}.
 However, we observed that the most tedious and error-prone step in the manual definition of the Editor aspect is the creation of cells for all literals (keywords), properties, children, and other fields of a given concept that should appear in its visual representation.
@@
 The resulting basic layout is illustrated on the example of the \mpsconcept{Element{\_}1} concept in the upper left corner of Figure~\ref{fig:EDITORADJUST}.
-Further adjustments of the layout, such as indentation and line breaks, can be done very efficiently in the MPS IDE.
+Further adjustments of the layout, such as indentation and line breaks, can be done very efficiently \todo{by the user} in the MPS IDE.
 The bottom right part of Figure~\ref{fig:EDITORADJUST} shows a fully customized layout for the \mpsconcept{Element{\_}1} concept.

### 2017-06-13 chapters/background.tex
making changes

@@
 Relate the MPS-specific terms to the large body of academic work.}
-\todo{Mozna rict jak se zakladni aspekty (Structure, Editor, TextGen) nazyvaji v jinych IDE (Spoofax, apod).}
+\todo{Mozna rict jak se zakladni aspekty (Structure, Editor, TextGen) nazyvaji v jinych IDE (MetaCase MWB, apod).}
 
@@
 JetBrains MPS is a language workbench --- an IDE that allows developers to create their own languages and use them to write code.
-The code can then be transformed into a target language, typically a GPL, such as Java or C, and eventually compiled into executable programs.
+The code can then be transformed into a target language, typically a GPPL, such as Java or C, and eventually compiled into executable programs.
 
@@
 DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since \todo{programs in these languages} are rarely converted to text directly.
-After TextGen has generated textual sources from an AST, a compiler for the particular GPL is invoked to compile the textual sources into binary code.
+After TextGen has generated textual sources from an AST, a compiler for the particular GPPL is invoked to compile the textual sources into binary code.
 The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes at the right moments.
@@
 
-Besides that, we are aware of three other projects that provide certain support for GPLs in the context of MPS --- namely PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
+Besides that, we are aware of three other projects that provide certain support for GPPLs within projectional editors, however all of them target MPS.
+The projects are named PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
 We describe their main features and limitations in this section.
-\todo{Check what has been done for Spoofax.}
 
@@
 
-\todo{Najit dalsi podobne systemy (treba Spoofax) a kratce popsat existujici reseni podobnych uloh (automated import of general-purpose languages) v jejich kontextu.}
+\todo{Najit dalsi podobne systemy a kratce popsat existujici reseni podobnych uloh (automated import of general-purpose languages) v jejich kontextu.}
 
@@
 \todo{
-Mention other state-of-the-art language workbenches: Spoofax.
-	Reference: L.C.L. Kats and E. Visser. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. OOPSLA 2010
-	Eclipse-based IDE support for SDF-based languages.
-	Parser-based language workbench.
-		Syntax essentially restricted to linear text.
-		But it supports language extension and composition.
+\paragraph{Other IDEs.}
+There also exist several language workbenches that are based on parser generators (such as ANTLR) and allow free editing of textual source code (unlike the projectional editor in MPS).
+A prominent example/tooo in this category is Spoofax [ref].
+	Reference: L.C.L. Kats and E. Visser. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. OOPSLA 2010, ACM.
+	Syntax of DSLs is essentially restricted to linear text.
+INGRID is not applicable to such tool/workbenches (for creating languages inside them).
 }

### 2017-06-13 chapters/introduction.tex
making changes

--- 2017-06-10 chapters/introduction.tex
@@
 
-An alternative approach is projectional (or structural) editing~\cite{ref:VWK15} \todo{Pridat citaci: nedavno precteny clanek o srovnani projectional editors}, where a developer directly manipulates the AST representation of the source code instead of plain text.
+An alternative approach is projectional (or structural) editing~\cite{ref:VWK15,ref:VSB14}, where a developer directly manipulates the AST representation of the source code instead of plain text.
 This idea emerged as early as in 1970s, but it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
 A recent revival of projectional editing has been observed in the area of language workbenches --- IDE-like tools that enable the developers to manipulate the actual language definition.
-Popular IDEs (examples) include JetBrains MPS~\cite{ref:MPS,ref:MPSBOOK} and Spoofax \todo{Pridat referenci}.
+Popular IDEs (examples) include JetBrains MPS~\cite{ref:MPS,ref:MPSBOOK} and Spoofax~\cite{ref:KV10}.
 In particular, JetBrains MPS (Meta Programming System) is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
 MPS provides the whole IDE infrastructure that enables developers to design custom languages, use them to write program code, and compile the code into executables.
-We provide more technical details about MPS in Section~\ref{sect:MPS}.
-In the rest of the paper, we will often use \todo{examples from} MPS to illustrate the key concepts of our approach and related issues.
+%We provide more technical details about MPS in Section~\ref{sect:MPS}.
+In the rest of the paper, we will often use examples from MPS to illustrate the key concepts of our approach and related issues.
 
@@
 		%An AST created by the editor can represent only a syntactically valid code.
-	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together or one can extend another, while parsers seriously limit language modularity for traditional parsed languages.
-	\item The languages may support contextual or non-parseable notations, such as tables, diagrams and positional syntax. \todo{Zminit layout.}
+	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together in a single program or one can extend another, while parsers seriously limit language modularity for traditional parsed languages.
+	\item The languages may support diverse contextual or non-parseable notations with complex layout, such as tables, mathematical expressions, diagrams and positional syntax.
 	\item Since the projection is detached from the physical representation of code (AST), authors of languages can define multiple notations and allow the developers to switch between different representations of the code on the screen.
@@
 The mbeddr project~\cite{ref:MBEDDR}, which extends C with numerous domain-specific constructs and data types, illustrates the abilities of language workbenches in general and of JetBrains MPS in particular.
-\todo{Popsat dalsi benefits of projectional editors/IDEs (and of MPS in particular): Projectional editors never have to parse the concrete syntax, so they can support/use more complicated syntax/layout.
-Support for non-linear notations (mathematical expressions, graphical or tabular layout), notations that use two-dimensional layout.
-Projectional editors mix different notations/styles.}
-\todo{Nekam pridat text: Benefits and limitations of projectional editors are thoroughly discussed in [M.
-Voelter, J.
-Siegmund, T.
-Berger, and B.
-Kolb.
-Towards User-Friendly Projectional Editors.
-SLE 2014, LNCS 8706].}
+Voelter et al.~\cite{ref:VSB14} discuss the benefits and limitations of projectional editors.
 
-On the other hand, the usage of projectional editors introduces some new problems.
+Nevertheless, the usage of projectional editors introduces some new problems.
 Before a language can be used inside editors\todo{/IDEs} such as MPS, it has to be defined through the \todo{corresponding/specific} infrastructure, so that the IDE can understand the language and work with it.
@@
 \todo{IDE systems based on projectional editors are} typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
-Before a program written in a rich DSL can be executed, it is \todo{has to be} transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose language (GPL), such as C or Java.
+Before a program written in a rich DSL can be executed, it is \todo{has to be} transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose language (GPPL), such as C or Java.
 This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
-Therefore, the target GPL must also be defined by the MPS means in order to allow the DSLs to have their models transformed into a model in the GPL.
+Therefore, the target GPPL must also be defined by the MPS means in order to allow the DSLs to have their models transformed into a model in the GPPL.
 
-However, very few mainstream GPLs are now fully supported by MPS \todo{Zjistit jako to maji dalsi language workbenches (Spoofax, apod)}, because it requires substantial effort to manually create a full definition of a GPL in MPS.
-Only Java and C have been implemented to date, which limits the authors of DSLs to only target these supported GPLs.
+However, very few mainstream GPPLs are now fully supported by MPS \todo{(and other language workbenches based on the same principles, such as MetaCase MetaEdit+ WorkBench (MWB) \url{http://www.metacase.com/mwb/})}, because it requires substantial effort to manually create a full definition of a GPPL in MPS.
+Only Java and C have been implemented to date, which limits the authors of DSLs to only target these supported GPPLs.
 The overall goal of our project is to automatize the process of language definition in MPS as much as possible, exploiting the fact that a big part of this process is quite straightforward.
-We believe that such an automated process would encourage and speed-up the migration of more GPLs into MPS, thus giving the authors of DSLs more options.
-Other possible applications include the support for writing programs in GPLs using the projectional editor and combination of GPLs with domain-specific languages natively created in MPS.
+We believe that such an automated process would encourage and speed-up the migration of more GPPLs into MPS, thus giving the authors of DSLs more options.
+Other possible applications include the support for writing programs in GPPLs using the projectional editor and combination of GPPLs with domain-specific languages natively created in MPS.
 
@@
 
-\todo{Although we define the INGRID method in the context of MPS, it can be easily adapted for/towards construction of language definitions usable in (suitable for) other IDEs and language workbenches based on projectional editing.}
+\todo{Although we define the INGRID method in the context of MPS, it can be easily adapted for/towards construction of language definitions usable in (suitable for) other IDEs and language workbenches (i) based on projectional editing (which let the user edit the AST representation (abstract syntax) directly) and (ii) where language definition is based on structural object-oriented approach.}
 \todo{Zmenit: we explain our tool using examples from MPS and ANTLR} Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges and principles behind our approach are general and they apply also to other language workbenches.
@@
 		\item Metoda pro konstrukci language definitions usable/suitable for IDEs based on projectional editing, in particular for JetBrains MPS, from ANTLR grammars, ktera je pouzitelna pro (1) import of general-purpose programming languages a (2) efficient construction of DSLs.
-To our belief, the method can be easily adapted to other language workbenches (e.g., Spoofax).
+To our belief, the method can be easily adapted to other language workbenches based on similar principles as MPS.
 		\item Discussion of general challenges that apply to differences between (1) the definition of a programming language that uses grammars in an EBNF-like notation and (2) the structured object-oriented approach used in projectional editors.
+		\item We discuss how practical usability of generated MPS languages depends on the specific definition of ANTLR grammar and its manual adjustments that improve usability (details in Section 3.6).
 	\end{itemize}
@@
 The rest of the paper has the following structure.
-In the next section, we provide necessary information about MPS and a brief overview of related work, including projects with similar goals as \textsc{Ingrid}.
+In the next section, we provide (i) all information about MPS that is necessary to understand the examples from MPS presented in the paper and (ii) a brief overview of related work, including projects with similar goals as \textsc{Ingrid}.
 Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.

### 2017-06-13 chapters/introduction.tex
editing

@@
 Nevertheless, the usage of projectional editors introduces some new problems.
-Before a language can be used inside editors\todo{/IDEs} such as MPS, it has to be defined through the \todo{corresponding/specific} infrastructure, so that the IDE can understand the language and work with it.
-Specifically, an author of a language has to create:
+Before a language can be used inside an editor such as MPS, it has to be defined through the specific infrastructure, so that the IDE can understand the language and work with it.
+More precisely, an author of a language has to create:
 \begin{enumerate}
@@
 
-\todo{IDE systems based on projectional editors are} typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
-Before a program written in a rich DSL can be executed, it is \todo{has to be} transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose language (GPPL), such as C or Java.
+IDE tools based on projectional editors are typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
+In the specific case of MPS, before a program written in a rich DSL can be executed, it is transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose programming language (GPPL), such as C or Java.
 This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
@@
 
-However, very few mainstream GPPLs are now fully supported by MPS \todo{(and other language workbenches based on the same principles, such as MetaCase MetaEdit+ WorkBench (MWB) \url{http://www.metacase.com/mwb/})}, because it requires substantial effort to manually create a full definition of a GPPL in MPS.
+However, very few mainstream GPPLs are now fully supported by MPS, because it requires substantial effort to manually create a full definition of a GPPL in MPS.
 Only Java and C have been implemented to date, which limits the authors of DSLs to only target these supported GPPLs.
@@
 In this paper we present \textsc{Ingrid}, a method and a tool for semi-automated construction of the definition of a programming language in MPS that uses an ANTLR grammar of the language as a description of its syntax.
-We discuss the main challenges that we encountered during the development of \textsc{Ingrid} and our solution to them.
-They apply to the relation between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
+The process of construction is semi-automated for two reasons: (1) input ANTLR grammars have to be adjusted before \textsc{Ingrid} can process them, and (2) some aspects of the generated MPS languages typically must be tweaked or created manually in order to improve usability of the language.
+
+We also discuss the main challenges that we encountered during the development of \textsc{Ingrid} and our solution to them.
+They are related to the principal differences between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
 Specifically, the description of a language in the form of a grammar does not hold any information about the code layout, and the grammar typically contains many rules that do not directly correspond to AST nodes and programming language constructs.
-We provide more details in Section~\ref{sect:LANGDEF}.
-
-\todo{
-Napsat ze input ANTLR grammar has to be tweaked before INGRID can process them.
-Za tim bude "We evaluated INGRID on complex mainstream languages, such as JavaScript and C\#. Time needed for adjusting of the mainstream languages is short ... (less than one hour).
-Rict nejak zkracene nasledujici text/vetu: Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the human users must also define the remaining aspects not yet supported by \textsc{Ingrid}.
-}
 
@@
 
-\todo{Although we define the INGRID method in the context of MPS, it can be easily adapted for/towards construction of language definitions usable in (suitable for) other IDEs and language workbenches (i) based on projectional editing (which let the user edit the AST representation (abstract syntax) directly) and (ii) where language definition is based on structural object-oriented approach.}
-\todo{Zmenit: we explain our tool using examples from MPS and ANTLR} Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges and principles behind our approach are general and they apply also to other language workbenches.
-\todo{Although we present INGRID on MPS, we believe it can be easily adapted to other language workbenches.}
+Although we define the \textsc{Ingrid} method in the context of MPS, and illustrate everything using examples from MPS and ANTLR grammars, it can be easily adapted to construction of language definitions in other language workbenches (i) that use projectional editing and (ii) where the language definition is based on a structural object-oriented approach.
+Most of the challenges that we discuss and key principles behind our method apply also to such other language workbenches.
 
-\todo{Zduraznit (vyzdvihnout) general research contribution (applicable to MPS and other platforms).
-	Mozna rict neco jako "INGRID efektivne resi problem XY".
-	We summarize/highlight our contribution:
+We make the following key contributions:
 \begin{itemize}
-		\item Metoda pro konstrukci language definitions usable/suitable for IDEs based on projectional editing, in particular for JetBrains MPS, from ANTLR grammars, ktera je pouzitelna pro (1) import of general-purpose programming languages a (2) efficient construction of DSLs.
-To our belief, the method can be easily adapted to other language workbenches based on similar principles as MPS.
-		\item Discussion of general challenges that apply to differences between (1) the definition of a programming language that uses grammars in an EBNF-like notation and (2) the structured object-oriented approach used in projectional editors.
-		\item We discuss how practical usability of generated MPS languages depends on the specific definition of ANTLR grammar and its manual adjustments that improve usability (details in Section 3.6).
+	\item The \textsc{Ingrid} method for construction of MPS language definitions from ANTLR grammars, which can be used (1) for the import of general-purpose programming languages into MPS and (2) for efficient construction of DSLs.
+Besides MPS, the method can be adapted to other language workbenches based on similar principles as MPS (especially projectional editing).
+	\item Identification and discussion of general challenges related to the differences between (1) definitions of programming languages that use grammars in an EBNF-like notation and (2) the structured object-oriented approach used in projectional editors such as MPS.
+	\item Discussion about the dependence of practical usability of generated MPS languages on (1) the specific form of ANTLR grammars and (2) the manual adjustments of input grammars and specific aspects of MPS languages (e.g., with respect to code layout).
+Details are provided in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:REMARKSGRAMMARS}.
 \end{itemize}
-}
 
@@
 Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
-We also discuss our experience with usage of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.
+We also discuss our experience with application of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.

### 2017-06-14 chapters/background.tex
editing

--- 2017-06-10 chapters/background.tex
@@
 
-In this section, first we introduce basic features of the MPS platform, and then we discuss related work, including other existing projects that aim to add support for general purpose languages into MPS \todo{IDE based on projectional editor}.
+In this section, first we introduce basic features of the MPS platform, and then we discuss related work, including other existing projects that aim to add support for general purpose languages into MPS.
 
@@
 
-\todo{For each MPS term (Structure, Editor, TextGen, concept, and so on) say which generally/widely known terms (from related academic work) are similar/analogous.
-Relate the MPS-specific terms to the large body of academic work.}
-
-\todo{Napsat sem nebo pripadne do Introduction (ale radsi sem): MPS platform, and tools built upon it (e.g., mbeddr) addresses to a large degree known limitations of projectional editors, such as usability of editors for languages based on textual notation (e.g., non-linear typing) and support for version control [citovat "M.
-Voelter, J.
-Siegmund, T.
-Berger, and B.
-Kolb.
-Towards User-Friendly Projectional Editors.
-SLE 2014, LNCS 8706"].}
-
-\todo{Mozna pouzit termin "language concept" misto samotneho "concept".}
-
 JetBrains MPS is a language workbench --- an IDE that allows developers to create their own languages and use them to write code.
-The code can then be transformed into a target language, typically a GPL, such as Java or C, and eventually compiled into executable programs.
+The code can then be transformed into a target language, typically a GPPL, such as Java or C, and eventually compiled into executable programs.
 
@@
 The building blocks of MPS models are called nodes.
-Code of any program in MPS is built from nodes, which represent instances of concepts defined in one of the languages that the program declares to be using.
-\todo{zvyraznit slovo "concept"} In the context of MPS, a concept is, in fact, a building block of a language definition.
+Code of any program in MPS is built from nodes, which represent instances of concepts from the languages that the program declares to be using.
+In the context of MPS, a \emph{concept} is a language element, i.e. a building block of a language definition.
 We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.
@@
 The definition of an MPS concept (language element) consists of several aspects, where each aspect codifies a different part of the AST nodes' behavior.
-The essential aspects are the following: \emph{Structure} (abstract syntax), \emph{Editor} (concrete syntax) and \emph{TextGen}.
-Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
+The essential aspects are the following: \emph{Structure}, \emph{Editor} and \emph{TextGen}.
+Structure represents the abstract syntax (types and hierarchy of AST nodes), Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
 If, instead of generating text directly, programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
@@
 It is typically needed only for the bottom-line base languages.
-DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since these are rarely converted to text directly.
-After TextGen has generated textual sources from an AST, a compiler for the particular GPL is invoked to compile the textual sources into binary code.
+DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since programs in such languages are rarely converted to text directly.
+After TextGen has generated textual sources from an AST, a compiler for the particular GPPL is invoked to compile the textual sources into binary code.
 The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes at the right moments.
@@
 
-\subsection{Related Projects}
+\subsection{Related Work}
 \label{sect:RELATED}
 
-The BaseLanguage~\cite{ref:BASELANG} is an almost full port of Java, which was created manually by JetBrains and extended with MPS-specific features.
+We are aware just of a few attempts to create ports of general-purpose languages into IDE tools based on projectional editing.
+All the ports that we mention below were conducted manually.
+
+The BaseLanguage~\cite{ref:BASELANG} is an almost full port of Java, which was created by JetBrains and extended with MPS-specific features.
 % and it is still undergoing changes as Java itself is evolving.
@@
 
-Besides that, we are aware of three other projects that provide certain support for GPLs in the context of MPS --- namely PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
+The following three projects --- PE4MPS, ANTLR{\_}MPS, and mps-metabnf --- aim to provide certain support for GPPLs within the MPS platform.
 We describe their main features and limitations in this section.
@@
 
+We discuss application of PE4MPS on JavaScript in Section~\ref{sect:EVAL}, and compare its results to our \textsc{Ingrid} approach.
+
 \paragraph{ANTLR{\_}MPS.}
@@
 
-\todo{
-S. Erdweg, T. Rendel, C. K\"{a}stner, and K. Ostermann. Layout-Sensitive Generalized Parsing. In Proceedings of SLE 2012, LNCS 7745.
-	Source code layout specified declaratively through constraints on shape and relative positions of syntax trees (AST).
-	Layout constraints are represented/encoded using annotations on production rules of context-free grammars.
-	Useful for layout-sensitive programming language (e.g., Python).
-	Their proposed approach: grammar rule applies only if the input satisfied the constraint (on shape and positioning of source code fragments).
-		Parsing algorithm that accepts only inputs with valid layout (that satisfies the constraints attached to grammar rules).
-	Constraints support (can express): indentation, number of line breaks, alignment.
-	Hand-coded approach used in Python (how the annotated grammar is processed) --- mentioned for comparison.
-		Special tokens that indicate layout aspects (indentation, alignment) are added to stream of tokens by preprocessor.
-		Layout handled by algorithm that works with syntax trees (ASTs) and token streams.
-	The main difference from our approach (INGRID): layout inferred automatically (via automated learning) from several input files.
-		No layout annotations in the grammar are used.
-		Currently, simple heuristics are used to create editor and textgen.
-}
-
-\todo{
-Mention other state-of-the-art language workbenches: Spoofax.
-	Reference: L.C.L. Kats and E. Visser. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. OOPSLA 2010
-	Eclipse-based IDE support for SDF-based languages.
-	Parser-based language workbench.
-		Syntax essentially restricted to linear text.
-		But it supports language extension and composition.
-}
+\paragraph{Other IDEs.}
+Note that while the \textsc{Ingrid} method has been designed primarily for MPS, and can be easily adapted to other IDE tools based on projectional editing, there also exist several language workbenches that use parser generators (such as ANTLR) and support free editing of textual source code.
+A prominent member of the latter category of tools is Spoofax~\cite{ref:KV10}, which essentially restricts the syntax of DSLs to linear text.
+\textsc{Ingrid} is not applicable for creating languages inside such tools.

### 2017-06-14 chapters/language_definition.tex
editing

--- 2017-06-05 chapters/language_definition.tex
@@
 The proposed \textsc{Ingrid} method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
-\todo{Zmenit nasledujici vetu: There exist ANTLR v4 grammars for all widely used programming languages ...}
-All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
+Syntax definition in the form of an ANTLR v4 grammar exist for all widely used programming languages~\footnote{https://github.com/antlr/grammars-v4}.
 Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.
@@
 
-Our approach differs from the related projects (Section~\ref{sect:RELATED}) especially in the level of automation. It also has much better support for the definition of Editor and TextGen.
+Our approach differs from the existing projects with similar goals (Section~\ref{sect:RELATED}) especially in the level of automation.
+It also has much better support for the definition of Editor and TextGen.
 
@@
 A sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.
-\todo{Algorithm seems to fail if the grammar has right-recursion (according to one reviewer) --- check that and describe how we handle that case.}
 
@@
 However, presence of the intermediate layers significantly complicates usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
-We show examples illustrating this problem, which we call a \emph{layer problem}, later in the subsection where we describe our solution \todo{Pridat odkaz na konkretni section a opravit gramatickou chybu}.
+We show examples illustrating this problem, which we call a \emph{layer problem}, later in this section at a point where we describe our solution.
 
@@
 
-The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree and consider just the leaf nodes, e.g., to be directly offered to the user through the auto-completion menu.
-Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, which is given in Figure~\ref{fig:CONTENTHIERARCHY} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the leaf nodes that are highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.
+The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree, and consider just the nodes to be directly offered to the user through the auto-completion menu.
+Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, which is given in Figure~\ref{fig:CONTENTHIERARCHY} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the nodes that are highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.
 
@@
 
-For the input that consists of a particular AST node $N$ and the grammar rule $R$ that expands $N$, the procedure implementing the shortcut approach systematically traverses the parser tree (AST) built in the earlier phases in order to identify each AST node that may appear at the end of some derivation chain starting by the rule $R$ from the node $N$.
-Such nodes cannot expand further based on any rule in the grammar, and for that reason we call them \emph{end nodes}.
-\todo{Srovnat pojem "end nodes" vzhledem ke terminal symbols from grammars (jako analogie).
-Mozna existuje dalsi vhodny termin.}
+For the input that consists of a particular AST node $N$ and the grammar rule $R$ that expands $N$, the procedure implementing the shortcut approach systematically traverses the parser tree (AST) built in the earlier phases in order to identify each AST node that (i) may appear in some derivation chain starting by the rule $R$ from the node $N$ and (ii) does not belong to any intermediary layer.
+We call them \emph{end nodes}.
+Here we must emphasize that end nodes do not correspond to terminal symbols from grammars, because they may represent AST nodes with children (i.e., syntax elements that can be further expanded using grammar rules).
 
@@
 
-We experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results \todo{Give examples or quantify this result}.
-However, we observed that the most tedious and error-prone step in the manual definition of the Editor aspect is the creation of cells for all literals (keywords), properties, children, and other fields of a given concept that should appear in its visual representation.
+We observed that the most tedious and error-prone step in the manual definition of the Editor aspect is the creation of cells for all literals (keywords), properties, children, and other fields of a given concept that should appear in its visual representation.
 This step can be very easily automated.
@@
 The resulting basic layout is illustrated on the example of the \mpsconcept{Element{\_}1} concept in the upper left corner of Figure~\ref{fig:EDITORADJUST}.
-Further adjustments of the layout, such as indentation and line breaks, can be done very efficiently in the MPS IDE.
+Further adjustments of the layout, such as indentation and line breaks, can be done very efficiently by the user in the MPS IDE.
 The bottom right part of Figure~\ref{fig:EDITORADJUST} shows a fully customized layout for the \mpsconcept{Element{\_}1} concept.
@@
 Based on our experience, it takes a very short time to manually adjust the layout into a form much better than any fully automated heuristic could achieve.
-We expect that a typical user of the \textsc{Ingrid} method will be able to use the projectional editor in MPS quite efficiently.
+A typical user of the \textsc{Ingrid} method will be able to use the projectional editor in MPS quite efficiently.
+We actually experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results --- it is quite hard to automatically identify language concepts for which the default layout should be refactored, and the user would have to adjust the layout by hand anyway.
 More details are provided in Section~\ref{sect:EVAL}, including our experience with several mainstream programming languages.
@@
 \subsection{Remarks about Grammars}
+\label{sect:REMARKSGRAMMARS}
 

### 2017-06-15 chapters/conclusion.tex
reducing the length

--- 2017-06-05 chapters/conclusion.tex
@@
 
-In the future, we would like to address the remaining open problems (there are many of them) and increase the level of automation.
-Our primary research goal is to investigate the usage of machine learning for the purpose of generating editors that support nice and readable code layout.
+In the future, we would like to add support for generating other aspects of MPS languages and increase the level of automation.
+Our primary research goal is to investigate the usage of automated (machine) learning for the purpose of generating editors that support nice and readable code layout.
+First, we would have to collect a set of input source code files to be used for training.
+This step might involve a user study designed in order to identify examples of layout that a majority considers as nice and readable.
 The information about code layout, once it is available, will be then leveraged and used also to improve TextGen.
-Another subject for future work is generating parsers that would allow users to import existing source code into MPS \todo{Vysvetlit podrobneji ze chceme automaticky generovat parsery textovych zdrojaku ktere budou vytvaret interni reprezentace (ASTs) postavene na zaklade MPS concepts pro dany jazyk}.
+Another subject for future work is automated construction of parser-based tools that could be used for translating source code in plain-text to instances of MPS languages.

### 2017-06-15 chapters/evaluation.tex
reducing the length

--- 2016-11-12 chapters/evaluation.tex
@@
 
-The main advantage of \textsc{Ingrid} over ECMAScript4MPS is that, despite its current limitations, \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful.
-\textsc{Ingrid} achieved a very good result especially in the case of language structure, concept aliases, and support for auto-completion.
+The main advantage of \textsc{Ingrid} over ECMAScript4MPS is that, despite its current limitations, \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful --- for example, the language designer has to define actions that improve editing experience by automatically transforming the AST in the background.
+\textsc{Ingrid} achieved a very good result, when compared to PE4MPS, especially in the case of language structure, concept aliases, and support for auto-completion.
 The Structure aspect generated by \textsc{Ingrid} is very similar to that of the ECMAScript4MPS project, which was created manually over a large number of hours.
@@
 Another very complex programming language that we used for experiments is C\#~\cite{ref:CSHARP}.
-Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces at least a reasonable structure of the MPS language.
+Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces at least a reasonable structure (hierarchy of concepts) that can be practically used in MPS just with minor alterations.
+Necessary adjustments include, for example, removal of intermediary rules such as \antlrparserrule{statement\_list} : \antlrparserrule{statement*}~\antlrliteral{;} by inlining the right-hand side.
 The Editor aspect was modified afterwards, but only for some of the concepts generated by \textsc{Ingrid}. 
 Roughly one hour of manual effort was needed in total.
-Nevertheless, additional modifications of the grammar and selected aspects of the MPS language (Editor, TextGen) are still needed to get an optimal result.
+Nevertheless, additional modifications of the grammar and selected aspects of the MPS language (Editor, TextGen) are still needed.
 A great flexibility and complexity of the C\# language is the main reason behind all the necessary changes to the language definition in MPS and to the ANTLR grammar.
-The definition is quite large, involving more than 800 concepts, and therefore building of the language takes a rather long time.
+The definition is quite large, involving more than 800 concepts.
+%, and therefore compilation (building) of the language in MPS takes a rather long time.
 
@@
 For example, the structure and hierarchy of rules in the Python grammar are quite different from JavaScript or JSON, and consequently the language definition created by \textsc{Ingrid} is rather badly organized.
-We plan to address this limitation in the future.
-However, significant modifications of the grammar would probably be necessary anyway.
+We plan to address this limitation in the future and extend \textsc{Ingrid} such that very few modifications of the grammar would be necessary even for these languages.
 
@@
 
-On the other hand, our experiments with complex languages show that, in the case of the Editor and TextGen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, much less time than we expected.
+On the other hand, our experiments with complex languages show that, in the case of the Editor and TextGen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, by an order of magnitude less time than we initiall expected.
 The first author spent between 20 and 60 minutes of work on each language, when adjusting the result of automated generation into a more useful and readable form.
-MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing users to reach optimal results very fast.
-All of this justifies some of our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
-Despite that, we also tried to design some automated heuristics during our work on the \textsc{Ingrid} project, but so far all yield rather average results when compared to what human users can achieve efficiently instead.
+MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing designers to produce really useful languages very fast.
+All of this justifies our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
+Despite that, we also tried to design some automated heuristics during our work on the \textsc{Ingrid} project, but so far all yield rather mediocre results when compared to what human users can achieve efficiently instead.

### 2017-06-15 chapters/introduction.tex
reducing the length

@@
 A recent revival of projectional editing has been observed in the area of language workbenches --- IDE-like tools that enable the developers to manipulate the actual language definition.
-Popular IDEs (examples) include JetBrains MPS~\cite{ref:MPS,ref:MPSBOOK} and Spoofax~\cite{ref:KV10}.
+Popular examples are JetBrains MPS~\cite{ref:MPS,ref:MPSBOOK} and Spoofax~\cite{ref:KV10}.
 In particular, JetBrains MPS (Meta Programming System) is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
@@
 \begin{itemize}
-	\item The developers practically cannot enter syntactically invalid code, because a projectional editor controls the interaction between the user and the code.
+	\item The developers practically cannot enter syntactically invalid code, because a projectional editor controls the interaction between the user and the program code.
 		%An AST created by the editor can represent only a syntactically valid code.
-	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together in a single program or one can extend another, while parsers seriously limit language modularity for traditional parsed languages.
-	\item The languages may support diverse contextual or non-parseable notations with complex layout, such as tables, mathematical expressions, diagrams and positional syntax.
-	\item Since the projection is detached from the physical representation of code (AST), authors of languages can define multiple notations and allow the developers to switch between different representations of the code on the screen.
+	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together in a single program or one can extend another.
+		%, while parsers seriously limit language modularity for traditional parsed languages.
+	\item The languages may support diverse contextual or non-parseable notations with complex layout, such as tables, diagrams, and mathematical expressions.
+	\item Since the projection is detached from the physical representation of code (AST), authors of languages can define multiple notations and allow the developers to switch between them on the screen.
 \end{itemize}
 All of this is useful especially for DSLs, which are frequently used by domain experts who may not be professional software developers.
-The mbeddr project~\cite{ref:MBEDDR}, which extends C with numerous domain-specific constructs and data types, illustrates the abilities of language workbenches in general and of JetBrains MPS in particular.
-Voelter et al.~\cite{ref:VSB14} discuss the benefits and limitations of projectional editors.
+The mbeddr project~\cite{ref:MBEDDR} illustrates the abilities of language workbenches in general and of JetBrains MPS in particular.
+Voelter et al.~\cite{ref:VSB14} discuss the benefits and limitations of projectional editors in more detail.
 
 Nevertheless, the usage of projectional editors introduces some new problems.
-Before a language can be used inside an editor such as MPS, it has to be defined through the specific infrastructure, so that the IDE can understand the language and work with it.
+Before a language can be used inside an editor, it has to be defined through the specific infrastructure, so that the IDE can understand the language and work with it.
 More precisely, an author of a language has to create:
@@
 	\item the abstract syntax (structure of the language), which defines the types of allowed AST nodes,
-	\item the concrete syntax used for editing, i.e. projection of the AST on the screen and its interactions with the user, and
-	\item text generation scripts to enable creation of plain text files used as input for compilers.
+	\item the concrete syntax for editing, i.e. projection of the AST on the screen and interactions with the user, and
+	\item text generation scripts to enable creation of plain text representation used as input for compilers.
 \end{enumerate}
@@
 
-IDE tools based on projectional editors are typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
-In the specific case of MPS, before a program written in a rich DSL can be executed, it is transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose programming language (GPPL), such as C or Java.
+IDE tools based on projectional editors are typically used for syntactically rich DSLs.
+%, which are likely to benefit most from overcoming the limitations of parser-based languages.
+In the specific case of MPS, before a program written in a DSL can be executed, it is transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose programming language (GPPL), such as C or Java.
 This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
@@
 
-However, very few mainstream GPPLs are now fully supported by MPS, because it requires substantial effort to manually create a full definition of a GPPL in MPS.
-Only Java and C have been implemented to date, which limits the authors of DSLs to only target these supported GPPLs.
-The overall goal of our project is to automatize the process of language definition in MPS as much as possible, exploiting the fact that a big part of this process is quite straightforward.
-We believe that such an automated process would encourage and speed-up the migration of more GPPLs into MPS, thus giving the authors of DSLs more options.
-Other possible applications include the support for writing programs in GPPLs using the projectional editor and combination of GPPLs with domain-specific languages natively created in MPS.
+However, very few mainstream GPPLs are now supported by MPS, because it requires substantial effort to manually create a full definition of a language in MPS.
+Only Java and C have been implemented to date.
+%, which limits the authors of DSLs to only target these supported GPPLs.
+The overall goal of our project is to automatize the process of language definition in MPS as much as possible.
+%, exploiting the fact that a big part of this process is quite straightforward.
+We believe that such an automated process would encourage and speed-up the migration of more GPPLs into MPS, thus giving the authors of DSLs more options regarding target general languages.
+Another possible application is the combination of general-purpose languages with DSLs natively created in MPS.
+%include the support for writing programs in GPPLs using the projectional editor and
 
-In this paper we present \textsc{Ingrid}, a method and a tool for semi-automated construction of the definition of a programming language in MPS that uses an ANTLR grammar of the language as a description of its syntax.
-The process of construction is semi-automated for two reasons: (1) input ANTLR grammars have to be adjusted before \textsc{Ingrid} can process them, and (2) some aspects of the generated MPS languages typically must be tweaked or created manually in order to improve usability of the language.
+In this paper we present \textsc{Ingrid}, a method and a tool for semi-automated construction of a language definition in MPS that uses an ANTLR grammar as a description of its syntax.
+The process of construction is semi-automated for two reasons: (1) ANTLR grammars have to be adjusted before \textsc{Ingrid} can process them, and (2) some aspects of the generated MPS languages typically must be tweaked or created manually in order to improve languages' usability.
 
 We also discuss the main challenges that we encountered during the development of \textsc{Ingrid} and our solution to them.
-They are related to the principal differences between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
+They are related to principal differences between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
 Specifically, the description of a language in the form of a grammar does not hold any information about the code layout, and the grammar typically contains many rules that do not directly correspond to AST nodes and programming language constructs.
 
-The \textsc{Ingrid} tool can help also to make the construction of DSLs more efficient.
-Based on our experience, we believe that, especially for simple languages, it is less time consuming to write their grammar by hand and then create the language definition in MPS using \textsc{Ingrid}, than to create the language manually using the MPS GUI.
+\textsc{Ingrid} can also make the construction of DSLs more efficient.
+Based on our experience, we believe that, especially for simple languages, it is less time consuming to write their grammar by hand and then create the language definition in MPS with the help of \textsc{Ingrid}, than to create everything manually using the MPS GUI.
 
@@
 
-Although we define the \textsc{Ingrid} method in the context of MPS, and illustrate everything using examples from MPS and ANTLR grammars, it can be easily adapted to construction of language definitions in other language workbenches (i) that use projectional editing and (ii) where the language definition is based on a structural object-oriented approach.
-Most of the challenges that we discuss and key principles behind our method apply also to such other language workbenches.
+Although we define the \textsc{Ingrid} method in the context of MPS, and illustrate everything using examples from MPS and ANTLR grammars, it can be easily adapted to other language workbenches (i) that use projectional editing and (ii) where the language definition is based on a structural object-oriented approach.
+Most of the challenges that we discuss and key principles behind our method are general and therefore apply also to other language workbenches.
 
@@
 \begin{itemize}
-	\item The \textsc{Ingrid} method for construction of MPS language definitions from ANTLR grammars, which can be used (1) for the import of general-purpose programming languages into MPS and (2) for efficient construction of DSLs.
-Besides MPS, the method can be adapted to other language workbenches based on similar principles as MPS (especially projectional editing).
-	\item Identification and discussion of general challenges related to the differences between (1) definitions of programming languages that use grammars in an EBNF-like notation and (2) the structured object-oriented approach used in projectional editors such as MPS.
-	\item Discussion about the dependence of practical usability of generated MPS languages on (1) the specific form of ANTLR grammars and (2) the manual adjustments of input grammars and specific aspects of MPS languages (e.g., with respect to code layout).
+	\item The \textsc{Ingrid} method for construction of MPS language definitions from ANTLR grammars, which can be used (1) for the import of general-purpose languages into MPS and (2) for efficient construction of DSLs.
+\textsc{Ingrid} can be adapted also to other language workbenches based on similar principles as MPS.
+	\item Identification and discussion of general challenges related to the differences between (1) language definitions that use grammars and (2) the structured object-oriented approach used in projectional editors.
+	\item Discussion of the practical usability of generated MPS languages and its dependence on (1) the specific form of ANTLR grammars and (2) the manual adjustments of input grammars and specific aspects of MPS languages (e.g., with respect to code layout).
 Details are provided in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:REMARKSGRAMMARS}.
@@
 The rest of the paper has the following structure.
-In the next section, we provide (i) all information about MPS that is necessary to understand the examples from MPS presented in the paper and (ii) a brief overview of related work, including projects with similar goals as \textsc{Ingrid}.
-Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
-We also discuss our experience with application of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.
+In the next section, we provide (i) all information about MPS that is necessary to understand the examples presented in the paper and (ii) a brief overview of related work, including projects with similar goals as \textsc{Ingrid}.
+Then, in Section~\ref{sect:LANGDEF}, we describe in detail our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
+We also discuss our experience with application of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\# (Section~\ref{sect:EVAL}), and then we conclude.

### 2017-06-15 chapters/language_definition.tex
reducing the length

@@
 A sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.
+Our implementation of the algorithm detects rules defined in a recursive way (e.g., using the pattern \antlrlexerrule{N} : \antlrliteral{text} \antlrlexerrule{N}) and rejects the input grammar if it contains such rules.
+However, note that we use the flatenning algorithm only for lexer rules, where recursion is quite rare --- specifically, the algorithm is not applied to parser rules.
 
@@
 Nevertheless, since the Editor aspect has to be adjusted manually, we decided to use a different approach for TextGen.
-We designed a procedure based on a simple fully automated heuristic that provides surprisingly good results \todo{Quantify this}.
+We designed a procedure based on a simple fully automated heuristic that provides acceptable results --- the output is human-readable and close to standard formatting of source code used by popular IDE tools, although minor tweaks are often needed.
 
@@
 
-\todo{Figure 15: whitespace is just blank space instead of quoted string, mozna jsem neco smazal omylem (za "with ").}
 Figure~\ref{fig:TEXTGENFINAL} contains the complete, automatically generated, TextGen aspect for the \mpsconcept{Element{\_}1} concept, which represents the full SimpleXML element.

### 2017-06-15 chapters/background.tex
reducing length and polishing - intro and background done

@@
 
-In this section, first we introduce basic features of the MPS platform, and then we discuss related work, including other existing projects that aim to add support for general purpose languages into MPS.
+In this section, we introduce basic features of the MPS platform, and then we discuss other existing projects that aim to add support for general purpose languages into MPS.
 
@@
 JetBrains MPS is a language workbench --- an IDE that allows developers to create their own languages and use them to write code.
-The code can then be transformed into a target language, typically a GPPL, such as Java or C, and eventually compiled into executable programs.
+The code can then be transformed into a target language, typically a GPPL such as Java or C, and eventually compiled into executable programs.
 
-As we already indicated, MPS differs from typical IDEs in one important aspect --- the projectional editor.
-The developer does not work with the textual representation of the source code directly, but rather with its AST that is the model of the code.
-Basically, when using the MPS projectional editor, programs are created by assembling the tree (AST) out of predefined building blocks of selected languages.
+When using the MPS projectional editor, the developer does not work with the textual representation of the source code, but rather directly with its AST that is the model of the code.
+Programs are created by assembling the tree (AST) out of predefined building blocks of selected languages.
 The definition of a language in MPS dictates, where in the AST can certain elements be placed and how they can be nested inside other elements.
-On the other hand, in traditional IDEs it is the parser that constructs an AST of a program using the language's elements.
 
 The building blocks of MPS models are called nodes.
-Code of any program in MPS is built from nodes, which represent instances of concepts from the languages that the program declares to be using.
-In the context of MPS, a \emph{concept} is a language element, i.e. a building block of a language definition.
+Code of any program in MPS is built from nodes, which represent instances of concepts from the languages that the program is written in.
+In MPS, a \emph{concept} is a language element, i.e. a building block of a language definition.
 We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.
@@
 The editor can take any visual form and shape.
-The language author can define multiple alternative visualizations and let the developer choose one that fits best the task at hand --- for example, editing, debugging, reviewing, resolving merge conflicts, etc.
+The language author can define multiple alternative visualizations and let the developer choose one that fits best the task at hand.
+%--- for example, editing, debugging, reviewing, resolving merge conflicts, etc.
 In particular, the visual representations are not bound to be just textual at all.
-For example, blocks corresponding to branches of an \code{if-then-else} statement can be aligned next to each other and displayed with different background colors.
+%For example, blocks corresponding to branches of an \code{if-then-else} statement can be aligned next to each other and displayed with different background colors.
 
@@
 If, instead of generating text directly, programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
-Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we describe these three below in more detail, and neglect other aspects such as type system and data-flow.
+Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we describe them below in more detail, and neglect other aspects such as type constraints.
 
@@
 The fundamental aspect of any MPS language is Structure.
-It must be created first for each intended language element (concept).
-Structure specifies core attributes of an MPS concept such as the name, inheritance relationships, child concepts (including their types and cardinalities), the implemented interfaces, and references to other AST nodes.
+It must be created first for each intended language concept.
+Structure specifies core attributes of an MPS concept such as the name, inheritance relationships, child concepts (their types and cardinalities), implemented interfaces, and references to other AST nodes.
 % other properties (fields of any type that can hold values), etc.
-Figure~\ref{fig:if_statement_structure} shows definition of the Structure aspect for the \code{if-then-else} statement.
+Figure~\ref{fig:if_statement_structure} shows the Structure aspect for the \code{if-then-else} statement.
 
@@
 \begin{figure}[ht]
+\vspace{-2mm}
 \centering
@@
 \label{fig:if_statement_structure}
+\vspace{-4mm}
 \end{figure}
@@
 \paragraph{Editor.}
-The Editor aspect is where the language designer specifies what the projectional editor representation of a code fragment (an AST) looks like on the screen and how the user interacts with the code.
+The Editor aspect is where the language designer specifies what the projectional representation of a code fragment (an AST) looks like on the screen and how the user interacts with the code.
 JetBrains have developed a cellular system that enables placing properties and children of a node (concept) into different cells.
-The author usually incorporates all of the node's children, references, and properties inside the representation, so that the future users of the language can insert all values that the node expects.
+The author usually incorporates all of the node's children, references, and properties inside the representation, so that future users of the language can insert all values that the node expects.
 Additionally, cells of the editor can be styled using a language similar to CSS.
-The supported visual characteristics include text color and indentation.
+Supported visual characteristics include color and indentation.
 
 Figure~\ref{fig:if_editor_definition} shows an example of what the definition of the Editor aspect for the \code{if-then-else} statement might look like.
-While here we indicate positions of cell borders on each line by spaces (just for the purpose of illustration), MPS GUI actually uses a graphically much more appealing way of displaying the Editor aspects, which involves vertical and horizontal lines of different colors and also background colors other than white for some cells.
-The symbols \verb|[-| and \verb|-]| represent layout information (cells), which define rules for mutual positioning of the contained cells (vertical, horizontal, indentation).
+While here we indicate positions of cell borders on each line by spaces (just for illustration), MPS GUI actually uses a graphically much more appealing way of displaying the Editor aspects, which involves vertical and horizontal lines of different colors and also background colors other than white for some cells.
+The symbols \verb|[-| and \verb|-]| represent layout information, which specifies mutual positioning of the contained cells (vertical, horizontal, indentation).
 
 \begin{figure}[ht]
+\vspace{-1mm}
 \centering
@@
 \end{alltt}
+\vspace{-1mm}
 \caption{Editor aspect for the \code{if-else} statement}
 \label{fig:if_editor_definition}
+\vspace{-2mm}
 \end{figure}
@@
 \paragraph{BaseLanguage.}
-Another important feature of MPS that we need to describe in more detail is \emph{BaseLanguage}~\cite{ref:BASELANG}.
-It is a clone of Java implemented using the MPS constructs.
+Another important feature of MPS that we need to describe is \emph{BaseLanguage}~\cite{ref:BASELANG}, a clone of Java implemented using the MPS constructs.
 BaseLanguage was developed in the early days of MPS in order to implement MPS itself and also to support the basic set of language-definition DSLs,
@@
 Similarly, all the custom DSLs that are meant to be generated into Java choose BaseLanguage as their generation target.
-The conversion to textual Java is handled by BaseLanguage without any further manual effort, since BaseLanguage has a TextGen aspect defined, which translates BaseLanguage code into textual Java sources.
+The conversion to textual Java sources is handled by BaseLanguage (without any further manual effort), through its TextGen aspect.
 
@@
 \end{alltt}
+\vspace{-1mm}
 \caption{TextGen aspect definition for the \code{if-else} statement}
 \label{fig:if_statement_textgen}
+\vspace{-2mm}
 \end{figure}
@@
 
-\subsection{Related Work}
+\subsection{Related Projects with Similar Goals}
 \label{sect:RELATED}
@@
 We are aware just of a few attempts to create ports of general-purpose languages into IDE tools based on projectional editing.
-All the ports that we mention below were conducted manually.
+All of them (described below) were conducted manually.
 
@@
 
-\paragraph{PE4MPS.}
-PE4MPS~\cite{ref:PE4MPS} is a project that addresses the lack of information about code layout in grammars by using so called PE grammars~\cite{ref:PE}, which is a new grammar notation proposed by the same author.
-The abbreviation PE stands for projectional editing.
+\paragraph{PE4MPS~\cite{ref:PE4MPS}.} This is a project that addresses the lack of information about code layout in grammars by using so called PE grammars~\cite{ref:PE}, which is a new notation proposed by the same author.
+PE stands for projectional editing.
 
@@
 The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
-In particular, language developers are forced to specify the code layout manually in PE grammars, through the corresponding extensions.
+Language developers are forced to specify the code layout manually in PE grammars, through the corresponding extensions.
 Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
-Another limitation of PE4MPS is that it does not implement the full ANTLR syntax.
-Every grammar might require non-trivial adjustments before it can be processed using the PE4MPS tool.
-To compare, the goals behind our \textsc{Ingrid} approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.
+Like for \textsc{Ingrid}, every grammar might require non-trivial adjustments before it can be processed using the PE4MPS tool.
+%To compare, the goals behind our \textsc{Ingrid} approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.
 
@@
 
-\paragraph{ANTLR{\_}MPS.}
-The ANTLR{\_}MPS project~\cite{ref:ANTLR2MPS} also uses the ANTLR v4 grammar notation, but otherwise works quite differently from PE4MPS.
-We provide only a brief mention here because this project is in an early stage of development.
+\paragraph{ANTLR{\_}MPS~\cite{ref:ANTLR2MPS}.} This project also uses the ANTLR v4 grammar notation, but otherwise works quite differently from PE4MPS.
+%We provide only a brief mention here because this project is in an early stage of development.
 Its author created an ANTLR v4 MPS language, which captures the syntax of ANTLR v4 notation using MPS concepts.
@@
 
-\paragraph{mps-metabnf.}
-The mps-metabnf tool~\cite{ref:MPSMETABNF}, implemented by members of the DSLFoundry group, builds upon similar ideas as the ANTLR{\_}MPS project described above.
+\paragraph{mps-metabnf~\cite{ref:MPSMETABNF}.} This tool, implemented by members of the DSLFoundry group, builds upon similar ideas as the ANTLR{\_}MPS project described above.
 Authors of this project, too, created an MPS language to capture ANTLR grammars.
@@
 A prominent member of the latter category of tools is Spoofax~\cite{ref:KV10}, which essentially restricts the syntax of DSLs to linear text.
-\textsc{Ingrid} is not applicable for creating languages inside such tools.
+\textsc{Ingrid} is not applicable for creating languages inside such language workbenches.

### 2017-06-16 chapters/conclusion.tex
ready for submission

@@
 
-Our contribution is the \textsc{Ingrid} method for constructing language definitions in MPS from their grammars in the ANTLR notation.
+Our contribution is the \textsc{Ingrid} method for constructing language definitions in MPS from grammars in the ANTLR notation.
 The method is only partially automated, but nevertheless it greatly reduces the amount of manual work that developers of MPS languages have to perform.
-Although we discussed all the challenges, details of our solution and general observations just in the context of ANTLR grammars and JetBrains MPS, we believe they are more general --- and in particular relevant to everyone tackling a similar problem in the field of DSLs and language workbenches.
+Although we discussed all challenges, details of our solution and general observations just in the context of ANTLR grammars and JetBrains MPS, we believe they are more general --- in particular, relevant to everyone tackling a similar problem in the field of DSLs and language workbenches.
 

### 2017-06-16 chapters/evaluation.tex
ready for submission

@@
 
-We implemented the proposed \textsc{Ingrid} method as an MPS plugin that allows users to generate languages from ANTLR v4 grammars.
-While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to bind with the MPS API, which is used to programatically generate elements of the new language and their aspects.
+We implemented the \textsc{Ingrid} method as an MPS plugin.
+While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to bind with the MPS API, which is used to programatically generate language elements and their aspects.
 The plugin uses the ANTLR library~\cite{ref:ANTLR} for parsing of ANTLR v4 grammar files, and several MPS libraries that implement the MPS API.
@@
 For the purpose of evaluation, we applied \textsc{Ingrid} to several well-established and widely used languages, including JSON, JavaScript (ECMAScript 5.1~\cite{ref:ECMASCRIPT51}), and C\#.
-MPS projects that contain definitions of all three languages are also released in the public repository (\url{https://github.com/premun/ingrid}).
+MPS projects that contain definitions of all three languages are also released at \url{https://github.com/premun/ingrid}.
 In the rest of this section, we discuss our experience with application of \textsc{Ingrid} to these languages, and then we highlight few general observations.
 
-However, first we must emphasize that MPS languages automatically produced by the \textsc{Ingrid} method, as defined in this paper, are not ready-to-use full-fledged MPS counterparts of the original input languages.
-The structure of such a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects (e.g., the Editor) have to be manually tweaked (or defined from scratch) by the end user.
-\textsc{Ingrid} also does not support advanced features of MPS, such as type checking.
+However, first we must emphasize that MPS languages automatically produced by the \textsc{Ingrid} method, are not ready-to-use full-fledged MPS counterparts of the original input languages.
+The structure of a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects have to be manually tweaked (e.g., the Editor) or defined from scratch by the end user.
+\textsc{Ingrid} also does not yet support advanced features of MPS, such as type checking.
 
-For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition in the form that incorporates all tweaks performed manually by authors of this paper and (2) the adjusted ANTLR v4 grammar used as input for \textsc{Ingrid}, both in the repository that contains also our implementation.
+For each of the three languages (JSON, JavaScript, C\#), we provide (1) its MPS definition in the form that incorporates all tweaks performed manually by authors of this paper and (2) the adjusted ANTLR v4 grammar used as input for \textsc{Ingrid}.
+%, both in the repository that contains also our implementation.
 
@@
 \paragraph{JavaScript.}
-In the case of JavaScript, which is an example of a widely-used complex general purpose language, automated generating of the language definition in MPS from the ANTLR grammar\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} and subsequent manual adjusting was done in less than one hour.
+In the case of JavaScript, which is an example of a widely-used complex general purpose programming language, automated generating of the language definition in MPS from the ANTLR grammar and subsequent manual adjusting was done in less than one hour.
+%\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}
 
@@
 Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces at least a reasonable structure (hierarchy of concepts) that can be practically used in MPS just with minor alterations.
-Necessary adjustments include, for example, removal of intermediary rules such as \antlrparserrule{statement\_list} : \antlrparserrule{statement*}~\antlrliteral{;} by inlining the right-hand side.
-The Editor aspect was modified afterwards, but only for some of the concepts generated by \textsc{Ingrid}. 
+Necessary adjustments include removal of intermediary rules such as \antlrparserrule{statement\_list} : \antlrparserrule{statement*}~\antlrliteral{;} by inlining the right-hand side.
+The Editor aspect was modified afterwards, but only for some of the concepts. 
 Roughly one hour of manual effort was needed in total.
-Nevertheless, additional modifications of the grammar and selected aspects of the MPS language (Editor, TextGen) are still needed.
+Nevertheless, additional modifications of the grammar and selected aspects of the MPS language are still needed.
 A great flexibility and complexity of the C\# language is the main reason behind all the necessary changes to the language definition in MPS and to the ANTLR grammar.
@@
 The main overall benefit of the \textsc{Ingrid} method is partial automation.
-Most of the languages discussed above, which we tried to create in MPS by using \textsc{Ingrid}, are quite complex regarding their structure and syntax variety.
+Most of the languages discussed above are quite complex regarding their structure and syntax variety.
 Therefore, completely manual definition would be a very time-consuming and error-prone process.
@@
 
-On the other hand, our experiments with complex languages show that, in the case of the Editor and TextGen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, by an order of magnitude less time than we initiall expected.
+On the other hand, our experiments with complex languages show that, in the case of the Editor and TextGen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, an order of magnitude less time than we initially expected.
 The first author spent between 20 and 60 minutes of work on each language, when adjusting the result of automated generation into a more useful and readable form.
 MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing designers to produce really useful languages very fast.
-All of this justifies our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
-Despite that, we also tried to design some automated heuristics during our work on the \textsc{Ingrid} project, but so far all yield rather mediocre results when compared to what human users can achieve efficiently instead.
+%All of this justifies our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
+Despite that, we also tried to design some automated heuristics, but so far all yield rather mediocre results when compared to what human users can achieve efficiently instead.

### 2017-06-16 chapters/language_definition.tex
ready for submission

@@
 The proposed \textsc{Ingrid} method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
-Syntax definition in the form of an ANTLR v4 grammar exist for all widely used programming languages~\footnote{https://github.com/antlr/grammars-v4}.
+Syntax definition in the form of an ANTLR v4 grammar exists for all widely used programming languages~\footnote{https://github.com/antlr/grammars-v4}.
 Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.
 
-The process of MPS language construction, as performed by \textsc{Ingrid}, consists of four phases --- the first is parsing of the input grammar, which is followed by definition of the essential aspects of the language (Structure, Editor, and TextGen, respectively).
+The process of language construction by \textsc{Ingrid} consists of four phases --- the first is parsing of the input grammar, followed by definition of the essential aspects of the language (Structure, Editor, and TextGen, respectively).
 %Each of the phases 2-4 is exclusively and fully responsible for one aspect.
 
-\textsc{Ingrid} is currently able to create (i) a full Structure aspect for each element (type of an AST node) of the given language, (ii) a very basic Editor, and (iii) a basic TextGen aspect.
-Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the human users must also define the remaining aspects not yet supported by \textsc{Ingrid}.
+\textsc{Ingrid} is currently able to create (i) a full Structure aspect for each element of the given language, (ii) a very basic Editor, and (iii) a basic TextGen aspect.
+Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the remaining aspects not yet supported by \textsc{Ingrid} must also be defined.
 
-Our approach differs from the existing projects with similar goals (Section~\ref{sect:RELATED}) especially in the level of automation.
-It also has much better support for the definition of Editor and TextGen.
+Our approach differs from the existing projects with similar goals (Section~\ref{sect:RELATED}) especially in the level of automation, and it also has much better support for Editor and TextGen.
 
@@
 
-We will describe the \textsc{Ingrid} method using the example of a simplified XML language that is defined in Figure~\ref{fig:SIMPLEXML}.
+We will describe the \textsc{Ingrid} method using a simplified XML language defined in Figure~\ref{fig:SIMPLEXML}.
 The \emph{SimpleXML} language is small but complex enough to be used for illustration of the main challenges and behavior of the proposed algorithms.
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
+\vspace{-2mm}
 \centering
-\begin{framed}
 \begin{alltt}
@@
 \antlrparserrule{document}  : \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
-\antlrparserrule{prolog}    : \antlrliteral{<?xml } \antlrparserrule{attrib}* \antlrliteral{?>} ;
+\antlrparserrule{prolog}    : \antlrliteral{<?xml } \antlrparserrule{attr}* \antlrliteral{?>} ;
 \antlrparserrule{comment}   : \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-\antlrparserrule{element}   : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{/>} ;
-\antlrparserrule{attrib}    : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
+\antlrparserrule{element}   : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attr}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
+          |  \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attr}* \antlrliteral{/>} ;
+\antlrparserrule{attr}      : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
 \antlrparserrule{content}   : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
-\antlrlexerrule{Name}      : \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
+\antlrlexerrule{Name}      : \antlrlexerrule{NameStCh} \antlrlexerrule{NameChar}* ;
 \antlrlexerrule{DIGIT}     : \antlrregex{[0-9]} ;
-\antlrlexerrule{NameChar}  : \antlrlexerrule{NameStartChar} | \antlrliteral{-} | \antlrliteral{\_} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
-\antlrlexerrule{NameStartChar} : \antlrregex{[:a-zA-Z]} ;
+\antlrlexerrule{NameChar}  : \antlrlexerrule{NameStCh} | \antlrliteral{-} | \antlrliteral{\_} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
+\antlrlexerrule{NameStCh}  : \antlrregex{[:a-zA-Z]} ;
 \antlrlexerrule{TEXT}      : \antlrregex{~[<"]*} ;
@@
 \end{alltt}
-\end{framed}
 \caption{Grammar of the SimpleXML language in the ANTLR v4 notation}
 \label{fig:SIMPLEXML}
-\end{figure*}
+\vspace{-2mm}
+\end{figure}
 
 The grammar of SimpleXML contains elements of the kinds that are listed below.
-Each color in Figure~\ref{fig:SIMPLEXML} corresponds to a different kind in a way that is indicated by the list item headers.
+Each color in Figure~\ref{fig:SIMPLEXML} corresponds to one kind in a way indicated by the list item headers.
 \begin{itemize}
@@
 		Alternatives on the right side of a rule are separated by the pipe character ($|$).
-	\item \antlrlexerrule{\textbf{Lexer rules}} describe terminal symbols that the parser matches against the input string.
-		A terminal symbol can be encoded as a string value or using a regular expression.
+	\item \antlrlexerrule{\textbf{Lexer rules}} describe terminal symbols that the parser matches against the input.
+		A terminal symbol can be encoded as a string value or a regular expression.
 	\item \antlrliteralnoap{\textbf{String literals}}, always written inside of a pair of single quote marks, also represent terminal symbols, but by exact match to a string constant.
-	\item \antlrregex{\textbf{Regular expressions}} describe string tokens to be matched against regular expressions with a special ANTLR v4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
+	\item \antlrregex{\textbf{String tokens}} are described by regular expressions with a special ANTLR v4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
 \end{itemize}
@@
 The representation of lexer rules (tokens) in the AST has to be simplified too.
-Note that, in the ANTLR notation, lexer rules can be built from alternatives just like the parser rules --- see, for example, the lexer rule \antlrlexerrule{Name} from our SimpleXML language (Figure~\ref{fig:SIMPLEXML}).
-For each lexer rule, the parser produces a tree that captures the structure of the rule.
-However, the lexer rules are, in fact, just regular expressions used to recognize tokens in the input string.
+In the ANTLR notation, lexer rules can be built from alternatives just like the parser rules --- see, for example, the lexer rule \antlrlexerrule{Name} from our SimpleXML language (Figure~\ref{fig:SIMPLEXML}).
+For each lexer rule, the parser produces a tree that captures its structure.
+The lexer rules are, in fact, just regular expressions used to recognize tokens in the input string.
 
@@
     \textbf{for each} element \textit{E} \textbf{of} \textit{A}:
-      \textbf{if} \textit{E} is not yet flattened \textbf{then}:
-        Flatten(\textit{E})
+      \textbf{if} \textit{E} is not yet flattened \textbf{then} Flatten(\textit{E})
       \textit{R}.append(\textit{E})
@@
 \label{fig:ALGFLATTEN}
+\vspace{-2mm}
 \end{figure}
@@
 
-When designing the process for translating AST nodes (i.e., grammar rules) into the MPS language structure, we faced several challenges.
-The main challenge related to the structure, as we already mentioned in the introduction, is that a grammar typically contains rules that do not directly correspond to programming language constructs.
+When designing the procedure for translating AST nodes (i.e., grammar rules) into the MPS language structure, we faced several challenges.
+The main challenge, as we already mentioned, is that a grammar typically contains rules that do not directly correspond to programming language constructs.
 Such rules exist at the intermediate layers of a syntax hierarchy.
 Their main purpose is to enable easier understanding and maintenance of the grammar by humans.
-However, presence of the intermediate layers significantly complicates usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
+Nevertheless, presence of the intermediate layers would significantly complicate usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
 We show examples illustrating this problem, which we call a \emph{layer problem}, later in this section at a point where we describe our solution.
-
 As a part of the \textsc{Ingrid} method, we have designed an approach to eliminate the unnecessary layers during construction of the Structure aspect --- we call it the \emph{shortcut} approach (Section~\ref{sect:SHORTCUT}).
@@
 Additionally, the object may implement any number of interfaces, and it may also contain references to objects representing other AST nodes.
-The parent-child relationships between objects that make the Structure aspect are derived from rules of the grammar (i.e., from the structure of the AST).
+The parent-child relationships between objects that make the Structure aspect are derived from rules of the grammar.
 For each rule, the object corresponding to the left-hand side is in the parent-child relationships with sets of objects that represent language elements referenced by the right-hand side.
@@
 \begin{figure}[ht]
+\vspace{-3mm}
 \centering
@@
 \label{fig:ELEMENTSTRUCT}
+\vspace{-1mm}
 \end{figure}
@@
 \begin{figure}[ht]
+\vspace{-1mm}
 \centering
@@
 \label{fig:CONTENTRULE}
+\vspace{-1mm}
 \end{figure}
@@
 \small
-\mpsinterface{IContent} : \mpsconcept{Content{\_}1} | \mpsconcept{Content{\_}2} |
-           \mpsconcept{Content{\_}3} | \mpsconcept{Content{\_}4}
+\mpsinterface{IContent} : \mpsconcept{Content{\_}1} | \mpsconcept{Content{\_}2} | \mpsconcept{Content{\_}3} | \mpsconcept{Content{\_}4}
 \end{alltt}
@@
 The \mpsconcept{Content{\_}2} object has only a single child node of the interface type \mpsinterface{IElement}.
-	\item Then, the user must trigger the auto-complete again and insert either an \mpsconcept{Element{\_}1} node or an \mpsconcept{Element{\_}2} node into the \mpsconcept{Content{\_}2} node.
+	\item Then, the user must again trigger auto-complete and insert either an \mpsconcept{Element{\_}1} node or \mpsconcept{Element{\_}2} into the \mpsconcept{Content{\_}2} node.
 See the right part of Figure~\ref{fig:LAYERPROBLEM}.
@@
 
-Similarly, if the user would like to replace a nested \mpsconcept{Element{\_}1} node with, for example, an XML comment (represented by the \mpsconcept{Comment} node), then both intermediary layers have to be deleted before she gets back to the original selection among the concepts \mpsconcept{Content{\_}1}, $\ldots$, \mpsconcept{Content{\_}4}.
+Similarly, if the user would like to replace a nested \mpsconcept{Element{\_}1} node with, for example, an XML comment (i.e., the \mpsconcept{Comment} node), then both intermediary layers have to be deleted before she gets back to the original selection among the concepts \mpsconcept{Content{\_}1}, $\ldots$, \mpsconcept{Content{\_}4}.
 
@@
 The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree, and consider just the nodes to be directly offered to the user through the auto-completion menu.
-Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, which is given in Figure~\ref{fig:CONTENTHIERARCHY} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the nodes that are highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.
+Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, given in Figure~\ref{fig:CONTENTHIERARCHY} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the nodes highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.
 
@@
 \centering
-\begin{framed}
 \begin{alltt}
@@
   \antlrparserrule{content} : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
-  \antlrparserrule{element} : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{/>} ;
+  \antlrparserrule{element} : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attr}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attr}* \antlrliteral{/>} ;
   \antlrparserrule{comment} : \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
 \end{alltt}
-\end{framed}
 \caption{The parser rule \antlrparserrule{content} with other rules that determine the relevant fragment of the syntax hierarchy}
@@
 \begin{figure}[ht]
-\begin{framed}
 \small
@@
 \ \ \ \textbf{Content{\_}4} (CDATA)
-\end{framed}
 \caption{Leaf nodes of the parser tree fragment that has the \antlrparserrule{content} rule as its root}
 \label{fig:CONTENTEXPAND}
+\vspace{-2mm}
 \end{figure}
@@
 \end{itemize}
-By appending the node corresponding to the current alternative (lines 10 and 15) and the rule that leads to the alternative (line 14), the algorithm creates a full path that contains the target end node as the last element of the chain.
+By appending the node corresponding to the current alternative (lines 10 and 15) and to the rule that leads to the alternative (line 14), the algorithm creates a full path that contains the target end node as the last element of the chain.
 
 We use the name \emph{shortcut approach} for this algorithm, because the paths collected by the algorithm provide shortcuts from the given rule to end nodes, by the virtue of hiding all intermediary layers.
-For example, the result of this algorithm for the \antlrparserrule{content} rule (Figure~\ref{fig:CONTENTHIERARCHY}) is the list of five items already shown in Figure~\ref{fig:CONTENTEXPAND}.
+For example, the result of this algorithm for the \antlrparserrule{content} rule (Figure~\ref{fig:CONTENTHIERARCHY}) is the list shown in Figure~\ref{fig:CONTENTEXPAND}.
 
 The primary use case for shortcuts is to generate options for the auto-completion menus, such that only the end nodes are offered.
-Shortcuts have to be considered not only when nodes are inserted, but also when they are deleted.
+Shortcuts have to be considered when nodes are inserted, and also when they are deleted.
 In each case, the whole chain including possibly multiple intermediary AST nodes (up to the end node) must be added, respectively deleted.
@@
 Cells of different types are supported by MPS --- for storing property values, references to child nodes, and keywords (string literals), and also cells that influence layout (e.g., indentation).
-The specific goal of this phase is to create the Editor aspect for each language concept (element), such that all attributes of the concept --- name, properties, children --- are projected using the respective cell types.
+The specific goal of this phase is to create the Editor aspect for each language concept, such that all its attributes --- name, properties, children --- are projected using the respective cell types.
 
-The main problem that we had to address is the absence of information about the code layout and whitespace in the ANTLR grammar of an input language.
+The main problem that we had to address is the absence of information about the code layout and whitespace in ANTLR grammars.
 Rules forming the grammar only tell what the syntax tree looks like and how the program code is decomposed into AST nodes.
@@
 A typical user of the \textsc{Ingrid} method will be able to use the projectional editor in MPS quite efficiently.
-We actually experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results --- it is quite hard to automatically identify language concepts for which the default layout should be refactored, and the user would have to adjust the layout by hand anyway.
+We actually experimented with several heuristics to derive a useful code layout automatically, but all of them produced rather suboptimal results --- it is quite hard to automatically identify language concepts for which the default layout should be refactored, and the user would have to adjust the layout by hand anyway.
 More details are provided in Section~\ref{sect:EVAL}, including our experience with several mainstream programming languages.
@@
 We have found that the combination of two steps, (1) automated placing of cells into a single row and (2) subsequent manual adjustment of the layout, is a very fast and efficient way of creating a nice code layout.
-Nevertheless, we plan to work on a more automated approach to the definition of Editor aspects in the future, using automated learning.
+Nevertheless, we plan to work on a more automated approach to the definition of Editor aspects in the future, based on machine learning.
 The key idea would be to derive the code layout from a set of valid input source files.
-We believe that the user would have to adjust the editor manually a little bit even in this case, because the results of a machine learning-based approach would not be perfect.
+%We believe that the user would have to adjust the editor manually a little bit even in this case, because the results of a machine learning-based approach would not be perfect.
 
@@
 We use the \mpsconcept{Element{\_}1} concept again for illustration.
-Figure~\ref{fig:TEXTGENBASIC} shows a very basic example of BaseLanguage code that can be used as its TextGen aspect.
-The code appends all literals, properties, and children of \mpsconcept{Element{\_}1} to the output buffer, and does that in the same order as they appear in the corresponding grammar rule.
+Figure~\ref{fig:TEXTGENBASIC} shows a fragment of BaseLanguage code that can be used as its TextGen aspect.
+The code appends all literals, properties, and children of \mpsconcept{Element{\_}1} to the output buffer.
+%, and does that in the same order as they appear in the corresponding grammar rule.
 
@@
 \label{fig:TEXTGENBASIC}
+\vspace{-3mm}
 \end{figure}
@@
 Like in the case of a projectional editor, the main challenge associated with TextGen is to produce valid code with a reasonable layout.
-An implementation of the TextGen aspect must determine properly where to put line breaks, indentation, and other whitespace characters.
-For example, in the case of the concept representing an XML element, there must be a space between the element's name and the first attribute, while it is not required between the opening bracket \antlrliteral{\textless} and the name (Figure~\ref{fig:TEXTGENBASIC}).
+An implementation of the TextGen aspect must determine properly where to put line breaks, spaces, and indentation.
+For example, in the case of the concept representing an XML element, there must be a space between the element's name and the first attribute, while it is not required between the opening bracket \antlrliteral{\textless} and the name.
+%(Figure~\ref{fig:TEXTGENBASIC}).
 
@@
 
-The procedure creates the TextGen aspect for a given language concept (AST node) in two steps.
+The procedure creates the TextGen aspect for a given language concept in two steps.
 First, it generates a basic variant that inserts spaces in between every two tokens of the textual representation of the concept.
-In the second step, which is the core of our heuristic, spaces are eliminated from places where they are not really needed.
-The main criterion is whether the generated plain-text output can be safely processed by a parser of the original language (using the ANTLR grammar).
+In the second step, spaces are eliminated from places where they are not really needed.
+The main criterion is whether the generated textual output can be accepted by a parser of the original language (using the ANTLR grammar).
 We discuss all the cases here:
@@
 		Based on this heuristic, redundant spaces inside of quotes will be eliminated, as well as spaces next to semicolons and around brackets.
-		On the other hand, it will distinguish language keywords (e.g., \code{function}, \code{var}, \code{in}) from other literals by preserving the space character in between them.
-	\item Spaces can be safely omitted also when specific optional child nodes are not present, and in the case of empty lists of child nodes, so that whitespace does not accumulate.
+		On the other hand, it will separate language keywords from other literals by preserving the space characters between them.
+	\item Spaces can be safely omitted also when optional child nodes are not present, and in the case of empty lists of child nodes, so that whitespace does not accumulate.
 \end{itemize}
@@
 \begin{figure}[ht]
+\vspace{-2mm}
 \begin{alltt}
@@
 \label{fig:TEXTGENFINAL}
+\vspace{-1mm}
 \end{figure}
@@
 \end{alltt}
+\vspace{-1mm}
 \caption{Fragment of the TextGen aspect of \mpsconcept{Element{\_}1} with adjusted indentation}
 \label{fig:TEXTGENADJUSTED}
+\vspace{-2mm}
 \end{figure}
 
-Figure~\ref{fig:TEXTGENFINAL} contains the complete, automatically generated, TextGen aspect for the \mpsconcept{Element{\_}1} concept, which represents the full SimpleXML element.
-Such code may be adjusted manually very easily in order to produce nicely indented XML documents.
+Figure~\ref{fig:TEXTGENFINAL} contains the complete, automatically generated, TextGen aspect for the \mpsconcept{Element{\_}1} concept.
+%, which represents the full SimpleXML element.
+Such code may be adjusted by hand very easily in order to produce nicely indented XML documents.
 We only need to wrap the \mpsconcept{Content{\_}2} child node with indentation and change the sequence separator to a new line character.
-A fragment of the resulting adjusted aspect (code) is in Figure~\ref{fig:TEXTGENADJUSTED}.
+A fragment of the resulting adjusted code is in Figure~\ref{fig:TEXTGENADJUSTED}.
 
@@
 \begin{figure}[ht]
-\begin{framed}
+\vspace{-1mm}
 \begin{alltt}
 \small
-  \antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
+  \antlrparserrule{attr} : \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
   \antlrlexerrule{STRING} : \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
@@
 \end{alltt}
-\end{framed}
 \caption{Definition of an XML attribute}
@@
 \begin{figure}[ht]
-\begin{framed}
 \begin{alltt}
 \small
-  \antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
+  \antlrparserrule{attr} : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
          | \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'} ;
@@
 \end{alltt}
-\end{framed}
 \caption{Adjusted fragment of the SimpleXML grammar}
@@
 \begin{figure}[ht]
-\begin{framed}
+\vspace{-1mm}
 \begin{alltt}
@@
 \end{alltt}
-\end{framed}
 \caption{Statements in JavaScript}
 \label{fig:JAVASCRIPTSTMT}
+\vspace{-1mm}
 \end{figure}
@@
 For example, we can put each statement on a distinct line, as is usual for JavaScript code, and keep just the semicolon as a fixed literal in the projectional editor.
-The \antlrparserrule{eos} rule must be changed to this form: \antlrparserrule{eos} : \antlrliteral{;}.
-A small adjustment of the grammar, like this one, is a very quick solution and definitely makes the generated language more usable in MPS.
+The \antlrparserrule{eos} rule would have to be changed to this form: \antlrparserrule{eos} : \antlrliteral{;}.
+A small adjustment of the grammar, like this one, is a very quick solution and makes the generated language more usable in MPS.
 On the other hand, this adjustment of the original ANTLR grammar, performed just for the purpose of creating the MPS language, changes the grammar in such a way that it does not precisely describe the standard JavaScript language.

### 2018-05-03 paper.tex
making changes
editing
updated abstract

@@
 \newcommand{\antlrlexerrule}[1]{\textcolor{red}{#1}}
-\newcommand{\antlrliteralnoap}[1]{\textcolor{orange}{#1}}
+\newcommand{\antlrliteralnoap}[1]{\textcolor{brown}{#1}}
 \newcommand{\antlrliteral}[1]{\antlrap\antlrliteralnoap{#1}\antlrap}
@@
 \begin{abstract}
-JetBrains MPS is a language workbench, an IDE that allows developers both to write code and create or modify language definitions.
+JetBrains MPS is a language workbench, an IDE that allows developers both to write code and create language definitions.
 It leverages the concept of projectional editing, where the developer directly manipulates the AST representation of program code.
-While MPS focuses on domain-specific languages (DSL), it needs to support also general-purpose programming languages (GPPLs) --- for example, to compile and run programs written in some DSL, or to enable combination of GPPLs with DSLs created in MPS.
+While MPS focuses on domain-specific languages (DSL), it needs to support also general-purpose programming languages (GPPLs) --- for example, to compile and run programs written in some DSL.
+%, or to enable combination of GPPLs with DSLs created in MPS.
 
 We present \textsc{Ingrid}, a method for construction of a language definition in MPS based on its ANTLR grammar.
-The structure of a language is generated automatically, but its projectional editor and other aspects must be adjusted manually by the user.
+The structure of a language is generated automatically, but its projectional editor and other aspects has to be adjusted manually.
+%by the user.
 During the development of \textsc{Ingrid}, we encountered several challenges related to the mapping between grammars and language definitions in MPS that are based on object-oriented principles.
@@
 
+\bibitem{ref:KV10} L.C.L. Kats and E. Visser. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. OOPSLA 2010, ACM.
+
 \bibitem{ref:ANTLRBOOK} T. Parr. The Definitive ANTLR 4 Reference. Pragmatic Bookshelf, 2013.
@@
 \bibitem{ref:MBEDDR} M. Voelter, D. Ratiu, B. Schaetz, and B. Kolb. mbeddr: an Extensible C-based Programming Language and IDE for Embedded Systems. SPLASH 2012, ACM.
+
+\bibitem{ref:VSB14} M. Voelter, J. Siegmund, T. Berger, and B. Kolb. Towards User-Friendly Projectional Editors. SLE 2014, LNCS, 8706.
 

### 2018-05-06 chapters/background.tex
minor changes

@@
 Programs are created by assembling the tree (AST) out of predefined building blocks of selected languages.
-The definition of a language in MPS dictates, where in the AST can certain elements be placed and how they can be nested inside other elements.
+The definition of a language in MPS dictates where in the AST certain elements can be placed and how they can be nested inside other elements.
 
@@
 The current version, as of November 2016, supports just horizontal lists, vertical lists, and some indentation rules.
-However, even these few features make the syntax much more complicated.
+However, even these few features make the extended syntax of ANTLR v4 much more complicated.
 
@@
 Its author created an ANTLR v4 MPS language, which captures the syntax of ANTLR v4 notation using MPS concepts.
-Given some language as input, the textual grammar of the language is imported automatically into MPS, taking the form of the ANTLR v4 MPS language.
+Given the textual grammar of some language as input, the grammar is imported automatically into MPS, taking the form of the ANTLR v4 MPS language.
 The next step would be to create a new MPS language from the grammar definition in MPS, but it is not implemented yet.

### 2018-05-06 chapters/introduction.tex
minor changes

@@
 
-An alternative approach is projectional (or structural) editing~\cite{ref:VWK15,ref:VSB14}, where a developer directly manipulates the AST representation of the source code instead of plain text.
+An alternative approach is projectional (or structural) editing~\cite{ref:VWK15,ref:VSB14,ref:DHL80}, where a developer directly manipulates the AST representation of the source code instead of plain text.
 This idea emerged as early as in 1970s, but it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
@@
 \begin{itemize}
-	\item The developers practically cannot enter syntactically invalid code, because a projectional editor controls the interaction between the user and the program code.
+	\item A projectional editor does not allow the developers to enter syntactically invalid code, because it controls the interaction between the user and the program code.
 		%An AST created by the editor can represent only a syntactically valid code.
@@
 This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
-Therefore, the target GPPL must also be defined by the MPS means in order to allow the DSLs to have their models transformed into a model in the GPPL.
+Therefore, the target GPPL must also be defined in MPS, using the respective infrastructure, in order to allow the DSLs to have their models transformed into a model in the GPPL.
 
@@
 In this paper we present \textsc{Ingrid}, a method and a tool for semi-automated construction of a language definition in MPS that uses an ANTLR grammar as a description of its syntax.
-The process of construction is semi-automated for two reasons: (1) ANTLR grammars have to be adjusted before \textsc{Ingrid} can process them, and (2) some aspects of the generated MPS languages typically must be tweaked or created manually in order to improve languages' usability.
+The process of construction is semi-automated for two reasons: (1) ANTLR grammars have to be adjusted before \textsc{Ingrid} can process them, and (2) some aspects of the language definitions in MPS typically must be tweaked or created manually in order to improve languages' usability.
 

### 2018-05-06 chapters/language_definition.tex
minor changes

@@
 It is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
-Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the \textsc{Ingrid} algorithm.
-In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST are performed in this phase.
-An output is a custom AST-like structure representing the grammar, especially the hierarchy of parser rules.
+Nevertheless, the full parse tree, which comes out of the parser, is quite complex and contains information not relevant for the \textsc{Ingrid} algorithm.
+In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the parse tree are performed in this phase.
+An output is a custom AST-like structure that represents the grammar, especially the hierarchy of parser rules, in a way more suitable for MPS.
 
-The representation of lexer rules (tokens) in the AST has to be simplified too.
+The representation of lexer rules (tokens) in the full parse tree has to be simplified too.
 In the ANTLR notation, lexer rules can be built from alternatives just like the parser rules --- see, for example, the lexer rule \antlrlexerrule{Name} from our SimpleXML language (Figure~\ref{fig:SIMPLEXML}).
@@
 The main problem that we had to address is the absence of information about the code layout and whitespace in ANTLR grammars.
-Rules forming the grammar only tell what the syntax tree looks like and how the program code is decomposed into AST nodes.
+Rules forming the grammar only tell what the syntax tree looks like and how the program text is decomposed into AST nodes.
 Here we present a solution that is only partially automated for reasons explained below.

### 2018-05-07 chapters/background.tex
shortening

@@
-\section{Background and Related Work}
-\label{sect:BACKGRELWORK}
+\section{Background}
+\label{sect:BACKGROUND}
 
-In this section, we introduce basic features of the MPS platform, and then we discuss other existing projects that aim to add support for general purpose languages into MPS.
+In this section, we introduce basic features of the MPS platform, and then we briefly discuss existing projects that aim to add support for general purpose languages into MPS.
 
@@
 
-JetBrains MPS is a language workbench --- an IDE that allows developers to create their own languages and use them to write code.
-The code can then be transformed into a target language, typically a GPPL such as Java or C, and eventually compiled into executable programs.
+MPS is a language workbench --- an IDE that allows developers to create new languages and use them to write code.
+The code can be transformed into a target language, typically a GPPL such as Java or C, and compiled into executables.
 
-When using the MPS projectional editor, the developer does not work with the textual representation of the source code, but rather directly with its AST that is the model of the code.
-Programs are created by assembling the tree (AST) out of predefined building blocks of selected languages.
+When using the MPS projectional editor, the developer creates programs by assembling the AST out of predefined building blocks of selected languages.
 The definition of a language in MPS dictates where in the AST certain elements can be placed and how they can be nested inside other elements.
 
-The building blocks of MPS models are called nodes.
 Code of any program in MPS is built from nodes, which represent instances of concepts from the languages that the program is written in.
 In MPS, a \emph{concept} is a language element, i.e. a building block of a language definition.
-We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.
 
 One of the key advantages of projectional editing stems from the separation of abstract and concrete syntax.
-While AST provides a complete and precise representation of the code, the way it is displayed on the screen and the way the user interacts with it are unconstrained.
-The editor can take any visual form and shape.
+While AST provides a complete and precise representation of the code, the way it is displayed on the screen and the way the user edits it are unconstrained.
+The editor can take any visual form and shape (i.e., not just textual).
 The language author can define multiple alternative visualizations and let the developer choose one that fits best the task at hand.
-%--- for example, editing, debugging, reviewing, resolving merge conflicts, etc.
-In particular, the visual representations are not bound to be just textual at all.
-%For example, blocks corresponding to branches of an \code{if-then-else} statement can be aligned next to each other and displayed with different background colors.
 
@@
 Structure represents the abstract syntax (types and hierarchy of AST nodes), Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
-If, instead of generating text directly, programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
-Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we describe them below in more detail, and neglect other aspects such as type constraints.
+If programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
+Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we neglect other aspects such as type constraints.
 
-Languages are built from the concepts using techniques known from object-oriented programming --- containment, inheritance, interfaces, and so on.
-Therefore, a definition of a whole language in MPS typically has an object-oriented and hierarchical nature.
+Languages are built from the concepts using techniques known from object-oriented programming --- containment, inheritance, and so on.
+Therefore, the definition of a language in MPS has an object-oriented and hierarchical nature.
 
@@
 The fundamental aspect of any MPS language is Structure.
-It must be created first for each intended language concept.
-Structure specifies core attributes of an MPS concept such as the name, inheritance relationships, child concepts (their types and cardinalities), implemented interfaces, and references to other AST nodes.
-% other properties (fields of any type that can hold values), etc.
-Figure~\ref{fig:if_statement_structure} shows the Structure aspect for the \code{if-then-else} statement.
+It specifies core attributes of a language concept such as the name, inheritance relationships, child concepts (their types and cardinalities), and references to other AST nodes.
+Figure~\ref{fig:if_statement_structure} shows the most important fragments of the Structure aspect for the \code{if-then-else} statement.
 
@@
         \mpsstkeyword{implements} IContainsStatementList
-                   IDontSubstituteByDefault
-                   IConditional
 
@@
   \mpsstkeyword{alias:} \mpsstalias{if}
-  \mpsstkeyword{short description:} \mpsstplaceholder{<no short description>}
-
-  \mpsstkeyword{properties:}
-  \mpsstproperty{forceOneLine}   : boolean
-  \mpsstproperty{forceMultiLine} : boolean
 
@@
   \mpsstproperty{elsifClauses}     : ElsifClause[\mpsstcardinality{0..n}]
-
-  \mpsstkeyword{references:}
-  \mpsstplaceholder{<< ... >>}
 \end{alltt}
@@
 JetBrains have developed a cellular system that enables placing properties and children of a node (concept) into different cells.
-The author usually incorporates all of the node's children, references, and properties inside the representation, so that future users of the language can insert all values that the node expects.
-Additionally, cells of the editor can be styled using a language similar to CSS.
+Individual cells of the editor can be styled using a language similar to CSS.
 Supported visual characteristics include color and indentation.
@@
 \paragraph{BaseLanguage.}
-Another important feature of MPS that we need to describe is \emph{BaseLanguage}~\cite{ref:BASELANG}, a clone of Java implemented using the MPS constructs.
-BaseLanguage was developed in the early days of MPS in order to implement MPS itself and also to support the basic set of language-definition DSLs,
-Although BaseLanguage is syntactically almost identical to Java, it is edited in a projectional editor, just like all the languages in MPS.
-The language-definition DSLs, used to define custom languages by their authors, are generated into the BaseLanguage.
-Similarly, all the custom DSLs that are meant to be generated into Java choose BaseLanguage as their generation target.
-The conversion to textual Java sources is handled by BaseLanguage (without any further manual effort), through its TextGen aspect.
+Another very important feature of MPS is \emph{BaseLanguage}~\cite{ref:BASELANG}, a clone of Java implemented using the MPS constructs.
+BaseLanguage is syntactically almost identical to Java, but it is edited in a projectional editor, just like all the languages in MPS.
+Custom DSLs meant to be generated into Java choose BaseLanguage as their generation target.
 
@@
 DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since programs in such languages are rarely converted to text directly.
-After TextGen has generated textual sources from an AST, a compiler for the particular GPPL is invoked to compile the textual sources into binary code.
-The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes at the right moments.
+The TextGen definition follows a straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes.
 MPS calls the corresponding method on the root AST nodes of the given program.
@@
 
-The BaseLanguage~\cite{ref:BASELANG} is an almost full port of Java, which was created by JetBrains and extended with MPS-specific features.
-% and it is still undergoing changes as Java itself is evolving.
+The BaseLanguage~\cite{ref:BASELANG} is an almost full port of Java, extended with MPS-specific features.
 The C language has also been manually tailored for MPS within the mbeddr project~\cite{ref:MBEDDR}.
@@
 
-\paragraph{PE4MPS~\cite{ref:PE4MPS}.} This is a project that addresses the lack of information about code layout in grammars by using so called PE grammars~\cite{ref:PE}, which is a new notation proposed by the same author.
+\paragraph{PE4MPS~\cite{ref:PE4MPS}.}
+This project addresses the lack of information about code layout in grammars by using so called PE grammars~\cite{ref:PE}, which is a new notation proposed by the same author.
 PE stands for projectional editing.
@@
 PE grammars extend the syntax of the ANTLR v4 notation by constructs that provide information about the possible layout of AST nodes.
-The current version, as of November 2016, supports just horizontal lists, vertical lists, and some indentation rules.
-However, even these few features make the extended syntax of ANTLR v4 much more complicated.
-
 The PE4MPS tool creates an MPS language in a single atomic step.
-An input PE file is loaded and processed by the PE parser, and then all concepts (AST nodes) and their aspects are directly generated inside MPS.
 Information about the code layout, extracted from the PE file, is used when generating the projectional editor for every AST node.
@@
 The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
-Language developers are forced to specify the code layout manually in PE grammars, through the corresponding extensions.
 Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
-Like for \textsc{Ingrid}, every grammar might require non-trivial adjustments before it can be processed using the PE4MPS tool.
-%To compare, the goals behind our \textsc{Ingrid} approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.
 
-We discuss application of PE4MPS on JavaScript in Section~\ref{sect:EVAL}, and compare its results to our \textsc{Ingrid} approach.
-
-\paragraph{ANTLR{\_}MPS~\cite{ref:ANTLR2MPS}.} This project also uses the ANTLR v4 grammar notation, but otherwise works quite differently from PE4MPS.
-%We provide only a brief mention here because this project is in an early stage of development.
-Its author created an ANTLR v4 MPS language, which captures the syntax of ANTLR v4 notation using MPS concepts.
+\paragraph{ANTLR{\_}MPS~\cite{ref:ANTLR2MPS}.}
+This project also uses the ANTLR v4 grammar notation, but otherwise works quite differently from PE4MPS.
+Its author created an MPS language that captures the syntax of ANTLR v4 notation using MPS concepts.
 Given the textual grammar of some language as input, the grammar is imported automatically into MPS, taking the form of the ANTLR v4 MPS language.
@@
 
-However, the project has also other important limitations.
-The tool does not generate any parent-child relationships in the Structure aspect, and it provides no support for the projectional editor at all (by completely neglecting the Editor and TextGen aspects).
+However, the project has several important limitations.
+The tool does not generate any parent-child relationships in the Structure aspect, and it completely neglects the Editor and TextGen aspects.
 
-\paragraph{mps-metabnf~\cite{ref:MPSMETABNF}.} This tool, implemented by members of the DSLFoundry group, builds upon similar ideas as the ANTLR{\_}MPS project described above.
-Authors of this project, too, created an MPS language to capture ANTLR grammars.
-Every imported grammar is represented using this ANTLR v4 MPS language.
+\paragraph{mps-metabnf~\cite{ref:MPSMETABNF}.}
+This tool builds upon similar ideas as the ANTLR{\_}MPS project.
+Every imported grammar is represented using an ANTLR v4 MPS language.
 A user is then able to adjust the grammar as he wishes, specifying the code layout in the projectional editor.
@@
 The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLR v4 MPS language into the definition of the actual target MPS language.
-
-\paragraph{Other IDEs.}
-Note that while the \textsc{Ingrid} method has been designed primarily for MPS, and can be easily adapted to other IDE tools based on projectional editing, there also exist several language workbenches that use parser generators (such as ANTLR) and support free editing of textual source code.
-A prominent member of the latter category of tools is Spoofax~\cite{ref:KV10}, which essentially restricts the syntax of DSLs to linear text.
-\textsc{Ingrid} is not applicable for creating languages inside such language workbenches.

### 2018-05-07 chapters/introduction.tex
shortening

@@
 
-The prevailing approach to define valid syntaxes for programming languages is through grammars, which are typically written in notations based on the Extended Backus-Naur Form (EBNF).
+The prevailing approach to define valid syntaxes for programming languages is through grammars written in notations based on the Extended Backus-Naur Form (EBNF).
 Many tools exist for automated generation of parsers from the grammar definitions --- for example, ANTLR~\cite{ref:ANTLRBOOK,ref:ANTLR} and Bison/Flex~\cite{ref:BISONFLEX}.
-The source code of programs is written in a text editor, usually embedded into an IDE. Parsers are used to create an abstract syntax tree (AST) representation of the code, e.g. inside the compiler.
+The source code of programs is written in a text editor, while parsers are used to create an abstract syntax tree (AST) representation of the code.
 
 An alternative approach is projectional (or structural) editing~\cite{ref:VWK15,ref:VSB14,ref:DHL80}, where a developer directly manipulates the AST representation of the source code instead of plain text.
-This idea emerged as early as in 1970s, but it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
-A recent revival of projectional editing has been observed in the area of language workbenches --- IDE-like tools that enable the developers to manipulate the actual language definition.
+Recently, projectional editing has become widely used in the area of language workbenches --- IDE-like tools that enable the developers to manipulate language definitions.
 Popular examples are JetBrains MPS~\cite{ref:MPS,ref:MPSBOOK} and Spoofax~\cite{ref:KV10}.
-In particular, JetBrains MPS (Meta Programming System) is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
+In particular, JetBrains MPS is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
 MPS provides the whole IDE infrastructure that enables developers to design custom languages, use them to write program code, and compile the code into executables.
-%We provide more technical details about MPS in Section~\ref{sect:MPS}.
-In the rest of the paper, we will often use examples from MPS to illustrate the key concepts of our approach and related issues.
-
-%MPS has put extra effort into making text manipulation feel like editing real text.
-%MPS models, although being persisted as XML or binary files, can be versioned using the industry standard source-control systems (Git, Subversion).
-%On-line model browsers have been incorporated into some of the popular public source code repositories (BitBucket) [reference] .
 
@@
 \begin{itemize}
-	\item A projectional editor does not allow the developers to enter syntactically invalid code, because it controls the interaction between the user and the program code.
-		%An AST created by the editor can represent only a syntactically valid code.
+	\item A projectional editor does not allow the developers to enter syntactically invalid code.
 	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together in a single program or one can extend another.
-		%, while parsers seriously limit language modularity for traditional parsed languages.
 	\item The languages may support diverse contextual or non-parseable notations with complex layout, such as tables, diagrams, and mathematical expressions.
@@
 \end{itemize}
-All of this is useful especially for DSLs, which are frequently used by domain experts who may not be professional software developers.
-The mbeddr project~\cite{ref:MBEDDR} illustrates the abilities of language workbenches in general and of JetBrains MPS in particular.
-Voelter et al.~\cite{ref:VSB14} discuss the benefits and limitations of projectional editors in more detail.
+All of this is useful for DSLs, which are often used by domain experts who may not be professional software developers.
 
@@
 
-%Again, we provide additional technical details in Section~\ref{sect:MPS}.
-
 IDE tools based on projectional editors are typically used for syntactically rich DSLs.
-%, which are likely to benefit most from overcoming the limitations of parser-based languages.
 In the specific case of MPS, before a program written in a DSL can be executed, it is transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose programming language (GPPL), such as C or Java.
 This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
-Therefore, the target GPPL must also be defined in MPS, using the respective infrastructure, in order to allow the DSLs to have their models transformed into a model in the GPPL.
+Therefore, the target GPPL must also be defined in MPS, using the respective infrastructure.
 
@@
 Only Java and C have been implemented to date.
-%, which limits the authors of DSLs to only target these supported GPPLs.
 The overall goal of our project is to automatize the process of language definition in MPS as much as possible.
-%, exploiting the fact that a big part of this process is quite straightforward.
-We believe that such an automated process would encourage and speed-up the migration of more GPPLs into MPS, thus giving the authors of DSLs more options regarding target general languages.
+Greater automation would encourage and speed-up the migration of more GPPLs into MPS, thus giving the authors of DSLs more options regarding target general languages.
 Another possible application is the combination of general-purpose languages with DSLs natively created in MPS.
-%include the support for writing programs in GPPLs using the projectional editor and
 
@@
 \textsc{Ingrid} can also make the construction of DSLs more efficient.
-Based on our experience, we believe that, especially for simple languages, it is less time consuming to write their grammar by hand and then create the language definition in MPS with the help of \textsc{Ingrid}, than to create everything manually using the MPS GUI.
-
-%\textsc{Ingrid} allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
-
-Although we define the \textsc{Ingrid} method in the context of MPS, and illustrate everything using examples from MPS and ANTLR grammars, it can be easily adapted to other language workbenches (i) that use projectional editing and (ii) where the language definition is based on a structural object-oriented approach.
-Most of the challenges that we discuss and key principles behind our method are general and therefore apply also to other language workbenches.
+Our experience shows that, especially for simple languages, it is less time consuming to write their grammar by hand and then create the language definition in MPS with the help of \textsc{Ingrid}, than to create everything manually using the MPS GUI.
 
@@
 \begin{itemize}
-	\item The \textsc{Ingrid} method for construction of MPS language definitions from ANTLR grammars, which can be used (1) for the import of general-purpose languages into MPS and (2) for efficient construction of DSLs.
+	\item The \textsc{Ingrid} method and tool for construction of MPS language definitions from ANTLR grammars, which can be used (1) for the import of general-purpose languages into MPS and (2) for efficient construction of DSLs.
 \textsc{Ingrid} can be adapted also to other language workbenches based on similar principles as MPS.
@@
 	\item Discussion of the practical usability of generated MPS languages and its dependence on (1) the specific form of ANTLR grammars and (2) the manual adjustments of input grammars and specific aspects of MPS languages (e.g., with respect to code layout).
-Details are provided in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:REMARKSGRAMMARS}.
 \end{itemize}
 
-The rest of the paper has the following structure.
-In the next section, we provide (i) all information about MPS that is necessary to understand the examples presented in the paper and (ii) a brief overview of related work, including projects with similar goals as \textsc{Ingrid}.
-Then, in Section~\ref{sect:LANGDEF}, we describe in detail our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
+In the next section, we provide (i) the necessary information about the MPS infrastructure for language definition and (ii) a brief overview of projects with similar goals as \textsc{Ingrid}.
+Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution.
 We also discuss our experience with application of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\# (Section~\ref{sect:EVAL}), and then we conclude.
+
+Note also that we omitted many important and interesting details from this paper because of limited space.
+Readers can find the additional details in the full version~\cite{ref:TRFULL}.

### 2018-05-08 chapters/language_definition.tex
shortening

@@
 The proposed \textsc{Ingrid} method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
-Syntax definition in the form of an ANTLR v4 grammar exists for all widely used programming languages~\footnote{https://github.com/antlr/grammars-v4}.
-Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.
+Syntax definition in the form of an ANTLR v4 grammar exists for all widely used programming languages. It is available at \url{https://github.com/antlr/grammars-v4}.
+Unlike some of the related projects, we did not extend the notation with any custom features.
 
-The process of language construction by \textsc{Ingrid} consists of four phases --- the first is parsing of the input grammar, followed by definition of the essential aspects of the language (Structure, Editor, and TextGen, respectively).
-%Each of the phases 2-4 is exclusively and fully responsible for one aspect.
+The process of language construction by \textsc{Ingrid} consists of four phases --- the first is parsing of the input grammar, followed by definition of the essential aspects of the language.
 
@@
 
-We will describe the \textsc{Ingrid} method using a simplified XML language defined in Figure~\ref{fig:SIMPLEXML}.
-The \emph{SimpleXML} language is small but complex enough to be used for illustration of the main challenges and behavior of the proposed algorithms.
+We describe the \textsc{Ingrid} method, and illustrate the main challenges, using a simplified XML language defined in Figure~\ref{fig:SIMPLEXML}.
 
@@
 	\item \antlrparserrule{\textbf{Parser rules}} describe the structure of a language.
-		Alternatives on the right side of a rule are separated by the pipe character ($|$).
 	\item \antlrlexerrule{\textbf{Lexer rules}} describe terminal symbols that the parser matches against the input.
 		A terminal symbol can be encoded as a string value or a regular expression.
-	\item \antlrliteralnoap{\textbf{String literals}}, always written inside of a pair of single quote marks, also represent terminal symbols, but by exact match to a string constant.
+	\item \antlrliteralnoap{\textbf{String literals}} also represent terminal symbols, but by exact match to a string constant.
 	\item \antlrregex{\textbf{String tokens}} are described by regular expressions with a special ANTLR v4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
@@
 It is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
-Nevertheless, the full parse tree, which comes out of the parser, is quite complex and contains information not relevant for the \textsc{Ingrid} algorithm.
-In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the parse tree are performed in this phase.
+The full parse tree, which comes out of the parser, is quite complex and contains information not relevant for \textsc{Ingrid}.
+In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the parse tree are performed.
 An output is a custom AST-like structure that represents the grammar, especially the hierarchy of parser rules, in a way more suitable for MPS.
@@
 In the ANTLR notation, lexer rules can be built from alternatives just like the parser rules --- see, for example, the lexer rule \antlrlexerrule{Name} from our SimpleXML language (Figure~\ref{fig:SIMPLEXML}).
-For each lexer rule, the parser produces a tree that captures its structure.
-The lexer rules are, in fact, just regular expressions used to recognize tokens in the input string.
-
-Every tree that represents a lexer rule is flattened into the equivalent regular expression by the recursive algorithm in Figure~\ref{fig:ALGFLATTEN}.
-A sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.
-Our implementation of the algorithm detects rules defined in a recursive way (e.g., using the pattern \antlrlexerrule{N} : \antlrliteral{text} \antlrlexerrule{N}) and rejects the input grammar if it contains such rules.
-However, note that we use the flatenning algorithm only for lexer rules, where recursion is quite rare --- specifically, the algorithm is not applied to parser rules.
-
-\begin{figure}[ht]
-\vspace{-2mm}
-\centering
-\begin{framed}
-\begin{alltt}
-\small
-Flatten(\textit{R}):
-  \textit{T} = empty list
-  \textbf{for each} alternative \textit{A} \textbf{in} rule \textit{R}:
-    \textit{R} = \antlrap\antlrap
-    \textbf{for each} element \textit{E} \textbf{of} \textit{A}:
-      \textbf{if} \textit{E} is not yet flattened \textbf{then} Flatten(\textit{E})
-      \textit{R}.append(\textit{E})
-      \textit{R}.append(\textit{E}.operator)
-    \textit{T}.add(\textit{R})
-  build string S from elements of \textit{T}:
-    \textit{S} = \(t\sb{1}\) | \(t\sb{2}\) | \(\ldots\) | \(t\sb{n}\)
-  \textbf{return} \textit{S}
-\end{alltt}
-\end{framed}
-\caption{Flattening algorithm}
-\label{fig:ALGFLATTEN}
-\vspace{-2mm}
-\end{figure}
-
-The output of \texttt{Flatten(Name)}, i.e. application of the algorithm to the \antlrlexerrule{Name} rule, is this regular expression:
-
-\begin{center}
-  \texttt{[:a-zA-Z]([:a-zA-Z]|\textbackslash-|{\_}|\textbackslash.|[0-9])*}
-\end{center}
-It defines the syntactically valid identifiers of elements and attributes in SimpleXML.
-Each identifier must begin with a letter, followed by any combination of letters, digits, underscore, dash, or a dot.
+Every tree that represents a lexer rule is flattened into the equivalent regular expression.
+We describe the flattening algorithm in the full version of this paper~\cite{ref:TRFULL}.
 
@@
 Nevertheless, presence of the intermediate layers would significantly complicate usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
-We show examples illustrating this problem, which we call a \emph{layer problem}, later in this section at a point where we describe our solution.
-As a part of the \textsc{Ingrid} method, we have designed an approach to eliminate the unnecessary layers during construction of the Structure aspect --- we call it the \emph{shortcut} approach (Section~\ref{sect:SHORTCUT}).
-However, before focusing on the intermediate layers, we describe the basic principles of translation from the AST into the language structure.
+We show examples illustrating this problem, which we call a \emph{layer problem}, later in this section.
+As a part of the \textsc{Ingrid} method, we have designed an approach to eliminate the unnecessary layers during construction of the Structure aspect --- we call it the \emph{shortcut} approach.
+However, before focusing on the intermediate layers, we describe the basic principles of translation from AST into Structure.
 
 In MPS, each concept of the language (i.e., every AST node) is represented by an object that may have a parent, some children, and properties of any data type.
-Additionally, the object may implement any number of interfaces, and it may also contain references to objects representing other AST nodes.
+Additionally, the object may also contain references to objects representing other AST nodes.
 The parent-child relationships between objects that make the Structure aspect are derived from rules of the grammar.
-For each rule, the object corresponding to the left-hand side is in the parent-child relationships with sets of objects that represent language elements referenced by the right-hand side.
 If a rule has multiple alternatives, then a distinct object (MPS concept) has to be created in the structure for each alternative.
-The name of a concept (object) in MPS is composed from (1) the name of the AST node, which is equivalent to the string encoding of the left-hand side of the corresponding rule, and (2) the number indicating the position of the respective alternative on the right-hand side of the rule.
+The name of a concept (object) in MPS is composed from (1) the name of the AST node and (2) the number indicating the position of the respective alternative on the right-hand side of the rule.
 
@@
         \mpsstkeyword{implements} IContent, IElement
-
-  \mpsstkeyword{instance can be root:} false
-  \mpsstkeyword{alias:} \mpsstalias{< > </ >}
-  \mpsstkeyword{short description:} \mpsstalias{Element}
 
@@
 Its first alternative represents the full XML element with content.
-Figure~\ref{fig:ELEMENTSTRUCT} shows the Structure aspect for the MPS concept (object) named \code{Element{\_}1} that corresponds to the alternative.
+Figure~\ref{fig:ELEMENTSTRUCT} shows the relevant fragments of the Structure aspect for the MPS concept (object) named \code{Element{\_}1} that corresponds to the alternative.
 The object contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name}.
-Values of these properties will be restricted using the regular expression that corresponds to the rule.
 String literals, such as the opening and closing brackets (\antlrliteral{\textless}, \antlrliteral{/\textgreater}) in XML, are omitted because they will be defined only in the Editor aspect for this concept.
@@
 Consider the \antlrparserrule{content} rule from the SimpleXML language.
-We show just the relevant fragment of the grammar again in Figure~\ref{fig:CONTENTRULE}.
 An object corresponding to any one of the four alternatives could be the actual value anywhere the \antlrparserrule{content} rule is referenced.
-
-\begin{figure}[ht]
-\vspace{-1mm}
-\centering
-\begin{alltt}
-\small
-\antlrparserrule{content} : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
-\end{alltt}
-\caption{Parser rule \antlrparserrule{content}}
-\label{fig:CONTENTRULE}
-\vspace{-1mm}
-\end{figure}
 
@@
 Now we can illustrate the layer problem on the behavior of auto-completion in MPS.
-Suppose that a user is creating a new document in the SimpleXML language, just inserted a fresh node of the type \mpsconcept{Element{\_1}} (see above), and would like to insert another nested XML element inside.
+Suppose that a user just inserted a fresh node of the type \mpsconcept{Element{\_1}}, and would like to insert another nested XML element inside.
 The auto-completion mechanism of MPS offers four options that are displayed in the left part of Figure~\ref{fig:LAYERPROBLEM}.
-Each option represents one of the MPS concepts that implement the interface \mpsinterface{IContent}, and therefore also one alternative of the \antlrparserrule{content} rule.
+Each option represents one of the MPS concepts that implement the interface \mpsinterface{IContent}.
 
@@
 Intermediary layers have no visual appearance, and therefore it is very difficult for users to see what is actually happening and they may get confused very easily.
-The layer problem is addressed by the shortcut approach, which we describe in the next subsection.
+The layer problem is addressed by the shortcut approach, which we describe briefly in the following paragraphs.
 
-\subsubsection{The Shortcut Approach}
-\label{sect:SHORTCUT}
-
+\paragraph{The Shortcut Approach.}
 The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree, and consider just the nodes to be directly offered to the user through the auto-completion menu.
-Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, given in Figure~\ref{fig:CONTENTHIERARCHY} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the nodes highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.
-
-\begin{figure*}[ht]
-\centering
-\begin{alltt}
-\small
-  \antlrparserrule{content} : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
-  \antlrparserrule{element} : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attr}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attr}* \antlrliteral{/>} ;
-  \antlrparserrule{comment} : \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
-\end{alltt}
-\caption{The parser rule \antlrparserrule{content} with other rules that determine the relevant fragment of the syntax hierarchy}
-\label{fig:CONTENTHIERARCHY}
-\end{figure*}
+Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar expands ultimately into the nodes highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.
 
@@
 We call them \emph{end nodes}.
-Here we must emphasize that end nodes do not correspond to terminal symbols from grammars, because they may represent AST nodes with children (i.e., syntax elements that can be further expanded using grammar rules).
+Our algorithm that can recognize intermediary layers is, too, described in the full version~\cite{ref:TRFULL}.
 
-\begin{figure*}[ht]
-\begin{framed}
-\begin{alltt}
-\small
- 1 FindAllPathsToEndNodes(\textit{R}):
- 2   \textit{CurPath} = empty list of rules and nodes
- 3   \textbf{return} FindPaths(\textit{R}, \textit{CurPath})
- 4
- 5 FindPaths(\textit{R}, \textit{CurPath}):
- 6  \textit{Paths} = empty list
- 7  \textbf{for each} alternative \textit{A} \textbf{in} rule \textit{R}:
- 8    \textit{NewCurPath} = Clone(\textit{CurPath})
- 9    \textbf{if} \textit{A} contains only a single element \textit{E}:
-10      \textit{NewCurPath}.Add(\(N\sb{E}\)) \textbf{where} \(N\sb{E}\) is the node representing \textit{E}
-11      \textit{P} = FindPaths(\(R\sb{E}\), \textit{NewCurPath}) \textbf{where} \(R\sb{E}\) is the rule that expands \textit{E}
-12      \textit{Paths} = Merge(\textit{Paths}, \textit{P})
-13    \textbf{else}:
-14      \textit{NewCurPath}.Add(\textit{R})
-15      \textit{NewCurPath}.Add(\(N\sb{A}\)) \textbf{where} \(N\sb{A}\) is the node representing \textit{A}
-16      \textit{Paths}.Add(\textit{NewCurPath})
-17  \textbf{return} \textit{Paths}
-\end{alltt}
-\end{framed}
-\caption{Algorithm to find all paths to end nodes for a parser rule}
-\label{fig:SHORTCUTALG}
-\end{figure*}
-
-Figure~\ref{fig:SHORTCUTALG} shows the algorithm that finds all paths to some end node from a given parser rule $R$.
-The algorithm is based on recursive traversal of the parser tree.
-At each level of recursion, it gathers all paths that lead from the current parser rule $R$ to some end node through its alternatives (line~7).
-Two cases may occur:
-\begin{itemize}
-	\item When a particular alternative $A$ of the rule $R$ contains only a single element $E$ that is a reference to another rule (line 9), the alternative $A$ is an intermediary layer that can be hidden from the user of the MPS language.
-		  A run of the algorithm continues, at line 11, by recursively processing alternatives of the rule corresponding to $E$, which lies at the next level of the parser tree.
-	\item Otherwise, an end node of a derivation chain was found and the recursion stops (lines 13-16).
-\end{itemize}
-By appending the node corresponding to the current alternative (lines 10 and 15) and to the rule that leads to the alternative (line 14), the algorithm creates a full path that contains the target end node as the last element of the chain.
-
-We use the name \emph{shortcut approach} for this algorithm, because the paths collected by the algorithm provide shortcuts from the given rule to end nodes, by the virtue of hiding all intermediary layers.
-For example, the result of this algorithm for the \antlrparserrule{content} rule (Figure~\ref{fig:CONTENTHIERARCHY}) is the list shown in Figure~\ref{fig:CONTENTEXPAND}.
+We use the name \emph{shortcut approach} for this algorithm, because it creates shortcuts from the given rule to end nodes, by the virtue of hiding all intermediary layers.
 
@@
 As we said in Section~\ref{sect:MPS}, MPS uses a cellular system that enables the language developer to arrange the children and properties of an AST node in a table-like manner.
-Cells of different types are supported by MPS --- for storing property values, references to child nodes, and keywords (string literals), and also cells that influence layout (e.g., indentation).
 The specific goal of this phase is to create the Editor aspect for each language concept, such that all its attributes --- name, properties, children --- are projected using the respective cell types.
@@
 The main problem that we had to address is the absence of information about the code layout and whitespace in ANTLR grammars.
-Rules forming the grammar only tell what the syntax tree looks like and how the program text is decomposed into AST nodes.
 Here we present a solution that is only partially automated for reasons explained below.
@@
 A typical user of the \textsc{Ingrid} method will be able to use the projectional editor in MPS quite efficiently.
-We actually experimented with several heuristics to derive a useful code layout automatically, but all of them produced rather suboptimal results --- it is quite hard to automatically identify language concepts for which the default layout should be refactored, and the user would have to adjust the layout by hand anyway.
-More details are provided in Section~\ref{sect:EVAL}, including our experience with several mainstream programming languages.
-
-We have found that the combination of two steps, (1) automated placing of cells into a single row and (2) subsequent manual adjustment of the layout, is a very fast and efficient way of creating a nice code layout.
-Nevertheless, we plan to work on a more automated approach to the definition of Editor aspects in the future, based on machine learning.
-The key idea would be to derive the code layout from a set of valid input source files.
-%We believe that the user would have to adjust the editor manually a little bit even in this case, because the results of a machine learning-based approach would not be perfect.
+We actually experimented with several heuristics to derive a useful code layout automatically, but all of them produced rather suboptimal results.
 

### 2018-05-09 chapters/appendix.tex
shortening

--- 2018-05-07 chapters/background.tex
@@
-\section{Background}
-\label{sect:BACKGROUND}
+\appendix
+\section{Examples of MPS Language Aspects}
+\label{sect:APX}
 
-In this section, we introduce basic features of the MPS platform, and then we briefly discuss existing projects that aim to add support for general purpose languages into MPS.
-
-\subsection{JetBrains MPS}
-\label{sect:MPS}
-
-MPS is a language workbench --- an IDE that allows developers to create new languages and use them to write code.
-The code can be transformed into a target language, typically a GPPL such as Java or C, and compiled into executables.
-
-When using the MPS projectional editor, the developer creates programs by assembling the AST out of predefined building blocks of selected languages.
-The definition of a language in MPS dictates where in the AST certain elements can be placed and how they can be nested inside other elements.
-
-Code of any program in MPS is built from nodes, which represent instances of concepts from the languages that the program is written in.
-In MPS, a \emph{concept} is a language element, i.e. a building block of a language definition.
-
-One of the key advantages of projectional editing stems from the separation of abstract and concrete syntax.
-While AST provides a complete and precise representation of the code, the way it is displayed on the screen and the way the user edits it are unconstrained.
-The editor can take any visual form and shape (i.e., not just textual).
-The language author can define multiple alternative visualizations and let the developer choose one that fits best the task at hand.
-
-The definition of an MPS concept (language element) consists of several aspects, where each aspect codifies a different part of the AST nodes' behavior.
-The essential aspects are the following: \emph{Structure}, \emph{Editor} and \emph{TextGen}.
-Structure represents the abstract syntax (types and hierarchy of AST nodes), Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
-If programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
-Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we neglect other aspects such as type constraints.
-
-Languages are built from the concepts using techniques known from object-oriented programming --- containment, inheritance, and so on.
-Therefore, the definition of a language in MPS has an object-oriented and hierarchical nature.
-
-\paragraph{Structure.}
-The fundamental aspect of any MPS language is Structure.
-It specifies core attributes of a language concept such as the name, inheritance relationships, child concepts (their types and cardinalities), and references to other AST nodes.
-Figure~\ref{fig:if_statement_structure} shows the most important fragments of the Structure aspect for the \code{if-then-else} statement.
-
-Structure also restricts the type of AST nodes that can appear at a particular place in the tree.
-For example, one can restrict the condition in \code{if-then-else} to be a boolean expression, and the \code{then}-block to be a list of statements.
+The first part of this appendix contains examples of core aspects for the \code{if-then-else} statement --- Structure (Figure~\ref{fig:if_statement_structure}), Editor (Figure~\ref{fig:if_editor_definition}), and TextGen (Figure~\ref{fig:if_statement_textgen}).
 
@@
 \end{figure}
-
-\paragraph{Editor.}
-The Editor aspect is where the language designer specifies what the projectional representation of a code fragment (an AST) looks like on the screen and how the user interacts with the code.
-JetBrains have developed a cellular system that enables placing properties and children of a node (concept) into different cells.
-Individual cells of the editor can be styled using a language similar to CSS.
-Supported visual characteristics include color and indentation.
 
@@
 
-\paragraph{BaseLanguage.}
-Another very important feature of MPS is \emph{BaseLanguage}~\cite{ref:BASELANG}, a clone of Java implemented using the MPS constructs.
-BaseLanguage is syntactically almost identical to Java, but it is edited in a projectional editor, just like all the languages in MPS.
-Custom DSLs meant to be generated into Java choose BaseLanguage as their generation target.
-
-\paragraph{TextGen.}
-The TextGen aspect specifies how a given AST node will be translated into plain text representation.
-It is typically needed only for the bottom-line base languages.
-DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since programs in such languages are rarely converted to text directly.
-The TextGen definition follows a straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes.
+TextGen aspect for each AST node (concept of the language) has to be defined using the BaseLanguage.
+The definition follows a straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes.
 MPS calls the corresponding method on the root AST nodes of the given program.
+Figure~\ref{fig:if_statement_textgen} shows an example for the \code{if-else} statement.
 
@@
 
-TextGen aspect for each AST node (concept of the language) has to be defined using the BaseLanguage.
-Again, we include an example for the \code{if-else} statement (Figure~\ref{fig:if_statement_textgen}).
+Figure~\ref{fig:EDITORADJUST} shows the Editor aspect for the \mpsconcept{Element{\_}1} concept --- the basic, automatically generated, layout in the upper left corner and a fully customized layout in the bottom right part.
 
-\subsection{Related Projects with Similar Goals}
-\label{sect:RELATED}
+\begin{figure*}[ht]
+	\centering
+	\includegraphics[scale=0.55]{./images/editor_adjustment.png}
+	\caption{Editor aspect of the \mpsconcept{Element{\_}1} concept}
+	\label{fig:EDITORADJUST}
+\end{figure*}
 
-We are aware just of a few attempts to create ports of general-purpose languages into IDE tools based on projectional editing.
-All of them (described below) were conducted manually.
+Figure~\ref{fig:TEXTGENFINAL} contains the complete, automatically generated, TextGen aspect for the \mpsconcept{Element{\_}1} concept.
+Such code may be adjusted by hand very easily in order to produce nicely indented XML documents.
+We only need to wrap the \mpsconcept{Content{\_}2} child node with indentation and change the sequence separator to a new line character.
+A fragment of the resulting adjusted code is in Figure~\ref{fig:TEXTGENADJUSTED}.
 
-The BaseLanguage~\cite{ref:BASELANG} is an almost full port of Java, extended with MPS-specific features.
-The C language has also been manually tailored for MPS within the mbeddr project~\cite{ref:MBEDDR}.
+\begin{figure}[ht]
+\vspace{-2mm}
+\begin{alltt}
+\small
+\mpstgkeyword{text gen component for concept} \mpstgtarget{Element{\_}1} \{
+  \mpstgparam{(context, buffer, node)->void} \{
+    \mpstgaction{append} \{\mpstgliteral{<}\};
+    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{1}.\textbf{isNotEmpty}) \{
+      \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{1}\};
+    \}
+    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Attribute}}{\_}\textit{1}.\textbf{size} > 0) \{
+      \mpstgaction{append} \{\ \};
+      \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Attribute}}{\_}\textit{1} with  \};
+    \}
+    \mpstgaction{append} \{\mpstgliteral{>}\};
+    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2}.\textbf{size} > 0) \{
+      \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2} with  \};
+    \}
+    \mpstgaction{append} \{\mpstgliteral{</}\};
+    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{2}.\textbf{isNotEmpty}) \{
+      \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{2}\};
+    \}
+    \mpstgaction{append} \{\mpstgliteral{>}\};
+  \}
+\}
+\end{alltt}
+\caption{Full TextGen aspect for the \mpsconcept{Element{\_}1} concept}
+\label{fig:TEXTGENFINAL}
+\vspace{-1mm}
+\end{figure}
 
-The following three projects --- PE4MPS, ANTLR{\_}MPS, and mps-metabnf --- aim to provide certain support for GPPLs within the MPS platform.
-We describe their main features and limitations in this section.
-
-\paragraph{PE4MPS~\cite{ref:PE4MPS}.}
-This project addresses the lack of information about code layout in grammars by using so called PE grammars~\cite{ref:PE}, which is a new notation proposed by the same author.
-PE stands for projectional editing.
-
-PE grammars extend the syntax of the ANTLR v4 notation by constructs that provide information about the possible layout of AST nodes.
-The PE4MPS tool creates an MPS language in a single atomic step.
-Information about the code layout, extracted from the PE file, is used when generating the projectional editor for every AST node.
-
-The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
-Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
-
-\paragraph{ANTLR{\_}MPS~\cite{ref:ANTLR2MPS}.}
-This project also uses the ANTLR v4 grammar notation, but otherwise works quite differently from PE4MPS.
-Its author created an MPS language that captures the syntax of ANTLR v4 notation using MPS concepts.
-Given the textual grammar of some language as input, the grammar is imported automatically into MPS, taking the form of the ANTLR v4 MPS language.
-The next step would be to create a new MPS language from the grammar definition in MPS, but it is not implemented yet.
-
-However, the project has several important limitations.
-The tool does not generate any parent-child relationships in the Structure aspect, and it completely neglects the Editor and TextGen aspects.
-
-\paragraph{mps-metabnf~\cite{ref:MPSMETABNF}.}
-This tool builds upon similar ideas as the ANTLR{\_}MPS project.
-Every imported grammar is represented using an ANTLR v4 MPS language.
-A user is then able to adjust the grammar as he wishes, specifying the code layout in the projectional editor.
-In the last step, the target MPS language is derived from the adjusted grammar definition.
-
-The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLR v4 MPS language into the definition of the actual target MPS language.
+\begin{figure}[ht]
+\begin{alltt}
+\small
+\mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2}.\textbf{size} > 0) \{
+  \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n;}}
+  \mpstgaction{indent buffer;}
+  \mpstgkeyword{with indent} \{
+    \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2} with  \};
+  \}
+  \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n;}}
+\} 
+\end{alltt}
+\vspace{-1mm}
+\caption{Fragment of the TextGen aspect of \mpsconcept{Element{\_}1} with adjusted indentation}
+\label{fig:TEXTGENADJUSTED}
+\vspace{-2mm}
+\end{figure}

### 2018-05-09 chapters/background.tex
shortening

@@
 
-\paragraph{Structure.}
+\noindent\textbf{Structure.}
 The fundamental aspect of any MPS language is Structure.
 It specifies core attributes of a language concept such as the name, inheritance relationships, child concepts (their types and cardinalities), and references to other AST nodes.
-Figure~\ref{fig:if_statement_structure} shows the most important fragments of the Structure aspect for the \code{if-then-else} statement.
+Figure~\ref{fig:if_statement_structure} in Appendix~\ref{sect:APX} shows the most important fragments of the Structure aspect for the \code{if-then-else} statement.
 
@@
 
-\begin{figure}[ht]
-\vspace{-2mm}
-\centering
-\begin{alltt}
-\small
-\mpsstkeyword{concept} IfStatement \mpsstkeyword{extends} Statement
-        \mpsstkeyword{implements} IContainsStatementList
-
-  \mpsstkeyword{instance can be root:} false
-  \mpsstkeyword{alias:} \mpsstalias{if}
-
-  \mpsstkeyword{children:}
-  \mpsstproperty{condition}        : Expression[\mpsstcardinality{1}]
-  \mpsstproperty{ifFalseStatement} : Statement[\mpsstcardinality{0..1}]
-  \mpsstproperty{ifTrue}           : StatementList[\mpsstcardinality{1}]
-  \mpsstproperty{elsifClauses}     : ElsifClause[\mpsstcardinality{0..n}]
-\end{alltt}
-\caption{Structure aspect of the \code{if-then-else} statement}
-\label{fig:if_statement_structure}
-\vspace{-4mm}
-\end{figure}
-
-\paragraph{Editor.}
+\noindent\textbf{Editor.}
 The Editor aspect is where the language designer specifies what the projectional representation of a code fragment (an AST) looks like on the screen and how the user interacts with the code.
@@
 
-Figure~\ref{fig:if_editor_definition} shows an example of what the definition of the Editor aspect for the \code{if-then-else} statement might look like.
-While here we indicate positions of cell borders on each line by spaces (just for illustration), MPS GUI actually uses a graphically much more appealing way of displaying the Editor aspects, which involves vertical and horizontal lines of different colors and also background colors other than white for some cells.
-The symbols \verb|[-| and \verb|-]| represent layout information, which specifies mutual positioning of the contained cells (vertical, horizontal, indentation).
-
-\begin{figure}[ht]
-\vspace{-1mm}
-\centering
-\begin{alltt}
-\small
-\mpsedannotation{<default>} \mpsedkeyword{editor for concept} \mpsedtarget{IfStatement}
-  \mpsedkeyword{node cell layout:}
-    [-
-      \mpsedkeyword{if} \textbf{(} \% conditions \% \textbf{)} [-
-      \textbf{\{}
-      [- \% ifTrue % -]
-      \textbf{\}}
-      -]
-      ?[- \mpsedkeyword{else} \% ifFalseStatement \% -]
-    -]
-\end{alltt}
-\vspace{-1mm}
-\caption{Editor aspect for the \code{if-else} statement}
-\label{fig:if_editor_definition}
-\vspace{-2mm}
-\end{figure}
-
-\paragraph{BaseLanguage.}
+\noindent\textbf{BaseLanguage.}
 Another very important feature of MPS is \emph{BaseLanguage}~\cite{ref:BASELANG}, a clone of Java implemented using the MPS constructs.
@@
 
-\paragraph{TextGen.}
+\noindent\textbf{TextGen.}
 The TextGen aspect specifies how a given AST node will be translated into plain text representation.
@@
 DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since programs in such languages are rarely converted to text directly.
-The TextGen definition follows a straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes.
-MPS calls the corresponding method on the root AST nodes of the given program.
 
-\begin{figure}[ht]
-\vspace{-2mm}
-\centering
-\begin{alltt}
-\small
-\mpstgkeyword{text gen component for concept} \mpstgtarget{IfStatement} \{
-  \mpstgparam{(context, buffer, node)->void} \{
-    \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n;}}
-    \mpstgaction{indent buffer;}
-    \mpstgaction{append} \{\mpstgliteral{if (}\} \$\{\mpstgparam{node}.\mpstgnodeprop{condition}\} \{\mpstgliteral{) \{}\};
-    \mpstgkeyword{with indent} \{
-      \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{ifTrue}\};
-    \}
-    \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n}} \{\mpstgliteral{\}}\} \$list\{\mpstgparam{node}.\mpstgnodeprop{elsifClauses}\};
-    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{ifFalseStatement}.\textbf{isNotNull}) \{
-      \mpstgaction{append} \{ \mpstgliteral{else}\} \$\{\mpstgparam{node}.\mpstgnodeprop{isFalseStatement}\};
-    \}
-  \}
-\}
-\end{alltt}
-\vspace{-1mm}
-\caption{TextGen aspect definition for the \code{if-else} statement}
-\label{fig:if_statement_textgen}
-\vspace{-2mm}
-\end{figure}
-
-TextGen aspect for each AST node (concept of the language) has to be defined using the BaseLanguage.
-Again, we include an example for the \code{if-else} statement (Figure~\ref{fig:if_statement_textgen}).
+We provide examples of the Editor and TextGen aspects for the \code{if-else} statement in the Appendix~\ref{sect:APX}.
 
@@
 
-\paragraph{PE4MPS~\cite{ref:PE4MPS}.}
+\noindent\textbf{PE4MPS~\cite{ref:PE4MPS}.}
 This project addresses the lack of information about code layout in grammars by using so called PE grammars~\cite{ref:PE}, which is a new notation proposed by the same author.
@@
 
-\paragraph{ANTLR{\_}MPS~\cite{ref:ANTLR2MPS}.}
+\noindent\textbf{ANTLR{\_}MPS~\cite{ref:ANTLR2MPS}.}
 This project also uses the ANTLR v4 grammar notation, but otherwise works quite differently from PE4MPS.
@@
 
-\paragraph{mps-metabnf~\cite{ref:MPSMETABNF}.}
+\noindent\textbf{mps-metabnf~\cite{ref:MPSMETABNF}.}
 This tool builds upon similar ideas as the ANTLR{\_}MPS project.

### 2018-05-09 chapters/language_definition.tex
shortening

@@
-\section{Creating Language Definition in MPS}
+\section{Creating Language in MPS}
 \label{sect:LANGDEF}
@@
 
-\paragraph{The Shortcut Approach.}
+\noindent\textbf{The Shortcut Approach.}
 The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree, and consider just the nodes to be directly offered to the user through the auto-completion menu.
@@
 Our solution that we implemented in the current version of the \textsc{Ingrid} method is to create all the cells and place them in a single row.
-The resulting basic layout is illustrated on the example of the \mpsconcept{Element{\_}1} concept in the upper left corner of Figure~\ref{fig:EDITORADJUST}.
 Further adjustments of the layout, such as indentation and line breaks, can be done very efficiently by the user in the MPS IDE.
-The bottom right part of Figure~\ref{fig:EDITORADJUST} shows a fully customized layout for the \mpsconcept{Element{\_}1} concept.
-
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[scale=0.55]{./images/editor_adjustment.png}
-	\caption{Editor aspect of the \mpsconcept{Element{\_}1} concept}
-	\label{fig:EDITORADJUST}
-\end{figure*}
+The resulting layout is illustrated on the example of the \mpsconcept{Element{\_}1} concept in Figure~\ref{fig:EDITORADJUST} (Appendix~\ref{sect:APX}).
 
@@
 The purpose of the last phase of the MPS language construction is to generate the TextGen aspect for each concept.
-We use the \mpsconcept{Element{\_}1} concept again for illustration.
-Figure~\ref{fig:TEXTGENBASIC} shows a fragment of BaseLanguage code that can be used as its TextGen aspect.
+Figure~\ref{fig:TEXTGENBASIC} shows a fragment of BaseLanguage code that can be used as the TextGen aspect for \mpsconcept{Element{\_}1}.
 The code appends all literals, properties, and children of \mpsconcept{Element{\_}1} to the output buffer.
-%, and does that in the same order as they appear in the corresponding grammar rule.
 
@@
   \mpstgparam{(context, buffer, node)->void} \{
-    \mpstgaction{append} \{\mpstgliteral{<}\};
-    \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{1}\};
-    \mpstgaction{append} \{\ \};
-    \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Attribute}}{\_}\textit{1}\};
-    \mpstgaction{append} \{\mpstgliteral{>}\};
+    \mpstgaction{append} \{\mpstgliteral{<}\}; \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{1}\}; \mpstgaction{append} \{\ \};
+    \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Attribute}}{\_}\textit{1}\}; \mpstgaction{append} \{\mpstgliteral{>}\};
     \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2}\};
-    \mpstgaction{append} \{\mpstgliteral{</}\};
-    \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{2}\};
-    \mpstgaction{append} \{\mpstgliteral{>}\};
+    \mpstgaction{append} \{\mpstgliteral{</}\}; \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{2}\}; \mpstgaction{append} \{\mpstgliteral{>}\};
   \}
@@
 For example, in the case of the concept representing an XML element, there must be a space between the element's name and the first attribute, while it is not required between the opening bracket \antlrliteral{\textless} and the name.
-%(Figure~\ref{fig:TEXTGENBASIC}).
 
@@
 The procedure creates the TextGen aspect for a given language concept in two steps.
-First, it generates a basic variant that inserts spaces in between every two tokens of the textual representation of the concept.
+First, it generates a basic variant that inserts spaces (or line breaks) in between every two tokens of the textual representation of the concept.
 In the second step, spaces are eliminated from places where they are not really needed.
 The main criterion is whether the generated textual output can be accepted by a parser of the original language (using the ANTLR grammar).
-We discuss all the cases here:
-\begin{itemize}
-	\item When there is a non-alphabetical literal that is used as a token in the grammar, and that might get recognized by the parser without the need for whitespace separators around it, then we can omit the spaces.
-An example of such literal is '\textless' in SimpleXML.
-	\item In the case of an arbitrary string, the whitespace may be omitted when the adjacent literal ends, respectively begins, with a non-alphabetical character.
-		This applies especially to the values of properties defined in Structure aspects, such as the name of an XML element.
-		Based on this heuristic, redundant spaces inside of quotes will be eliminated, as well as spaces next to semicolons and around brackets.
-		On the other hand, it will separate language keywords from other literals by preserving the space characters between them.
-	\item Spaces can be safely omitted also when optional child nodes are not present, and in the case of empty lists of child nodes, so that whitespace does not accumulate.
-\end{itemize}
-A space must be always inserted when two child nodes are next to each other.
-Elements from a sequence of child nodes are separated with spaces or line breaks.
-
-\begin{figure}[ht]
-\vspace{-2mm}
-\begin{alltt}
-\small
-\mpstgkeyword{text gen component for concept} \mpstgtarget{Element{\_}1} \{
-  \mpstgparam{(context, buffer, node)->void} \{
-    \mpstgaction{append} \{\mpstgliteral{<}\};
-    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{1}.\textbf{isNotEmpty}) \{
-      \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{1}\};
-    \}
-    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Attribute}}{\_}\textit{1}.\textbf{size} > 0) \{
-      \mpstgaction{append} \{\ \};
-      \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Attribute}}{\_}\textit{1} with  \};
-    \}
-    \mpstgaction{append} \{\mpstgliteral{>}\};
-    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2}.\textbf{size} > 0) \{
-      \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2} with  \};
-    \}
-    \mpstgaction{append} \{\mpstgliteral{</}\};
-    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{2}.\textbf{isNotEmpty}) \{
-      \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{\textit{Name}}{\_}\textit{2}\};
-    \}
-    \mpstgaction{append} \{\mpstgliteral{>}\};
-  \}
-\}
-\end{alltt}
-\caption{Full TextGen aspect for the \mpsconcept{Element{\_}1} concept}
-\label{fig:TEXTGENFINAL}
-\vspace{-1mm}
-\end{figure}
-
-\begin{figure}[ht]
-\begin{alltt}
-\small
-\mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2}.\textbf{size} > 0) \{
-  \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n;}}
-  \mpstgaction{indent buffer;}
-  \mpstgkeyword{with indent} \{
-    \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2} with  \};
-  \}
-  \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n;}}
-\} 
-\end{alltt}
-\vspace{-1mm}
-\caption{Fragment of the TextGen aspect of \mpsconcept{Element{\_}1} with adjusted indentation}
-\label{fig:TEXTGENADJUSTED}
-\vspace{-2mm}
-\end{figure}
-
-Figure~\ref{fig:TEXTGENFINAL} contains the complete, automatically generated, TextGen aspect for the \mpsconcept{Element{\_}1} concept.
-%, which represents the full SimpleXML element.
-Such code may be adjusted by hand very easily in order to produce nicely indented XML documents.
-We only need to wrap the \mpsconcept{Content{\_}2} child node with indentation and change the sequence separator to a new line character.
-A fragment of the resulting adjusted code is in Figure~\ref{fig:TEXTGENADJUSTED}.
+Again, the specific cases are discussed in the full version~\cite{ref:TRFULL}.
+We show the complete TextGen aspect for \mpsconcept{Element{\_}1} in the appendix.
 
@@
 
-During our work on the \textsc{Ingrid} method, we have also observed that practical usability of a resulting MPS language depends on the specific manner in which the input ANTLR grammar is defined.
-We discuss several issues and our workarounds in this section.
+During our work on the \textsc{Ingrid} method, we have observed that practical usability of a resulting MPS language depends on the specific manner in which the input ANTLR grammar is defined.
 
-\paragraph{Adjusting grammars.}
+\noindent\textbf{Adjusting grammars.}
 In some cases, a small adjustment of the input grammar before the run of \textsc{Ingrid} might yield a better and more useful MPS language.
-Here we illustrate this on two simple examples.
+Here we illustrate this on a simple example.
 
-For the first example, we use the definition of an XML attribute in Figure~\ref{fig:XMLATTRIB}, which is taken from the original XML grammar.
+Figure~\ref{fig:XMLATTRIB} shows the definition of an XML attribute, which is taken from the original XML grammar.
 The lexer rule \antlrlexerrule{STRING} actually says that quotes make a part of the attribute's value.
@@
   \antlrparserrule{attr} : \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
-  \antlrlexerrule{STRING} : \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
-         | \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'} ;
+  \antlrlexerrule{STRING} : \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"} | \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'} ;
 \end{alltt}
@@
 
+Additional examples are provided in~\cite{ref:TRFULL}.
+
 \begin{figure}[ht]
@@
 \small
-  \antlrparserrule{attr} : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
-         | \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'} ;
+  \antlrparserrule{attr} : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"} | \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'} ;
   \antlrlexerrule{TEXT1} : \antlrregex{~["]*} ;
@@
 
-The second example relates to a fragment of the JavaScript language, also known as ECMAScript.
-%\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}.
-Every statement in JavaScript needs to be terminated by a semicolon, newline, end of the file, or end of the block --- see the Figure~\ref{fig:JAVASCRIPTSTMT}, which contains the corresponding fragment of the ECMAScript grammar.
-Since there are multiple options, a user would have to select one of them for each statement in the editor.
-Technically, every language concept representing a statement would contain one child node of the interface type \mpsinterface{IEos}, which has to be assigned a proper object (i.e., an AST node corresponding to one of the options listed above).
-
-\begin{figure}[ht]
-\vspace{-1mm}
-\begin{alltt}
-\small
-\antlrparserrule{eos} : \antlrlexerrule{SemiColon} | \antlrlexerrule{EOF} | {lineBreakAhead()}? 
-    | {{\_}input.LT(1).type() == \antlrlexerrule{CloseBrace}}? ;
-\textcolor{gray}{// example reference to the eos rule}
-\antlrparserrule{breakStmt} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos} ;
-\end{alltt}
-\caption{Statements in JavaScript}
-\label{fig:JAVASCRIPTSTMT}
-\vspace{-1mm}
-\end{figure}
-
-Since MPS can differentiate between statements on the AST level, there is no need for an explicit separator.
-For example, we can put each statement on a distinct line, as is usual for JavaScript code, and keep just the semicolon as a fixed literal in the projectional editor.
-The \antlrparserrule{eos} rule would have to be changed to this form: \antlrparserrule{eos} : \antlrliteral{;}.
-A small adjustment of the grammar, like this one, is a very quick solution and makes the generated language more usable in MPS.
-On the other hand, this adjustment of the original ANTLR grammar, performed just for the purpose of creating the MPS language, changes the grammar in such a way that it does not precisely describe the standard JavaScript language.
-To ensure compatibility with JavaScript, the TextGen aspect in the MPS language should put the semicolon after each statement in the plain-text representation of a program code.
-
-\paragraph{Breaking original grammars and parsers.}
+\noindent\textbf{Breaking original grammars and parsers.}
 We also want to point out a general problem with grammar adjustments, which all potential users of the \textsc{Ingrid} method should be aware of.

### 2018-05-09 chapters/evaluation.tex
minor changes
shortening

@@
 
-However, first we must emphasize that MPS languages automatically produced by the \textsc{Ingrid} method, are not ready-to-use full-fledged MPS counterparts of the original input languages.
+However, first we must emphasize that MPS languages automatically produced by \textsc{Ingrid}, are not ready-to-use full-fledged MPS counterparts of the original input languages.
 The structure of a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects have to be manually tweaked (e.g., the Editor) or defined from scratch by the end user.
@@
 
-For each of the three languages (JSON, JavaScript, C\#), we provide (1) its MPS definition in the form that incorporates all tweaks performed manually by authors of this paper and (2) the adjusted ANTLR v4 grammar used as input for \textsc{Ingrid}.
-%, both in the repository that contains also our implementation.
-
-\paragraph{JSON.}
+\noindent\textbf{JSON.}
 The least amount of manual adjusting after the import into MPS was needed in case of the JSON language~\cite{ref:JSON}, because it is the simplest language from all that we used for our experiments.
@@
 
-\paragraph{JavaScript.}
-In the case of JavaScript, which is an example of a widely-used complex general purpose programming language, automated generating of the language definition in MPS from the ANTLR grammar and subsequent manual adjusting was done in less than one hour.
-%\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}
+\noindent\textbf{JavaScript.}
+In the case of JavaScript, automated generating of the language definition in MPS from the ANTLR grammar and subsequent manual adjusting was done in less than one hour.
+The current version of \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful --- for example, the language designer has to define actions that improve editing experience by automatically transforming the AST in the background.
 
-We are aware of other projects that aim to create a manual port of JavaScript into MPS.
-For example, there is ECMAScript4MPS~\cite{ref:ECMA4MPS} developed by the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section~\ref{sect:RELATED}.
-
-The main advantage of \textsc{Ingrid} over ECMAScript4MPS is that, despite its current limitations, \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful --- for example, the language designer has to define actions that improve editing experience by automatically transforming the AST in the background.
-\textsc{Ingrid} achieved a very good result, when compared to PE4MPS, especially in the case of language structure, concept aliases, and support for auto-completion.
-The Structure aspect generated by \textsc{Ingrid} is very similar to that of the ECMAScript4MPS project, which was created manually over a large number of hours.
-
-\paragraph{C\#.}
-Another very complex programming language that we used for experiments is C\#~\cite{ref:CSHARP}.
-Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces at least a reasonable structure (hierarchy of concepts) that can be practically used in MPS just with minor alterations.
+\noindent\textbf{C\#.}
+Before we could run \textsc{Ingrid} on C\#~\cite{ref:CSHARP}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces at least a reasonable structure (hierarchy of concepts) that can be practically used in MPS just with minor alterations.
 Necessary adjustments include removal of intermediary rules such as \antlrparserrule{statement\_list} : \antlrparserrule{statement*}~\antlrliteral{;} by inlining the right-hand side.
@@
 The definition is quite large, involving more than 800 concepts.
-%, and therefore compilation (building) of the language in MPS takes a rather long time.
 
-\paragraph{Other languages.}
+\noindent\textbf{Other languages.}
 We also tried to apply \textsc{Ingrid} on few other languages, such as Python and Ruby.
 Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with \textsc{Ingrid}.
-For example, the structure and hierarchy of rules in the Python grammar are quite different from JavaScript or JSON, and consequently the language definition created by \textsc{Ingrid} is rather badly organized.
-We plan to address this limitation in the future and extend \textsc{Ingrid} such that very few modifications of the grammar would be necessary even for these languages.
 
-\paragraph{General observations.}
+\noindent\textbf{General observations.}
 The main overall benefit of the \textsc{Ingrid} method is partial automation.
@@
 On the other hand, our experiments with complex languages show that, in the case of the Editor and TextGen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, an order of magnitude less time than we initially expected.
-The first author spent between 20 and 60 minutes of work on each language, when adjusting the result of automated generation into a more useful and readable form.
 MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing designers to produce really useful languages very fast.
-%All of this justifies our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
 Despite that, we also tried to design some automated heuristics, but so far all yield rather mediocre results when compared to what human users can achieve efficiently instead.
+
+Regarding future changes to the ANTLR grammar of a subject language, it is certainly possible to directly update the corresponding MPS definition, but in some cases it may be easier to run again the whole import process and overwrite the previous definition with a new one.

### 2018-05-09 chapters/futurework.tex
shortening

--- 2017-06-16 chapters/conclusion.tex
@@
-\section{Conclusion and Future Work}
+\section{Future Work}
 
-Our contribution is the \textsc{Ingrid} method for constructing language definitions in MPS from grammars in the ANTLR notation.
-The method is only partially automated, but nevertheless it greatly reduces the amount of manual work that developers of MPS languages have to perform.
-Although we discussed all challenges, details of our solution and general observations just in the context of ANTLR grammars and JetBrains MPS, we believe they are more general --- in particular, relevant to everyone tackling a similar problem in the field of DSLs and language workbenches.
-
-In the future, we would like to add support for generating other aspects of MPS languages and increase the level of automation.
+In the future, we would like to add support for generating other aspects of MPS languages and to increase the level of automation.
 Our primary research goal is to investigate the usage of automated (machine) learning for the purpose of generating editors that support nice and readable code layout.

### 2018-05-09 chapters/introduction.tex
shortening

@@
 Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution.
-We also discuss our experience with application of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\# (Section~\ref{sect:EVAL}), and then we conclude.
+We also discuss our experience with application of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\# (Section~\ref{sect:EVAL}), and then we outline our plans for the future work.
 

### 2018-05-09 paper.tex
minor changes
shortening

@@
-\documentclass[sigplan,10pt,preprint]{acmart}
+\documentclass[sigconf]{acmart}
 
@@
 
-\setcounter{tocdepth}{3}
+% TODO
+% copyright
+%\setcopyright{acmlicensed}
+%\setcopyright{rightsretained}
 
-\settopmatter{printfolios=true}
+% TODO
+%\acmDOI{10.475/123_4}
+
+% TODO
+%\acmISBN{123-4567-24-567/08/06}
+
+% TODO
+%\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El Paso, Texas USA}
+%\acmYear{2018}
+%\copyrightyear{2018}
+
+% TODO
+%\acmArticle{4}
+%\acmPrice{15.00}
 
@@
 \newcommand{\mpstgnodeprop}[1]{\textcolor{darkgray}{\textbf{#1}}}
-
-% TODO
-%\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-%\acmYear{2017}
-%\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-%\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-
-% TODO
-%\setcopyright{acmlicensed}
 
@@
 \input{chapters/evaluation}
-\input{chapters/conclusion}
+\input{chapters/futurework}
 
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{ACM-Reference-Format}
 
@@
 
-\bibitem{ref:MPSBOOK} F. Campagne. The MPS Language Workbench, Volume I. CreateSpace Independent Publishing Platform, 2014.
+\bibitem{ref:MPSBOOK} Fabien Campagne. 2014. The MPS Language Workbench, Volume I. CreateSpace Independent Publishing Platform.
 
-\bibitem{ref:BISONFLEX} J. Levine. Flex \& Bison: Text Processing Tools. O'Reilly Media, 2009.
+\bibitem{ref:DHL80} Veronique Donzeau-Gouge, Gerard Huet, Bernard Lang, and Gilles Kahn.
+1980.
+Programming Environments Based on Structured Editors: The Mentor Experience.
+Research Report RR-0026, INRIA.
 
-\bibitem{ref:KV10} L.C.L. Kats and E. Visser. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. OOPSLA 2010, ACM.
+\bibitem{ref:BISONFLEX} John Levine. 2009. Flex \& Bison: Text Processing Tools. O'Reilly Media.
 
-\bibitem{ref:ANTLRBOOK} T. Parr. The Definitive ANTLR 4 Reference. Pragmatic Bookshelf, 2013.
+\bibitem{ref:KV10} Lennart C.L. Kats and Eelco Visser. 2010. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. In Proc. of OOPSLA 2010, ACM.
 
-\bibitem{ref:MBEDDR} M. Voelter, D. Ratiu, B. Schaetz, and B. Kolb. mbeddr: an Extensible C-based Programming Language and IDE for Embedded Systems. SPLASH 2012, ACM.
+\bibitem{ref:ANTLRBOOK} Terence Parr. 2013. The Definitive ANTLR 4 Reference. Pragmatic Bookshelf.
 
-\bibitem{ref:VSB14} M. Voelter, J. Siegmund, T. Berger, and B. Kolb. Towards User-Friendly Projectional Editors. SLE 2014, LNCS, 8706.
+\bibitem{ref:MBEDDR} Markus Voelter, Daniel Ratiu, Bernhard Schaetz, and Bernd Kolb.
+2012.
+mbeddr: an Extensible C-based Programming Language and IDE for Embedded Systems.
+In Proc.
+of SPLASH 2012, ACM.
 
-\bibitem{ref:VWK15} M. Voelter, J. Warmer, and B. Kolb. Projecting a Modular Future. IEEE Software, 32(5), 2015.
+\bibitem{ref:VSB14} Markus Voelter, Janet Siegmund, Thorsten Berger, and Bernd Kolb. 2014. Towards User-Friendly Projectional Editors. In Proc. of SLE 2014, LNCS, 8706.
+
+\bibitem{ref:VWK15} Markus Voelter, Jos Warmer, and Bernd Kolb. 2015. Projecting a Modular Future. IEEE Software, 32(5).
+
+\bibitem{ref:TRFULL} Premysl Vysoky, Pavel Parizek, and Vaclav Pech.
+2018.
+INGRID: Creating Languages in MPS from ANTLR Grammars (Full version).
+Tech.
+Report No.
+D3S-TR-2018-01, Dep.
+of Distributed and Dependable Systems, Charles University.
 
@@
 
-\bibitem{ref:ANTLR2MPS} F. Campagne. The ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR_MPS}
+\bibitem{ref:ANTLR2MPS} Fabien Campagne. The ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR_MPS}
 
@@
 
-\bibitem{ref:ECMA4MPS} M. Lombardo. The ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}
+\bibitem{ref:PE4MPS} Marco Lombardo. The PE4MPS project, \url{https://github.com/mar9000/pe4mps}
 
-\bibitem{ref:PE4MPS} M. Lombardo. The PE4MPS project, \url{https://github.com/mar9000/pe4mps}
-
-\bibitem{ref:PE} M. Lombardo. PE grammars, \url{https://github.com/mar9000/pe}
+\bibitem{ref:PE} Marco Lombardo. PE grammars, \url{https://github.com/mar9000/pe}
 
@@
 
+\input{chapters/appendix}
+
 \end{document}

### 2018-05-10 chapters/background.tex
further shortening

@@
 
-In this section, we introduce basic features of the MPS platform, and then we briefly discuss existing projects that aim to add support for general purpose languages into MPS.
+Here we introduce basic features of the JetBrains MPS platform, and then we briefly describe other projects that aim to add support for general purpose languages into MPS.
 
@@
 
-MPS is a language workbench --- an IDE that allows developers to create new languages and use them to write code.
-The code can be transformed into a target language, typically a GPPL such as Java or C, and compiled into executables.
-
 When using the MPS projectional editor, the developer creates programs by assembling the AST out of predefined building blocks of selected languages.
 The definition of a language in MPS dictates where in the AST certain elements can be placed and how they can be nested inside other elements.
-
-Code of any program in MPS is built from nodes, which represent instances of concepts from the languages that the program is written in.
+Code of any program is actually built from AST nodes, which represent instances of concepts from the languages that the program is written in.
 In MPS, a \emph{concept} is a language element, i.e. a building block of a language definition.
 
-One of the key advantages of projectional editing stems from the separation of abstract and concrete syntax.
+A key advantage of projectional editing stems from the separation of abstract and concrete syntax.
 While AST provides a complete and precise representation of the code, the way it is displayed on the screen and the way the user edits it are unconstrained.
@@
 Structure represents the abstract syntax (types and hierarchy of AST nodes), Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
-If programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
-Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we neglect other aspects such as type constraints.
+Other aspects, such as type constraints and model-to-model transformation rules, are not relevant for the contribution of this paper, and therefore we do not describe them here.
 
@@
 It specifies core attributes of a language concept such as the name, inheritance relationships, child concepts (their types and cardinalities), and references to other AST nodes.
+Structure also restricts the type of AST nodes that can appear at a particular place in the tree.
+For example, one can restrict the condition in the \code{if-then-else} statement to be a boolean expression, and the \code{then}-block to be a list of statements.
 Figure~\ref{fig:if_statement_structure} in Appendix~\ref{sect:APX} shows the most important fragments of the Structure aspect for the \code{if-then-else} statement.
-
-Structure also restricts the type of AST nodes that can appear at a particular place in the tree.
-For example, one can restrict the condition in \code{if-then-else} to be a boolean expression, and the \code{then}-block to be a list of statements.
 
@@
 \noindent\textbf{BaseLanguage.}
-Another very important feature of MPS is \emph{BaseLanguage}~\cite{ref:BASELANG}, a clone of Java implemented using the MPS constructs.
-BaseLanguage is syntactically almost identical to Java, but it is edited in a projectional editor, just like all the languages in MPS.
+Another important feature of MPS is \emph{BaseLanguage}~\cite{ref:BASELANG}, a clone of Java implemented using the MPS constructs.
+BaseLanguage is syntactically almost identical to Java, but it is edited in a projectional editor.
 Custom DSLs meant to be generated into Java choose BaseLanguage as their generation target.
@@
 \noindent\textbf{TextGen.}
-The TextGen aspect specifies how a given AST node will be translated into plain text representation.
+The TextGen aspect specifies how an AST node is translated into plain text representation.
 It is typically needed only for the bottom-line base languages.
-DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since programs in such languages are rarely converted to text directly.
-
 We provide examples of the Editor and TextGen aspects for the \code{if-else} statement in the Appendix~\ref{sect:APX}.
@@
 The following three projects --- PE4MPS, ANTLR{\_}MPS, and mps-metabnf --- aim to provide certain support for GPPLs within the MPS platform.
-We describe their main features and limitations in this section.
+We describe their main features and limitations.
 
 \noindent\textbf{PE4MPS~\cite{ref:PE4MPS}.}
-This project addresses the lack of information about code layout in grammars by using so called PE grammars~\cite{ref:PE}, which is a new notation proposed by the same author.
-PE stands for projectional editing.
-
+This project addresses the lack of information about code layout in grammars by using so called PE grammars~\cite{ref:PE}, which is a notation proposed by the same author.
 PE grammars extend the syntax of the ANTLR v4 notation by constructs that provide information about the possible layout of AST nodes.
-The PE4MPS tool creates an MPS language in a single atomic step.
 Information about the code layout, extracted from the PE file, is used when generating the projectional editor for every AST node.
@@
 \noindent\textbf{ANTLR{\_}MPS~\cite{ref:ANTLR2MPS}.}
-This project also uses the ANTLR v4 grammar notation, but otherwise works quite differently from PE4MPS.
-Its author created an MPS language that captures the syntax of ANTLR v4 notation using MPS concepts.
+The author of this project created an MPS language that captures the syntax of ANTLR v4 notation using MPS concepts.
 Given the textual grammar of some language as input, the grammar is imported automatically into MPS, taking the form of the ANTLR v4 MPS language.
 The next step would be to create a new MPS language from the grammar definition in MPS, but it is not implemented yet.
-
-However, the project has several important limitations.
-The tool does not generate any parent-child relationships in the Structure aspect, and it completely neglects the Editor and TextGen aspects.
+Another important limitation of this project is the completely missing support for the Editor and TextGen aspects, and it also does not generate any parent-child relationships in the Structure aspect.
 
@@
 Every imported grammar is represented using an ANTLR v4 MPS language.
-A user is then able to adjust the grammar as he wishes, specifying the code layout in the projectional editor.
-In the last step, the target MPS language is derived from the adjusted grammar definition.
+A user is then able to adjust the grammar, specifying the code layout in the projectional editor.
+In the last step, the target MPS language is derived from the grammar.
 
-The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLR v4 MPS language into the definition of the actual target MPS language.
+The main limitation of this approach is the need to perform the last step manually, because MPS does not provide any support for automated transformation of a grammar captured by the ANTLR v4 MPS language into the definition of the actual target MPS language.

### 2018-05-10 chapters/introduction.tex
further shortening

@@
 
-An alternative approach is projectional (or structural) editing~\cite{ref:VWK15,ref:VSB14,ref:DHL80}, where a developer directly manipulates the AST representation of the source code instead of plain text.
+An alternative approach is projectional editing~\cite{ref:VWK15,ref:VSB14,ref:DHL80}, where a developer directly manipulates the AST representation of the source code.
 Recently, projectional editing has become widely used in the area of language workbenches --- IDE-like tools that enable the developers to manipulate language definitions.
 Popular examples are JetBrains MPS~\cite{ref:MPS,ref:MPSBOOK} and Spoofax~\cite{ref:KV10}.
-In particular, JetBrains MPS is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
+In particular, JetBrains MPS is an open-source language workbench that focuses on domain specific languages (DSL).
 MPS provides the whole IDE infrastructure that enables developers to design custom languages, use them to write program code, and compile the code into executables.
@@
 	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together in a single program or one can extend another.
-	\item The languages may support diverse contextual or non-parseable notations with complex layout, such as tables, diagrams, and mathematical expressions.
-	\item Since the projection is detached from the physical representation of code (AST), authors of languages can define multiple notations and allow the developers to switch between them on the screen.
+	\item The languages may support diverse notations with complex layout, such as tables and mathematical expressions.
+	\item Since the projection is detached from the physical representation of code (AST), authors of languages can define multiple notations and allow the developers to switch between them.
 \end{itemize}
@@
 Nevertheless, the usage of projectional editors introduces some new problems.
-Before a language can be used inside an editor, it has to be defined through the specific infrastructure, so that the IDE can understand the language and work with it.
-More precisely, an author of a language has to create:
+Languages have to be defined through the specific infrastructure, so that the IDE can understand them.
+More precisely, an author of a language has to define:
 \begin{enumerate}
-	\item the abstract syntax (structure of the language), which defines the types of allowed AST nodes,
+	\item the abstract syntax (structure of the language), which determines the types of allowed AST nodes,
 	\item the concrete syntax for editing, i.e. projection of the AST on the screen and interactions with the user, and
-	\item text generation scripts to enable creation of plain text representation used as input for compilers.
+	\item scripts that generate plain text representation of code, to be used as input for compilers.
 \end{enumerate}
 
-IDE tools based on projectional editors are typically used for syntactically rich DSLs.
 In the specific case of MPS, before a program written in a DSL can be executed, it is transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose programming language (GPPL), such as C or Java.
@@
 
-However, very few mainstream GPPLs are now supported by MPS, because it requires substantial effort to manually create a full definition of a language in MPS.
+However, very few GPPLs are now supported by MPS, because it requires substantial effort to manually create a full definition of a language in MPS.
 Only Java and C have been implemented to date.
 The overall goal of our project is to automatize the process of language definition in MPS as much as possible.
-Greater automation would encourage and speed-up the migration of more GPPLs into MPS, thus giving the authors of DSLs more options regarding target general languages.
+Greater automation would encourage and speed-up the migration of more GPPLs into MPS, thus giving the authors of DSLs more options regarding target languages.
 Another possible application is the combination of general-purpose languages with DSLs natively created in MPS.
@@
 In this paper we present \textsc{Ingrid}, a method and a tool for semi-automated construction of a language definition in MPS that uses an ANTLR grammar as a description of its syntax.
-The process of construction is semi-automated for two reasons: (1) ANTLR grammars have to be adjusted before \textsc{Ingrid} can process them, and (2) some aspects of the language definitions in MPS typically must be tweaked or created manually in order to improve languages' usability.
+The process of construction is semi-automated for two reasons: (1) ANTLR grammars have to be adjusted before \textsc{Ingrid} can process them, and (2) some aspects of the language definitions in MPS typically must be tweaked or created manually to improve languages' usability.
+We discuss the practical usability of generated MPS languages, especially in the context of necessary manual adjustments of input ANTLR grammars and specific aspects of MPS languages.
 
-We also discuss the main challenges that we encountered during the development of \textsc{Ingrid} and our solution to them.
+We also identify and discuss the main challenges that we encountered during the development of \textsc{Ingrid} and our solution to them.
 They are related to principal differences between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
@@
 
-\textsc{Ingrid} can also make the construction of DSLs more efficient.
+While the primary use case of the \textsc{Ingrid} method and tool is the import of general-purpose programming languages into MPS, it can also make the construction of DSLs more efficient.
 Our experience shows that, especially for simple languages, it is less time consuming to write their grammar by hand and then create the language definition in MPS with the help of \textsc{Ingrid}, than to create everything manually using the MPS GUI.
 
-We make the following key contributions:
-\begin{itemize}
-	\item The \textsc{Ingrid} method and tool for construction of MPS language definitions from ANTLR grammars, which can be used (1) for the import of general-purpose languages into MPS and (2) for efficient construction of DSLs.
-\textsc{Ingrid} can be adapted also to other language workbenches based on similar principles as MPS.
-	\item Identification and discussion of general challenges related to the differences between (1) language definitions that use grammars and (2) the structured object-oriented approach used in projectional editors.
-	\item Discussion of the practical usability of generated MPS languages and its dependence on (1) the specific form of ANTLR grammars and (2) the manual adjustments of input grammars and specific aspects of MPS languages (e.g., with respect to code layout).
-\end{itemize}
-
+The rest of the paper is organized as follows.
 In the next section, we provide (i) the necessary information about the MPS infrastructure for language definition and (ii) a brief overview of projects with similar goals as \textsc{Ingrid}.
@@
 
-Note also that we omitted many important and interesting details from this paper because of limited space.
-Readers can find the additional details in the full version~\cite{ref:TRFULL}.
+Note also that we omitted many details from this paper because of limited space.
+Readers can find them in the full version~\cite{ref:TRFULL}.

### 2018-05-11 chapters/evaluation.tex
ready

@@
 
-We implemented the \textsc{Ingrid} method as an MPS plugin.
+We implemented \textsc{Ingrid} as an MPS plugin.
 While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to bind with the MPS API, which is used to programatically generate language elements and their aspects.
@@
 
-For the purpose of evaluation, we applied \textsc{Ingrid} to several well-established and widely used languages, including JSON, JavaScript (ECMAScript 5.1~\cite{ref:ECMASCRIPT51}), and C\#.
+For the purpose of evaluation, we applied \textsc{Ingrid} to several widely used languages, including JSON, JavaScript (ECMAScript 5.1~\cite{ref:ECMASCRIPT51}), and C\#.
 MPS projects that contain definitions of all three languages are also released at \url{https://github.com/premun/ingrid}.
-In the rest of this section, we discuss our experience with application of \textsc{Ingrid} to these languages, and then we highlight few general observations.
+In the rest of this section, we discuss our experience with \textsc{Ingrid} on these languages, and then we highlight few general observations.
 
-However, first we must emphasize that MPS languages automatically produced by \textsc{Ingrid}, are not ready-to-use full-fledged MPS counterparts of the original input languages.
+However, first we must emphasize that MPS languages automatically produced by \textsc{Ingrid} are not ready-to-use full-fledged MPS counterparts of the original input languages.
 The structure of a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects have to be manually tweaked (e.g., the Editor) or defined from scratch by the end user.
@@
 \noindent\textbf{JSON.}
-The least amount of manual adjusting after the import into MPS was needed in case of the JSON language~\cite{ref:JSON}, because it is the simplest language from all that we used for our experiments.
+The least amount of manual adjusting after the import into MPS was needed for JSON~\cite{ref:JSON}, because it is the simplest language from all that we used.
 Specifically, the first author spent less than 20 minutes in order to get a language that is ready to use.
@@
 \noindent\textbf{JavaScript.}
-In the case of JavaScript, automated generating of the language definition in MPS from the ANTLR grammar and subsequent manual adjusting was done in less than one hour.
-The current version of \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful --- for example, the language designer has to define actions that improve editing experience by automatically transforming the AST in the background.
+In the case of JavaScript, automated generating of the language definition in MPS and subsequent manual adjusting was done in less than one hour.
+The current version of \textsc{Ingrid} automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful --- for example, the language designer has to define actions that improve editing experience by automatically transforming the AST in the background.
 
@@
 Necessary adjustments include removal of intermediary rules such as \antlrparserrule{statement\_list} : \antlrparserrule{statement*}~\antlrliteral{;} by inlining the right-hand side.
-The Editor aspect was modified afterwards, but only for some of the concepts. 
 Roughly one hour of manual effort was needed in total.
@@
 
-Regarding future changes to the ANTLR grammar of a subject language, it is certainly possible to directly update the corresponding MPS definition, but in some cases it may be easier to run again the whole import process and overwrite the previous definition with a new one.
+Regarding future changes to the ANTLR grammar of a given language, a user can directly update the corresponding MPS definition, but in some cases it may be easier to run again the whole import process and overwrite the previous definition with a new one.

### 2018-05-11 chapters/language_definition.tex
shortening
ready

@@
 The proposed \textsc{Ingrid} method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
-Syntax definition in the form of an ANTLR v4 grammar exists for all widely used programming languages. It is available at \url{https://github.com/antlr/grammars-v4}.
-Unlike some of the related projects, we did not extend the notation with any custom features.
+Syntax definition in the form of an ANTLR v4 grammar exists for all widely used programming languages --- it is available at \url{https://github.com/antlr/grammars-v4}.
+We did not extend the notation with any custom features.
 
@@
 \textsc{Ingrid} is currently able to create (i) a full Structure aspect for each element of the given language, (ii) a very basic Editor, and (iii) a basic TextGen aspect.
-Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the remaining aspects not yet supported by \textsc{Ingrid} must also be defined.
+Therefore, the resulting MPS language has to be adjusted manually to improve its usability, and the remaining aspects not yet supported by \textsc{Ingrid} must also be defined.
 
@@
 \small
-\textbf{grammar} \textit{SimpleXML};
 \antlrparserrule{document}  : \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
@@
 \end{alltt}
-\caption{Grammar of the SimpleXML language in the ANTLR v4 notation}
+\caption{ANTLR v4 grammar of the SimpleXML language}
 \label{fig:SIMPLEXML}
@@
 The grammar of SimpleXML contains elements of the kinds that are listed below.
-Each color in Figure~\ref{fig:SIMPLEXML} corresponds to one kind in a way indicated by the list item headers.
+Each color in Figure~\ref{fig:SIMPLEXML} corresponds to one kind, as indicated by the list item headers.
 \begin{itemize}
@@
 	\item \antlrparserrule{\textbf{Parser rules}} describe the structure of a language.
-	\item \antlrlexerrule{\textbf{Lexer rules}} describe terminal symbols that the parser matches against the input.
-		A terminal symbol can be encoded as a string value or a regular expression.
+	\item \antlrlexerrule{\textbf{Lexer rules}} describe terminal symbols.
+		Any of them can be encoded as a string value or a regular expression.
 	\item \antlrliteralnoap{\textbf{String literals}} also represent terminal symbols, but by exact match to a string constant.
@@
 \end{itemize}
-Elements on the right side of a rule can be annotated with standard EBNF operators (\code{?}, \code{+}, \code{*}) that specify the allowed number of occurrences.
+Elements on the right side of a rule can be annotated with EBNF operators (\code{?}, \code{+}, \code{*}) that specify the allowed number of occurrences.
 
@@
 
-The first phase in the process of MPS language construction is parsing of the input ANTLR v4 grammar.
-It is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
+Parsing of the input ANTLR v4 grammar of a given language is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
 The full parse tree, which comes out of the parser, is quite complex and contains information not relevant for \textsc{Ingrid}.
@@
 The representation of lexer rules (tokens) in the full parse tree has to be simplified too.
-In the ANTLR notation, lexer rules can be built from alternatives just like the parser rules --- see, for example, the lexer rule \antlrlexerrule{Name} from our SimpleXML language (Figure~\ref{fig:SIMPLEXML}).
+In the ANTLR notation, lexer rules can be built from alternatives just like the parser rules --- see, for example, the lexer rule \antlrlexerrule{Name} in Figure~\ref{fig:SIMPLEXML}.
 Every tree that represents a lexer rule is flattened into the equivalent regular expression.
@@
 In the next phase, the complete structure of the MPS language is automatically generated from the AST that represents syntax of the input language.
-Elements of the Structure aspect are derived from the AST nodes, and then linked appropriately.
-Therefore, structure of the MPS language is usually similar to the original ANTLR grammar.
+Elements of the Structure aspect are derived from the AST nodes, and linked appropriately.
+Therefore, structure of the MPS language is similar to the original ANTLR grammar.
 
 When designing the procedure for translating AST nodes (i.e., grammar rules) into the MPS language structure, we faced several challenges.
-The main challenge, as we already mentioned, is that a grammar typically contains rules that do not directly correspond to programming language constructs.
+The main challenge is that a grammar typically contains rules that do not directly correspond to programming language constructs.
 Such rules exist at the intermediate layers of a syntax hierarchy.
-Their main purpose is to enable easier understanding and maintenance of the grammar by humans.
-Nevertheless, presence of the intermediate layers would significantly complicate usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
+Their purpose is to enable easier understanding and maintenance of the grammar by humans.
+Nevertheless, presence of the intermediate layers would significantly complicate usage of the language in MPS, and the layers are actually not necessary for construction of an MPS language.
 We show examples illustrating this problem, which we call a \emph{layer problem}, later in this section.
-As a part of the \textsc{Ingrid} method, we have designed an approach to eliminate the unnecessary layers during construction of the Structure aspect --- we call it the \emph{shortcut} approach.
+As a part of the \textsc{Ingrid} method, we have designed an approach to eliminate the unnecessary layers --- we call it the \emph{shortcut} approach.
 However, before focusing on the intermediate layers, we describe the basic principles of translation from AST into Structure.
 
-In MPS, each concept of the language (i.e., every AST node) is represented by an object that may have a parent, some children, and properties of any data type.
+In MPS, each concept of the language (i.e., every AST node) is represented by an object that may have a parent, some children, and properties.
 Additionally, the object may also contain references to objects representing other AST nodes.
-The parent-child relationships between objects that make the Structure aspect are derived from rules of the grammar.
-If a rule has multiple alternatives, then a distinct object (MPS concept) has to be created in the structure for each alternative.
+The parent-child relationships between objects that make the Structure aspect are derived from the grammar rules.
+If a rule has multiple alternatives, then a distinct object (MPS concept) has to be created for each alternative.
 The name of a concept (object) in MPS is composed from (1) the name of the AST node and (2) the number indicating the position of the respective alternative on the right-hand side of the rule.
@@
 \label{fig:ELEMENTSTRUCT}
-\vspace{-1mm}
+\vspace{-2mm}
 \end{figure}
@@
 References to other parser rules are captured by pointers to child objects.
-The types of child objects, such as \code{IContent} in the \code{Element{\_}1} concept, are determined as follows.
-Consider the \antlrparserrule{content} rule from the SimpleXML language.
+The types of child objects, such as \code{IContent} in \code{Element{\_}1}, are determined as follows.
+Consider the \antlrparserrule{content} rule from SimpleXML.
 An object corresponding to any one of the four alternatives could be the actual value anywhere the \antlrparserrule{content} rule is referenced.
@@
 Our solution is to use interface concepts.
-For each rule with more than one alternative on the right side, first an interface concept is defined in the MPS language structure, and then one object that implements the given interface is created for each alternative.
-The resulting fragment of the language structure looks like the one in Figure~\ref{fig:ICONTENTITF}.
-It contains one interface \mpsinterface{IContent} that is implemented by four object concepts \mpsconcept{Content{\_}1}, $\ldots$, \mpsconcept{Content{\_}4}.
-
-\begin{figure}[ht]
-\vspace{-2mm}
-\centering
-\begin{alltt}
-\small
-\mpsinterface{IContent} : \mpsconcept{Content{\_}1} | \mpsconcept{Content{\_}2} | \mpsconcept{Content{\_}3} | \mpsconcept{Content{\_}4}
-\end{alltt}
-\caption{MPS interface \code{IContent} and the types of implementing objects}
-\label{fig:ICONTENTITF}
-\end{figure}
+For each rule with multiple alternatives on the right side, first an interface concept is defined, and then one object that implements the given interface is created for each alternative.
+The resulting fragment of the language structure contains one interface \mpsinterface{IContent} that is implemented by four object concepts \mpsconcept{Content{\_}1}, $\ldots$, \mpsconcept{Content{\_}4}.
 
 Now we can illustrate the layer problem on the behavior of auto-completion in MPS.
-Suppose that a user just inserted a fresh node of the type \mpsconcept{Element{\_1}}, and would like to insert another nested XML element inside.
+Suppose that a user inserted a fresh node of the type \mpsconcept{Element{\_1}}, and would like to insert another XML element inside.
 The auto-completion mechanism of MPS offers four options that are displayed in the left part of Figure~\ref{fig:LAYERPROBLEM}.
@@
 \end{enumerate}
-The main difficulty here is that, in the first step, the user has to (i) either correctly guess what option offered by the auto-complete menu to select, or (2) to remember the order of alternatives in the grammar rule.
+The difficulty here is that, in the first step, the user either (i) has to correctly guess what option to select in the auto-complete menu, or (2) to remember the order of alternatives in the grammar rule.
 
-Similarly, if the user would like to replace a nested \mpsconcept{Element{\_}1} node with, for example, an XML comment (i.e., the \mpsconcept{Comment} node), then both intermediary layers have to be deleted before she gets back to the original selection among the concepts \mpsconcept{Content{\_}1}, $\ldots$, \mpsconcept{Content{\_}4}.
+Similarly, if the user would like to replace a nested \mpsconcept{Element{\_}1} node with, for example, a \mpsconcept{Comment} node, then both intermediary layers have to be deleted before she gets back to the original selection among the concepts \mpsconcept{Content{\_}1}, $\ldots$, \mpsconcept{Content{\_}4}.
 
-Intermediary layers have no visual appearance, and therefore it is very difficult for users to see what is actually happening and they may get confused very easily.
-The layer problem is addressed by the shortcut approach, which we describe briefly in the following paragraphs.
+Intermediary layers have no visual appearance, and therefore it is very difficult for users to see what is actually happening and they may get confused easily.
+The layer problem is addressed by the shortcut approach, which we describe in the following paragraphs.
 
@@
 
-For the input that consists of a particular AST node $N$ and the grammar rule $R$ that expands $N$, the procedure implementing the shortcut approach systematically traverses the parser tree (AST) built in the earlier phases in order to identify each AST node that (i) may appear in some derivation chain starting by the rule $R$ from the node $N$ and (ii) does not belong to any intermediary layer.
+For the input that consists of an AST node $N$ and the grammar rule $R$ that expands $N$, the procedure implementing the shortcut approach systematically traverses the full tree (AST) in order to identify each AST node that (i) may appear in some derivation chain starting by the rule $R$ from $N$ and (ii) does not belong to any intermediary layer.
 We call them \emph{end nodes}.
-Our algorithm that can recognize intermediary layers is, too, described in the full version~\cite{ref:TRFULL}.
+Our algorithm that can recognize intermediary layers is described in the full version~\cite{ref:TRFULL}.
 
 We use the name \emph{shortcut approach} for this algorithm, because it creates shortcuts from the given rule to end nodes, by the virtue of hiding all intermediary layers.
-
 The primary use case for shortcuts is to generate options for the auto-completion menus, such that only the end nodes are offered.
 Shortcuts have to be considered when nodes are inserted, and also when they are deleted.
-In each case, the whole chain including possibly multiple intermediary AST nodes (up to the end node) must be added, respectively deleted.
-When the user wants to delete some end node from the AST of a program or document written in the MPS language, the effect of an insertion must be fully reversed.
+In each case, the whole chain including possibly multiple intermediary AST nodes must be added, respectively deleted.
 
@@
 The main problem that we had to address is the absence of information about the code layout and whitespace in ANTLR grammars.
-Here we present a solution that is only partially automated for reasons explained below.
+Here we present a solution that is only partially automated.
 
-We observed that the most tedious and error-prone step in the manual definition of the Editor aspect is the creation of cells for all literals (keywords), properties, children, and other fields of a given concept that should appear in its visual representation.
+We observed that the most tedious and error-prone step in the manual definition of the Editor aspect is the creation of cells for all literals (keywords), properties, children, and other fields of a given concept.
 This step can be very easily automated.
 
-Our solution that we implemented in the current version of the \textsc{Ingrid} method is to create all the cells and place them in a single row.
+Our solution that we implemented in the current version of \textsc{Ingrid} is to create all the cells and place them in a single row.
 Further adjustments of the layout, such as indentation and line breaks, can be done very efficiently by the user in the MPS IDE.
@@
 Based on our experience, it takes a very short time to manually adjust the layout into a form much better than any fully automated heuristic could achieve.
-A typical user of the \textsc{Ingrid} method will be able to use the projectional editor in MPS quite efficiently.
 We actually experimented with several heuristics to derive a useful code layout automatically, but all of them produced rather suboptimal results.
@@
 \label{fig:TEXTGENBASIC}
-\vspace{-3mm}
+\vspace{-4mm}
 \end{figure}
@@
 Like in the case of a projectional editor, the main challenge associated with TextGen is to produce valid code with a reasonable layout.
-An implementation of the TextGen aspect must determine properly where to put line breaks, spaces, and indentation.
+An implementation of TextGen must determine properly where to put line breaks, spaces, and indentation.
 For example, in the case of the concept representing an XML element, there must be a space between the element's name and the first attribute, while it is not required between the opening bracket \antlrliteral{\textless} and the name.
@@
 \noindent\textbf{Adjusting grammars.}
-In some cases, a small adjustment of the input grammar before the run of \textsc{Ingrid} might yield a better and more useful MPS language.
+In some cases, a small adjustment of the grammar before the run of \textsc{Ingrid} might yield a better and more useful MPS language.
 Here we illustrate this on a simple example.
@@
 Figure~\ref{fig:XMLATTRIB} shows the definition of an XML attribute, which is taken from the original XML grammar.
-The lexer rule \antlrlexerrule{STRING} actually says that quotes make a part of the attribute's value.
-If the MPS language would faithfully match the grammar, the user would have to always input the leading and trailing quote together with the value in the projectional editor.
+The lexer rule \antlrlexerrule{STRING} says that quotes make a part of the attribute's value.
+If the MPS language would faithfully match the grammar, the user would have to always input both quotes together with the value in the projectional editor.
 
 \begin{figure}[ht]
-\vspace{-1mm}
+\vspace{-2mm}
 \begin{alltt}
@@
 
-In our SimpleXML language, we adjusted the grammar easily in a way that we show in Figure~\ref{fig:XMLADJUST}.
+In SimpleXML, we adjusted the grammar in a way that we show in Figure~\ref{fig:XMLADJUST}.
 We turned quotes into literals, ensuring that they will only appear in the projectional editor as string constant cells.
-
-Additional examples are provided in~\cite{ref:TRFULL}.
 
@@
 \noindent\textbf{Breaking original grammars and parsers.}
-We also want to point out a general problem with grammar adjustments, which all potential users of the \textsc{Ingrid} method should be aware of.
-The original ANTLR grammar for an input language can be very easily changed in a way that, at first, seems harmless and valid inside MPS, but the parser generated out of the adjusted grammar stops accepting the original language.
-In particular, adjustments needed to improve the resulting MPS language can break down the parser.
+We also want to point out a general problem with grammar adjustments, which all potential users of \textsc{Ingrid} should be aware of.
+The original ANTLR grammar for an input language can be easily changed in a way that, at first, seems harmless and valid in MPS, but the parser generated out of the adjusted grammar stops accepting the original language.
 
-This problem has two main causes: (1) low-level implementation details of token matching in the ANTLR parser and (2) usage of ANTLR grammars for a different purpose (that was not expected) in our project.
+This problem has two main causes: (1) low-level implementation details of token matching in the ANTLR parser and (2) usage of ANTLR grammars for a different purpose in our project.
 Since, for practical reasons, it is very important that a parser works for programs written according to the original grammar, adjustments have to be performed carefully by the users of \textsc{Ingrid}.
